---
title: "Detecting departure and destination aerodromes from ADS-B"
subtitle: "Version 5 — why arrivals are hard, and what recovers part of them"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every number comes from a committed CSV under data/, produced by
# benchmarks/{abstained_vertical,vertical_measure,motion_model_compare}.py in
# the opdi repository.
cache <- function(f) {
  p <- file.path("data", f)
  if (!file.exists(p)) return(NULL)
  read.csv(p, stringsAsFactors = FALSE)
}
modes <- cache("mode_comparison.csv")
typ   <- cache("per_type.csv")
sweep <- cache("angle_sweep.csv")
prof  <- cache("motion_profile.csv")
vm    <- cache("vertical_measure.csv")
cmp   <- cache("model_compare.csv")

pct <- function(x) ifelse(is.na(x), "—", sprintf("%.2f%%", 100 * x))
pp  <- function(x) ifelse(is.na(x), "—", sprintf("%+.2f pp", 100 * x))
num <- function(x) format(round(x), big.mark = ",", trim = TRUE)

m <- function(mode, field) {
  v <- modes[[field]][modes$mode == mode]
  if (length(v) == 0) NA_real_ else v[1]
}
N_GT   <- modes$n_ground_truth[1]
OOA_SH <- m("endpoint", "ades_gt_ooa_share")
IN_ARR <- round(N_GT * (1 - OOA_SH))

# The in-area arrival decomposition, from the two modes' in-area figures.
ne_ans <- round(m("nearest",  "ades_inarea_coverage") * IN_ARR)
ne_cor <- round(m("nearest",  "ades_inarea_coverage") * m("nearest",  "ades_inarea_accuracy") * IN_ARR)
ep_ans <- round(m("endpoint", "ades_inarea_coverage") * IN_ARR)
ep_cor <- round(m("endpoint", "ades_inarea_coverage") * m("endpoint", "ades_inarea_accuracy") * IN_ARR)
NO_CAND   <- IN_ARR - ne_ans
ABSTAINED <- ne_ans - ep_ans
WRONG     <- ep_ans - ep_cor
GUESS_OK  <- ne_cor - ep_cor          # nearest's correct answers on the abstained block

# Model comparison, in flights.
C <- function(model, role, field) cmp[[paste0(role, "_", field)]][cmp$model == model]
BASE <- "base (30 NM / 15,000 ft)"
cnt <- function(model, role) {
  ans <- round(C(model, role, "coverage") * N_GT)
  cor <- round(C(model, role, "overall")  * N_GT)
  list(ans = ans, cor = cor, wrong = ans - cor)
}
delta <- function(model, role) {
  b <- cnt(BASE, role); x <- cnt(model, role)
  dc <- x$cor - b$cor; dw <- x$wrong - b$wrong
  list(dc = dc, dw = dw, ratio = if (dw > 0 && dc > 0) dc / dw else NA_real_)
}
BEAR <- "+bearing 7 min / 3 deg"
BEAR30 <- "+bearing 30 min / 1 deg"
RATE <- "+bearing 7 min / 3 deg +rate"
```

## Summary {.unnumbered}

Version 4 established that arrivals are harder than departures on every measure
and recommended not changing the arrivals algorithm. It did not say *why*, and
"we do not know why" is a poor basis for leaving a quarter of the data
unlabelled. This version takes the arrival deficit apart.

**Three findings.**

**The arrival deficit is silence, not error.** Of `r num(IN_ARR)` in-area
arrivals, `r num(WRONG)` are answered wrongly and **`r num(ABSTAINED + NO_CAND)`
are not answered at all**. Version 4's emphasis on misnaming — built on a vivid
case, Bilbao — described `r pct(WRONG / (ABSTAINED + NO_CAND + WRONG))` of the
problem.

**Most of that silence is ours, but it is not cheap to break.** `r num(ABSTAINED)`
flights had a usable candidate that the abstention refused; only `r num(NO_CAND)`
had nothing at all. So the recoverable share is nearly twice the permanently
lost one. But guessing on the refused flights scores
`r pct(GUESS_OK / ABSTAINED)` — four right for eleven wrong. The abstention is
doing its job.

**One new signal beats that, and it is horizontal.** Whether the aircraft's
course *points at* the candidate aerodrome recovers
`r num(delta(BEAR, "ades")$dc)` correct arrivals for
`r num(delta(BEAR, "ades")$dw)` new errors —
`r sprintf("%.2f:1", delta(BEAR, "ades")$ratio)`. Vertical rate recovers
nothing, and adding it to the bearing test makes things worse.

There is a single reason for that asymmetry, and it is the most useful thing in
this paper:

::: {.callout-important}
## Positions survive staleness; rates do not

The OpenSky archive fills every second by carrying the last known position
forward, and in this population roughly half the rows are repeats. A **bearing
between two genuinely measured fixes is correct regardless of when they were
measured**. A **rate divides by an elapsed time the timestamps get wrong** — the
clock says ten minutes, the two real measurements may be ninety seconds apart.

That one sentence explains why bearing alignment works, and why vertical rate,
closing rate and the approach cone of version 4.5 all failed.
:::

## Where the arrival deficit sits {#sec-deficit}

```{r deficit}
t <- typ[typ$role == "arrivals", ]
t <- t[order(-t$n_truth), ]
t$missing <- t$n_truth - t$n_correct
kable(data.frame(
  Class = gsub("_", " ", t$apt_type),
  Flights = num(t$n_truth),
  Coverage = pct(t$coverage),
  Accuracy = pct(t$accuracy),
  Correct = num(t$n_correct),
  Missing = num(t$missing),
  `Share of deficit` = pct(t$missing / sum(t$missing)),
  check.names = FALSE), row.names = FALSE, align = "lrrrrrr",
  caption = sprintf("Arrivals not correctly labelled, by class of the true aerodrome. %s of %s flights over 2025-06-05 to 07.",
                    num(sum(t$missing)), num(N_GT)))
```

Two thirds of the deficit is at large airports and a quarter is out-of-area.
This paper addresses the first; the second remains the largest untouched block
and is left for a successor.

## Silence, not error {#sec-silence}

"Not labelled" hides two situations that need opposite responses: **no candidate
existed**, which is a receiver-coverage fact, and **a candidate existed and the
rule refused it**, which is a choice. `nearest` never abstains, so its coverage
is exactly the rate at which a candidate existed — no new computation is needed
to separate them.

```{r split}
kable(data.frame(
  Outcome = c("correctly labelled", "answered, wrong",
              "candidate existed, abstained", "no candidate anywhere"),
  Flights = num(c(ep_cor, WRONG, ABSTAINED, NO_CAND)),
  Share = pct(c(ep_cor, WRONG, ABSTAINED, NO_CAND) / IN_ARR),
  check.names = FALSE), row.names = FALSE, align = "lrr",
  caption = sprintf("The %s in-area arrivals, decomposed. In-area means the true destination is a real aerodrome, so out-of-area flights are excluded.", num(IN_ARR)))
```

**`r pct(NO_CAND / IN_ARR)` of in-area arrivals have no candidate at all.** That
is the hard ceiling: no endpoint-based method can exceed
`r pct(1 - NO_CAND / IN_ARR)` coverage on this data, whatever rule it uses.

## The abstained block is hard, not neglected {#sec-hard}

The `r num(ABSTAINED)` refused flights look like free coverage. They are not.

`endpoint` answers a subset of what `nearest` answers, and on shared flights
both rank candidates by the same effective distance, so they return identical
answers there. `nearest`'s *extra* correct answers are therefore exactly its
score on the refused block: **`r num(GUESS_OK)` of `r num(ABSTAINED)`, or
`r pct(GUESS_OK / ABSTAINED)`.**

Answering them all would add `r num(GUESS_OK)` correct labels and
`r num(ABSTAINED - GUESS_OK)` wrong ones —
`r sprintf("%.2f:1", GUESS_OK / (ABSTAINED - GUESS_OK))`. Against the exchange
rate version 4.5 argued for, that is not close. **"Relax the thresholds for
arrivals" is now priced, and the price is bad.**

Why are they hard? Because of what an arrival endpoint is. The state-vector
decimation study measured that an arrival's last fix carries a position
**measured a median of 190 s earlier** — at approach speed, roughly eight
nautical miles back up the approach. The endpoint is not where the flight ended.
Position alone cannot identify an aerodrome from eight miles out on a busy
terminal area, and the abstention correctly declines to try.

## Bearing alignment {#sec-bearing}

If position is exhausted, motion might not be. The measure is the angle between
**the direction the aircraft actually travelled** over a window and **the
direction of the candidate aerodrome**, seen from the far end of that window.
Zero degrees means the track runs straight at it.

This involves no extrapolated position — it is an angle between two observed
bearings — so the documented failures of trajectory extrapolation in version 2
and the 2025 PRC report do not bind it.

```{r sweep-matrix}
s <- sweep[sweep$side == "arrival", ]
A <- c(1, 2, 3, 5, 10, 20, 30)
W <- sort(unique(s$window_min))
mat <- do.call(rbind, lapply(W, function(w) {
  data.frame(Window = paste0(w, " min"),
             setNames(lapply(A, function(a) {
               v <- s$ratio[s$window_min == w & s$angle_deg == a]
               if (length(v) == 0 || is.na(v)) "—" else sprintf("%.2f", v)
             }), paste0(A, "°")),
             check.names = FALSE)
}))
kable(mat, row.names = FALSE, align = "lrrrrrrr",
      caption = "Correct labels per error on the abstained arrival block, by window and angle gate. The block's own baseline is 0.37. Note that these are computed on in-area truth only, and so overstate the population-level trade -- see the next section.")
```

Three things are visible and all three make physical sense.

**Short windows carry no information.** At 2–3 minutes every cell sits near the
0.37 baseline. Over two minutes the aircraft's displacement is small against its
position error, so the bearing is noise.

**It switches on at 5 minutes and peaks at 7.** A metric that were an artefact
would not have an interior optimum with noise on both sides.

**The gate should tighten as the window lengthens.** At loose gates short
windows win, because over half an hour everything points roughly everywhere. At
1° that reverses — a long baseline gives a more precise bearing, so a tight gate
becomes meaningful rather than noise.

**Accuracy plateaus at 70–76%** however tight the gate. Roughly one arrival in
four that points straight at an aerodrome lands somewhere else. Bearing alone
has a ceiling, and tuning it further is close to exhausted.

## The comparison that matters {#sec-compare}

The sweep above scores the abstained block with out-of-area truth excluded.
That flatters it: rescuing a flight whose true destination is outside the
observed area produces a wrong answer the block analysis never counted. The
honest test is the whole population.

```{r compare}
mk <- function(role) do.call(rbind, lapply(cmp$model, function(md) {
  x <- cnt(md, role); d <- delta(md, role)
  data.frame(Model = md,
             Coverage = pct(C(md, role, "coverage")),
             Accuracy = pct(C(md, role, "accuracy")),
             Overall  = pct(C(md, role, "overall")),
             Correct = num(x$cor), Wrong = num(x$wrong),
             `d correct` = if (md == BASE) "—" else sprintf("%+s", num(d$dc)),
             `d wrong`   = if (md == BASE) "—" else sprintf("%+s", num(d$dw)),
             Ratio = if (md == BASE) "—" else
               if (is.na(d$ratio)) "worse" else sprintf("%.2f:1", d$ratio),
             check.names = FALSE)
}))
kable(mk("ades"), row.names = FALSE, align = "lrrrrrrrr",
      caption = sprintf("Arrivals, all %s ground-truth flights. Ratio is extra correct labels per extra wrong label against the base rule.", num(N_GT)))
kable(mk("adep"), row.names = FALSE, align = "lrrrrrrrr",
      caption = "Departures, the same models. Bearing is actively harmful here.")
```

**Arrivals gain.** `r BEAR` adds `r num(delta(BEAR, "ades")$dc)` correct labels
for `r num(delta(BEAR, "ades")$dw)` errors,
`r sprintf("%.2f:1", delta(BEAR, "ades")$ratio)`. The wider
`r BEAR30` adds `r num(delta(BEAR30, "ades")$dc)` for
`r num(delta(BEAR30, "ades")$dw)`,
`r sprintf("%.2f:1", delta(BEAR30, "ades")$ratio)` — more volume, worse rate.

**Departures lose.** The same test adds `r num(delta(BEAR, "adep")$dc)` correct
departures for `r num(delta(BEAR, "adep")$dw)` errors,
`r sprintf("%.2f:1", delta(BEAR, "adep")$ratio)`. Consistent with everything
this study has found: departures are not short of evidence, and adding evidence
to a rule that is not short of it costs accuracy. **Bearing should not be
applied to departures.**

::: {.callout-warning}
## The block-level ratios were optimistic, and this is why

The abstained-block sweep reports `r sprintf("%.2f:1", sweep$ratio[sweep$side=="arrival" & sweep$window_min==7 & sweep$angle_deg==3])`
at 7 min / 3°. The same rule on the full population gives
`r sprintf("%.2f:1", delta(BEAR, "ades")$ratio)`. The difference is entirely the
out-of-area flights the block analysis filtered out and the population cannot.

The lesson generalises: a rule evaluated on the subset it was designed for will
beat its own field performance, and by a wide margin. Every ratio quoted in
@sec-bearing should be read as an upper bound.
:::

## Why the vertical channel is dead {#sec-vertical}

Vertical rate is the obvious second signal, and it does not work. Two
independent measures of the same quantity — the broadcast `vert_rate`, and an
ordinary least squares fit of altitude on time across the window — agree in sign
at close to chance.

```{r vertical}
v <- vm[!duplicated(vm$window_min), ]
kable(data.frame(
  Window = paste0(v$window_min, " min"),
  `vert_rate vs OLS` = pct(v$overall_vr_vs_ols),
  `2-point vs OLS` = pct(v$overall_t2pt_vs_ols),
  `rows with vert_rate = 0` = pct(v$vr_zero_share),
  `rows fresher than 5 s` = pct(v$fresh_share),
  check.names = FALSE), row.names = FALSE, align = "lrrrr",
  caption = "Sign agreement between vertical measures on the abstained arrival block. 50% is chance.")
```

```{r vertical-bands}
b <- vm[vm$window_min == 20, ]
kable(data.frame(
  `Window freshness` = substring(b$freshness, 4),
  Flights = num(b$flights),
  `vert_rate vs OLS` = pct(b$vr_vs_ols),
  `2-point vs OLS` = pct(b$t2pt_vs_ols),
  check.names = FALSE), row.names = FALSE, align = "lrrr",
  caption = "The same agreement at 20 minutes, banded by the share of rows carrying a position fresher than 5 s. Agreement tracks freshness, which identifies the cause rather than inferring it.")
```

Agreement rises with freshness, and `r pct(v$vr_zero_share[1])` of rows carry a
vertical rate of exactly zero — the signature of a value held over rather than
measured. **There is no vertical signal to extract here in either form.**

The measured consequence is in the comparison table: adding the vertical test to
the bearing test cuts the arrival gain from `r num(delta(BEAR, "ades")$dc)`
correct to `r num(delta(RATE, "ades")$dc)`, and the ratio from
`r sprintf("%.2f:1", delta(BEAR, "ades")$ratio)` to
`r sprintf("%.2f:1", delta(RATE, "ades")$ratio)`. It does not merely fail to
help — **it discards flights the bearing test correctly rescued.**

## Recommendation {#sec-recommendation}

**Do not adopt the bearing test yet.** At
`r sprintf("%.2f:1", delta(BEAR, "ades")$ratio)` for arrivals it sits below the
2:1 exchange rate version 4.5 argued for, on the reasoning that a wrong
aerodrome corrupts two movement counts while a silence corrupts one and is
visible. It is the closest anything has come, and the first signal to beat the
abstained block at all, but it is not over the line.

**Do not apply it to departures under any circumstances.**
`r sprintf("%.2f:1", delta(BEAR, "adep")$ratio)` is a clear loss.

**Do not pursue vertical rate on this data.** The measurement is not marginal;
the channel is noise.

**Refine the bearing estimate before re-testing.** The current course is a
two-point bearing across the window. Fitting it by regression, and gating on the
stability of that fit rather than on window length, should raise the ratio — the
1° gate is currently winning despite a noisy input, not because of one. If that
brings arrivals above 2:1, the case changes.

**And do out-of-area first regardless.** It is `r pct(t$missing[t$apt_type == "out of area"] / sum(t$missing))`
of the arrival deficit against the `r pct(GUESS_OK / sum(t$missing))` this
recovers, and unlike the abstained block nobody has shown it is hard.

## What is now ruled out {#sec-ruled-out}

Recorded so it is not re-attempted:

* **Relaxing the abstention thresholds** — priced at
  `r sprintf("%.2f:1", GUESS_OK / (ABSTAINED - GUESS_OK))` (@sec-hard).
* **The approach cone** — version 4.5 measured it against the rectangle at
  matched coverage; departures lost up to 2.55 pp of accuracy, arrivals were a
  wash.
* **Bucket decimation** — the state-vector study measured +55 correct arrivals
  per day for a change that rewrites `track_id` for every published row.
* **Vertical rate, in any form** (@sec-vertical).
* **Trajectory extrapolation** — version 2 and the 2025 PRC report,
  independently.

## Limitations {#sec-limitations}

* **One sample.** Three days, 2025-06-05 to 07. The operating points were chosen
  as the argmax over 120 grid cells of that same sample, so the ratios are
  optimistic. The 2° and 3° columns peak cleanly at 7 minutes, which is why the
  *region* is credible; the specific numbers are not validated.
* **Two measurement bugs were found in this work**, both by sanity checks rather
  than by anything failing: `angle_between` used `%`, and Spark's remainder
  keeps the sign of the dividend, so bearings 20° apart measured 340° whenever
  the pair straddled north — the tell was a median alignment of 186° on a
  0–180° scale. And the altitude slope divided an altitude change by an elapsed
  time drawn from different rows, because `min_by` skips nulls where `min` does
  not. Both are fixed; both had already produced numbers that were reported
  before being caught.
* **The `>=90% fresh` band reverses the freshness trend** for the 2-point
  measure, on about a thousand flights. Either small-sample noise or something
  real about very short tracks. Not resolved.
* **`nearest` and `endpoint` are assumed to answer identically where both
  answer**, which underpins @sec-hard. It follows from both ranking on the same
  effective distance, but it is an argument, not a measurement.

## Reproducing {#sec-reproducing}

```bash
# the abstained block, its motion profile, and the angle x window sweep
python benchmarks/abstained_vertical.py --results-dir <dir>

# which vertical measure to trust, banded by freshness
python benchmarks/vertical_measure.py --results-dir <dir>

# base against the motion variants, whole population
python benchmarks/motion_model_compare.py --results-dir <dir>
```

Rendering this page needs no credentials, no database and no cluster.
