---
title: "Detecting departure and destination aerodromes from ADS-B"
subtitle: "Version 5.1 — why arrivals are hard, and what recovers part of them"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every number comes from a committed CSV under data/, produced by
# benchmarks/{abstained_vertical,vertical_measure,motion_model_compare}.py in
# the opdi repository.
cache <- function(f) {
  p <- file.path("data", f)
  if (!file.exists(p)) return(NULL)
  read.csv(p, stringsAsFactors = FALSE)
}
modes <- cache("mode_comparison.csv")
typ   <- cache("per_type.csv")
sweep <- cache("angle_sweep.csv")
prof  <- cache("motion_profile.csv")
vm    <- cache("vertical_measure.csv")
cmp   <- cache("model_compare.csv")
whole <- cache("whole_sample.csv")

# The whole-sample experiment, scored the same way.
ws <- function(model, role, field) whole[[paste0(role, "_", field)]][whole$model == model]
wcnt <- function(model, role) {
  ans <- round(ws(model, role, "coverage") * N_GT)
  cor <- round(ws(model, role, "overall")  * N_GT)
  list(ans = ans, cor = cor, wrong = ans - cor)
}
wdelta <- function(model, role) {
  b <- wcnt("base", role); x <- wcnt(model, role)
  list(dc = x$cor - b$cor, dw = x$wrong - b$wrong)
}

pct <- function(x) ifelse(is.na(x), "—", sprintf("%.2f%%", 100 * x))
pp  <- function(x) ifelse(is.na(x), "—", sprintf("%+.2f pp", 100 * x))
num <- function(x) format(round(x), big.mark = ",", trim = TRUE)

m <- function(mode, field) {
  v <- modes[[field]][modes$mode == mode]
  if (length(v) == 0) NA_real_ else v[1]
}
N_GT   <- modes$n_ground_truth[1]
OOA_SH <- m("endpoint", "ades_gt_ooa_share")
IN_ARR <- round(N_GT * (1 - OOA_SH))

# The in-area arrival decomposition, from the two modes' in-area figures.
ne_ans <- round(m("nearest",  "ades_inarea_coverage") * IN_ARR)
ne_cor <- round(m("nearest",  "ades_inarea_coverage") * m("nearest",  "ades_inarea_accuracy") * IN_ARR)
ep_ans <- round(m("endpoint", "ades_inarea_coverage") * IN_ARR)
ep_cor <- round(m("endpoint", "ades_inarea_coverage") * m("endpoint", "ades_inarea_accuracy") * IN_ARR)
NO_CAND   <- IN_ARR - ne_ans
ABSTAINED <- ne_ans - ep_ans
WRONG     <- ep_ans - ep_cor
GUESS_OK  <- ne_cor - ep_cor          # nearest's correct answers on the abstained block

# Model comparison, in flights.
C <- function(model, role, field) cmp[[paste0(role, "_", field)]][cmp$model == model]
BASE <- "base (30 NM / 15,000 ft)"
cnt <- function(model, role) {
  ans <- round(C(model, role, "coverage") * N_GT)
  cor <- round(C(model, role, "overall")  * N_GT)
  list(ans = ans, cor = cor, wrong = ans - cor)
}
delta <- function(model, role) {
  b <- cnt(BASE, role); x <- cnt(model, role)
  dc <- x$cor - b$cor; dw <- x$wrong - b$wrong
  list(dc = dc, dw = dw, ratio = if (dw > 0 && dc > 0) dc / dw else NA_real_)
}
BEAR   <- "+bearing 7 min / 0.10 deg"     # the recommended gate
BEAR1  <- "+bearing 7 min / 1 deg"
BEAR3  <- "+bearing 7 min / 3 deg"
BEAR30 <- "+bearing 30 min / 1 deg"
RATE   <- "+bearing 7 min / 1 deg +rate"
# Net value of a gate when an error is priced at `k` correct labels -- the
# exchange-rate reasoning of version 4.5, applied to choose among gates.
net <- function(model, role, k = 2) {
  d <- delta(model, role); d$dc - k * d$dw
}
```

## Summary {.unnumbered}

Version 4 established that arrivals are harder than departures on every measure
and recommended not changing the arrivals algorithm. It did not say *why*, and
"we do not know why" is a poor basis for leaving a quarter of the data
unlabelled. This version takes the arrival deficit apart.

**Three findings.**

**The arrival deficit is silence, not error.** Of `r num(IN_ARR)` in-area
arrivals, `r num(WRONG)` are answered wrongly and **`r num(ABSTAINED + NO_CAND)`
are not answered at all**. Version 4's emphasis on misnaming — built on a vivid
case, Bilbao — described `r pct(WRONG / (ABSTAINED + NO_CAND + WRONG))` of the
problem.

**Most of that silence is ours, but it is not cheap to break.** `r num(ABSTAINED)`
flights had a usable candidate that the abstention refused; only `r num(NO_CAND)`
had nothing at all. So the recoverable share is nearly twice the permanently
lost one. But guessing on the refused flights scores
`r pct(GUESS_OK / ABSTAINED)` — four right for eleven wrong. The abstention is
doing its job.

**One new signal beats that, and it is horizontal.** Whether the aircraft's
course *points at* the candidate aerodrome, measured tightly enough, recovers
`r num(delta(BEAR, "ades")$dc)` correct arrivals for only
`r num(delta(BEAR, "ades")$dw)` new errors —
**`r sprintf("%.2f:1", delta(BEAR, "ades")$ratio)`**, comfortably past the 2:1
exchange rate version 4.5 argued for. This is the first change in the study to
clear that bar. It applies to arrivals only: on departures the same test loses
at `r sprintf("%.2f:1", delta(BEAR, "adep")$ratio)`. Vertical rate recovers
nothing, and adding it to the bearing test destroys most of the gain.

There is a single reason for that asymmetry, and it is the most useful thing in
this paper:

::: {.callout-important}
## Positions survive staleness; rates do not

The OpenSky archive fills every second by carrying the last known position
forward, and in this population roughly half the rows are repeats. A **bearing
between two genuinely measured fixes is correct regardless of when they were
measured**. A **rate divides by an elapsed time the timestamps get wrong** — the
clock says ten minutes, the two real measurements may be ninety seconds apart.

That one sentence explains why bearing alignment works, and why vertical rate,
closing rate and the approach cone of version 4.5 all failed.
:::

## Where the arrival deficit sits {#sec-deficit}

```{r deficit}
t <- typ[typ$role == "arrivals", ]
t <- t[order(-t$n_truth), ]
t$missing <- t$n_truth - t$n_correct
kable(data.frame(
  Class = gsub("_", " ", t$apt_type),
  Flights = num(t$n_truth),
  Coverage = pct(t$coverage),
  Accuracy = pct(t$accuracy),
  Correct = num(t$n_correct),
  Missing = num(t$missing),
  `Share of deficit` = pct(t$missing / sum(t$missing)),
  check.names = FALSE), row.names = FALSE, align = "lrrrrrr",
  caption = sprintf("Arrivals not correctly labelled, by class of the true aerodrome. %s of %s flights over 2025-06-05 to 07.",
                    num(sum(t$missing)), num(N_GT)))
```

Two thirds of the deficit is at large airports and a quarter is out-of-area.
This paper addresses the first; the second remains the largest untouched block
and is left for a successor.

## Silence, not error {#sec-silence}

"Not labelled" hides two situations that need opposite responses: **no candidate
existed**, which is a receiver-coverage fact, and **a candidate existed and the
rule refused it**, which is a choice. `nearest` never abstains, so its coverage
is exactly the rate at which a candidate existed — no new computation is needed
to separate them.

```{r split}
kable(data.frame(
  Outcome = c("correctly labelled", "answered, wrong",
              "candidate existed, abstained", "no candidate anywhere"),
  Flights = num(c(ep_cor, WRONG, ABSTAINED, NO_CAND)),
  Share = pct(c(ep_cor, WRONG, ABSTAINED, NO_CAND) / IN_ARR),
  check.names = FALSE), row.names = FALSE, align = "lrr",
  caption = sprintf("The %s in-area arrivals, decomposed. In-area means the true destination is a real aerodrome, so out-of-area flights are excluded.", num(IN_ARR)))
```

**`r pct(NO_CAND / IN_ARR)` of in-area arrivals have no candidate at all.** That
is the hard ceiling: no endpoint-based method can exceed
`r pct(1 - NO_CAND / IN_ARR)` coverage on this data, whatever rule it uses.

## The abstained block is hard, not neglected {#sec-hard}

The `r num(ABSTAINED)` refused flights look like free coverage. They are not.

`endpoint` answers a subset of what `nearest` answers, and on shared flights
both rank candidates by the same effective distance, so they return identical
answers there. `nearest`'s *extra* correct answers are therefore exactly its
score on the refused block: **`r num(GUESS_OK)` of `r num(ABSTAINED)`, or
`r pct(GUESS_OK / ABSTAINED)`.**

Answering them all would add `r num(GUESS_OK)` correct labels and
`r num(ABSTAINED - GUESS_OK)` wrong ones —
`r sprintf("%.2f:1", GUESS_OK / (ABSTAINED - GUESS_OK))`. Against the exchange
rate version 4.5 argued for, that is not close. **"Relax the thresholds for
arrivals" is now priced, and the price is bad.**

Why are they hard? Because of what an arrival endpoint is. The state-vector
decimation study measured that an arrival's last fix carries a position
**measured a median of 190 s earlier** — at approach speed, roughly eight
nautical miles back up the approach. The endpoint is not where the flight ended.
Position alone cannot identify an aerodrome from eight miles out on a busy
terminal area, and the abstention correctly declines to try.

## Bearing alignment {#sec-bearing}

If position is exhausted, motion might not be. The measure is the angle between
**the direction the aircraft actually travelled** over a window and **the
direction of the candidate aerodrome**, seen from the far end of that window.
Zero degrees means the track runs straight at it.

This involves no extrapolated position — it is an angle between two observed
bearings — so the documented failures of trajectory extrapolation in version 2
and the 2025 PRC report do not bind it.

::: {.callout-note collapse="true"}
## Why departures need no 180-degree correction

A departing aircraft flies *away* from its aerodrome, so it looks as though the
measure must be inverted for departures. It must not, because the origin of the
angle already flips with the role.

The window is measured from the arrival's **last** fix and the departure's
**first**, so its far end is earlier in time for an arrival and later for a
departure. The bearing from that far end toward the near end is therefore the
course in one case and the reverse course in the other. The question becomes
*"is the aerodrome in the direction I am heading"* for an arrival and *"is the
aerodrome in the direction I came from"* for a departure, and both read zero for
a correct match.

The distinction matters because the two vertical and horizontal *rates* in the
same analysis do carry an explicit per-role sign — they are signed along the
time axis, and an angle between two bearings from a common origin is not.
Confirmed empirically: correctly identified departures sit at a median of 2–9
degrees, not 171–178.

It also explains part of why the measure is weak for departures. By 30 minutes
down-track the separation is 6 degrees against 2 — from far enough away, the
true origin aerodrome and its neighbours all lie in much the same direction, so
the angle stops discriminating. An arrival endpoint sits a median of eight
nautical miles short of the runway, where the direction along the track is still
informative.
:::

```{r sweep-matrix}
s <- sweep[sweep$side == "arrival", ]
# Local names, deliberately not single capitals: chunks share one
# environment, and a bare W here once clobbered an accessor defined in setup.
angles_sw <- c(1, 2, 3, 5, 10, 20, 30)
wins_sw <- sort(unique(s$window_min))
mat <- do.call(rbind, lapply(wins_sw, function(w) {
  data.frame(Window = paste0(w, " min"),
             setNames(lapply(angles_sw, function(a) {
               v <- s$ratio[s$window_min == w & s$angle_deg == a]
               if (length(v) == 0 || is.na(v)) "—" else sprintf("%.2f", v)
             }), paste0(angles_sw, "°")),
             check.names = FALSE)
}))
kable(mat, row.names = FALSE, align = "lrrrrrrr",
      caption = "Correct labels per error on the abstained arrival block, by window and angle gate. The block's own baseline is 0.37. Note that these are computed on in-area truth only, and so overstate the population-level trade -- see the next section.")
```

Three things are visible and all three make physical sense.

**Short windows carry no information.** At 2–3 minutes every cell sits near the
0.37 baseline. Over two minutes the aircraft's displacement is small against its
position error, so the bearing is noise.

**It switches on at 5 minutes and peaks at 7.** A metric that were an artefact
would not have an interior optimum with noise on both sides.

**The gate should tighten as the window lengthens.** At loose gates short
windows win, because over half an hour everything points roughly everywhere. At
1° that reverses — a long baseline gives a more precise bearing, so a tight gate
becomes meaningful rather than noise.

**Accuracy plateaus at 70–76%** however tight the gate. Roughly one arrival in
four that points straight at an aerodrome lands somewhere else. Bearing alone
has a ceiling, and tuning it further is close to exhausted.

## The comparison that matters {#sec-compare}

The sweep above scores the abstained block with out-of-area truth excluded.
That flatters it: rescuing a flight whose true destination is outside the
observed area produces a wrong answer the block analysis never counted. The
honest test is the whole population.

```{r compare}
mk <- function(role) do.call(rbind, lapply(cmp$model, function(md) {
  x <- cnt(md, role); d <- delta(md, role)
  data.frame(Model = sub("^\\+bearing ", "", md),
             Coverage = pct(C(md, role, "coverage")),
             Accuracy = pct(C(md, role, "accuracy")),
             Overall  = pct(C(md, role, "overall")),
             `d correct` = if (md == BASE) "—" else sprintf("%+s", num(d$dc)),
             `d wrong`   = if (md == BASE) "—" else sprintf("%+s", num(d$dw)),
             Ratio = if (md == BASE) "—" else
               if (is.na(d$ratio)) "worse" else sprintf("%.2f:1", d$ratio),
             `Net @2:1` = if (md == BASE) "—" else sprintf("%+s", num(net(md, role))),
             check.names = FALSE)
}))
kable(mk("ades"), row.names = FALSE, align = "lrrrrrrr",
      caption = sprintf("Arrivals, all %s ground-truth flights. *Net @2:1* prices each new error at two correct labels, the exchange rate of version 4.5; positive means the gate is worth adopting on that reasoning.", num(N_GT)))
kable(mk("adep"), row.names = FALSE, align = "lrrrrrrr",
      caption = "Departures, the same models. Every gate is a loss.")
```

Because the bearing test only ever *adds* answers — a flight that already passed
the abstention is untouched, and the aerodrome named is the same candidate
either way — every newly answered flight falls in exactly one of the two delta
columns. Their sum is the number of flights the gate rescued, and

$$\text{accuracy on the rescued} = \frac{\Delta\text{correct}}{\Delta\text{correct} + \Delta\text{wrong}}$$

is how often the gate is right *when it overrides the abstention*. It is not the
model's accuracy, which is swamped by the flights the change never touched: at
3° the gate is right `r pct(delta(BEAR3,"ades")$dc / (delta(BEAR3,"ades")$dc + delta(BEAR3,"ades")$dw))`
of the time on the `r num(delta(BEAR3,"ades")$dc + delta(BEAR3,"ades")$dw)` flights
it rescues, while the model's accuracy reads `r pct(C(BEAR3,"ades","accuracy"))`.
The benchmark to hold it against is `r pct(GUESS_OK / ABSTAINED)` —
what guessing blind on that population scores (@sec-hard).

**The gate must be tight, and tighter than seems physical.** The ratio climbs
monotonically as the gate narrows — `r sprintf("%.2f:1", delta(BEAR3,"ades")$ratio)`
at 3°, `r sprintf("%.2f:1", delta(BEAR1,"ades")$ratio)` at 1°,
`r sprintf("%.2f:1", delta(BEAR,"ades")$ratio)` at 0.1° — and accuracy on the
rescued flights rises with it, from
`r pct(delta(BEAR3,"ades")$dc / (delta(BEAR3,"ades")$dc + delta(BEAR3,"ades")$dw))`
to `r pct(delta(BEAR,"ades")$dc / (delta(BEAR,"ades")$dc + delta(BEAR,"ades")$dw))`.
Under 2:1 pricing the best gate is the tightest tested.

::: {.callout-note}
## A prediction this refuted

Before running the sub-degree gates we argued they could not help. A runway
complex spans 2–4 km, which from 40 NM subtends 2–3°, so a gate below a degree
is tighter than the target: it should stop selecting for "heading to this
airport" and start selecting for "aimed at the arbitrary reference coordinate
OurAirports publishes". Accuracy was predicted to plateau and counts to collapse
into noise.

Accuracy did not plateau. It rose from
`r pct(delta(BEAR3,"ades")$dc / (delta(BEAR3,"ades")$dc + delta(BEAR3,"ades")$dw))`
at 3° to
`r pct(delta(BEAR,"ades")$dc / (delta(BEAR,"ades")$dc + delta(BEAR,"ades")$dw))`
at 0.1°, monotonically, on `r num(delta(BEAR,"ades")$dc + delta(BEAR,"ades")$dw)`
flights — not a handful.

The reasoning was wrong about what the gate selects. It is not resolving which
runway an aircraft is aimed at; it is separating tracks *deliberately* pointed
at a field, over a long straight leg, from tracks that happen to line up by
coincidence. Coincidental alignment is common at 3° and rare at 0.1°, and that
is the whole of the discrimination.
:::

**Volume and rate trade against each other, and the rate wins.** The widest
gate tested, `r BEAR30`, recovers `r num(delta(BEAR30, "ades")$dc)` correct
labels — five times the tightest — but at
`r sprintf("%.2f:1", delta(BEAR30, "ades")$ratio)` it costs
`r num(-net(BEAR30, "ades"))` on the 2:1 reckoning. Chasing volume here is what
the exchange rate exists to prevent.

**Departures lose.** The same test adds `r num(delta(BEAR, "adep")$dc)` correct
departures for `r num(delta(BEAR, "adep")$dw)` errors,
`r sprintf("%.2f:1", delta(BEAR, "adep")$ratio)`. Consistent with everything
this study has found: departures are not short of evidence, and adding evidence
to a rule that is not short of it costs accuracy. **Bearing should not be
applied to departures.**

::: {.callout-warning}
## The block-level ratios were optimistic, and this is why

The abstained-block sweep reports `r sprintf("%.2f:1", sweep$ratio[sweep$side=="arrival" & sweep$window_min==7 & sweep$angle_deg==3])`
at 7 min / 3°. The same rule on the full population gives
`r sprintf("%.2f:1", delta(BEAR3, "ades")$ratio)` — the same gate, the same
rule, a third less. The difference is entirely the out-of-area flights the block
analysis filtered out and the population cannot.

The lesson generalises: a rule evaluated on the subset it was designed for will
beat its own field performance, and by a wide margin. Every ratio quoted in
@sec-bearing should be read as an upper bound.
:::

## Why the vertical channel is dead {#sec-vertical}

Vertical rate is the obvious second signal, and it does not work. Two
independent measures of the same quantity — the broadcast `vert_rate`, and an
ordinary least squares fit of altitude on time across the window — agree in sign
at close to chance.

```{r vertical}
v <- vm[!duplicated(vm$window_min), ]
kable(data.frame(
  Window = paste0(v$window_min, " min"),
  `vert_rate vs OLS` = pct(v$overall_vr_vs_ols),
  `2-point vs OLS` = pct(v$overall_t2pt_vs_ols),
  `rows with vert_rate = 0` = pct(v$vr_zero_share),
  `rows fresher than 5 s` = pct(v$fresh_share),
  check.names = FALSE), row.names = FALSE, align = "lrrrr",
  caption = "Sign agreement between vertical measures on the abstained arrival block. 50% is chance.")
```

```{r vertical-bands}
b <- vm[vm$window_min == 20, ]
kable(data.frame(
  `Window freshness` = substring(b$freshness, 4),
  Flights = num(b$flights),
  `vert_rate vs OLS` = pct(b$vr_vs_ols),
  `2-point vs OLS` = pct(b$t2pt_vs_ols),
  check.names = FALSE), row.names = FALSE, align = "lrrr",
  caption = "The same agreement at 20 minutes, banded by the share of rows carrying a position fresher than 5 s. Agreement tracks freshness, which identifies the cause rather than inferring it.")
```

Agreement rises with freshness, and `r pct(v$vr_zero_share[1])` of rows carry a
vertical rate of exactly zero — the signature of a value held over rather than
measured. **There is no vertical signal to extract here in either form.**

The measured consequence is in the comparison table: adding the vertical test to
the bearing test cuts the arrival gain from `r num(delta(BEAR, "ades")$dc)`
correct to `r num(delta(RATE, "ades")$dc)`, and the ratio from
`r sprintf("%.2f:1", delta(BEAR, "ades")$ratio)` to
`r sprintf("%.2f:1", delta(RATE, "ades")$ratio)`. It does not merely fail to
help — **it discards flights the bearing test correctly rescued.**

## Recommendation {#sec-recommendation}

**Adopt the bearing test for arrivals, at a tight gate.** `r BEAR` recovers
`r num(delta(BEAR, "ades")$dc)` correct arrival labels for
`r num(delta(BEAR, "ades")$dw)` new errors —
`r sprintf("%.2f:1", delta(BEAR, "ades")$ratio)`, against the 2:1 bar. Arrival
coverage rises `r pp(C(BEAR,"ades","coverage") - C(BASE,"ades","coverage"))`
and overall correctness
`r pp(C(BEAR,"ades","overall") - C(BASE,"ades","overall"))`, for
`r pp(C(BEAR,"ades","accuracy") - C(BASE,"ades","accuracy"))` of accuracy.

This is the first change this study has recommended adopting since version 4
moved detection into the pipeline, and the first to clear the exchange rate on
its own terms rather than by argument.

**Do not apply it to departures.** Every gate tested is a loss, the best at
`r sprintf("%.2f:1", max(sapply(cmp$model[grepl("bearing", cmp$model) & !grepl("rate", cmp$model)], function(m) delta(m,"adep")$ratio), na.rm=TRUE))`.
Departures are not short of evidence, and adding evidence to a rule that is not
short of it costs accuracy.

**Do not pursue vertical rate on this data.** Adding it to the bearing test cuts
the arrival gain from `r num(delta(BEAR1, "ades")$dc)` correct to
`r num(delta(RATE, "ades")$dc)` and the ratio from
`r sprintf("%.2f:1", delta(BEAR1, "ades")$ratio)` to
`r sprintf("%.2f:1", delta(RATE, "ades")$ratio)`. It discards flights the
bearing test correctly rescued.

**Validate on a second period before publishing the change.** Everything here is
one three-day sample, and the gate was chosen from that sample.
`r pct(1 - NO_CAND / IN_ARR)` remains the ceiling regardless.

**And out-of-area is still the larger prize.** It is
`r pct(t$missing[t$apt_type == "out of area"] / sum(t$missing))` of the arrival
deficit against the `r pct(delta(BEAR, "ades")$dc / sum(t$missing))` this
recovers. The bearing test is worth having; it is not the main event.

## Bearing on the whole sample {#sec-whole}

Used as a rescue, the bearing test can only *add* answers. That is a structural
limit rather than a property of the data: a flight the abstention already
accepts is untouched, so the
`r num(round(C(BASE,"ades","coverage")*N_GT) - round(C(BASE,"ades","overall")*N_GT))`
arrivals the base rule gets **wrong** were unreachable by every variant in
@sec-compare. No rescue can raise accuracy.

Applying the test to every flight opens three questions the rescue framing
cannot ask.

* **Veto** — require alignment on flights the gate accepts, so bearing can
  remove answers. The only variant that can raise accuracy.
* **Replace** — alignment instead of the distance and height gate, testing
  whether that box carries anything the course does not.
* **Rerank** — choose *which* aerodrome by alignment rather than by effective
  distance. Version 4.5 argued the abstention is a gate and never a naming
  rule, and that arrivals fail by misnaming; this is the naming rule that
  argument implies.

```{r whole-table}
sel <- c("base",
         "rescue: gate OR align<=0.1", "rescue: gate OR align<=3.0",
         "veto: gate AND align<=3.0", "veto: gate AND align<=0.1",
         "replace: align<=3.0 only", "replace: align<=0.1 only",
         "rerank by alignment, then gate", "rerank, gate OR align<=0.1")
kable(do.call(rbind, lapply(sel, function(md) {
  d <- wdelta(md, "ades")
  data.frame(Model = md,
             Coverage = pct(ws(md, "ades", "coverage")),
             Accuracy = pct(ws(md, "ades", "accuracy")),
             Overall  = pct(ws(md, "ades", "overall")),
             `d correct` = if (md == "base") "—" else sprintf("%+s", num(d$dc)),
             `d wrong`   = if (md == "base") "—" else sprintf("%+s", num(d$dw)),
             check.names = FALSE)
})), row.names = FALSE, align = "lrrrrr",
  caption = "Arrivals: the bearing test used four ways. Only the rescue helps; the other three are catastrophic.")
```

**All three fail, and not marginally.** The veto costs
`r num(-wdelta("veto: gate AND align<=3.0", "ades")$dc)` correct arrivals at 3°.
Replace costs `r num(-wdelta("replace: align<=3.0 only", "ades")$dc)`. Rerank
costs `r num(-wdelta("rerank by alignment, then gate", "ades")$dc)` correct
labels *and* adds `r num(wdelta("rerank by alignment, then gate", "ades")$dw)`
wrong ones, dropping accuracy from `r pct(ws("base","ades","accuracy"))` to
`r pct(ws("rerank by alignment, then gate","ades","accuracy"))`.

### Why: the measure is out of domain on flights that landed {#sec-domain}

The diagnosis is not that alignment is a weak signal. It is that the flights the
gate already accepts are the wrong population to ask.

| | no course computable | median alignment | within 3° | on ground at the endpoint |
|---|---|---|---|---|
| gate accepts | 23.8% | **21°** | 18.4% | **43.1%** |
| gate refuses | 24.3% | 53° | 5.2% | 0.6% |

**43% of accepted arrivals are on the ground at their endpoint.** For those, the
last seven minutes are the approach, the landing roll and the taxi to stand —
and the "course" over that window is dominated by ground manoeuvring, which
points wherever the taxiways happen to run. A further quarter have no computable
course at all. A veto at 3° therefore rejects roughly four in five flights it
should keep.

The bearing test was developed on the abstained block, where every endpoint is
airborne and en route by construction. **Its domain is airborne endpoints, and
the flights the gate accepts are mostly not that.** Used as a rescue it never
leaves its domain, which is why it works there and nowhere else.

### Rerank refutes the naming-rule argument {#sec-naming}

The reranking result deserves its own note, because version 4.5 predicted the
opposite. Its argument was that the abstention decides only *whether* to answer
and never *what* to answer, that arrivals fail predominantly by misnaming, and
that a trajectory-based naming rule was therefore the most promising fix
available.

Ranking candidates by alignment instead of distance is that rule, and it is far
worse than ranking by distance. The reason is geometric: **an aerodrome 200 NM
directly ahead is perfectly aligned.** Alignment has no sense of scale, so as a
ranking criterion it systematically prefers distant aerodromes that happen to lie
on the track. Distance was doing enormous work that the argument did not credit.

Alignment is a **filter over distance-ranked candidates**, not a ranking
criterion. That is what the rescue variant does, and it is the only form in
which the signal is worth anything.

::: {.callout-note}
## A small correction to @sec-compare

The whole-sample run computes alignment without requiring an altitude, because
alignment is purely horizontal; @sec-compare's figures came from a routine that
also fitted the vertical slope and so discarded samples with no altitude. The
rescue numbers here are consequently a little higher —
`r num(wdelta("rescue: gate OR align<=0.1", "ades")$dc)` correct against
`r num(delta(BEAR, "ades")$dc)` at the same gate. The direction and the
recommendation are unchanged; the earlier figures were conservative.
:::

## What is now ruled out {#sec-ruled-out}

Recorded so it is not re-attempted:

* **Relaxing the abstention thresholds** — priced at
  `r sprintf("%.2f:1", GUESS_OK / (ABSTAINED - GUESS_OK))` (@sec-hard).
* **The approach cone** — version 4.5 measured it against the rectangle at
  matched coverage; departures lost up to 2.55 pp of accuracy, arrivals were a
  wash.
* **Bucket decimation** — the state-vector study measured +55 correct arrivals
  per day for a change that rewrites `track_id` for every published row.
* **Vertical rate, in any form** (@sec-vertical).
* **Trajectory extrapolation** — version 2 and the 2025 PRC report,
  independently.
* **Bearing as a veto, a replacement, or a naming rule** (@sec-whole). Only the
  rescue form works, and only because it stays inside the measure's domain.

## Limitations {#sec-limitations}

* **One sample.** Three days, 2025-06-05 to 07. The operating points were chosen
  as the argmax over 120 grid cells of that same sample, so the ratios are
  optimistic. The 2° and 3° columns peak cleanly at 7 minutes, which is why the
  *region* is credible; the specific numbers are not validated.
* **Two measurement bugs were found in this work**, both by sanity checks rather
  than by anything failing: `angle_between` used `%`, and Spark's remainder
  keeps the sign of the dividend, so bearings 20° apart measured 340° whenever
  the pair straddled north — the tell was a median alignment of 186° on a
  0–180° scale. And the altitude slope divided an altitude change by an elapsed
  time drawn from different rows, because `min_by` skips nulls where `min` does
  not. Both are fixed; both had already produced numbers that were reported
  before being caught.
* **The `>=90% fresh` band reverses the freshness trend** for the 2-point
  measure, on about a thousand flights. Either small-sample noise or something
  real about very short tracks. Not resolved.
* **`nearest` and `endpoint` are assumed to answer identically where both
  answer**, which underpins @sec-hard. It follows from both ranking on the same
  effective distance, but it is an argument, not a measurement.

## Reproducing {#sec-reproducing}

```bash
# the abstained block, its motion profile, and the angle x window sweep
python benchmarks/abstained_vertical.py --results-dir <dir>

# which vertical measure to trust, banded by freshness
python benchmarks/vertical_measure.py --results-dir <dir>

# base against the motion variants, whole population
python benchmarks/motion_model_compare.py --results-dir <dir>
```

Rendering this page needs no credentials, no database and no cluster.
