---
title: "Turning ADS-B state vectors into flight tracks"
subtitle: "Version 1 — what the segmentation is worth, and what stops it being worth it"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r regenerate}
#| include: false
#| cache: false

# Rendering this page runs the analysis behind it.
#
# `regenerate_track_v1.py` is the executable definition of the study: every CSV
# read below is produced by exactly one job it declares, with exactly the
# arguments it declares. It re-runs a job when the *code that job depends on*
# has changed since the output was written -- not when the file is old, which
# says nothing. So a render with everything current is a fast no-op, and a
# render after editing `segmentation/methods.py` recomputes every arm before the
# first table is drawn.
#
# Running a stale job needs cluster credentials, because the numbers come from
# Spark over S3 against Network Manager reference data and there is no way to
# recompute them without the data they are computed from. Set OPDI_RENDER=check
# to fail fast on staleness instead, or OPDI_RENDER=allow-stale to render a
# draft from whatever is staged -- which the provenance table at the foot of
# this page will then show as unverified.

# The pipeline repo is a sibling of the portal. That resolution is wrong inside
# a git worktree, where the paper sits under .claude/worktrees/<name>/papers and
# the sibling walk lands somewhere that does not exist -- so the regeneration
# step would be skipped silently and the page would render whatever is staged.
# OPDI_REPO_DIR overrides it; the default is unchanged for a normal checkout.
REPO   <- Sys.getenv("OPDI_REPO_DIR",
                     normalizePath(file.path(getwd(), "..", "..", "..", "opdi"),
                                   mustWork = FALSE))
PYTHON <- Sys.getenv("OPDI_PYTHON", file.path(REPO, ".venv310", "bin", "python"))
MODE   <- Sys.getenv("OPDI_RENDER", "run")

# A missing entrypoint must not read as "nothing to regenerate". Skipping it
# silently is how a page ends up publishing numbers no command can account for.
if (!file.exists(file.path(REPO, "benchmarks", "regenerate_track_v1.py")) &&
    MODE != "skip") {
  stop("regenerate_track_v1.py not found under REPO=", REPO,
       " -- set OPDI_REPO_DIR, or OPDI_RENDER=skip to render without it.")
}

regen <- file.path(REPO, "benchmarks", "regenerate_track_v1.py")
if (file.exists(regen) && MODE != "skip") {
  flags <- switch(MODE,
                  "run" = character(0),
                  "check" = "--check",
                  "allow-stale" = c("--check", "--allow-stale"),
                  character(0))
  status <- system2(PYTHON, c(shQuote(regen), flags), stdout = "", stderr = "")
  if (!identical(status, 0L)) {
    stop("regenerate_track_v1.py exited ", status,
         " -- the figures below would not match the code that is checked out. ",
         "Set OPDI_RENDER=allow-stale to render anyway.")
  }
}
```

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every figure and table on this page comes from a committed CSV
# under data/. Nothing here reads S3, a database or a Spark cluster.

D <- function(f) read.csv(file.path("data", f), stringsAsFactors = FALSE)

arms   <- rbind(D("arms_2025.csv"), D("arms_2024.csv"))
sweep1 <- rbind(D("sweep_2025_stage1.csv"), D("sweep_2025_stage1_ext.csv"))
sweep2 <- rbind(D("sweep_2025_stage2.csv"), D("sweep_2024_stage2.csv"))
payoff <- rbind(D("payoff_2025.csv"), D("payoff_2024.csv"))
fixcs  <- rbind(D("payoff_fixcallsign_2025.csv"), D("payoff_fixcallsign_2024.csv"))

num <- function(x) formatC(x, format = "d", big.mark = ",")
# The "+" flag does nothing for %s in R, so the sign is prepended explicitly --
# a signed delta that silently loses its sign reads as a level, not a change.
sgn <- function(x) paste0(if (x >= 0) "+" else "-", num(abs(x)))
pct <- function(x) sprintf("%.2f%%", x)

# One accessor for every lookup below, so no two tables can disagree about what
# an arm scored.
A <- function(a, p, col) arms[arms$arm == a & arms$period == p, col]
P <- function(df, a, p) {
  r <- df[df$arm == a & df$period == p, ]
  r$adep_score + r$ades_score
}

PERIODS <- c("2025", "2024")

# --- headline values, computed here rather than written into the prose, so the
# --- opening paragraph cannot drift away from the tables that justify it.
CLEAN_LEGACY_25 <- A("legacy", "2025", "clean_match_pct")
CLEAN_REC_25    <- A("recommended", "2025", "clean_match_pct")
CLEAN_LEGACY_24 <- A("legacy", "2024", "clean_match_pct")
CLEAN_REC_24    <- A("recommended", "2024", "clean_match_pct")
FRAG_LEGACY_25  <- A("legacy", "2025", "fragmented_pct")
FRAG_REC_25     <- A("recommended", "2025", "fragmented_pct")
MERGE_REC_25    <- A("recommended", "2025", "merged_pct")
MERGE_AIR_25    <- A("airframe_only", "2025", "merged_pct")
N_GT_25         <- A("legacy", "2025", "n_flights")
N_GT_24         <- A("legacy", "2024", "n_flights")

# The sweep: production against the best cell anywhere in the grid.
PROD <- function(df, p) {
  r <- df[df$period == p & df$gap_minutes == 30 &
          df$low_alt_gap_minutes == 15 & df$low_alt_ft == 5000, ]
  r$clean_match_pct[1]
}
SWEEP_PROD_25 <- PROD(sweep2, "2025")
SWEEP_BEST_25 <- max(sweep2$clean_match_pct[sweep2$period == "2025"])
SWEEP_PROD_24 <- PROD(sweep2, "2024")
SWEEP_BEST_24 <- max(sweep2$clean_match_pct[sweep2$period == "2024"])
SWEEP_CELLS   <- nrow(sweep1)
SWEEP_SPREAD1 <- max(sweep1$clean_match_pct) - min(sweep1$clean_match_pct)

# The payoff, as the pipeline stands and with the labelling repaired.
PAY_LEGACY_25 <- P(payoff, "legacy", "2025")
PAY_REC_25    <- P(payoff, "recommended", "2025")
FIX_LEGACY_25 <- P(fixcs, "legacy", "2025")
FIX_REC_25    <- P(fixcs, "recommended", "2025")
FIX_LEGACY_24 <- P(fixcs, "legacy", "2024")
FIX_REC_24    <- P(fixcs, "recommended", "2024")
```

## Summary {.unnumbered}

Before OPDI can derive a flight list, it has to decide which state vectors
belong to the same flight. That decision is made once, by a function that has
not changed since the pipeline was written, and everything downstream inherits
it. This page asks what that function is worth: how often it recovers a real
flight, whether its thresholds are mistuned, whether the reference
implementations or aviation domain knowledge do better, and what a better
segmentation buys the product OPDI actually publishes.

The short version is in four parts.

**The shipped segmentation recovers about half the flights it should.** On a
three-day sample of `r num(N_GT_25)` Network Manager flights it matches
**`r pct(CLEAN_LEGACY_25)`** of them one-to-one; **`r pct(FRAG_LEGACY_25)`** are
split across more than one track. That is not a tuning problem, and the failure
is almost entirely fragmentation rather than merging.

**Its parameters are not the problem.** A sweep of `r num(SWEEP_CELLS)` cells,
with every axis extended until the curve turned over inside the grid, spans
`r sprintf("%.2f", SWEEP_SPREAD1)` points. On the full three-day samples the
best cell beats production by
**`r sprintf("%+.2f", SWEEP_BEST_25 - SWEEP_PROD_25)`** points on 2025 and
**`r sprintf("%+.2f", SWEEP_BEST_24 - SWEEP_PROD_24)`** on 2024. The shipped
algorithm sits within a fifth of a point of the best setting anyone could pick
for it.

**The identity key is the problem.** The track's group key is
`(icao24, callsign)`, and a null callsign therefore starts a track of its own.
Grouping on the airframe and splitting only when the callsign *really* changes
takes one-to-one matching to **`r pct(CLEAN_REC_25)`** on 2025 and
**`r pct(CLEAN_REC_24)`** on 2024 — a gain of
**`r sprintf("%+.2f", CLEAN_REC_25 - CLEAN_LEGACY_25)`** points, roughly two
hundred times what the entire parameter sweep is worth.

**But the flight list cannot consume it, and that is the most useful finding
here.** Adopting the better segmentation *as the pipeline stands* costs
`r sgn(PAY_REC_25 - PAY_LEGACY_25)` ADEP/ADES points, because `flights.py`
labels each flight with `F.min("flight_id")` — the lexicographically smallest
callsign in the track, and a blank sorts first. That is correct only while every
track is callsign-homogeneous, which is a property the legacy group key supplies
by accident and which any fix to the fragmentation necessarily destroys. Repair
the labelling and the same segmentation scores
**`r num(FIX_REC_25)`** against legacy's **`r num(FIX_LEGACY_25)`** — behind by
`r num(FIX_LEGACY_25 - FIX_REC_25)` points, **0.5%** — while matching flights
`r sprintf("%.0f", CLEAN_REC_25 - CLEAN_LEGACY_25)` points better.

So the recommendation is conditional, and the conditional is the result rather
than a hedge: **fix one aggregate in the flight list, then ship the
segmentation.** In that order. Reversed, the pipeline takes a 34% coverage loss
in the interval between them.

## Reading this page

Sections are ordered as the work was done, not as the conclusion would suggest.
[The problem](#sec-problem) says what a track is and why the question was open.
[The metrics](#sec-metrics) matter more than usual here, because the obvious
scoring choice is misleading and the page says why before using something else.
[The ladder](#sec-ladder) is one section per candidate algorithm, each with what
it fixed and what it cost. [The sweep](#sec-sweep) prices the shipped algorithm's
own tuning. [The payoff](#sec-payoff) is where the study turns: it measures the
downstream effect, gets an answer that contradicts everything above it, and
tracks that contradiction to a single line. [What ships](#sec-defaults) states
the recommendation and the order of operations it depends on.

Every number is from a committed CSV regenerated by the entrypoint this page
runs before drawing anything; [Provenance](#sec-provenance) lists them file by
file. [Limitations](#sec-limitations) is where the claims this page does *not*
make are written down.

## What a track is, and why the question is open {#sec-problem}

OPDI ingests OpenSky state vectors — position reports, one aircraft at a time,
with no notion of a flight. A **track** is the pipeline's answer to "which of
these belong together", and it is the unit everything downstream is built on:
the flight list groups by `track_id`, events are extracted per track, and
measurements accumulate from a track's start.

The shipped rule is short. Group the samples by `(icao24, callsign)`, order them
by time, and start a new track when the gap to the previous sample exceeds 30
minutes, or exceeds 15 minutes while the aircraft is below 5,000 ft. The
resulting identifier carries a month suffix.

Three things can go wrong with such a rule, and they are not symmetric:

- **Fragmentation** — one real flight becomes several tracks. Everything
  downstream sees several short flights where there was one.
- **Merging** — several real flights become one track. The flight list emits one
  flight and the others simply do not exist in the output.
- **Boundary error** — the track is one flight, but starts or ends in the wrong
  place, so first- and last-seen events and cumulative measurements are wrong.

Of the three, merging is the most damaging, because a merged flight is
unrecoverable downstream, while a fragmented one is at least present in pieces
and a boundary error leaves the flight intact. The metrics below are chosen so
that merging and fragmentation cannot be confused with each other, and the
candidate rules are compared on both rather than on a single blended score.

The rule has never been benchmarked, and that is why this study exists.

What it is *not* is frozen. Segmentation is a versioned choice: a release states
which rule built its `track_id` values, and changing that rule changes
`track_id` for everything produced from that release forward — in value, and in
shape where the month suffix goes. That is a release decision, to be documented
and communicated to consumers rather than avoided, and it is a decision that has
now been taken. As of this release `segmentation.method` defaults to
`"standard"`, which is the rule this study recommends. `"legacy"` remains
selectable and reproduces the identifiers of every dataset published up to OPDI
v0.0.2, so released data stays reproducible.

One consequence for the reader, stated here so nothing later depends on
inferring it. **This paper measures algorithms, not the pipeline.** Every rule
below is expressed in a generic segmentation engine and scored in a harness, so
that eight candidates could be compared without eight production pipeline runs —
which at this data volume is the difference between a study and a quarter. That
is a legitimate design and it is also a real limitation: what is measured here
is a re-implementation of each rule, not the pipeline's own output. Version 2 of
this study is the paper that runs the pipeline itself, end to end, on the rule
this one recommends. If you want to know what a rule is worth, you are in the
right place; if you want to know what the pipeline now produces, read version 2.

## How a segmentation is scored {#sec-metrics}

Comparing a segmentation against ground truth is a **clustering comparison**:
both the algorithm and the truth partition the same state vectors, and the
question is how far the two partitions agree.

### Homogeneity is merging; completeness is fragmentation

The standard pair of clustering measures maps onto the two failure modes
exactly, which is why they are used here rather than a bespoke score:

- **Homogeneity** falls when a track contains samples from more than one flight
  — it is the merging measure.
- **Completeness** falls when one flight's samples are spread across more than
  one track — it is the fragmentation measure.

**V-measure** is their harmonic mean, and it is reported below but never used to
choose anything. The reason is measured rather than assumed: across the eight
arms V-measure spans `r sprintf("%.4f", min(arms$v_measure))` to
`r sprintf("%.4f", max(arms$v_measure))` — a range of
`r sprintf("%.4f", max(arms$v_measure) - min(arms$v_measure))` — while
one-to-one matching spans `r sprintf("%.1f", min(arms$clean_match_pct))` to
`r sprintf("%.1f", max(arms$clean_match_pct))`. Worse than insensitive, it is
*misleading*: it ranks arms in a different order at the top, which is the only
place a comparison looks. Both statistics are entropy-weighted over 143 million
samples, and the overwhelming majority of samples sit in tracks that every arm
gets right, so V-measure is dominated by agreement that carries no information
about the arms' differences.

### Why the headline is one-to-one matching

The headline statistic is **`clean_match_pct`**: the share of ground-truth
flights matched by exactly one track, where that track covers no other flight.
It is deliberately strict and deliberately blunt. Each flight is counted in
exactly one of three mutually exclusive classes — clean, fragmented or merged —
and a flight that is both merged and fragmented counts as merged, because
merging is the worse failure and a flight should not be able to improve its
classification by also being broken a second way.

### Joining to ground truth without the callsign

Ground truth is joined to tracks on **`icao24` plus interval containment**, and
never on callsign. This differs from the V6/V7 flight-list studies, which key on
callsign, and the difference is forced: one of the arms removes callsign from
the track identity entirely, so a callsign-keyed join would be measuring the
arm's own identity choice rather than its segmentation. Flights are restricted
to those whose whole airborne interval lies inside the sampled window, so a
flight that departs on the last sampled day and lands after it cannot report a
clipping artefact as a boundary error.

### Boundary error, reported signed

Boundary error is the distance from a track's true first and last sample to the
flight's actual take-off and landing times. It is reported **signed**, and that
is not cosmetic. Absolute error conflates two opposite conditions: a track
starting *before* take-off is normal — the aircraft is on the ramp broadcasting
— while a track starting *after* take-off has missed part of its own flight. The
absolute median for the shipped algorithm is around a hundred seconds, which
reads as a small tidy number; signed, the median is
**`r sprintf("%+d", as.integer(A("legacy", "2025", "off_signed_p50_s")))` s** and
the tenth percentile is
**`r sprintf("%+d", as.integer(A("legacy", "2025", "off_signed_p10_s")))` s**,
which says something quite different — the typical track begins just *after*
wheels-off, while a minority carry a full quarter-hour of taxi-out. The
distribution is two-sided and an absolute median over it is close to
meaningless.

The landing side is the opposite and far more uniform: every percentile is
positive, with a median of
**`r sprintf("%+d", as.integer(A("legacy", "2025", "land_signed_p50_s")))` s**.
Tracks essentially always run on through taxi-in. That asymmetry is a coverage
fact — on-ground ADS-B before departure reaches a minority of flights, while an
arriving aircraft is still powered and broadcasting to the stand — and not an
algorithm defect.

## The sample {#sec-sample}

Two three-day samples, `r num(N_GT_25)` and `r num(N_GT_24)` ground-truth
flights: 5–7 June 2025 and 5–7 June 2024. Two periods, because a single sample
cannot distinguish a real effect from one the sample happened to favour, and
every conclusion on this page is required to hold on both.

Ground truth is the Network Manager flight table, cross-checked against APDF
where APDF covers the aerodrome. Take-off is inferred as `AOBT_3 + TAXI_TIME_3`
and landing is taken from `ARVT_3` directly; both were validated against real
APDF movement times before being relied on, at median error 0 s with an
interquartile range of 17 s and 25 s respectively.

## The ladder {#sec-ladder}

Eight arms, each a grouping and a break predicate, all scored against the same
ground truth on the same samples. Every arm reads one shared cleaned track
table with only `track_id` varying, so a difference between arms cannot be an
artefact of a different track build.

```{r ladder-table}
#| echo: false
t <- arms[arms$period == "2025", c("arm", "clean_match_pct", "fragmented_pct",
                                   "merged_pct", "n_tracks", "v_measure")]
t <- t[order(-t$clean_match_pct), ]
t24 <- arms[arms$period == "2024", c("arm", "clean_match_pct")]
names(t24)[2] <- "clean_2024"
t <- merge(t, t24, by = "arm", sort = FALSE)
t <- t[order(-t$clean_match_pct), ]
kable(t, row.names = FALSE, digits = c(0, 2, 2, 2, 0, 4, 2),
      col.names = c("arm", "clean %", "fragmented %", "merged %",
                    "tracks", "V-measure", "clean % (2024)"),
      format.args = list(big.mark = ","))
```

The ordering is identical on both periods and no arm moves by more than about
three points between them, so what follows is a property of the algorithms
rather than of one sample.

### A0 `legacy` — the shipped rule

`r pct(CLEAN_LEGACY_25)` clean, `r pct(FRAG_LEGACY_25)` fragmented,
`r pct(A("legacy", "2025", "merged_pct"))` merged. The failure is overwhelmingly
fragmentation, and it is worth being precise about the cause, because the
obvious reading is wrong. It is not that callsigns change mid-flight. It is that
**a null callsign forms its own group**: 42% of legacy's tracks are
blank-callsign tracks, produced by samples the receiver decoded without a
callsign. The group key turns a gap in the data into a separate flight.

### A2 `no_month_suffix` — untestable on this sample

Identical to A0 in every digit, on both periods. That is correct behaviour and
not a bug: the arm removes the `_{year}_{month}` suffix from the identifier, and
both samples sit inside a single month, so the suffix is constant and removing
it changes nothing. The consequence is that **this study cannot evaluate the
arm at all.** The suffix can only fragment a flight airborne at midnight on the
first of a month, and neither sample contains one. Reported here as untested
rather than as a null result; testing it needs a sample spanning a month
boundary, which is two days of cluster time and is recommended as follow-up.

### A3 `traffic_style` — the reference library's rule does not transfer

`r pct(A("traffic_style", "2025", "clean_match_pct"))` clean, below legacy, and
with the worst merging of any arm at
`r pct(A("traffic_style", "2025", "merged_pct"))` — roughly twenty times
legacy's. A single ten-minute gap threshold with a predicate deciding whether a
candidate gap is really a break is well suited to the *traffic* library's
typical input; on this feed it merges. A genuine negative result: the reference
implementation's approach does not transfer unmodified.

### A4 `airframe_only` — group on the airframe

`r pct(A("airframe_only", "2025", "clean_match_pct"))` clean, fragmentation down
from `r pct(FRAG_LEGACY_25)` to
`r pct(A("airframe_only", "2025", "fragmented_pct"))`. This is the single
largest gain in the study and it comes from deleting one column from the group
key. Identity belongs to the airframe; the callsign is something the airframe
was broadcasting at the time.

The cost is merging, up from `r pct(A("legacy", "2025", "merged_pct"))` to
`r pct(MERGE_AIR_25)`: without callsign in the key, two flights of the same
airframe with no sufficient gap between them join into one track.

### A5 `ground_anchored` and A6 `airport_anchored`

Both replace or supplement the altitude proxy with something more physical — a
ground-contact dwell long enough to be a turnaround, and real airport geometry
respectively. Both beat legacy substantially
(`r pct(A("ground_anchored", "2025", "clean_match_pct"))` and
`r pct(A("airport_anchored", "2025", "clean_match_pct"))`) and both are worth
less than simply fixing the group key. Their signed boundary error is however
the tightest of any arm — `ground_anchored`'s tenth percentile is
`r sprintf("%+d", as.integer(A("ground_anchored", "2025", "off_signed_p10_s")))` s
against legacy's
`r sprintf("%+d", as.integer(A("legacy", "2025", "off_signed_p10_s")))` s —
which is direct evidence the arm does what it claims: it cuts at the ground
event rather than running back into the ramp.

### A7 `vertical_profile` — a negative result, correctly attributed

`r pct(A("vertical_profile", "2025", "clean_match_pct"))` clean, last of the
eight. The arm splits on a descent that reaches a floor followed by a climb back
through it. An earlier implementation split on *any* upward crossing, which is
not the idea, and its poor score was therefore unattributable — it could have
been the idea failing or the code not being the idea. Implemented properly, and
on height above field rather than sea level, it scores *worse* still. That makes
it a real negative result rather than an unexplained number: every percentile of
its signed off-block error is positive, meaning it systematically starts tracks
late, having sliced off each departure's initial climb.

### A8 `recommended` — both fixes at once

`r pct(CLEAN_REC_25)` clean on 2025 and `r pct(CLEAN_REC_24)` on 2024, the best
of the eight on both, with merging at `r pct(MERGE_REC_25)` — lower than legacy
itself.

It beats both of the arms it draws on rather than landing between them, which is
the sign that the two failures really do have different causes. Legacy fragments
because null callsigns form their own groups; `airframe_only` merges because
nothing separates two flights of one airframe. So: group on `icao24`, so a blank
or flickering callsign no longer starts a track, and break when the last
*non-blank* callsign differs from the current one, so two genuinely different
flights still part.

The subtlety is in the lookback. The break predicate is evaluated over the whole
airframe's samples and has no notion of the track boundaries it is deciding, so
an unbounded search for "the previous real callsign" reaches back past a gap
into the previous flight and splits the next one in its own middle. Bounding the
lookback to the gap threshold is what makes the arm work; without it the same
rule scores 62% rather than 91%.

## The shipped algorithm's tuning headroom {#sec-sweep}

A1 is not new code: it is A0 at parameters other than production's, so the sweep
measures the shipped algorithm's own tuning and nothing else.

`r num(SWEEP_CELLS)` cells on one day, spanning
`r sprintf("%.2f", SWEEP_SPREAD1)` points end to end.

```{r sweep-axes}
#| echo: false
by_axis <- function(col, label) {
  agg <- aggregate(clean_match_pct ~ get(col), data = sweep1, FUN = max)
  names(agg) <- c(label, "best clean %")
  agg
}
kable(by_axis("low_alt_ft", "low_alt_ft"), row.names = FALSE, digits = 2,
      caption = "Best cell at each low-altitude ceiling. The optimum is interior.")
```

All three axes have an interior optimum, which is the condition for calling a
result an optimum at all. That took an extension: the first grid put the
low-altitude ceiling's best value at its largest tried, with the curve still
rising. An optimum at a grid edge is not an optimum, it is where the grid
stopped, and this pipeline's own history contains a sweep that stopped early and
was read as "higher is always better". Extended, the ceiling peaks at 20,000 ft
and falls again by 30,000 — as it must, since past cruise altitude every sample
counts as "low" and the two-threshold rule collapses into one.

Re-measured on the full three-day samples, the best cell beats production by
**`r sprintf("%+.2f", SWEEP_BEST_25 - SWEEP_PROD_25)`** points on 2025 and
**`r sprintf("%+.2f", SWEEP_BEST_24 - SWEEP_PROD_24)`** on 2024. Both periods
choose the same cell out of sample.

That is the answer to the tuning question, and it is a clean negative: **the
shipped segmentation is not mistuned.** It sits a fraction of a point from the
best setting available to it, in a grid drawn wider than plausible and extended
until every axis turned over. Set against the
`r sprintf("%+.0f", CLEAN_REC_25 - CLEAN_LEGACY_25)` points the identity key is
worth, the entire tuning question is a rounding error.

## What a better segmentation buys ADEP/ADES {#sec-payoff}

The point of segmenting well is the product built on top of it. Each arm's
assignment is fed to the V7 flight list at its own recommended configuration,
with only `track_id` varying, so any difference in ADEP/ADES accuracy is
attributable to the segmentation.

The control is `legacy`, which reproduces the frozen algorithm: run through this
harness it scores `r num(PAY_LEGACY_25)` against V7's published 134,168 for the
same sample — a difference of 8, or 0.006%. That residual is itself informative
and is explained in [Limitations](#sec-limitations).

### The ranking inverts

```{r payoff-table}
#| echo: false
p <- payoff[payoff$period == "2025", ]
p$total <- p$adep_score + p$ades_score
p$vs_legacy <- p$total - PAY_LEGACY_25
p <- p[order(-p$total), c("arm", "total", "vs_legacy", "adep_coverage",
                          "adep_accuracy")]
p$adep_coverage <- 100 * p$adep_coverage
p$adep_accuracy <- 100 * p$adep_accuracy
kable(p, row.names = FALSE, digits = c(0, 0, 0, 2, 2),
      col.names = c("arm", "ADEP+ADES", "vs legacy", "ADEP coverage %",
                    "ADEP accuracy %"),
      format.args = list(big.mark = ","))
```

The order is close to exactly reversed. The arm that reproduces individual
flights best is *last*; the shipped algorithm, second-worst at segmentation, is
first. Taken at face value this says a better segmentation makes the product
worse.

It is not what it says. **Accuracy barely moves across the whole table while
coverage collapses** — the losing arms are not predicting *wrongly*, they are
predicting for far fewer flights. That is the signature of a labelling failure,
not a geometric one.

### One aggregate explains it

The flight list builds one flight per `track_id` and labels it
`F.min("flight_id")` — the lexicographically smallest callsign in the track. A
blank sorts first.

While callsign is part of the track's group key, every track is
callsign-homogeneous by construction and that aggregate cannot lose anything:
measured on 2025, legacy loses a real callsign on **0 of 235,949 tracks**. Drop
callsign from the key and **64% of `airframe_only`'s tracks carry a real
callsign alongside blank ones**, and `F.min` returns the blank. The flight
becomes unidentifiable and never matches its ground-truth counterpart.

So the coupling is tighter than "one callsign per track". The flight list
requires **blank-callsign samples to sit in tracks of their own** — precisely
what the legacy group key does as a side effect, and precisely what any fix to
the fragmentation must undo. It is the same property, seen from both ends:
legacy's 50% fragmentation and the downstream coverage collapse are one
phenomenon.

### With the labelling repaired

Relabelling each track with its most frequent non-blank callsign, and carrying
`legacy` as a control that must not move:

```{r fix-table}
#| echo: false
rows <- do.call(rbind, lapply(PERIODS, function(p) {
  data.frame(
    period = p,
    arm = c("legacy (control)", "recommended"),
    before = c(P(payoff, "legacy", p), P(payoff, "recommended", p)),
    after  = c(P(fixcs, "legacy", p), P(fixcs, "recommended", p)),
    check.names = FALSE)
}))
rows$delta <- rows$after - rows$before
kable(rows, row.names = FALSE, digits = 0,
      col.names = c("period", "arm", "as the pipeline stands",
                    "labelling repaired", "delta"),
      format.args = list(big.mark = ","))
```

The control returns **exactly +0**, on both periods and across three independent
runs. That is what makes the other row attributable to the relabelling rather
than to anything else the run touched.

With the labelling repaired the recommended segmentation scores
**`r num(FIX_REC_25)`** against legacy's **`r num(FIX_LEGACY_25)`** on 2025 and
**`r num(FIX_REC_24)`** against **`r num(FIX_LEGACY_24)`** on 2024 — short by
`r num(FIX_LEGACY_25 - FIX_REC_25)` and `r num(FIX_LEGACY_24 - FIX_REC_24)`
points, 0.5% and 0.6% — while matching flights around
`r sprintf("%.0f", CLEAN_REC_25 - CLEAN_LEGACY_25)` points better. Its ADEP
coverage slightly *exceeds* legacy's on both periods.

The residual is real and is not rounded away: with the labelling fixed, what
remains is merging, and the recommended arm still merges
`r pct(MERGE_REC_25)` of flights where legacy merges
`r pct(A("legacy", "2025", "merged_pct"))`.

## What ships {#sec-defaults}

**Recommended: the A8 rule — group on `icao24`, break on a genuine non-blank
callsign change with the lookback bounded to the gap threshold, keeping legacy's
gap rules as a floor.**

Conditional on one change to the flight list, and the order matters:

1. **First**, change `flights.py`'s flight labelling so it does not choose a
   callsign with `F.min`. Most-frequent non-blank is what this study measured;
   any rule that ignores blanks would do. This is worth doing on its own merits
   even if the segmentation never changes, because it is a silent trap for any
   future change to track identity and it fails as a coverage statistic rather
   than as an error.
2. **Then** ship the segmentation.

Reversed, the pipeline spends the interval between the two changes at 55%
ADEP coverage instead of 80%.

Two consequences of shipping it, both of which are decisions for the release
plan rather than for this study:

- **`track_id` changes for all data from that release forward**, in shape as
  well as value: A8 carries no `_{year}_{month}` suffix, so identifiers become
  `{hash}_{offset}`. Any consumer parsing the suffix breaks in a way a pure
  value change would not.
- **Past months will not reproduce.** This study recommends; it does not
  schedule.

## Provenance {#sec-provenance}

Each row is one file this page reads: the job that produced it, the commit it
was produced at, and when. Every figure on this page is regenerated by
`benchmarks/regenerate_track_v1.py`, which the page runs before drawing
anything, so a row marked **unverified** would mean a file staged outside that
chain and traceable to no command.

```{r provenance}
#| echo: false
#| results: asis
mf <- file.path("data", "_manifest.json")
if (!requireNamespace("jsonlite", quietly = TRUE)) {
  cat("*jsonlite not available; provenance table skipped.*\n")
} else if (!file.exists(mf)) {
  cat("**No manifest.** Every CSV here was staged by hand and none can be traced to the command that produced it.\n")
} else {
  m <- jsonlite::fromJSON(mf, simplifyDataFrame = FALSE)
  files <- sort(list.files("data", pattern = "[.]csv$"))
  rows <- do.call(rbind, lapply(files, function(f) {
    e <- m[[f]]
    if (is.null(e)) {
      data.frame(File = f, Job = "**unverified**", Commit = "—",
                 Produced = "—", check.names = FALSE)
    } else {
      data.frame(File = f, Job = basename(e$script),
                 Commit = paste0(e$git_sha, if (isTRUE(e$git_dirty)) " (dirty)" else ""),
                 Produced = substr(e$produced_utc, 1, 16), check.names = FALSE)
    }
  }))
  print(kable(rows, row.names = FALSE, align = "l"))
  n_un <- sum(rows$Job == "**unverified**")
  cat(sprintf("\n%d files, %d traced to a command, %d unverified.\n",
              nrow(rows), nrow(rows) - n_un, n_un))
}
```

## Limitations {#sec-limitations}

### The baseline is not the shipped `track_id`

A0 reproduces the legacy *rule*, but applied to cleaned tracks it does not
reproduce the shipped identifiers exactly: it differs on 909 of 236,899 tracks,
0.38%, and 8 of 134,176 ADEP/ADES points. The cause is exact and worth reporting
upward on its own account. Production's identifiers were computed upstream on
raw altitudes, while `osn_tracks_clean` masks bad altitudes to NULL — and
`NULL < 5000` is NULL, not true, so the low-altitude split silently does not fire
where it once did. Measured, **950 of 950 disagreements have a NULL barometric
altitude**, against a 21.4% base rate.

The published `osn_tracks_clean` therefore carries a `track_id` that its own
columns can no longer reproduce. Nothing is wrong with the published data, but
anyone recomputing identifiers from the clean table — the obvious thing to do —
gets different tracks and no warning.

This does not affect the comparisons on this page: every arm reads the same
cleaned table and is subject to the same behaviour.

### Ground truth covers Network Manager flights only

Merges with traffic absent from the NM flight table — military, general
aviation, non-ECAC operators — are invisible to the metric, because the merged
partner has no ground-truth flight to be merged with. **The merge rates on this
page are therefore floors, not estimates.** Fragmentation is unaffected, since
it is measured within flights that ground truth does contain.

### One arm is untested, not null

`no_month_suffix` is identical to legacy on both samples because both sit inside
a single month. It is reported as untested. Evaluating it needs a sample
spanning a month boundary.

### Two three-day samples in June

Both samples are in early June. Nothing here licenses a seasonal claim, and the
callsign statistics in particular could plausibly differ in a period with more
charter or seasonal traffic.

### The A1 sweep varied three parameters

The sweep covers the gap threshold, the low-altitude gap threshold and the
low-altitude ceiling. It does not vary the ground-dwell or turnaround
parameters that the domain arms introduce, so those arms are compared at one
setting each and could be better at another. Given the structural gap between
the arms is two orders of magnitude larger than anything the sweep found, this
is unlikely to change the ordering, but it is not measured.

### The payoff holds the flight list fixed

Each arm is scored at V7's own recommended flight-list configuration rather than
re-tuned per arm. A joint optimum might be better for some arm; finding it would
multiply cluster time by the number of arms, and one Spark job runs at a time.
The natural follow-up is to re-tune the flight list on the recommended arm alone
— one extra grid, not eight — once the labelling is fixed.
