---
title: "Turning ADS-B state vectors into flight tracks"
subtitle: "Version 1 — what the segmentation is worth, and what had to change downstream before a better one could ship"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r regenerate}
#| include: false
#| cache: false

# Rendering this page runs the analysis behind it.
#
# `regenerate_track_v1.py` is the executable definition of the study: every CSV
# read below is produced by exactly one job it declares, with exactly the
# arguments it declares. It re-runs a job when the *code that job depends on*
# has changed since the output was written -- not when the file is old, which
# says nothing. So a render with everything current is a fast no-op, and a
# render after editing `segmentation/methods.py` recomputes every arm before the
# first table is drawn.
#
# Running a stale job needs cluster credentials, because the numbers come from
# Spark over S3 against Network Manager reference data and there is no way to
# recompute them without the data they are computed from. Set OPDI_RENDER=check
# to fail fast on staleness instead, or OPDI_RENDER=allow-stale to render a
# draft from whatever is staged -- which the provenance table at the foot of
# this page will then show as unverified.

# The pipeline repo is a sibling of the portal. That resolution is wrong inside
# a git worktree, where the paper sits under .claude/worktrees/<name>/papers and
# the sibling walk lands somewhere that does not exist -- so the regeneration
# step would be skipped silently and the page would render whatever is staged.
# OPDI_REPO_DIR overrides it; the default is unchanged for a normal checkout.
REPO   <- Sys.getenv("OPDI_REPO_DIR",
                     normalizePath(file.path(getwd(), "..", "..", "..", "opdi"),
                                   mustWork = FALSE))
PYTHON <- Sys.getenv("OPDI_PYTHON", file.path(REPO, ".venv310", "bin", "python"))
MODE   <- Sys.getenv("OPDI_RENDER", "run")

# A missing entrypoint must not read as "nothing to regenerate". Skipping it
# silently is how a page ends up publishing numbers no command can account for.
if (!file.exists(file.path(REPO, "benchmarks", "regenerate_track_v1.py")) &&
    MODE != "skip") {
  stop("regenerate_track_v1.py not found under REPO=", REPO,
       " -- set OPDI_REPO_DIR, or OPDI_RENDER=skip to render without it.")
}

regen <- file.path(REPO, "benchmarks", "regenerate_track_v1.py")
if (file.exists(regen) && MODE != "skip") {
  flags <- switch(MODE,
                  "run" = character(0),
                  "check" = "--check",
                  "allow-stale" = c("--check", "--allow-stale"),
                  character(0))
  status <- system2(PYTHON, c(shQuote(regen), flags), stdout = "", stderr = "")
  if (!identical(status, 0L)) {
    stop("regenerate_track_v1.py exited ", status,
         " -- the figures below would not match the code that is checked out. ",
         "Set OPDI_RENDER=allow-stale to render anyway.")
  }
}
```

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every figure and table on this page comes from a committed CSV
# under data/. Nothing here reads S3, a database or a Spark cluster.

D <- function(f) read.csv(file.path("data", f), stringsAsFactors = FALSE)

arms   <- rbind(D("arms_2025.csv"), D("arms_2024.csv"))
sweep1 <- rbind(D("sweep_recommended_2025_stage1.csv"),
                D("sweep_recommended_2025_stage1_ext.csv"))
payoff <- rbind(D("payoff_2025.csv"), D("payoff_2024.csv"))
fixcs  <- rbind(D("payoff_fixcallsign_2025.csv"), D("payoff_fixcallsign_2024.csv"))
cont   <- rbind(D("containment_2025.csv"), D("containment_2024.csv"))
tfill  <- rbind(D("traffic_fill_2025.csv"), D("traffic_fill_2024.csv"))

# The boundary histograms carry no period column -- one file per period *is* the
# period -- so it is attached on read rather than assumed by anything later.
readhist <- function(p) {
  x <- D(paste0("boundary_hist_", p, ".csv"))
  x$period <- p
  x
}
bhist <- rbind(readhist("2025"), readhist("2024"))

num <- function(x) formatC(x, format = "d", big.mark = ",")
# The "+" flag does nothing for %s in R, so the sign is prepended explicitly --
# a signed delta that silently loses its sign reads as a level, not a change.
sgn <- function(x) paste0(if (x >= 0) "+" else "-", num(abs(x)))
pct <- function(x) sprintf("%.2f%%", x)
pp  <- function(x) sprintf("%+.2f", x)
sec <- function(x) sprintf("%+d", as.integer(x))

# One accessor per file, so no two tables on this page can disagree about what
# an arm scored.
A  <- function(a, p, col) arms[arms$arm == a & arms$period == p, col]
CT <- function(p, col) cont[cont$period == p, col]
TF <- function(p, col) tfill[tfill$period == p, col]
# Of the pre-gap samples that are not already NULL, what share are exact repeats
# of their predecessor. Both rates are over every pre-gap sample and the two
# classes are disjoint, so this is the conditional share and not a ratio of
# unrelated denominators.
TF_REPEAT_OF_LIVE <- function(p) 100 * TF(p, "pregap_repeat_pct") /
                                 (100 - TF(p, "pregap_null_pct"))
PR <- function(df, a, p, col) df[df$arm == a & df$period == p, col]
P  <- function(df, a, p) {
  r <- df[df$arm == a & df$period == p, ]
  r$adep_score + r$ades_score
}

PERIODS <- c("2025", "2024")

# --- headline values, computed here rather than written into the prose, so the
# --- opening paragraph cannot drift away from the tables that justify it.
CLEAN_LEGACY_25 <- A("legacy", "2025", "clean_match_pct")
CLEAN_REC_25    <- A("recommended", "2025", "clean_match_pct")
CLEAN_LEGACY_24 <- A("legacy", "2024", "clean_match_pct")
CLEAN_REC_24    <- A("recommended", "2024", "clean_match_pct")
FRAG_LEGACY_25  <- A("legacy", "2025", "fragmented_pct")
FRAG_REC_25     <- A("recommended", "2025", "fragmented_pct")
MERGE_LEGACY_25 <- A("legacy", "2025", "merged_pct")
MERGE_REC_25    <- A("recommended", "2025", "merged_pct")
MERGE_AIR_25    <- A("airframe_only", "2025", "merged_pct")
GAIN_25         <- CLEAN_REC_25 - CLEAN_LEGACY_25
GAIN_24         <- CLEAN_REC_24 - CLEAN_LEGACY_24
N_GT_25         <- A("legacy", "2025", "n_flights")
N_GT_24         <- A("legacy", "2024", "n_flights")
N_APDF_25       <- A("legacy", "2025", "n_apdf_flights")

# The same three arms scored against the gate-to-gate interval rather than the
# airborne one. Held beside the airborne values so the headline cannot quote one
# pair and the table the other.
GATE_LEGACY_25  <- A("legacy", "2025", "gate_clean_match_pct")
GATE_REC_25     <- A("recommended", "2025", "gate_clean_match_pct")
GATE_LEGACY_24  <- A("legacy", "2024", "gate_clean_match_pct")
GATE_REC_24     <- A("recommended", "2024", "gate_clean_match_pct")
GATE_GAIN_25    <- GATE_REC_25 - GATE_LEGACY_25
GATE_GAIN_24    <- GATE_REC_24 - GATE_LEGACY_24
N_GATE_25       <- A("legacy", "2025", "gate_n_flights")
# What the wider interval adds to an arm, signed. Negative is the usual case:
# the gate interval can only add samples, and added samples can only fragment.
GATE_DROP <- function(a, p) A(a, p, "gate_clean_match_pct") - A(a, p, "clean_match_pct")
# Which arms carry callsign in the group key. Written once, read by the table
# that reports what the gate interval detects.
GROUPKEY <- c(legacy = "icao24 + callsign", no_month_suffix = "icao24 + callsign",
              traffic_style = "icao24 + callsign", airframe_only = "icao24",
              recommended = "icao24", ground_anchored = "icao24",
              airport_anchored = "icao24", vertical_profile = "icao24")

# The sweep of the rule that ships. The shipped cell is named once, from the
# config values quoted in the ladder, and every rank and delta below is located
# by matching on it -- so a change of default moves the whole section rather
# than leaving one stale sentence behind.
SHIP_GAP    <- 30
SHIP_LOWGAP <- 15
SHIP_LOWFT  <- 5000
SHIP_LOOKBACK <- NA   # None: the lookback follows gap_minutes

# The full three-axis grid, one day. `sweep1` is stage 1 plus its extension.
SWEEP_CELLS   <- nrow(D("sweep_recommended_2025_stage1.csv"))
SWEEP_MIN1    <- min(sweep1$clean_match_pct)
SWEEP_MAX1    <- max(sweep1$clean_match_pct)
SWEEP_SPREAD1 <- SWEEP_MAX1 - SWEEP_MIN1
s1 <- D("sweep_recommended_2025_stage1.csv")
s1 <- s1[order(-s1$clean_match_pct), ]
SHIP_ROW      <- which(s1$gap_minutes == SHIP_GAP &
                       s1$low_alt_gap_minutes == SHIP_LOWGAP &
                       s1$low_alt_ft == SHIP_LOWFT)[1]
SHIP_RANK     <- SHIP_ROW
SHIP_CLEAN_25 <- s1$clean_match_pct[SHIP_ROW]
SWEEP_GAIN_25 <- max(s1$clean_match_pct) - SHIP_CLEAN_25

# The lookback axis, swept at two cells: the cell 2025's grid preferred, and the
# cell that ships. `LB_GAIN_*` is what decoupling the lookback from gap_minutes
# would buy -- the number that decides whether the parameter ships as None.
LB <- function(f) {
  d <- D(f)
  d[order(-d$clean_match_pct), ]
}
lb_t <- LB("sweep_recommended_2025_stage2.csv")
lb_s <- LB("sweep_recommended_2025_stage2b_shipped.csv")
LB_GAP_TUNED  <- lb_t$gap_minutes[1]
LB_GAP_SHIP   <- lb_s$gap_minutes[1]
LB_PEAK_TUNED <- lb_t$callsign_lookback_minutes[1]
LB_PEAK_SHIP  <- lb_s$callsign_lookback_minutes[1]
# The coupled setting is the row whose lookback equals the cell's gap_minutes.
LB_COUPLED <- function(d) d$clean_match_pct[d$callsign_lookback_minutes ==
                                            d$gap_minutes[1]][1]
LB_GAIN_TUNED <- lb_t$clean_match_pct[1] - LB_COUPLED(lb_t)
LB_GAIN_SHIP  <- lb_s$clean_match_pct[1] - LB_COUPLED(lb_s)

# 2024, where the gap ordering reverses. GAP30/GAP40 are the best cell at each
# gap value, which is what "the ordering reverses" has to be measured on.
s3      <- D("sweep_recommended_2024_stage3.csv")
SHIP_24 <- s3$clean_match_pct[s3$gap_minutes == SHIP_GAP &
                              s3$low_alt_gap_minutes == SHIP_LOWGAP &
                              s3$low_alt_ft == SHIP_LOWFT][1]
BEST_24  <- max(s3$clean_match_pct)
GAP30_24 <- max(s3$clean_match_pct[s3$gap_minutes == 30])
GAP40_24 <- max(s3$clean_match_pct[s3$gap_minutes == 40])
# The largest movement of any arm's clean-match rate between the two periods.
# Quoted rather than asserted, because "the ordering holds on both samples" is
# the claim the second period was bought to support.
ARM_SHIFT <- max(abs(sapply(unique(arms$arm), function(a)
  A(a, "2025", "clean_match_pct") - A(a, "2024", "clean_match_pct"))))

# The headline payoff is the two-arm run: `legacy` against `recommended`, both
# carrying production's callsign fix and the harness's own repair, which is the
# configuration production now matches. The eight-arm run is kept separately,
# for the ranking table and as the convergence check's other side, and is never
# the headline.
PAY_LEGACY_25 <- P(fixcs, "legacy", "2025")
PAY_REC_25    <- P(fixcs, "recommended", "2025")
PAY_LEGACY_24 <- P(fixcs, "legacy", "2024")
PAY_REC_24    <- P(fixcs, "recommended", "2024")
PAY_SHORT_25  <- PAY_LEGACY_25 - PAY_REC_25
PAY_SHORT_24  <- PAY_LEGACY_24 - PAY_REC_24
PROD_LEGACY_25 <- P(payoff, "legacy", "2025")
PROD_REC_25    <- P(payoff, "recommended", "2025")

# The one figure on this page that comes from another paper rather than from a
# CSV here: the V7 flight-list study's own published ADEP+ADES total for the same
# sample. Held as a named constant so the difference below is arithmetic on it
# rather than a second hand-typed number.
V7_PUBLISHED <- 134168

# Where each swept axis peaks, read off the grid rather than described from
# memory. The low-altitude ceiling is the axis that needed the grid extended.
axis_peak <- function(col) {
  a <- aggregate(clean_match_pct ~ get(col), data = sweep1, FUN = max)
  a[[1]][which.max(a[[2]])]
}
LOALT_PEAK <- axis_peak("low_alt_ft")
LOALT_MAX  <- max(sweep1$low_alt_ft)
LOALT_BASE_MAX <- max(D("sweep_recommended_2025_stage1.csv")$low_alt_ft)
```

## Summary {.unnumbered}

**The rule that built every `track_id` OPDI published up to v0.0.2 recovers
about half the flights it should, the cause is one column in a grouping key
rather than a mistuned threshold, and a rule that fixes it has now shipped.**

Before OPDI can derive a flight list it has to decide which state vectors belong
to the same flight. That decision — **segmentation** — is made once, and
everything downstream inherits it: the flight list groups by `track_id`, events
are extracted per track, measurements accumulate from a track's start. What
follows is what the decision is worth, and what had to be repaired in the flight
list before the better rule could be adopted without making the published product
worse.

**The old rule recovers about half the flights it should.** On a three-day
sample of `r num(N_GT_25)` Network Manager flights it matches
**`r pct(CLEAN_LEGACY_25)`** of them one-to-one, meaning one flight, one track,
that track holding nothing else; **`r pct(FRAG_LEGACY_25)`** of flights are split
across more than one track. The failure is almost entirely that splitting, and
almost never the opposite error of running two flights together. Count the
turnaround as part of the flight, which is what a consumer of taxi phases needs,
and the rate falls to **`r pct(GATE_LEGACY_25)`**.

**The identity key is the problem.** The old group key is
`(icao24, callsign)`, so a state vector decoded without a callsign forms a group
— and therefore a track — of its own. Grouping on the airframe alone, and
splitting only when the callsign *genuinely* changes, takes one-to-one matching
to **`r pct(CLEAN_REC_25)`** on 2025 and **`r pct(CLEAN_REC_24)`** on 2024: a
gain of **`r pp(GAIN_25)`** points, and **`r pp(GATE_GAIN_25)`** once the
turnaround is scored as part of the flight. This is the rule that now ships, as
`segmentation.method = "standard"`.

**Its thresholds are the smaller question, and they were swept anyway.** The
shipped rule's three gap parameters were run over `r num(SWEEP_CELLS)`
combinations, every axis widened until the curve turned over inside the grid
rather than at its edge. The settings that ship rank `r num(SHIP_RANK)` and sit
`r sprintf("%.2f", SWEEP_GAIN_25)` points from the best cell anywhere on 2025 —
and on 2024 the ordering of the one axis 2025 wanted to change reverses, so
nothing was re-shipped.

**The flight list could not consume it until one aggregate was fixed, and that
is the most useful finding here.** `flights.py` labelled each flight with the
lexicographically smallest callsign in its track, and a blank sorts first — which
is harmless only while every track carries exactly one callsign, a property the
old group key supplied by accident and any fix to the fragmentation destroys.
Against the pipeline as it then stood the better segmentation therefore made the
published product *worse*, not by naming wrong aerodromes but by producing
flights that could no longer be identified at all. That labelling is now fixed in
production and the numbers on this page are the fixed ones.

With it fixed, the better segmentation costs
**`r num(PAY_SHORT_25)`** ADEP/ADES points on 2025 and **`r num(PAY_SHORT_24)`**
on 2024 — `r pct(100 * PAY_SHORT_25 / PAY_LEGACY_25)` and
`r pct(100 * PAY_SHORT_24 / PAY_LEGACY_24)` of the total, where a *point* is the
flight list's own score for one correctly named aerodrome and a wrong one costs
two — while recovering `r sprintf("%.0f", GAIN_25)` points more flights. The
residual is not accuracy, which is unchanged, and not departures, where the new
rule is slightly ahead. It is arrival coverage, and the one failure mode the new
rule makes worse is merging.

## How to read this page

Sections are ordered as the work was done, not as the conclusion would suggest,
because two of the results only make sense as answers to questions asked in
order.

[What a track is](#sec-problem) defines the object.
[What has already happened to a sample](#sec-pipeline) maps the pipeline steps
that run before any segmentation sees a state vector; skip it if you know them.
[How a segmentation is scored](#sec-metrics) defines every term the later
sections use. [The sample](#sec-sample) says what ground truth is.
[The ladder](#sec-ladder) is one subsection per candidate rule.
[The sweep](#sec-sweep) says what the shipped rule's four parameters are worth.
[The payoff](#sec-payoff) measures the downstream effect on ADEP/ADES and why it
does not track segmentation quality. [What shipped](#sec-defaults) states what
was adopted and what consumers have to know.
[Limitations](#sec-limitations) is where the claims this page does *not* make
are written down.

Every number here is an expression evaluated at render time over a committed CSV
in `data/`, each regenerated by `benchmarks/regenerate_track_v1.py`, which the
page runs before drawing anything and which stamps every output with the code
that produced it in `data/_manifest.json`. Nothing is typed into the prose, which
is what stops the text and the tables drifting apart.

## What a track is, and why the question was open {#sec-problem}

OPDI ingests OpenSky state vectors — position reports, one aircraft at a time,
with no notion of a flight. A **track** is the pipeline's answer to "which of
these belong together". The procedure that produces tracks is a
**segmentation**, and a **track identifier** (`track_id`) is what it stamps on
every sample it decides belongs to the same flight.

The rule this study set out to evaluate — called **`legacy`** throughout, and
the rule behind every dataset OPDI published up to v0.0.2 — is short. Group the
samples by `(icao24, callsign)`, order them by time, and start a new track when
the gap to the previous sample exceeds 30 minutes, or exceeds 15 minutes while
the aircraft is below 5,000 ft. The resulting identifier carries a
`_{year}_{month}` suffix.

Three things can go wrong with such a rule, and they are not symmetric:

- **Fragmentation** — one real flight becomes several tracks. Everything
  downstream sees several short flights where there was one.
- **Merging** — several real flights become one track. The flight list emits one
  flight and the others simply do not exist in the output.
- **Boundary error** — the track is one flight, but starts or ends in the wrong
  place, so first- and last-seen events and cumulative measurements are wrong.

Merging is the most damaging of the three, because a merged flight is
unrecoverable downstream while a fragmented one is at least present in pieces.
The metrics below therefore keep merging and fragmentation separate rather than
blending them into one score. The rule had never been benchmarked before this
study, which is why the study exists.

Segmentation is a versioned choice, not a frozen one: changing the rule changes
`track_id` for everything produced from that release forward, in value and in
shape where the month suffix goes. That decision has now been taken.
`segmentation.method` defaults to `"standard"`, the rule this study recommends,
and `"legacy"` remains selectable and reproduces the identifiers of every dataset
published up to OPDI v0.0.2.

**This paper measures algorithms, not the pipeline.** Every rule below is
expressed in a generic segmentation engine and scored in a harness, so that eight
candidates could be compared without eight production pipeline runs — at this
data volume, the difference between a study and a quarter. What is measured is
therefore a re-implementation of each rule rather than the pipeline's own output.
Version 2 of this study runs the pipeline itself, end to end, on the rule this
one recommends.

## What has already happened to a sample {#sec-pipeline}

*Side information: if you know the pipeline steps, skip to
[the metrics](#sec-metrics).*

Segmentation is step 02 of a chain, and the samples reaching it are not raw
OpenSky rows. Two earlier decisions in particular shape everything measured on
this page: the ingestion filter decides which samples exist at all, and the
**thinning** rule — which sample survives out of each short interval — decides
how densely a flight is described.

```{r pipeline-diagram}
#| echo: false
#| fig-width: 7
#| fig-height: 4.2
#| fig-cap: "The pipeline steps that run before, and after, the decision this page studies. Step 02 is the one under test; everything above it has already happened to a sample by then, and everything below it inherits the result."
# Drawn with ggplot rather than as a mermaid block, deliberately. A mermaid
# diagram written as an executable cell renders in HTML but needs a headless
# browser to rasterise for PDF, and that browser does not start in this
# environment -- the render then hangs with no error. A mermaid fence emitted
# from knitr avoids the hang but arrives without the JavaScript that draws it,
# so it shows as literal text. A plot goes through whatever device Quarto has
# already selected and is correct in both formats.
nodes <- data.frame(
  y = 6:1,
  label = c(
    "step 00 — reference tables (airports, aircraft)",
    "step 01 — ingest OSN state vectors: bbox + thinning",
    "step 02 — tracks: assign track_id",
    "step 02a — cleaning: mask bad values to NULL",
    "step 03 — flight list: ADEP / ADES per track",
    "step 04 — events and measurements per track"),
  under_test = c(FALSE, FALSE, TRUE, FALSE, FALSE, FALSE))
h <- 0.34
ggplot(nodes) +
  geom_rect(aes(xmin = 0, xmax = 1, ymin = y - h, ymax = y + h,
                fill = under_test, colour = under_test, linewidth = under_test)) +
  geom_text(aes(x = 0.5, y = y, label = label, fontface = ifelse(under_test, 2, 1)),
            size = 3.1) +
  geom_segment(data = nodes[nodes$y > 1, ],
               aes(x = 0.5, xend = 0.5, y = y - h, yend = y - 1 + h),
               arrow = arrow(length = unit(1.8, "mm"), type = "closed"),
               colour = "grey40", linewidth = 0.3) +
  annotate("text", x = 1.04, y = 4, hjust = 0, size = 2.9, colour = "grey25",
           label = "this study replaces\nthe rule inside step 02") +
  scale_fill_manual(values = c(`FALSE` = "white", `TRUE` = "grey90")) +
  scale_colour_manual(values = c(`FALSE` = "grey55", `TRUE` = "grey15")) +
  scale_linewidth_manual(values = c(`FALSE` = 0.3, `TRUE` = 0.8)) +
  coord_cartesian(xlim = c(0, 1.5), ylim = c(0.4, 6.6), clip = "off") +
  theme_void() +
  theme(legend.position = "none")
```

```{r pipeline-table}
#| echo: false
steps <- data.frame(
  Step = c("00", "01", "02", "02a", "03", "04"),
  Does = c(
    "Builds reference tables: H3 airport detection zones, HexAero airport ground layouts, OurAirports, the OpenSky aircraft database. Read later; never rewritten by a pipeline run.",
    "Downloads OpenSky state vectors, restricts them to the OPDI bounding box, and thins them to one sample per aircraft per interval.",
    "Groups samples into tracks and stamps `track_id`. **This is the step under study.** Also attaches the H3 index each later join uses.",
    "Masks values that cannot be right — duplicates, out-of-range positions, stale broadcasts, derivative spikes, isolated points — to NULL, keeping the row. Writes a second table rather than editing the first.",
    "One flight per `track_id`: departure and destination aerodrome, and the flight's callsign label.",
    "Milestones per track — take-off, top-of-climb, level segments, landing — and the measurements attached to them."
  ),
  Output = c("reference tables", "`osn_statevectors`", "`osn_tracks`",
             "`osn_tracks_clean`", "`opdi_flight_list`", "events, measurements"),
  check.names = FALSE)
kable(steps, row.names = FALSE, align = "lll",
      col.names = c("Step", "What it does", "Writes"))
```

Every count on this page is conditioned on three parameters of step 01.

```{r ingestion-params}
#| echo: false
ing <- data.frame(
  Parameter = c("bounding box (min lon, min lat, max lon, max lat)",
                "thinning interval",
                "thinning rule"),
  Value = c("(-25.86653, 26.74617, 49.65699, 70.25976)",
            "5 s",
            "`bucket`"),
  Consequence = c(
    "A flight is observed only inside this box. A track that begins at its edge may have entered from outside, which is why an out-of-area label exists at all.",
    "Consecutive samples of one aircraft are at least this far apart, so every gap threshold in this study is measured against a 5 s grid rather than a 1 Hz one.",
    "One row kept per aircraft per interval bin — the last, so a track's final observation survives exactly. The earlier `modulo` rule kept the row whose epoch second divides the interval, and built every dataset published before it."
  ),
  check.names = FALSE)
kable(ing, row.names = FALSE, align = "lll",
      col.names = c("Parameter", "Value", "Why it matters here"))
```

The diagram's ordering is production's, and this study reverses it. In
production step 02 builds tracks and step 02a cleans them; here every candidate
rule is applied to the *cleaned* table, so that all of them see the same input
and a difference between two arms cannot be an artefact of different cleaning.

That has one consequence. Cleaning masks bad values to NULL rather than deleting
rows, and `NULL < 5000` is NULL rather than true, so legacy's low-altitude split
does not fire on a sample whose altitude the cleaner rejected, where once it did.
Every arm is subject to that equally, so no comparison here is affected — but the
identifiers in published data cannot be recomputed from the published clean
table, which [Limitations](#sec-limitations) picks up.

## How a segmentation is scored {#sec-metrics}

*Nothing after this section introduces a metric.*

Comparing a segmentation against ground truth is a **clustering comparison**:
both the algorithm and the truth partition the same state vectors — the
algorithm into tracks, the truth into flights — and the question is how far the
two partitions agree.

### Homogeneity is merging; completeness is fragmentation

Two standard clustering measures map onto the two failure modes exactly, which is
why they are used rather than a score invented for the occasion:

- **Homogeneity** asks whether each track contains samples from only one flight.
  It falls when a track holds more than one flight — so it is the *merging*
  measure.
- **Completeness** asks whether each flight's samples all landed in one track.
  It falls when a flight is spread over several tracks — so it is the
  *fragmentation* measure.

Both are entropy-based and both run from 0 to 1, with 1 meaning perfect
agreement with ground truth on that axis alone. Both are also entropy-weighted
over every sample in the period, and the overwhelming majority of samples sit in
tracks that every candidate gets right: across the eight rules they span
`r sprintf("%.4f", min(arms$homogeneity))`–`r sprintf("%.4f", max(arms$homogeneity))`
and `r sprintf("%.4f", min(arms$completeness))`–`r sprintf("%.4f", max(arms$completeness))`
respectively, where the one-to-one matching rate defined next spans
`r sprintf("%.1f", min(arms$clean_match_pct))` to
`r sprintf("%.1f", max(arms$clean_match_pct))` points. They name the two failure
modes; they do not rank the candidates. That is what the next statistic is for.

### The headline: one-to-one matching

The headline statistic on this page is **`clean_match_pct`**, and it answers the
question a data consumer actually has: *what share of real flights came out as
exactly one track, holding nothing but that flight?*

Every ground-truth flight is placed in exactly one of three mutually exclusive
classes, which sum to 100%:

- **clean** — the flight's samples are in one track, and that track carries no
  other flight;
- **fragmented** — the flight's samples span several tracks, none of which
  carries another flight;
- **merged** — the flight's dominant track also carries at least one other
  flight.

A flight that is both merged and fragmented counts as **merged**, because
merging is the worse failure. A single stray sample in a second track costs a
flight its clean match,[^tolerant] so `clean_match_pct` is a lower bound and is
reported as one.

[^tolerant]: The tolerant alternative — a dominant track covering some fraction
of the flight — was rejected because that fraction is a free parameter chosen
after seeing the results.

### Matching a track to a flight: interval containment {#sec-containment}

Every statistic above needs a rule saying which ground-truth flight a given
state vector belongs to. That rule is **interval containment**, and it is set out
fully because it is where a comparison of this kind is most easily rigged without
anyone noticing.

**A ground-truth flight is an airframe and a span of time.** The airframe is
`icao24` — the same 24-bit address ADS-B broadcasts, which the Network Manager
flight table carries as `AIRCRAFT_ADDRESS`. Two spans are defined, and every arm
is scored against both. The first is the flight's airborne interval, from
take-off to landing:

- where APDF covers the aerodrome, from APDF's measured movement times, the real
  ATOT and ALDT;
- elsewhere, inferred from Network Manager as `AOBT_3 + TAXI_TIME_3` for
  take-off and `ARVT_3` for landing. Both inferences were validated against real
  APDF movements before being relied on (see [The sample](#sec-sample)).

Each row carries which of the two it came from, so a statistic that needs
measured times rather than inferred ones — boundary error does — can restrict
itself to the APDF-backed subset.

**The join is on `icao24` and time overlap, and never on callsign.** This
differs from the ADEP/ADES flight-list studies, which key on callsign, and the
difference is forced rather than stylistic: one of the candidate rules below
removes callsign from track identity entirely, and another splits on it. Joining
on callsign would therefore score each candidate against a key it deliberately
does or does not have, which is prejudging the very thing under comparison.

**Containment, concretely.** A state vector is attributed to the ground-truth
flight of the same `icao24` whose interval holds its timestamp. A track and a
flight are then *matched* when at least one of the track's samples is attributed
to that flight, and the contingency of tracks against flights is what the three
classes above are counted from. Where two intervals of the same airframe touch,
the boundary sample goes to the earlier one, so back-to-back legs cannot
double-count it and inflate the merge rate.

A worked example, one airframe with two rotations in an afternoon:

```{r containment-example}
#| echo: false
ex <- data.frame(
  Flight = c("leg 1", "leg 1", "gap", "leg 2", "leg 2"),
  Interval = c("12:00–13:30", "12:00–13:30", "—", "14:10–15:40", "14:10–15:40"),
  Sample = c("12:05", "13:25", "13:50 (on stand)", "14:15", "15:35"),
  Attributed = c("leg 1", "leg 1", "no flight — outside both intervals",
                 "leg 2", "leg 2"),
  check.names = FALSE)
kable(ex, row.names = FALSE, align = "llll",
      col.names = c("Ground-truth flight", "Its interval", "A sample",
                    "Attributed to"))
```

If one track holds the 12:05 and 15:35 samples, both flights are **merged** into
it. If leg 1's two samples land in different tracks, leg 1 is **fragmented**.
The samples on stand at 13:50 belong to no flight and are scored against
nothing. That is the airborne metric's blind spot, and the second interval,
[below](#sec-gate), is what measures it.

**What the restriction costs.** Ground truth is further restricted to flights
whose *whole* airborne interval lies inside the sampled window: a flight observed
only for its last twenty minutes would score as truncated however well the
segmentation performed, and its boundary error would be a fact about where the
window fell. That argument justifies the restriction without pricing it, so the
price was measured.

```{r containment-table}
#| echo: false
ct <- data.frame(
  period = PERIODS,
  overlapping = sapply(PERIODS, function(p) CT(p, "n_gt_flights")),
  inside      = sapply(PERIODS, function(p) CT(p, "n_wholly_inside")),
  cs          = sapply(PERIODS, function(p) CT(p, "n_clipped_start")),
  ce          = sapply(PERIODS, function(p) CT(p, "n_clipped_end")),
  kept        = sapply(PERIODS, function(p) CT(p, "pct_kept")),
  frac        = sapply(PERIODS, function(p) CT(p, "median_observed_fraction_clipped")),
  check.names = FALSE)
kable(ct, row.names = FALSE, digits = c(0, 0, 0, 0, 0, 2, 3),
      col.names = c("period", "any airborne time in window", "wholly inside",
                    "clipped at start", "clipped at end", "kept %",
                    "median observed fraction of an excluded flight"),
      format.args = list(big.mark = ","))
```

The restriction keeps `r sprintf("%.2f%%", CT("2025", "pct_kept"))` of the 2025
flights with any airborne time in the window and
`r sprintf("%.2f%%", CT("2024", "pct_kept"))` of the 2024 ones, and the exclusion
is near-symmetric between the two window edges, which is what a window artefact
should look like rather than a rule that systematically drops long-haul arrivals.

Nor is it discarding nearly-complete flights on a technicality. The median
excluded flight had
`r sprintf("%.0f%%", 100 * CT("2025", "median_observed_fraction_clipped"))` of its
airborne time inside the window on 2025 and
`r sprintf("%.0f%%", 100 * CT("2024", "median_observed_fraction_clipped"))` on
2024 — about half observed, and unscoreable.

One further gap explains a difference between two counts on this page. Of the
`r num(CT("2025", "n_wholly_inside"))` flights wholly inside the 2025 window,
`r num(N_GT_25)` are scored — the remaining
`r num(CT("2025", "n_wholly_inside") - N_GT_25)` produced no ADS-B sample at all
in the feed and so appear in no contingency. They are absent from the
denominator for the same reason, and no candidate rule is credited or penalised
for them.

### The second interval: gate to gate {#sec-gate}

Containment as defined above matches on the airborne interval, so a sample taken
on stand or on a taxiway is attributed to no flight and leaves every metric where
it found it. That defensible line hides the failure a data consumer meets first.
An OPDI track is supposed to carry the turnaround: first- and last-seen events
fire on it, taxi phases are extracted from it, and every cumulative measurement
starts at its first sample. A rule that shatters a flight into three tracks
between the gate and the runway threshold scores a clean airborne match and hands
over a broken product. So each arm is scored twice over the same partition: once
against the airborne interval, once against the flight's **gate-to-gate
interval**, off-block to in-block.

**Where the block times come from.** `aobt` prefers APDF's measured off-block
time and falls back to Network Manager's `AOBT_3`, so it exists for substantially
every flight. `aibt` has no fallback and cannot have one — `ARVT_3` is a landing
time, not a gate arrival — so the arrival side is APDF-only, and about 62% of
scored flights carry a measured in-block time on each sample (measured in the
ground-truth loader, not read from a CSV here). Where a block time is missing the
interval is extended past the airborne endpoint by the median measured taxi on
that side, over the flights of that period that did measure one. Taking the
modelled half's duration from the measured half is the honest version of a
buffer; ten minutes because it sounds like a taxi is not.

**The gate interval is clamped to contain the airborne one.** APDF carries rows
whose block time falls the wrong side of its own movement time. Unclamped, such a
row gives a gate interval *narrower* than the airborne one, and the wider metric
would drop samples the narrower one kept. Clamped, the two differ only by the
samples the gate interval adds, which is what makes the difference between the
two rates readable as one quantity.

One qualification is exposed rather than filtered away. The sample window bounds
the airborne interval and cannot bound the gate interval, which is attached after
that filter, so a flight can be wholly in-window airborne and still push back
before the window opens. About a third of a percent are, and ground truth carries
the condition per flight as `gate_in_window`. Their taxi samples are clipped by
the day filter, in the flattering direction, since fewer samples is fewer chances
to fragment. Filtering on the column would change which flights the gate metric
covers without saying so, so it is carried and not applied.

The two metrics score `r num(N_GATE_25)` and `r num(N_GT_25)` flights on 2025,
one flight apart, so nothing below is the wider interval scoring a different
population.

### Boundary error, reported signed

Boundary error is the distance in time from a track's first and last sample to
the flight's actual take-off and landing. It is reported **signed**, with the
convention that

> **`off` = track start − take-off, `land` = track end − landing.** A negative
> value means the track began (or ended) *before* the aircraft did.

That is not cosmetic. An absolute error conflates two opposite conditions. A
track that starts *before* take-off has picked up the taxi-out, and that
overhang is wanted rather than tolerated: those samples are the flight's own,
[the gate interval](#sec-gate) scores them, and the taxi phases OPDI publishes
are extracted from them. A track that starts *after* take-off has missed part of
its own flight. `legacy`'s absolute median is about a hundred seconds, a small
tidy number. Signed, its median is
**`r sec(A("legacy", "2025", "off_signed_p50_s"))` s** and its tenth percentile
**`r sec(A("legacy", "2025", "off_signed_p10_s"))` s**: the typical track begins
just after wheels-off, while a minority carry a full quarter-hour of taxi-out.
An absolute median over a two-sided distribution is close to meaningless.

The landing side is one-sided, every percentile positive, median
**`r sec(A("legacy", "2025", "land_signed_p50_s"))` s**. That asymmetry is a
coverage fact rather than an algorithm defect: on-ground ADS-B before departure
reaches a minority of flights, while an arriving aircraft is still powered and
broadcasting to the stand.

The percentiles are computed over the flights whose ground-truth times are
APDF-measured rather than NM-inferred — `r num(N_APDF_25)` of the
`r num(N_GT_25)` scored flights on 2025 — because a boundary error against an
inferred time would partly be measuring the inference.

```{r boundary-percentiles}
#| echo: false
bp <- arms[arms$period == "2025",
           c("arm", "off_signed_p10_s", "off_signed_p50_s", "off_signed_p90_s",
             "land_signed_p10_s", "land_signed_p50_s", "land_signed_p90_s")]
bp <- bp[order(bp$off_signed_p50_s), ]
kable(bp, row.names = FALSE, digits = 0,
      col.names = c("arm", "dep p10", "dep p50", "dep p90",
                    "arr p10", "arr p50", "arr p90"),
      caption = "Signed boundary error in seconds, 2025. Negative means the track began or ended before the aircraft did.")
```

The shape is what the percentiles cannot show.

```{r boundary-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 3.6
#| fig-cap: "Signed boundary error on the 2025 sample, over the flights whose ground-truth times are APDF-measured. **Negative means the track began, or ended, before the aircraft did.** Departure: `legacy` and `recommended` both show a two-sided distribution — a spike just after wheels-off plus a long left tail of tracks that started on the ramp — while `ground_anchored` concentrates hard at zero, which is the arm cutting at the ground event rather than running back into the ramp. Arrival: all three are one-sided and nearly identical, because tracks essentially always run on through taxi-in. The outermost bin on each side is an overflow bin and carries everything beyond half an hour."
b <- bhist[bhist$period == "2025", ]
b$mid  <- (b$bin_lower_s + b$bin_upper_s) / 2
b$edge <- factor(b$edge, levels = c("off", "land"),
                 labels = c("departure: track start − take-off",
                            "arrival: track end − landing"))
b$arm  <- factor(b$arm, levels = c("legacy", "recommended", "ground_anchored"))
ggplot(b, aes(x = mid / 60, y = n, colour = arm)) +
  geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey55") +
  geom_step(direction = "mid", linewidth = 0.5) +
  facet_wrap(~edge, ncol = 2) +
  labs(x = "signed offset (minutes)", y = "flights", colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

## The sample {#sec-sample}

Two three-day samples, `r num(N_GT_25)` and `r num(N_GT_24)` scored ground-truth
flights: 5–7 June 2025 and 5–7 June 2024. Two periods rather than one, because a
single sample cannot distinguish a real effect from one that sample happened to
favour. Every conclusion on this page has to hold on both.

Ground truth is the Network Manager flight table, which covers all of ECAC,
cross-checked against APDF where APDF covers the aerodrome. Take-off is inferred
as `AOBT_3 + TAXI_TIME_3` — off-block plus taxi-out — and landing is taken from
`ARVT_3` directly, which is already a landing time rather than a gate arrival.
Both inferences were validated against real APDF movement times before being
relied on, at median error 0 s with an interquartile range of 17 s and 25 s
respectively; that is what licenses using NM-inferred intervals for matching.
Boundary error is nevertheless reported only over the APDF-measured subset,
which is conservatism rather than necessity.

## The ladder: eight candidate rules {#sec-ladder}

A **ladder** here is a set of candidate segmentation rules run against the same
ground truth, on the same samples, one change at a time. Each candidate is
called an **arm**, borrowing the term from a controlled trial, and each arm is
exactly two things: a **grouping** — which samples are even eligible to share a
track — and a **break predicate** — a test evaluated at each sample in time
order, which when true starts a new track.

Every arm reads one shared cleaned track table, and only `track_id` varies
between them, so a difference between two arms cannot be an artefact of a
different track build. Arms are numbered A0 to A8. A1 was never an arm: it was a
parameter sweep, and [it now sweeps the rule that ships](#sec-sweep) rather than
the one that does not.

Every domain-knowledge arm (A5, A6, A7) keeps a legacy gap rule as a **floor**:
it can add splits to legacy's, and in one deliberate exception replace one of
them, but it cannot split less than a long silence demands. Two reasons, both of
which matter for reading the numbers. The ladder measures one change at a time,
so an arm that also silently removed a legacy split would report a delta mixing
two changes; and an arm whose inputs are missing must not degrade into assigning
an airframe's whole month to one track.

```{r ladder-table}
#| echo: false
t <- arms[arms$period == "2025", c("arm", "clean_match_pct",
                                   "gate_clean_match_pct", "fragmented_pct",
                                   "merged_pct", "n_tracks")]
t$gate_delta <- t$gate_clean_match_pct - t$clean_match_pct
t24 <- arms[arms$period == "2024", c("arm", "clean_match_pct")]
names(t24)[2] <- "clean_2024"
t <- merge(t, t24, by = "arm", sort = FALSE)
t <- t[order(-t$clean_match_pct),
       c("arm", "clean_match_pct", "gate_clean_match_pct", "gate_delta",
         "fragmented_pct", "merged_pct", "n_tracks", "clean_2024")]
kable(t, row.names = FALSE, digits = c(0, 2, 2, 2, 2, 2, 0, 2),
      col.names = c("arm", "clean %", "gate clean %", "gate − air",
                    "fragmented %", "merged %", "tracks", "clean % (2024)"),
      format.args = list(big.mark = ","))
```

The ordering is identical on both periods, and no arm's clean-match rate moves
by more than `r sprintf("%.1f", ARM_SHIFT)` points between them, so what follows
is a property of the algorithms rather than of one sample.

The gate columns are the [second interval](#sec-gate), and they widen the result
rather than repeat it. `recommended` beats `legacy` by
**`r pp(GAIN_25)`** points airborne and **`r pp(GATE_GAIN_25)`** gate to gate on
2025, and by **`r pp(GAIN_24)`** and **`r pp(GATE_GAIN_24)`** on 2024. Scoring
only the airborne leg understated what the shipped change is worth by about
`r sprintf("%.0f", GATE_GAIN_25 - GAIN_25)` points.

What the wider interval detects is a property of the group key, not of the break
predicate:

```{r gate-by-groupkey}
#| echo: false
gk <- data.frame(
  arm = names(GROUPKEY),
  key = as.character(GROUPKEY),
  d25 = sapply(names(GROUPKEY), function(a) GATE_DROP(a, "2025")),
  d24 = sapply(names(GROUPKEY), function(a) GATE_DROP(a, "2024")),
  check.names = FALSE)
gk <- gk[order(gk$d25), ]
kable(gk, row.names = FALSE, digits = c(0, 0, 2, 2),
      col.names = c("arm", "group key", "gate − air (2025)",
                    "gate − air (2024)"),
      caption = "What each arm loses when the turnaround is scored as part of the flight.")
```

Every arm carrying callsign in its group key loses between fourteen and eighteen
points. A callsign goes blank, or flips to the next rotation's value, while the
aircraft is parked — so a key containing it splits the taxi phases into separate
tracks, in exactly the place the airborne metric cannot see. `ground_anchored`
loses as much for a different reason: it cuts through the turnaround by
construction. The arms that group on the airframe alone lose under a point,
because nothing they do fires at the stand. `airport_anchored` is the one arm whose drop moves
between periods, `r pp(GATE_DROP("airport_anchored", "2025"))` against
`r pp(GATE_DROP("airport_anchored", "2024"))`: it groups on the airframe, but its
stationary-at-an-aerodrome break can fire during the dwell the gate interval
scores, and how often it does depends on which aerodromes the sample contains.

Each arm below has the same four parts: what it changes relative to `legacy`,
what it is trying to fix, the rule, and what the measurement said.

### A0 `legacy` — the baseline

**What it changes.** Nothing. This is the rule that built every published
`track_id` up to v0.0.2, and everything else is measured against it.

**The rule.** Group the samples by `(icao24, callsign)` — in production, by a
SHA-2 hash of the two concatenated. Order each group by time. Break when the
gap to the previous sample exceeds 30 minutes, or exceeds 15 minutes while
barometric altitude is below 5,000 ft. The identifier carries a
`_{year}_{month}` suffix.

**What the measurement said.** `r pct(CLEAN_LEGACY_25)` clean,
`r pct(FRAG_LEGACY_25)` fragmented, `r pct(MERGE_LEGACY_25)` merged on 2025;
`r pct(CLEAN_LEGACY_24)` clean on 2024. The failure is overwhelmingly
fragmentation, and almost never merging.

The cause is not the obvious one. It is *not* that callsigns change in
mid-flight; it is that **a sample with no callsign forms a group, and therefore a
track, of its own.** Measured while the arm was implemented, 42.4% of `legacy`'s
tracks carry no callsign at all — they are made of samples the receiver decoded
without one. The group key turns a hole in the data into a separate flight, and
that single mechanism is what the rest of this ladder chases.

### A2 `no_month_suffix` — untestable on this sample

**What it changes.** Removes the `_{year}_{month}` suffix from the identifier.
Grouping and break predicate are `legacy`'s, untouched.

**What it is trying to fix.** The suffix splits any flight airborne at midnight
on the first of a month, by construction and regardless of the data. That is a
correctness defect rather than a threshold, which is why no amount of parameter
sweeping can reach it.

**What the measurement said.** Identical to A0 in every digit, on both periods,
which is correct behaviour rather than a bug: both samples sit inside a single
month, so the suffix is constant. **This study cannot evaluate the arm at all**,
and reports it as untested rather than as a null result. Testing it needs a
sample spanning a month boundary, about two days of cluster time.

### A3 `traffic_style` — the reference library's rule, outside its own preprocessing

**What it changes.** Replaces legacy's two hard thresholds with the semantics of
`Flight.split` in the *traffic* library, the widely used open-source reference
for this kind of work.

**What it is trying to fix.** Nothing specific to OPDI. The question this arm
answers is whether the established reference implementation, applied to this
feed, would simply have done better.

**The rule.** Group on `(icao24, callsign)`. Break when the gap exceeds a single
threshold — *traffic*'s documented default is ten minutes — *unless* a predicate
says the candidate gap is not really a break. The predicate used here is the
aviation-sensible inverse of *traffic*'s own documented example: do not split
when the samples on both sides of the gap are above the low-altitude ceiling,
because a reception hole in the cruise is a coverage gap, not a landing.

**What the measurement said.**
`r pct(A("traffic_style", "2025", "clean_match_pct"))` clean, below `legacy`,
and the worst merging of any arm at
`r pct(A("traffic_style", "2025", "merged_pct"))` — about
`r sprintf("%.0f", A("traffic_style", "2025", "merged_pct") / MERGE_LEGACY_25)`
times legacy's. Two things are behind that, and only the second is the reason it
lost.

#### The predicate reads a column OPDI has taken away

`Flight.split` fills nothing: it cuts on gaps and hands back the pieces. It is
one method of a class whose neighbours do fill, though — `Flight.filter` defaults
to `bfill().ffill()` and `Flight.resample` to interpolation — and a *traffic*
pipeline conventionally runs one of them first. The predicate is written for a
frame with no holes in it.

OPDI hands it a frame that is almost all hole in the one place the predicate
looks. The test asks whether the samples either side of a candidate gap are above
the low-altitude ceiling, and `NULL > 10000` evaluates to NULL rather than to
false, so a boundary with a masked altitude on either side is one the arm cannot
decide.

```{r traffic-fill}
#| echo: false
tf <- data.frame(
  period = PERIODS,
  gaps   = sapply(PERIODS, function(p) TF(p, "n_gaps")),
  clean  = sapply(PERIODS, function(p) TF(p, "null_pct")),
  raw    = sapply(PERIODS, function(p) TF(p, "raw_null_pct")),
  nogap  = sapply(PERIODS, function(p) TF(p, "n_no_gap_turnarounds")),
  check.names = FALSE)
kable(tf, row.names = FALSE, digits = c(0, 0, 2, 2, 0),
      col.names = c("period", "candidate gaps",
                    "undecidable, cleaned %", "undecidable, raw %",
                    "turnarounds with no gap at all"),
      format.args = list(big.mark = ","))
```

On the cleaned table the arm reads, `r sprintf("%.2f%%", TF("2025", "null_pct"))`
of candidate gaps on 2025 and `r sprintf("%.2f%%", TF("2024", "null_pct"))` on
2024 are undecidable. On the raw table the same count is
`r sprintf("%.2f%%", TF("2025", "raw_null_pct"))` and
`r sprintf("%.2f%%", TF("2024", "raw_null_pct"))` — a third rather than
everything. **The missingness is not in the feed; OPDI's cleaning makes it**, and
one rule in particular. `mask_stale_broadcasts` masks an altitude that exactly
repeats the previous sample's, because ADS-B sends position and velocity in
separate message types and an unchanged consecutive value is a repeat rather than
a measurement. A coverage fade loses fresh altitude before it loses messages
altogether, so the last sample before a silence is precisely the sample most
likely to be a repeat: on 2025,
`r sprintf("%.2f%%", TF("2025", "pregap_null_pct"))` of pre-gap samples are
already NULL in the raw feed and
`r sprintf("%.1f%%", TF_REPEAT_OF_LIVE("2025"))` of the rest repeat their
predecessor, against `r sprintf("%.2f%%", TF("2025", "other_null_pct"))` and
`r sprintf("%.2f%%", TF("2025", "other_repeat_pct"))` away from a boundary. The
masking is correct and the predicate is correct. They cannot be used together.

#### That is not why the arm lost

A3's failure is merging, and merging is under-splitting. Filling the altitude
column could only let the predicate *decline* to split; nothing it does can
manufacture a split that was not offered. And the splits A3 needs were never
offered: `r num(TF("2025", "n_no_gap_turnarounds"))` turnarounds on 2025 and
`r num(TF("2024", "n_no_gap_turnarounds"))` on 2024 produce no candidate gap at
all, because the aircraft broadcast straight through the stand. No fill reaches
them. Legacy's second rule — the shorter threshold below 5,000 ft — is what
catches them, and a single gap threshold has no equivalent.

So the conclusion is narrower than "the reference approach does not transfer",
and neither half of it is a defect in the reference. The rule is used outside the
preprocessing it assumes, and even inside that preprocessing one threshold cannot
see a turnaround that never went quiet.

### A4 `airframe_only` — delete callsign from the group key

**What it changes.** One column. Group on `icao24` alone; callsign becomes an
attribute of a sample rather than part of the track's identity. The break
predicate is `legacy`'s, untouched.

**What it is trying to fix.** A0's diagnosis directly: if a blank callsign is
what starts a spurious track, removing callsign from the key stops it. Identity
belongs to the airframe; the callsign is what the airframe happened to be
broadcasting at the time.

**What the measurement said.**
`r pct(A("airframe_only", "2025", "clean_match_pct"))` clean, with fragmentation
down from `r pct(FRAG_LEGACY_25)` to
`r pct(A("airframe_only", "2025", "fragmented_pct"))`. This is the single
largest movement anywhere in the study, and it comes from deleting one column.

**But it is not free**, and the cost is exactly the failure legacy did not have:
merging rises from `r pct(MERGE_LEGACY_25)` to `r pct(MERGE_AIR_25)`. Without
callsign in the key, two consecutive flights of the same airframe with no
sufficient gap between them become one track. Measured during implementation,
3.3% of this arm's tracks contain more than one genuine callsign.

### A5 `ground_anchored` — split at a turnaround, using the on-ground flag

**What it changes.** Groups on `icao24`, like A4, and *adds* a break condition
to legacy's gap rules rather than replacing them.

**What it is trying to fix.** Legacy's "below 5,000 ft" test is a proxy for "the
aircraft is on the ground somewhere", used because barometric altitude was all
the rule had. The OSN schema carries an `on_ground` flag that step 02 reads
nothing of, and this arm uses it.

**The rule.** A break is declared at the *first airborne sample after* a run of
consecutive on-ground samples lasting at least the ground-dwell threshold. The
dwell requirement is what separates a turnaround from a touch-and-go. Placing
the break at the first airborne sample, rather than at the start of the ground
run, means the departing leg begins at the rotation instead of inheriting the
previous arrival's taxi-in. Legacy's two gap rules are kept as a floor, because
an aircraft that stops broadcasting entirely still has to be split somehow.

**What the measurement said.**
`r pct(A("ground_anchored", "2025", "clean_match_pct"))` clean — well above
`legacy`, well below A4. Its distinguishing result is not the headline but the
boundary error: its departure tenth percentile is
`r sec(A("ground_anchored", "2025", "off_signed_p10_s"))` s against legacy's
`r sec(A("legacy", "2025", "off_signed_p10_s"))` s, and the figure in
[the metrics section](#sec-metrics) shows the whole left tail collapsing. The arm
cuts at the ground event rather than running back into the ramp.

Scored [gate to gate](#sec-gate) that same behaviour is its worst result:
`r pct(A("ground_anchored", "2025", "gate_clean_match_pct"))`, a drop of
`r pp(GATE_DROP("ground_anchored", "2025"))` points, the largest of any arm on
that sample. Cutting at the ground event means cutting through the turnaround, so
one flight's taxi-in and taxi-out land in different tracks. The arm locates
rotation precisely and delivers a flight poorly.

### A6 `airport_anchored` — split only where an aerodrome actually is

**What it changes.** Groups on `icao24`, and replaces legacy's low-altitude gap
rule with real airport geometry. It is the one arm that takes only *half* of
legacy as its floor.

**What it is trying to fix.** Two errors that no altitude threshold can reach.
A split legacy *misses*: an aircraft parked at a 6,000 ft aerodrome never
satisfies "below 5,000 ft", so the turnaround is invisible and two flights merge.
And a split legacy *invents*: a long slow low transit in the middle of nowhere
satisfies the low-altitude rule and is split, when nothing happened at all.

**The rule.** A sample is *stationary at an aerodrome* when it is inside a known
airport detection zone, below a turnaround height measured **above field
elevation** rather than above sea level, and slower than a turnaround speed. A
break is then declared two ways, and the arm fires on either: at the sample that
ends a reception gap, when the previous sample was stationary at an aerodrome
and at least the dwell threshold has passed; or at the first sample that moves
again after a continuous stationary run lasting at least the dwell threshold.
Where the airport join finds nothing, "near an aerodrome" is false and the arm
cannot fire, falling back to its floor. Because the arm's entire thesis is that
legacy's low-altitude rule is the wrong test, that rule is deliberately *not*
inherited — only the general "a long enough silence is a new track" rule is,
which no arm disputes and which guarantees no airframe can end up with a
month-long track.

**What the measurement said.**
`r pct(A("airport_anchored", "2025", "clean_match_pct"))` clean on 2025 and
`r pct(A("airport_anchored", "2024", "clean_match_pct"))` on 2024 — the closest
any domain-knowledge arm gets to A4, and still short of it. Real geometry is
worth a lot; it is worth less than not fragmenting on blanks.

### A7 `vertical_profile` — a negative result, correctly attributed

**What it changes.** Groups on `icao24`, and adds a break based purely on the
altitude trace, with legacy's gap rules as a floor.

**What it is trying to fix.** The case where the transponder never reports
`on_ground` at all. If a flight is one climb–cruise–descent cycle, then a
descent to circuit height followed by a climb back through it is a new sortie,
whether or not there was a reception gap and whether or not ground contact was
ever broadcast.

**The rule.** Both halves are required. First, *the descent*: some earlier
sample of the group crossed the floor downwards — the previous sample above it,
this one at or below. Second, *the climb back*: this sample is above the floor
and the previous one is not. The break is placed at the upward crossing, so the
departing leg begins there. The floor is measured **above field elevation**;
comparing against sea level made the arm degenerate to bare legacy at any
aerodrome above the floor's own height, since a parked aircraft there already
reads above it and no crossing is ever seen.

**What the measurement said.**
`r pct(A("vertical_profile", "2025", "clean_match_pct"))` clean, last of the eight
on both periods. An earlier implementation split on *any* upward crossing, which
is not the idea, so its poor score could have been the idea failing or the code
not being the idea. Implemented properly, on height above field, it scores
*worse* still — which is what makes this a negative result rather than an
unexplained number. The mechanism is visible in the boundary error: every
percentile of its signed departure offset is positive
(tenth percentile `r sec(A("vertical_profile", "2025", "off_signed_p10_s"))` s),
meaning it systematically starts tracks late, having sliced off each departure's
initial climb.

Scored [gate to gate](#sec-gate) it loses only
`r pp(GATE_DROP("vertical_profile", "2025"))` points on 2025 and
`r pp(GATE_DROP("vertical_profile", "2024"))` on 2024, among the smallest drops
of any arm. An earlier draft of this study predicted the reverse — that an arm
this poor airborne would be worse again once the turnaround counted — and the
measurement refuses it. A7 groups on `icao24`, so no blank callsign at the stand
can split its ground samples, and its own break fires at a vertical crossing,
which happens in the air. It is already last of the eight before any taxi sample
is scored, and the taxi adds almost nothing to that. A7 is a bad rule for a
reason the gate interval cannot see, which is worth knowing about the gate
interval as well as about A7.

### A8 `recommended` — the two fixes, kept separate {#sec-a8}

This is the rule that now ships as `segmentation.method = "standard"`. **Its two
changes are independent, they fix different failures, and the arm only works
because they are kept separate**, so they are set out one at a time before any
number.

**What it changes.** The group key loses its callsign column, and the break
predicate gains a callsign test. Legacy's two gap rules are kept underneath both,
unchanged. Nothing about the gap behaviour of published tracks changes; what
changes is which samples may share a track, and where a shared group is cut.

#### Fix one: take callsign out of the group key

*The failure it addresses* is A0's: a sample with no callsign forms its own
group, so a hole in the data becomes a separate flight. That is where legacy's
fragmentation comes from, and A4 showed that deleting the column removes it.

*The change* is exactly A4's: group on `icao24` alone. A blank or flickering
callsign can no longer start a track, because callsign is no longer part of what
makes a track a track.

*What it costs on its own* is A4's cost: two consecutive flights of one airframe
now have nothing to separate them but a gap, and where the aircraft kept
broadcasting through the turnaround there is no sufficient gap. They merge.

#### Fix two: break when the callsign genuinely changes

*The failure it addresses* is the one fix one creates. Legacy separated those
two flights for the wrong reason — callsign was in the key, so any change of it
started a new group — but it did separate them.

*The change* restores that separation as an explicit break rather than as a side
effect of the key. Walking an airframe's samples in time order, carry forward
the last **non-blank** callsign seen. Break at a sample when that sample's
callsign is itself non-blank, the carried-forward one is non-blank, and the two
differ. Comparing the raw callsign column instead would break at every blank
sample and reproduce legacy's fragmentation exactly; it is carrying the last
*real* value across the blanks that makes the test mean "the callsign changed"
rather than "the callsign was missing". Callsigns are compared trimmed, because
ADS-B pads them to eight characters and `ABC123  ` is not a different flight
from `ABC123`.

*Why the two are genuinely independent:* one is about which samples may share a
track, the other about where a shared group is cut. The evidence that they are
is that the combination beats **both** of the arms it draws on, rather than
landing between them — which is what would happen if they were two attempts at
the same thing.

#### The subtlety: the lookback must be bounded

The break predicate runs over the whole airframe's samples and has no notion of
the track boundaries it is in the middle of deciding, so "the last non-blank
callsign" is by default the last one *anywhere earlier in the month*. That
reaches back across a gap break into the previous flight, and the next flight —
whose first samples happen to be blank — is compared against a callsign belonging
to a flight that already ended, and split in its own middle.

The fix bounds the lookback: the carried-forward callsign counts only if it is no
older than the general gap threshold. That bound is the right one rather than an
arbitrary one, because a silence longer than the gap threshold *is already a
break*, so there is nothing legitimate to carry across it.

It is not a detail. Measured during implementation, the unbounded version
fragmented 37.75% of flights, where `airframe_only` — which does no callsign
splitting at all — fragments
`r pct(A("airframe_only", "2025", "fragmented_pct"))`, and where only 3.2% of
tracks contain a genuine callsign transition to fire on. The mismatch between
those two is what exposed the bug: a rule reaching one track in thirty cannot
legitimately cost thirty points of fragmentation.

#### What the measurement said

`r pct(CLEAN_REC_25)` clean on 2025 and `r pct(CLEAN_REC_24)` on 2024 — the best
of the eight on both periods, and `r pp(GAIN_25)` and `r pp(GAIN_24)` points
above `legacy`. Fragmentation falls from `r pct(FRAG_LEGACY_25)` to
`r pct(FRAG_REC_25)`.

The result that settles it is merging: `r pct(MERGE_REC_25)`, against
`airframe_only`'s `r pct(MERGE_AIR_25)`. Fix two recovers almost all of the
merging fix one introduces, but not all: the remainder is still above `legacy`'s
`r pct(MERGE_LEGACY_25)`. That residual is small, real, and where the downstream
cost measured in [the payoff](#sec-payoff) comes from.

## Tuning what actually ships {#sec-sweep}

A **sweep** is a re-run of one rule at many parameter settings; a **cell** is one
such setting. Earlier drafts of this study swept `legacy`, which answered a
question about a rule nobody now runs. This sweep is of `recommended`, at the
same four parameters production exposes:

- **`gap_minutes`** — a gap longer than this starts a new track, at any
  altitude.
- **`low_alt_gap_minutes`** — a shorter gap that also starts a new track, but
  only when the aircraft is low.
- **`low_alt_ft`** — the altitude below which the shorter gap applies.
- **`callsign_lookback_minutes`** — how far back the carried-forward non-blank
  callsign may reach before it stops counting. This axis exists only on
  `recommended`, and only since it was made settable; before that it was pinned
  to `gap_minutes` in the code.

### The grid, and where each axis peaks

Stage 1 evaluated `r num(SWEEP_CELLS)` cells across the first three axes on one
day, with the lookback left following `gap_minutes`. Locating an optimum and
quoting it are different claims, so the location is done on one day and the
quoting on three.

End to end the grid spans `r sprintf("%.2f", SWEEP_SPREAD1)` points of clean
matching, from `r sprintf("%.2f", SWEEP_MIN1)`% to
`r sprintf("%.2f", SWEEP_MAX1)`%. The parameters are not decorative: a rule
handed the worst corner of its own grid gives up
`r sprintf("%.0f%%", 100 * SWEEP_SPREAD1 / GAIN_25)` of what the identity fix
bought.

```{r sweep-axes}
#| echo: false
by_axis <- function(col) {
  agg <- aggregate(clean_match_pct ~ get(col), data = sweep1, FUN = max)
  names(agg) <- c("value", "best")
  agg[order(agg$value), ]
}
axes <- list(gap_minutes = "gap_minutes (min)",
             low_alt_gap_minutes = "low_alt_gap_minutes (min)",
             low_alt_ft = "low_alt_ft (ft)")
prof <- do.call(rbind, lapply(names(axes), function(a) {
  d <- by_axis(a)
  data.frame(Axis = axes[[a]], Value = d$value, Best = d$best,
             Peak = ifelse(d$best == max(d$best), "*", ""),
             check.names = FALSE)
}))
kable(prof, row.names = FALSE, digits = c(0, 0, 3, 0),
      col.names = c("axis", "value", "best clean % at this value", ""),
      format.args = list(big.mark = ","),
      caption = "For each value of each parameter, the best clean-match rate achieved anywhere in the grid at that value. A star marks each axis's peak.")
```

Read down each block: every axis rises to a peak and falls again on the far
side. That is what an **interior optimum** means, and it is the condition for
calling a result an optimum at all. An optimum sitting on the edge of a grid is
not an optimum — it is where you stopped looking.

### The extension, and why it was necessary

`low_alt_ft` did not start out interior. The first grid ran it to
`r num(LOALT_BASE_MAX)` ft, and its best value was the largest tried with the
curve still rising, which is the edge case above. This pipeline's own history
contains a sweep that stopped early and was read as "higher is always better", so
the grid was extended rather than the result reported.

```{r sweep-extension}
#| echo: false
base <- D("sweep_recommended_2025_stage1.csv")
lo <- function(d) {
  a <- aggregate(clean_match_pct ~ low_alt_ft, data = d, FUN = max)
  a[order(a$low_alt_ft), ]
}
b <- lo(base); e <- lo(sweep1)
tab <- data.frame(low_alt_ft = e$low_alt_ft,
                  grid = ifelse(e$low_alt_ft %in% b$low_alt_ft,
                                "original", "extension"),
                  best = e$clean_match_pct, check.names = FALSE)
kable(tab, row.names = FALSE, digits = c(0, 0, 3),
      col.names = c("low_alt_ft", "grid", "best clean %"),
      format.args = list(big.mark = ","),
      caption = "The low-altitude ceiling before and after the grid was extended. On the original grid the best value was the largest tried; extended, the curve turns over.")
```

Extended, the ceiling peaks at `r num(LOALT_PEAK)` ft and falls away to
`r num(LOALT_MAX)` — as it must, since past cruise altitude every sample counts
as "low", the two-threshold rule collapses into one, and
`low_alt_gap_minutes` simply replaces `gap_minutes` everywhere.

### The lookback axis

`callsign_lookback_minutes` is the axis unique to `recommended`, and until
recently it was not an axis at all: the implementation reached back exactly
`gap_minutes` because [that bound is the defensible one](#sec-a8) — a silence
longer than the gap threshold is already a break, so there is nothing legitimate
to carry across it. The argument is sound and it was never measured. Making the
parameter settable turns it into a question with an answer.

It was swept twice, at two different cells, because a lookback that is best at
one gap threshold need not be best at another.

```{r sweep-lookback}
#| echo: false
lb <- rbind(
  transform(D("sweep_recommended_2025_stage2.csv"), cell = "gap 40 / low 20 / 10,000 ft"),
  transform(D("sweep_recommended_2025_stage2b_shipped.csv"), cell = "gap 30 / low 15 / 5,000 ft (shipped)"))
lb <- lb[order(lb$cell, lb$callsign_lookback_minutes),
         c("cell", "callsign_lookback_minutes", "clean_match_pct")]
kable(lb, row.names = FALSE, digits = c(0, 0, 3),
      col.names = c("cell", "callsign_lookback_minutes", "clean %"),
      caption = "The lookback axis at two cells of the three-parameter grid, on the full 2025 sample.")
```

Both curves peak at, or one step above, their own `gap_minutes`:
`r num(LB_PEAK_TUNED)` minutes at the tuned cell's gap of
`r num(LB_GAP_TUNED)`, and `r num(LB_PEAK_SHIP)` minutes at the shipped cell's
gap of `r num(LB_GAP_SHIP)`. Decoupling the two buys
`r sprintf("%.3f", LB_GAIN_SHIP)` points at the shipped cell and
`r sprintf("%.3f", LB_GAIN_TUNED)` at the tuned one. So the coupling that was in
the code by accident was very nearly the right choice, and
`callsign_lookback_minutes` ships as `None`, meaning *follow `gap_minutes`*. The
argument for the bound now has a number beside it.

### What the sweep is worth against what shipped

Stage 2 and the 2024 confirmation re-measured the interesting cells on the full
three-day samples, so a winner is chosen out of the day it was tuned on.

The shipped setting — gap `r num(SHIP_GAP)`, low-altitude gap
`r num(SHIP_LOWGAP)`, ceiling `r num(SHIP_LOWFT)` ft — ranks
`r num(SHIP_RANK)` of `r num(SWEEP_CELLS)` on the 2025 grid, and the best cell
anywhere in it beats the shipped one by
**`r sprintf("%.2f", SWEEP_GAIN_25)`** points. **The sweep does not beat the
defaults by enough to move them.** That is the result, and a sweep confirming a
setting is as much a measurement as a sweep overturning one.

Then 2024, which is why nothing was re-shipped.

```{r sweep-2024}
#| echo: false
s3 <- D("sweep_recommended_2024_stage3.csv")
s3$shipped <- ifelse(s3$gap_minutes == SHIP_GAP &
                     s3$low_alt_gap_minutes == SHIP_LOWGAP &
                     s3$low_alt_ft == SHIP_LOWFT, "shipped", "")
s3 <- s3[order(-s3$clean_match_pct),
         c("gap_minutes", "low_alt_gap_minutes", "low_alt_ft",
           "clean_match_pct", "shipped")]
kable(s3, row.names = FALSE, digits = c(0, 0, 0, 3, 0),
      col.names = c("gap_minutes", "low_alt_gap_minutes", "low_alt_ft",
                    "clean %", ""),
      format.args = list(big.mark = ","),
      caption = "2024, the eight cells spanning the choice 2025 wanted to change. Every cell at gap 30 beats every cell at gap 40 — the reverse of 2025's ordering.")
```

On 2024 the `gap_minutes` ordering **reverses**: every cell at 30 minutes beats
every cell at 40, and the shipped cell scores
`r sprintf("%.2f", SHIP_24)`% against `r sprintf("%.2f", BEST_24)`% for the best
of the eight — `r sprintf("%.2f", BEST_24 - SHIP_24)` points, on the same
`low_alt_gap_minutes` and a different ceiling. Retuning to 2025's preferred gap
would have cost 2024 about `r sprintf("%.2f", GAP30_24 - GAP40_24)` points.

That is the closing argument. The gain 2025 offers is
`r sprintf("%.2f", SWEEP_GAIN_25)` points and does not survive the second
sample, where the same change costs `r sprintf("%.2f", GAP30_24 - GAP40_24)`.
The shipped setting is within a fifth of a point on the period that prefers
something else, and preferred outright on the period that does not.


## What a better segmentation buys ADEP/ADES {#sec-payoff}

Each arm's assignment is fed to the V7 flight list with **only `track_id`
varying** — same input table, same detection configuration, same reference
data — so any difference in the result is attributable to the segmentation.

Three terms, because the flight list scores differently from the ladder:

- **coverage** — the share of ground-truth flights the flight list gives *any*
  aerodrome for, in that role;
- **accuracy** — of those it does answer, the share it gets right;
- **score** — `correct − k × wrong` at `k = 2`, so a wrong answer costs twice
  what a right one earns and a silence costs nothing. `ADEP + ADES` is the sum
  over both roles, and it is the single number the tables below rank on.

The ground truth here is the flight-list studies' own, keyed on callsign and
day, which is why its denominator (`r num(payoff$n_ground_truth[1])` flights on
2025) differs from the ladder's. The two are not interchangeable.

As a control, `legacy` run through this harness scores `r num(PROD_LEGACY_25)`
against the V7 flight-list study's own published `r num(V7_PUBLISHED)` for the
same sample — a difference of `r num(abs(PROD_LEGACY_25 - V7_PUBLISHED))`, or
`r sprintf("%.3f%%", 100 * abs(PROD_LEGACY_25 - V7_PUBLISHED) / V7_PUBLISHED)`.
That residual is itself informative and is explained in
[Limitations](#sec-limitations).

### The flight-list configuration these numbers were produced at {#sec-detection}

Every arm is scored at one fixed ADEP/ADES configuration. Naming it matters,
because "the V7 recommended configuration" is a moving target and these numbers
belong to one point on it. The table is read from the run's own recorded
parameters rather than typed, so it cannot drift from what was executed.

```{r detection-params}
#| echo: false
cfg <- payoff[payoff$period == "2025", ]
fields <- c("adep_mode", "ades_mode",
            "endpoint_radius_nm", "endpoint_height_ft",
            "endpoint_sched_penalty_nm", "endpoint_candidate_radius_nm",
            "trend_max_fl", "trend_radius_nm", "trend_radius_exact",
            "trend_rank_by", "trend_vote_margin", "trend_sched_penalty_nm",
            "trend_smooth_half_window", "trend_smooth_before_cut",
            "trend_bearing_tiebreak_nm", "trend_ooa", "k")
# Every arm must have been run at the same configuration, or the comparison is
# not a comparison. Asserted rather than assumed: a silent disagreement here
# would attribute a detection change to a segmentation.
stopifnot(all(sapply(fields, function(f) length(unique(cfg[[f]])) == 1)))
meaning <- c(
  adep_mode = "Which algorithm names the departure aerodrome.",
  ades_mode = "Which algorithm names the destination aerodrome.",
  endpoint_radius_nm = "`endpoint`: how close the track's first or last fix must be to an aerodrome.",
  endpoint_height_ft = "`endpoint`: how high that fix may be, above field elevation.",
  endpoint_sched_penalty_nm = "Distance added to an aerodrome without scheduled service, so a military field must be clearly nearest.",
  endpoint_candidate_radius_nm = "How far the cached candidate table reaches. Wider than any detection radius, so the radius stays sweepable.",
  trend_max_fl = "`trend`: only samples below this flight level vote.",
  trend_radius_nm = "`trend`: how close a sample must be to an aerodrome to vote for it.",
  trend_radius_exact = "Cut that radius on exact great-circle distance rather than on H3 band.",
  trend_rank_by = "How `trend` chooses among candidates: exact distance, not H3 ring count.",
  trend_vote_margin = "How far one direction must beat the other before the vote counts. Zero is a simple majority.",
  trend_sched_penalty_nm = "The scheduled-service penalty, on `trend`'s side.",
  trend_smooth_half_window = "Half-width in samples of the rolling mean over barometric altitude.",
  trend_smooth_before_cut = "Smooth altitude before applying the flight-level cut, not after.",
  trend_bearing_tiebreak_nm = "Within this distance band, candidates are separated by course alignment rather than distance.",
  trend_ooa = "Whether `trend` may emit the out-of-area marker. Off: on arrivals it is right barely half the time.",
  k = "The scoring exchange rate: one wrong answer costs `k` right ones.")
dp <- data.frame(
  Parameter = paste0("`", fields, "`"),
  Value = sapply(fields, function(f) as.character(cfg[[f]][1])),
  Meaning = meaning[fields], check.names = FALSE)
kable(dp, row.names = FALSE, align = "lll",
      col.names = c("parameter", "value", "what it does"),
      caption = "The ADEP/ADES configuration every arm in this section was scored at, identical across all eight arms and both periods.")
```

**Cross-checked, not copied.** Every value above was compared field by field
against `DetectionConfig()` in the checkout these runs were made from, and they
agree. Compared against the *v6.2* study's shipped configuration, one difference
is real and is reported rather than smoothed over:

> v6.2 gives `trend`'s altitude cut a **datum**, defaulting to field elevation:
> the cut becomes `trend_max_height_ft` above the field rather than
> `trend_max_fl` above mean sea level, and `trend_max_fl` is retained but inert.
> That change landed on `main` **after** this study's base commit and is not
> present in the checkout these payoff numbers were produced from. The runs
> above therefore cut on FL60 above sea level.

This does not affect any comparison on this page: every arm is scored at the same
configuration, so the datum is held constant across all of them. It does mean
these absolute ADEP/ADES totals are not comparable with figures published by v6.2
or later, and it is the reason the payoff figures on this page are **not
regenerated at this revision**. The V7 ladder that produced them cannot currently
be re-run — its last rung no longer matches the shipped `DetectionConfig()` — so
the choice is between quoting a measurement made at a configuration that has
since moved and quoting one made at a configuration that answers a different
question. The first is quoted, and labelled. [Limitations](#sec-limitations)
carries it as a standing caveat rather than a footnote here.

**Everything else on this page did move at this revision, and not because of any
segmentation change.** Ground truth improved underneath it: the APDF join now
keys on the Samad flight identifier and the arrival-time correction is applied,
which took the APDF-measured subset from about 22,000 flights to
`r num(N_APDF_25)` on 2025. Every rate computed against it shifted slightly — the
`recommended` − `legacy` gap by under 0.6 points, changing no ordering and no
conclusion. The payoff figures alone are frozen at the older ground truth as well
as at the older datum, which is what "retained historical" means where it appears
below.

### Segmentation quality and downstream score do not line up

This subsection is the one place all eight arms still appear. It reads the
eight-arm CSVs already in `data/`, recomputes nothing, and is retained because it
is the evidence for not optimising `clean_match_pct` directly — a historical
measurement kept for what it rules out, not re-run.

```{r payoff-table}
#| echo: false
p <- payoff[payoff$period == "2025", ]
p$total <- p$adep_score + p$ades_score
p$vs_legacy <- p$total - PROD_LEGACY_25
p$clean <- sapply(p$arm, function(a) A(a, "2025", "clean_match_pct"))
p <- p[order(-p$total), ]
out <- data.frame(
  arm = p$arm,
  clean = p$clean,
  total = p$total,
  vs = p$vs_legacy,
  adep_cov = 100 * p$adep_coverage,
  adep_acc = 100 * p$adep_accuracy,
  ades_cov = 100 * p$ades_coverage,
  ades_acc = 100 * p$ades_accuracy,
  check.names = FALSE)
kable(out, row.names = FALSE, digits = c(0, 2, 0, 0, 2, 2, 2, 2),
      col.names = c("arm", "clean % (ladder)", "ADEP+ADES", "vs legacy",
                    "ADEP cov %", "ADEP acc %", "ADES cov %", "ADES acc %"),
      format.args = list(big.mark = ","),
      caption = "2025. Ranked on ADEP+ADES score. The second column is the same arm's clean-match rate from the ladder, so the two rankings can be read against each other in one place.")
```

The two rankings disagree. `recommended` is first on the ladder and third here;
`vertical_profile` is last on the ladder and has the **highest ADEP score of any
arm**. A reader could take that to mean better segmentation makes the product
worse. It does not, and the table says why in its last four columns.

**Accuracy barely moves.** Every arm lands within a couple of points on both
roles. No arm is naming *wrong* aerodromes at a materially different rate.

**Coverage moves, and almost all of the movement is on the arrival side.** ADEP
coverage spans a narrow band across all eight arms; ADES coverage runs from
`r sprintf("%.1f", min(100 * p$ades_coverage))`% to
`r sprintf("%.1f", max(100 * p$ades_coverage))`%. So the whole ranking is
decided by how often the arrival algorithm can answer at all.

Why the arrival side is the sensitive one is not settled here. What the data says
is consistent across both periods: the two arms with the worst arrival coverage
by a wide margin, `ground_anchored` and `vertical_profile`, are the two whose
breaks are placed at *low-altitude events*, while `legacy` and `recommended`
place theirs at reception gaps and land close together. The destination algorithm
reads its evidence from the low-altitude part of a track, so an arm that cuts
there is cutting into the evidence. Isolating the mechanism would need a
per-flight diagnosis this study did not run, so it is recorded as a hypothesis.

The lesson does not depend on that hypothesis: **an arm that is good at
recovering flights can still be bad at preserving the evidence a downstream
detector needs**, and the two have to be measured separately.

### The labelling bug this study found, and why it is now history {#sec-labelling}

The measurement above could not be made until a defect in the flight list was
fixed. It is recorded because anyone meeting an OPDI flight list built before the
fix needs to recognise it.

**What it was.** The flight list built one flight per `track_id` and labelled it
with `F.min("flight_id")` — the lexicographically smallest callsign in the
track. A blank sorts first.

**Why it was harmless, until it was not.** While callsign is part of the track's
group key, every track carries exactly one callsign by construction, so that
aggregate can never lose anything. It is correct only because of a property
`legacy` supplies *by accident*: blank-callsign samples sit in tracks of their
own. Any fix to legacy's fragmentation necessarily destroys that property —
which is the whole point of the fix. Drop callsign from the key and a track that
carries a real callsign alongside blank ones is labelled blank; the flight
becomes unidentifiable and matches no ground-truth counterpart.

**What it cost.** Measured against the pipeline as it then stood, adopting the
better segmentation *lost* ADEP/ADES points, and the losing arms lost them by
coverage collapse with accuracy unmoved — the signature of a labelling failure
rather than a geometric one. Legacy's fragmentation and the downstream coverage
collapse were one phenomenon seen from two ends, so the recommendation was
conditional: fix the label first, then ship the segmentation.

**What was actually wrong was larger than the label.** Tracing it found that
`flight_id` is used as a grouping or join key in a dozen places downstream and
is never aggregated to one value per track — so the flight list is written on
the invariant *one `flight_id` per `track_id`*, which `legacy` guarantees and
the new segmentation breaks. A track carrying several callsigns does not merely
get labelled wrongly; it *fans out into several rows* at each of those keys. The
same latent fan-out was present in the event extraction of step 04.

**It is fixed.** Every point where the flight list and the event extraction read
the track table now resolves `flight_id` to the track's dominant non-blank
callsign, over the unfiltered frame, before anything groups or joins on it. Runs
that stamp a legacy version string skip the resolution, so re-running a released
month still reproduces what was released, column for column. The numbers
throughout this page are from the fixed pipeline; there is no repaired-versus-
unrepaired pairing to interpret any more.

### The convergence check

The benchmark kept its own independent repair, relabelling each track with its
most frequent non-blank callsign inside the harness, and both jobs are still run.
Now that production does the same thing, their **agreement** verifies that the
production fix and the benchmark repair are the same operation. A disagreement
would be a finding rather than a measurement.

```{r fix-table}
#| echo: false
rows <- do.call(rbind, lapply(PERIODS, function(p) {
  data.frame(
    period = p,
    arm = c("legacy (control)", "recommended"),
    production = c(P(payoff, "legacy", p), P(payoff, "recommended", p)),
    harness    = c(P(fixcs, "legacy", p), P(fixcs, "recommended", p)),
    check.names = FALSE)
}))
rows$delta <- rows$harness - rows$production
kable(rows, row.names = FALSE, digits = 0,
      col.names = c("period", "arm", "production fix only",
                    "production fix + harness repair", "difference"),
      format.args = list(big.mark = ","),
      caption = "The two jobs agree where it matters. `recommended` comes back identical on every recorded field in both periods; before the production fix the two differed by tens of thousands of flights.")
```

`recommended` agrees exactly, on both periods, which is the check passing.

`legacy` does not, on 2025 only, and by
`r num(abs(PAY_LEGACY_25 - PROD_LEGACY_25))` points confined entirely to the
arrival side — the same flights answered, a handful labelled with a different
aerodrome. That was traced, and it is not the callsign repair: it is a
non-deterministic aggregate in shipped ADES detection, which is reported in
[Limitations](#sec-limitations) because it is a real defect and it is not this
study's to fix.

### Where the residual comes from

With the labelling fixed in production, the recommended segmentation scores
**`r num(PAY_REC_25)`** against legacy's **`r num(PAY_LEGACY_25)`** on 2025 and
**`r num(PAY_REC_24)`** against **`r num(PAY_LEGACY_24)`** on 2024 — short by
`r num(PAY_SHORT_25)` and `r num(PAY_SHORT_24)` points,
`r pct(100 * PAY_SHORT_25 / PAY_LEGACY_25)` and
`r pct(100 * PAY_SHORT_24 / PAY_LEGACY_24)` — while recovering
`r sprintf("%.0f", GAIN_25)` points more flights on the ladder. Its **ADEP**
coverage slightly *exceeds* legacy's on both periods:
`r sprintf("%.2f", 100 * PR(fixcs, "recommended", "2025", "adep_coverage"))`%
against `r sprintf("%.2f", 100 * PR(fixcs, "legacy", "2025", "adep_coverage"))`%.

The residual is not accuracy and it is not departures. It is arrival coverage,
and the only failure mode that grew is the one [A8](#sec-a8) said would grow:
`recommended` merges `r pct(MERGE_REC_25)` of flights where `legacy` merges
`r pct(MERGE_LEGACY_25)`, and a merged track has one arrival where there should
be two, so one of the two flights loses its destination outright. The two
statistics are measured against different ground truths — the ladder's and the
flight list's — so they cannot be reconciled flight for flight, and merging on
its own is smaller than the coverage difference it is offered to explain. What
can be said without qualification is that merging is the only thing that moved
in the wrong direction, and that it moved.

That is the price of grouping on the airframe. It is stated rather than rounded
away, and it is small enough that the `r sprintf("%.0f", GAIN_25)`-point gain in
recovered flights buys it many times over.

## What shipped {#sec-defaults}

**The A8 rule ships, as `segmentation.method = "standard"`:** group on `icao24`,
break when the last non-blank callsign genuinely changes with the lookback
bounded to the gap threshold, and keep legacy's gap rules as a floor.

The recommendation was conditional when this study first reported, and the
condition was an ordering rather than a hedge. Both steps have now been taken,
in this order:

1. **First, the flight-list labelling was fixed.** `flights.py` no longer picks
   a track's callsign with `F.min`; every read of the track table resolves
   `flight_id` to the track's dominant non-blank callsign before anything groups
   or joins on it, and the same fix was applied to step 04's event extraction.
   This was worth doing on its own merits even had the segmentation never
   changed, because it is a silent trap for any future change to track identity
   and it fails as a *coverage* statistic rather than as an error — nothing
   looks broken, there is simply less data.
2. **Then the segmentation was switched.** Reversed, the pipeline would have
   spent the interval between the two changes at substantially reduced ADEP
   coverage.

Three things a consumer needs to know, all of them consequences of the release
rather than open questions:

- **`track_id` changes for all data produced from this release forward**, in
  shape as well as value: `standard` carries no `_{year}_{month}` suffix, so
  identifiers become `{hash}_{offset}`. A consumer joining on `track_id` across
  the boundary gets an empty join rather than an error, which reads as missing
  data; a consumer parsing the suffix breaks outright.
- **Past months do not reproduce under the new rule, and are not meant to.**
  Setting `segmentation.method` back to `"legacy"` reproduces the published
  identifiers, and the version-guarded paths in the flight list and the event
  extraction reproduce the published labels alongside them.
- **Nothing in the published data records which rule produced a row.**
  `osn_tracks` carries no version column, so the only way to tell is to know
  when the row was published. This is a deliberate decision and its cost is
  written down in [Limitations](#sec-limitations).

## Provenance {#sec-provenance}

*This section says where every number on this page comes from.*

Each row is one file this page reads: the job that produced it, the commit it
was produced at, and when. Every figure on this page is regenerated by
`benchmarks/regenerate_track_v1.py`, which the page runs before drawing
anything, so a row marked **unverified** would mean a file staged outside that
chain and traceable to no command.

```{r provenance}
#| echo: false
#| results: asis
mf <- file.path("data", "_manifest.json")
if (!requireNamespace("jsonlite", quietly = TRUE)) {
  cat("*jsonlite not available; provenance table skipped.*\n")
} else if (!file.exists(mf)) {
  cat("**No manifest.** Every CSV here was staged by hand and none can be traced to the command that produced it.\n")
} else {
  m <- jsonlite::fromJSON(mf, simplifyDataFrame = FALSE)
  files <- sort(list.files("data", pattern = "[.]csv$"))
  rows <- do.call(rbind, lapply(files, function(f) {
    e <- m[[f]]
    if (is.null(e)) {
      data.frame(File = f, Job = "**unverified**", Commit = "—",
                 Produced = "—", check.names = FALSE)
    } else {
      data.frame(File = f, Job = basename(e$script),
                 Commit = paste0(e$git_sha, if (isTRUE(e$git_dirty)) " (dirty)" else ""),
                 Produced = substr(e$produced_utc, 1, 16), check.names = FALSE)
    }
  }))
  print(kable(rows, row.names = FALSE, align = "l"))
  n_un <- sum(rows$Job == "**unverified**")
  cat(sprintf("\n%d files, %d traced to a command, %d unverified.\n",
              nrow(rows), nrow(rows) - n_un, n_un))
}
```

Two classes of figure on this page are **not** read from those files, and both
are marked as such where they appear.

- A handful of diagnostic measurements made while the arms were implemented —
  the share of `legacy` tracks carrying no callsign, the share of
  `airframe_only` tracks carrying more than one, and the cost of an unbounded
  callsign lookback. They are quoted with their source named in the text
  ("measured while the arm was implemented", "measured during implementation")
  and they are not regenerated by this page.
- The V7 flight-list study's own published total, used once as an external
  control, and the ground-truth validation figures quoted in
  [The sample](#sec-sample). Both belong to other papers and are cited, not
  recomputed here.

Everything else — every rate, count, percentile, table cell and plotted bar — is
an expression evaluated over the files above at render time. None of it is typed
into the prose, which is what stops the text and the tables drifting apart.

The re-run behind the current numbers is itself worth a note on method. All
fourteen files were regenerated after two defects were fixed in the ground-truth
loader, including six jobs that were confidently predicted not to move: they had
been reasoned about through two of their three declared dependencies, and the
third had changed. Every cell of all four sweep files moved, by up to 0.023 pp,
relocating no optimum and changing no conclusion — but the numbers on the page
are different from the ones that were there. Narrowing a declared dependency to
avoid a re-run is exactly how a page ends up publishing figures that nothing can
account for.

## Limitations {#sec-limitations}

*The claims this page does **not** make, and the defects it found and did not
fix.*

### The baseline is a re-implementation, not the published `track_id`

Two distinct things are meant by "the baseline", and neither is what production
now runs.

First, this study measures algorithms in a harness rather than the pipeline
itself. A0 is a re-implementation of the legacy *rule* inside the generic
segmentation engine, which is asserted by test to reproduce the frozen
algorithm's partition under production parameters. That contract is on the
partition, not a claim to have run the pipeline.

Second, applied to *cleaned* tracks, the legacy rule does not reproduce the
published identifiers exactly: it differs on 909 of 236,899 tracks, 0.38%, and
on `r num(abs(PROD_LEGACY_25 - V7_PUBLISHED))` of this page's
`r num(PROD_LEGACY_25)` ADEP/ADES points — the control residual noted in
[the payoff](#sec-payoff). Production's identifiers were computed upstream on raw
altitudes, while
`osn_tracks_clean` masks bad altitudes to NULL — and `NULL < 5000` is NULL, not
true, so the low-altitude split silently does not fire where it once did.
Measured, **950 of 950 disagreements have a NULL barometric altitude**, against
a 21.4% base rate. The published `osn_tracks_clean` therefore carries a
`track_id` that its own columns can no longer reproduce: nothing is wrong with
the published data, but anyone recomputing identifiers from the clean table —
the obvious thing to do — gets different tracks and no warning.

Neither point affects the comparisons on this page, because every arm reads the
same cleaned table through the same engine and is subject to both behaviours
equally. Version 2 of this study is the one that runs the pipeline end to end.

### Published data carries no marker of which segmentation produced it

`osn_tracks` has no version column, so **no row of published OPDI data says
which segmentation rule built its `track_id`**. Distinguishing the two regimes
depends entirely on knowing the release date.

The decision not to add such a column was deliberate and its cost is concrete.
The guards that let a legacy re-run reproduce released labels — in `flights.py`
and in `events.py` — key on the *run's configuration*, not on the data: they ask
"is this run stamping a legacy version string?" when the question they need
answered is "were the tracks I am reading built with the legacy rule?". Nothing
in the data can tell them. So a
legacy-configured run over tracks that were rebuilt with `standard` will take
the legacy branch and publish under a frozen version string, and no check
anywhere can catch it. The same blind spot reaches the flight-events ladder,
whose lower rungs carry a legacy events version and would silently measure the
new segmentation if re-run after the switch.

### ADES values are not reproducible run to run

**This is a defect in shipped ADEP/ADES detection, it predates this study, and
it is not fixed.** It is recorded here because both this page and the
flight-list studies quote ADES percentages to two decimals, and a figure that is
not reproducible must say so.

`flights.py:_rank_airports` ranks candidate aerodromes correctly with
`row_number()`, and then names the winner with `first(apt_ident)` — an
aggregate with no defined order. Adding a join upstream can change the answer.
The defect is self-detecting, because the *runner-up* list is built from the
rank rather than from `first()`, so a wrongly named aerodrome shows up in its
own "not most likely" list.

Measured across repeated runs on identical code and data: roughly 28 of 88,737
named ADES per run, about 0.03%, come from the undefined-order aggregate, and
about 20 can flip between two runs. The affected flights are short hops between
neighbouring aerodromes. **ADEP is immune** — the departure path filters on the
rank instead of aggregating, which is the correct pattern and already exists in
the same file. `ades_mode` defaults to `trend`, so this is the shipped arrival
path.

It is the explanation for the one disagreement in
[the convergence check](#sec-payoff): the control differs by a handful of ADES
flights on one period and by none on the other, with coverage identical to six
decimals.

### Two ground-truth intervals over twenty hours survive in the 2024 sample

Ground truth is built from Network Manager's own `AOBT_3`/`ARVT_3` pairing, and
two intervals in the 2024 sample exceed twenty hours — the longest being 27.9 h,
THY801 LTFM→SKBO. Because samples are attributed to a flight by containment, an
over-long interval can absorb samples belonging to a neighbouring leg of the
same airframe.

These are **not** the defect that was fixed. A separate loader defect, since
corrected, windowed ground truth on the off-block day rather than on the flight
interval and keyed the APDF departure join on the off-block day rather than the
take-off day; the second let a nightly service take its take-off time from
yesterday's rotation, producing intervals of up to 26.9 hours. Fixing it removed
every such interval from 2025 and all but these two from 2024, and moved the
scored flight counts by about a tenth of a percent with every rate moving by no
more than 0.038 pp.

The two survivors are NM's own pairing and are undiagnosed. They were
deliberately not filtered out: a length filter would change the denominator of
every metric on this page, which is a larger and less examined intervention than
the contamination it removes. The effect is carried, stated, and small.

### The ADEP/ADES configuration predates the v6.2 datum change

As [§ the flight-list configuration](#sec-detection) records, the payoff runs were
made from a checkout that does not contain v6.2's altitude-cut datum, so `trend`
cuts at FL60 above mean sea level rather than at a height above field elevation.
The configuration is identical across all eight arms, so every comparison in
[the payoff](#sec-payoff) is unaffected; the absolute totals are not comparable
with figures published by v6.2 or later.

They are also not regenerated, and will not be until the V7 ladder can run
again — its last rung no longer matches the shipped `DetectionConfig()`. So the
payoff chapter alone quotes a measurement made against the older ground truth and
the older datum, while every other number on this page comes from the current
one. That is a real inconsistency inside one paper. It is carried rather than
hidden because the alternative is a re-run at a configuration that answers a
different question, and because the quantity the chapter reports — the difference
between two arms at one fixed configuration — is the quantity least disturbed by
holding that configuration still.

### Ground truth covers Network Manager flights only

Merges with traffic absent from the NM flight table — military, general
aviation, non-ECAC operators — are invisible to the metric, because the merged
partner has no ground-truth flight to be merged with. **The merge rates on this
page are therefore floors, not estimates.** Fragmentation is unaffected, since
it is measured within flights that ground truth does contain.

### One arm is untested, not null

`no_month_suffix` is identical to legacy on both samples because both sit inside
a single month. It is reported as untested. Evaluating it needs a sample
spanning a month boundary.

### Two three-day samples in June

Both samples are in early June. Nothing here licenses a seasonal claim, and the
callsign statistics in particular could plausibly differ in a period with more
charter or seasonal traffic.

### The sweep covers one arm's parameters, not every arm's

[The sweep](#sec-sweep) varies the four parameters of `recommended`. It does not
vary the ground-dwell or turnaround parameters that the domain arms introduce, so
those arms are each compared at one setting and could be better at another. Nor
does it sweep `legacy`: an earlier version of this study did, and found it within
a fraction of a point of its own best cell, but those runs are not in this
revision and the claim is not made here. What the ladder measures is each rule at
one setting; what the sweep measures is how much one rule's setting is worth.

### The payoff holds the flight list fixed

Each arm is scored at one flight-list configuration — the one tabulated in
[§ the flight-list configuration](#sec-detection) — rather than re-tuned per
arm. A joint optimum might be better for some arm; finding it would multiply
cluster time by the number of arms, and one Spark job runs at a time. The
natural follow-up, now that the labelling is fixed, is to re-tune the flight
list on the recommended arm alone: one extra grid, not eight.
