---
title: "Turning ADS-B state vectors into flight tracks"
subtitle: "Version 1 — what the segmentation is worth, and what had to change downstream before a better one could ship"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r regenerate}
#| include: false
#| cache: false

# Rendering this page runs the analysis behind it.
#
# `regenerate_track_v1.py` is the executable definition of the study: every CSV
# read below is produced by exactly one job it declares, with exactly the
# arguments it declares. It re-runs a job when the *code that job depends on*
# has changed since the output was written -- not when the file is old, which
# says nothing. So a render with everything current is a fast no-op, and a
# render after editing `segmentation/methods.py` recomputes every arm before the
# first table is drawn.
#
# Running a stale job needs cluster credentials, because the numbers come from
# Spark over S3 against Network Manager reference data and there is no way to
# recompute them without the data they are computed from. Set OPDI_RENDER=check
# to fail fast on staleness instead, or OPDI_RENDER=allow-stale to render a
# draft from whatever is staged -- which the provenance table at the foot of
# this page will then show as unverified.

# The pipeline repo is a sibling of the portal. That resolution is wrong inside
# a git worktree, where the paper sits under .claude/worktrees/<name>/papers and
# the sibling walk lands somewhere that does not exist -- so the regeneration
# step would be skipped silently and the page would render whatever is staged.
# OPDI_REPO_DIR overrides it; the default is unchanged for a normal checkout.
REPO   <- Sys.getenv("OPDI_REPO_DIR",
                     normalizePath(file.path(getwd(), "..", "..", "..", "opdi"),
                                   mustWork = FALSE))
PYTHON <- Sys.getenv("OPDI_PYTHON", file.path(REPO, ".venv310", "bin", "python"))
MODE   <- Sys.getenv("OPDI_RENDER", "run")

# A missing entrypoint must not read as "nothing to regenerate". Skipping it
# silently is how a page ends up publishing numbers no command can account for.
if (!file.exists(file.path(REPO, "benchmarks", "regenerate_track_v1.py")) &&
    MODE != "skip") {
  stop("regenerate_track_v1.py not found under REPO=", REPO,
       " -- set OPDI_REPO_DIR, or OPDI_RENDER=skip to render without it.")
}

regen <- file.path(REPO, "benchmarks", "regenerate_track_v1.py")
if (file.exists(regen) && MODE != "skip") {
  flags <- switch(MODE,
                  "run" = character(0),
                  "check" = "--check",
                  "allow-stale" = c("--check", "--allow-stale"),
                  character(0))
  status <- system2(PYTHON, c(shQuote(regen), flags), stdout = "", stderr = "")
  if (!identical(status, 0L)) {
    stop("regenerate_track_v1.py exited ", status,
         " -- the figures below would not match the code that is checked out. ",
         "Set OPDI_RENDER=allow-stale to render anyway.")
  }
}
```

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every figure and table on this page comes from a committed CSV
# under data/. Nothing here reads S3, a database or a Spark cluster.

D <- function(f) read.csv(file.path("data", f), stringsAsFactors = FALSE)

arms   <- rbind(D("arms_2025.csv"), D("arms_2024.csv"))
sweep1 <- rbind(D("sweep_2025_stage1.csv"), D("sweep_2025_stage1_ext.csv"))
sweep2 <- rbind(D("sweep_2025_stage2.csv"), D("sweep_2024_stage2.csv"))
payoff <- rbind(D("payoff_2025.csv"), D("payoff_2024.csv"))
fixcs  <- rbind(D("payoff_fixcallsign_2025.csv"), D("payoff_fixcallsign_2024.csv"))
cont   <- rbind(D("containment_2025.csv"), D("containment_2024.csv"))

# The boundary histograms carry no period column -- one file per period *is* the
# period -- so it is attached on read rather than assumed by anything later.
readhist <- function(p) {
  x <- D(paste0("boundary_hist_", p, ".csv"))
  x$period <- p
  x
}
bhist <- rbind(readhist("2025"), readhist("2024"))

num <- function(x) formatC(x, format = "d", big.mark = ",")
# The "+" flag does nothing for %s in R, so the sign is prepended explicitly --
# a signed delta that silently loses its sign reads as a level, not a change.
sgn <- function(x) paste0(if (x >= 0) "+" else "-", num(abs(x)))
pct <- function(x) sprintf("%.2f%%", x)
pp  <- function(x) sprintf("%+.2f", x)
sec <- function(x) sprintf("%+d", as.integer(x))

# One accessor per file, so no two tables on this page can disagree about what
# an arm scored.
A  <- function(a, p, col) arms[arms$arm == a & arms$period == p, col]
CT <- function(p, col) cont[cont$period == p, col]
PR <- function(df, a, p, col) df[df$arm == a & df$period == p, col]
P  <- function(df, a, p) {
  r <- df[df$arm == a & df$period == p, ]
  r$adep_score + r$ades_score
}

PERIODS <- c("2025", "2024")

# --- headline values, computed here rather than written into the prose, so the
# --- opening paragraph cannot drift away from the tables that justify it.
CLEAN_LEGACY_25 <- A("legacy", "2025", "clean_match_pct")
CLEAN_REC_25    <- A("recommended", "2025", "clean_match_pct")
CLEAN_LEGACY_24 <- A("legacy", "2024", "clean_match_pct")
CLEAN_REC_24    <- A("recommended", "2024", "clean_match_pct")
FRAG_LEGACY_25  <- A("legacy", "2025", "fragmented_pct")
FRAG_REC_25     <- A("recommended", "2025", "fragmented_pct")
MERGE_LEGACY_25 <- A("legacy", "2025", "merged_pct")
MERGE_REC_25    <- A("recommended", "2025", "merged_pct")
MERGE_AIR_25    <- A("airframe_only", "2025", "merged_pct")
GAIN_25         <- CLEAN_REC_25 - CLEAN_LEGACY_25
GAIN_24         <- CLEAN_REC_24 - CLEAN_LEGACY_24
N_GT_25         <- A("legacy", "2025", "n_flights")
N_GT_24         <- A("legacy", "2024", "n_flights")
N_APDF_25       <- A("legacy", "2025", "n_apdf_flights")

# The sweep: legacy at its production parameters, against the best cell anywhere
# in the grid.
PROD <- function(df, p) {
  r <- df[df$period == p & df$gap_minutes == 30 &
          df$low_alt_gap_minutes == 15 & df$low_alt_ft == 5000, ]
  r$clean_match_pct[1]
}
SWEEP_PROD_25 <- PROD(sweep2, "2025")
SWEEP_BEST_25 <- max(sweep2$clean_match_pct[sweep2$period == "2025"])
SWEEP_PROD_24 <- PROD(sweep2, "2024")
SWEEP_BEST_24 <- max(sweep2$clean_match_pct[sweep2$period == "2024"])
SWEEP_CELLS   <- nrow(sweep1)
SWEEP_SPREAD1 <- max(sweep1$clean_match_pct) - min(sweep1$clean_match_pct)
SWEEP_GAIN_25 <- SWEEP_BEST_25 - SWEEP_PROD_25
SWEEP_GAIN_24 <- SWEEP_BEST_24 - SWEEP_PROD_24
# How many times larger the identity fix is than the entire tuning question.
SWEEP_RATIO   <- GAIN_25 / SWEEP_GAIN_25

# The payoff, and the convergence check that now sits beside it.
PAY_LEGACY_25 <- P(payoff, "legacy", "2025")
PAY_REC_25    <- P(payoff, "recommended", "2025")
PAY_LEGACY_24 <- P(payoff, "legacy", "2024")
PAY_REC_24    <- P(payoff, "recommended", "2024")
PAY_SHORT_25  <- PAY_LEGACY_25 - PAY_REC_25
PAY_SHORT_24  <- PAY_LEGACY_24 - PAY_REC_24
FIX_LEGACY_25 <- P(fixcs, "legacy", "2025")
FIX_REC_25    <- P(fixcs, "recommended", "2025")
FIX_LEGACY_24 <- P(fixcs, "legacy", "2024")
FIX_REC_24    <- P(fixcs, "recommended", "2024")
```

## Summary {.unnumbered}

**The rule that built every `track_id` OPDI published up to v0.0.2 recovers
about half the flights it should, the cause is one column in a grouping key
rather than a mistuned threshold, and a rule that fixes it has now shipped.**
That is the whole finding. The rest of this page is how it was established, and
what had to be repaired in the flight list before the better rule could be
adopted without making the published product worse.

Before OPDI can derive a flight list it has to decide which state vectors belong
to the same flight. That decision — **segmentation** — is made once, and
everything downstream inherits it: the flight list groups by `track_id`, events
are extracted per track, measurements accumulate from a track's start. This page
asks what the decision is worth, in four parts.

**The old rule recovers about half the flights it should.** On a three-day
sample of `r num(N_GT_25)` Network Manager flights it matches
**`r pct(CLEAN_LEGACY_25)`** of them one-to-one, meaning one flight, one track,
that track holding nothing else; **`r pct(FRAG_LEGACY_25)`** of flights are split
across more than one track. The failure is almost entirely that splitting, and
almost never the opposite error of running two flights together.

**Its parameters are not the problem.** Its three thresholds were swept over
`r num(SWEEP_CELLS)` combinations, every axis widened until the curve turned over
inside the grid rather than at its edge. The whole grid spans
`r sprintf("%.2f", SWEEP_SPREAD1)` points of one-to-one matching, and on the full
three-day samples the best combination beats the production setting by only
**`r pp(SWEEP_GAIN_25)`** points on 2025 and **`r pp(SWEEP_GAIN_24)`** on 2024.
The rule was already at, or a fraction of a point from, the best setting anyone
could have picked for it.

**The identity key is the problem.** The old group key is
`(icao24, callsign)`, so a state vector decoded without a callsign forms a group
— and therefore a track — of its own. Grouping on the airframe alone, and
splitting only when the callsign *genuinely* changes, takes one-to-one matching
to **`r pct(CLEAN_REC_25)`** on 2025 and **`r pct(CLEAN_REC_24)`** on 2024: a
gain of **`r pp(GAIN_25)`** points, about `r sprintf("%.0f", SWEEP_RATIO)` times
everything the parameter sweep had to offer. This is the rule that now ships, as
`segmentation.method = "standard"`.

**The flight list could not consume it until one aggregate was fixed, and that
is the most useful finding here.** `flights.py` labelled each flight with the
lexicographically smallest callsign in its track, and a blank sorts first. That
is harmless only while every track carries exactly one callsign — a property the
old group key supplied by accident, and which any fix to the fragmentation
necessarily destroys. Measured against the pipeline as it then stood, the better
segmentation therefore made the published product *worse*: not by predicting
wrong aerodromes, but by producing flights that could no longer be identified at
all. That labelling is now fixed in production, and the numbers on this page are
the fixed ones.

With it fixed, the better segmentation costs
**`r num(PAY_SHORT_25)`** ADEP/ADES points on 2025 and **`r num(PAY_SHORT_24)`**
on 2024 — `r pct(100 * PAY_SHORT_25 / PAY_LEGACY_25)` and
`r pct(100 * PAY_SHORT_24 / PAY_LEGACY_24)` of the total — while recovering
`r sprintf("%.0f", GAIN_25)` points more flights. The residual is not accuracy,
which is unchanged; it is arrival coverage, and it traces to the one failure mode
the new rule makes slightly worse: merging.

## How to read this page

Sections are ordered as the work was done, not as the conclusion would suggest,
because two of the results only make sense as answers to questions asked in
order.

[What a track is](#sec-problem) defines the object and says why the question was
open. [What has already happened to a sample](#sec-pipeline) is a map of the
pipeline steps that run before any segmentation sees a state vector; skip it if
you already know them. [How a segmentation is scored](#sec-metrics) matters more
than usual here, because the obvious scoring choice is misleading and this page
says why before using something else; it also defines every term the later
sections use. [The sample](#sec-sample) says what ground truth is and where it
comes from. [The ladder](#sec-ladder) is one subsection per candidate rule, each
with what it changes, what it is trying to fix, the rule itself in enough detail
to reimplement, and what it scored. [The sweep](#sec-sweep) asks whether the old
rule was merely mistuned, and answers no. [The payoff](#sec-payoff) measures the
downstream effect on ADEP/ADES and explains why it does not track segmentation
quality. [What shipped](#sec-defaults) states what was adopted and what
consumers have to know about it.

Every number on this page is read from a committed CSV, regenerated by the
entrypoint the page runs before drawing anything;
[Provenance](#sec-provenance) lists the files one by one.
[Limitations](#sec-limitations) is where the claims this page does *not* make
are written down.

## What a track is, and why the question was open {#sec-problem}

*This section defines the object under study and the three ways a rule for
building it can fail. By the end of it you will know what the rule under test
was, and what "better" is going to mean.*

OPDI ingests OpenSky state vectors — position reports, one aircraft at a time,
with no notion of a flight. A **track** is the pipeline's answer to "which of
these belong together", and it is the unit everything downstream is built on:
the flight list groups by `track_id`, events are extracted per track, and
measurements accumulate from a track's start. The procedure that produces tracks
is called a **segmentation**, and a **track identifier** (`track_id`) is what it
stamps on every sample it decides belongs to the same flight.

The rule this study set out to evaluate — called **`legacy`** throughout, and
the rule behind every dataset OPDI published up to v0.0.2 — is short. Group the
samples by `(icao24, callsign)`, order them by time, and start a new track when
the gap to the previous sample exceeds 30 minutes, or exceeds 15 minutes while
the aircraft is below 5,000 ft. The resulting identifier carries a
`_{year}_{month}` suffix.

Three things can go wrong with such a rule, and they are not symmetric:

- **Fragmentation** — one real flight becomes several tracks. Everything
  downstream sees several short flights where there was one.
- **Merging** — several real flights become one track. The flight list emits one
  flight and the others simply do not exist in the output.
- **Boundary error** — the track is one flight, but starts or ends in the wrong
  place, so first- and last-seen events and cumulative measurements are wrong.

Of the three, merging is the most damaging, because a merged flight is
unrecoverable downstream, while a fragmented one is at least present in pieces
and a boundary error leaves the flight intact. The metrics below are chosen so
that merging and fragmentation cannot be confused with each other, and the
candidate rules are compared on both rather than on a single blended score.

The rule has never been benchmarked, and that is why this study exists.

What it is *not* is frozen. Segmentation is a versioned choice: a release states
which rule built its `track_id` values, and changing that rule changes
`track_id` for everything produced from that release forward — in value, and in
shape where the month suffix goes. That is a release decision, to be documented
and communicated to consumers rather than avoided, and it is a decision that has
now been taken. As of this release `segmentation.method` defaults to
`"standard"`, which is the rule this study recommends. `"legacy"` remains
selectable and reproduces the identifiers of every dataset published up to OPDI
v0.0.2, so released data stays reproducible.

One consequence for the reader, stated here so nothing later depends on
inferring it. **This paper measures algorithms, not the pipeline.** Every rule
below is expressed in a generic segmentation engine and scored in a harness, so
that eight candidates could be compared without eight production pipeline runs —
which at this data volume is the difference between a study and a quarter. That
is a legitimate design and it is also a real limitation: what is measured here
is a re-implementation of each rule, not the pipeline's own output. Version 2 of
this study is the paper that runs the pipeline itself, end to end, on the rule
this one recommends. If you want to know what a rule is worth, you are in the
right place; if you want to know what the pipeline now produces, read version 2.

## What has already happened to a sample {#sec-pipeline}

*Side information. This section says what the `/opdi` pipeline has already done
to a state vector before any segmentation sees it, so that nothing later in this
page is a surprise. If you know the pipeline steps, skip to
[the metrics](#sec-metrics).*

Segmentation is step 02 of a chain, and the samples reaching it are not raw
OpenSky rows. Two earlier decisions in particular shape everything measured on
this page: the ingestion filter decides which samples exist at all, and the
decimation rule decides how densely a flight is described.

```{r pipeline-diagram}
#| echo: false
#| fig-width: 7
#| fig-height: 4.2
#| fig-cap: "The pipeline steps that run before, and after, the decision this page studies. Step 02 is the one under test; everything above it has already happened to a sample by then, and everything below it inherits the result."
# Drawn with ggplot rather than as a mermaid block, deliberately. A mermaid
# diagram written as an executable cell renders in HTML but needs a headless
# browser to rasterise for PDF, and that browser does not start in this
# environment -- the render then hangs with no error. A mermaid fence emitted
# from knitr avoids the hang but arrives without the JavaScript that draws it,
# so it shows as literal text. A plot goes through whatever device Quarto has
# already selected and is correct in both formats.
nodes <- data.frame(
  y = 6:1,
  label = c(
    "step 00 — reference tables (airports, aircraft)",
    "step 01 — ingest OSN state vectors: bbox + thinning",
    "step 02 — tracks: assign track_id",
    "step 02a — cleaning: mask bad values to NULL",
    "step 03 — flight list: ADEP / ADES per track",
    "step 04 — events and measurements per track"),
  under_test = c(FALSE, FALSE, TRUE, FALSE, FALSE, FALSE))
h <- 0.34
ggplot(nodes) +
  geom_rect(aes(xmin = 0, xmax = 1, ymin = y - h, ymax = y + h,
                fill = under_test, colour = under_test, linewidth = under_test)) +
  geom_text(aes(x = 0.5, y = y, label = label, fontface = ifelse(under_test, 2, 1)),
            size = 3.1) +
  geom_segment(data = nodes[nodes$y > 1, ],
               aes(x = 0.5, xend = 0.5, y = y - h, yend = y - 1 + h),
               arrow = arrow(length = unit(1.8, "mm"), type = "closed"),
               colour = "grey40", linewidth = 0.3) +
  annotate("text", x = 1.04, y = 4, hjust = 0, size = 2.9, colour = "grey25",
           label = "this study replaces\nthe rule inside step 02") +
  scale_fill_manual(values = c(`FALSE` = "white", `TRUE` = "grey90")) +
  scale_colour_manual(values = c(`FALSE` = "grey55", `TRUE` = "grey15")) +
  scale_linewidth_manual(values = c(`FALSE` = 0.3, `TRUE` = 0.8)) +
  coord_cartesian(xlim = c(0, 1.5), ylim = c(0.4, 6.6), clip = "off") +
  theme_void() +
  theme(legend.position = "none")
```

```{r pipeline-table}
#| echo: false
steps <- data.frame(
  Step = c("00", "01", "02", "02a", "03", "04"),
  Does = c(
    "Builds reference tables: H3 airport detection zones, HexAero airport ground layouts, OurAirports, the OpenSky aircraft database. Read later; never rewritten by a pipeline run.",
    "Downloads OpenSky state vectors, restricts them to the OPDI bounding box, and thins them to one sample per aircraft per interval.",
    "Groups samples into tracks and stamps `track_id`. **This is the step under study.** Also attaches the H3 index each later join uses.",
    "Masks values that cannot be right — duplicates, out-of-range positions, stale broadcasts, derivative spikes, isolated points — to NULL, keeping the row. Writes a second table rather than editing the first.",
    "One flight per `track_id`: departure and destination aerodrome, and the flight's callsign label.",
    "Milestones per track — take-off, top-of-climb, level segments, landing — and the measurements attached to them."
  ),
  Output = c("reference tables", "`osn_statevectors`", "`osn_tracks`",
             "`osn_tracks_clean`", "`opdi_flight_list`", "events, measurements"),
  check.names = FALSE)
kable(steps, row.names = FALSE, align = "lll",
      col.names = c("Step", "What it does", "Writes"))
```

Two parameters of step 01 are worth naming, because every count on this page is
conditioned on them.

```{r ingestion-params}
#| echo: false
ing <- data.frame(
  Parameter = c("bounding box (min lon, min lat, max lon, max lat)",
                "thinning interval",
                "thinning rule"),
  Value = c("(-25.86653, 26.74617, 49.65699, 70.25976)",
            "5 s",
            "`bucket`"),
  Consequence = c(
    "A flight is observed only inside this box. A track that begins at its edge may have entered from outside, which is why an out-of-area label exists at all.",
    "Consecutive samples of one aircraft are at least this far apart, so every gap threshold in this study is measured against a 5 s grid rather than a 1 Hz one.",
    "One row kept per aircraft per interval bin — the last, so a track's final observation survives exactly. The earlier `modulo` rule kept the row whose epoch second divides the interval, and built every dataset published before it."
  ),
  check.names = FALSE)
kable(ing, row.names = FALSE, align = "lll",
      col.names = c("Parameter", "Value", "Why it matters here"))
```

Step 02a matters for a different reason, and it comes back in
[Limitations](#sec-limitations): it masks bad values to NULL rather than
deleting rows, and `NULL < 5000` is NULL rather than true. Every arm in this
study reads the *cleaned* table, so all of them are subject to that behaviour
equally — but the identifiers in published data were computed before it, which
is a discrepancy worth knowing about.

## How a segmentation is scored {#sec-metrics}

*This section defines every measurement used later: the two failure measures,
the headline statistic, the rule that decides which track belongs to which
flight, and the sign convention for boundary error. Nothing after this section
introduces a metric.*

Comparing a segmentation against ground truth is a **clustering comparison**:
both the algorithm and the truth partition the same state vectors — the
algorithm into tracks, the truth into flights — and the question is how far the
two partitions agree.

### Homogeneity is merging; completeness is fragmentation

Two standard clustering measures map onto the two failure modes exactly, which
is why they are used here rather than a score invented for the occasion:

- **Homogeneity** asks whether each track contains samples from only one flight.
  It falls when a track holds more than one flight — so it is the *merging*
  measure.
- **Completeness** asks whether each flight's samples all landed in one track.
  It falls when a flight is spread over several tracks — so it is the
  *fragmentation* measure.

Both are entropy-based and both run from 0 to 1, with 1 meaning perfect
agreement with ground truth on that axis alone.

**V-measure** is their harmonic mean — a single number summarising both. It is
reported below and used to choose nothing, and the reason is measured rather
than assumed. Across the eight candidate rules V-measure spans
`r sprintf("%.4f", min(arms$v_measure))` to
`r sprintf("%.4f", max(arms$v_measure))`, a range of
`r sprintf("%.4f", max(arms$v_measure) - min(arms$v_measure))`, while the
one-to-one matching rate defined next spans
`r sprintf("%.1f", min(arms$clean_match_pct))` to
`r sprintf("%.1f", max(arms$clean_match_pct))` points. Worse than insensitive,
it is *misleading*: it ranks the candidates in a different order at the top,
which is the only place a comparison ever looks. The cause is that both
statistics are entropy-weighted over every sample in the period, and the
overwhelming majority of samples sit in tracks that every candidate gets right —
so V-measure is dominated by agreement that carries no information about where
the candidates differ.

### The headline: one-to-one matching

The headline statistic on this page is **`clean_match_pct`**, and it answers the
question a data consumer actually has: *what share of real flights came out as
exactly one track, holding nothing but that flight?*

Every ground-truth flight is placed in exactly one of three mutually exclusive
classes, which sum to 100%:

- **clean** — the flight's samples are in one track, and that track carries no
  other flight;
- **fragmented** — the flight's samples span several tracks, none of which
  carries another flight;
- **merged** — the flight's dominant track also carries at least one other
  flight.

A flight that is both merged and fragmented counts as **merged**, because
merging is the worse failure and a flight should not be able to improve its
classification by also being broken a second way. The definition is deliberately
strict: a single stray sample in a second track costs a flight its clean match.
The tolerant alternative — a dominant track covering some fraction of the flight
— was rejected because that fraction would be a free parameter chosen after
seeing the results. `clean_match_pct` is therefore a lower bound, and is
reported as one.

### Matching a track to a flight: interval containment {#sec-containment}

Every statistic above needs one thing first: a rule saying which ground-truth
flight a given state vector belongs to. That rule is **interval containment**,
and it is worth setting out fully, because it is where a comparison of this kind
is most easily rigged without anyone noticing.

**A ground-truth flight is an airframe and a span of time.** The airframe is
`icao24` — the same 24-bit address ADS-B broadcasts, which the Network Manager
flight table carries as `AIRCRAFT_ADDRESS`. The span is the flight's airborne
interval, from take-off to landing:

- where APDF covers the aerodrome, from APDF's measured movement times, the real
  ATOT and ALDT;
- elsewhere, inferred from Network Manager as `AOBT_3 + TAXI_TIME_3` for
  take-off and `ARVT_3` for landing. Both inferences were validated against real
  APDF movements before being relied on (see [The sample](#sec-sample)).

Each row carries which of the two it came from, so a statistic that needs
measured times rather than inferred ones — boundary error does — can restrict
itself to the APDF-backed subset.

**The join is on `icao24` and time overlap, and never on callsign.** This
differs from the ADEP/ADES flight-list studies, which key on callsign, and the
difference is forced rather than stylistic: one of the candidate rules below
removes callsign from track identity entirely, and another splits on it. Joining
on callsign would therefore score each candidate against a key it deliberately
does or does not have, which is prejudging the very thing under comparison.

**Containment, concretely.** A state vector is attributed to the ground-truth
flight of the same `icao24` whose interval holds its timestamp. A track and a
flight are then *matched* when at least one of the track's samples is attributed
to that flight, and the contingency of tracks against flights is what the three
classes above are counted from. Where two intervals of the same airframe touch,
the boundary sample goes to the earlier one, so back-to-back legs cannot
double-count it and inflate the merge rate.

A worked example, one airframe with two rotations in an afternoon:

```{r containment-example}
#| echo: false
ex <- data.frame(
  Flight = c("leg 1", "leg 1", "gap", "leg 2", "leg 2"),
  Interval = c("12:00–13:30", "12:00–13:30", "—", "14:10–15:40", "14:10–15:40"),
  Sample = c("12:05", "13:25", "13:50 (on stand)", "14:15", "15:35"),
  Attributed = c("leg 1", "leg 1", "no flight — outside both intervals",
                 "leg 2", "leg 2"),
  check.names = FALSE)
kable(ex, row.names = FALSE, align = "llll",
      col.names = c("Ground-truth flight", "Its interval", "A sample",
                    "Attributed to"))
```

If one track holds the 12:05 and 15:35 samples, both flights are **merged** into
it. If leg 1's two samples land in different tracks, leg 1 is **fragmented**.
The samples on stand at 13:50 belong to no flight and are scored against
nothing — segmentation is not penalised for what it does with them.

**What the restriction costs.** Ground truth is further restricted to flights
whose *whole* airborne interval lies inside the sampled window. The reason is
that a flight observed only for its last twenty minutes has no samples for the
rest of itself: it would score as truncated however well the segmentation
performed, and its boundary error would be a fact about where the window fell.
That argument justifies the restriction but does not price it, so the price was
measured.

```{r containment-table}
#| echo: false
ct <- data.frame(
  period = PERIODS,
  overlapping = sapply(PERIODS, function(p) CT(p, "n_gt_flights")),
  inside      = sapply(PERIODS, function(p) CT(p, "n_wholly_inside")),
  cs          = sapply(PERIODS, function(p) CT(p, "n_clipped_start")),
  ce          = sapply(PERIODS, function(p) CT(p, "n_clipped_end")),
  kept        = sapply(PERIODS, function(p) CT(p, "pct_kept")),
  frac        = sapply(PERIODS, function(p) CT(p, "median_observed_fraction_clipped")),
  check.names = FALSE)
kable(ct, row.names = FALSE, digits = c(0, 0, 0, 0, 0, 2, 3),
      col.names = c("period", "any airborne time in window", "wholly inside",
                    "clipped at start", "clipped at end", "kept %",
                    "median observed fraction of an excluded flight"),
      format.args = list(big.mark = ","))
```

So the restriction keeps `r sprintf("%.2f%%", CT("2025", "pct_kept"))` of the
2025 flights that have any airborne time in the window, and
`r sprintf("%.2f%%", CT("2024", "pct_kept"))` of the 2024 ones. The exclusion is
near-symmetric between the two window edges, which is what a window artefact
should look like and is evidence the rule is not systematically dropping one
kind of flight — long-haul arrivals, for instance.

The last column answers the obvious objection, that the rule might be discarding
nearly-complete flights on a technicality. It does not: the median excluded
flight had only about
`r sprintf("%.0f%%", 100 * CT("2025", "median_observed_fraction_clipped"))` of
its airborne time inside the window on 2025, and
`r sprintf("%.0f%%", 100 * CT("2024", "median_observed_fraction_clipped"))` on
2024. A typical excluded flight really was about half observed, and could not
have been scored fairly.

One further gap is worth stating, because it explains a difference between two
counts on this page. Of the
`r num(CT("2025", "n_wholly_inside"))` flights wholly inside the 2025 window,
`r num(N_GT_25)` are scored — the remaining
`r num(CT("2025", "n_wholly_inside") - N_GT_25)` produced no ADS-B sample at all
in the feed and so appear in no contingency. They are absent from the
denominator for the same reason, and no candidate rule is credited or penalised
for them.

### Boundary error, reported signed

Boundary error is the distance in time from a track's first and last sample to
the flight's actual take-off and landing. It is reported **signed**, with the
convention that

> **`off` = track start − take-off, `land` = track end − landing.** A negative
> value means the track began (or ended) *before* the aircraft did.

That is not cosmetic. An absolute error conflates two opposite conditions: a
track starting *before* take-off is normal and often desirable — the aircraft is
on the ramp broadcasting — while a track starting *after* take-off has missed
part of its own flight. `legacy`'s absolute median is about a hundred seconds,
which reads as a small tidy number. Signed, its median is
**`r sec(A("legacy", "2025", "off_signed_p50_s"))` s** and its tenth percentile
is **`r sec(A("legacy", "2025", "off_signed_p10_s"))` s** — which says something
quite different: the typical track begins just *after* wheels-off, while a
minority carry a full quarter-hour of taxi-out. The distribution is two-sided,
and an absolute median over it is close to meaningless.

The landing side is the opposite and far more uniform: every percentile is
positive, with a median of
**`r sec(A("legacy", "2025", "land_signed_p50_s"))` s**. Tracks essentially
always run on through taxi-in. That asymmetry is a coverage fact — on-ground
ADS-B before departure reaches a minority of flights, while an arriving aircraft
is still powered and broadcasting to the stand — and not an algorithm defect.

The percentiles below are what the text quotes. They are computed over the
flights whose ground-truth times are APDF-measured rather than NM-inferred —
`r num(N_APDF_25)` of the `r num(N_GT_25)` scored flights on 2025 — because a
boundary error against an inferred time would partly be measuring the inference.

```{r boundary-percentiles}
#| echo: false
bp <- arms[arms$period == "2025",
           c("arm", "off_signed_p10_s", "off_signed_p50_s", "off_signed_p90_s",
             "land_signed_p10_s", "land_signed_p50_s", "land_signed_p90_s")]
bp <- bp[order(bp$off_signed_p50_s), ]
kable(bp, row.names = FALSE, digits = 0,
      col.names = c("arm", "dep p10", "dep p50", "dep p90",
                    "arr p10", "arr p50", "arr p90"),
      caption = "Signed boundary error in seconds, 2025. Negative means the track began or ended before the aircraft did.")
```

The shape is what the percentiles cannot show.

```{r boundary-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 3.6
#| fig-cap: "Signed boundary error on the 2025 sample, over the flights whose ground-truth times are APDF-measured. **Negative means the track began, or ended, before the aircraft did.** Departure: `legacy` and `recommended` both show a two-sided distribution — a spike just after wheels-off plus a long left tail of tracks that started on the ramp — while `ground_anchored` concentrates hard at zero, which is the arm cutting at the ground event rather than running back into the ramp. Arrival: all three are one-sided and nearly identical, because tracks essentially always run on through taxi-in. The outermost bin on each side is an overflow bin and carries everything beyond half an hour."
b <- bhist[bhist$period == "2025", ]
b$mid  <- (b$bin_lower_s + b$bin_upper_s) / 2
b$edge <- factor(b$edge, levels = c("off", "land"),
                 labels = c("departure: track start − take-off",
                            "arrival: track end − landing"))
b$arm  <- factor(b$arm, levels = c("legacy", "recommended", "ground_anchored"))
ggplot(b, aes(x = mid / 60, y = n, colour = arm)) +
  geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey55") +
  geom_step(direction = "mid", linewidth = 0.5) +
  facet_wrap(~edge, ncol = 2) +
  labs(x = "signed offset (minutes)", y = "flights", colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

## The sample {#sec-sample}

Two three-day samples, `r num(N_GT_25)` and `r num(N_GT_24)` ground-truth
flights: 5–7 June 2025 and 5–7 June 2024. Two periods, because a single sample
cannot distinguish a real effect from one the sample happened to favour, and
every conclusion on this page is required to hold on both.

Ground truth is the Network Manager flight table, cross-checked against APDF
where APDF covers the aerodrome. Take-off is inferred as `AOBT_3 + TAXI_TIME_3`
and landing is taken from `ARVT_3` directly; both were validated against real
APDF movement times before being relied on, at median error 0 s with an
interquartile range of 17 s and 25 s respectively.

## The ladder {#sec-ladder}

Eight arms, each a grouping and a break predicate, all scored against the same
ground truth on the same samples. Every arm reads one shared cleaned track
table with only `track_id` varying, so a difference between arms cannot be an
artefact of a different track build.

```{r ladder-table}
#| echo: false
t <- arms[arms$period == "2025", c("arm", "clean_match_pct", "fragmented_pct",
                                   "merged_pct", "n_tracks", "v_measure")]
t <- t[order(-t$clean_match_pct), ]
t24 <- arms[arms$period == "2024", c("arm", "clean_match_pct")]
names(t24)[2] <- "clean_2024"
t <- merge(t, t24, by = "arm", sort = FALSE)
t <- t[order(-t$clean_match_pct), ]
kable(t, row.names = FALSE, digits = c(0, 2, 2, 2, 0, 4, 2),
      col.names = c("arm", "clean %", "fragmented %", "merged %",
                    "tracks", "V-measure", "clean % (2024)"),
      format.args = list(big.mark = ","))
```

The ordering is identical on both periods and no arm moves by more than about
three points between them, so what follows is a property of the algorithms
rather than of one sample.

### A0 `legacy` — the shipped rule

`r pct(CLEAN_LEGACY_25)` clean, `r pct(FRAG_LEGACY_25)` fragmented,
`r pct(A("legacy", "2025", "merged_pct"))` merged. The failure is overwhelmingly
fragmentation, and it is worth being precise about the cause, because the
obvious reading is wrong. It is not that callsigns change mid-flight. It is that
**a null callsign forms its own group**: 42% of legacy's tracks are
blank-callsign tracks, produced by samples the receiver decoded without a
callsign. The group key turns a gap in the data into a separate flight.

### A2 `no_month_suffix` — untestable on this sample

Identical to A0 in every digit, on both periods. That is correct behaviour and
not a bug: the arm removes the `_{year}_{month}` suffix from the identifier, and
both samples sit inside a single month, so the suffix is constant and removing
it changes nothing. The consequence is that **this study cannot evaluate the
arm at all.** The suffix can only fragment a flight airborne at midnight on the
first of a month, and neither sample contains one. Reported here as untested
rather than as a null result; testing it needs a sample spanning a month
boundary, which is two days of cluster time and is recommended as follow-up.

### A3 `traffic_style` — the reference library's rule does not transfer

`r pct(A("traffic_style", "2025", "clean_match_pct"))` clean, below legacy, and
with the worst merging of any arm at
`r pct(A("traffic_style", "2025", "merged_pct"))` — roughly twenty times
legacy's. A single ten-minute gap threshold with a predicate deciding whether a
candidate gap is really a break is well suited to the *traffic* library's
typical input; on this feed it merges. A genuine negative result: the reference
implementation's approach does not transfer unmodified.

### A4 `airframe_only` — group on the airframe

`r pct(A("airframe_only", "2025", "clean_match_pct"))` clean, fragmentation down
from `r pct(FRAG_LEGACY_25)` to
`r pct(A("airframe_only", "2025", "fragmented_pct"))`. This is the single
largest gain in the study and it comes from deleting one column from the group
key. Identity belongs to the airframe; the callsign is something the airframe
was broadcasting at the time.

The cost is merging, up from `r pct(A("legacy", "2025", "merged_pct"))` to
`r pct(MERGE_AIR_25)`: without callsign in the key, two flights of the same
airframe with no sufficient gap between them join into one track.

### A5 `ground_anchored` and A6 `airport_anchored`

Both replace or supplement the altitude proxy with something more physical — a
ground-contact dwell long enough to be a turnaround, and real airport geometry
respectively. Both beat legacy substantially
(`r pct(A("ground_anchored", "2025", "clean_match_pct"))` and
`r pct(A("airport_anchored", "2025", "clean_match_pct"))`) and both are worth
less than simply fixing the group key. Their signed boundary error is however
the tightest of any arm — `ground_anchored`'s tenth percentile is
`r sprintf("%+d", as.integer(A("ground_anchored", "2025", "off_signed_p10_s")))` s
against legacy's
`r sprintf("%+d", as.integer(A("legacy", "2025", "off_signed_p10_s")))` s —
which is direct evidence the arm does what it claims: it cuts at the ground
event rather than running back into the ramp.

### A7 `vertical_profile` — a negative result, correctly attributed

`r pct(A("vertical_profile", "2025", "clean_match_pct"))` clean, last of the
eight. The arm splits on a descent that reaches a floor followed by a climb back
through it. An earlier implementation split on *any* upward crossing, which is
not the idea, and its poor score was therefore unattributable — it could have
been the idea failing or the code not being the idea. Implemented properly, and
on height above field rather than sea level, it scores *worse* still. That makes
it a real negative result rather than an unexplained number: every percentile of
its signed off-block error is positive, meaning it systematically starts tracks
late, having sliced off each departure's initial climb.

### A8 `recommended` — both fixes at once

`r pct(CLEAN_REC_25)` clean on 2025 and `r pct(CLEAN_REC_24)` on 2024, the best
of the eight on both, with merging at `r pct(MERGE_REC_25)` — lower than legacy
itself.

It beats both of the arms it draws on rather than landing between them, which is
the sign that the two failures really do have different causes. Legacy fragments
because null callsigns form their own groups; `airframe_only` merges because
nothing separates two flights of one airframe. So: group on `icao24`, so a blank
or flickering callsign no longer starts a track, and break when the last
*non-blank* callsign differs from the current one, so two genuinely different
flights still part.

The subtlety is in the lookback. The break predicate is evaluated over the whole
airframe's samples and has no notion of the track boundaries it is deciding, so
an unbounded search for "the previous real callsign" reaches back past a gap
into the previous flight and splits the next one in its own middle. Bounding the
lookback to the gap threshold is what makes the arm work; without it the same
rule scores 62% rather than 91%.

## The shipped algorithm's tuning headroom {#sec-sweep}

A1 is not new code: it is A0 at parameters other than production's, so the sweep
measures the shipped algorithm's own tuning and nothing else.

`r num(SWEEP_CELLS)` cells on one day, spanning
`r sprintf("%.2f", SWEEP_SPREAD1)` points end to end.

```{r sweep-axes}
#| echo: false
by_axis <- function(col, label) {
  agg <- aggregate(clean_match_pct ~ get(col), data = sweep1, FUN = max)
  names(agg) <- c(label, "best clean %")
  agg
}
kable(by_axis("low_alt_ft", "low_alt_ft"), row.names = FALSE, digits = 2,
      caption = "Best cell at each low-altitude ceiling. The optimum is interior.")
```

All three axes have an interior optimum, which is the condition for calling a
result an optimum at all. That took an extension: the first grid put the
low-altitude ceiling's best value at its largest tried, with the curve still
rising. An optimum at a grid edge is not an optimum, it is where the grid
stopped, and this pipeline's own history contains a sweep that stopped early and
was read as "higher is always better". Extended, the ceiling peaks at 20,000 ft
and falls again by 30,000 — as it must, since past cruise altitude every sample
counts as "low" and the two-threshold rule collapses into one.

Re-measured on the full three-day samples, the best cell beats production by
**`r sprintf("%+.2f", SWEEP_BEST_25 - SWEEP_PROD_25)`** points on 2025 and
**`r sprintf("%+.2f", SWEEP_BEST_24 - SWEEP_PROD_24)`** on 2024. Both periods
choose the same cell out of sample.

That is the answer to the tuning question, and it is a clean negative: **the
shipped segmentation is not mistuned.** It sits a fraction of a point from the
best setting available to it, in a grid drawn wider than plausible and extended
until every axis turned over. Set against the
`r sprintf("%+.0f", CLEAN_REC_25 - CLEAN_LEGACY_25)` points the identity key is
worth, the entire tuning question is a rounding error.

## What a better segmentation buys ADEP/ADES {#sec-payoff}

The point of segmenting well is the product built on top of it. Each arm's
assignment is fed to the V7 flight list at its own recommended configuration,
with only `track_id` varying, so any difference in ADEP/ADES accuracy is
attributable to the segmentation.

The control is `legacy`, which reproduces the frozen algorithm: run through this
harness it scores `r num(PAY_LEGACY_25)` against V7's published 134,168 for the
same sample — a difference of 8, or 0.006%. That residual is itself informative
and is explained in [Limitations](#sec-limitations).

### The ranking inverts

```{r payoff-table}
#| echo: false
p <- payoff[payoff$period == "2025", ]
p$total <- p$adep_score + p$ades_score
p$vs_legacy <- p$total - PAY_LEGACY_25
p <- p[order(-p$total), c("arm", "total", "vs_legacy", "adep_coverage",
                          "adep_accuracy")]
p$adep_coverage <- 100 * p$adep_coverage
p$adep_accuracy <- 100 * p$adep_accuracy
kable(p, row.names = FALSE, digits = c(0, 0, 0, 2, 2),
      col.names = c("arm", "ADEP+ADES", "vs legacy", "ADEP coverage %",
                    "ADEP accuracy %"),
      format.args = list(big.mark = ","))
```

The order is close to exactly reversed. The arm that reproduces individual
flights best is *last*; the shipped algorithm, second-worst at segmentation, is
first. Taken at face value this says a better segmentation makes the product
worse.

It is not what it says. **Accuracy barely moves across the whole table while
coverage collapses** — the losing arms are not predicting *wrongly*, they are
predicting for far fewer flights. That is the signature of a labelling failure,
not a geometric one.

### One aggregate explains it

The flight list builds one flight per `track_id` and labels it
`F.min("flight_id")` — the lexicographically smallest callsign in the track. A
blank sorts first.

While callsign is part of the track's group key, every track is
callsign-homogeneous by construction and that aggregate cannot lose anything:
measured on 2025, legacy loses a real callsign on **0 of 235,949 tracks**. Drop
callsign from the key and **64% of `airframe_only`'s tracks carry a real
callsign alongside blank ones**, and `F.min` returns the blank. The flight
becomes unidentifiable and never matches its ground-truth counterpart.

So the coupling is tighter than "one callsign per track". The flight list
requires **blank-callsign samples to sit in tracks of their own** — precisely
what the legacy group key does as a side effect, and precisely what any fix to
the fragmentation must undo. It is the same property, seen from both ends:
legacy's 50% fragmentation and the downstream coverage collapse are one
phenomenon.

### With the labelling repaired

Relabelling each track with its most frequent non-blank callsign, and carrying
`legacy` as a control that must not move:

```{r fix-table}
#| echo: false
rows <- do.call(rbind, lapply(PERIODS, function(p) {
  data.frame(
    period = p,
    arm = c("legacy (control)", "recommended"),
    before = c(P(payoff, "legacy", p), P(payoff, "recommended", p)),
    after  = c(P(fixcs, "legacy", p), P(fixcs, "recommended", p)),
    check.names = FALSE)
}))
rows$delta <- rows$after - rows$before
kable(rows, row.names = FALSE, digits = 0,
      col.names = c("period", "arm", "as the pipeline stands",
                    "labelling repaired", "delta"),
      format.args = list(big.mark = ","))
```

The control returns **exactly +0**, on both periods and across three independent
runs. That is what makes the other row attributable to the relabelling rather
than to anything else the run touched.

With the labelling repaired the recommended segmentation scores
**`r num(FIX_REC_25)`** against legacy's **`r num(FIX_LEGACY_25)`** on 2025 and
**`r num(FIX_REC_24)`** against **`r num(FIX_LEGACY_24)`** on 2024 — short by
`r num(FIX_LEGACY_25 - FIX_REC_25)` and `r num(FIX_LEGACY_24 - FIX_REC_24)`
points, 0.5% and 0.6% — while matching flights around
`r sprintf("%.0f", CLEAN_REC_25 - CLEAN_LEGACY_25)` points better. Its ADEP
coverage slightly *exceeds* legacy's on both periods.

The residual is real and is not rounded away: with the labelling fixed, what
remains is merging, and the recommended arm still merges
`r pct(MERGE_REC_25)` of flights where legacy merges
`r pct(A("legacy", "2025", "merged_pct"))`.

## What ships {#sec-defaults}

**Recommended: the A8 rule — group on `icao24`, break on a genuine non-blank
callsign change with the lookback bounded to the gap threshold, keeping legacy's
gap rules as a floor.**

Conditional on one change to the flight list, and the order matters:

1. **First**, change `flights.py`'s flight labelling so it does not choose a
   callsign with `F.min`. Most-frequent non-blank is what this study measured;
   any rule that ignores blanks would do. This is worth doing on its own merits
   even if the segmentation never changes, because it is a silent trap for any
   future change to track identity and it fails as a coverage statistic rather
   than as an error.
2. **Then** ship the segmentation.

Reversed, the pipeline spends the interval between the two changes at 55%
ADEP coverage instead of 80%.

Two consequences of shipping it, both of which are decisions for the release
plan rather than for this study:

- **`track_id` changes for all data from that release forward**, in shape as
  well as value: A8 carries no `_{year}_{month}` suffix, so identifiers become
  `{hash}_{offset}`. Any consumer parsing the suffix breaks in a way a pure
  value change would not.
- **Past months will not reproduce.** This study recommends; it does not
  schedule.

## Provenance {#sec-provenance}

Each row is one file this page reads: the job that produced it, the commit it
was produced at, and when. Every figure on this page is regenerated by
`benchmarks/regenerate_track_v1.py`, which the page runs before drawing
anything, so a row marked **unverified** would mean a file staged outside that
chain and traceable to no command.

```{r provenance}
#| echo: false
#| results: asis
mf <- file.path("data", "_manifest.json")
if (!requireNamespace("jsonlite", quietly = TRUE)) {
  cat("*jsonlite not available; provenance table skipped.*\n")
} else if (!file.exists(mf)) {
  cat("**No manifest.** Every CSV here was staged by hand and none can be traced to the command that produced it.\n")
} else {
  m <- jsonlite::fromJSON(mf, simplifyDataFrame = FALSE)
  files <- sort(list.files("data", pattern = "[.]csv$"))
  rows <- do.call(rbind, lapply(files, function(f) {
    e <- m[[f]]
    if (is.null(e)) {
      data.frame(File = f, Job = "**unverified**", Commit = "—",
                 Produced = "—", check.names = FALSE)
    } else {
      data.frame(File = f, Job = basename(e$script),
                 Commit = paste0(e$git_sha, if (isTRUE(e$git_dirty)) " (dirty)" else ""),
                 Produced = substr(e$produced_utc, 1, 16), check.names = FALSE)
    }
  }))
  print(kable(rows, row.names = FALSE, align = "l"))
  n_un <- sum(rows$Job == "**unverified**")
  cat(sprintf("\n%d files, %d traced to a command, %d unverified.\n",
              nrow(rows), nrow(rows) - n_un, n_un))
}
```

## Limitations {#sec-limitations}

### The baseline is not the shipped `track_id`

A0 reproduces the legacy *rule*, but applied to cleaned tracks it does not
reproduce the shipped identifiers exactly: it differs on 909 of 236,899 tracks,
0.38%, and 8 of 134,176 ADEP/ADES points. The cause is exact and worth reporting
upward on its own account. Production's identifiers were computed upstream on
raw altitudes, while `osn_tracks_clean` masks bad altitudes to NULL — and
`NULL < 5000` is NULL, not true, so the low-altitude split silently does not fire
where it once did. Measured, **950 of 950 disagreements have a NULL barometric
altitude**, against a 21.4% base rate.

The published `osn_tracks_clean` therefore carries a `track_id` that its own
columns can no longer reproduce. Nothing is wrong with the published data, but
anyone recomputing identifiers from the clean table — the obvious thing to do —
gets different tracks and no warning.

This does not affect the comparisons on this page: every arm reads the same
cleaned table and is subject to the same behaviour.

### Ground truth covers Network Manager flights only

Merges with traffic absent from the NM flight table — military, general
aviation, non-ECAC operators — are invisible to the metric, because the merged
partner has no ground-truth flight to be merged with. **The merge rates on this
page are therefore floors, not estimates.** Fragmentation is unaffected, since
it is measured within flights that ground truth does contain.

### One arm is untested, not null

`no_month_suffix` is identical to legacy on both samples because both sit inside
a single month. It is reported as untested. Evaluating it needs a sample
spanning a month boundary.

### Two three-day samples in June

Both samples are in early June. Nothing here licenses a seasonal claim, and the
callsign statistics in particular could plausibly differ in a period with more
charter or seasonal traffic.

### The A1 sweep varied three parameters

The sweep covers the gap threshold, the low-altitude gap threshold and the
low-altitude ceiling. It does not vary the ground-dwell or turnaround
parameters that the domain arms introduce, so those arms are compared at one
setting each and could be better at another. Given the structural gap between
the arms is two orders of magnitude larger than anything the sweep found, this
is unlikely to change the ordering, but it is not measured.

### The payoff holds the flight list fixed

Each arm is scored at V7's own recommended flight-list configuration rather than
re-tuned per arm. A joint optimum might be better for some arm; finding it would
multiply cluster time by the number of arms, and one Spark job runs at a time.
The natural follow-up is to re-tune the flight list on the recommended arm alone
— one extra grid, not eight — once the labelling is fixed.
