---
title: "Turning ADS-B state vectors into flight tracks"
subtitle: "Version 1 — what the segmentation is worth, and what had to change downstream before a better one could ship"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r regenerate}
#| include: false
#| cache: false

# Rendering this page runs the analysis behind it.
#
# `regenerate_track_v1.py` is the executable definition of the study: every CSV
# read below is produced by exactly one job it declares, with exactly the
# arguments it declares. It re-runs a job when the *code that job depends on*
# has changed since the output was written -- not when the file is old, which
# says nothing. So a render with everything current is a fast no-op, and a
# render after editing `segmentation/methods.py` recomputes every arm before the
# first table is drawn.
#
# Running a stale job needs cluster credentials, because the numbers come from
# Spark over S3 against Network Manager reference data and there is no way to
# recompute them without the data they are computed from. Set OPDI_RENDER=check
# to fail fast on staleness instead, or OPDI_RENDER=allow-stale to render a
# draft from whatever is staged -- which the provenance table at the foot of
# this page will then show as unverified.

# The pipeline repo is a sibling of the portal. That resolution is wrong inside
# a git worktree, where the paper sits under .claude/worktrees/<name>/papers and
# the sibling walk lands somewhere that does not exist -- so the regeneration
# step would be skipped silently and the page would render whatever is staged.
# OPDI_REPO_DIR overrides it; the default is unchanged for a normal checkout.
REPO   <- Sys.getenv("OPDI_REPO_DIR",
                     normalizePath(file.path(getwd(), "..", "..", "..", "opdi"),
                                   mustWork = FALSE))
PYTHON <- Sys.getenv("OPDI_PYTHON", file.path(REPO, ".venv310", "bin", "python"))
MODE   <- Sys.getenv("OPDI_RENDER", "run")

# A missing entrypoint must not read as "nothing to regenerate". Skipping it
# silently is how a page ends up publishing numbers no command can account for.
if (!file.exists(file.path(REPO, "benchmarks", "regenerate_track_v1.py")) &&
    MODE != "skip") {
  stop("regenerate_track_v1.py not found under REPO=", REPO,
       " -- set OPDI_REPO_DIR, or OPDI_RENDER=skip to render without it.")
}

regen <- file.path(REPO, "benchmarks", "regenerate_track_v1.py")
if (file.exists(regen) && MODE != "skip") {
  flags <- switch(MODE,
                  "run" = character(0),
                  "check" = "--check",
                  "allow-stale" = c("--check", "--allow-stale"),
                  character(0))
  status <- system2(PYTHON, c(shQuote(regen), flags), stdout = "", stderr = "")
  if (!identical(status, 0L)) {
    stop("regenerate_track_v1.py exited ", status,
         " -- the figures below would not match the code that is checked out. ",
         "Set OPDI_RENDER=allow-stale to render anyway.")
  }
}
```

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every figure and table on this page comes from a committed CSV
# under data/. Nothing here reads S3, a database or a Spark cluster.

D <- function(f) read.csv(file.path("data", f), stringsAsFactors = FALSE)

arms   <- rbind(D("arms_2025.csv"), D("arms_2024.csv"))
sweep1 <- rbind(D("sweep_2025_stage1.csv"), D("sweep_2025_stage1_ext.csv"))
sweep2 <- rbind(D("sweep_2025_stage2.csv"), D("sweep_2024_stage2.csv"))
payoff <- rbind(D("payoff_2025.csv"), D("payoff_2024.csv"))
fixcs  <- rbind(D("payoff_fixcallsign_2025.csv"), D("payoff_fixcallsign_2024.csv"))
cont   <- rbind(D("containment_2025.csv"), D("containment_2024.csv"))

# The boundary histograms carry no period column -- one file per period *is* the
# period -- so it is attached on read rather than assumed by anything later.
readhist <- function(p) {
  x <- D(paste0("boundary_hist_", p, ".csv"))
  x$period <- p
  x
}
bhist <- rbind(readhist("2025"), readhist("2024"))

num <- function(x) formatC(x, format = "d", big.mark = ",")
# The "+" flag does nothing for %s in R, so the sign is prepended explicitly --
# a signed delta that silently loses its sign reads as a level, not a change.
sgn <- function(x) paste0(if (x >= 0) "+" else "-", num(abs(x)))
pct <- function(x) sprintf("%.2f%%", x)
pp  <- function(x) sprintf("%+.2f", x)
sec <- function(x) sprintf("%+d", as.integer(x))

# One accessor per file, so no two tables on this page can disagree about what
# an arm scored.
A  <- function(a, p, col) arms[arms$arm == a & arms$period == p, col]
CT <- function(p, col) cont[cont$period == p, col]
PR <- function(df, a, p, col) df[df$arm == a & df$period == p, col]
P  <- function(df, a, p) {
  r <- df[df$arm == a & df$period == p, ]
  r$adep_score + r$ades_score
}

PERIODS <- c("2025", "2024")

# --- headline values, computed here rather than written into the prose, so the
# --- opening paragraph cannot drift away from the tables that justify it.
CLEAN_LEGACY_25 <- A("legacy", "2025", "clean_match_pct")
CLEAN_REC_25    <- A("recommended", "2025", "clean_match_pct")
CLEAN_LEGACY_24 <- A("legacy", "2024", "clean_match_pct")
CLEAN_REC_24    <- A("recommended", "2024", "clean_match_pct")
FRAG_LEGACY_25  <- A("legacy", "2025", "fragmented_pct")
FRAG_REC_25     <- A("recommended", "2025", "fragmented_pct")
MERGE_LEGACY_25 <- A("legacy", "2025", "merged_pct")
MERGE_REC_25    <- A("recommended", "2025", "merged_pct")
MERGE_AIR_25    <- A("airframe_only", "2025", "merged_pct")
GAIN_25         <- CLEAN_REC_25 - CLEAN_LEGACY_25
GAIN_24         <- CLEAN_REC_24 - CLEAN_LEGACY_24
N_GT_25         <- A("legacy", "2025", "n_flights")
N_GT_24         <- A("legacy", "2024", "n_flights")
N_APDF_25       <- A("legacy", "2025", "n_apdf_flights")

# The sweep: legacy at its production parameters, against the best cell anywhere
# in the grid.
PROD <- function(df, p) {
  r <- df[df$period == p & df$gap_minutes == 30 &
          df$low_alt_gap_minutes == 15 & df$low_alt_ft == 5000, ]
  r$clean_match_pct[1]
}
SWEEP_PROD_25 <- PROD(sweep2, "2025")
SWEEP_BEST_25 <- max(sweep2$clean_match_pct[sweep2$period == "2025"])
SWEEP_PROD_24 <- PROD(sweep2, "2024")
SWEEP_BEST_24 <- max(sweep2$clean_match_pct[sweep2$period == "2024"])
SWEEP_CELLS   <- nrow(sweep1)
SWEEP_SPREAD1 <- max(sweep1$clean_match_pct) - min(sweep1$clean_match_pct)
SWEEP_GAIN_25 <- SWEEP_BEST_25 - SWEEP_PROD_25
SWEEP_GAIN_24 <- SWEEP_BEST_24 - SWEEP_PROD_24
# How many times larger the identity fix is than the entire tuning question.
SWEEP_RATIO   <- GAIN_25 / SWEEP_GAIN_25
# The largest movement of any arm's clean-match rate between the two periods.
# Quoted rather than asserted, because "the ordering holds on both samples" is
# the claim the second period was bought to support.
ARM_SHIFT <- max(abs(sapply(unique(arms$arm), function(a)
  A(a, "2025", "clean_match_pct") - A(a, "2024", "clean_match_pct"))))

# The payoff, and the convergence check that now sits beside it.
PAY_LEGACY_25 <- P(payoff, "legacy", "2025")
PAY_REC_25    <- P(payoff, "recommended", "2025")
PAY_LEGACY_24 <- P(payoff, "legacy", "2024")
PAY_REC_24    <- P(payoff, "recommended", "2024")
PAY_SHORT_25  <- PAY_LEGACY_25 - PAY_REC_25
PAY_SHORT_24  <- PAY_LEGACY_24 - PAY_REC_24
FIX_LEGACY_25 <- P(fixcs, "legacy", "2025")
FIX_REC_25    <- P(fixcs, "recommended", "2025")
FIX_LEGACY_24 <- P(fixcs, "legacy", "2024")
FIX_REC_24    <- P(fixcs, "recommended", "2024")

# The one figure on this page that comes from another paper rather than from a
# CSV here: the V7 flight-list study's own published ADEP+ADES total for the same
# sample. Held as a named constant so the difference below is arithmetic on it
# rather than a second hand-typed number.
V7_PUBLISHED <- 134168

# Where each swept axis peaks, read off the grid rather than described from
# memory. The low-altitude ceiling is the axis that needed the grid extended.
axis_peak <- function(col) {
  a <- aggregate(clean_match_pct ~ get(col), data = sweep1, FUN = max)
  a[[1]][which.max(a[[2]])]
}
LOALT_PEAK <- axis_peak("low_alt_ft")
LOALT_MAX  <- max(sweep1$low_alt_ft)
LOALT_BASE_MAX <- max(D("sweep_2025_stage1.csv")$low_alt_ft)
```

## Summary {.unnumbered}

**The rule that built every `track_id` OPDI published up to v0.0.2 recovers
about half the flights it should, the cause is one column in a grouping key
rather than a mistuned threshold, and a rule that fixes it has now shipped.**
That is the whole finding. The rest of this page is how it was established, and
what had to be repaired in the flight list before the better rule could be
adopted without making the published product worse.

Before OPDI can derive a flight list it has to decide which state vectors belong
to the same flight. That decision — **segmentation** — is made once, and
everything downstream inherits it: the flight list groups by `track_id`, events
are extracted per track, measurements accumulate from a track's start. This page
asks what the decision is worth, in four parts.

**The old rule recovers about half the flights it should.** On a three-day
sample of `r num(N_GT_25)` Network Manager flights it matches
**`r pct(CLEAN_LEGACY_25)`** of them one-to-one, meaning one flight, one track,
that track holding nothing else; **`r pct(FRAG_LEGACY_25)`** of flights are split
across more than one track. The failure is almost entirely that splitting, and
almost never the opposite error of running two flights together.

**Its parameters are not the problem.** Its three thresholds were swept over
`r num(SWEEP_CELLS)` combinations, every axis widened until the curve turned over
inside the grid rather than at its edge. The whole grid spans
`r sprintf("%.2f", SWEEP_SPREAD1)` points of one-to-one matching, and on the full
three-day samples the best combination beats the production setting by only
**`r pp(SWEEP_GAIN_25)`** points on 2025 and **`r pp(SWEEP_GAIN_24)`** on 2024.
The rule was already at, or a fraction of a point from, the best setting anyone
could have picked for it.

**The identity key is the problem.** The old group key is
`(icao24, callsign)`, so a state vector decoded without a callsign forms a group
— and therefore a track — of its own. Grouping on the airframe alone, and
splitting only when the callsign *genuinely* changes, takes one-to-one matching
to **`r pct(CLEAN_REC_25)`** on 2025 and **`r pct(CLEAN_REC_24)`** on 2024: a
gain of **`r pp(GAIN_25)`** points, about `r sprintf("%.0f", SWEEP_RATIO)` times
everything the parameter sweep had to offer. This is the rule that now ships, as
`segmentation.method = "standard"`.

**The flight list could not consume it until one aggregate was fixed, and that
is the most useful finding here.** `flights.py` labelled each flight with the
lexicographically smallest callsign in its track, and a blank sorts first. That
is harmless only while every track carries exactly one callsign — a property the
old group key supplied by accident, and which any fix to the fragmentation
necessarily destroys. Measured against the pipeline as it then stood, the better
segmentation therefore made the published product *worse*: not by predicting
wrong aerodromes, but by producing flights that could no longer be identified at
all. That labelling is now fixed in production, and the numbers on this page are
the fixed ones.

With it fixed, the better segmentation costs
**`r num(PAY_SHORT_25)`** ADEP/ADES points on 2025 and **`r num(PAY_SHORT_24)`**
on 2024 — `r pct(100 * PAY_SHORT_25 / PAY_LEGACY_25)` and
`r pct(100 * PAY_SHORT_24 / PAY_LEGACY_24)` of the total, where a *point* is the
flight list's own score for one correctly named aerodrome and a wrong one costs
two — while recovering `r sprintf("%.0f", GAIN_25)` points more flights. The
residual is not accuracy, which is unchanged, and it is not departures, where
the new rule is slightly ahead. It is arrival coverage, and the one failure mode
the new rule makes worse is merging.

## How to read this page

Sections are ordered as the work was done, not as the conclusion would suggest,
because two of the results only make sense as answers to questions asked in
order.

[What a track is](#sec-problem) defines the object and says why the question was
open. [What has already happened to a sample](#sec-pipeline) is a map of the
pipeline steps that run before any segmentation sees a state vector; skip it if
you already know them. [How a segmentation is scored](#sec-metrics) matters more
than usual here, because the obvious scoring choice is misleading and this page
says why before using something else; it also defines every term the later
sections use. [The sample](#sec-sample) says what ground truth is and where it
comes from. [The ladder](#sec-ladder) is one subsection per candidate rule, each
with what it changes, what it is trying to fix, the rule itself in enough detail
to reimplement, and what it scored. [The sweep](#sec-sweep) asks whether the old
rule was merely mistuned, and answers no. [The payoff](#sec-payoff) measures the
downstream effect on ADEP/ADES and explains why it does not track segmentation
quality. [What shipped](#sec-defaults) states what was adopted and what
consumers have to know about it.

Every number on this page is read from a committed CSV, regenerated by the
entrypoint the page runs before drawing anything;
[Provenance](#sec-provenance) lists the files one by one.
[Limitations](#sec-limitations) is where the claims this page does *not* make
are written down.

## What a track is, and why the question was open {#sec-problem}

*This section defines the object under study and the three ways a rule for
building it can fail. By the end of it you will know what the rule under test
was, and what "better" is going to mean.*

OPDI ingests OpenSky state vectors — position reports, one aircraft at a time,
with no notion of a flight. A **track** is the pipeline's answer to "which of
these belong together", and it is the unit everything downstream is built on:
the flight list groups by `track_id`, events are extracted per track, and
measurements accumulate from a track's start. The procedure that produces tracks
is called a **segmentation**, and a **track identifier** (`track_id`) is what it
stamps on every sample it decides belongs to the same flight.

The rule this study set out to evaluate — called **`legacy`** throughout, and
the rule behind every dataset OPDI published up to v0.0.2 — is short. Group the
samples by `(icao24, callsign)`, order them by time, and start a new track when
the gap to the previous sample exceeds 30 minutes, or exceeds 15 minutes while
the aircraft is below 5,000 ft. The resulting identifier carries a
`_{year}_{month}` suffix.

Three things can go wrong with such a rule, and they are not symmetric:

- **Fragmentation** — one real flight becomes several tracks. Everything
  downstream sees several short flights where there was one.
- **Merging** — several real flights become one track. The flight list emits one
  flight and the others simply do not exist in the output.
- **Boundary error** — the track is one flight, but starts or ends in the wrong
  place, so first- and last-seen events and cumulative measurements are wrong.

Of the three, merging is the most damaging, because a merged flight is
unrecoverable downstream, while a fragmented one is at least present in pieces
and a boundary error leaves the flight intact. The metrics below are chosen so
that merging and fragmentation cannot be confused with each other, and the
candidate rules are compared on both rather than on a single blended score.

Until this study, the rule had never been benchmarked, and that is why the study
exists.

What it is *not* is frozen. Segmentation is a versioned choice: a release states
which rule built its `track_id` values, and changing that rule changes
`track_id` for everything produced from that release forward — in value, and in
shape where the month suffix goes. That is a release decision, to be documented
and communicated to consumers rather than avoided, and it is a decision that has
now been taken. As of this release `segmentation.method` defaults to
`"standard"`, which is the rule this study recommends. `"legacy"` remains
selectable and reproduces the identifiers of every dataset published up to OPDI
v0.0.2, so released data stays reproducible.

One consequence for the reader, stated here so nothing later depends on
inferring it. **This paper measures algorithms, not the pipeline.** Every rule
below is expressed in a generic segmentation engine and scored in a harness, so
that eight candidates could be compared without eight production pipeline runs —
which at this data volume is the difference between a study and a quarter. That
is a legitimate design and it is also a real limitation: what is measured here
is a re-implementation of each rule, not the pipeline's own output. Version 2 of
this study is the paper that runs the pipeline itself, end to end, on the rule
this one recommends. If you want to know what a rule is worth, you are in the
right place; if you want to know what the pipeline now produces, read version 2.

## What has already happened to a sample {#sec-pipeline}

*Side information. This section says what the `/opdi` pipeline has already done
to a state vector before any segmentation sees it, so that nothing later in this
page is a surprise. If you know the pipeline steps, skip to
[the metrics](#sec-metrics).*

Segmentation is step 02 of a chain, and the samples reaching it are not raw
OpenSky rows. Two earlier decisions in particular shape everything measured on
this page: the ingestion filter decides which samples exist at all, and the
**thinning** rule — which sample survives out of each short interval — decides
how densely a flight is described.

```{r pipeline-diagram}
#| echo: false
#| fig-width: 7
#| fig-height: 4.2
#| fig-cap: "The pipeline steps that run before, and after, the decision this page studies. Step 02 is the one under test; everything above it has already happened to a sample by then, and everything below it inherits the result."
# Drawn with ggplot rather than as a mermaid block, deliberately. A mermaid
# diagram written as an executable cell renders in HTML but needs a headless
# browser to rasterise for PDF, and that browser does not start in this
# environment -- the render then hangs with no error. A mermaid fence emitted
# from knitr avoids the hang but arrives without the JavaScript that draws it,
# so it shows as literal text. A plot goes through whatever device Quarto has
# already selected and is correct in both formats.
nodes <- data.frame(
  y = 6:1,
  label = c(
    "step 00 — reference tables (airports, aircraft)",
    "step 01 — ingest OSN state vectors: bbox + thinning",
    "step 02 — tracks: assign track_id",
    "step 02a — cleaning: mask bad values to NULL",
    "step 03 — flight list: ADEP / ADES per track",
    "step 04 — events and measurements per track"),
  under_test = c(FALSE, FALSE, TRUE, FALSE, FALSE, FALSE))
h <- 0.34
ggplot(nodes) +
  geom_rect(aes(xmin = 0, xmax = 1, ymin = y - h, ymax = y + h,
                fill = under_test, colour = under_test, linewidth = under_test)) +
  geom_text(aes(x = 0.5, y = y, label = label, fontface = ifelse(under_test, 2, 1)),
            size = 3.1) +
  geom_segment(data = nodes[nodes$y > 1, ],
               aes(x = 0.5, xend = 0.5, y = y - h, yend = y - 1 + h),
               arrow = arrow(length = unit(1.8, "mm"), type = "closed"),
               colour = "grey40", linewidth = 0.3) +
  annotate("text", x = 1.04, y = 4, hjust = 0, size = 2.9, colour = "grey25",
           label = "this study replaces\nthe rule inside step 02") +
  scale_fill_manual(values = c(`FALSE` = "white", `TRUE` = "grey90")) +
  scale_colour_manual(values = c(`FALSE` = "grey55", `TRUE` = "grey15")) +
  scale_linewidth_manual(values = c(`FALSE` = 0.3, `TRUE` = 0.8)) +
  coord_cartesian(xlim = c(0, 1.5), ylim = c(0.4, 6.6), clip = "off") +
  theme_void() +
  theme(legend.position = "none")
```

```{r pipeline-table}
#| echo: false
steps <- data.frame(
  Step = c("00", "01", "02", "02a", "03", "04"),
  Does = c(
    "Builds reference tables: H3 airport detection zones, HexAero airport ground layouts, OurAirports, the OpenSky aircraft database. Read later; never rewritten by a pipeline run.",
    "Downloads OpenSky state vectors, restricts them to the OPDI bounding box, and thins them to one sample per aircraft per interval.",
    "Groups samples into tracks and stamps `track_id`. **This is the step under study.** Also attaches the H3 index each later join uses.",
    "Masks values that cannot be right — duplicates, out-of-range positions, stale broadcasts, derivative spikes, isolated points — to NULL, keeping the row. Writes a second table rather than editing the first.",
    "One flight per `track_id`: departure and destination aerodrome, and the flight's callsign label.",
    "Milestones per track — take-off, top-of-climb, level segments, landing — and the measurements attached to them."
  ),
  Output = c("reference tables", "`osn_statevectors`", "`osn_tracks`",
             "`osn_tracks_clean`", "`opdi_flight_list`", "events, measurements"),
  check.names = FALSE)
kable(steps, row.names = FALSE, align = "lll",
      col.names = c("Step", "What it does", "Writes"))
```

Two parameters of step 01 are worth naming, because every count on this page is
conditioned on them.

```{r ingestion-params}
#| echo: false
ing <- data.frame(
  Parameter = c("bounding box (min lon, min lat, max lon, max lat)",
                "thinning interval",
                "thinning rule"),
  Value = c("(-25.86653, 26.74617, 49.65699, 70.25976)",
            "5 s",
            "`bucket`"),
  Consequence = c(
    "A flight is observed only inside this box. A track that begins at its edge may have entered from outside, which is why an out-of-area label exists at all.",
    "Consecutive samples of one aircraft are at least this far apart, so every gap threshold in this study is measured against a 5 s grid rather than a 1 Hz one.",
    "One row kept per aircraft per interval bin — the last, so a track's final observation survives exactly. The earlier `modulo` rule kept the row whose epoch second divides the interval, and built every dataset published before it."
  ),
  check.names = FALSE)
kable(ing, row.names = FALSE, align = "lll",
      col.names = c("Parameter", "Value", "Why it matters here"))
```

One ordering detail matters, because the diagram would otherwise mislead. In
production, step 02 builds tracks and step 02a cleans them, in that order. This
study runs the other way round: every candidate rule is applied to the *cleaned*
table, so that all of them see exactly the same input and a difference between
two of them cannot be an artefact of different cleaning.

That is the right choice for a comparison and it has one consequence worth
knowing. Cleaning masks bad values to NULL rather than deleting rows, and
`NULL < 5000` evaluates to NULL rather than to true — so legacy's low-altitude
split does not fire on a sample whose altitude the cleaner rejected, where it
once did. Every arm is subject to that equally, so no comparison here is
affected; but it means the identifiers in published data cannot be recomputed
from the published clean table. That is picked up again in
[Limitations](#sec-limitations).

## How a segmentation is scored {#sec-metrics}

*This section defines every measurement used later: the two failure measures,
the headline statistic, the rule that decides which track belongs to which
flight, and the sign convention for boundary error. Nothing after this section
introduces a metric.*

Comparing a segmentation against ground truth is a **clustering comparison**:
both the algorithm and the truth partition the same state vectors — the
algorithm into tracks, the truth into flights — and the question is how far the
two partitions agree.

### Homogeneity is merging; completeness is fragmentation

Two standard clustering measures map onto the two failure modes exactly, which
is why they are used here rather than a score invented for the occasion:

- **Homogeneity** asks whether each track contains samples from only one flight.
  It falls when a track holds more than one flight — so it is the *merging*
  measure.
- **Completeness** asks whether each flight's samples all landed in one track.
  It falls when a flight is spread over several tracks — so it is the
  *fragmentation* measure.

Both are entropy-based and both run from 0 to 1, with 1 meaning perfect
agreement with ground truth on that axis alone.

**V-measure** is their harmonic mean — a single number summarising both. It is
reported below and used to choose nothing, and the reason is measured rather
than assumed. Across the eight candidate rules V-measure spans
`r sprintf("%.4f", min(arms$v_measure))` to
`r sprintf("%.4f", max(arms$v_measure))`, a range of
`r sprintf("%.4f", max(arms$v_measure) - min(arms$v_measure))`, while the
one-to-one matching rate defined next spans
`r sprintf("%.1f", min(arms$clean_match_pct))` to
`r sprintf("%.1f", max(arms$clean_match_pct))` points. Worse than insensitive,
it is *misleading*: it ranks the candidates in a different order at the top,
which is the only place a comparison ever looks. The cause is that both
statistics are entropy-weighted over every sample in the period, and the
overwhelming majority of samples sit in tracks that every candidate gets right —
so V-measure is dominated by agreement that carries no information about where
the candidates differ.

### The headline: one-to-one matching

The headline statistic on this page is **`clean_match_pct`**, and it answers the
question a data consumer actually has: *what share of real flights came out as
exactly one track, holding nothing but that flight?*

Every ground-truth flight is placed in exactly one of three mutually exclusive
classes, which sum to 100%:

- **clean** — the flight's samples are in one track, and that track carries no
  other flight;
- **fragmented** — the flight's samples span several tracks, none of which
  carries another flight;
- **merged** — the flight's dominant track also carries at least one other
  flight.

A flight that is both merged and fragmented counts as **merged**, because
merging is the worse failure and a flight should not be able to improve its
classification by also being broken a second way. The definition is deliberately
strict: a single stray sample in a second track costs a flight its clean match.
The tolerant alternative — a dominant track covering some fraction of the flight
— was rejected because that fraction would be a free parameter chosen after
seeing the results. `clean_match_pct` is therefore a lower bound, and is
reported as one.

### Matching a track to a flight: interval containment {#sec-containment}

Every statistic above needs one thing first: a rule saying which ground-truth
flight a given state vector belongs to. That rule is **interval containment**,
and it is worth setting out fully, because it is where a comparison of this kind
is most easily rigged without anyone noticing.

**A ground-truth flight is an airframe and a span of time.** The airframe is
`icao24` — the same 24-bit address ADS-B broadcasts, which the Network Manager
flight table carries as `AIRCRAFT_ADDRESS`. The span is the flight's airborne
interval, from take-off to landing:

- where APDF covers the aerodrome, from APDF's measured movement times, the real
  ATOT and ALDT;
- elsewhere, inferred from Network Manager as `AOBT_3 + TAXI_TIME_3` for
  take-off and `ARVT_3` for landing. Both inferences were validated against real
  APDF movements before being relied on (see [The sample](#sec-sample)).

Each row carries which of the two it came from, so a statistic that needs
measured times rather than inferred ones — boundary error does — can restrict
itself to the APDF-backed subset.

**The join is on `icao24` and time overlap, and never on callsign.** This
differs from the ADEP/ADES flight-list studies, which key on callsign, and the
difference is forced rather than stylistic: one of the candidate rules below
removes callsign from track identity entirely, and another splits on it. Joining
on callsign would therefore score each candidate against a key it deliberately
does or does not have, which is prejudging the very thing under comparison.

**Containment, concretely.** A state vector is attributed to the ground-truth
flight of the same `icao24` whose interval holds its timestamp. A track and a
flight are then *matched* when at least one of the track's samples is attributed
to that flight, and the contingency of tracks against flights is what the three
classes above are counted from. Where two intervals of the same airframe touch,
the boundary sample goes to the earlier one, so back-to-back legs cannot
double-count it and inflate the merge rate.

A worked example, one airframe with two rotations in an afternoon:

```{r containment-example}
#| echo: false
ex <- data.frame(
  Flight = c("leg 1", "leg 1", "gap", "leg 2", "leg 2"),
  Interval = c("12:00–13:30", "12:00–13:30", "—", "14:10–15:40", "14:10–15:40"),
  Sample = c("12:05", "13:25", "13:50 (on stand)", "14:15", "15:35"),
  Attributed = c("leg 1", "leg 1", "no flight — outside both intervals",
                 "leg 2", "leg 2"),
  check.names = FALSE)
kable(ex, row.names = FALSE, align = "llll",
      col.names = c("Ground-truth flight", "Its interval", "A sample",
                    "Attributed to"))
```

If one track holds the 12:05 and 15:35 samples, both flights are **merged** into
it. If leg 1's two samples land in different tracks, leg 1 is **fragmented**.
The samples on stand at 13:50 belong to no flight and are scored against
nothing — segmentation is not penalised for what it does with them.

**What the restriction costs.** Ground truth is further restricted to flights
whose *whole* airborne interval lies inside the sampled window. The reason is
that a flight observed only for its last twenty minutes has no samples for the
rest of itself: it would score as truncated however well the segmentation
performed, and its boundary error would be a fact about where the window fell.
That argument justifies the restriction but does not price it, so the price was
measured.

```{r containment-table}
#| echo: false
ct <- data.frame(
  period = PERIODS,
  overlapping = sapply(PERIODS, function(p) CT(p, "n_gt_flights")),
  inside      = sapply(PERIODS, function(p) CT(p, "n_wholly_inside")),
  cs          = sapply(PERIODS, function(p) CT(p, "n_clipped_start")),
  ce          = sapply(PERIODS, function(p) CT(p, "n_clipped_end")),
  kept        = sapply(PERIODS, function(p) CT(p, "pct_kept")),
  frac        = sapply(PERIODS, function(p) CT(p, "median_observed_fraction_clipped")),
  check.names = FALSE)
kable(ct, row.names = FALSE, digits = c(0, 0, 0, 0, 0, 2, 3),
      col.names = c("period", "any airborne time in window", "wholly inside",
                    "clipped at start", "clipped at end", "kept %",
                    "median observed fraction of an excluded flight"),
      format.args = list(big.mark = ","))
```

So the restriction keeps `r sprintf("%.2f%%", CT("2025", "pct_kept"))` of the
2025 flights that have any airborne time in the window, and
`r sprintf("%.2f%%", CT("2024", "pct_kept"))` of the 2024 ones. The exclusion is
near-symmetric between the two window edges, which is what a window artefact
should look like and is evidence the rule is not systematically dropping one
kind of flight — long-haul arrivals, for instance.

The last column answers the obvious objection, that the rule might be discarding
nearly-complete flights on a technicality. It does not: the median excluded
flight had only about
`r sprintf("%.0f%%", 100 * CT("2025", "median_observed_fraction_clipped"))` of
its airborne time inside the window on 2025, and
`r sprintf("%.0f%%", 100 * CT("2024", "median_observed_fraction_clipped"))` on
2024. A typical excluded flight really was about half observed, and could not
have been scored fairly.

One further gap is worth stating, because it explains a difference between two
counts on this page. Of the
`r num(CT("2025", "n_wholly_inside"))` flights wholly inside the 2025 window,
`r num(N_GT_25)` are scored — the remaining
`r num(CT("2025", "n_wholly_inside") - N_GT_25)` produced no ADS-B sample at all
in the feed and so appear in no contingency. They are absent from the
denominator for the same reason, and no candidate rule is credited or penalised
for them.

### Boundary error, reported signed

Boundary error is the distance in time from a track's first and last sample to
the flight's actual take-off and landing. It is reported **signed**, with the
convention that

> **`off` = track start − take-off, `land` = track end − landing.** A negative
> value means the track began (or ended) *before* the aircraft did.

That is not cosmetic. An absolute error conflates two opposite conditions: a
track starting *before* take-off is normal and often desirable — the aircraft is
on the ramp broadcasting — while a track starting *after* take-off has missed
part of its own flight. `legacy`'s absolute median is about a hundred seconds,
which reads as a small tidy number. Signed, its median is
**`r sec(A("legacy", "2025", "off_signed_p50_s"))` s** and its tenth percentile
is **`r sec(A("legacy", "2025", "off_signed_p10_s"))` s** — which says something
quite different: the typical track begins just *after* wheels-off, while a
minority carry a full quarter-hour of taxi-out. The distribution is two-sided,
and an absolute median over it is close to meaningless.

The landing side is the opposite and far more uniform: every percentile is
positive, with a median of
**`r sec(A("legacy", "2025", "land_signed_p50_s"))` s**. Tracks essentially
always run on through taxi-in. That asymmetry is a coverage fact — on-ground
ADS-B before departure reaches a minority of flights, while an arriving aircraft
is still powered and broadcasting to the stand — and not an algorithm defect.

The percentiles below are what the text quotes. They are computed over the
flights whose ground-truth times are APDF-measured rather than NM-inferred —
`r num(N_APDF_25)` of the `r num(N_GT_25)` scored flights on 2025 — because a
boundary error against an inferred time would partly be measuring the inference.

```{r boundary-percentiles}
#| echo: false
bp <- arms[arms$period == "2025",
           c("arm", "off_signed_p10_s", "off_signed_p50_s", "off_signed_p90_s",
             "land_signed_p10_s", "land_signed_p50_s", "land_signed_p90_s")]
bp <- bp[order(bp$off_signed_p50_s), ]
kable(bp, row.names = FALSE, digits = 0,
      col.names = c("arm", "dep p10", "dep p50", "dep p90",
                    "arr p10", "arr p50", "arr p90"),
      caption = "Signed boundary error in seconds, 2025. Negative means the track began or ended before the aircraft did.")
```

The shape is what the percentiles cannot show.

```{r boundary-figure}
#| echo: false
#| fig-width: 8
#| fig-height: 3.6
#| fig-cap: "Signed boundary error on the 2025 sample, over the flights whose ground-truth times are APDF-measured. **Negative means the track began, or ended, before the aircraft did.** Departure: `legacy` and `recommended` both show a two-sided distribution — a spike just after wheels-off plus a long left tail of tracks that started on the ramp — while `ground_anchored` concentrates hard at zero, which is the arm cutting at the ground event rather than running back into the ramp. Arrival: all three are one-sided and nearly identical, because tracks essentially always run on through taxi-in. The outermost bin on each side is an overflow bin and carries everything beyond half an hour."
b <- bhist[bhist$period == "2025", ]
b$mid  <- (b$bin_lower_s + b$bin_upper_s) / 2
b$edge <- factor(b$edge, levels = c("off", "land"),
                 labels = c("departure: track start − take-off",
                            "arrival: track end − landing"))
b$arm  <- factor(b$arm, levels = c("legacy", "recommended", "ground_anchored"))
ggplot(b, aes(x = mid / 60, y = n, colour = arm)) +
  geom_vline(xintercept = 0, linewidth = 0.3, colour = "grey55") +
  geom_step(direction = "mid", linewidth = 0.5) +
  facet_wrap(~edge, ncol = 2) +
  labs(x = "signed offset (minutes)", y = "flights", colour = NULL) +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        panel.grid.minor = element_blank())
```

## The sample {#sec-sample}

*This section says what is being measured against, and over what period.*

Two three-day samples, `r num(N_GT_25)` and `r num(N_GT_24)` scored ground-truth
flights: 5–7 June 2025 and 5–7 June 2024. Two periods rather than one, because a
single sample cannot distinguish a real effect from one that sample happened to
favour — and every conclusion on this page is required to hold on both.

Ground truth is the Network Manager flight table, which covers all of ECAC,
cross-checked against APDF where APDF covers the aerodrome. Take-off is inferred
as `AOBT_3 + TAXI_TIME_3` — off-block plus taxi-out — and landing is taken from
`ARVT_3` directly, which is already a landing time rather than a gate arrival.
Both inferences were validated against real APDF movement times before being
relied on, at median error 0 s with an interquartile range of 17 s and 25 s
respectively; that is what licenses using NM-inferred intervals for matching.
Boundary error is nevertheless reported only over the APDF-measured subset,
which is conservatism rather than necessity.

## The ladder: eight candidate rules {#sec-ladder}

*This section is the study proper. It defines what a candidate rule is, then
takes each in turn: what it changes, what it is trying to fix, the rule itself,
and what it scored. By the end you will know why one of them was recommended and
why the other seven were not.*

A **ladder** here is a set of candidate segmentation rules run against the same
ground truth, on the same samples, one change at a time. Each candidate is
called an **arm**, borrowing the term from a controlled trial, and each arm is
exactly two things: a **grouping** — which samples are even eligible to share a
track — and a **break predicate** — a test evaluated at each sample in time
order, which when true starts a new track.

Every arm reads one shared cleaned track table, and only `track_id` varies
between them, so a difference between two arms cannot be an artefact of a
different track build. Arms are numbered A0 to A8; A1 is not an arm but a
parameter sweep of A0, and has [its own section](#sec-sweep).

Every domain-knowledge arm (A5, A6, A7) keeps a legacy gap rule as a **floor** —
it can add splits to legacy's, and in one deliberate exception replace one of
them, but it cannot end up splitting less than a long silence demands. There are
two reasons, and both matter for reading the numbers. The ladder measures one
change at a time, so an arm that also silently removed a legacy split would
report a delta mixing two changes; and an arm whose inputs are missing must not
degrade into assigning an airframe's whole month to a single track.

```{r ladder-table}
#| echo: false
t <- arms[arms$period == "2025", c("arm", "clean_match_pct", "fragmented_pct",
                                   "merged_pct", "n_tracks", "v_measure")]
t <- t[order(-t$clean_match_pct), ]
t24 <- arms[arms$period == "2024", c("arm", "clean_match_pct")]
names(t24)[2] <- "clean_2024"
t <- merge(t, t24, by = "arm", sort = FALSE)
t <- t[order(-t$clean_match_pct), ]
kable(t, row.names = FALSE, digits = c(0, 2, 2, 2, 0, 4, 2),
      col.names = c("arm", "clean %", "fragmented %", "merged %",
                    "tracks", "V-measure", "clean % (2024)"),
      format.args = list(big.mark = ","))
```

The ordering is identical on both periods, and no arm's clean-match rate moves
by more than `r sprintf("%.1f", ARM_SHIFT)` points between them, so what follows
is a property of the algorithms rather than of one sample.

Each arm below is given in the same four parts: what it changes relative to
`legacy`, what it is trying to fix, the rule itself, and what the measurement
said.

### A0 `legacy` — the baseline

**What it changes.** Nothing. This is the rule that built every published
`track_id` up to v0.0.2, and everything else is measured against it.

**The rule.** Group the samples by `(icao24, callsign)` — in production, by a
SHA-2 hash of the two concatenated. Order each group by time. Break when the
gap to the previous sample exceeds 30 minutes, or exceeds 15 minutes while
barometric altitude is below 5,000 ft. The identifier carries a
`_{year}_{month}` suffix.

**What the measurement said.** `r pct(CLEAN_LEGACY_25)` clean,
`r pct(FRAG_LEGACY_25)` fragmented, `r pct(MERGE_LEGACY_25)` merged on 2025;
`r pct(CLEAN_LEGACY_24)` clean on 2024. The failure is overwhelmingly
fragmentation, and almost never merging.

It is worth being precise about the cause, because the obvious reading is wrong.
It is *not* that callsigns change in mid-flight. It is that **a sample with no
callsign forms a group, and therefore a track, of its own.** Measured while the
arm was implemented, 42.4% of `legacy`'s tracks carry no callsign at all: they
are made of samples the receiver decoded without one. The group key turns a hole
in the data into a separate flight. That single mechanism is what the rest of
this ladder is chasing.

### A2 `no_month_suffix` — untestable on this sample

**What it changes.** Removes the `_{year}_{month}` suffix from the identifier.
Grouping and break predicate are `legacy`'s, untouched.

**What it is trying to fix.** The suffix splits any flight airborne at midnight
on the first of a month, by construction and regardless of the data. That is a
correctness defect rather than a threshold, which is why no amount of parameter
sweeping can reach it.

**What the measurement said.** Identical to A0 in every digit, on both periods.
That is correct behaviour and not a bug: both samples sit inside a single month,
so the suffix is constant and removing it changes nothing. The consequence is
that **this study cannot evaluate the arm at all.** It is reported as untested
rather than as a null result. Testing it needs a sample spanning a month
boundary — about two days of cluster time — and is recommended as follow-up.

### A3 `traffic_style` — the reference library's rule does not transfer

**What it changes.** Replaces legacy's two hard thresholds with the semantics of
`Flight.split` in the *traffic* library, the widely used open-source reference
for this kind of work.

**What it is trying to fix.** Nothing specific to OPDI. The question this arm
answers is whether the established reference implementation, applied to this
feed, would simply have done better.

**The rule.** Group on `(icao24, callsign)`. Break when the gap exceeds a single
threshold — *traffic*'s documented default is ten minutes — *unless* a predicate
says the candidate gap is not really a break. The predicate used here is the
aviation-sensible inverse of *traffic*'s own documented example: do not split
when the samples on both sides of the gap are above the low-altitude ceiling,
because a reception hole in the cruise is a coverage gap, not a landing.

**What the measurement said.**
`r pct(A("traffic_style", "2025", "clean_match_pct"))` clean, below `legacy`,
and the worst merging of any arm at
`r pct(A("traffic_style", "2025", "merged_pct"))` — about
`r sprintf("%.0f", A("traffic_style", "2025", "merged_pct") / MERGE_LEGACY_25)`
times legacy's. A single short gap threshold suits *traffic*'s typical input;
on a 5-second decimated continental feed it runs flights together. A genuine
negative result: the reference approach does not transfer unmodified.

### A4 `airframe_only` — delete callsign from the group key

**What it changes.** One column. Group on `icao24` alone; callsign becomes an
attribute of a sample rather than part of the track's identity. The break
predicate is `legacy`'s, untouched.

**What it is trying to fix.** A0's diagnosis directly: if a blank callsign is
what starts a spurious track, then removing callsign from the key stops it. The
underlying claim is one of principle — identity belongs to the airframe; the
callsign is something the airframe happened to be broadcasting at the time.

**What the measurement said.**
`r pct(A("airframe_only", "2025", "clean_match_pct"))` clean, with fragmentation
down from `r pct(FRAG_LEGACY_25)` to
`r pct(A("airframe_only", "2025", "fragmented_pct"))`. This is the single
largest movement anywhere in the study, and it comes from deleting one column.

**But it is not free**, and the cost is exactly the failure legacy did not have:
merging rises from `r pct(MERGE_LEGACY_25)` to `r pct(MERGE_AIR_25)`. Without
callsign in the key, two consecutive flights of the same airframe with no
sufficient gap between them become one track. Measured during implementation,
3.3% of this arm's tracks contain more than one genuine callsign.

### A5 `ground_anchored` — split at a turnaround, using the on-ground flag

**What it changes.** Groups on `icao24`, like A4, and *adds* a break condition
to legacy's gap rules rather than replacing them.

**What it is trying to fix.** Legacy's "below 5,000 ft" test is a proxy for
"the aircraft is on the ground somewhere", used because barometric altitude was
all the rule had. But the OSN schema carries an `on_ground` flag that step 02
reads nothing of. This arm uses it.

**The rule.** A break is declared at the *first airborne sample after* a run of
consecutive on-ground samples lasting at least the ground-dwell threshold. The
dwell requirement is what separates a turnaround from a touch-and-go. Placing
the break at the first airborne sample, rather than at the start of the ground
run, means the departing leg begins at the rotation instead of inheriting the
previous arrival's taxi-in. Legacy's two gap rules are kept as a floor, because
an aircraft that stops broadcasting entirely still has to be split somehow.

**What the measurement said.**
`r pct(A("ground_anchored", "2025", "clean_match_pct"))` clean — well above
`legacy`, well below A4. Its distinguishing result is not the headline but the
boundary error: its departure tenth percentile is
`r sec(A("ground_anchored", "2025", "off_signed_p10_s"))` s against legacy's
`r sec(A("legacy", "2025", "off_signed_p10_s"))` s, and the figure in
[the metrics section](#sec-metrics) shows the whole left tail collapsing. That
is direct evidence the arm does what it claims: it cuts at the ground event
rather than running back into the ramp. It is the arm to reach for if track
*edges* matter more than track *counts*.

### A6 `airport_anchored` — split only where an aerodrome actually is

**What it changes.** Groups on `icao24`, and replaces legacy's low-altitude gap
rule with real airport geometry. It is the one arm that takes only *half* of
legacy as its floor.

**What it is trying to fix.** Two errors that no altitude threshold can reach.
A split legacy *misses*: an aircraft parked at a 6,000 ft aerodrome never
satisfies "below 5,000 ft", so the turnaround is invisible and two flights merge.
And a split legacy *invents*: a long slow low transit in the middle of nowhere
satisfies the low-altitude rule and is split, when nothing happened at all.

**The rule.** A sample is *stationary at an aerodrome* when it is inside a known
airport detection zone, below a turnaround height measured **above field
elevation** rather than above sea level, and slower than a turnaround speed. A
break is then declared two ways, and the arm fires on either: at the sample that
ends a reception gap, when the previous sample was stationary at an aerodrome
and at least the dwell threshold has passed; or at the first sample that moves
again after a continuous stationary run lasting at least the dwell threshold.
Where the airport join finds nothing, "near an aerodrome" is false and the arm
cannot fire, falling back to its floor. Because the arm's entire thesis is that
legacy's low-altitude rule is the wrong test, that rule is deliberately *not*
inherited — only the general "a long enough silence is a new track" rule is,
which no arm disputes and which guarantees no airframe can end up with a
month-long track.

**What the measurement said.**
`r pct(A("airport_anchored", "2025", "clean_match_pct"))` clean on 2025 and
`r pct(A("airport_anchored", "2024", "clean_match_pct"))` on 2024 — the closest
any domain-knowledge arm gets to A4, and still short of it. Real geometry is
worth a lot; it is worth less than not fragmenting on blanks.

### A7 `vertical_profile` — a negative result, correctly attributed

**What it changes.** Groups on `icao24`, and adds a break based purely on the
altitude trace, with legacy's gap rules as a floor.

**What it is trying to fix.** The case where the transponder never reports
`on_ground` at all. If a flight is one climb–cruise–descent cycle, then a
descent to circuit height followed by a climb back through it is a new sortie,
whether or not there was a reception gap and whether or not ground contact was
ever broadcast.

**The rule.** Both halves are required. First, *the descent*: some earlier
sample of the group crossed the floor downwards — the previous sample above it,
this one at or below. Second, *the climb back*: this sample is above the floor
and the previous one is not. The break is placed at the upward crossing, so the
departing leg begins there. The floor is measured **above field elevation**;
comparing against sea level made the arm degenerate to bare legacy at any
aerodrome above the floor's own height, since a parked aircraft there already
reads above it and no crossing is ever seen.

**What the measurement said.**
`r pct(A("vertical_profile", "2025", "clean_match_pct"))` clean, last of the
eight on both periods. An earlier implementation split on *any* upward crossing,
which is not the idea, and its poor score was therefore unattributable — it
could have been the idea failing or the code not being the idea. Implemented
properly, and on height above field, it scores *worse* still. That is what makes
this a real negative result rather than an unexplained number. The mechanism is
visible in the boundary error: every percentile of its signed departure offset
is positive
(tenth percentile `r sec(A("vertical_profile", "2025", "off_signed_p10_s"))` s),
meaning it systematically starts tracks late, having sliced off each departure's
initial climb.

### A8 `recommended` — the two fixes, kept separate {#sec-a8}

This is the arm the study recommends and the rule that now ships as
`segmentation.method = "standard"`, so it gets the most space. **Its two changes
are independent, they fix different failures, and the arm only works because
they are kept separate — so they are set out one at a time, before any number.**

**What it changes.** Two things, both relative to `legacy`: the group key loses
its callsign column, and the break predicate gains a callsign test. Legacy's two
gap rules are kept underneath both, unchanged — a long enough silence still
starts a new track, at any altitude or a shorter one when low. Nothing about the
gap behaviour of published tracks changes; what changes is which samples are
allowed to share one, and where a shared group is cut.

#### Fix one: take callsign out of the group key

*The failure it addresses* is A0's: a sample with no callsign forms its own
group, so a hole in the data becomes a separate flight. That is where legacy's
fragmentation comes from, and A4 showed that deleting the column removes it.

*The change* is exactly A4's: group on `icao24` alone. A blank or flickering
callsign can no longer start a track, because callsign is no longer part of what
makes a track a track.

*What it costs on its own* is A4's cost: two consecutive flights of one airframe
now have nothing to separate them but a gap, and where the aircraft kept
broadcasting through the turnaround there is no sufficient gap. They merge.

#### Fix two: break when the callsign genuinely changes

*The failure it addresses* is the one fix one creates. Legacy separated those
two flights for the wrong reason — callsign was in the key, so any change of it
started a new group — but it did separate them.

*The change* restores that separation as an explicit break rather than as a side
effect of the key. Walking an airframe's samples in time order, carry forward
the last **non-blank** callsign seen. Break at a sample when that sample's
callsign is itself non-blank, the carried-forward one is non-blank, and the two
differ. Comparing the raw callsign column instead would break at every blank
sample and reproduce legacy's fragmentation exactly; it is carrying the last
*real* value across the blanks that makes the test mean "the callsign changed"
rather than "the callsign was missing". Callsigns are compared trimmed, because
ADS-B pads them to eight characters and `ABC123  ` is not a different flight
from `ABC123`.

*Why the two are genuinely independent:* one is about which samples may share a
track, the other about where a shared group is cut. The evidence that they are
is that the combination beats **both** of the arms it draws on, rather than
landing between them — which is what would happen if they were two attempts at
the same thing.

#### The subtlety: the lookback must be bounded

The break predicate is evaluated over the whole airframe's samples, and it has
no notion of the track boundaries it is in the middle of deciding. So "the last
non-blank callsign" is, by default, the last one *anywhere earlier in the
month*. That reaches back across a gap break into the previous flight — and the
next flight, whose first samples happen to be blank, is then compared against a
callsign belonging to a flight that already ended, and is split in its own
middle.

The fix is to bound the lookback: the carried-forward callsign only counts if it
is no older than the general gap threshold. That is exactly the right bound
rather than an arbitrary one, because a silence longer than the gap threshold
*is already a break* — so there is nothing legitimate to carry across it.

This is not a detail. Measured during implementation, the unbounded version
fragmented 37.75% of flights, where `airframe_only` — which does no callsign
splitting at all — fragments
`r pct(A("airframe_only", "2025", "fragmented_pct"))`, and where only 3.2% of
tracks contain a genuine callsign transition to fire on. The mismatch between
those two is what exposed the bug: a rule reaching one track in thirty cannot
legitimately cost thirty points of fragmentation.

#### What the measurement said

`r pct(CLEAN_REC_25)` clean on 2025 and `r pct(CLEAN_REC_24)` on 2024 — the best
of the eight on both periods, and `r pp(GAIN_25)` and `r pp(GAIN_24)` points
above `legacy`. Fragmentation falls from `r pct(FRAG_LEGACY_25)` to
`r pct(FRAG_REC_25)`.

The result that settles it is merging: `r pct(MERGE_REC_25)`, against
`airframe_only`'s `r pct(MERGE_AIR_25)`. Fix two recovers almost all of the
merging fix one introduces — though not quite all of it, and the remainder is
still above `legacy`'s `r pct(MERGE_LEGACY_25)`. That residual is small, it is
real, and it is where the downstream cost measured in [the payoff](#sec-payoff)
comes from.

## Was `legacy` simply mistuned? {#sec-sweep}

*This section exists to close off one explanation of A0's result, and it is
worth reading even though it tunes a rule this page does not recommend. By the
end you will know that legacy's shortfall is structural rather than a matter of
badly chosen thresholds.*

The ladder says legacy recovers about half the flights it should. There is an
obvious and much less interesting explanation for that: its three thresholds
were never measured, so perhaps they are simply set wrong, and legacy at better
settings would close most of the gap. If that were true, everything the ladder
concludes would be an artefact of comparing a tuned rule against an untuned one.

So it was tested. A **sweep** is a re-run of the same rule at many different
parameter settings; a **cell** is one such setting. This one is labelled A1, and
it is not new code — it is A0 evaluated at settings other than production's, so
it measures legacy's own tuning and nothing else.

### The three axes

`legacy` has exactly three parameters, and all three were swept:

- **`gap_minutes`** — a gap longer than this starts a new track, at any
  altitude. Production: 30.
- **`low_alt_gap_minutes`** — a shorter gap that also starts a new track, but
  only when the aircraft is low. Production: 15.
- **`low_alt_ft`** — the altitude below which the shorter gap applies.
  Production: 5,000.

Stage 1 evaluated `r num(SWEEP_CELLS)` cells across the three axes on one day.
End to end, the entire grid spans `r sprintf("%.2f", SWEEP_SPREAD1)` points of
clean matching.

```{r sweep-axes}
#| echo: false
by_axis <- function(col) {
  agg <- aggregate(clean_match_pct ~ get(col), data = sweep1, FUN = max)
  names(agg) <- c("value", "best")
  agg[order(agg$value), ]
}
axes <- list(gap_minutes = "gap_minutes (min)",
             low_alt_gap_minutes = "low_alt_gap_minutes (min)",
             low_alt_ft = "low_alt_ft (ft)")
prof <- do.call(rbind, lapply(names(axes), function(a) {
  d <- by_axis(a)
  data.frame(Axis = axes[[a]], Value = d$value, Best = d$best,
             Peak = ifelse(d$best == max(d$best), "*", ""),
             check.names = FALSE)
}))
kable(prof, row.names = FALSE, digits = c(0, 0, 3, 0),
      col.names = c("axis", "value", "best clean % at this value", ""),
      format.args = list(big.mark = ","),
      caption = "Stage 1: for each value of each parameter, the best clean-match rate achieved anywhere in the grid at that value. A star marks each axis's peak. Every peak has lower values on both sides of it.")
```

Read down each block: every axis rises to a peak and falls again on the far
side. That is what an **interior optimum** means, and it is the condition for
calling a result an optimum at all. An optimum sitting on the edge of a grid is
not an optimum — it is where you stopped looking.

### The extension, and why it was necessary

`low_alt_ft` did not start out interior. The first grid ran it to
`r num(LOALT_BASE_MAX)` ft, and its best value was the largest tried, with the
curve still rising — precisely the edge case above. This pipeline's own history
contains a sweep that stopped early and was read as "higher is always better",
so the grid was extended rather than the result reported.

```{r sweep-extension}
#| echo: false
base <- D("sweep_2025_stage1.csv")
ext  <- D("sweep_2025_stage1_ext.csv")
lo <- function(d) {
  a <- aggregate(clean_match_pct ~ low_alt_ft, data = d, FUN = max)
  a[order(a$low_alt_ft), ]
}
b <- lo(base); e <- lo(sweep1)
tab <- data.frame(low_alt_ft = e$low_alt_ft,
                  grid = ifelse(e$low_alt_ft %in% b$low_alt_ft,
                                "original", "extension"),
                  best = e$clean_match_pct, check.names = FALSE)
kable(tab, row.names = FALSE, digits = c(0, 0, 3),
      col.names = c("low_alt_ft", "grid", "best clean %"),
      format.args = list(big.mark = ","),
      caption = "The low-altitude ceiling before and after the grid was extended. On the original grid the best value was the largest tried; extended, the curve turns over.")
```

Extended, the ceiling peaks at `r num(LOALT_PEAK)` ft and falls again by
`r num(LOALT_MAX)` — as it must, since past cruise altitude every sample counts
as "low", the two-threshold rule collapses into a single one, and
`low_alt_gap_minutes` simply replaces `gap_minutes` everywhere.

### What it is worth

Stage 2 re-measured a shortlist of cells on the full three-day samples, so that
the winner is chosen out of sample rather than on the day it was tuned.

```{r sweep-stage2}
#| echo: false
s2 <- sweep2[, c("period", "gap_minutes", "low_alt_gap_minutes", "low_alt_ft",
                 "clean_match_pct")]
s2$production <- ifelse(s2$gap_minutes == 30 & s2$low_alt_gap_minutes == 15 &
                        s2$low_alt_ft == 5000, "production", "")
s2 <- s2[order(s2$period, -s2$clean_match_pct), ]
kable(s2, row.names = FALSE, digits = c(0, 0, 0, 0, 3, 0),
      col.names = c("period", "gap_minutes", "low_alt_gap_minutes",
                    "low_alt_ft", "clean %", ""),
      format.args = list(big.mark = ","))
```

The best cell beats production by **`r pp(SWEEP_GAIN_25)`** points on 2025 and
**`r pp(SWEEP_GAIN_24)`** on 2024, and both periods choose the same cell.

That is the answer, and it is a clean negative: **`legacy` is not mistuned.** It
sits a fraction of a point from the best setting available to it, in a grid
drawn wider than plausible and extended until every axis turned over. Set
against the `r pp(GAIN_25)` points the identity key is worth, the entire tuning
question is roughly `r sprintf("%.0f", SWEEP_RATIO)` times too small to matter —
which is why the rest of this page is about the group key and not about
thresholds.

## What a better segmentation buys ADEP/ADES {#sec-payoff}

*This section measures the thing that actually matters: what a segmentation does
to the product OPDI publishes. It also records a defect the study found in the
flight list, which has since been fixed, and why the ADEP/ADES ranking does not
follow the segmentation ranking even now that it is.*

The point of segmenting well is the product built on top of it. Each arm's
assignment is fed to the V7 flight list with **only `track_id` varying** — same
input table, same detection configuration, same reference data — so any
difference in the result is attributable to the segmentation and to nothing
else.

Three terms, because the flight list scores differently from the ladder:

- **coverage** — the share of ground-truth flights the flight list gives *any*
  aerodrome for, in that role;
- **accuracy** — of those it does answer, the share it gets right;
- **score** — `correct − k × wrong` at `k = 2`, so a wrong answer costs twice
  what a right one earns and a silence costs nothing. `ADEP + ADES` is the sum
  over both roles, and it is the single number the tables below rank on.

The ground truth here is the flight-list studies' own, keyed on callsign and
day, which is why its denominator (`r num(payoff$n_ground_truth[1])` flights on
2025) differs from the ladder's. The two are measuring different things and are
not interchangeable.

As a control, `legacy` run through this harness scores `r num(PAY_LEGACY_25)`
against the V7 flight-list study's own published `r num(V7_PUBLISHED)` for the
same sample — a difference of `r num(abs(PAY_LEGACY_25 - V7_PUBLISHED))`, or
`r sprintf("%.3f%%", 100 * abs(PAY_LEGACY_25 - V7_PUBLISHED) / V7_PUBLISHED)`.
That residual is itself informative and is explained in
[Limitations](#sec-limitations).

### The flight-list configuration these numbers were produced at {#sec-detection}

Every arm is scored at one fixed ADEP/ADES configuration. Naming it matters,
because "the V7 recommended configuration" is a moving target and these numbers
belong to one point on it. The table is read from the run's own recorded
parameters rather than typed, so it cannot drift from what was executed.

```{r detection-params}
#| echo: false
cfg <- payoff[payoff$period == "2025", ]
fields <- c("adep_mode", "ades_mode",
            "endpoint_radius_nm", "endpoint_height_ft",
            "endpoint_sched_penalty_nm", "endpoint_candidate_radius_nm",
            "trend_max_fl", "trend_radius_nm", "trend_radius_exact",
            "trend_rank_by", "trend_vote_margin", "trend_sched_penalty_nm",
            "trend_smooth_half_window", "trend_smooth_before_cut",
            "trend_bearing_tiebreak_nm", "trend_ooa", "k")
# Every arm must have been run at the same configuration, or the comparison is
# not a comparison. Asserted rather than assumed: a silent disagreement here
# would attribute a detection change to a segmentation.
stopifnot(all(sapply(fields, function(f) length(unique(cfg[[f]])) == 1)))
meaning <- c(
  adep_mode = "Which algorithm names the departure aerodrome.",
  ades_mode = "Which algorithm names the destination aerodrome.",
  endpoint_radius_nm = "`endpoint`: how close the track's first or last fix must be to an aerodrome.",
  endpoint_height_ft = "`endpoint`: how high that fix may be, above field elevation.",
  endpoint_sched_penalty_nm = "Distance added to an aerodrome without scheduled service, so a military field must be clearly nearest.",
  endpoint_candidate_radius_nm = "How far the cached candidate table reaches. Wider than any detection radius, so the radius stays sweepable.",
  trend_max_fl = "`trend`: only samples below this flight level vote.",
  trend_radius_nm = "`trend`: how close a sample must be to an aerodrome to vote for it.",
  trend_radius_exact = "Cut that radius on exact great-circle distance rather than on H3 band.",
  trend_rank_by = "How `trend` chooses among candidates: exact distance, not H3 ring count.",
  trend_vote_margin = "How far one direction must beat the other before the vote counts. Zero is a simple majority.",
  trend_sched_penalty_nm = "The scheduled-service penalty, on `trend`'s side.",
  trend_smooth_half_window = "Half-width in samples of the rolling mean over barometric altitude.",
  trend_smooth_before_cut = "Smooth altitude before applying the flight-level cut, not after.",
  trend_bearing_tiebreak_nm = "Within this distance band, candidates are separated by course alignment rather than distance.",
  trend_ooa = "Whether `trend` may emit the out-of-area marker. Off: on arrivals it is right barely half the time.",
  k = "The scoring exchange rate: one wrong answer costs `k` right ones.")
dp <- data.frame(
  Parameter = paste0("`", fields, "`"),
  Value = sapply(fields, function(f) as.character(cfg[[f]][1])),
  Meaning = meaning[fields], check.names = FALSE)
kable(dp, row.names = FALSE, align = "lll",
      col.names = c("parameter", "value", "what it does"),
      caption = "The ADEP/ADES configuration every arm in this section was scored at, identical across all eight arms and both periods.")
```

**Cross-checked, not copied.** Every value above was compared field by field
against `DetectionConfig()` in the checkout these runs were made from, and they
agree. Compared against the *v6.2* study's shipped configuration, one difference
is real and is reported rather than smoothed over:

> v6.2 gives `trend`'s altitude cut a **datum**, defaulting to field elevation:
> the cut becomes `trend_max_height_ft` above the field rather than
> `trend_max_fl` above mean sea level, and `trend_max_fl` is retained but inert.
> That change landed on `main` **after** this study's base commit and is not
> present in the checkout these payoff numbers were produced from. The runs
> above therefore cut on FL60 above sea level.

This does not affect any comparison on this page — every arm is scored at the
same configuration, so the datum is held constant across all of them — but it
does mean these absolute ADEP/ADES totals are not directly comparable with
figures published by v6.2 or later.

### Segmentation quality and downstream score do not line up

```{r payoff-table}
#| echo: false
p <- payoff[payoff$period == "2025", ]
p$total <- p$adep_score + p$ades_score
p$vs_legacy <- p$total - PAY_LEGACY_25
p$clean <- sapply(p$arm, function(a) A(a, "2025", "clean_match_pct"))
p <- p[order(-p$total), ]
out <- data.frame(
  arm = p$arm,
  clean = p$clean,
  total = p$total,
  vs = p$vs_legacy,
  adep_cov = 100 * p$adep_coverage,
  adep_acc = 100 * p$adep_accuracy,
  ades_cov = 100 * p$ades_coverage,
  ades_acc = 100 * p$ades_accuracy,
  check.names = FALSE)
kable(out, row.names = FALSE, digits = c(0, 2, 0, 0, 2, 2, 2, 2),
      col.names = c("arm", "clean % (ladder)", "ADEP+ADES", "vs legacy",
                    "ADEP cov %", "ADEP acc %", "ADES cov %", "ADES acc %"),
      format.args = list(big.mark = ","),
      caption = "2025. Ranked on ADEP+ADES score. The second column is the same arm's clean-match rate from the ladder, so the two rankings can be read against each other in one place.")
```

The two rankings disagree. `recommended` is first on the ladder and third here;
`vertical_profile` is last on the ladder and has the **highest ADEP score of any
arm**. A reader could take that to mean better segmentation makes the product
worse. It does not, and the table says why in its last four columns.

**Accuracy barely moves.** Every arm lands within a couple of points on both
roles. No arm is naming *wrong* aerodromes at a materially different rate.

**Coverage moves, and almost all of the movement is on the arrival side.** ADEP
coverage spans a narrow band across all eight arms; ADES coverage runs from
`r sprintf("%.1f", min(100 * p$ades_coverage))`% to
`r sprintf("%.1f", max(100 * p$ades_coverage))`%. So the whole ranking is
decided by how often the arrival algorithm can answer at all.

Why the arrival side is so much more sensitive than the departure side is not
settled by this study, and the page does not pretend otherwise. What the data
does say is suggestive and consistent across both periods: the two arms with the
worst arrival coverage by a wide margin — `ground_anchored` and
`vertical_profile` — are exactly the two whose breaks are placed at
*low-altitude events*, while `legacy` and `recommended` place theirs at
reception gaps and land close together. The destination algorithm reads its
evidence from the low-altitude part of a track, so an arm that cuts there is
cutting into the evidence. The correlation is strong and the direction is
plausible; isolating the mechanism would need a per-flight diagnosis this study
did not run, and it is recorded as a hypothesis rather than a finding.

The general lesson does not depend on that hypothesis being right: **an arm that
is good at recovering flights can still be bad at preserving the evidence a
downstream detector needs**, and the two have to be measured separately.

### The labelling bug this study found, and why it is now history {#sec-labelling}

The measurement above could not be made until a defect in the flight list was
fixed, and the defect is worth recording because it explains why an earlier
version of this page reported the ranking as inverted, and because anyone
meeting an OPDI flight list built before the fix needs to recognise it.

**What it was.** The flight list built one flight per `track_id` and labelled it
with `F.min("flight_id")` — the lexicographically smallest callsign in the
track. A blank sorts first.

**Why it was harmless, until it was not.** While callsign is part of the track's
group key, every track carries exactly one callsign by construction, so that
aggregate can never lose anything. It is correct only because of a property
`legacy` supplies *by accident*: blank-callsign samples sit in tracks of their
own. Any fix to legacy's fragmentation necessarily destroys that property —
which is the whole point of the fix. Drop callsign from the key and a track that
carries a real callsign alongside blank ones is labelled blank; the flight
becomes unidentifiable and matches no ground-truth counterpart.

**What it cost.** Measured against the pipeline as it then stood, adopting the
better segmentation *lost* ADEP/ADES points, and the losing arms lost them by
coverage collapse with accuracy essentially unmoved — the signature of a
labelling failure rather than a geometric one. So legacy's fragmentation and the
downstream coverage collapse were one phenomenon seen from two ends, and the
recommendation was, at that point, conditional: fix the label first, then ship
the segmentation. Reversed, the pipeline would have spent the interval between
the two changes at substantially reduced ADEP coverage.

**What was actually wrong was larger than the label.** Tracing it found that
`flight_id` is used as a grouping or join key in a dozen places downstream and
is never aggregated to one value per track — so the flight list is written on
the invariant *one `flight_id` per `track_id`*, which `legacy` guarantees and
the new segmentation breaks. A track carrying several callsigns does not merely
get labelled wrongly; it *fans out into several rows* at each of those keys. The
same latent fan-out was present in the event extraction of step 04.

**It is fixed.** Every point where the flight list and the event extraction read
the track table now resolves `flight_id` to the track's dominant non-blank
callsign, over the unfiltered frame, before anything groups or joins on it. Runs
that stamp a legacy version string skip the resolution, so re-running a released
month still reproduces what was released, column for column. The numbers
throughout this page are from the fixed pipeline; there is no repaired-versus-
unrepaired pairing to interpret any more.

### The convergence check

The benchmark kept its own independent repair — relabelling each track with its
most frequent non-blank callsign, applied inside the harness — and both jobs are
still run. Now that production does the same thing, their **agreement** is the
verification that the production fix and the benchmark repair are the same
operation. A disagreement would be a finding rather than a measurement.

```{r fix-table}
#| echo: false
rows <- do.call(rbind, lapply(PERIODS, function(p) {
  data.frame(
    period = p,
    arm = c("legacy (control)", "recommended"),
    production = c(P(payoff, "legacy", p), P(payoff, "recommended", p)),
    harness    = c(P(fixcs, "legacy", p), P(fixcs, "recommended", p)),
    check.names = FALSE)
}))
rows$delta <- rows$harness - rows$production
kable(rows, row.names = FALSE, digits = 0,
      col.names = c("period", "arm", "production fix only",
                    "production fix + harness repair", "difference"),
      format.args = list(big.mark = ","),
      caption = "The two jobs agree where it matters. `recommended` comes back identical on every recorded field in both periods; before the production fix the two differed by tens of thousands of flights.")
```

`recommended` agrees exactly, on both periods, which is the check passing.

`legacy` does not, on 2025 only, and by
`r num(abs(FIX_LEGACY_25 - PAY_LEGACY_25))` points confined entirely to the
arrival side — the same flights answered, a handful labelled with a different
aerodrome. That was traced, and it is not the callsign repair: it is a
non-deterministic aggregate in shipped ADES detection, which is reported in
[Limitations](#sec-limitations) because it is a real defect and it is not this
study's to fix.

### Where the residual comes from

With the labelling fixed in production, the recommended segmentation scores
**`r num(PAY_REC_25)`** against legacy's **`r num(PAY_LEGACY_25)`** on 2025 and
**`r num(PAY_REC_24)`** against **`r num(PAY_LEGACY_24)`** on 2024 — short by
`r num(PAY_SHORT_25)` and `r num(PAY_SHORT_24)` points,
`r pct(100 * PAY_SHORT_25 / PAY_LEGACY_25)` and
`r pct(100 * PAY_SHORT_24 / PAY_LEGACY_24)` — while recovering
`r sprintf("%.0f", GAIN_25)` points more flights on the ladder. Its **ADEP**
coverage slightly *exceeds* legacy's on both periods:
`r sprintf("%.2f", 100 * PR(payoff, "recommended", "2025", "adep_coverage"))`%
against `r sprintf("%.2f", 100 * PR(payoff, "legacy", "2025", "adep_coverage"))`%.

The residual is not accuracy and it is not departures. It is arrival coverage,
and the only failure mode that grew is the one [A8](#sec-a8) said would grow:
`recommended` merges `r pct(MERGE_REC_25)` of flights where `legacy` merges
`r pct(MERGE_LEGACY_25)`, and a merged track has one arrival where there should
be two, so one of the two flights loses its destination outright. The two
statistics are measured against different ground truths — the ladder's and the
flight list's — so they cannot be reconciled flight for flight, and merging on
its own is smaller than the coverage difference it is offered to explain. What
can be said without qualification is that merging is the only thing that moved
in the wrong direction, and that it moved.

That is the price of grouping on the airframe. It is stated rather than rounded
away, and it is small enough that the `r sprintf("%.0f", GAIN_25)`-point gain in
recovered flights buys it many times over.

## What shipped {#sec-defaults}

*This section says what was adopted, in what order, and what a consumer of OPDI
data has to know about it.*

**The A8 rule ships, as `segmentation.method = "standard"`:** group on `icao24`,
break when the last non-blank callsign genuinely changes with the lookback
bounded to the gap threshold, and keep legacy's gap rules as a floor.

The recommendation was conditional when this study first reported, and the
condition was an ordering rather than a hedge. Both steps have now been taken,
in this order:

1. **First, the flight-list labelling was fixed.** `flights.py` no longer picks
   a track's callsign with `F.min`; every read of the track table resolves
   `flight_id` to the track's dominant non-blank callsign before anything groups
   or joins on it, and the same fix was applied to step 04's event extraction.
   This was worth doing on its own merits even had the segmentation never
   changed, because it is a silent trap for any future change to track identity
   and it fails as a *coverage* statistic rather than as an error — nothing
   looks broken, there is simply less data.
2. **Then the segmentation was switched.** Reversed, the pipeline would have
   spent the interval between the two changes at substantially reduced ADEP
   coverage.

Three things a consumer needs to know, all of them consequences of the release
rather than open questions:

- **`track_id` changes for all data produced from this release forward**, in
  shape as well as value: `standard` carries no `_{year}_{month}` suffix, so
  identifiers become `{hash}_{offset}`. A consumer joining on `track_id` across
  the boundary gets an empty join rather than an error, which reads as missing
  data; a consumer parsing the suffix breaks outright.
- **Past months do not reproduce under the new rule, and are not meant to.**
  Setting `segmentation.method` back to `"legacy"` reproduces the published
  identifiers, and the version-guarded paths in the flight list and the event
  extraction reproduce the published labels alongside them.
- **Nothing in the published data records which rule produced a row.**
  `osn_tracks` carries no version column, so the only way to tell is to know
  when the row was published. This is a deliberate decision and its cost is
  written down in [Limitations](#sec-limitations).

## Provenance {#sec-provenance}

*This section says where every number on this page comes from.*

Each row is one file this page reads: the job that produced it, the commit it
was produced at, and when. Every figure on this page is regenerated by
`benchmarks/regenerate_track_v1.py`, which the page runs before drawing
anything, so a row marked **unverified** would mean a file staged outside that
chain and traceable to no command.

```{r provenance}
#| echo: false
#| results: asis
mf <- file.path("data", "_manifest.json")
if (!requireNamespace("jsonlite", quietly = TRUE)) {
  cat("*jsonlite not available; provenance table skipped.*\n")
} else if (!file.exists(mf)) {
  cat("**No manifest.** Every CSV here was staged by hand and none can be traced to the command that produced it.\n")
} else {
  m <- jsonlite::fromJSON(mf, simplifyDataFrame = FALSE)
  files <- sort(list.files("data", pattern = "[.]csv$"))
  rows <- do.call(rbind, lapply(files, function(f) {
    e <- m[[f]]
    if (is.null(e)) {
      data.frame(File = f, Job = "**unverified**", Commit = "—",
                 Produced = "—", check.names = FALSE)
    } else {
      data.frame(File = f, Job = basename(e$script),
                 Commit = paste0(e$git_sha, if (isTRUE(e$git_dirty)) " (dirty)" else ""),
                 Produced = substr(e$produced_utc, 1, 16), check.names = FALSE)
    }
  }))
  print(kable(rows, row.names = FALSE, align = "l"))
  n_un <- sum(rows$Job == "**unverified**")
  cat(sprintf("\n%d files, %d traced to a command, %d unverified.\n",
              nrow(rows), nrow(rows) - n_un, n_un))
}
```

Two classes of figure on this page are **not** read from those files, and both
are marked as such where they appear.

- A handful of diagnostic measurements made while the arms were implemented —
  the share of `legacy` tracks carrying no callsign, the share of
  `airframe_only` tracks carrying more than one, and the cost of an unbounded
  callsign lookback. They are quoted with their source named in the text
  ("measured while the arm was implemented", "measured during implementation")
  and they are not regenerated by this page.
- The V7 flight-list study's own published total, used once as an external
  control, and the ground-truth validation figures quoted in
  [The sample](#sec-sample). Both belong to other papers and are cited, not
  recomputed here.

Everything else — every rate, count, percentile, table cell and plotted bar — is
an expression evaluated over the files above at render time. None of it is typed
into the prose, which is what stops the text and the tables drifting apart.

The re-run behind the current numbers is itself worth a note on method. All
fourteen files were regenerated after two defects were fixed in the ground-truth
loader, including six jobs that were confidently predicted not to move: they had
been reasoned about through two of their three declared dependencies, and the
third had changed. Every cell of all four sweep files moved, by up to 0.023 pp,
relocating no optimum and changing no conclusion — but the numbers on the page
are different from the ones that were there. Narrowing a declared dependency to
avoid a re-run is exactly how a page ends up publishing figures that nothing can
account for.

## Limitations {#sec-limitations}

*This section is where the claims this page does **not** make are written down,
together with the defects it found and did not fix.*

### The baseline is a re-implementation, not the published `track_id`

Two distinct things are meant by "the baseline", and neither is what production
now runs.

First, and as [§ what a track is](#sec-problem) states plainly, this study
measures algorithms in a harness rather than the pipeline itself. A0 is a
re-implementation of the legacy *rule* inside the generic segmentation engine.
The engine is asserted by test to reproduce the frozen algorithm's partition
under production parameters, which is the contract that makes A0 a fair
baseline — but it is a contract on the partition, not a claim to have run the
pipeline.

Second, applied to *cleaned* tracks, the legacy rule does not reproduce the
published identifiers exactly: it differs on 909 of 236,899 tracks, 0.38%, and
on `r num(abs(PAY_LEGACY_25 - V7_PUBLISHED))` of this page's
`r num(PAY_LEGACY_25)` ADEP/ADES points — which is the control residual noted in
[the payoff](#sec-payoff). The cause is exact and worth reporting on its own
account. Production's identifiers were computed upstream on raw altitudes, while
`osn_tracks_clean` masks bad altitudes to NULL — and `NULL < 5000` is NULL, not
true, so the low-altitude split silently does not fire where it once did.
Measured, **950 of 950 disagreements have a NULL barometric altitude**, against
a 21.4% base rate. The published `osn_tracks_clean` therefore carries a
`track_id` that its own columns can no longer reproduce: nothing is wrong with
the published data, but anyone recomputing identifiers from the clean table —
the obvious thing to do — gets different tracks and no warning.

Neither point affects the comparisons on this page, because every arm reads the
same cleaned table through the same engine and is subject to both behaviours
equally. Version 2 of this study is the one that runs the pipeline end to end.

### Published data carries no marker of which segmentation produced it

`osn_tracks` has no version column, so **no row of published OPDI data says
which segmentation rule built its `track_id`**. Distinguishing the two regimes
depends entirely on knowing the release date.

This is a consequence of a deliberate decision not to add such a column, and its
cost is concrete rather than abstract. The guards that let a legacy re-run
reproduce released labels — in `flights.py` and in `events.py` — key on the
*run's configuration*, not on the data: they ask "is this run stamping a legacy
version string?" when the question they need answered is "were the tracks I am
reading built with the legacy rule?". Nothing in the data can tell them. So a
legacy-configured run over tracks that were rebuilt with `standard` will take
the legacy branch and publish under a frozen version string, and no check
anywhere can catch it. The same blind spot reaches the flight-events ladder,
whose lower rungs carry a legacy events version and would silently measure the
new segmentation if re-run after the switch.

### ADES values are not reproducible run to run

**This is a defect in shipped ADEP/ADES detection, it predates this study, and
it is not fixed.** It is recorded here because both this page and the
flight-list studies quote ADES percentages to two decimals, and a figure that is
not reproducible must say so.

`flights.py:_rank_airports` ranks candidate aerodromes correctly with
`row_number()`, and then names the winner with `first(apt_ident)` — an
aggregate with no defined order. Adding a join upstream can change the answer.
The defect is self-detecting, because the *runner-up* list is built from the
rank rather than from `first()`, so a wrongly named aerodrome shows up in its
own "not most likely" list.

Measured across repeated runs on identical code and data: roughly 28 of 88,737
named ADES per run, about 0.03%, come from the undefined-order aggregate, and
about 20 can flip between two runs. The affected flights are short hops between
neighbouring aerodromes. **ADEP is immune** — the departure path filters on the
rank instead of aggregating, which is the correct pattern and already exists in
the same file. `ades_mode` defaults to `trend`, so this is the shipped arrival
path.

It is the explanation for the one disagreement in
[the convergence check](#sec-payoff): the control differs by a handful of ADES
flights on one period and by none on the other, with coverage identical to six
decimals.

### Two ground-truth intervals over twenty hours survive in the 2024 sample

Ground truth is built from Network Manager's own `AOBT_3`/`ARVT_3` pairing, and
two intervals in the 2024 sample exceed twenty hours — the longest being 27.9 h,
THY801 LTFM→SKBO. Because samples are attributed to a flight by containment, an
over-long interval can absorb samples belonging to a neighbouring leg of the
same airframe.

These are **not** the defect that was fixed. A separate loader defect, since
corrected, windowed ground truth on the off-block day rather than on the flight
interval and keyed the APDF departure join on the off-block day rather than the
take-off day; the second let a nightly service take its take-off time from
yesterday's rotation, producing intervals of up to 26.9 hours. Fixing it removed
every such interval from 2025 and all but these two from 2024, and moved the
scored flight counts by about a tenth of a percent with every rate moving by no
more than 0.038 pp.

The two survivors are NM's own pairing and are undiagnosed. They were
deliberately not filtered out: a length filter would change the denominator of
every metric on this page, which is a larger and less examined intervention than
the contamination it removes. The effect is carried, stated, and small.

### The ADEP/ADES configuration predates the v6.2 datum change

As [§ the flight-list configuration](#sec-detection) records, the payoff runs
were made from a checkout that does not contain v6.2's altitude-cut datum, so
`trend` cuts at FL60 above mean sea level rather than at a height above field
elevation. The configuration is identical across all eight arms, so every
comparison on this page is unaffected; the absolute ADEP/ADES totals are not
comparable with figures published by v6.2 or later.

### Ground truth covers Network Manager flights only

Merges with traffic absent from the NM flight table — military, general
aviation, non-ECAC operators — are invisible to the metric, because the merged
partner has no ground-truth flight to be merged with. **The merge rates on this
page are therefore floors, not estimates.** Fragmentation is unaffected, since
it is measured within flights that ground truth does contain.

### One arm is untested, not null

`no_month_suffix` is identical to legacy on both samples because both sit inside
a single month. It is reported as untested. Evaluating it needs a sample
spanning a month boundary.

### Two three-day samples in June

Both samples are in early June. Nothing here licenses a seasonal claim, and the
callsign statistics in particular could plausibly differ in a period with more
charter or seasonal traffic.

### The A1 sweep varied three parameters

The sweep covers the gap threshold, the low-altitude gap threshold and the
low-altitude ceiling. It does not vary the ground-dwell or turnaround
parameters that the domain arms introduce, so those arms are compared at one
setting each and could be better at another. Given the structural gap between
the arms is two orders of magnitude larger than anything the sweep found, this
is unlikely to change the ordering, but it is not measured.

### The payoff holds the flight list fixed

Each arm is scored at one flight-list configuration — the one tabulated in
[§ the flight-list configuration](#sec-detection) — rather than re-tuned per
arm. A joint optimum might be better for some arm; finding it would multiply
cluster time by the number of arms, and one Spark job runs at a time. The
natural follow-up, now that the labelling is fixed, is to re-tune the flight
list on the recommended arm alone: one extra grid, not eight.
