---
title: Open Performance Data Initiative (OPDI)
toc: false
filters:
  - acronyms
acronyms:
  insert_loa: false

---

```{r}
library(bslib)
library(bsicons)
```



::: {layout-ncol=2}
```{r}
bslib::value_box(
  title = "Period",
  value = "01/2022 - 12/2024",
  showcase = bs_icon("calendar-range"),
  theme = "primary"
)
```

```{r}
bslib::value_box(
  title = "Flights",
  value = "43.54M", 
  showcase = bs_icon("airplane"),
  theme = "primary"
)
```
:::

::: {layout-ncol=2}
```{r}
bslib::value_box(
  title = "Events",
  value = "1.13B",
  showcase = bs_icon("calendar4-event"),
  theme = "primary"
)
```

```{r}
bslib::value_box(
  title = "Measurements",
  value = "2.26B",
  showcase = bs_icon("pencil-square"),
  theme = "primary"
)
```
:::

::: {.callout-tip}
## Second version available

The second iteration of the OPDI flight list, flight events and measurements is now available! Now extended to up until December 2024. See the description in [Methodology](/methodology.html) and download the data under [Data](/data.html).

In addition, get a preview of the data under [Data Preview](/data-preview.html).
:::

## Data

The flight data made available is published under the **[Open Performance Data Initiative (OPDI)](https://www.opdi.aero/)** as sponsored by
the **[Performance Review Commission](https://ansperformance.eu/about/prc/)** and in collaboration with the **[OpenSky Network (OSN)](https://opensky-network.org/)**. 

Download the data [here](https://opdi.aero/data).

## Concepts

To learn more about the concepts used within the OPDI data tables, see the [Concepts](https://opdi.aero/concepts).

## Methodology
The description of the algorithms used to identify flights, flight events and compute the measurements are described under [Methodology](https://opdi.aero/concepts).

## About

To learn more about the initiative, see [About](https://opdi.aero/concepts).
