---
title: Open Performance Data Initiative (OPDI)
execute:
  echo: false
  warning: false
---

```{r}
library(bslib)
library(bsicons)
```

::: {layout-ncol=4}
```{r}
bslib::value_box(
  title = "Period",
  value = "April 20xx",
  showcase = bs_icon("calendar-range")
)
```

```{r}
bslib::value_box(
  title = "Flights",
  value = "xx Million",
  showcase = bs_icon("airplane")
)
```

```{r}
bslib::value_box(
  title = "Events",
  value = "xx Million",
  showcase = bs_icon("calendar4-event")
)
```

```{r}
bslib::value_box(
  title = "Measurements",
  value = "xx Million",
  showcase = bs_icon("pencil-square")
)
```
:::

## Data

The flight data made available is published under the **[Open Performance Data Initiative (OPDI)](https://www.opdi.aero/)** as sponsored by
the **[Performance Review Commission](https://ansperformance.eu/about/prc/)** and in collaboration with the **[OpenSky Network (OSN)](https://opensky-network.org/)**. 

Download the data [here](https://opdi.aero/data).

## Concepts

To learn more about the concepts used within the OPDI data tables, see the [Concepts](https://opdi.aero/concepts).

## Methodology
The description of the algorithms used to identify flights, flight events and compute the measurements are described under [Methodology](https://opdi.aero/concepts).

## About

To learn more about the initiative, see [About](https://opdi.aero/concepts).
