---
title: Open Performance Data Initiative
filters:
  - acronyms
---

```{r setup}
#| include: false
library(bslib)
library(bsicons)
source("R/portal_config.R")
cfg <- portal_cfg
labs <- portal_labels
```

The **\acr{OPDI}** is an initiative of the \acr{PRC} to advance transparency, reproducibility, and democratisation of aviation performance monitoring. In collaboration with the [\acr{OSN}](https://opensky-network.org/), the \acr{OPDI} pursues three complementary objectives: publishing open flight performance data with transparent and reproducible methodologies, engaging the research community to develop and validate novel analytical approaches through data challenges, and expanding ground-based \acr{ADS-B} surveillance coverage by supporting crowd-sourced receiver infrastructure at underserved airports.

---

::: {.pillar-card .pillar-data}

### {{< iconify fa6-solid database >}} OPDI Data

Open flight performance data — flight lists, events, and measurements derived from \acr{ADS-B} data.

::: {layout-ncol=2}
```{r}
bslib::value_box(
  title = "Flights",
  value = cfg$counts$flights,
  showcase = bs_icon("airplane"),
  theme = value_box_theme(bg = "#003399")
)
```

```{r}
bslib::value_box(
  title = "Period",
  value = labs$short_month_range(cfg$coverage$start, cfg$coverage$end),
  showcase = bs_icon("calendar-range"),
  theme = value_box_theme(bg = "#003399")
)
```
:::

::: {.callout-note appearance="simple" style="border-left-color: #003399;"}
**Latest release (`r cfg$version`):** `r labs$month_range(cfg$coverage$start, cfg$coverage$end)`, refreshed `r cfg$refresh_label`. See [Methodology](content/methodology.qmd) for details.
:::

[Explore the data &rarr;](content/data.qmd){.btn .btn-sm style="background-color:#003399;color:white;border:none;"}

:::

::: {.pillar-card .pillar-challenges}

### {{< iconify fa6-solid trophy >}} PRC Data Challenges

Engaging data scientists worldwide to develop transparent, reproducible methods for aviation performance analysis.

::: {layout-ncol=2}
```{r}
bslib::value_box(
  title = "Editions",
  value = cfg$challenges$editions,
  showcase = bs_icon("trophy"),
  theme = value_box_theme(bg = "#00857c")
)
```

```{r}
bslib::value_box(
  title = "Teams",
  value = cfg$challenges$total_teams,
  showcase = bs_icon("people"),
  theme = value_box_theme(bg = "#00857c")
)
```
:::

[View challenges &rarr;](content/data-challenges.qmd){.btn .btn-sm style="background-color:#00857c;color:white;border:none;"}

:::

::: {.pillar-card .pillar-sensors}

### {{< iconify fa6-solid satellite-dish >}} OPDI Sensors

Extending the [OpenSky Network (OSN)](https://opensky-network.org/) crowd-sourced receiver network to improve ground-based and low-altitude \acr{ADS-B} coverage around airports, especially where existing coverage is limited.

::: {layout-ncol=2}
```{r}
bslib::value_box(
  title = "Sensors",
  value = cfg$sensors$total_sensors,
  showcase = bs_icon("broadcast"),
  theme = value_box_theme(bg = "#c35a00")
)
```

```{r}
bslib::value_box(
  title = "Airports",
  value = cfg$sensors$total_airports,
  showcase = bs_icon("geo-alt"),
  theme = value_box_theme(bg = "#c35a00")
)
```
:::

[Learn more &rarr;](content/sensors.qmd){.btn .btn-sm style="background-color:#c35a00;color:white;border:none;"}
[Extend the network &rarr;](content/sensors-contribute.qmd){.btn .btn-sm .btn-outline-secondary}
[Access sensor data &rarr;](content/sensors-data.qmd){.btn .btn-sm .btn-outline-secondary}

:::

::: {.pillar-card .pillar-community}

### {{< iconify fa6-solid handshake >}} OPDI Community

Working with research institutions, air navigation service providers, and open-data enthusiasts to advance open aviation performance data.

::: {layout-ncol=3}
```{r}
bslib::value_box(
  title = "Members",
  value = cfg$community$members,
  showcase = bs_icon("building"),
  theme = value_box_theme(bg = "#6a3d9a")
)
```

```{r}
bslib::value_box(
  title = "Countries",
  value = cfg$community$countries,
  showcase = bs_icon("globe2"),
  theme = value_box_theme(bg = "#6a3d9a")
)
```
:::

[Meet the community &rarr;](content/community.qmd){.btn .btn-sm style="background-color:#6a3d9a;color:white;border:none;"}

:::

::: {.pillar-card .pillar-resources}

### {{< iconify fa6-solid book-open >}} Resources

Peer-reviewed papers, open datasets, and the open-source software behind the \acr{OPDI}.

::: {layout-ncol=2}
```{r}
bslib::value_box(
  title = "Publications",
  value = cfg$resources$publications,
  showcase = bs_icon("journal-text"),
  theme = value_box_theme(bg = "#37536b")
)
```

```{r}
bslib::value_box(
  title = "Software Projects",
  value = cfg$resources$software,
  showcase = bs_icon("code-slash"),
  theme = value_box_theme(bg = "#37536b")
)
```
:::

[Browse resources &rarr;](content/resources.qmd){.btn .btn-sm style="background-color:#37536b;color:white;border:none;"}

:::
