---
title: Open Performance Data Initiative (OPDI)
filters:
  - acronyms
---

```{r setup}
library(bslib)
library(bsicons)
source("R/portal_config.R")
cfg <- portal_cfg
labs <- portal_labels
```



::: {layout-ncol=2}
```{r value_period}
bslib::value_box(
  title = "Period",
  value = labs$short_month_range(cfg$coverage$start, cfg$coverage$end),
  showcase = bs_icon("calendar-range"),
  theme = "primary"
)
```

```{r value_flights}
bslib::value_box(
  title = "Flights",
  value = cfg$counts$flights, 
  showcase = bs_icon("airplane"),
  theme = "primary"
)
```
:::

::: {layout-ncol=2}
```{r value_events}
bslib::value_box(
  title = "Events",
  value = cfg$counts$events,
  showcase = bs_icon("calendar4-event"),
  theme = "primary"
)
```

```{r value_measurements}
bslib::value_box(
  title = "Measurements",
  value = cfg$counts$measurements,
  showcase = bs_icon("pencil-square"),
  theme = "primary"
)
```
:::

::: {.callout-tip}
## Second version available

The second iteration of the OPDI flight list, flight events and measurements is now available! Now extended to up until `r format(cfg$coverage$end, "%B %Y")` (incl.). See the description in [Methodology](/methodology.html) and download the data under [Data](/data.html).

In addition, get a preview of the data under [Data Preview](/data-preview.html).
:::

::: {.callout-note appearance="simple"}
**Coverage:** `r labs$month_range(cfg$coverage$start, cfg$coverage$end)` (`r cfg$version` datasets, refreshed `r cfg$refresh_label`).
:::

## Data

The flight data made available is published under the **[Open Performance Data Initiative (OPDI)](https://www.opdi.aero/)** as sponsored by
the **[Performance Review Commission](https://ansperformance.eu/about/prc/)** and in collaboration with the **[OpenSky Network (OSN)](https://opensky-network.org/)**. 

Download the data [here](/data.html).

## Concepts

To learn more about the concepts used within the OPDI data tables, see [Concepts](/concepts.html).

## Methodology
The description of the algorithms used to identify flights, flight events and compute the measurements are described under [Methodology](/methodology.html).

## About

To learn more about the initiative, see [About](/about.html).
