---
title: "PRC Data Challenges"
---

```{r setup}
#| include: false
library(bslib)
library(bsicons)
source("../R/portal_config.R")
cfg <- portal_cfg
```

::: {layout-ncol=2}
```{r}
bslib::value_box(
  title = "Editions",
  value = cfg$challenges$editions,
  showcase = bs_icon("trophy"),
  theme = value_box_theme(bg = "#00857c")
)
```

```{r}
bslib::value_box(
  title = "Participating Teams",
  value = cfg$challenges$total_teams,
  showcase = bs_icon("people"),
  theme = value_box_theme(bg = "#00857c")
)
```

```{r}
bslib::value_box(
  title = "Countries",
  value = cfg$challenges$total_countries,
  showcase = bs_icon("globe2"),
  theme = value_box_theme(bg = "#00857c")
)
```

```{r}
bslib::value_box(
  title = "Submissions",
  value = cfg$challenges$total_submissions,
  showcase = bs_icon("upload"),
  theme = value_box_theme(bg = "#00857c")
)
```
:::

## About

The **PRC Data Challenges** invite data scientists, researchers, and aviation enthusiasts to tackle real-world aviation performance questions using open data. Organised by the [Performance Review Commission (PRC)](https://ansperformance.eu/about/prc/) of EUROCONTROL in collaboration with the [OpenSky Network](https://opensky-network.org/), each challenge defines a specific prediction or estimation task, provides curated open datasets, and evaluates submissions on a public leaderboard.

A core objective of the PRC Data Challenges is to foster **transparency and reproducibility** in aviation analytics. All participating teams are required to publish their code in open Git repositories, making their methodologies fully inspectable and reproducible. This approach aligns with the PRC's broader mission to democratise access to performance data and analytical methods, enabling independent validation and building a shared body of open, peer-reviewed techniques for the aviation community.

For more information, visit the [PRC Data Challenge programme page](https://ansperformance.eu/study/data-challenge/).


## Challenge Editions

:::: {.challenge-card .challenge-upcoming}

### 2026 PRC Data Challenge

[Coming soon]{.pill .pill-upcoming}

The next edition of the PRC Data Challenge is in development and will launch in 2026. The topic and dataset are yet to be announced.

Follow the [PRC Data Challenge programme page](https://ansperformance.eu/study/data-challenge/) for the announcement, or [get in touch](#get-involved) to be notified.

::::

:::: {.challenge-card}

### 2025 PRC Data Challenge — Fuel Flow Estimation

Participants estimated per-second fuel flow along flight trajectories, enabling granular emissions analysis using open data sources. The challenge attracted 179 teams from 48 countries, with 2,127 total submissions; 53 teams reached the final phase.

**Winners** (ranked by RMSE on the held-out test set):

| Place | Team | RMSE | Members | Affiliation | Country | Code |
|-------|------|-----:|---------|-------------|:-------:|:----:|
| 1st | [resourceful-quiver](https://ansperformance.eu/study/data-challenge/dc2025/teams/resourceful-quiver.html) | 200.83 kg/h | M.F. Rahman et al. | [TU Delft](https://www.tudelft.nl/) | NL | [{{< iconify fa6-brands github >}}](https://github.com/PRC-Data-Challenge-2025/resourceful-quiver) |
| 2nd | [jubilant-vase](https://ansperformance.eu/study/data-challenge/dc2025/teams/jubilant-vase.html) | 211.99 kg/h | A. Vernin et al. | [Sapienza](https://www.uniroma1.it/) / [ENSAI](https://ensai.fr/) | FR | [{{< iconify fa6-brands github >}}](https://github.com/PRC-Data-Challenge-2025/jubilant-vase) |
| 3rd | [outspoken-tornado](https://ansperformance.eu/study/data-challenge/dc2025/teams/outspoken-tornado.html) | 219.85 kg/h | B. Ryelandt et al. | [Euranova](https://euranova.eu/) | BE | [{{< iconify fa6-brands github >}}](https://github.com/PRC-Data-Challenge-2025/outspoken-tornado) |

::: {.callout-note appearance="simple" icon=false}
**Challenge dataset(s)**
<br>
{{< iconify fa6-solid book-open >}} [How the dataset was built](https://journals.open.tudelft.nl/joas/article/view/8750) — *Journal of Open Aviation Science*
<br>
{{< iconify fa6-solid database >}} [Download the dataset](https://doi.org/10.5281/zenodo.19184661) — Zenodo
:::

[Challenge details &rarr;](https://ansperformance.eu/study/data-challenge/dc2025/){.btn .btn-sm .btn-outline-secondary}

::::

:::: {.challenge-card}

### 2024 PRC Data Challenge — Actual Takeoff Weight Prediction

Participants predicted the actual takeoff weight (ATOW) of flights using open ADS-B trajectory data, weather information, and aircraft characteristics from a dataset of 369,013 European flights. The challenge attracted 132 teams, with around 550 submissions; 43 teams from 22 countries reached the final phase.

**Winners** (announced at the 12th OpenSky Symposium, Hamburg, November 2024):

| Place | Team | RMSE | Members | Affiliation | Country | Code |
|-------|------|-----:|---------|-------------|:-------:|:----:|
| 1st | [likable_jelly](https://ansperformance.eu/study/data-challenge/dc2024/teams.html#team_likable_jelly) | 1561.63 kg | R. Alligier et al. | [ENAC](https://www.enac.fr/) | FR | [{{< iconify fa6-brands github >}}](https://github.com/PRC-Data-Challenge-2024/team_likable_jelly) |
| 2nd | [tiny_rainbow](https://ansperformance.eu/study/data-challenge/dc2024/teams.html#team_tiny_rainbow) | 2217.75 kg | M. Murca et al. | [ITA](https://www.ita.br/) | BR | [{{< iconify fa6-brands github >}}](https://github.com/PRC-Data-Challenge-2024/team_tiny_rainbow) |
| 3rd | [brave_pillow](https://ansperformance.eu/study/data-challenge/dc2024/teams.html#team_brave_pillow) | 2270.06 kg | A. Tassanbi et al. | [TU Delft](https://www.tudelft.nl/) | NL | [{{< iconify fa6-brands github >}}](https://github.com/PRC-Data-Challenge-2024/team_brave_pillow) |

::: {.callout-note appearance="simple" icon=false}
**Challenge dataset(s)**
<br>
{{< iconify fa6-solid book-open >}} [How the dataset was built](https://journals.open.tudelft.nl/joas/article/view/8252) — *Journal of Open Aviation Science*
<br>
{{< iconify fa6-solid database >}} [Download the dataset](https://doi.org/10.4121/8cb8484b-dbe7-4750-8b87-a5b1dbc621b4) — 4TU.ResearchData
:::

[Challenge details &rarr;](https://ansperformance.eu/study/data-challenge/dc2024/){.btn .btn-sm .btn-outline-secondary}

::::

## Get Involved

Interested in participating in a future PRC Data Challenge? Follow the [PRC Data Challenge page](https://ansperformance.eu/study/data-challenge/) for announcements, or contact us at [pru-support@eurocontrol.int](mailto:pru-support@eurocontrol.int).
