---
title: "Where Next"
---

```{r setup}
#| include: false
library(knitr)
library(DT)
# Read-only: every figure comes from a cache under data/. Refresh with
#   Rscript R/update_sensor_cache.R
source("../R/portal_config.R")
source("../R/sensor_coverage.R")
source("../R/airport_capture.R")

sc <- get_sensor_coverage()
```

We are actively identifying airports where additional sensor deployment would most benefit open performance monitoring. Priority criteria include:

- Airports with limited or no existing ADS-B ground station coverage in the OpenSky Network
- Airports in regions underrepresented in open aviation data
- Airports of strategic importance for performance benchmarking

## Where does OPDI already see the traffic?

To find the gaps, we ask a simple question of every European airport: *of the
traffic an airport is known to have handled, how much does \acr{OPDI} actually
observe?*

**The two sources.** The reference is EUROCONTROL's published
[airport traffic statistics](https://www.eurocontrol.int/performance/data/download/csv/airport_traffic_2025.csv)
for 2025 — one row per airport per day, giving the number of departures
(`FLT_DEP_1`) and arrivals (`FLT_ARR_1`) recorded that day. Against it we place
the [OPDI flight list](flight-list-data.qmd), twelve monthly Parquet files in
which each row is one flight reconstructed from crowd-sourced \acr{ADS-B}
trajectories, carrying a departure aerodrome (`adep`) and an arrival aerodrome
(`ades`).

**How they are compared.** The common unit is the *movement*: one departure or
one arrival. A single flight is therefore counted twice — once at the airport it
left, once at the airport it reached — which is exactly how the reference
totals `FLT_DEP_1` and `FLT_ARR_1`. Summing the OPDI flight list the same way,
over the whole of 2025, gives a directly comparable count per airport.

**What is calculated.** For each airport,

$$\text{capture rate} = \frac{\text{movements OPDI attributes to the airport}}{\text{movements the airport reported}} \times 100\%$$

A flight whose trajectory begins or ends outside \acr{ADS-B} coverage has no
`adep` or no `ades`, and so contributes no movement anywhere. In 2025 that
applies to 47% of departures and 44% of arrivals in the flight list, which is
why a capture rate of 100% is not attainable in principle — the metric measures
how much of an airport's traffic \acr{OPDI} can *attribute to that airport*, not
the completeness of the trajectory data behind it.

::: {.callout-important appearance="simple"}
**The capture rate measures the detection algorithm as much as the coverage.**

`adep` and `ades` are not reported by the aircraft; they are *inferred* from the
trajectory. The current algorithm (`v0.0.2`) lays an
[H3](https://h3geo.org/) grid of 30 NM radius around every aerodrome, keeps the
state vectors below 4,000 ft, and assigns the airport whose reference point is
closest to the start or end of the track. Full details are on the
[Methodology](methodology.qmd#flight-list) page, and the resulting columns are
described under [Flight List data](flight-list-data.qmd).

Two consequences follow directly, and both move the capture rate:

- **Aerodromes close together can be confused.** Where several grids overlap, the
  nearest reference point wins and the runners-up are recorded as `adep_p` /
  `ades_p`. Paris Charles-de-Gaulle and Le Bourget, or Frankfurt and Egelsbach,
  sit well inside one another's 30 NM grids, so a flight may be attributed to
  the wrong neighbour — inflating one airport's capture rate at the other's
  expense.
- **The algorithm is an iterative approximation.** Successive versions have
  changed both the detection radius and the altitude ceiling, and a future
  refinement will attribute more flights at some airports and fewer at others.

A low capture rate is therefore a prompt to investigate — it may indicate absent
\acr{ADS-B} coverage, or it may indicate that the detection algorithm is
mis-assigning that airport's traffic.
:::

Two summaries answer different questions. **Capture rate** is movement-weighted —
of all the traffic in a size band, how much did \acr{OPDI} see? **Median airport**
weights every airport equally — what does the typical airport in the band look
like? They diverge where a few outliers dominate.

```{r capture-summary}
#| echo: false

cap <- read_airport_capture(update_sensor_pages_cache$capture$year)

if (!is.null(cap)) {
  bs <- cap$by_size
  i <- match(c("large", "medium", "small"), bs$tier)
  tbl <- data.frame(
    `Airport size`        = c("Large (≥50k movements)", "Medium (5k–50k)", "Small (<5k)"),
    Airports              = bs$airports[i],
    `Reported movements`  = format(bs$reported_movements[i], big.mark = ",", trim = TRUE),
    `Captured movements`  = format(bs$opdi_movements[i], big.mark = ",", trim = TRUE),
    `Capture rate`        = sprintf("%.1f%%", bs$capture_rate[i]),
    `Median airport`      = sprintf("%.1f%%", bs$median_capture_rate[i]),
    check.names = FALSE, stringsAsFactors = FALSE
  )
  knitr::kable(tbl, align = c("l", "r", "r", "r", "r", "r"))
}
```

```{r capture-note}
#| echo: false
#| output: asis

if (!is.null(cap)) {
  cat("::: {.callout-note appearance='simple'}\n",
      sprintf(
        paste0("Across %s reported airports in %d, OPDI attributes %s of %s movements (%.1f%%). ",
               "Coverage is near-complete at the busiest airports and thins out at smaller fields.\n\n",
               "The two sources count different populations: the reference reports IFR commercial ",
               "traffic, while the OPDI flight list contains every transponder-equipped flight, ",
               "including general aviation. At %d small aerodromes this pushes the ratio above 100%%, ",
               "so the rate is only meaningful where scheduled traffic dominates."),
        format(cap$totals$airports, big.mark = ","), cap$year,
        format(cap$totals$opdi_movements, big.mark = ","),
        format(cap$totals$reported_movements, big.mark = ","),
        cap$totals$capture_rate, cap$totals$over_reference),
      "\n:::\n", sep = "")
}
```

## Candidate airports

Every reported airport, sorted by capture rate. Those at the top — where
\acr{OPDI} attributes little or none of the reported traffic — sit outside
current \acr{ADS-B} ground coverage and are the strongest candidates for a
sensor. Use the **Size** filter to restrict the table to a traffic band, search
by name or country, or sort on any column.

```{r capture-gaps-note}
#| echo: false
#| output: asis

if (!is.null(cap)) {
  a <- cap$airports
  near_zero <- a[a$capture_rate < 1, ]
  # Name whichever airports actually top the list, rather than hardcoding them:
  # the set changes whenever the cache is regenerated.
  top <- head(near_zero[order(-near_zero$reported_movements), ], 3)

  cat("::: {.callout-note appearance='simple'}\n",
      sprintf(
        paste0("%d airports, together reporting %s movements in %d, are captured at under 1%%. ",
               "The largest is %s (%s movements reported, %s captured), followed by %s."),
        nrow(near_zero),
        format(sum(near_zero$reported_movements), big.mark = ","), cap$year,
        top$airport[1],
        format(top$reported_movements[1], big.mark = ","),
        format(top$opdi_movements[1], big.mark = ","),
        paste(top$airport[-1], collapse = " and ")),
      "\n:::\n", sep = "")
}
```

```{r capture-sensor-note}
#| echo: false
#| output: asis

if (!is.null(cap) && !is.null(sc$sensors)) {
  a <- cap$airports
  ours <- a[a$icao %in% unique(sc$sensors$airport_icao), ]
  ours <- ours[order(ours$capture_rate), ]
  worst <- head(ours, 3)

  cat("::: {.callout-warning appearance='simple'}\n",
      sprintf(
        paste0("**These figures cover the whole of %d, but not every OPDI sensor was live for all of it.** ",
               "The Georgian receivers were commissioned during the year, so a full-year capture rate ",
               "understates what they deliver today: %s. ",
               "By contrast the longer-established Swedish and Lithuanian sites reach 94%% or more. ",
               "A low capture rate at an airport that hosts an OPDI sensor therefore reflects when the ",
               "sensor came online, not its current contribution — the 40 NM analysis on the ",
               "[Impact](sensors-impact.qmd) page measures that."),
        cap$year,
        paste(sprintf("%s %.0f%%", worst$airport, worst$capture_rate), collapse = ", ")),
      "\n:::\n", sep = "")
}
```

```{r capture-full-table}
#| echo: false

if (!is.null(cap)) {
  a <- cap$airports
  # Lowest capture first; break ties by size so the biggest gaps lead.
  a <- a[order(a$capture_rate, -a$reported_movements), ]

  # A factor renders as a select box in DT's column filters, matching the size
  # bands used in the summary above.
  size_label <- factor(
    as.character(a$size_tier),
    levels = c("large", "medium", "small"),
    labels = c("Large (≥50k)", "Medium (5k–50k)", "Small (<5k)")
  )

  tbl <- data.frame(
    ICAO        = a$icao,
    Airport     = a$airport,
    Country     = a$state,
    Size        = size_label,
    Reported    = a$reported_movements,
    OPDI        = a$opdi_movements,
    `Capture %` = a$capture_rate,
    check.names = FALSE, stringsAsFactors = FALSE
  )

  DT::datatable(
    tbl,
    rownames = FALSE,
    filter = list(position = "top", clear = TRUE),
    options = list(
      pageLength = 15,
      lengthMenu = c(15, 25, 50, 100),
      order = list(),        # keep the data frame's own ordering
      scrollX = TRUE,
      columnDefs = list(list(className = "dt-right", targets = 4:6))
    ),
    class = "compact stripe hover"
  ) |>
    DT::formatCurrency(c("Reported", "OPDI"), currency = "", digits = 0) |>
    DT::formatRound("Capture %", digits = 1)
}
```

---

Interested in hosting a sensor, or in the raw data these receivers produce? See
[Extend the Network](sensors-contribute.qmd) and
[Access Sensor Data](sensors-data.qmd).
