---
title: "Impact on Airport Monitoring"
---

```{r setup}
#| include: false
library(knitr)
library(ggplot2)
# Read-only: every figure comes from a cache under data/. Refresh with
#   Rscript R/update_sensor_cache.R
source("../R/portal_config.R")
source("../R/airport_impact.R")
source("../R/airport_capture_daily.R")
```

Deploying a sensor is one thing; showing that it changed what can be observed is
another. This page answers that question twice over — once as a snapshot across
every instrumented airport, and once as a time series in which a single sensor
switching on is visible to the day.

## What the sensors add

How much aircraft observation would be lost without the \acr{OPDI} sensors? For
each airport we take every state vector within **40 NM** of the airport
reference point and ask which receivers in the \acr{OSN} reported it. A fix is
*exclusive* when only \acr{OPDI} sensors saw it: without them, that moment of
flight would go unobserved.

```{r impact-table}
#| echo: false

imp <- read_airport_impact()

if (is.null(imp)) {
  cat("::: {.callout-note appearance='simple'}\n",
      "Coverage improvement metrics are being recomputed.\n:::\n", sep = "")
} else {
  a <- imp$airports
  fmt_pct <- function(p, status) {
    ifelse(status == "commissioning", "—", sprintf("%.1f%%", p))
  }
  tbl <- data.frame(
    Airport            = sprintf("%s (%s)", a$name, a$icao),
    Country            = a$country,
    `Flight hours`     = ifelse(a$status == "commissioning", "—",
                                sprintf("%.0f", a$total_hours)),
    `OPDI-only hours`  = ifelse(a$status == "commissioning", "—",
                                sprintf("%.0f", a$exclusive_hours)),
    `Added coverage`   = fmt_pct(a$added_pct, a$status),
    check.names = FALSE, stringsAsFactors = FALSE
  )
  knitr::kable(tbl, align = c("l", "l", "r", "r", "r"))
}
```

```{r impact-note}
#| echo: false
#| output: asis

if (!is.null(imp)) {
  a <- imp$airports
  rep <- a[a$status == "reporting", ]
  commissioning <- a[a$status == "commissioning", ]

  lines <- sprintf(
    "Measured over a single day (%s) within %d NM of each airport reference point.",
    imp$analysis_date, imp$radius_nm)

  if (nrow(rep)) {
    lines <- c(lines, sprintf(
      "Across the reporting sites, OPDI sensors exclusively observed between %.0f%% and %.0f%% of flight hours.",
      min(rep$added_pct), max(rep$added_pct)))
  }
  if (nrow(commissioning)) {
    lines <- c(lines, sprintf(
      "%s is still being commissioned and is not yet reporting.",
      paste(commissioning$name, collapse = ", ")))
  }

  cat("::: {.callout-note appearance='simple'}\n",
      paste(lines, collapse = " "),
      "\n\nFlight hours count the time between consecutive receptions of the same aircraft. ",
      "State vectors carrying no receiver attribution (about 8% of fixes) are excluded, ",
      "so the totals describe traffic observed by an identified receiver rather than all traffic aloft.\n:::\n",
      sep = "")
}
```


## A sensor switching on

The 40 NM figures above measure a single day. Tracking the same airports over
time shows the effect of a deployment directly.

The chart below plots the **daily capture rate** at every airport hosting
\acr{OPDI} sensors: the share of an airport's reported movements — one
departure or one arrival — that the \acr{OPDI} flight list attributes to it,
measured against EUROCONTROL's published traffic statistics. A rate near 100%
means \acr{OPDI} sees essentially all of the airport's traffic. The full
method, and the caveats that come with it, are set out under
[Where Next](sensors-expansion.qmd).

```{r daily-capture-plot}
#| echo: false
#| out-width: "100%"
#| fig-height: 6
#| warning: false

dc <- read_airport_capture_daily()

if (!is.null(dc)) {
  library(ggplot2)
  m <- dc$daily
  # Ingest gaps are days the flight list is missing, not coverage events; drop
  # them from the line rather than draw a plunge to zero that means nothing.
  m$capture_rate[m$data_gap] <- NA_real_
  m$label <- sprintf("%s (%s)", m$airport, m$icao)

  labels <- unique(m[, c("icao", "label")])

  # Airport-specific events, detected in R/airport_capture_daily.R: an
  # activation is the day a sensor first reports; an outage is a span during
  # which one airport goes quiet while the others keep reporting.
  ev <- dc$events
  act <- out <- NULL
  if (length(ev) && is.data.frame(ev) && nrow(ev)) {
    ev <- merge(ev, labels, by = "icao")
    act <- subset(ev, kind == "activation" & !is.na(resumed))
    if (nrow(act)) act$at <- as.Date(act$resumed)
    out <- subset(ev, kind == "outage")
    if (nrow(out)) {
      out$xmin <- as.Date(out$from)
      # Extend to the last quiet day so the band covers the whole outage.
      out$xmax <- as.Date(out$to) + 1
    }
  }

  p <- ggplot(m, aes(x = day))

  if (!is.null(out) && nrow(out)) {
    p <- p + geom_rect(data = out, inherit.aes = FALSE,
                       aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
                       fill = "#c0392b", alpha = 0.10)
  }

  p <- p +
    geom_hline(yintercept = 100, linetype = "dotted",
               colour = "grey60", linewidth = 0.3) +
    geom_point(aes(y = capture_rate), colour = "#c35a00",
               alpha = 0.25, size = 0.5) +
    geom_line(aes(y = capture_rate_7d), colour = "#c35a00", linewidth = 0.7)

  if (!is.null(act) && nrow(act)) {
    p <- p + geom_vline(data = act, inherit.aes = FALSE,
                        aes(xintercept = at),
                        linetype = "dashed", colour = "#37536b", linewidth = 0.5)
  }

  p <- p +
    facet_wrap(~ label, ncol = 2) +
    # Palanga's general-aviation traffic runs past 150%; clip rather than let
    # one facet stretch the axis for all six.
    coord_cartesian(ylim = c(0, 130)) +
    scale_y_continuous(labels = function(x) paste0(x, "%"),
                       breaks = c(0, 50, 100)) +
    scale_x_date(date_labels = "%b", date_breaks = "1 month") +
    labs(x = NULL, y = "Daily capture rate",
         caption = paste0(
           "Points: daily rate. Line: 7-day mean. Dashed line: sensor activation.\n",
           "Red band: airport-specific outage. Days on which the OPDI flight list\n",
           "is missing network-wide are omitted.")) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank(),
          strip.text = element_text(face = "bold", size = 9),
          plot.caption = element_text(colour = "grey40", hjust = 0, size = 8))
  p
}
```

```{r daily-capture-note}
#| echo: false
#| output: asis

if (!is.null(dc)) {
  m <- dc$daily[!dc$daily$data_gap, ]
  ev <- dc$events

  act <- if (length(ev) && is.data.frame(ev)) subset(ev, kind == "activation") else NULL
  out <- if (length(ev) && is.data.frame(ev)) subset(ev, kind == "outage") else NULL

  lines <- character()

  if (!is.null(act) && nrow(act)) {
    a <- act[1, ]
    g <- m[m$icao == a$icao, ]
    on <- as.Date(a$resumed)
    before <- mean(g$capture_rate[g$day < on], na.rm = TRUE)
    after  <- mean(g$capture_rate[g$day >= on], na.rm = TRUE)
    nm <- unique(g$airport)[1]
    lines <- c(lines, sprintf(
      paste0("**%s is the clearest case.** Its capture rate sat at %.1f%% until %s, then stepped to ",
             "%.1f%% from the day its OPDI sensor began reporting. The other airports stay flat across ",
             "the same window, which is what makes the step attributable to the sensor rather than to a ",
             "change in traffic or in the flight-list pipeline."),
      nm, before, format(on - 1, "%d %B"), after))
  }

  if (!is.null(out) && nrow(out)) {
    o <- out[1, ]
    nm <- unique(m$airport[m$icao == o$icao])[1]
    lines <- c(lines, sprintf(
      paste0("The red band marks a %d-day outage at **%s**, from %s to %s, with reporting resuming on %s. ",
             "Only that airport goes quiet — its neighbours continue reporting throughout — so the cause ",
             "is local to the receiver, not the data pipeline."),
      o$days, nm, format(as.Date(o$from), "%d %B"), format(as.Date(o$to), "%d %B"),
      format(as.Date(o$resumed), "%d %B")))
  }

  # With no activation and no outage detected there is nothing to annotate, and
  # an empty callout would render as a bare box.
  if (length(lines)) {
    cat("::: {.callout-note appearance='simple'}\n",
        paste(lines, collapse = "\n\n"), "\n:::\n", sep = "")
  }
}
```

---

The sensors deployed so far close real gaps. Finding the next ones means asking
which airports are still unobserved: see [Where Next](sensors-expansion.qmd).
