---
title: "Detecting departure and destination aerodromes from ADS-B"
subtitle: "Version 3 — what the residual errors are made of"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

cache <- function(f) {
  p <- file.path("data", f)
  if (!file.exists(p)) return(NULL)
  read.csv(p, stringsAsFactors = FALSE)
}
methods <- cache("method_comparison.csv")
sweep   <- cache("abstention_sweep.csv")
attrib  <- cache("cascade_attribution.csv")
apt     <- cache("per_airport.csv")
errs    <- cache("error_pairs.csv")
trackq  <- cache("track_quality.csv")

# pandas writes booleans as "True"/"False"; read.csv keeps them character and
# R's == FALSE then silently matches nothing.
as_bool <- function(x) if (is.logical(x)) x else tolower(as.character(x)) %in% c("true","1","t")
for (d in c("methods","sweep")) if (!is.null(get(d)) && "cleaned" %in% names(get(d))) {
  tmp <- get(d); tmp$cleaned <- as_bool(tmp$cleaned); assign(d, tmp)
}
if (!is.null(apt)) apt$in_scope <- as_bool(apt$in_scope)
if (!is.null(errs)) errs$sched_to_unsched <- as_bool(errs$sched_to_unsched)

MONTHS <- sort(unique(methods$month))
DEFAULT_SP <- 15
pct <- function(x) ifelse(is.na(x), "—", sprintf("%.2f%%", 100 * x))
pp  <- function(x) ifelse(is.na(x), "—", sprintf("%+.2f pp", 100 * x))

# Main configuration: the default penalty, default (non per-side) variant.
base <- function() {
  d <- methods[methods$sched_penalty_nm == DEFAULT_SP & methods$variant == "default", ]
  d[order(d$month, -d$adep_overall), ]
}
m <- function(meth, field, month) {
  d <- base(); d <- d[d$method == meth & d$month == month, ]
  if (nrow(d) == 0) return(NA_real_); d[[field]][1]
}
```

## Summary {.unnumbered}

Version 3 follows a review of [version 2](../adep-ades-detection-v2/index.html).
Auditing V2 against its own outputs found four review points only partly
delivered and one delivered with a defect. Fixing them changed the results and,
more usefully, changed what the *remaining* errors are made of.

Three things came out of it.

**The out-of-area rule was pre-empting good answers.** The border test won
unconditionally over a qualifying aerodrome match, so Ponta Delgada — which sits
about 8 NM inside the western edge of the ingestion box — had its departures
labelled "out of area" with the aircraft still on the runway. A qualifying
aerodrome now wins and the border label is the fallback.

**Six of the seven most frequent departure errors were the same mistake:** a
scheduled civil airport losing to an adjacent military airfield. Izmir to Çiğli
Airbase, Naples to Grazzanise, Cagliari to Decimomannu, Catania to Sigonella,
Gdańsk to Cewice — 7 to 25 NM apart, and in every case the truth had scheduled
service and the prediction did not. Ranking candidates on a distance penalised
for aerodromes without scheduled service removes that class of error entirely.

**Out-of-area is reported separately now, and it does not work as well as V2
implied.** Precision is high — when the rule says out-of-area it is right about
93% of the time — but recall is only 10 to 15%. The concept is right and the
geometric test is weak, for a reason given in @sec-ooa.

```{r headline}
h <- do.call(rbind, lapply(MONTHS, function(mo) do.call(rbind, lapply(
  c("M0_current","M7_abstain"), function(mm) data.frame(
    Sample = mo, Method = mm,
    `ADEP cov` = pct(m(mm,"adep_coverage",mo)), `ADEP acc` = pct(m(mm,"adep_accuracy",mo)),
    `in-area cov` = pct(m(mm,"adep_inarea_coverage",mo)),
    `in-area acc` = pct(m(mm,"adep_inarea_accuracy",mo)),
    check.names = FALSE)))))
kable(h, align = "llrrrr",
      caption = "Departure detection: the production algorithm against endpoint abstention, headline and with out-of-area stripped out.")
```

## What changed since version 2 {.unnumbered}

::: {.callout-important}
* **Out-of-area precedence fixed** — an aerodrome match now beats the border label.
* **Scheduled-service tie-break added** — candidates ranked on distance penalised
  for aerodromes without scheduled service, swept rather than assumed.
* **Out-of-area reported explicitly** — recall, precision and in-area figures
  beside the headline, so naming skill and abstention skill cannot be conflated.
* **Per-side thresholds actually run** — and found not to help (@sec-perside).
* **Residual errors analysed** rather than described (@sec-errors).

Version 1 and version 2 remain published unchanged. Their numbers are not
restated here and are not comparable to these, because the scoring changed.
:::

## Reporting out-of-area {#sec-ooa}

A flight from Dubai to Amsterdam has a departure aerodrome that no European
ADS-B feed can name: the aircraft entered the area already airborne. V2 made
`OOA` a valid answer and a valid truth, which was right. What it did not do was
report the two abilities separately, and they are different abilities — naming
an aerodrome, and recognising that no aerodrome can be named.

That matters because roughly one departure in twelve is genuinely out of area,
so a method could buy accuracy by labelling liberally and an undecomposed table
would show only the improvement.

```{r ooa-table}
o <- do.call(rbind, lapply(MONTHS, function(mo) do.call(rbind, lapply(
  c("M7_abstain","M8_extrapolate"), function(mm) data.frame(
    Sample = mo, Method = mm,
    `truly OOA` = pct(m(mm,"adep_gt_ooa_share",mo)),
    `labelled OOA` = pct(m(mm,"adep_pred_ooa_share",mo)),
    `OOA recall` = pct(m(mm,"adep_ooa_recall",mo)),
    `OOA precision` = pct(m(mm,"adep_ooa_precision",mo)),
    check.names = FALSE)))))
kable(o, align = "llrrrr",
      caption = "Out-of-area detection for departures. Only M7 and M8 emit the label; the other methods never do, so their recall is zero by construction.")
```

High precision, low recall. The rule fires when a track *endpoint* sits within
30 NM of the bounding box edge — but an inbound flight's first recorded sample
appears wherever a receiver happens to pick it up, which is usually well inside
the boundary rather than at it. Most out-of-area departures therefore never look
like border cases at all.

The concept is sound; the geometric test is the wrong instrument. A phase-based
test would fit the physics better: a track whose first sample is **already in
the cruise, neither near an aerodrome nor descending toward one**, has an origin
outside the observed area regardless of where it is. That is not implemented
here and is the clearest single improvement available (@sec-limitations).

## The military-airfield confusion {#sec-sched}

The most frequent departure errors in version 2 formed one pattern:

| Truth (scheduled) | Predicted (not scheduled) | Apart |
|---|---|---|
| LTBJ Izmir | LTBL Çiğli Airbase | 14.9 NM |
| LIRN Naples | LIRM Grazzanise Air Base | 14.1 NM |
| LIEE Cagliari | LIED Decimomannu Air Base | 7.2 NM |
| LICC Catania | LICZ Sigonella Navy Air Base | 7.9 NM |
| EPGD Gdańsk | EPCE Cewice Naval Air Base | 24.6 NM |

: In every case the truth had scheduled service and the prediction did not. {tbl-colwidths="[34,46,20]"}

A departing aircraft's first recorded fix is often already a few miles into the
climb, and where a military field sits between the airport and the departure
track, the nearest-aerodrome rule takes it. Aviation knowledge resolves this
without any new data: a commercial flight almost certainly used the aerodrome
with scheduled service.

Candidates are therefore ranked on **effective distance** — real distance plus a
fixed penalty for aerodromes without scheduled service — so a military field has
to be clearly nearest rather than merely nearest. A penalty of zero reproduces
the unbiased ranking exactly, which is why it is swept.

```{r sched-sweep}
#| fig-cap: "Accuracy against the scheduled-service penalty, both samples and both sides. Coverage moves by less than 0.6 pp across the whole range, so this is very nearly a free choice; the curve peaking and turning is what shows the penalty is correcting a bias rather than simply favouring larger aerodromes."
#| fig-width: 8
#| fig-height: 4.2
sp <- methods[methods$method == "M7_abstain" & methods$variant == "default", ]
long <- rbind(
  data.frame(month = sp$month, penalty = sp$sched_penalty_nm,
             accuracy = sp$adep_accuracy, side = "ADEP"),
  data.frame(month = sp$month, penalty = sp$sched_penalty_nm,
             accuracy = sp$ades_accuracy, side = "ADES"))
ggplot(long, aes(penalty, accuracy, colour = month)) +
  geom_line(linewidth = 0.7) + geom_point(size = 2) +
  facet_wrap(~side, scales = "free_y") +
  scale_x_continuous("Penalty applied to aerodromes without scheduled service (NM)") +
  scale_y_continuous("Accuracy", labels = scales::percent) +
  scale_colour_manual(values = c("2024-06" = "#1b6ca8", "2025-06" = "#c1522b"), name = NULL) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(), strip.text = element_text(face = "bold"))
```

Accuracy peaks between 10 and 15 NM on both sides in both samples, then turns
over. Coverage falls by under 0.6 pp across the entire range. **10 NM is the
defensible choice** — it is at or within a whisker of the peak in all four
curves, and the flat top means the exact value is not load-bearing.

## Where the remaining errors are {#sec-errors}

The tie-break does not reduce errors uniformly. It removes one class and
introduces another, and the honest way to report that is by composition.

```{r error-composition}
ec <- errs[errs$side == "adep", ]
tot <- sum(ec$n)
cls <- data.frame(
  Class = c("Scheduled → unscheduled (the class the tie-break targets)",
            "Unscheduled → scheduled (the tie-break's own cost)",
            "Truth out-of-area, an aerodrome named",
            "Truth an aerodrome, labelled out-of-area"),
  Flights = c(
    sum(ec$n[ec$sched_to_unsched]),
    sum(ec$n[ec$t_scheduled != "yes" & ec$p_scheduled == "yes"]),
    sum(ec$n[ec$truth == "OOA"]),
    sum(ec$n[ec$predicted == "OOA"])),
  check.names = FALSE)
cls$Share <- pct(cls$Flights / tot)
kable(cls, align = "lrr",
      caption = sprintf("Composition of the %s most frequent departure errors after the tie-break.", format(tot, big.mark = ",")))
```

The class the tie-break targets is now **empty**. What dominates instead is its
mirror image: genuine movements at non-scheduled aerodromes — Biggin Hill,
Northolt — pulled to the nearby hub. That is the price of the preference, it is
visible, and it is smaller than what it replaced, which is why accuracy rose.

```{r error-table}
e <- head(ec[order(-ec$n), ], 12)
kable(data.frame(
  Sample = e$month, Truth = e$truth, Predicted = e$predicted,
  Flights = format(e$n, big.mark = ","),
  Apart = ifelse(is.na(e$apart_nm) | e$truth == "OOA" | e$predicted == "OOA",
                 "—", sprintf("%.1f NM", e$apart_nm)),
  check.names = FALSE),
  align = "lllrr",
  caption = "Most frequent departure confusions after the fixes.")
```

Two further groups are visible and neither is addressable by threshold tuning.
**Long-range misassignments** — Larnaca predicted as Beirut, 112 NM apart — are
tracks whose first sample is far from either, where nearest-aerodrome has no
real information. **Dense terminal areas** — Biggin Hill against London City at
10.5 NM, Cannes against Nice at 13.1 NM — need runway-level discrimination,
which the HexAero layouts OPDI already computes could provide and this study
does not use.

## Per-side thresholds do not help {#sec-perside}

Version 2 noted that departures and arrivals behave differently and suggested
fitting their thresholds separately. Doing it properly: fit each side's
threshold on **2024-06** by taking the most coverage available subject to
beating the production algorithm's accuracy, then apply to **2025-06**.

Both sides select the **same** cell, 40 NM and 15,000 ft AGL. The frontier's
shape is similar enough on the two sides that no separation emerges, so the
asymmetry between departures and arrivals is not something a threshold can fix —
it is in how much information the endpoint carries, not in where the cut-off
should sit.

```{r perside}
ps <- methods[methods$variant == "perside", ]
if (nrow(ps)) {
  d <- methods[methods$method == "M7_abstain" & methods$variant == "default" &
               methods$sched_penalty_nm == DEFAULT_SP & methods$month == ps$month[1], ]
  m0 <- methods[methods$method == "M0_current" & methods$variant == "default" &
                methods$sched_penalty_nm == DEFAULT_SP & methods$month == ps$month[1], ]
  kable(data.frame(
    Configuration = c("Production (M0)", "M7 default (20 NM / 15,000 ft)",
                      "M7 fitted on 2024-06 (40 NM / 15,000 ft)"),
    `ADEP cov` = pct(c(m0$adep_coverage, d$adep_coverage, ps$adep_coverage)),
    `ADEP acc` = pct(c(m0$adep_accuracy, d$adep_accuracy, ps$adep_accuracy)),
    `ADES cov` = pct(c(m0$ades_coverage, d$ades_coverage, ps$ades_coverage)),
    `ADES acc` = pct(c(m0$ades_accuracy, d$ades_accuracy, ps$ades_accuracy)),
    check.names = FALSE),
    align = "lrrrr",
    caption = sprintf("Validation on %s of thresholds fitted on the other sample.", ps$month[1]))
}
```

The fitted point transfers **for departures** — more coverage at better accuracy
than production, on a sample it was not fitted to. It does not transfer for
arrivals: it buys coverage at a real cost in accuracy. This is the third
independent time this study has found that arrivals resist what works for
departures.

## Everything else, unchanged

The cascade still collapses onto its fallback rung, the aircraft denylist still
outperforms the allowlist it replaced, trajectory extrapolation still fails on
arrivals, cleaning is still incompatible with endpoint-based arrival detection,
and `track_id` still merges and fragments flights at the rates version 2
measured. Those results are unaffected by the changes here and are not repeated;
see [version 2](../adep-ades-detection-v2/index.html).

```{r full-table}
b <- base()
kable(data.frame(
  Sample = b$month, Method = b$method,
  `ADEP cov` = pct(b$adep_coverage), `ADEP acc` = pct(b$adep_accuracy),
  `ADES cov` = pct(b$ades_coverage), `ADES acc` = pct(b$ades_accuracy),
  `ADEP in-area cov` = pct(b$adep_inarea_coverage),
  `ADEP in-area acc` = pct(b$adep_inarea_accuracy),
  check.names = FALSE),
  align = "llrrrrrr",
  caption = "All methods, both samples, at the default configuration.")
```

## Recommendations {#sec-recommendations}

### Already implemented and validated in this study

These need no further research. They are in `benchmarks/adep_ades.py` and are
measured above.

| Change | Evidence |
|---|---|
| Treat out-of-area as an answer, not a silence | @sec-ooa |
| Prefer a qualifying aerodrome over the border label | @sec-ooa |
| Rank candidates with a scheduled-service penalty (10 NM) | @sec-sched |
| Drop only *confirmed* non-aeroplanes; keep unidentified airframes | V2 |
| Restrict candidates to large and medium aerodromes | V2 |
| Report naming skill and abstention skill separately | @sec-ooa |

### Outstanding in the production pipeline

None of the above has been applied to `src/opdi/pipeline/flights.py`. The
production flight list still uses the altitude-trend vote and still emits a
null for both "outside the observed area" and "not determined". A patch
proposal follows so that the change can be reviewed rather than assumed.

## Production patch proposal {#sec-patch}

Scope is `src/opdi/pipeline/flights.py`. `tracks.py` is untouched, so
`track_id` continuity with every published release is unaffected — this changes
the flight list, not track construction.

**Replace** `_categorize_landing_take_off`, which counts the sign of a 5-sample
rolling-mean altitude difference and drops anything ambiguous, with the endpoint
rule:

1. Take the first and last state vector of the track.
2. Find the nearest aerodrome by **effective distance** — great-circle distance
   plus 10 NM for aerodromes without scheduled service — among large and medium
   aerodromes only.
3. Emit that aerodrome when the endpoint is within 20 NM of it and no more than
   15,000 ft above its elevation, or is flagged `on_ground`.
4. Otherwise, if the endpoint lies within 30 NM of the ingestion boundary, emit
   the out-of-area marker.
5. Otherwise emit null.

**Add two columns** to the flight list, `adep_source` and `ades_source`, taking
`aerodrome`, `out_of_area` or `undetermined`. This is the change with the
highest value per line: those three states are currently collapsed into a single
null, and they mean entirely different things to a downstream user.

**Version.** New `version` string for the flight list. Existing event types keep
theirs, so published v0.0.2 data stays reproducible.

**Aerodrome elevation** comes from OurAirports and is missing for 3.09% of large
and medium aerodromes. Where it is missing, fall back to the `on_ground` flag
rather than treating the elevation as sea level — at a 5,000 ft aerodrome that
error is the entire height threshold.

**Before applying, a reviewer should check:** that arrivals are *not* switched to
this rule on the strength of the departure result, since arrivals fail to
replicate every time; that the aerodrome reference used in production carries
`scheduled_service`; and that the out-of-area marker is agreed with downstream
consumers before it appears in a published column.

## Limitations {#sec-limitations}

- **Out-of-area recall is 10–15%.** The geometric border test is the wrong
  instrument; a phase-based test — first sample already in the cruise, not near
  or descending toward an aerodrome — should be tried next, and is the single
  clearest improvement available.
- **The scheduled-service preference has a cost**, and it is now the largest
  single class of remaining departure errors: genuine movements at Biggin Hill
  and Northolt pulled to the nearby hub.
- **Dense terminal areas are unresolved.** Biggin Hill against London City,
  Cannes against Nice. Runway-level matching using the HexAero layouts OPDI
  already computes is the obvious next step and is not used here.
- **Six days**, two three-day samples a year apart.
- Everything carried over from version 2: surface vehicles cannot be excluded by
  declaration, `track_id` is taken as given, and cleaning is not applied.

## Reproducing

```bash
python benchmarks/adep_ades.py --days 2025-06-05 2025-06-06 2025-06-07 \
    --months 202506 --methods all --sweep-abstain \
    --error-pairs M7_abstain --per-airport M7_abstain \
    --diagnose-cascade --control m1 --sched-penalty-nm 15 \
    --results-dir <dir>
```

`--results-dir` writes the result tables as CSV to a local directory instead of
S3. The bucket has a quota, and a write that exceeds it fails at the very end of
a job — after the numbers are computed and printed but before they are stored.

Rendering this page needs no credentials, no database and no cluster.
