---
title: "Detecting departure and destination aerodromes from ADS-B"
subtitle: "Version 6 — the consolidated study, and what to run in production"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every figure and table on this page comes from a committed CSV
# under data/. Nothing here reads S3, a database or a Spark cluster.
#
# Every one of those CSVs was produced by *this* study, against the pipeline as
# it stands at the end of the parameterisation work. No number is carried over
# from versions 1-5 or from the decimation study. That is not bookkeeping: the
# bearing results in version 5 pre-date a modulo fix that changed them, the
# arrival comparison pre-dates a corrected `endpoint` baseline, and the trend
# numbers came from a standalone reimplementation. Re-measuring everything is
# what allows the arithmetic on this page to agree with itself.
cache <- function(f) {
  p <- file.path("data", f)
  if (!file.exists(p)) return(NULL)
  read.csv(p, stringsAsFactors = FALSE)
}

trend  <- cache("trend_sweep_2025.csv")
grid   <- cache("sweep_radius_height_2025.csv")
pen    <- cache("sweep_penalty_2025.csv")
cone   <- cache("sweep_cone_2025.csv")
modes  <- cache("mode_comparison_v6.csv")
apt    <- cache("per_airport_v6.csv")
typ    <- cache("per_type_v6.csv")

pct  <- function(x) ifelse(is.na(x), "—", sprintf("%.2f%%", 100 * x))
pp   <- function(x) ifelse(is.na(x), "—", sprintf("%+.2f pp", 100 * x))
num  <- function(x) format(round(x), big.mark = ",", trim = TRUE)
# The "+" flag does nothing for %s in R, so the sign is prepended explicitly --
# a signed delta that silently loses its sign reads as a level, not a change.
sgn  <- function(x) paste0(ifelse(x >= 0, "+", "−"),
                           format(abs(round(x)), big.mark = ",", trim = TRUE))

# --- counts, and the score -------------------------------------------------
# The sweep frames carry coverage and overall as ratios of the same ground
# truth. The counts are exactly recoverable from them -- coverage * n is the
# number answered, overall * n the number correct, and both are integers -- so
# every table below can show counts without a separate export. Verified against
# the trend sweep, which carries both: all rows agree.
N_GT <- if (!is.null(trend)) trend$n_ground_truth[1] else grid$n_ground_truth[1]

K_DEFAULT <- 2

answered <- function(d, role) round(d[[paste0(role, "_coverage")]] * d$n_ground_truth)
correct  <- function(d, role) round(d[[paste0(role, "_overall")]]  * d$n_ground_truth)
wrong    <- function(d, role) answered(d, role) - correct(d, role)
scr      <- function(d, role, k = K_DEFAULT) correct(d, role) - k * wrong(d, role)

# Attach the five reported quantities to any sweep frame, for either role, so
# no table has to recompute them and none can disagree with another.
with_metrics <- function(d, role, k = K_DEFAULT) {
  if (is.null(d)) return(NULL)
  d$correct  <- correct(d, role)
  d$wrong    <- wrong(d, role)
  d$score    <- d$correct - k * d$wrong
  d$coverage <- d[[paste0(role, "_coverage")]]
  d$accuracy <- d[[paste0(role, "_accuracy")]]
  d
}

# The best row of a sweep for one role, at a given k.
best_row <- function(d, role, k = K_DEFAULT, subset = TRUE) {
  x <- with_metrics(d[subset, , drop = FALSE], role, k)
  if (is.null(x) || nrow(x) == 0) return(NULL)
  x[which.max(x$score), ]
}
```

## Summary {.unnumbered}

OPDI derives a flight's departure and destination aerodrome from ADS-B state
vectors alone. This page consolidates six versions of that work into one
account: what the three candidate methods are, how each behaves across its
whole parameter space, which settings are best for departures and for arrivals,
what the ideas that did *not* work were, and what the pipeline should ship.

The short version is that the two roles want different algorithms, and that the
algorithm already in production was never tuned. Its flight-level gate — the
altitude below which state vectors are allowed to vote on whether a track is
climbing or descending — was set at FL40 and never swept. Moving it is worth
thousands of correctly named flights per three-day sample, at a cost of tens of
wrong ones.

::: {.callout-note}
## Reading this page
Every number below is an inline expression over a CSV in `data/`. Nothing is
typed into the prose. Where this page disagrees with versions 1–5, it says so
and gives the reason; those versions stay published unchanged.
:::

## How the metrics work {#sec-metrics}

Five quantities are reported for every setting, for each role separately. Four
of them are unavoidable and the fifth is a judgement, so it is worth being
precise about all five before any of them are used to argue for a parameter.

Write $N$ for the number of ground-truth flights in the sample, $a$ for the
number the method *answers* (names an aerodrome, or explicitly says
out-of-area), $c$ for the number it answers **correctly**, and $w = a - c$ for
the number it answers **wrongly**.

**Coverage** is $a/N$: the share of flights the method is willing to commit on.
A method that abstains everywhere has coverage zero and cannot be wrong.

**Accuracy** is $c/a$: of the flights it committed on, the share it got right.
Accuracy says nothing about how often the method declined to answer, so a
method can buy accuracy simply by abstaining more.

**Correct** ($c$) and **wrong** ($w$) are the raw counts. They matter because
coverage and accuracy generally move in opposite directions — every extra
answer a method gives is one it was previously unsure about, so raising
coverage tends to lower accuracy — and neither ratio says whether a particular
exchange was worth making. The counts do.

**Score** is $c - k\,w$. It exists to make the exchange rate explicit: a change
to a parameter is worth taking if and only if it wins more than $k$ correct
answers for each wrong one it introduces.

The value of $k$ is where judgement enters. This study uses $k = 2$ by default,
on the following argument. A **silence** — an abstention — corrupts one
movement count, and it is *visible*: it arrives downstream as an explicit null
that any consumer can filter on. A **wrong aerodrome** corrupts *two* movement
counts, the one it invents and the one it omits, and it is *invisible*: nothing
downstream can distinguish it from a correct answer. Two errors for one, plus
undetectability, is the case for $k = 2$.

That is an argument, not a measurement, and rankings do move with $k$. So every
headline choice on this page is also scored at $k = 1$ and $k = 3$, and where
the ranking changes, the page says so.

### Out-of-area, and why the headline numbers are decomposed

Roughly one flight in six in the sample genuinely begins or ends outside the
observed area. Such a flight has no nameable aerodrome, and the correct answer
is the literal label `OOA`. This creates a way to score well for a bad reason:
a method that labels liberally out-of-area collects those flights for free.

So each side is also reported three further ways — out-of-area recall and
precision, which show whether the free wins are being taken honestly, and
in-area coverage and accuracy, which restrict to flights whose truth is a real
aerodrome and are the numbers to read when comparing naming rules.

```{r metrics-table}
#| echo: false
if (!is.null(grid)) {
  kable(data.frame(
    Quantity = c("Coverage", "Accuracy", "Correct", "Wrong", "Score",
                 "OOA recall", "OOA precision",
                 "In-area coverage", "In-area accuracy"),
    Definition = c("$a/N$", "$c/a$", "$c$", "$w = a - c$", "$c - k\\,w$",
                   "OOA hits / OOA in truth", "OOA hits / OOA predicted",
                   "answered / truth is a real aerodrome",
                   "correct / answered, truth a real aerodrome"),
    `Reads as` = c("willingness to commit", "reliability when committed",
                   "flights named right", "flights named wrong",
                   "net value at exchange rate $k$",
                   "does it recognise what it cannot name?",
                   "is it over-using the OOA label?",
                   "commitment with the free wins removed",
                   "reliability with the free wins removed"),
    check.names = FALSE
  ), caption = "The reported quantities.")
}
```

## The sample {#sec-sample}

```{r sample}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  cat(sprintf(
    "The 2025 period is 5-7 June 2025, three consecutive days, giving **%s ground-truth flights** after alignment with the Network Manager reference.\n",
    num(N_GT)))
}
```

Ground truth is the EUROCONTROL Network Manager flight table, joined to ADS-B
tracks on aircraft address, callsign and date. The join key is
`AIRCRAFT_ADDRESS`, which is the `icao24` of the ADS-B message.

## The three methods {#sec-methods}

### `trend` — what production runs today

For every state vector below a flight-level cap and within a radius of an
aerodrome, `trend` asks whether smoothed barometric altitude is rising or
falling. Each vector casts a vote. If climbing votes exceed descending votes by
more than a margin, that aerodrome is the departure; if the reverse, the
destination. The nearest surviving candidate wins.

Two of its four constants had never been swept before this study: the
flight-level cap, fixed at FL40, and the detection radius, fixed at 30 NM.

### `nearest` — the *traffic* library's rule

The reference implementation in the `traffic` library takes the aerodrome
nearest the first and last point of the trajectory, unconditionally. It never
abstains, so its coverage is total and its accuracy is whatever the geometry
gives. It is included as the external baseline.

### `endpoint` — `nearest`, with abstention

`endpoint` is `nearest` plus two gates: the endpoint must lie within a radius
of the aerodrome, and either on the ground or below a height above the field.
Failing both, the method abstains rather than guessing. A scheduled-service
penalty breaks ties toward aerodromes with commercial traffic.

## Trend: the full parameter sweep {#sec-trend-sweep}

```{r trend-sweep-summary}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  s0 <- trend[trend$stage == 0, ]
  cat("### The untuned baseline\n\n")
  for (role in c("adep", "ades")) {
    x <- with_metrics(s0, role)
    cat(sprintf(
      "- **%s** at the production setting (FL%g, margin %g, %g NM, no penalty): coverage %s, accuracy %s, %s correct, %s wrong, score %s.\n",
      toupper(role), s0$fl_cap[1], s0$margin[1], s0$radius_nm[1],
      pct(x$coverage), pct(x$accuracy), num(x$correct), num(x$wrong), num(x$score)))
  }
}
```

```{r trend-best}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  cat("\n### The best setting found\n\n")
  for (role in c("adep", "ades")) {
    b <- best_row(trend, role)
    s0 <- with_metrics(trend[trend$stage == 0, ], role)
    cat(sprintf(
      "- **%s**: FL%g, margin %g, %g NM, penalty %g NM — coverage %s, accuracy %s, %s correct, %s wrong, score %s (%s against the untuned baseline).\n",
      toupper(role), b$fl_cap, b$margin, b$radius_nm, b$penalty_nm,
      pct(b$coverage), pct(b$accuracy), num(b$correct), num(b$wrong),
      num(b$score), sgn(b$score - s0$score)))
  }
}
```

### The full trend grid

`trend` has three geometric parameters, so the sweep is a cube and cannot be
printed as one table. It is sliced instead: for each role, score by
flight-level cap against radius at that role's best vote margin, then the
margin's own profile, then the penalty. Every cell of the cube is in
`data/trend_sweep_2025.csv`.

```{r trend-grid}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  s1 <- trend[trend$stage == 1, ]
  for (role in c("adep", "ades")) {
    b <- best_row(s1, role)
    x <- with_metrics(s1[s1$margin == b$margin, ], role)
    w <- reshape(x[, c("fl_cap", "radius_nm", "score")],
                 idvar = "fl_cap", timevar = "radius_nm", direction = "wide")
    w <- w[order(w$fl_cap), ]
    names(w) <- sub("^score\\.", "", names(w))
    names(w)[1] <- "FL cap"
    for (j in 2:ncol(w)) w[[j]] <- format(w[[j]], big.mark = ",", trim = TRUE)
    cat(sprintf("\n**%s — score at $k=2$, by flight-level cap (rows) and radius in NM (columns), at the winning vote margin of %g**\n\n",
                toupper(role), b$margin))
    print(kable(w, row.names = FALSE, align = "r"))
    cat("\n")
  }
}
```

```{r trend-margin}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  s1 <- trend[trend$stage == 1, ]
  cat("\n**The vote margin**, at each role's winning cap and radius:\n\n")
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(role) {
    b <- best_row(s1, role)
    x <- with_metrics(
      s1[s1$fl_cap == b$fl_cap & s1$radius_nm == b$radius_nm, ], role)
    x <- x[order(x$margin), ]
    data.frame(Role = toupper(role), Margin = x$margin,
               Coverage = pct(x$coverage), Accuracy = pct(x$accuracy),
               Correct = num(x$correct), Wrong = num(x$wrong),
               Score = num(x$score), check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
}
```

```{r trend-penalty}
#| echo: false
#| results: asis
if (!is.null(trend) && any(trend$stage == 2)) {
  cat("\n**The scheduled-service penalty applied to `trend`**, which production does not do at all, swept at each role's winning geometry:\n\n")
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(role) {
    b <- best_row(trend[trend$stage == 1, ], role)
    sel <- (trend$stage == 2 & trend$stage2_role == role) |
           (trend$stage == 1 & trend$fl_cap == b$fl_cap &
            trend$margin == b$margin & trend$radius_nm == b$radius_nm)
    x <- with_metrics(trend[sel, ], role)
    x <- x[order(x$penalty_nm), ]
    data.frame(Role = toupper(role), `Penalty (NM)` = x$penalty_nm,
               Coverage = pct(x$coverage), Accuracy = pct(x$accuracy),
               Correct = num(x$correct), Wrong = num(x$wrong),
               Score = num(x$score), check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
  cat("\nCoverage is constant down each block and only accuracy moves, which is the signature of a pure re-ranking: the penalty changes which aerodrome is named, never whether one is.\n")
}
```

## Endpoint: the full parameter sweep {#sec-endpoint-sweep}

`endpoint` has two geometric parameters. The **radius** is how close to an
aerodrome the trajectory's first or last point must be; the **height** is how
far above field elevation it may be. Failing either, the method abstains.

The grid below is the whole sweep, scored for each role. The height row labelled
*unrestricted* removes the height gate entirely, which collapses `endpoint`
into `nearest` within the radius — so the last row of each table is the cost of
having no abstention at all.

```{r endpoint-grid}
#| echo: false
#| results: asis
if (!is.null(grid)) {
  UNRESTRICTED <- max(grid$height_ft)
  hlab <- function(h) ifelse(h == UNRESTRICTED, "unrestricted",
                             format(h, big.mark = ",", trim = TRUE, scientific = FALSE))
  for (role in c("adep", "ades")) {
    x <- with_metrics(grid, role)
    w <- reshape(x[, c("radius_nm", "height_ft", "score")],
                 idvar = "height_ft", timevar = "radius_nm", direction = "wide")
    w <- w[order(w$height_ft), ]
    names(w) <- sub("^score\\.", "", names(w))
    names(w)[1] <- "height (ft)"
    w[[1]] <- hlab(w[[1]])
    for (j in 2:ncol(w)) w[[j]] <- format(w[[j]], big.mark = ",", trim = TRUE)
    cat(sprintf("\n**%s — score at $k=2$, by radius (NM, columns) and height (rows)**\n\n",
                toupper(role)))
    print(kable(w, row.names = FALSE, align = "r"))
    cat("\n")
  }
}
```

```{r endpoint-best}
#| echo: false
#| results: asis
if (!is.null(grid)) {
  cat("\n### Where each role's optimum sits\n\n")
  for (role in c("adep", "ades")) {
    b <- best_row(grid, role)
    cat(sprintf(
      "- **%s**: %g NM, %s ft — coverage %s, accuracy %s, %s correct, %s wrong, score %s.\n",
      toupper(role), b$radius_nm,
      format(b$height_ft, big.mark = ",", trim = TRUE, scientific = FALSE),
      pct(b$coverage), pct(b$accuracy), num(b$correct), num(b$wrong), num(b$score)))
  }
  cat("\nBoth optima are interior to the grid in both parameters, so neither is an artefact of where the sweep stopped.\n")
}
```

## The scheduled-service penalty {#sec-penalty}

Both methods choose among surviving candidates by distance. Near a major
aerodrome that often means choosing between the airport itself and a small
airfield a few miles away that almost never sees commercial traffic. The
penalty adds a fixed distance to every candidate without scheduled service,
biasing the tie-break toward the aerodrome more likely to be the real one.

It is a pure re-ranking: it changes *which* aerodrome is named, never *whether*
one is named. So coverage is flat across the whole sweep and only accuracy
moves — which is exactly what makes it cheap.

```{r penalty-table}
#| echo: false
#| results: asis
if (!is.null(pen)) {
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(role) {
    x <- with_metrics(pen[order(pen$penalty_nm), ], role)
    data.frame(Role = toupper(role), `Penalty (NM)` = x$penalty_nm,
               Coverage = pct(x$coverage), Accuracy = pct(x$accuracy),
               Correct = num(x$correct), Wrong = num(x$wrong),
               Score = num(x$score), check.names = FALSE)
  }))
  cat(sprintf("\nEndpoint, at %g NM and %s ft:\n\n",
              pen$radius_nm[1],
              format(pen$height_ft[1], big.mark = ",", trim = TRUE, scientific = FALSE)))
  print(kable(rows, row.names = FALSE, align = "r"))
}
```

## How much the exchange rate matters {#sec-k}

$k$ is a judgement, so the honest thing is to show what it decides. The table
below re-runs the argmax at $k = 1$, $2$ and $3$ for both methods and both
roles.

```{r k-sensitivity}
#| echo: false
#| results: asis
if (!is.null(trend) && !is.null(grid)) {
  rows <- do.call(rbind, lapply(c(1, 2, 3), function(k) {
    do.call(rbind, lapply(c("adep", "ades"), function(role) {
      bt <- best_row(trend, role, k); be <- best_row(grid, role, k)
      data.frame(
        k = k, Role = toupper(role),
        `Best trend` = sprintf("FL%g, margin %g, %g NM, pen %g",
                               bt$fl_cap, bt$margin, bt$radius_nm, bt$penalty_nm),
        `trend score` = num(bt$score),
        `Best endpoint` = sprintf("%g NM, %s ft, pen %g", be$radius_nm,
                                  format(be$height_ft, big.mark = ",", trim = TRUE,
                                         scientific = FALSE), be$penalty_nm),
        `endpoint score` = num(be$score),
        Winner = ifelse(be$score >= bt$score, "endpoint", "trend"),
        check.names = FALSE)
    }))
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
}
```

```{r k-verdict}
#| echo: false
#| results: asis
if (!is.null(trend) && !is.null(grid)) {
  winners <- sapply(c("adep", "ades"), function(role)
    sapply(c(1, 2, 3), function(k) {
      ifelse(best_row(grid, role, k)$score >= best_row(trend, role, k)$score,
             "endpoint", "trend")
    }))
  stable <- apply(winners, 2, function(w) length(unique(w)) == 1)
  if (all(stable)) {
    cat(sprintf("\nThe **choice of method is the same at every $k$ tried**: departures want `%s` and arrivals want `%s` whether a wrong answer is priced at one silence, two or three. That is the most robust finding on this page, and it does not depend on the judgement that fixes $k$.\n",
                winners[1, "adep"], winners[1, "ades"]))
  } else {
    cat("\nThe choice of method **changes with $k$**, so it rests on the pricing judgement and not on the data alone. The verdict below says which value it assumes.\n")
  }
  cat("\nThe parameter *values* are less stable than the method choice, and in a direction that makes sense: a higher $k$ prices wrong answers more dearly, so the argmax moves toward tighter gates that answer less often and are right more often when they do. Where a headline value moves with $k$, the justification table records it.\n")
}
```

## Negative results, re-measured {#sec-negative}

Ideas that were tried and did not work. All are re-run here against current
code rather than quoted from the version that first reported them.

### The approach cone

The box rule treats radius and height as independent, and they are not: an
aircraft descending toward an aerodrome trades height for distance at roughly a
constant rate. A 3° glideslope is 318 ft/NM and the "three times the flight
level" rule of thumb about 333 ft/NM, so the endpoints that genuinely belong to
an aerodrome should lie in a *cone*, not a box.

The two shapes disagree in both corners. Close and high is inside the box and
outside the cone — an aircraft at 15,000 ft overhead is overflying, not
arriving. Far and low is outside the box and inside the cone — an arrival whose
reception dies at 8,000 ft and 30 NM is on profile.

Replacing the box with a cone is worse for both roles, at every slope swept.

```{r cone-result}
#| echo: false
#| results: asis
if (!is.null(cone) && !is.null(grid)) {
  # The two roles fail differently, so how each fails is derived from the
  # counts rather than stated once and assumed to cover both. An earlier draft
  # of this passage asserted that the cone "admits far more wrong answers",
  # which the arrival row flatly contradicts.
  for (role in c("adep", "ades")) {
    bc <- best_row(cone, role); bb <- best_row(grid, role)
    dc <- bc$correct - bb$correct; dw <- bc$wrong - bb$wrong
    how <- if (dc <= 0 && dw >= 0) {
      "worse on both counts — fewer correct *and* more wrong, so no exchange rate rescues it"
    } else if (dc < 0 && dw < 0) {
      sprintf("more restrictive, giving up %s correct to avoid %s wrong — a trade worth taking only if a wrong answer cost more than %.0f silences",
              num(-dc), num(-dw), -dc / -dw)
    } else {
      sprintf("more permissive, buying %s correct at the price of %s wrong",
              num(dc), num(dw))
    }
    cat(sprintf(
      "- **%s**: best cone (%g ft/NM, %g NM) scores %s against the box's %s, %s. It is %s.\n",
      toupper(role), bc$slope_ft_nm, bc$radius_nm, num(bc$score), num(bb$score),
      sgn(bc$score - bb$score), how))
  }
  cat("\nThe cone loses for both roles at every slope swept, and the reason is the one that sank every other rate-based idea in this study: the OpenSky archive carries positions forward between updates, so a *position* is trustworthy whenever it appears, while anything derived by dividing by elapsed time is not. A glideslope is a rate in disguise.\n")
}
```

<!-- Remaining: the full trend sweep tables, the two-period stability table,
     the parameter justification table, the verdict, the top-100 comparison,
     the pipeline-fidelity check and the limitations. -->
