---
title: "Detecting departure and destination aerodromes from ADS-B"
subtitle: "Version 6 — the consolidated study, and what to run in production"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every figure and table on this page comes from a committed CSV
# under data/. Nothing here reads S3, a database or a Spark cluster.
#
# Every one of those CSVs was produced by *this* study, against the pipeline as
# it stands at the end of the parameterisation work. No number is carried over
# from versions 1-5 or from the decimation study. That is not bookkeeping: the
# bearing results in version 5 pre-date a modulo fix that changed them, the
# arrival comparison pre-dates a corrected `endpoint` baseline, and the trend
# numbers came from a standalone reimplementation. Re-measuring everything is
# what allows the arithmetic on this page to agree with itself.
cache <- function(f) {
  p <- file.path("data", f)
  if (!file.exists(p)) return(NULL)
  read.csv(p, stringsAsFactors = FALSE)
}

trend  <- cache("trend_sweep_2025.csv")
grid   <- cache("sweep_radius_height_2025.csv")
pen    <- cache("sweep_penalty_2025.csv")
cone   <- cache("sweep_cone_2025.csv")
modes  <- cache("mode_comparison_v6.csv")
apt    <- cache("per_airport_v6.csv")
typ    <- cache("per_type_v6.csv")

pct  <- function(x) ifelse(is.na(x), "—", sprintf("%.2f%%", 100 * x))
pp   <- function(x) ifelse(is.na(x), "—", sprintf("%+.2f pp", 100 * x))
num  <- function(x) format(round(x), big.mark = ",", trim = TRUE)
# The "+" flag does nothing for %s in R, so the sign is prepended explicitly --
# a signed delta that silently loses its sign reads as a level, not a change.
sgn  <- function(x) paste0(ifelse(x >= 0, "+", "−"),
                           format(abs(round(x)), big.mark = ",", trim = TRUE))

# --- counts, and the score -------------------------------------------------
# The sweep frames carry coverage and overall as ratios of the same ground
# truth. The counts are exactly recoverable from them -- coverage * n is the
# number answered, overall * n the number correct, and both are integers -- so
# every table below can show counts without a separate export. Verified against
# the trend sweep, which carries both: all rows agree.
N_GT <- if (!is.null(trend)) trend$n_ground_truth[1] else grid$n_ground_truth[1]

K_DEFAULT <- 2

answered <- function(d, role) round(d[[paste0(role, "_coverage")]] * d$n_ground_truth)
correct  <- function(d, role) round(d[[paste0(role, "_overall")]]  * d$n_ground_truth)
wrong    <- function(d, role) answered(d, role) - correct(d, role)
scr      <- function(d, role, k = K_DEFAULT) correct(d, role) - k * wrong(d, role)

# Attach the five reported quantities to any sweep frame, for either role, so
# no table has to recompute them and none can disagree with another.
with_metrics <- function(d, role, k = K_DEFAULT) {
  if (is.null(d)) return(NULL)
  d$correct  <- correct(d, role)
  d$wrong    <- wrong(d, role)
  d$score    <- d$correct - k * d$wrong
  d$coverage <- d[[paste0(role, "_coverage")]]
  d$accuracy <- d[[paste0(role, "_accuracy")]]
  d
}

# The best row of a sweep for one role, at a given k.
best_row <- function(d, role, k = K_DEFAULT, subset = TRUE) {
  x <- with_metrics(d[subset, , drop = FALSE], role, k)
  if (is.null(x) || nrow(x) == 0) return(NULL)
  x[which.max(x$score), ]
}

# --- headline values, for the summary -------------------------------------
# Computed here rather than written into the prose, so the opening paragraph
# cannot drift away from the tables that justify it.
mv <- function(run, field) {
  if (is.null(modes)) return(NA_real_)
  v <- modes[[field]][modes$run == run]
  if (length(v) == 0) NA_real_ else v[1]
}
ADEP_GAIN <- mv("recommended", "adep_score") - mv("legacy", "adep_score")

top100_recall <- function(run, rl) {
  if (is.null(apt)) return(NA_real_)
  x <- apt[apt$run == run & apt$role == rl & apt$airport != "OOA", ]
  x <- x[order(-x$n_truth), ][1:min(100, nrow(x)), ]
  sum(x$n_correct) / sum(x$n_truth)
}
```

## Summary {.unnumbered}

OPDI derives a flight's departure and destination aerodrome from ADS-B state
vectors alone. This page consolidates six versions of that work into one
account: what the three candidate methods are, how each behaves across its
whole parameter space, which settings are best for departures and for arrivals,
what the ideas that did *not* work were, and what the pipeline should ship.

The short version is in three parts.

**The two roles want different algorithms.** Departures are best served by
`endpoint` — take the aerodrome nearest the first fix, but only commit when the
fix is close and low, and abstain otherwise. Arrivals are best served by
`trend`, the algorithm already in production. That split holds at every
exchange rate tested.

**Departures improve substantially.** Tuning `endpoint` and using it for the
departure half is worth **`r sgn(ADEP_GAIN)`** on the study's score over what
production does today, on a three-day sample of `r num(N_GT)` flights, lifting
recall across the hundred busiest aerodromes from
`r pct(top100_recall("legacy", "departures"))` to
`r pct(top100_recall("recommended", "departures"))`.

**The arrival tuning turned out to be an artefact of how it was measured**, and
this is the result worth carrying away. Sweeping `trend` suggested its
never-tuned flight-level gate should move from FL40 to FL120, apparently worth
thousands of flights. Run through the actual pipeline, that same change *lost*
ground. The sweep ranks candidate aerodromes by exact distance; the pipeline
ranks them by H3 ring count and tie-breaks within the ring. For `trend`, that
difference is larger than the effect being tuned. One arrival change survived —
a scheduled-service tie-break the algorithm never had — and it is the only one
adopted.

::: {.callout-note}
## Reading this page
Every number below is an inline expression over a CSV in `data/`. Nothing is
typed into the prose. Where this page disagrees with versions 1–5, it says so
and gives the reason; those versions stay published unchanged.
:::

## How the metrics work {#sec-metrics}

Five quantities are reported for every setting, for each role separately. Four
of them are unavoidable and the fifth is a judgement, so it is worth being
precise about all five before any of them are used to argue for a parameter.

Write $N$ for the number of ground-truth flights in the sample, $a$ for the
number the method *answers* (names an aerodrome, or explicitly says
out-of-area), $c$ for the number it answers **correctly**, and $w = a - c$ for
the number it answers **wrongly**.

**Coverage** is $a/N$: the share of flights the method is willing to commit on.
A method that abstains everywhere has coverage zero and cannot be wrong.

**Accuracy** is $c/a$: of the flights it committed on, the share it got right.
Accuracy says nothing about how often the method declined to answer, so a
method can buy accuracy simply by abstaining more.

**Correct** ($c$) and **wrong** ($w$) are the raw counts. They matter because
coverage and accuracy generally move in opposite directions — every extra
answer a method gives is one it was previously unsure about, so raising
coverage tends to lower accuracy — and neither ratio says whether a particular
exchange was worth making. The counts do.

**Score** is $c - k\,w$. It exists to make the exchange rate explicit: a change
to a parameter is worth taking if and only if it wins more than $k$ correct
answers for each wrong one it introduces.

The value of $k$ is where judgement enters. This study uses $k = 2$ by default,
on the following argument. A **silence** — an abstention — corrupts one
movement count, and it is *visible*: it arrives downstream as an explicit null
that any consumer can filter on. A **wrong aerodrome** corrupts *two* movement
counts, the one it invents and the one it omits, and it is *invisible*: nothing
downstream can distinguish it from a correct answer. Two errors for one, plus
undetectability, is the case for $k = 2$.

That is an argument, not a measurement, and rankings do move with $k$. So every
headline choice on this page is also scored at $k = 1$ and $k = 3$, and where
the ranking changes, the page says so.

### Out-of-area, and why the headline numbers are decomposed

Roughly one flight in six in the sample genuinely begins or ends outside the
observed area. Such a flight has no nameable aerodrome, and the correct answer
is the literal label `OOA`. This creates a way to score well for a bad reason:
a method that labels liberally out-of-area collects those flights for free.

So each side is also reported three further ways — out-of-area recall and
precision, which show whether the free wins are being taken honestly, and
in-area coverage and accuracy, which restrict to flights whose truth is a real
aerodrome and are the numbers to read when comparing naming rules.

```{r metrics-table}
#| echo: false
if (!is.null(grid)) {
  kable(data.frame(
    Quantity = c("Coverage", "Accuracy", "Correct", "Wrong", "Score",
                 "OOA recall", "OOA precision",
                 "In-area coverage", "In-area accuracy"),
    Definition = c("$a/N$", "$c/a$", "$c$", "$w = a - c$", "$c - k\\,w$",
                   "OOA hits / OOA in truth", "OOA hits / OOA predicted",
                   "answered / truth is a real aerodrome",
                   "correct / answered, truth a real aerodrome"),
    `Reads as` = c("willingness to commit", "reliability when committed",
                   "flights named right", "flights named wrong",
                   "net value at exchange rate $k$",
                   "does it recognise what it cannot name?",
                   "is it over-using the OOA label?",
                   "commitment with the free wins removed",
                   "reliability with the free wins removed"),
    check.names = FALSE
  ), caption = "The reported quantities.")
}
```

## The sample {#sec-sample}

```{r sample}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  cat(sprintf(
    "The 2025 period is 5-7 June 2025, three consecutive days, giving **%s ground-truth flights** after alignment with the Network Manager reference.\n",
    num(N_GT)))
}
```

Ground truth is the EUROCONTROL Network Manager flight table, joined to ADS-B
tracks on aircraft address, callsign and date. The join key is
`AIRCRAFT_ADDRESS`, which is the `icao24` of the ADS-B message.

## The three methods {#sec-methods}

### `trend` — what production runs today

For every state vector below a flight-level cap and within a radius of an
aerodrome, `trend` asks whether smoothed barometric altitude is rising or
falling. Each vector casts a vote. If climbing votes exceed descending votes by
more than a margin, that aerodrome is the departure; if the reverse, the
destination. The nearest surviving candidate wins.

Two of its four constants had never been swept before this study: the
flight-level cap, fixed at FL40, and the detection radius, fixed at 30 NM.

### `nearest` — the *traffic* library's rule

The reference implementation in the `traffic` library takes the aerodrome
nearest the first and last point of the trajectory, unconditionally. It never
abstains, so its coverage is total and its accuracy is whatever the geometry
gives. It is included as the external baseline.

### `endpoint` — `nearest`, with abstention

`endpoint` is `nearest` plus two gates: the endpoint must lie within a radius
of the aerodrome, and either on the ground or below a height above the field.
Failing both, the method abstains rather than guessing. A scheduled-service
penalty breaks ties toward aerodromes with commercial traffic.

## Trend: the full parameter sweep {#sec-trend-sweep}

```{r trend-sweep-summary}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  s0 <- trend[trend$stage == 0, ]
  cat("### The untuned baseline\n\n")
  for (role in c("adep", "ades")) {
    x <- with_metrics(s0, role)
    cat(sprintf(
      "- **%s** at the production setting (FL%g, margin %g, %g NM, no penalty): coverage %s, accuracy %s, %s correct, %s wrong, score %s.\n",
      toupper(role), s0$fl_cap[1], s0$margin[1], s0$radius_nm[1],
      pct(x$coverage), pct(x$accuracy), num(x$correct), num(x$wrong), num(x$score)))
  }
}
```

```{r trend-best}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  cat("\n### The best setting found\n\n")
  for (role in c("adep", "ades")) {
    b <- best_row(trend, role)
    s0 <- with_metrics(trend[trend$stage == 0, ], role)
    cat(sprintf(
      "- **%s**: FL%g, margin %g, %g NM, penalty %g NM — coverage %s, accuracy %s, %s correct, %s wrong, score %s (%s against the untuned baseline).\n",
      toupper(role), b$fl_cap, b$margin, b$radius_nm, b$penalty_nm,
      pct(b$coverage), pct(b$accuracy), num(b$correct), num(b$wrong),
      num(b$score), sgn(b$score - s0$score)))
  }
}
```

### The full trend grid

`trend` has three geometric parameters, so the sweep is a cube and cannot be
printed as one table. It is sliced instead: for each role, score by
flight-level cap against radius at that role's best vote margin, then the
margin's own profile, then the penalty. Every cell of the cube is in
`data/trend_sweep_2025.csv`.

```{r trend-grid}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  s1 <- trend[trend$stage == 1, ]
  for (role in c("adep", "ades")) {
    b <- best_row(s1, role)
    x <- with_metrics(s1[s1$margin == b$margin, ], role)
    w <- reshape(x[, c("fl_cap", "radius_nm", "score")],
                 idvar = "fl_cap", timevar = "radius_nm", direction = "wide")
    w <- w[order(w$fl_cap), ]
    names(w) <- sub("^score\\.", "", names(w))
    names(w)[1] <- "FL cap"
    for (j in 2:ncol(w)) w[[j]] <- format(w[[j]], big.mark = ",", trim = TRUE)
    cat(sprintf("\n**%s — score at $k=2$, by flight-level cap (rows) and radius in NM (columns), at the winning vote margin of %g**\n\n",
                toupper(role), b$margin))
    print(kable(w, row.names = FALSE, align = "r"))
    cat("\n")
  }
}
```

```{r trend-margin}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  s1 <- trend[trend$stage == 1, ]
  cat("\n**The vote margin**, at each role's winning cap and radius:\n\n")
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(role) {
    b <- best_row(s1, role)
    x <- with_metrics(
      s1[s1$fl_cap == b$fl_cap & s1$radius_nm == b$radius_nm, ], role)
    x <- x[order(x$margin), ]
    data.frame(Role = toupper(role), Margin = x$margin,
               Coverage = pct(x$coverage), Accuracy = pct(x$accuracy),
               Correct = num(x$correct), Wrong = num(x$wrong),
               Score = num(x$score), check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
}
```

```{r trend-penalty}
#| echo: false
#| results: asis
if (!is.null(trend) && any(trend$stage == 2)) {
  cat("\n**The scheduled-service penalty applied to `trend`**, which production does not do at all, swept at each role's winning geometry:\n\n")
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(role) {
    b <- best_row(trend[trend$stage == 1, ], role)
    sel <- (trend$stage == 2 & trend$stage2_role == role) |
           (trend$stage == 1 & trend$fl_cap == b$fl_cap &
            trend$margin == b$margin & trend$radius_nm == b$radius_nm)
    x <- with_metrics(trend[sel, ], role)
    x <- x[order(x$penalty_nm), ]
    data.frame(Role = toupper(role), `Penalty (NM)` = x$penalty_nm,
               Coverage = pct(x$coverage), Accuracy = pct(x$accuracy),
               Correct = num(x$correct), Wrong = num(x$wrong),
               Score = num(x$score), check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
  cat("\nCoverage is constant down each block and only accuracy moves, which is the signature of a pure re-ranking: the penalty changes which aerodrome is named, never whether one is.\n")
}
```

## Endpoint: the full parameter sweep {#sec-endpoint-sweep}

`endpoint` has two geometric parameters. The **radius** is how close to an
aerodrome the trajectory's first or last point must be; the **height** is how
far above field elevation it may be. Failing either, the method abstains.

The grid below is the whole sweep, scored for each role. The height row labelled
*unrestricted* removes the height gate entirely, which collapses `endpoint`
into `nearest` within the radius — so the last row of each table is the cost of
having no abstention at all.

```{r endpoint-grid}
#| echo: false
#| results: asis
if (!is.null(grid)) {
  UNRESTRICTED <- max(grid$height_ft)
  hlab <- function(h) ifelse(h == UNRESTRICTED, "unrestricted",
                             format(h, big.mark = ",", trim = TRUE, scientific = FALSE))
  for (role in c("adep", "ades")) {
    x <- with_metrics(grid, role)
    w <- reshape(x[, c("radius_nm", "height_ft", "score")],
                 idvar = "height_ft", timevar = "radius_nm", direction = "wide")
    w <- w[order(w$height_ft), ]
    names(w) <- sub("^score\\.", "", names(w))
    names(w)[1] <- "height (ft)"
    w[[1]] <- hlab(w[[1]])
    for (j in 2:ncol(w)) w[[j]] <- format(w[[j]], big.mark = ",", trim = TRUE)
    cat(sprintf("\n**%s — score at $k=2$, by radius (NM, columns) and height (rows)**\n\n",
                toupper(role)))
    print(kable(w, row.names = FALSE, align = "r"))
    cat("\n")
  }
}
```

```{r endpoint-best}
#| echo: false
#| results: asis
if (!is.null(grid)) {
  cat("\n### Where each role's optimum sits\n\n")
  for (role in c("adep", "ades")) {
    b <- best_row(grid, role)
    cat(sprintf(
      "- **%s**: %g NM, %s ft — coverage %s, accuracy %s, %s correct, %s wrong, score %s.\n",
      toupper(role), b$radius_nm,
      format(b$height_ft, big.mark = ",", trim = TRUE, scientific = FALSE),
      pct(b$coverage), pct(b$accuracy), num(b$correct), num(b$wrong), num(b$score)))
  }
  cat("\nBoth optima are interior to the grid in both parameters, so neither is an artefact of where the sweep stopped.\n")
}
```

## The scheduled-service penalty {#sec-penalty}

Both methods choose among surviving candidates by distance. Near a major
aerodrome that often means choosing between the airport itself and a small
airfield a few miles away that almost never sees commercial traffic. The
penalty adds a fixed distance to every candidate without scheduled service,
biasing the tie-break toward the aerodrome more likely to be the real one.

It is a pure re-ranking: it changes *which* aerodrome is named, never *whether*
one is named. So coverage is flat across the whole sweep and only accuracy
moves — which is exactly what makes it cheap.

```{r penalty-table}
#| echo: false
#| results: asis
if (!is.null(pen)) {
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(role) {
    x <- with_metrics(pen[order(pen$penalty_nm), ], role)
    data.frame(Role = toupper(role), `Penalty (NM)` = x$penalty_nm,
               Coverage = pct(x$coverage), Accuracy = pct(x$accuracy),
               Correct = num(x$correct), Wrong = num(x$wrong),
               Score = num(x$score), check.names = FALSE)
  }))
  cat(sprintf("\nEndpoint, at %g NM and %s ft:\n\n",
              pen$radius_nm[1],
              format(pen$height_ft[1], big.mark = ",", trim = TRUE, scientific = FALSE)))
  print(kable(rows, row.names = FALSE, align = "r"))
}
```

## How much the exchange rate matters {#sec-k}

$k$ is a judgement, so the honest thing is to show what it decides. The table
below re-runs the argmax at $k = 1$, $2$ and $3$ for both methods and both
roles.

```{r k-sensitivity}
#| echo: false
#| results: asis
if (!is.null(trend) && !is.null(grid)) {
  rows <- do.call(rbind, lapply(c(1, 2, 3), function(k) {
    do.call(rbind, lapply(c("adep", "ades"), function(role) {
      bt <- best_row(trend, role, k); be <- best_row(grid, role, k)
      data.frame(
        k = k, Role = toupper(role),
        `Best trend` = sprintf("FL%g, margin %g, %g NM, pen %g",
                               bt$fl_cap, bt$margin, bt$radius_nm, bt$penalty_nm),
        `trend score` = num(bt$score),
        `Best endpoint` = sprintf("%g NM, %s ft, pen %g", be$radius_nm,
                                  format(be$height_ft, big.mark = ",", trim = TRUE,
                                         scientific = FALSE), be$penalty_nm),
        `endpoint score` = num(be$score),
        Winner = ifelse(be$score >= bt$score, "endpoint", "trend"),
        check.names = FALSE)
    }))
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
}
```

```{r k-verdict}
#| echo: false
#| results: asis
if (!is.null(trend) && !is.null(grid)) {
  winners <- sapply(c("adep", "ades"), function(role)
    sapply(c(1, 2, 3), function(k) {
      ifelse(best_row(grid, role, k)$score >= best_row(trend, role, k)$score,
             "endpoint", "trend")
    }))
  stable <- apply(winners, 2, function(w) length(unique(w)) == 1)
  if (all(stable)) {
    cat(sprintf("\nThe **choice of method is the same at every $k$ tried**: departures want `%s` and arrivals want `%s` whether a wrong answer is priced at one silence, two or three. That is the most robust finding on this page, and it does not depend on the judgement that fixes $k$.\n",
                winners[1, "adep"], winners[1, "ades"]))
  } else {
    cat("\nThe choice of method **changes with $k$**, so it rests on the pricing judgement and not on the data alone. The verdict below says which value it assumes.\n")
  }
  cat("\nThe parameter *values* are less stable than the method choice, and in a direction that makes sense: a higher $k$ prices wrong answers more dearly, so the argmax moves toward tighter gates that answer less often and are right more often when they do. Where a headline value moves with $k$, the justification table records it.\n")
}
```

## Negative results, re-measured {#sec-negative}

Ideas that were tried and did not work. All are re-run here against current
code rather than quoted from the version that first reported them.

### The approach cone

The box rule treats radius and height as independent, and they are not: an
aircraft descending toward an aerodrome trades height for distance at roughly a
constant rate. A 3° glideslope is 318 ft/NM and the "three times the flight
level" rule of thumb about 333 ft/NM, so the endpoints that genuinely belong to
an aerodrome should lie in a *cone*, not a box.

The two shapes disagree in both corners. Close and high is inside the box and
outside the cone — an aircraft at 15,000 ft overhead is overflying, not
arriving. Far and low is outside the box and inside the cone — an arrival whose
reception dies at 8,000 ft and 30 NM is on profile.

Replacing the box with a cone is worse for both roles, at every slope swept.

```{r cone-result}
#| echo: false
#| results: asis
if (!is.null(cone) && !is.null(grid)) {
  # The two roles fail differently, so how each fails is derived from the
  # counts rather than stated once and assumed to cover both. An earlier draft
  # of this passage asserted that the cone "admits far more wrong answers",
  # which the arrival row flatly contradicts.
  for (role in c("adep", "ades")) {
    bc <- best_row(cone, role); bb <- best_row(grid, role)
    dc <- bc$correct - bb$correct; dw <- bc$wrong - bb$wrong
    how <- if (dc <= 0 && dw >= 0) {
      "worse on both counts — fewer correct *and* more wrong, so no exchange rate rescues it"
    } else if (dc < 0 && dw < 0) {
      sprintf("more restrictive, giving up %s correct to avoid %s wrong — a trade worth taking only if a wrong answer cost more than %.0f silences",
              num(-dc), num(-dw), -dc / -dw)
    } else {
      sprintf("more permissive, buying %s correct at the price of %s wrong",
              num(dc), num(dw))
    }
    cat(sprintf(
      "- **%s**: best cone (%g ft/NM, %g NM) scores %s against the box's %s, %s. It is %s.\n",
      toupper(role), bc$slope_ft_nm, bc$radius_nm, num(bc$score), num(bb$score),
      sgn(bc$score - bb$score), how))
  }
  cat("\nThe cone loses for both roles at every slope swept, and the reason is the one that sank every other rate-based idea in this study: the OpenSky archive carries positions forward between updates, so a *position* is trustworthy whenever it appears, while anything derived by dividing by elapsed time is not. A glideslope is a rate in disguise.\n")
}
```

### The bearing test

Version 5 proposed asking whether the trajectory *points at* the aerodrome: take
the track's course near its endpoint, take the bearing from that endpoint to the
candidate, and require the two to agree within some angle. It was the first
signal found that could add answers the abstention refused, and it was proposed
in four forms — **rescue** (answer anyway when aligned), **veto** (require
alignment even when the gate accepts), **replace** (alignment instead of the
gate) and **rerank** (choose *which* aerodrome by alignment).

All four are re-measured here against the current `endpoint` baseline.

```{r bearing}
#| echo: false
#| results: asis
bw <- cache("bearing_whole_sample_v6.csv")
if (!is.null(bw)) {
  n <- bw$n_ground_truth[1]
  sc <- function(role) {
    ans <- round(bw[[paste0(role, "_coverage")]] * n)
    cor <- round(bw[[paste0(role, "_overall")]] * n)
    cor - K_DEFAULT * (ans - cor)
  }
  bw$adep_s <- sc("adep"); bw$ades_s <- sc("ades")
  b <- bw[bw$model == "base", ]
  print(kable(data.frame(
    Variant = bw$model,
    `ADEP coverage` = pct(bw$adep_coverage), `ADEP accuracy` = pct(bw$adep_accuracy),
    `ADEP score` = num(bw$adep_s), `Δ` = c("—", sgn(bw$adep_s[-1] - b$adep_s)),
    `ADES coverage` = pct(bw$ades_coverage), `ADES accuracy` = pct(bw$ades_accuracy),
    `ADES score` = num(bw$ades_s), `Δ ` = c("—", sgn(bw$ades_s[-1] - b$ades_s)),
    check.names = FALSE), row.names = FALSE, align = "r"))

  best <- bw[which.max(bw$ades_s), ]
  cat(sprintf("
The best variant of any kind is `%s`, worth %s on arrivals — and %s on departures, which is the role `endpoint` is actually used for. Nothing here is worth adopting.
",
              best$model, sgn(best$ades_s - b$ades_s), sgn(best$adep_s - b$adep_s)))
  cat(sprintf("
The three variants that let bearing *remove* or *choose* answers are not close: the best of them gives up %s on departures. Alignment is far too blunt to name an aerodrome — at 40 NM a runway subtends a couple of degrees, but so does every other aerodrome behind it on the same radial, and a track on final approach points at all of them equally.
",
              sgn(max(bw$adep_s[grepl("^(veto|replace|rerank)", bw$model)]) - b$adep_s)))
  cat("
This supersedes version 5's recommendation. That version scored the bearing rescue against `endpoint` alone and never against `trend`, which was already better than both — the comparison that mattered was the one not made.
")
}
```

## What the pipeline actually produces {#sec-pipeline}

Every number so far comes from a sweep harness, not from `src/opdi`. The trend
sweep re-applies the vote rule to a cached vote table and the endpoint sweeps
filter and re-rank a cached candidate table — which is what makes hundreds of
parameter cells affordable, and also what makes them unfit, on their own, to
justify changing a production default.

So the recommended settings were run through `FlightListProcessor.process_dai`
itself and scored the same way. Three differences between harness and pipeline
are known in advance and are not bugs:

1. the pipeline selects among candidate aerodromes by **H3 ring count**
   (`distance_from_center`, an integer stepping about 5.2 km at resolution 7),
   while the harness uses exact haversine distance;
2. the pipeline filters zones by radius **band**, the harness by exact radius;
3. the harness smooths barometric altitude across the flight-level boundary and
   the pipeline does not.

The gap those produce is the fidelity of every tuned number on this page.

```{r pipeline-fidelity}
#| echo: false
#| results: asis
if (!is.null(modes) && !is.null(trend) && !is.null(grid)) {
  rows <- do.call(rbind, lapply(seq_len(nrow(modes)), function(i) {
    r <- modes[i, ]
    do.call(rbind, lapply(c("adep", "ades"), function(role) data.frame(
      Run = r$run, Role = toupper(role),
      Mode = r[[paste0(role, "_mode")]],
      Coverage = pct(r[[paste0(role, "_coverage")]]),
      Accuracy = pct(r[[paste0(role, "_accuracy")]]),
      Correct  = num(r[[paste0(role, "_correct")]]),
      Wrong    = num(r[[paste0(role, "_wrong")]]),
      Score    = num(r[[paste0(role, "_score")]]),
      check.names = FALSE)))
  }))
  print(kable(rows, row.names = FALSE, align = "r"))

  # The harness's own figure for the same setting, so the residual is visible
  # rather than implied.
  cat("\n**Harness against pipeline, at the same parameters:**\n\n")
  cmp <- do.call(rbind, list(
    local({
      h <- best_row(grid, "adep"); p <- modes[modes$run == "endpoint", ]
      data.frame(Role = "ADEP", Method = "endpoint",
                 `Harness score` = num(h$score),
                 `Pipeline score` = num(p$adep_score),
                 Residual = sgn(p$adep_score - h$score),
                 `Relative` = sprintf("%+.2f%%", 100 * (p$adep_score - h$score) / h$score),
                 check.names = FALSE)
    }),
    local({
      # From the `combined` run, not the `trend` run. A DetectionConfig holds
      # one trend setting, so the single-mode `trend` run has both roles at the
      # *departure* optimum and its arrival figure describes a setting nobody
      # recommends. `combined` is where trend runs at the arrival optimum.
      h <- best_row(trend[trend$stage == 1, ], "ades"); p <- modes[modes$run == "combined", ]
      data.frame(Role = "ADES", Method = "trend",
                 `Harness score` = num(h$score),
                 `Pipeline score` = num(p$ades_score),
                 Residual = sgn(p$ades_score - h$score),
                 `Relative` = sprintf("%+.2f%%", 100 * (p$ades_score - h$score) / h$score),
                 check.names = FALSE)
    })))
  print(kable(cmp, row.names = FALSE, align = "r"))
  cat("\nThe two rows could hardly be more different, and the difference is the single most important result on this page.\n")
  cat("\n`endpoint` **reproduces exactly**. That is not luck: the endpoint sweep filters the same cached candidate table the pipeline reads, applying the same rule, so harness and pipeline are the same computation and agree to the flight.\n")
  cat("\n`trend` **does not**. The harness ranks candidate aerodromes by exact haversine distance. The pipeline first keeps only the candidates at the minimum H3 *ring count* — an integer stepping about 5.2 km at resolution 7 — and applies the haversine tie-break, and the scheduled-service penalty, only within that surviving set. An aerodrome one ring further out is gone before the penalty can promote it. So a penalty tuned against exact distance does something weaker in production, and the harness's trend optimum is optimistic by the margin shown.\n")
}
```

### What running the pipeline caught

The first pipeline path run produced a result that no amount of sweeping could
have: the scheduled-service penalty step scored **exactly** what the step before
it did, to the flight. An identical score is not a weak effect, it is no effect,
and a parameter with no effect is a parameter that is not reaching the code.

It was not. The penalty was added to the ranking in `_compute_flight_table`,
but the projection a few lines above it dropped the `apt_scheduled` column, so
the ranking looked the column up, failed to find it, and fell back to raw
distance. A penalty of ten nautical miles and a penalty of zero produced
byte-identical flight lists.

Two things made this visible rather than invisible. The pipeline run put the
parameter in front of real data, where a sweep harness — which reads a
*different* table that does carry the column — could not. And the fallback
printed a note saying the column was missing instead of quietly ranking on
distance, because a missing column and a zero penalty look identical in the
output. The numbers below are from after the fix.

### The trend tuning re-walked inside the pipeline

Because the harness overstates `trend`, the path decomposition was run again
through `process_dai` itself — the same four steps, each one a real pipeline
run scored against the same ground truth. This is the version to trust.

```{r pipeline-path}
#| echo: false
#| results: asis
ppath <- cache("pipeline_path_v6.csv")
if (!is.null(ppath)) {
  lab <- c(path0_legacy = "production constants",
           path1_penalty = "+ scheduled-service penalty",
           path2_flcap = "+ flight-level cap",
           path3_margin = "+ vote margin",
           path4_radius = "+ radius")
  ppath <- ppath[order(match(ppath$run, names(lab))), ]
  sc <- ppath$ades_score
  print(kable(data.frame(
    Step = lab[ppath$run],
    Coverage = pct(ppath$ades_coverage), Accuracy = pct(ppath$ades_accuracy),
    Correct = num(ppath$ades_correct), Wrong = num(ppath$ades_wrong),
    Score = num(sc),
    `Gain from this step` = c("—", sgn(diff(sc))),
    check.names = FALSE), row.names = FALSE, align = "r"))

  net <- sc[length(sc)] - sc[1]
  if (net < 0) {
    hp <- best_row(trend[trend$stage == 1, ], "ades")$score -
          with_metrics(trend[trend$stage == 0, ], "ades")$score
    cat(sprintf("\n**The tuning does not transfer.** Inside the pipeline the full recommended arrival setting scores %s against the production constants, where the harness predicted %s — opposite signs. Every number in @sec-trend-sweep is measured on a ranking rule the pipeline does not use, and for `trend` that difference is larger than the effect being tuned.\n",
                sgn(net), sgn(hp)))

    # Which steps survive is the actionable part, and it is read off the table
    # rather than asserted: a step is adopted only if it gains ground here.
    gains <- diff(sc); names(gains) <- lab[ppath$run][-1]
    good <- names(gains)[gains > 0]; bad <- names(gains)[gains <= 0]
    cat(sprintf("\nThe steps do not fail together. Taken in order, these gain ground: %s; and these lose it: %s. Only the first step's figure is a standalone verdict, though — each later step is measured on top of the ones before it, so the %s and %s gains are recovering part of what the cap gave away rather than showing that either is worth adopting on its own.\n",
                paste(sub("^\\+ ", "", good), collapse = ", "),
                paste(sub("^\\+ ", "", bad), collapse = ", "),
                sub("^\\+ ", "", lab[["path3_margin"]]),
                sub("^\\+ ", "", lab[["path4_radius"]])))
    cat(sprintf("\nThe one that matters is the %s, worth %s on its own — and the flight-level cap, which the harness liked most of all, is precisely the step that loses most here (%s). Raising the cap admits state vectors from higher up, where the aircraft is further out, more aerodromes fall in the same H3 ring, and the tie-break the pipeline uses to separate them is coarser than the one the sweep was scored on. So it answers more often (%s coverage) and is right less often (%s accuracy) — a trade the harness's exact-distance ranking never had to make.\n",
                sub("^\\+ ", "", names(gains)[which.max(gains)]),
                sgn(max(gains)), sgn(min(gains)),
                pp(ppath$ades_coverage[3] - ppath$ades_coverage[2]),
                pp(ppath$ades_accuracy[3] - ppath$ades_accuracy[2])))
    cat("\n**What is adopted for arrivals, therefore, is the production geometry with the scheduled-service penalty added** — the only tuned value that is worth anything in the code that will actually run.\n")
  } else {
    cat(sprintf("\nThe tuning transfers: inside the pipeline the recommended arrival setting scores %s against the production constants, against the harness's prediction of %s.\n",
                sgn(net), sgn(best_row(trend[trend$stage == 1, ], "ades")$score -
                              with_metrics(trend[trend$stage == 0, ], "ades")$score)))
  }
}
```

## Why each value, and not its neighbour {#sec-justification}

A recommendation is only as good as the evidence that picked it, so every value
the sweeps select appears below with the score at that value, the score at the
adjacent grid points, and the score at whatever production uses today. The last
column is the one that matters most for trusting the number: a value on a broad
**plateau** is a safe recommendation even from a single sample, because its
neighbours score nearly the same and a little noise cannot move it far. A value
on a sharp isolated **peak** may be partly the sample's noise, and is flagged
as such however good its score.

::: {.callout-important}
## These are the sweeps' choices, not the shipped ones
The `Chosen` column below is the **harness argmax**, which is what this section
exists to scrutinise. For `trend` it is superseded by @sec-pipeline, where the
same values are run through the pipeline and mostly fail. What actually ships
is in @sec-defaults. The two are deliberately kept apart: the sweep's
preference is evidence, and evidence that was overruled is still worth showing.
:::

```{r justification}
#| echo: false
#| results: asis

# The score at one grid point, holding every other swept parameter at the
# chosen value. This is what makes the neighbour columns meaningful: they vary
# one parameter at a time rather than reporting the best cell that happens to
# contain that value.
at <- function(d, role, fixed, param, value, k = K_DEFAULT) {
  sel <- rep(TRUE, nrow(d))
  for (nm in names(fixed)) if (nm != param) sel <- sel & d[[nm]] == fixed[[nm]]
  sel <- sel & d[[param]] == value
  x <- with_metrics(d[sel, , drop = FALSE], role, k)
  if (is.null(x) || nrow(x) == 0) return(NA_real_)
  x$score[1]
}

# One row of the table: chosen value, its two neighbours on the grid, and the
# shape that follows from them.
just_row <- function(d, role, chosen, param, label, orig_value, k = K_DEFAULT,
                     fixed_names) {
  fixed <- as.list(chosen[fixed_names])
  vals  <- sort(unique(d[[param]]))
  i     <- match(chosen[[param]], vals)
  s_ch  <- at(d, role, fixed, param, vals[i], k)
  s_lo  <- if (i > 1) at(d, role, fixed, param, vals[i - 1], k) else NA_real_
  s_hi  <- if (i < length(vals)) at(d, role, fixed, param, vals[i + 1], k) else NA_real_
  s_or  <- at(d, role, fixed, param, orig_value, k)

  # Plateau if the better neighbour is within 0.5% of the chosen score --
  # a margin far smaller than any effect the report argues from.
  nb   <- c(s_lo, s_hi); nb <- nb[!is.na(nb)]
  gap  <- if (length(nb)) (s_ch - max(nb)) / s_ch else NA_real_
  shape <- if (is.na(gap)) "—" else if (gap < 0.005) "plateau" else "peak"

  data.frame(
    Role = toupper(role), Parameter = label,
    Production = orig_value, `Sweep argmax` = chosen[[param]],
    `Score at production` = ifelse(is.na(s_or), "—", num(s_or)),
    `Score at argmax`     = num(s_ch),
    `Neighbour below`   = ifelse(is.na(s_lo), "—", num(s_lo)),
    `Neighbour above`   = ifelse(is.na(s_hi), "—", num(s_hi)),
    `Δ vs production`   = ifelse(is.na(s_or), "—", sgn(s_ch - s_or)),
    Shape = shape, check.names = FALSE)
}

if (!is.null(trend) && !is.null(grid)) {
  s1 <- trend[trend$stage == 1, ]
  tf <- c("fl_cap", "margin", "radius_nm")
  ef <- c("radius_nm", "height_ft")

  rows <- list()
  # trend, for the role it is recommended for (arrivals), against the
  # production constants: FL40, margin 4, 30 NM.
  bt <- best_row(s1, "ades")
  rows[[length(rows) + 1]] <- just_row(s1, "ades", bt, "fl_cap",    "trend FL cap",       40, fixed_names = tf)
  rows[[length(rows) + 1]] <- just_row(s1, "ades", bt, "margin",    "trend vote margin",   4, fixed_names = tf)
  rows[[length(rows) + 1]] <- just_row(s1, "ades", bt, "radius_nm", "trend radius (NM)",  30, fixed_names = tf)

  # endpoint, for departures, against the version 4.5 operating point that the
  # pipeline currently defaults to: 40 NM, 15,000 ft.
  be <- best_row(grid, "adep")
  rows[[length(rows) + 1]] <- just_row(grid, "adep", be, "radius_nm", "endpoint radius (NM)",    40, fixed_names = ef)
  rows[[length(rows) + 1]] <- just_row(grid, "adep", be, "height_ft", "endpoint height (ft)", 15000, fixed_names = ef)

  print(kable(do.call(rbind, rows), row.names = FALSE, align = "r"))
  cat("\nThe penalty is treated separately because it is not a geometric parameter: it re-ranks candidates without changing how many are answered, so its sweep holds the geometry fixed and appears in @sec-penalty.\n")
}
```

### Which change is actually doing the work

The table above varies one parameter at a time around the chosen setting, which
shows whether each value is defensible but not how much each contributes to the
total gain. This walks the whole distance instead: from the production
constants to the recommended setting, changing one parameter per step.

```{r trend-path}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  s1 <- trend[trend$stage == 1, ]
  b  <- best_row(s1, "ades")
  s0 <- with_metrics(trend[trend$stage == 0, ], "ades")

  # Each step adopts one recommended value and holds the rest at production's.
  # The order is deliberate: the penalty first, because it is the parameter
  # production does not apply at all, and the geometry after.
  cellv <- function(fl, mg, rd, pn) {
    x <- with_metrics(trend[trend$fl_cap == fl & trend$margin == mg &
                            trend$radius_nm == rd & trend$penalty_nm == pn, ],
                      "ades")
    if (is.null(x) || nrow(x) == 0) NA else x$score[1]
  }
  steps <- list(
    list(lab = "production constants (FL40, margin 4, 30 NM, no penalty)",
         s = s0$score),
    list(lab = sprintf("+ scheduled-service penalty of %g NM", b$penalty_nm),
         s = cellv(40, 4, 30, b$penalty_nm)),
    list(lab = sprintf("+ flight-level cap FL%g", b$fl_cap),
         s = cellv(b$fl_cap, 4, 30, b$penalty_nm)),
    list(lab = sprintf("+ vote margin %g", b$margin),
         s = cellv(b$fl_cap, b$margin, 30, b$penalty_nm)),
    list(lab = sprintf("+ radius %g NM", b$radius_nm),
         s = cellv(b$fl_cap, b$margin, b$radius_nm, b$penalty_nm)))

  ok <- !sapply(steps, function(x) is.na(x$s))
  steps <- steps[ok]
  sc <- sapply(steps, function(x) x$s)
  tab <- data.frame(
    Step = sapply(steps, function(x) x$lab),
    Score = num(sc),
    `Gain from this step` = c("—", sgn(diff(sc))),
    `Share of total gain` = c("—", sprintf("%.0f%%", 100 * diff(sc) / (sc[length(sc)] - sc[1]))),
    check.names = FALSE)
  print(kable(tab, row.names = FALSE, align = "r"))

  dom <- which.max(diff(sc))
  cat(sprintf("\nThe gain is not spread evenly. **%s** accounts for %.0f%% of it on its own. That is worth stating plainly: the single most valuable change to arrival detection is not a tuned threshold at all, but applying to `trend` the scheduled-service tie-break that `endpoint` already had and `trend` never did.\n",
              sub("^\\+ ", "", steps[[dom + 1]]$lab),
              100 * max(diff(sc)) / (sc[length(sc)] - sc[1])))
  cat(sprintf("\nBy contrast the radius change is worth %s flights out of %s — well inside what one three-day sample can resolve. It is reported as *not motivated*: the sweep's argmax prefers it, but the evidence does not justify moving a published constant for it.\n",
              num(sc[length(sc)] - sc[length(sc) - 1]), num(sc[length(sc)])))
}
```

## The verdict {#sec-verdict}

```{r verdict}
#| echo: false
#| results: asis
if (!is.null(modes)) {
  m <- function(run, field) {
    v <- modes[[field]][modes$run == run]
    if (length(v) == 0) NA_real_ else v[1]
  }
  tot <- function(run) m(run, "adep_score") + m(run, "ades_score")
  runs <- modes$run[order(-(modes$adep_score + modes$ades_score))]

  print(kable(data.frame(
    Configuration = modes$run,
    ADEP = paste0(modes$adep_mode, ""), ADES = paste0(modes$ades_mode, ""),
    `ADEP coverage` = pct(modes$adep_coverage),
    `ADEP accuracy` = pct(modes$adep_accuracy),
    `ADEP score` = num(modes$adep_score),
    `ADES coverage` = pct(modes$ades_coverage),
    `ADES accuracy` = pct(modes$ades_accuracy),
    `ADES score` = num(modes$ades_score),
    Total = num(modes$adep_score + modes$ades_score),
    check.names = FALSE)[order(-(modes$adep_score + modes$ades_score)), ],
    row.names = FALSE, align = "r"))

  cat(sprintf("\nThe recommended configuration scores **%s against production's %s, a gain of %s** on a three-day sample of %s flights. Every figure in that table is a real pipeline run scored against Network Manager ground truth.\n",
              num(tot("recommended")), num(tot("legacy")),
              sgn(tot("recommended") - tot("legacy")), num(N_GT)))
}
```

### What is being recommended, and why

```{r verdict-detail}
#| echo: false
#| results: asis
if (!is.null(modes) && !is.null(grid) && !is.null(ppath)) {
  m <- function(run, field) modes[[field]][modes$run == run][1]
  cat(sprintf("**Departures — `endpoint`, %g NM, %s ft, %g NM scheduled-service penalty.** Score %s against production's %s, a gain of %s. This is the clearest result in the study: the setting is an interior optimum in both parameters, the method beats `trend` at every exchange rate tried, and the pipeline reproduces the sweep exactly, so the tuned figure is the production figure.\n\n",
              best_row(grid, "adep")$radius_nm,
              format(best_row(grid, "adep")$height_ft, big.mark = ",", trim = TRUE, scientific = FALSE),
              best_row(grid, "adep")$penalty_nm,
              num(m("recommended", "adep_score")), num(m("legacy", "adep_score")),
              sgn(m("recommended", "adep_score") - m("legacy", "adep_score"))))

  alone <- ppath$ades_score[ppath$run == "path1_penalty"]
  base  <- ppath$ades_score[ppath$run == "path0_legacy"]
  cat(sprintf("**Arrivals — `trend` at the production geometry, with the scheduled-service penalty added.** Measured on its own, that is %s against production's %s, a gain of %s. None of the other tuned arrival values is adopted, because none of them survives the pipeline's ranking rule.\n\n",
              num(alone), num(base), sgn(alone - base)))

  cat(sprintf("Reported in the combined flight list the arrival half comes out at %s, which is %s against production rather than the %s just quoted. The difference is not the algorithm. Taking departures from `endpoint` adds tracks that `trend` alone never produced — the merged list holds %s rows against %s — and where two tracks share an aircraft, callsign and date, the scoring alignment awards the flight to whichever starts closest to the actual off-block time. Sometimes that is now an endpoint-only track carrying no arrival. The merge itself loses nothing: both tables name the same %s arrivals, and they disagree on none of them.\n\n",
              num(m("recommended", "ades_score")),
              sgn(m("recommended", "ades_score") - m("legacy", "ades_score")),
              sgn(alone - base),
              num(cache("merge_shapes_v6.csv")$rows[1]),
              num(cache("merge_shapes_v6.csv")$rows[2]),
              num(cache("merge_agreement_v6.csv")$agree[1])))

  cat("That distinction is worth being exact about, because it decides whether the arrival figure is a cost or an artefact. The two tables carry **the same arrival column**: every track id that has an ADES in one has the same ADES in the other, and there are no disagreements in either direction. So the arrival movements OPDI would publish are identical — an analyst counting arrivals at any aerodrome gets the same number from either table. What changes is only which track this benchmark pairs with each reference flight when several are eligible.\n\n")

  cat(sprintf("So the recommendation gains %s on the departure side and leaves the arrival side materially unchanged: %s from the penalty, measured like for like, and no change at all to the arrivals actually named. The %s reported against the combined list is the benchmark's alignment, not the algorithm's output, and it is shown rather than suppressed because the same artefact appears in the per-aerodrome table below.\n",
              sgn(m("recommended", "adep_score") - m("legacy", "adep_score")),
              sgn(alone - base),
              sgn(m("recommended", "ades_score") - m("legacy", "ades_score"))))
}
```

```{r verdict-penalty-check}
#| echo: false
#| results: asis
ppa <- cache("per_airport_path_v6.csv")
if (!is.null(ppa)) {
  cat("\nThe penalty was checked separately at aerodrome level, where a re-ranking rule could easily rob one aerodrome to pay another. Across the hundred busiest it does not:\n\n")
  rows <- do.call(rbind, lapply(c("departures", "arrivals"), function(rl) {
    b <- ppa[ppa$run == "path0_legacy"  & ppa$role == rl & ppa$airport != "OOA", ]
    p <- ppa[ppa$run == "path1_penalty" & ppa$role == rl & ppa$airport != "OOA", ]
    b <- b[order(-b$n_truth), ][1:min(100, nrow(b)), ]
    p <- p[match(b$airport, p$airport), ]
    data.frame(Role = tools::toTitleCase(rl),
               `Recall without penalty` = pct(sum(b$n_correct) / sum(b$n_truth)),
               `Recall with penalty` = pct(sum(p$n_correct) / sum(p$n_truth)),
               Improved = sum(p$recall > b$recall, na.rm = TRUE),
               Worsened = sum(p$recall < b$recall, na.rm = TRUE),
               Unchanged = sum(p$recall == b$recall, na.rm = TRUE),
               check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
  cat("\nNo aerodrome in the hundred is made worse by it. That is the signature of a tie-break that is breaking ties in the right direction rather than trading one error for another.\n")
}
```

### The `traffic` library's rule, for reference

```{r verdict-traffic}
#| echo: false
#| results: asis
if (!is.null(modes)) {
  m <- function(run, field) modes[[field]][modes$run == run][1]
  cat(sprintf("`nearest` — the unconditional rule the `traffic` library uses — answers far more often than anything else here (%s of departures, %s of arrivals) and is right far less often (%s and %s). Scored, it is last by a wide margin: %s against the recommendation's %s.\n",
              pct(m("nearest", "adep_coverage")), pct(m("nearest", "ades_coverage")),
              pct(m("nearest", "adep_accuracy")), pct(m("nearest", "ades_accuracy")),
              num(m("nearest", "adep_score") + m("nearest", "ades_score")),
              num(m("recommended", "adep_score") + m("recommended", "ades_score"))))
  cat("\nThat is not a criticism of `traffic`, which is solving a different problem: it names the nearest aerodrome and leaves the judgement to the caller. It is, however, the measurement that justifies OPDI carrying its own rule rather than adopting that one.\n")
}
```

## By aerodrome class {#sec-per-type}

The candidate set is large and medium aerodromes only, so a flight whose true
aerodrome is a small airfield or a heliport cannot be named correctly by any
method on this page. This is the size of that population, and of everything
else.

```{r per-type}
#| echo: false
#| results: asis
if (!is.null(typ)) {
  for (rl in c("departures", "arrivals")) {
    x <- typ[typ$run == "recommended" & typ$role == rl, ]
    if (nrow(x) == 0) next
    x <- x[order(-x$n_truth), ]
    cat(sprintf("\n**%s**\n\n", tools::toTitleCase(rl)))
    print(kable(data.frame(
      Class = x$apt_type, `Flights in truth` = num(x$n_truth),
      Aerodromes = num(x$n_aerodromes), Answered = num(x$n_answered),
      Correct = num(x$n_correct), Coverage = pct(x$coverage),
      Accuracy = pct(x$accuracy), check.names = FALSE),
      row.names = FALSE, align = "r"))
  }
  unn <- typ[typ$run == "recommended" &
             !typ$apt_type %in% c("large_airport", "medium_airport"), ]
  if (nrow(unn)) {
    cat(sprintf("\nOutside the candidate set — %s classes, %s flights across both roles — the method is working against a reference it was never given. Those flights are counted in the denominator throughout this page, so every coverage figure here already carries their cost.\n",
                num(length(unique(unn$apt_type))), num(sum(unn$n_truth))))
  }
}
```

## The busiest 100 aerodromes {#sec-top100}

Movement counts are what OPDI is for, so the last check is per aerodrome rather
than per flight: for each of the hundred busiest in the reference, how many
movements the recommended methodology finds, against how many the Network
Manager data says there were, and how the algorithm production runs today
compares on the same aerodromes.

`Count ratio` is movements found over movements in truth — the number a
capacity or delay analysis would be distorted by. `Recall` is the share of that
aerodrome's true movements correctly attributed to it.

```{r top100}
#| echo: false
#| results: asis
if (!is.null(apt)) {
  for (rl in c("departures", "arrivals")) {
    r <- apt[apt$run == "recommended" & apt$role == rl & apt$airport != "OOA", ]
    l <- apt[apt$run == "legacy"      & apt$role == rl & apt$airport != "OOA", ]
    if (nrow(r) == 0) next
    r <- r[order(-r$n_truth), ][1:min(100, nrow(r)), ]
    l <- l[match(r$airport, l$airport), ]
    cat(sprintf("\n### %s\n\n", tools::toTitleCase(rl)))
    print(kable(data.frame(
      `#` = seq_len(nrow(r)),
      Aerodrome = r$airport,
      `Movements (truth)` = num(r$n_truth),
      `Found` = num(r$n_predicted_here),
      `Correct` = num(r$n_correct),
      `Count ratio` = sprintf("%.3f", r$count_ratio),
      `Recall` = pct(r$recall),
      `Recall, production` = pct(l$recall),
      `Δ recall` = pp(r$recall - l$recall),
      check.names = FALSE), row.names = FALSE, align = "r"))
  }
}
```

```{r top100-summary}
#| echo: false
#| results: asis
if (!is.null(apt)) {
  cat("\n### Across the hundred\n\n")
  rows <- do.call(rbind, lapply(c("departures", "arrivals"), function(rl) {
    r <- apt[apt$run == "recommended" & apt$role == rl & apt$airport != "OOA", ]
    l <- apt[apt$run == "legacy"      & apt$role == rl & apt$airport != "OOA", ]
    r <- r[order(-r$n_truth), ][1:min(100, nrow(r)), ]
    l <- l[match(r$airport, l$airport), ]
    data.frame(
      Role = tools::toTitleCase(rl),
      `Movements in truth` = num(sum(r$n_truth)),
      `Recall, recommended` = pct(sum(r$n_correct) / sum(r$n_truth)),
      `Recall, production`  = pct(sum(l$n_correct, na.rm = TRUE) / sum(l$n_truth, na.rm = TRUE)),
      `Aerodromes improved` = sprintf("%d of %d", sum(r$recall > l$recall, na.rm = TRUE), nrow(r)),
      `Median count ratio`  = sprintf("%.3f", median(r$count_ratio, na.rm = TRUE)),
      check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
  cat("\nA count ratio below one means the aerodrome is under-counted — the usual case, since every abstention is a movement not attributed. Above one means movements are being attributed to an aerodrome that did not see them, which is the more damaging error and the one the scoring penalises at rate $k$.\n")
  cat("\nThe departure row is the result: recall rises by nearly nine points and the median aerodrome is within about one percent of its true movement count.\n")
  cat("\n**The arrival row must be read with @sec-verdict in hand.** It falls, and on almost every aerodrome, which looks like a broad regression and is not one. The arrival column of the recommended flight list is identical to production's, track for track — the same aerodromes named for the same flights. What changes is that the merged list also contains departure-only tracks, and this table pairs reference flights with tracks by start time, so some arrivals are now counted against a track that carries none. The measurement moved; the data did not. It is left in the table rather than corrected away, because a benchmark that quietly adjusts its own alignment to flatter a result is worth less than one that shows where it is fragile.\n")
}
```

## Does any of it hold on a second period? {#sec-stability}

Everything above is one three-day sample. An argmax over hundreds of cells on a
single sample always flatters itself: part of the winning cell's margin is that
sample's noise, and noise does not reproduce. The only way to find out which
part is real is to measure a different period with the same code.

The second period is **5–7 June 2024**, chosen because the Network Manager
reference for that month is committed to the repository and the matching tracks
already exist from earlier versions of this study.

```{r stability}
#| echo: false
#| results: asis
t24 <- cache("trend_sweep_2024.csv")
if (is.null(t24)) {
  cat("*The second-period sweep has not been staged yet.*\n")
} else {
  cat(sprintf("The 2024 sample carries %s ground-truth flights against 2025's %s.\n\n",
              num(t24$n_ground_truth[1]), num(N_GT)))

  s1_25 <- trend[trend$stage == 1, ]; s1_24 <- t24[t24$stage == 1, ]
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(role) {
    b25 <- best_row(s1_25, role); b24 <- best_row(s1_24, role)
    do.call(rbind, lapply(list(
      list("trend FL cap", "fl_cap"), list("trend vote margin", "margin"),
      list("trend radius (NM)", "radius_nm")), function(pp_) {
        data.frame(Role = toupper(role), Parameter = pp_[[1]],
                   `2025 optimum` = b25[[pp_[[2]]]],
                   `2024 optimum` = b24[[pp_[[2]]]],
                   Agrees = ifelse(b25[[pp_[[2]]]] == b24[[pp_[[2]]]], "yes", "**no**"),
                   check.names = FALSE)
      }))
  }))
  print(kable(rows, row.names = FALSE, align = "r"))

  agree <- sum(rows$Agrees == "yes")
  cat(sprintf("\n%d of %d parameter optima agree across the two periods.\n",
              agree, nrow(rows)))

  # The question that actually matters is not whether the argmax moved but
  # whether the *decision* would have changed -- so the rejected values are
  # re-checked on the second period too.
  b24 <- best_row(s1_24, "ades"); b25 <- best_row(s1_25, "ades")
  leg24 <- with_metrics(t24[t24$stage == 0, ], "ades")
  cat(sprintf("\nOn the arrival side, where this study rejected the sweep's preferences, the 2024 sweep prefers FL%g against 2025's FL%g — and its own gain over the production constants is %s, against %s on 2025. Both are harness figures, and @sec-pipeline showed harness gains of this size on `trend` do not survive the pipeline's ranking. A parameter that is neither stable across periods nor reproducible in production is not one to ship, which is what the defaults reflect.\n",
              b24$fl_cap, b25$fl_cap,
              sgn(b24$score - leg24$score),
              sgn(b25$score - with_metrics(trend[trend$stage == 0, ], "ades")$score)))
}
```

```{r stability-endpoint}
#| echo: false
#| results: asis
e24 <- cache("sweep_radius_height_2024.csv")
if (is.null(e24)) {
  cat("\n**The endpoint rule was not re-swept on the second period.** Its sweep runs over the cached candidate table, which exists for 2025 and not for 2024; building it for 2024 would mean either materialising an H3 index over the whole second period or reimplementing the pipeline's endpoint extraction in the harness. The second option is precisely the kind of shortcut @sec-pipeline exists to warn about, so neither was taken.\n")
  cat("\nThe departure recommendation therefore rests on a different guarantee, and a stronger one for this particular purpose: it was validated *in the pipeline itself* on 2025, where the sweep and `process_dai` agree to the flight. What is missing is a second sample, not a second implementation. It is listed in the limitations as the outstanding check.\n")
} else if (!is.null(grid)) {
  cat("\nFor the endpoint rule, which is what departures actually adopt:\n\n")
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(role) {
    a <- best_row(grid, role); b <- best_row(e24, role)
    data.frame(Role = toupper(role),
               `2025 radius / height` = sprintf("%g NM / %s ft", a$radius_nm,
                    format(a$height_ft, big.mark = ",", trim = TRUE, scientific = FALSE)),
               `2024 radius / height` = sprintf("%g NM / %s ft", b$radius_nm,
                    format(b$height_ft, big.mark = ",", trim = TRUE, scientific = FALSE)),
               Agrees = ifelse(a$radius_nm == b$radius_nm & a$height_ft == b$height_ft,
                               "yes", "**no**"),
               check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
}
```

## What ships {#sec-defaults}

The pipeline's `DetectionConfig` defaults are now the values this study
supports. Two were changed, three were deliberately left where production had
them, and the difference between those two groups is the whole argument of the
page: a value is adopted when a **pipeline** run confirms it, not when a sweep
prefers it.

| Parameter | Production | Now ships | Verdict |
|---|---|---|---|
| `trend_sched_penalty_nm` | 0 | **10** | **Changed.** The only tuned trend value the pipeline confirms: +234 score, improves 9 of the hundred busiest aerodromes, worsens none. |
| `endpoint_radius_nm` | 40 | **30** | **Changed.** Interior optimum, and the pipeline reproduces the sweep exactly. |
| `endpoint_height_ft` | 15,000 | 15,000 | **Confirmed.** The sweep's argmax lands on the value already in use. |
| `trend_max_fl` | 40 | 40 | **Rejected.** The sweep preferred FL120 by a wide margin; through the pipeline that setting lost 1,741. |
| `trend_radius_nm` | 30 | 30 | **Rejected.** The argmax at 20 NM was worth 15 flights out of 61,038 on one sample. |
| `trend_vote_margin` | 4 | 4 | **Rejected.** Its gain was measured on top of a flight-level cap that is not being adopted. |

Every OPDI flight list published before this change was built with the
production column. `DetectionConfig.legacy()` still returns exactly those
values, so anything already released stays reproducible by name — the same
principle as the frozen `version` strings on published events. A regression
test pins the preset, and a second pins the two rejected values, so restoring
an optimum on sweep evidence alone has to argue with a test rather than with a
comment.

## Limitations {#sec-limitations}

### Parameters that were not swept

Five parameters were held fixed by decision rather than by evidence. Naming
them matters: a reader who sees seven tuned values and no caveat will
reasonably assume the whole configuration was optimised, and it was not.

| Parameter | Where | Held at | Why not swept |
|---|---|---|---|
| Smoothing half-window | `flights.py` | 2 samples | Parameterised in config so it is visible and changeable, but a smoothing width interacts with the sample rate rather than the geometry, and the decimation study settled the sample rate separately. |
| Classification dead band | `flights.py` | none exists | There is no dead band to sweep. The vote margin already plays that role. |
| Out-of-area border margin | `flights.py` | 30 NM | Fixes what counts as "outside the observed area". Changing it redefines the ground truth rather than the method, so a sweep would move the target as well as the aim. |
| Aerodrome type set | `h3_airport_zones.py` | large + medium | The candidate set. Widening it is a data-volume decision about the zone table, not a detection threshold. |
| H3 resolution | `config.py` | 7 (±0.76 NM) | Changing it invalidates every cached zone table and the `distance_from_center` ring counts that production selects on. |

Two of these carry a cost that this study measures rather than waves at, since
"not swept" is not the same as "no consequence".

```{r unswept-cost}
#| echo: false
#| results: asis
if (!is.null(typ) && !is.null(modes)) {
  out <- typ[typ$run == "recommended" &
             !typ$apt_type %in% c("large_airport", "medium_airport", "out of area"), ]
  ooa <- typ[typ$run == "recommended" & typ$apt_type == "out of area", ]
  m <- function(run, f) modes[[f]][modes$run == run][1]

  # Aggregated over roles: the frame carries one row per (class, role), and
  # listing them separately reads as four classes rather than two.
  byclass <- tapply(out$n_truth, out$apt_type, sum)
  cat(sprintf("**The type set.** %s flights across both roles have a true aerodrome outside the large/medium candidate set (%s). No method on this page can name them, whatever its thresholds, because the aerodrome is not among the candidates. They are counted in every denominator here, so each coverage figure already carries their cost: they are worth about %s of headline coverage.\n\n",
              num(sum(out$n_truth)),
              paste(sprintf("%s at %ss", num(byclass), gsub("_", " ", names(byclass))),
                    collapse = ", "),
              pct(sum(out$n_truth) / (2 * N_GT))))

  cat(sprintf("**The border margin.** Fixed at 30 NM, it decides which flights are labelled out-of-area. %s of ground-truth departures genuinely begin outside the observed area, and the recommended configuration recognises %s of them — at %s precision, so the label is trustworthy when it appears but is rarely reached. For arrivals the recall is %s: `trend` emits no out-of-area label at all, which is a structural property of the vote rule rather than a tuning choice.\n\n",
              pct(m("recommended", "adep_gt_ooa_share")),
              pct(m("recommended", "adep_ooa_recall")),
              pct(m("recommended", "adep_ooa_precision")),
              pct(m("recommended", "ades_ooa_recall"))))

  cat(sprintf("Both costs fall on coverage rather than accuracy, which is the benign direction: with the free out-of-area wins removed, in-area accuracy is %s for departures and %s for arrivals.\n",
              pct(m("recommended", "adep_inarea_accuracy")),
              pct(m("recommended", "ades_inarea_accuracy"))))
}
```

### The outstanding check

The endpoint radius and height — the values that actually change what the
pipeline ships for departures — have been validated against the pipeline but on
**one** period. `trend`, whose tuned values were mostly rejected, is the one
measured on two. That is the wrong way round, and it is the first thing a
follow-up should fix: build the endpoint candidate cache for a second period
and re-run the radius x height grid against it.

Two things limit how much that weakens the recommendation. The departure gain
is large — thousands of flights, not tens — so it is not the kind of margin a
single sample's noise produces. And the optimum sits on a plateau in both
parameters, so even a moderate shift in the argmax would cost little.

### One sample, and what that does and does not license

The parameters are chosen as the argmax over a grid evaluated on a single
three-day period. An argmax over hundreds of cells on one sample will always
overstate the gain slightly: some of the winning cell's margin is the sample's
noise, and it will not reproduce.

That is the reason for the second period, and for reporting the shape of each
optimum rather than only its location. A parameter sitting on a broad plateau
is a safe recommendation even from one sample; one sitting on a sharp isolated
peak is not, however good its score.

<!-- Remaining: the two-period stability table, the parameter justification
     table, the verdict, the top-100 comparison, and the pipeline-fidelity
     check. -->
