---
title: "Detecting departure and destination aerodromes from ADS-B"
subtitle: "Version 6 — the consolidated study, and what to run in production"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every figure and table on this page comes from a committed CSV
# under data/. Nothing here reads S3, a database or a Spark cluster.
#
# Every one of those CSVs was produced by *this* study, against the pipeline as
# it stands at the end of the parameterisation work. No number is carried over
# from versions 1-5 or from the decimation study. That is not bookkeeping: the
# bearing results in version 5 pre-date a modulo fix that changed them, the
# arrival comparison pre-dates a corrected `endpoint` baseline, and the trend
# numbers came from a standalone reimplementation. Re-measuring everything is
# what allows the arithmetic on this page to agree with itself.
cache <- function(f) {
  p <- file.path("data", f)
  if (!file.exists(p)) return(NULL)
  read.csv(p, stringsAsFactors = FALSE)
}

trend  <- cache("trend_sweep_2025.csv")
grid   <- cache("sweep_radius_height_2025.csv")
pen    <- cache("sweep_penalty_2025.csv")
cone   <- cache("sweep_cone_2025.csv")
modes  <- cache("mode_comparison_v6.csv")
apt    <- cache("per_airport_v6.csv")
typ    <- cache("per_type_v6.csv")

pct  <- function(x) ifelse(is.na(x), "—", sprintf("%.2f%%", 100 * x))
pp   <- function(x) ifelse(is.na(x), "—", sprintf("%+.2f pp", 100 * x))
num  <- function(x) format(round(x), big.mark = ",", trim = TRUE)
# The "+" flag does nothing for %s in R, so the sign is prepended explicitly --
# a signed delta that silently loses its sign reads as a level, not a change.
sgn  <- function(x) paste0(ifelse(x >= 0, "+", "−"),
                           format(abs(round(x)), big.mark = ",", trim = TRUE))

# --- counts, and the score -------------------------------------------------
# The sweep frames carry coverage and overall as ratios of the same ground
# truth. The counts are exactly recoverable from them -- coverage * n is the
# number answered, overall * n the number correct, and both are integers -- so
# every table below can show counts without a separate export. Verified against
# the trend sweep, which carries both: all rows agree.
N_GT <- if (!is.null(trend)) trend$n_ground_truth[1] else grid$n_ground_truth[1]

K_DEFAULT <- 2

answered <- function(d, role) round(d[[paste0(role, "_coverage")]] * d$n_ground_truth)
correct  <- function(d, role) round(d[[paste0(role, "_overall")]]  * d$n_ground_truth)
wrong    <- function(d, role) answered(d, role) - correct(d, role)
scr      <- function(d, role, k = K_DEFAULT) correct(d, role) - k * wrong(d, role)

# Attach the five reported quantities to any sweep frame, for either role, so
# no table has to recompute them and none can disagree with another.
with_metrics <- function(d, role, k = K_DEFAULT) {
  if (is.null(d)) return(NULL)
  d$correct  <- correct(d, role)
  d$wrong    <- wrong(d, role)
  d$score    <- d$correct - k * d$wrong
  d$coverage <- d[[paste0(role, "_coverage")]]
  d$accuracy <- d[[paste0(role, "_accuracy")]]
  d
}

# The best row of a sweep for one role, at a given k.
best_row <- function(d, role, k = K_DEFAULT, subset = TRUE) {
  x <- with_metrics(d[subset, , drop = FALSE], role, k)
  if (is.null(x) || nrow(x) == 0) return(NULL)
  x[which.max(x$score), ]
}
```

## Summary {.unnumbered}

OPDI derives a flight's departure and destination aerodrome from ADS-B state
vectors alone. This page consolidates six versions of that work into one
account: what the three candidate methods are, how each behaves across its
whole parameter space, which settings are best for departures and for arrivals,
what the ideas that did *not* work were, and what the pipeline should ship.

The short version is that the two roles want different algorithms, and that the
algorithm already in production was never tuned. Its flight-level gate — the
altitude below which state vectors are allowed to vote on whether a track is
climbing or descending — was set at FL40 and never swept. Moving it is worth
thousands of correctly named flights per three-day sample, at a cost of tens of
wrong ones.

::: {.callout-note}
## Reading this page
Every number below is an inline expression over a CSV in `data/`. Nothing is
typed into the prose. Where this page disagrees with versions 1–5, it says so
and gives the reason; those versions stay published unchanged.
:::

## How the metrics work {#sec-metrics}

Five quantities are reported for every setting, for each role separately. Four
of them are unavoidable and the fifth is a judgement, so it is worth being
precise about all five before any of them are used to argue for a parameter.

Write $N$ for the number of ground-truth flights in the sample, $a$ for the
number the method *answers* (names an aerodrome, or explicitly says
out-of-area), $c$ for the number it answers **correctly**, and $w = a - c$ for
the number it answers **wrongly**.

**Coverage** is $a/N$: the share of flights the method is willing to commit on.
A method that abstains everywhere has coverage zero and cannot be wrong.

**Accuracy** is $c/a$: of the flights it committed on, the share it got right.
Accuracy says nothing about how often the method declined to answer, so a
method can buy accuracy simply by abstaining more.

**Correct** ($c$) and **wrong** ($w$) are the raw counts. They matter because
coverage and accuracy generally move in opposite directions — every extra
answer a method gives is one it was previously unsure about, so raising
coverage tends to lower accuracy — and neither ratio says whether a particular
exchange was worth making. The counts do.

**Score** is $c - k\,w$. It exists to make the exchange rate explicit: a change
to a parameter is worth taking if and only if it wins more than $k$ correct
answers for each wrong one it introduces.

The value of $k$ is where judgement enters. This study uses $k = 2$ by default,
on the following argument. A **silence** — an abstention — corrupts one
movement count, and it is *visible*: it arrives downstream as an explicit null
that any consumer can filter on. A **wrong aerodrome** corrupts *two* movement
counts, the one it invents and the one it omits, and it is *invisible*: nothing
downstream can distinguish it from a correct answer. Two errors for one, plus
undetectability, is the case for $k = 2$.

That is an argument, not a measurement, and rankings do move with $k$. So every
headline choice on this page is also scored at $k = 1$ and $k = 3$, and where
the ranking changes, the page says so.

### Out-of-area, and why the headline numbers are decomposed

Roughly one flight in six in the sample genuinely begins or ends outside the
observed area. Such a flight has no nameable aerodrome, and the correct answer
is the literal label `OOA`. This creates a way to score well for a bad reason:
a method that labels liberally out-of-area collects those flights for free.

So each side is also reported three further ways — out-of-area recall and
precision, which show whether the free wins are being taken honestly, and
in-area coverage and accuracy, which restrict to flights whose truth is a real
aerodrome and are the numbers to read when comparing naming rules.

```{r metrics-table}
#| echo: false
if (!is.null(grid)) {
  kable(data.frame(
    Quantity = c("Coverage", "Accuracy", "Correct", "Wrong", "Score",
                 "OOA recall", "OOA precision",
                 "In-area coverage", "In-area accuracy"),
    Definition = c("$a/N$", "$c/a$", "$c$", "$w = a - c$", "$c - k\\,w$",
                   "OOA hits / OOA in truth", "OOA hits / OOA predicted",
                   "answered / truth is a real aerodrome",
                   "correct / answered, truth a real aerodrome"),
    `Reads as` = c("willingness to commit", "reliability when committed",
                   "flights named right", "flights named wrong",
                   "net value at exchange rate $k$",
                   "does it recognise what it cannot name?",
                   "is it over-using the OOA label?",
                   "commitment with the free wins removed",
                   "reliability with the free wins removed"),
    check.names = FALSE
  ), caption = "The reported quantities.")
}
```

## The sample {#sec-sample}

```{r sample}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  cat(sprintf(
    "The 2025 period is 5-7 June 2025, three consecutive days, giving **%s ground-truth flights** after alignment with the Network Manager reference.\n",
    num(N_GT)))
}
```

Ground truth is the EUROCONTROL Network Manager flight table, joined to ADS-B
tracks on aircraft address, callsign and date. The join key is
`AIRCRAFT_ADDRESS`, which is the `icao24` of the ADS-B message.

## The three methods {#sec-methods}

### `trend` — what production runs today

For every state vector below a flight-level cap and within a radius of an
aerodrome, `trend` asks whether smoothed barometric altitude is rising or
falling. Each vector casts a vote. If climbing votes exceed descending votes by
more than a margin, that aerodrome is the departure; if the reverse, the
destination. The nearest surviving candidate wins.

Two of its four constants had never been swept before this study: the
flight-level cap, fixed at FL40, and the detection radius, fixed at 30 NM.

### `nearest` — the *traffic* library's rule

The reference implementation in the `traffic` library takes the aerodrome
nearest the first and last point of the trajectory, unconditionally. It never
abstains, so its coverage is total and its accuracy is whatever the geometry
gives. It is included as the external baseline.

### `endpoint` — `nearest`, with abstention

`endpoint` is `nearest` plus two gates: the endpoint must lie within a radius
of the aerodrome, and either on the ground or below a height above the field.
Failing both, the method abstains rather than guessing. A scheduled-service
penalty breaks ties toward aerodromes with commercial traffic.

## Trend: the full parameter sweep {#sec-trend-sweep}

```{r trend-sweep-summary}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  s0 <- trend[trend$stage == 0, ]
  cat("### The untuned baseline\n\n")
  for (role in c("adep", "ades")) {
    x <- with_metrics(s0, role)
    cat(sprintf(
      "- **%s** at the production setting (FL%g, margin %g, %g NM, no penalty): coverage %s, accuracy %s, %s correct, %s wrong, score %s.\n",
      toupper(role), s0$fl_cap[1], s0$margin[1], s0$radius_nm[1],
      pct(x$coverage), pct(x$accuracy), num(x$correct), num(x$wrong), num(x$score)))
  }
}
```

```{r trend-best}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  cat("\n### The best setting found\n\n")
  for (role in c("adep", "ades")) {
    b <- best_row(trend, role)
    s0 <- with_metrics(trend[trend$stage == 0, ], role)
    cat(sprintf(
      "- **%s**: FL%g, margin %g, %g NM, penalty %g NM — coverage %s, accuracy %s, %s correct, %s wrong, score %s (%s against the untuned baseline).\n",
      toupper(role), b$fl_cap, b$margin, b$radius_nm, b$penalty_nm,
      pct(b$coverage), pct(b$accuracy), num(b$correct), num(b$wrong),
      num(b$score), sgn(b$score - s0$score)))
  }
}
```

<!-- Sections below are completed as the remaining measurements land:
     the full sweep tables, the endpoint sweeps, the penalty treatment across
     both methods, the consolidated negative results, the parameter
     justification table, the two-period stability table, the verdict, the
     top-100 comparison and the limitations. -->
