---
title: "Detecting departure and destination aerodromes from ADS-B"
subtitle: "A methodology comparison against EUROCONTROL APDF ground truth"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every figure and table on this page comes from a committed CSV
# cache under data/. Refresh it from the opdi repository with
#   python benchmarks/export_results.py
# which is the only step that touches S3.
cache <- function(f) read.csv(file.path("data", f), stringsAsFactors = FALSE)

methods   <- cache("method_comparison.csv")
sweep     <- cache("abstention_sweep.csv")
attrib    <- cache("cascade_attribution.csv")

pct <- function(x) sprintf("%.2f%%", 100 * x)
```

## Summary {.unnumbered}

The OPDI flight list assigns each flight a departure aerodrome (ADEP) and a
destination aerodrome (ADES) from ADS-B alone. The production algorithm does
this by testing the sign of a smoothed altitude trend near each candidate
aerodrome and discarding every flight whose trend is ambiguous. It is accurate
where it speaks — `r pct(methods$adep_accuracy[methods$method == "M0_current"])`
on ADEP — but it stays silent on
`r pct(1 - methods$adep_coverage[methods$method == "M0_current"])` of flights.

We compared seven methodologies against EUROCONTROL APDF ground truth. Three
findings, in increasing order of consequence:

1. **Coverage is the problem, not accuracy.** Every trend-based method scores
   above 96% where it fires. None of them covers more than 58% of flights.

2. **The vertical-trend machinery earns nothing.** Attributing a cascade's
   answers to the rung that supplied them shows that the simplest possible
   rule — *the nearest aerodrome to the track's first and last sample* — is at
   least as accurate as every trend-based method **on that method's own
   flights**. A five-stage cascade built on those methods collapses exactly
   onto the one-line rule.

3. **What the trend methods actually supply is abstention, badly.** Replacing
   them with an explicit test on the endpoint — is it close to an aerodrome,
   and is it low over it? — beats the production algorithm on coverage and
   accuracy simultaneously:
   `r pct(methods$adep_coverage[methods$method == "M7_abstain"])` of flights at
   `r pct(methods$adep_accuracy[methods$method == "M7_abstain"])` accuracy,
   against `r pct(methods$adep_coverage[methods$method == "M0_current"])` at
   `r pct(methods$adep_accuracy[methods$method == "M0_current"])`.

::: {.callout-important}
## Scope of the results on this page

Every figure below is computed on a **single day, 2025-06-05**
(`r format(methods$n_ground_truth[1], big.mark = ",")` ground-truth flights,
35.4 million state vectors). Three-day samples in 2025-06 and 2024-06 are being
scored and will replace these numbers.

The sample is small in *days*, not in flights: three days is roughly 97,000
ground-truth flights, which is ample for fitting the two thresholds this study
ends up recommending. The real exposure is representativeness — day of week,
weather, seasonal traffic mix — which is why the second sample is a year apart
rather than adjacent. Read the conclusions as what these days support.
:::

## Why this is hard

An ADS-B trajectory does not carry its origin or destination. Both must be
inferred from where the aircraft was and what it was doing, and the inference
fails in a specific, asymmetric way: reception near the ground is patchy.
A receiver 50 NM from a regional aerodrome sees an aircraft in the cruise
perfectly and never sees it below 2,000 ft. So the evidence that would settle
the question — surface movement, the take-off roll, the flare — is exactly the
evidence most often missing.

Three consequences shape everything below:

- **Tracks frequently begin or end in mid-air.** A track that starts in the
  cruise has no departure evidence at all, and any method that names an
  aerodrome for it is guessing.
- **Departure and destination are not symmetric.** Departure evidence is a
  climb from a known point; destination evidence is a descent to one. Coverage
  differs between the two, and so does accuracy.
- **Nearby aerodromes compete.** Within 30 NM of a major hub there are often
  several licensed aerodromes and, if heliports are included, dozens.

## Data

### Trajectories

State vectors are read from the OpenSky Network historical archive, filtered to
the OPDI European bounding box and decimated to 5 s — byte-for-byte the same
filter the production ingestion applies, so the benchmark evaluates the same
population the pipeline ingests.

Tracks are built with OPDI's own track-splitting algorithm, imported and called
rather than reimplemented: `track_id` continuity with published data depends on
it and it is frozen.

### Aircraft selection

Only aeroplanes are in scope. Helicopters in particular defeat every
aerodrome-based method — they depart from and arrive at places that are not in
any aerodrome database — and they are not a rounding error: they are 9.8% of
classified airframes in the OpenSky aircraft database.

Filtering uses the ICAO Doc 8643 aircraft class, keeping `L` (landplane) and
`A` (amphibian) and dropping `H` (helicopter), `G` (gyrocopter) and `T`
(tiltrotor). The field is present on ~73% of the database, so *unknown*
handling is a real trade-off: the results here **exclude** unknowns, which
protects precision at some cost to coverage. Two further layers — the ADS-B
emitter category, which is the only field naming surface vehicles, and a
behavioural filter requiring a sustained airborne period — are described under
[Limitations](#limitations) and not yet applied.

### Ground truth

APDF `flights_tidy` provides ADEP, ADES and off-block time per flight.
`AIRCRAFT_ADDRESS` is the ICAO 24-bit address, which is the join key to ADS-B;
flights are matched on `(icao24, callsign, date)`. 2.42% of those keys collide —
same-day multi-leg rotations — and are resolved by proximity to the actual
off-block time rather than left to chance.

## How methods are scored

Ground truth is the denominator throughout. For a set of real flights $G$ and a
method that emits an answer on the subset $A \subseteq G$:

- **Coverage** $= |A| / |G|$ — the fraction of real flights the method answers
  at all. A flight whose trajectory was never seen counts as a miss.
- **Accuracy** $= |\text{correct}| / |A|$ — how often it is right *when it
  speaks*.
- **Overall** $= |\text{correct}| / |G|$ — coverage times accuracy; the fraction
  of real flights correctly labelled.

Reporting all three matters, because the production algorithm and its
alternatives sit at opposite ends of the same trade: one is nearly always right
and usually silent, the other nearly always speaks and is often wrong. Neither
"accuracy" nor "coverage" alone can distinguish them.

::: {.callout-note}
## The coverage ceiling

No method exceeds ~73% coverage, because that is roughly the share of APDF
flights with a usable ADS-B track in the sample at all. The remaining ~27% is a
data-availability limit, not an algorithmic one, and no change to the naming
rule can recover it.
:::

## The candidate methods

| | Method | Signal |
|---|---|---|
| **M0** | *Current production* | Sign of a 5-sample rolling-mean altitude difference near each candidate aerodrome, requiring a margin of 4; ambiguous flights dropped |
| **M1** | *traffic library* | Nearest aerodrome to the track's first sample (ADEP) and last sample (ADES). No altitude, no trend, no ground state |
| **M2** | *On-ground* | Aerodrome at the surface samples bracketing the airborne phase |
| **M3** | *Vertical rate* | Sign of the measured `vert_rate` near each candidate aerodrome |
| **M5** | *Lowest and nearest* | Aerodrome where the track is lowest and closest, split at the track midpoint |
| **M6** | *Cascade* | Take the most precise available answer, falling through when a rung is silent |
| **M7** | *Endpoint with abstention* | M1's naming rule, but silent unless the endpoint is close to the aerodrome and low over it |

M1 recreates `TakeoffAirportInference` / `LandingAirportInference` from the
`traffic` library, which sorts by timestamp and calls `guess_airport` on the
first and last row. M2 and M3 both use columns the production algorithm already
selects and then never reads: `on_ground` and `vert_rate`.

Aerodromes come from OurAirports, restricted to large and medium airports.
Including heliports is a swept parameter rather than a detail — `traffic`'s own
documentation shows `infer_airport` returning a helipad instead of Amsterdam
Schiphol when they are left in.

## Results

```{r results-table}
m <- methods[order(-methods$adep_overall), ]
kable(
  data.frame(
    Method       = m$method,
    `ADEP cov`   = pct(m$adep_coverage),
    `ADEP acc`   = pct(m$adep_accuracy),
    `ADEP overall` = pct(m$adep_overall),
    `ADES cov`   = pct(m$ades_coverage),
    `ADES acc`   = pct(m$ades_accuracy),
    `ADES overall` = pct(m$ades_overall),
    check.names  = FALSE
  ),
  align = "lrrrrrr",
  caption = "Coverage, accuracy and overall correctness by method, 2025-06-05, sorted by ADEP overall correctness."
)
```

The shape of the table is the first finding. The production algorithm M0 is the
**second most accurate** method tested and the **fifth** by overall
correctness, because it declines to answer nearly half the flights. `traffic`'s
one-line nearest-aerodrome rule, with no aviation logic in it whatsoever, gets
more flights right in absolute terms than any trend-based method.

That is not an argument for adopting M1 as it stands — an 87.5% ADEP accuracy
would be a serious regression in data quality for OPDI, which publishes a
flight list that downstream users treat as authoritative. It is an argument
that the trend logic is not where the value is.

## The cascade, and why it fails {#cascade}

The natural response to the table above is a cascade: take M0's answer when it
speaks, fall through to M3, then M2, and use M1 only as a last resort. The
strong signals keep their precision and M1 supplies coverage where they are
silent.

It does not work, and the way it fails is instructive.

A cascade's accuracy is *not* bounded below by its rungs' accuracies, because
each rung is scored only on the flights the rungs above it could not answer,
which are the harder ones. So a cascade legitimately scoring below every method
it is built from is not by itself evidence of a bug. What it must not do is
score below its own fallback rung used alone — that means some rung is actively
overwriting better answers.

To test that, every cascade answer is attributed to the rung that supplied it,
and that rung is scored against a control on **the same flights**:

```{r attribution-table}
a <- attrib[attrib$side == "adep", ]
a <- a[order(a$ladder_pos), ]
kable(
  data.frame(
    Rung         = a$rung,
    `Flights answered` = format(a$n_answered, big.mark = ","),
    `Share of all` = pct(a$share_of_gt),
    `Its accuracy` = pct(a$accuracy),
    `M1 on the same flights` = pct(a$fallback_accuracy),
    Lift         = sprintf("%+.2f%%", 100 * a$lift),
    check.names  = FALSE
  ),
  align = "lrrrrr",
  caption = "ADEP attribution for the precision-ordered cascade M0 -> M3 -> M2 -> M1 -> M5, with M1 as the control. A negative lift means the rung overwrote a better answer."
)
```

The lift is negative in every row. On the 17,751 flights M0 answers, M0 is
`r pct(attrib$accuracy[attrib$rung == "m0" & attrib$side == "adep"])` accurate
and M1 is
`r pct(attrib$fallback_accuracy[attrib$rung == "m0" & attrib$side == "adep"])`.
On M3's flights the gap is 8 points; on M2's, 39.

So the trend methods do not look precise because they reason well about
trajectories. They look precise because they *fire on easy flights* — those
with unambiguous low-altitude coverage near a single aerodrome — and on exactly
those flights, "which aerodrome is nearest the endpoint" is already the right
answer, slightly more often than the trend vote is. Where the trend vote
differs, it is usually because a neighbouring aerodrome produced a stronger
vote than the correct one.

Running the same cascade with M1 **first** confirms it: the remaining four
rungs together contribute 564 additional flights, of which **none** are
correct, and the cascade's score becomes indistinguishable from M1 alone.

::: {.callout-tip}
## The reframing

The five methods differ in *when they are willing to answer*, not in *what they
answer*. Their apparent precision is an abstention policy wearing a naming
rule's clothes — and an expensive one, since it discards half the flights to
buy accuracy it does not deliver.

So the design question is not which signal should name the aerodrome. It is
when to stay silent. And that is a property of the endpoint, not of the
trajectory's shape.
:::

## Abstention on the endpoint

M7 keeps M1's naming rule and adds an explicit two-parameter test. An ADEP is
emitted only if the track's first sample is within $d$ NM of the nearest
aerodrome and no more than $h$ ft above that aerodrome's elevation; ADES uses
the last sample. Height is measured above field elevation rather than above the
ellipsoid, because a fixed 1,000 ft cut-off means nothing at an aerodrome
sitting at 5,000 ft. A sample flagged `on_ground` satisfies the height test
outright, since a missing barometric altitude on the surface is normal.

```{r frontier-plot}
#| fig-cap: "Coverage/accuracy frontier for endpoint abstention, ADEP, 2025-06-05. Each line is one distance threshold; points along it are height thresholds from on-ground only (bottom left) to unrestricted (top right). The production algorithm M0 and the unrestricted endpoint rule M1 are marked. The region above and to the right of M0 is where a method beats it on both axes at once."
#| fig-width: 8
#| fig-height: 5.2

s <- sweep[sweep$max_endpoint_dist_nm < 1e8, ]
s$dist <- factor(
  s$max_endpoint_dist_nm,
  levels = sort(unique(s$max_endpoint_dist_nm)),
  labels = paste0(sort(unique(s$max_endpoint_dist_nm)), " NM")
)

ref <- methods[methods$method %in% c("M0_current", "M1_traffic"), ]
ref$label <- ifelse(ref$method == "M0_current", "M0 production", "M1 no abstention")

ggplot(s, aes(adep_coverage, adep_accuracy, colour = dist, group = dist)) +
  geom_hline(
    yintercept = methods$adep_accuracy[methods$method == "M0_current"],
    linetype = "dotted", colour = "grey45"
  ) +
  geom_vline(
    xintercept = methods$adep_coverage[methods$method == "M0_current"],
    linetype = "dotted", colour = "grey45"
  ) +
  geom_line(linewidth = 0.6) +
  geom_point(size = 1.7) +
  geom_point(data = ref, aes(adep_coverage, adep_accuracy),
             inherit.aes = FALSE, size = 3.2, shape = 18) +
  # Plain geom_text rather than ggrepel: the portal's render environment is
  # fixed and this page must not add a package to it for two labels.
  geom_text(
    data = ref, aes(adep_coverage, adep_accuracy, label = label),
    inherit.aes = FALSE, size = 3.4, hjust = 1.1, vjust = -0.9
  ) +
  scale_x_continuous("ADEP coverage", labels = scales::percent) +
  scale_y_continuous("ADEP accuracy", labels = scales::percent) +
  scale_colour_viridis_d("Endpoint within", end = 0.9) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank())
```

Two things are visible immediately.

**The height threshold does nearly all the work.** Moving the radius from 10 NM
to 40 NM changes ADEP coverage by 0.2 percentage points. Moving the height cut
from on-ground-only to 5,000 ft AGL moves it by 34 points. This makes physical
sense: an aircraft below 5,000 ft AGL is in the terminal area of the aerodrome
it is nearest to, and being 30 NM out at that height does not make it ambiguous.
It also means the radius parameter, which is where the production algorithm
spends its tuning effort, is close to irrelevant.

**There is a region that dominates the production algorithm outright.** The
dotted lines mark M0's operating point. Every point up and to the right of
their intersection beats it on coverage *and* accuracy at the same time:

```{r operating-points}
sel <- sweep[
  sweep$max_endpoint_dist_nm %in% c(5, 10, 20, 40) &
    sweep$max_endpoint_agl_ft %in% c(1000, 2000, 5000), ]
sel <- sel[order(sel$max_endpoint_agl_ft, sel$max_endpoint_dist_nm), ]
kable(
  data.frame(
    `Within`     = paste0(sel$max_endpoint_dist_nm, " NM"),
    `Below`      = paste0(format(sel$max_endpoint_agl_ft, big.mark = ","), " ft AGL"),
    `ADEP cov`   = pct(sel$adep_coverage),
    `ADEP acc`   = pct(sel$adep_accuracy),
    `ADES cov`   = pct(sel$ades_coverage),
    `ADES acc`   = pct(sel$ades_accuracy),
    check.names  = FALSE
  ),
  align = "llrrrr",
  caption = "Selected operating points. Compare against M0 production: 54.78% / 98.25% for ADEP and 52.66% / 96.68% for ADES."
)
```

At 10 NM and 5,000 ft AGL the rule answers
`r pct(methods$adep_coverage[methods$method == "M7_abstain"])` of ADEPs at
`r pct(methods$adep_accuracy[methods$method == "M7_abstain"])` accuracy — four
points more coverage than production, at higher accuracy. Tightening to 2 NM
and 1,000 ft trades coverage down to 43% for 99.69% accuracy, which is
essentially error-free labelling. The parameters give OPDI a dial where it
currently has a fixed point, and every position on the dial between 1,000 and
5,000 ft is above 99% accurate.

ADES is consistently weaker than ADEP at the same thresholds — about 6 points
less coverage at 5,000 ft AGL. Arrivals are harder: a descent can be
interrupted by a go-around or a diversion, and the last recorded sample of a
track is more often a loss of reception than a landing.

## Recommendation

On the evidence of one day, OPDI should replace the altitude-trend vote with
endpoint proximity plus explicit abstention. This is a smaller algorithm than
the one it replaces — no smoothing window, no margin parameter, no per-aerodrome
vote — and it is strictly better on both axes.

The two parameters should be published alongside the flight list, and the
abstention reason retained per flight, so that downstream users can distinguish
"no departure aerodrome" from "departure aerodrome not determinable". The
current algorithm collapses those two cases into a null, which is a loss of
information that costs nothing to preserve.

This should not be implemented until the full-month results are in.

## Limitations {#limitations}

- **One day, becoming three plus three.** Everything here rests on 2025-06-05.
  Three-day samples in 2025-06 and 2024-06 are being scored; fitting the
  thresholds on one and validating on the other, a year apart, is the only
  honest way to report a tuned parameter. Full-month sampling was considered
  and dropped: it costs ~12 hours of cluster time and ~100 GB to reduce a
  sampling error that is already small relative to the effect sizes here — the
  gap between M7 and the production algorithm is four percentage points of
  coverage and one of accuracy, not a fraction of a point.
- **No trajectory cleaning.** The state vectors are raw apart from the bounding
  box and decimation. OPDI's cleaning stage has not been applied, and whether
  it helps here is an open question rather than a formality: the stale-broadcast
  filter nulls repeated positions, which is precisely what a stationary aircraft
  at a gate transmits. It could remove the very samples the endpoint rule
  depends on. This will be measured as an ablation, not assumed.
- **Aircraft filtering is one layer of three.** The ADS-B emitter category —
  the only field that names surface vehicles — is dropped by the reference
  pipeline stage and would need to be retained. A behavioural filter requiring
  a sustained airborne period is database-independent and would catch the 27%
  of airframes with no recorded class at all.
- **Per-aerodrome counts are not yet reported.** Comparing arrival and
  departure counts per aerodrome against APDF is a stricter test than
  flight-level accuracy and is only meaningful at the 90 aerodromes APDF
  covers. It is not in this version.
- **Aerodrome set is fixed.** All results use large and medium airports.
  Sweeping small airports and heliports is implemented but not yet reported.

## Reproducing

Everything is in the `opdi` repository under `benchmarks/`, and every dataset
written is registered in `benchmarks/DATASETS.md`.

```bash
# 1. Sample state vectors from the OpenSky archive into S3
python benchmarks/osn_sample.py 2025-06-05 2025-06-08

# 2. Score every method
python benchmarks/adep_ades.py --days 2025-06-05 --months 202506 --methods all

# 3. Attribute the cascade's answers, with M1 as the control
python benchmarks/adep_ades.py --days 2025-06-05 --months 202506 \
    --methods M6_cascade --diagnose-cascade --control m1

# 4. Sweep the abstention grid
python benchmarks/adep_ades.py --days 2025-06-05 --months 202506 \
    --methods none --sweep-abstain

# 5. Refresh this page's CSV cache
python benchmarks/export_results.py
```

Steps 1–4 run on the OpenSky Kubernetes cluster and need credentials. Step 5
needs S3 read access. Rendering this page needs neither.
