---
title: "Thinning state vectors, and what the archive actually contains"
subtitle: "Why the 5 s decimation rule is not the problem, and what is"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every number comes from a committed CSV under data/, produced by
# benchmarks/decimation_experiment.py in the opdi repository.
cache <- function(f) {
  p <- file.path("data", f)
  if (!file.exists(p)) return(NULL)
  read.csv(p, stringsAsFactors = FALSE)
}
tim  <- cache("timing.csv")
alt  <- cache("by_altitude.csv")
den  <- cache("by_density.csv")
age  <- cache("age_by_altitude.csv")
ends <- cache("endpoint_age.csv")
lag  <- cache("endpoint_lag.csv")
arm  <- cache("arm_comparison.csv")
armt <- cache("arm_per_type.csv")

pct <- function(x) ifelse(is.na(x), "—", sprintf("%.2f%%", 100 * x))
num <- function(x) format(x, big.mark = ",", trim = TRUE)
secs <- function(x) sprintf("%.0f s", x)

r <- function(rule, f) tim[[f]][tim$rule == rule]
N_BASE <- r("bbox_only", "rows"); N_MOD <- r("modulo", "rows"); N_BUC <- r("bucket", "rows")
T_BASE <- r("bbox_only", "seconds"); T_MOD <- r("modulo", "seconds"); T_BUC <- r("bucket", "seconds")
EXTRA <- N_BUC - N_MOD

e <- function(which, f) ends[[f]][ends$endpoint == paste(which, "row of aircraft")]
HOURS <- 3
```

## Summary {.unnumbered}

The ingest thins OpenSky state vectors to roughly one sample every 5 s with a
modulo filter, `event_time % 5 == 0`. That is a **fixed-phase** sampler: it
keeps a row only if one exists at the single second per window congruent to
zero. Where reception is intermittent — low altitude, long range, the edge of
the receiver network — the phase of the data is arbitrary, so the filter should
delete far more than it is supposed to. Since version 4 of the
[ADEP/ADES study](../adep-ades-detection-v4/index.qmd) found arrivals much
harder than departures, and arrivals are exactly the flights whose evidence
sits in that sparse region, the filter was a plausible cause.

**It is not.** Measured on three hours of raw archive, a correct bin-based
sampler keeps `r sprintf("%.3f", N_BUC / N_MOD)`× the rows of the modulo filter
— `r num(EXTRA)` rows out of `r num(N_MOD)`, or `r pct(EXTRA / N_MOD)`. The
hypothesis fails because its premise does: the OpenSky `state_vectors` table is
not a record of receptions but a **complete 1 Hz grid**, with the last known
position carried forward when nothing is heard.

The experiment then found what the row counts were hiding. Positions in that
grid are often old, and the staleness is wildly asymmetric between the two ends
of a track:

| | median position age |
|---|---|
| **first** row of an aircraft | **`r secs(e("first","p50_s"))`** |
| **last** row of an aircraft | **`r secs(e("last","p50_s"))`** |

A departure's first fix is fresh by construction. An arrival's last fix, at the
median, carries a position measured **`r secs(e("last","p50_s"))` earlier** —
at approach speed, five to eight nautical miles back down the approach.
`r pct(e("last","share_gt30s"))` of last fixes are more than 30 s stale.

That is not a bug to fix in the decimation rule. It is a measurement of how
early reception ends, and it quantifies the asymmetry version 4 could only
assert. It also suggests something the detection can act on, because
`last_pos_update` is already ingested and read by nothing.

## What was measured

Three hourly partitions of `s3a://opensky-hdfs-backup/tables_v4/state_vectors`,
2025-06-05 06:00–09:00 UTC, filtered to the OPDI bounding box and nothing else:
**`r num(N_BASE)` rows**. Both rules were then applied to identical input, with
a cold read from S3 each time so the comparison includes the I/O the production
job actually pays.

**Modulo (current).** Keep the row where `event_time % 5 == 0`. A narrow filter;
no shuffle.

**Bucket (proposed).** Bin time into `floor(t/5)` and keep one row per
non-empty bin, via `max` over a struct with `event_time` first — a hash
aggregate with map-side partial aggregation, native Spark throughout.

## Result 1: the two rules are indistinguishable

```{r timing-table}
kable(data.frame(
  Rule = c("bbox only (baseline)", "modulo (current)", "bucket (proposed)"),
  Rows = num(c(N_BASE, N_MOD, N_BUC)),
  `Wall clock` = sprintf("%.1f s", c(T_BASE, T_MOD, T_BUC)),
  `Net of the read` = c("—", sprintf("%+.1f s", T_MOD - T_BASE), sprintf("%+.1f s", T_BUC - T_BASE)),
  check.names = FALSE), align = "lrrr",
  caption = sprintf("Both rules on %s rows, three hours of archive. The S3 read is common to both, so the difference net of it is the cost of the shuffle.", num(N_BASE)))
```

The bucket rule keeps `r sprintf("%.3f", N_BUC / N_MOD)`× the rows. The modulo
filter retains `r sprintf("1/%.3f", N_BASE / N_MOD)` of its input — within a
whisker of the exactly one-in-five it would keep if every second carried a row.

The density measurement confirms why:

```{r density-table}
d <- den[order(den$pb), ]
kable(data.frame(
  `Rows per aircraft-minute` = substring(d$pb, 4),
  `Aircraft-minutes` = num(d$aircraft_minutes),
  `Share` = pct(d$aircraft_minutes / sum(d$aircraft_minutes)),
  `Mean density` = sprintf("%.3f", d$mean_p),
  check.names = FALSE), row.names = FALSE, align = "lrrr",
  caption = "Fraction of each minute's 60 seconds carrying a row, per aircraft-minute. The archive is a dense grid, not a log of receptions.")
```

`r pct(den$aircraft_minutes[den$pb == "6. 0.9-1.0 (dense)"] / sum(den$aircraft_minutes))`
of aircraft-minutes are at least 90% populated. The sparse-coverage model that
motivated this experiment applies to a few percent of the data.

::: {.callout-important}
## The arithmetic was right; the premise was wrong

The predicted gain was 2–5×, derived from a model in which a second carries a
row with probability *p* < 1. Measured *p* is ≈ 1.0. Everything downstream of
that assumption — retention ratios, endpoint-lag formulae, loss probabilities —
described a feed that does not exist. The model was not tested before it was
believed; this experiment is what testing it looks like.
:::

## Result 2: the effect exists, in the right place, at the wrong size

The extra rows are not distributed evenly. They concentrate exactly where the
hypothesis said they would.

```{r altitude-table}
a <- alt[order(alt$b), ]
a$extra <- a$bucket_rows - a$modulo_rows
a$rate <- a$extra / a$modulo_rows
a$share_rows <- a$modulo_rows / sum(a$modulo_rows)
a$share_extra <- a$extra / sum(a$extra)
kable(data.frame(
  `Band (ft)` = substring(a$b, 4),
  `Rows` = num(a$modulo_rows),
  `Extra` = num(a$extra),
  `Rescue rate` = pct(a$rate),
  `% of rows` = pct(a$share_rows),
  `% of extra` = pct(a$share_extra),
  `Over-rep` = sprintf("%.2fx", a$share_extra / a$share_rows),
  check.names = FALSE), row.names = FALSE, align = "lrrrrrr",
  caption = "Rows the bucket rule rescues, by altitude. *Rescue rate* is extra rows over retained rows; *over-rep* is the share of rescued rows against the share of all rows, so above 1 means they concentrate there.")
```

The rescue rate runs monotonically from
**`r pct(a$rate[a$b == "1. 0-2,000"])` at 0–2,000 ft** down to
**`r pct(a$rate[a$b == "5. 20,000-30,000"])` at 20–30,000 ft** — a
`r sprintf("%.1f", a$rate[a$b == "1. 0-2,000"] / a$rate[a$b == "5. 20,000-30,000"])`×
gradient from the surface to cruise. The lowest band is
`r sprintf("%.2f", (a$share_extra / a$share_rows)[a$b == "1. 0-2,000"])`× over-represented
among the rescued rows and the null-altitude band — largely surface traffic —
`r sprintf("%.2f", (a$share_extra / a$share_rows)[a$b == "0. (null)"])`×. Every
band above 5,000 ft is at or below parity.

So the mechanism is real and it is altitude-dependent in the predicted
direction. It is also **`r pct(EXTRA / N_MOD)` of the data**, which cannot
explain an eleven-point coverage gap. Direction confirmed, magnitude
irrelevant.

The endpoint check agrees: of `r num(lag$aircraft)` aircraft, modulo erased
`r num(lag$erased_by_modulo)` (`r pct(lag$erased_share)`) entirely, and its last
retained row sits a median of `r secs(lag$lag_p50_s)` behind the true last row —
about a quarter of a nautical mile.

**On this evidence the decimation rule should not be changed.** It costs
`r sprintf("%+.0f%%", 100 * (T_BUC - T_MOD) / T_MOD)` in wall clock, it would
break `track_id` continuity with every published dataset, and it buys
`r pct(EXTRA / N_MOD)` more rows.

## Result 3: what the row counts were hiding

The archive fills every second, but it fills them by **carrying the last known
position forward**. `lastPosUpdate` records when the position was actually
measured, so `age = event_time − lastPosUpdate` is how old the position in a
row is. A row can be present and carry nothing new.

```{r age-table}
# The band label lost its sort prefix when the measurement was parsed out of
# the run log, so order explicitly rather than alphabetically.
BAND_ORDER <- c("(null)", "0-2,000", "2,000-5,000", "5,000-10,000",
                "10,000-20,000", "20,000-30,000", "30,000+")
g <- age[match(BAND_ORDER, age$band), ]
kable(data.frame(
  `Band (ft)` = g$band,
  Rows = num(g$n),
  `Median age` = secs(g$p50_s),
  `p90` = secs(g$p90_s),
  `p99` = secs(g$p99_s),
  `> 5 s` = pct(g$share_gt5s),
  `> 30 s` = pct(g$share_gt30s),
  check.names = FALSE), row.names = FALSE, align = "lrrrrrr",
  caption = "Age of the position carried by each row, by altitude. The mean is not shown: the distribution is bimodal -- mostly fresh, with a heavy stale tail -- and the mean describes neither part.")
```

At cruise, positions are fresh: `r pct(g$share_gt30s[g$band == "20,000-30,000"])`
of rows above 20,000 ft are more than 30 s old. **Below 2,000 ft that rises to
`r pct(g$share_gt30s[g$band == "0-2,000"])`, and on the surface to
`r pct(g$share_gt30s[g$band == "(null)"])`.** A third of all low-altitude rows
carry a position measured more than half a minute earlier.

This is the user-visible form of intermittent low-altitude coverage. It does not
appear as missing rows because OpenSky fills the gap for us.

### The asymmetry between the two ends of a track

```{r ends-table}
kable(data.frame(
  Endpoint = c("first row of an aircraft", "last row of an aircraft"),
  N = num(c(e("first","n"), e("last","n"))),
  `Median age` = secs(c(e("first","p50_s"), e("last","p50_s"))),
  `p90` = secs(c(e("first","p90_s"), e("last","p90_s"))),
  `p99` = secs(c(e("first","p99_s"), e("last","p99_s"))),
  `> 30 s` = pct(c(e("first","share_gt30s"), e("last","share_gt30s"))),
  check.names = FALSE), row.names = FALSE, align = "lrrrrr",
  caption = "Position age at the two ends of each aircraft's presence in the window. This is the sharpest asymmetry in the study.")
```

**A departure's first fix is fresh by construction** — an aircraft enters the
table when it is first heard, so there is no earlier position to carry forward.
Median age `r secs(e("first","p50_s"))`,
`r pct(e("first","share_gt30s"))` older than 30 s.

**An arrival's last fix is stale by construction** — the aircraft stays in the
table after it stops being heard, padded with its last known position until it
times out. Median age **`r secs(e("last","p50_s"))`**, and
**`r pct(e("last","share_gt30s"))`** older than 30 s.

At 150 kt on final approach, `r secs(e("last","p50_s"))` is about 8 NM.

::: {.callout-warning}
## What this does and does not mean

It does **not** mean the endpoint position is wrong. The carried-forward
position is the last position genuinely measured, which is still the best
available estimate of where the track ended. Deduplicating the stale rows would
not move it.

What it means is that **reception ends a median of `r secs(e("last","p50_s"))`
before the aircraft leaves the data**, and the aircraft keeps flying through
that interval. Version 4 asserted that an arrival's last sample is "where
reception died rather than where the flight ended"; this puts a number on it,
and the number is minutes rather than seconds.

It also means three things downstream are measurably wrong, none of them
detection: `last_seen` event times are late by the same interval; gap-based
track splitting sees padded gaps rather than real ones; and any derivative
computed across carried-forward rows is diluted by runs of zero change.
:::

## Result 4: the end-to-end test agrees

Rows are a proxy. What matters is whether detection changes, so the whole chain
was re-run on bucket-sampled input — ingest, tracks, endpoint candidates — and
scored against the same ground truth at the same operating point. One day,
2025-06-05, both arms; `endpoint` mode at 30 NM, 15,000 ft and a 10 NM
scheduled-service penalty, applied to **both** roles.

```{r arm-setup}
#| include: false
a <- function(k, f) arm[[f]][arm$arm == k]
NG <- a("modulo (published)", "n_ground_truth")
cnt <- function(k, role, m) round(a(k, paste0(role, "_", m)) * NG)
armrow <- function(role, label) {
  ans_m <- cnt("modulo (published)", role, "coverage"); ans_b <- cnt("bucket (proposed)", role, "coverage")
  cor_m <- cnt("modulo (published)", role, "overall");  cor_b <- cnt("bucket (proposed)", role, "overall")
  data.frame(
    Movements = label,
    `Answered` = sprintf("%s -> %s (%+d)", num(ans_m), num(ans_b), ans_b - ans_m),
    `Correct`  = sprintf("%s -> %s (%+d)", num(cor_m), num(cor_b), cor_b - cor_m),
    `Wrong`    = sprintf("%s -> %s (%+d)", num(ans_m - cor_m), num(ans_b - cor_b),
                         (ans_b - cor_b) - (ans_m - cor_m)),
    check.names = FALSE)
}
```

```{r arm-rates}
mk <- function(role, label) data.frame(
  Movements = label,
  Coverage = sprintf("%s -> %s", pct(a("modulo (published)", paste0(role,"_coverage"))),
                     pct(a("bucket (proposed)", paste0(role,"_coverage")))),
  `d cov` = sprintf("%+.2f pp", 100 * (a("bucket (proposed)", paste0(role,"_coverage")) -
                                       a("modulo (published)", paste0(role,"_coverage")))),
  Accuracy = sprintf("%s -> %s", pct(a("modulo (published)", paste0(role,"_accuracy"))),
                     pct(a("bucket (proposed)", paste0(role,"_accuracy")))),
  `d acc` = sprintf("%+.2f pp", 100 * (a("bucket (proposed)", paste0(role,"_accuracy")) -
                                       a("modulo (published)", paste0(role,"_accuracy")))),
  Overall = sprintf("%s -> %s", pct(a("modulo (published)", paste0(role,"_overall"))),
                    pct(a("bucket (proposed)", paste0(role,"_overall")))),
  `d ovr` = sprintf("%+.2f pp", 100 * (a("bucket (proposed)", paste0(role,"_overall")) -
                                       a("modulo (published)", paste0(role,"_overall")))),
  check.names = FALSE)
kable(rbind(mk("adep","departures"), mk("ades","arrivals")), row.names = FALSE,
      align = "lllllll",
      caption = sprintf("Detection on modulo- against bucket-sampled input, %s ground-truth flights on 2025-06-05. `endpoint` at 30 NM / 15,000 ft / 10 NM, both roles. Each cell is modulo then bucket.", num(NG)))
```

```{r arm-counts}
kable(rbind(armrow("adep","departures"), armrow("ades","arrivals")), row.names = FALSE,
      align = "llll",
      caption = "The same comparison in flights.")
```

The candidate cache grew from `r num(a("modulo (published)","candidates"))` to
`r num(a("bucket (proposed)","candidates"))` rows,
`r sprintf("%+.2f%%", 100 * (a("bucket (proposed)","candidates") / a("modulo (published)","candidates") - 1))`
— about twice the `r pct(EXTRA / N_MOD)` gain in the state vectors themselves.
That is the one place the rescued rows are worth more than average rows: they
sit at track boundaries, and boundaries are exactly what the candidate stage
extracts.

**Departures do not move**: `r sprintf("%+d", cnt("bucket (proposed)","adep","overall") - cnt("modulo (published)","adep","overall"))`
correct labels, which is noise. **Arrivals move slightly and in the predicted
direction**: `r sprintf("%+d", cnt("bucket (proposed)","ades","coverage") - cnt("modulo (published)","ades","coverage"))`
answered and `r sprintf("%+d", cnt("bucket (proposed)","ades","overall") - cnt("modulo (published)","ades","overall"))`
correct, for
`r sprintf("%+d", (cnt("bucket (proposed)","ades","coverage") - cnt("bucket (proposed)","ades","overall")) - (cnt("modulo (published)","ades","coverage") - cnt("modulo (published)","ades","overall")))`
wrong. That is `r sprintf("%+.2f pp", 100 * (a("bucket (proposed)","ades_overall") - a("modulo (published)","ades_overall")))`
of arrival overall correctness.

The asymmetry is the theory working: the bucket rule keeps the *last* row of
each bin, so a track's final observation survives exactly, while the modulo rule
lands a median of `r secs(lag$lag_p50_s)` short of it. Departures have nothing
to gain from that — their first fix is already well located — and arrivals have
a little.

**A little is the operative word.** Fifty-five additional correct arrivals in a
day, against thirty additional errors, for a change that rewrites `track_id` for
every published row. The end-to-end test confirms the row-count prediction
rather than overturning it.

## What to do with this

**Do not change the decimation rule.** `r pct(EXTRA / N_MOD)` of rows, at
`r sprintf("%+.0f%%", 100 * (T_BUC - T_MOD) / T_MOD)` wall clock and the loss of
`track_id` continuity, is not a trade worth making.

**Use position age as an abstention signal.** This is the actionable finding.
`last_pos_update` is already ingested, already carried into `osn_tracks`
(`tracks.py:455`), and read by nothing. A track whose final fix has an age of
300 s has an endpoint that is several miles from where the flight actually
ended, and the detection has no way to know it. Age at the endpoint is a direct,
free measure of how much to trust that endpoint — and unlike the height
threshold of the ADEP/ADES version 4 rule, it measures the thing that actually
goes wrong.

That gives a candidate rule worth testing against the version 4 operating point:
abstain when the endpoint's position age exceeds a threshold, instead of, or in
addition to, when its height exceeds one. It is a one-column change to the
candidate cache.

**Consider dropping stale rows at ingest.** Not to improve detection, but for
storage and for honesty: `r pct(g$share_gt30s[g$band == "0-2,000"])` of
low-altitude rows are repeats of a position already recorded. Removing them
would shrink the store and would make gap-based track splitting see the gaps
that are really there. It would also change `track_id`, so it belongs in the
same versioned re-ingest as any other ingest change.

## Limitations

* **Three hours, one day, one bounding box.** 2025-06-05 06:00–09:00 UTC is a
  busy European morning; a night sample or a different region may differ.
* **"Last row of an aircraft" is not "last row of a flight."** The window's end
  truncates aircraft still airborne, and an aircraft leaving the bounding box
  exits with a fresh position. Both dilute the staleness figure, so the true
  landing-endpoint staleness is likely worse than reported, not better.
* **Position age is not position error.** A stale position is exactly right for
  the moment it was measured. The error it causes downstream depends on what the
  aircraft did next, which is unobserved by construction.
* **The bucket rule keeps the last row per bin, not the freshest.** Given the
  measured penalty of under 2 s of staleness this makes no practical difference,
  but a rule optimising freshness would use `min_by(age)`.

## Reproducing

```bash
python benchmarks/decimation_experiment.py \
    --date 2025-06-05 --start-hour 6 --hours 3 --results-dir <dir>
```

Rendering this page needs no credentials, no database and no cluster.
