---
title: "Detecting departure and destination aerodromes from ADS-B"
subtitle: "Version 6.1 — the altitude cut, and which datum it is measured against"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r regenerate}
#| include: false
#| cache: false

# Rendering this page runs the analysis behind it.
#
# `regenerate_v61.py` is the executable definition of the study: every CSV read
# below is produced by exactly one job it declares, with exactly the arguments
# it declares. It re-runs a job when the *code that job depends on* has changed
# since the output was written -- not when the file is old, which says nothing.
#
# Running a stale job needs cluster credentials, because the numbers come from
# Spark over S3 against Network Manager reference data and there is no way to
# recompute them without the data they are computed from. Set OPDI_RENDER=check
# to fail fast on staleness instead, OPDI_RENDER=allow-stale to render a draft
# from whatever is staged -- which the provenance table at the foot of this page
# will then show as unverified -- or OPDI_RENDER=skip to render the prose alone.
REPO   <- normalizePath(file.path(getwd(), "..", "..", "..", "opdi"), mustWork = FALSE)
PYTHON <- Sys.getenv("OPDI_PYTHON", file.path(REPO, ".venv310", "bin", "python"))
MODE   <- Sys.getenv("OPDI_RENDER", "run")

regen <- file.path(REPO, "benchmarks", "regenerate_v61.py")
if (file.exists(regen) && MODE != "skip") {
  flags <- switch(MODE,
                  "run" = character(0),
                  "check" = "--check",
                  "allow-stale" = c("--check", "--allow-stale"),
                  character(0))
  status <- system2(PYTHON, c(shQuote(regen), flags), stdout = "", stderr = "")
  if (!identical(status, 0L)) {
    stop("regenerate_v61.py exited ", status,
         " -- the figures below would not match the code that is checked out. ",
         "Set OPDI_RENDER=allow-stale to render anyway.")
  }
}
```

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

DATA <- file.path(getwd(), "data")

read_if <- function(name) {
  path <- file.path(DATA, name)
  if (file.exists(path)) read.csv(path, stringsAsFactors = FALSE) else NULL
}

# Emit a mermaid diagram, but only into an HTML render.
#
# Not `::: {.content-visible when-format="html"}`. That hides the block from the
# PDF *after* Quarto has already tried to rasterise it, and rasterising mermaid
# needs headless Chromium -- which does not start in this environment. The PDF
# render then waits on it forever: no error, no output, just a stalled process
# and a trail of defunct chrome-headless children.
#
# Writing the fence from knitr means that in a PDF render the mermaid block
# never exists at all, so nothing tries to draw it. Each diagram supplies a
# table alongside, which is what the PDF shows instead -- the same two-form
# arrangement version 7 uses, made robust to a broken Chromium.
mermaid <- function(label, caption, lines) {
  if (!knitr::is_html_output()) return(invisible(NULL))
  cat("```{mermaid}\n")
  cat("%%| label: ", label, "\n", sep = "")
  cat('%%| fig-cap: "', caption, '"\n', sep = "")
  cat(paste(lines, collapse = "\n"), "\n", sep = "")
  cat("```\n")
}
```

## What this version changes

Version 6 tuned `trend`'s altitude cut and left its *datum* alone. This version
changes the datum, and measures what that is worth.

`trend` admitted a state vector to its vote when the aircraft was below a
**flight level** — pressure altitude above the standard datum, the same number
everywhere. `endpoint` had never worked that way: since version 6 its height
gate has been measured **above the aerodrome's own field elevation**, on the
stated grounds that "a fixed altitude cut-off means nothing at an aerodrome
sitting at 5,000 ft". Step 04's ground-phase detection moved to the same datum
for the same reason.

`trend` was the last of the three still referenced to sea level. That is the
asymmetry this version removes.

The consequence is not inaccuracy but **silence**, and it falls unevenly. At
FL60 an aerodrome at sea level offers 6,000 ft of climb and descent for the
vote to read. Madrid, at 1,998 ft, offers about 4,100. Ankara, at 3,125 ft,
about 2,975. The altitude trace has progressively less room to rise or fall
*clearly*, so the method abstains more often — and it abstains in proportion to
how high the aerodrome sits.

::: {.callout-note}
## What this change cannot do

`trend` votes on the **sign** of a smoothed altitude change. Adding a constant
offset to every altitude in a track cannot flip a sign, so the field elevation
changes only *which samples are admitted to vote*, never how a vote is cast.

This is therefore a **coverage** result first. Any movement in accuracy is a
second-order consequence of answering more flights, not of discriminating
better between them. The report is written to keep those apart.
:::

## How a flight gets its aerodromes {#sec-method}

Version 6 describes the three methods and the reasoning behind them; this
section does not repeat that. What it adds is the part version 6 states only in
prose — the **order** the pipeline does things in, and which datum each
altitude gate is measured against — because the change this version makes is a
change to that order.

Both methods answer two questions per track, separately: which aerodrome the
flight departed from (`ADEP`) and which it arrived at (`ADES`). The shipped
configuration takes departures from `endpoint` and arrivals from `trend`.

### The steps

Every box in the diagrams below is one of these.

**S1 — Read the track.** One aircraft's state vectors for one flight, already
split into a track by step 02 and cleaned by step 02a.

**S2 — Smooth the altitude.** A centred five-sample rolling mean over
barometric altitude, computed over the *whole* track before any altitude cut.
Smoothing after the cut would truncate the window exactly where altitude is
changing fastest, which is the worst place to lose half of it.

**S3 — Pre-filter on pressure altitude.** A cheap, deliberately loose cut whose
only job is to keep the aerodrome join affordable. Its ceiling is the
configured cap *plus the highest field elevation in the reference*, so it
cannot discard a sample that the exact test at S6 would have kept.

**S4 — Join the aerodrome zones.** Each sample is matched to candidate
aerodromes through an H3 index at resolution 7, then narrowed to those within
the detection radius by exact great-circle distance.

**S5 — Attach the field elevation.** Each candidate aerodrome's elevation, from
OurAirports. A candidate whose elevation is unknown gets zero, which reduces the
next step to the sea-level behaviour rather than dropping the aerodrome.

**S6 — Cut on altitude.** *This is the step this version changes.* Samples
above the ceiling are excluded from the vote. `trend` now measures that ceiling
**above field elevation**; before this version it was a flight level.

**S7 — Vote.** Each surviving sample votes on the sign of its smoothed altitude
change: rising for a departure, falling for an arrival.

**S8 — Decide, or abstain.** If one direction leads the other by more than the
margin, the track is departing or arriving. Otherwise `trend` says nothing.

**E1 — Take the endpoint fix.** `endpoint` reads one sample, not many: the
track's first fix for a departure, its last for an arrival.

**E2 — Test the gates.** The fix must be within the radius, and either flagged
on the ground or no higher than the height gate **above field elevation**.

**E3 — Rank the survivors.** Nearest wins, on exact great-circle distance, plus
a penalty for aerodromes without scheduled service so that a military or
general-aviation field must be *clearly* nearest rather than merely nearest.

**E4 — Or say where it came from.** Where the gates fail and the fix sits at
the edge of the observed area, `endpoint` reports **out-of-area** rather than
guessing. Precedence is aerodrome first, border second: Ponta Delgada lies
about 8 NM inside the western edge, and the other order labelled its departures
out-of-area with the aircraft still on the runway.

### `trend`, arrivals

```{r fig-trend-arrival-src}
#| echo: false
#| results: asis
mermaid(
  "fig-trend-arrival",
  "`trend` naming an arrival aerodrome (ADES). S6 is the step this version changes.",
  c(
    'flowchart TD',
    '    T[("S1 · track<br/>state vectors")] --> SM["S2 · smooth baro_altitude<br/><i>centred 5-sample mean</i>"]',
    '    SM --> PRE["S3 · pre-filter<br/><i>cap + highest field elevation</i>"]',
    '    PRE --> J{{"S4 · join H3 aerodrome zones<br/><i>res 7, then exact distance</i>"}}',
    '    J --> EL["S5 · attach field elevation<br/><i>OurAirports; unknown &rarr; 0</i>"]',
    '    EL --> CUT["<b>S6 · altitude cut</b><br/><b><i>height above field &le; cap</i></b>"]',
    '    CUT --> V["S7 · vote per sample<br/><i>sign of smoothed change</i>"]',
    '    V --> M{"S8 · descending votes<br/>lead by the margin?"}',
    '    M -->|yes| R["E3 · rank survivors<br/><i>distance + scheduled penalty</i>"]',
    '    M -->|no| U(["undetermined"])',
    '    R --> A(["ADES = nearest survivor"])'
  )
)
```

:::: {.content-visible when-format="pdf"}
| Step | What it does |
|---|---|
| S1 | Read the track's state vectors |
| S2 | Smooth `baro_altitude` over a centred 5-sample window, before any cut |
| S3 | Pre-filter on pressure altitude at the cap plus the highest field elevation |
| S4 | Join the H3 aerodrome zones at resolution 7, then cut on exact distance |
| S5 | Attach the candidate aerodrome's field elevation; unknown becomes zero |
| **S6** | **Cut on height above field elevation** — the step this version changes |
| S7 | Each surviving sample votes on the sign of its smoothed altitude change |
| S8 | If descending votes lead by the margin the track is arriving; else *undetermined* |
| E3 | Rank survivors on distance plus the scheduled-service penalty; nearest becomes `ADES` |

::::

### `trend`, departures

The same machinery reading the same samples; only the direction of the winning
vote differs. `trend` has **no out-of-area concept** — it names an aerodrome or
says nothing — which is why the shipped configuration does not use it for
departures.

```{r fig-trend-departure-src}
#| echo: false
#| results: asis
mermaid(
  "fig-trend-departure",
  "`trend` naming a departure aerodrome (ADEP). Identical to the arrival case except at S8.",
  c(
    'flowchart TD',
    '    T[("S1 · track<br/>state vectors")] --> SM["S2 · smooth baro_altitude<br/><i>centred 5-sample mean</i>"]',
    '    SM --> PRE["S3 · pre-filter<br/><i>cap + highest field elevation</i>"]',
    '    PRE --> J{{"S4 · join H3 aerodrome zones<br/><i>res 7, then exact distance</i>"}}',
    '    J --> EL["S5 · attach field elevation<br/><i>OurAirports; unknown &rarr; 0</i>"]',
    '    EL --> CUT["<b>S6 · altitude cut</b><br/><b><i>height above field &le; cap</i></b>"]',
    '    CUT --> V["S7 · vote per sample<br/><i>sign of smoothed change</i>"]',
    '    V --> M{"S8 · climbing votes<br/>lead by the margin?"}',
    '    M -->|yes| R["E3 · rank survivors<br/><i>distance + scheduled penalty</i>"]',
    '    M -->|no| U(["undetermined<br/><i>never out-of-area</i>"])',
    '    R --> A(["ADEP = nearest survivor"])'
  )
)
```

:::: {.content-visible when-format="pdf"}
Identical to the arrival table above, with one difference: at **S8** the track
is a departure when *climbing* votes lead by the margin. Failing that, `trend`
returns *undetermined* — it has no out-of-area answer available to it.

::::

### `endpoint`, departures

```{r fig-endpoint-departure-src}
#| echo: false
#| results: asis
mermaid(
  "fig-endpoint-departure",
  "`endpoint` naming a departure aerodrome (ADEP). One fix, two gates, three possible answers.",
  c(
    'flowchart TD',
    '    T[("S1 · track<br/>state vectors")] --> E1["E1 · take the FIRST fix"]',
    '    E1 --> J{{"S4 · candidate aerodromes<br/><i>within the radius</i>"}}',
    '    J --> EL["S5 · attach field elevation<br/><i>OurAirports</i>"]',
    '    EL --> G{"E2 · within radius AND<br/>(on ground OR<br/><b>height above field &le; gate</b>)?"}',
    '    G -->|yes| R["E3 · rank survivors<br/><i>distance + scheduled penalty</i>"]',
    '    G -->|no| B{"E4 · fix at the edge<br/>of the observed area?"}',
    '    R --> A(["ADEP = nearest survivor"])',
    '    B -->|yes| OOA(["out-of-area"])',
    '    B -->|no| U(["undetermined"])'
  )
)
```

:::: {.content-visible when-format="pdf"}
| Step | What it does |
|---|---|
| S1 | Read the track's state vectors |
| E1 | Take the **first** fix — one sample, not many |
| S4 | Find candidate aerodromes within the radius |
| S5 | Attach each candidate's field elevation |
| **E2** | Gate: within the radius **and** either on the ground or **no higher than the gate above field elevation** |
| E3 | Passing: rank on distance plus the scheduled-service penalty; nearest becomes `ADEP` |
| E4 | Failing, with the fix at the edge of the observed area: **out-of-area**. Otherwise *undetermined* |

Precedence is aerodrome first, border second — the other order labelled Ponta
Delgada's departures out-of-area with the aircraft still on the runway.

::::

### `endpoint`, arrivals

```{r fig-endpoint-arrival-src}
#| echo: false
#| results: asis
mermaid(
  "fig-endpoint-arrival",
  "`endpoint` naming an arrival aerodrome (ADES). The last fix rather than the first.",
  c(
    'flowchart TD',
    '    T[("S1 · track<br/>state vectors")] --> E1["E1 · take the LAST fix"]',
    '    E1 --> J{{"S4 · candidate aerodromes<br/><i>within the radius</i>"}}',
    '    J --> EL["S5 · attach field elevation<br/><i>OurAirports</i>"]',
    '    EL --> G{"E2 · within radius AND<br/>(on ground OR<br/><b>height above field &le; gate</b>)?"}',
    '    G -->|yes| R["E3 · rank survivors<br/><i>distance + scheduled penalty</i>"]',
    '    G -->|no| B{"E4 · fix at the edge<br/>of the observed area?"}',
    '    R --> A(["ADES = nearest survivor"])',
    '    B -->|yes| OOA(["out-of-area"])',
    '    B -->|no| U(["undetermined"])'
  )
)
```

:::: {.content-visible when-format="pdf"}
Identical to the departure table above, with one difference: at **E1** the
method reads the track's **last** fix rather than its first.

::::

### Where the two datums part company

Only **S6** and **E2** read an altitude, and until this version they read it
against different references:

| Test | Threshold | Datum, before v6.1 | Datum, from v6.1 |
|---|---|---|---|
| `trend` S6 | `trend_max_height_ft` | flight level (sea level) | **above field elevation** |
| `endpoint` E2 | `endpoint_height_ft` | above field elevation | above field elevation |
| step 04 ground phase | `phase_ground_ceiling_ft` | above field elevation | above field elevation |
| `nearest` | *none* | — | — |

`nearest` applies no altitude condition at all, on any datum. That is
deliberate: it is the naive `traffic`-faithful baseline the other methods are
scored against, and giving it a gate would defeat its purpose.

::: {.callout-important}
## A flight level is not a height

`FL60` is not the same cut as `6,000 ft`. The pipeline derives the flight level
with an integer cast, so `flight_level <= 60` admits everything below **6,100
ft**. The datum comparison in this report therefore holds the ceiling at 6,100
ft above field, not 6,000 — otherwise it would move the ceiling and the datum
at the same time and could not attribute the result to either.
:::

## Results

```{r results-placeholder}
#| echo: false
#| results: asis
if (is.null(read_if("datum_swap_2025.csv"))) {
  cat("::: {.callout-warning}\n## Not yet measured\n\n",
      "The arms defined in `regenerate_v61.py` have not been run against this ",
      "checkout, so this section is deliberately empty rather than populated ",
      "with numbers carried over from another study.\n:::\n", sep = "")
}
```
