---
title: "Detecting departure and destination aerodromes from ADS-B"
subtitle: "Version 6.1 — the altitude cut, and which datum it is measured against"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r regenerate}
#| include: false
#| cache: false

# Rendering this page runs the analysis behind it.
#
# `regenerate_v61.py` is the executable definition of the study: every CSV read
# below is produced by exactly one job it declares, with exactly the arguments
# it declares. It re-runs a job when the *code that job depends on* has changed
# since the output was written -- not when the file is old, which says nothing.
#
# Running a stale job needs cluster credentials, because the numbers come from
# Spark over S3 against Network Manager reference data and there is no way to
# recompute them without the data they are computed from. Set OPDI_RENDER=check
# to fail fast on staleness instead, OPDI_RENDER=allow-stale to render a draft
# from whatever is staged -- which the provenance table at the foot of this page
# will then show as unverified -- or OPDI_RENDER=skip to render the prose alone.
REPO   <- normalizePath(file.path(getwd(), "..", "..", "..", "opdi"), mustWork = FALSE)
PYTHON <- Sys.getenv("OPDI_PYTHON", file.path(REPO, ".venv310", "bin", "python"))
MODE   <- Sys.getenv("OPDI_RENDER", "run")

regen <- file.path(REPO, "benchmarks", "regenerate_v61.py")
if (file.exists(regen) && MODE != "skip") {
  flags <- switch(MODE,
                  "run" = character(0),
                  "check" = "--check",
                  "allow-stale" = c("--check", "--allow-stale"),
                  character(0))
  status <- system2(PYTHON, c(shQuote(regen), flags), stdout = "", stderr = "")
  if (!identical(status, 0L)) {
    stop("regenerate_v61.py exited ", status,
         " -- the figures below would not match the code that is checked out. ",
         "Set OPDI_RENDER=allow-stale to render anyway.")
  }
}
```

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

DATA <- file.path(getwd(), "data")

read_if <- function(name) {
  path <- file.path(DATA, name)
  if (file.exists(path)) read.csv(path, stringsAsFactors = FALSE) else NULL
}

# Emit a mermaid diagram, but only into an HTML render.
#
# Not `::: {.content-visible when-format="html"}`. That hides the block from the
# PDF *after* Quarto has already tried to rasterise it, and rasterising mermaid
# needs headless Chromium -- which does not start in this environment. The PDF
# render then waits on it forever: no error, no output, just a stalled process
# and a trail of defunct chrome-headless children.
#
# Writing the fence from knitr means that in a PDF render the mermaid block
# never exists at all, so nothing tries to draw it. Each diagram supplies a
# table alongside, which is what the PDF shows instead -- the same two-form
# arrangement version 7 uses, made robust to a broken Chromium.
mermaid <- function(label, caption, lines) {
  if (!knitr::is_html_output()) return(invisible(NULL))
  cat("```{mermaid}\n")
  cat("%%| label: ", label, "\n", sep = "")
  cat('%%| fig-cap: "', caption, '"\n', sep = "")
  cat(paste(lines, collapse = "\n"), "\n", sep = "")
  cat("```\n")
}
```

## What this version changes

Version 6 tuned `trend`'s altitude cut and left its *datum* alone. This version
changes the datum, and measures what that is worth.

`trend` admitted a state vector to its vote when the aircraft was below a
**flight level** — pressure altitude above the standard datum, the same number
everywhere. `endpoint` had never worked that way: since version 6 its height
gate has been measured **above the aerodrome's own field elevation**, on the
stated grounds that "a fixed altitude cut-off means nothing at an aerodrome
sitting at 5,000 ft". Step 04's ground-phase detection moved to the same datum
for the same reason.

`trend` was the last of the three still referenced to sea level. That is the
asymmetry this version removes.

The consequence is not inaccuracy but **silence**, and it falls unevenly. At
FL60 an aerodrome at sea level offers 6,000 ft of climb and descent for the
vote to read. Madrid, at 1,998 ft, offers about 4,100. Ankara, at 3,125 ft,
about 2,975. The altitude trace has progressively less room to rise or fall
*clearly*, so the method abstains more often — and it abstains in proportion to
how high the aerodrome sits.

::: {.callout-note}
## What this change cannot do

`trend` votes on the **sign** of a smoothed altitude change. Adding a constant
offset to every altitude in a track cannot flip a sign, so the field elevation
changes only *which samples are admitted to vote*, never how a vote is cast.

This is therefore a **coverage** result first. Any movement in accuracy is a
second-order consequence of answering more flights, not of discriminating
better between them. The report is written to keep those apart.
:::

## How a flight gets its aerodromes {#sec-method}

Version 6 describes the three methods and the reasoning behind them; this
section does not repeat that. What it adds is the part version 6 states only in
prose — the **order** the pipeline does things in, and which datum each
altitude gate is measured against — because the change this version makes is a
change to that order.

Both methods answer two questions per track, separately: which aerodrome the
flight departed from (`ADEP`) and which it arrived at (`ADES`). The shipped
configuration takes departures from `endpoint` and arrivals from `trend`.

### The steps

Every box in the diagrams below is one of these.

**S1 — Read the track.** One aircraft's state vectors for one flight, already
split into a track by step 02 and cleaned by step 02a.

**S2 — Smooth the altitude.** A centred five-sample rolling mean over
barometric altitude, computed over the *whole* track before any altitude cut.
Smoothing after the cut would truncate the window exactly where altitude is
changing fastest, which is the worst place to lose half of it.

**S3 — Pre-filter on pressure altitude.** A cheap, deliberately loose cut whose
only job is to keep the aerodrome join affordable. Its ceiling is the
configured cap *plus the highest field elevation in the reference*, so it
cannot discard a sample that the exact test at S6 would have kept.

**S4 — Join the aerodrome zones.** Each sample is matched to candidate
aerodromes through an H3 index at resolution 7, then narrowed to those within
the detection radius by exact great-circle distance.

**S5 — Attach the field elevation.** Each candidate aerodrome's elevation, from
OurAirports. A candidate whose elevation is unknown gets zero, which reduces the
next step to the sea-level behaviour rather than dropping the aerodrome.

**S6 — Cut on altitude.** *This is the step this version changes.* Samples
above the ceiling are excluded from the vote. `trend` now measures that ceiling
**above field elevation**; before this version it was a flight level.

**S7 — Vote.** Each surviving sample votes on the sign of its smoothed altitude
change: rising for a departure, falling for an arrival.

**S8 — Decide, or abstain.** If one direction leads the other by more than the
margin, the track is departing or arriving. Otherwise `trend` says nothing.

**E1 — Take the endpoint fix.** `endpoint` reads one sample, not many: the
track's first fix for a departure, its last for an arrival.

**E2 — Test the gates.** The fix must be within the radius, and either flagged
on the ground or no higher than the height gate **above field elevation**.

**E3 — Rank the survivors.** Nearest wins, on exact great-circle distance, plus
a penalty for aerodromes without scheduled service so that a military or
general-aviation field must be *clearly* nearest rather than merely nearest.

**E4 — Or say where it came from.** Where the gates fail and the fix sits at
the edge of the observed area, `endpoint` reports **out-of-area** rather than
guessing. Precedence is aerodrome first, border second: Ponta Delgada lies
about 8 NM inside the western edge, and the other order labelled its departures
out-of-area with the aircraft still on the runway.

### `trend`, arrivals

```{r fig-trend-arrival-src}
#| echo: false
#| results: asis
mermaid(
  "fig-trend-arrival",
  "`trend` naming an arrival aerodrome (ADES). S6 is the step this version changes.",
  c(
    'flowchart TD',
    '    T[("S1 · track<br/>state vectors")] --> SM["S2 · smooth baro_altitude<br/><i>centred 5-sample mean</i>"]',
    '    SM --> PRE["S3 · pre-filter<br/><i>cap + highest field elevation</i>"]',
    '    PRE --> J{{"S4 · join H3 aerodrome zones<br/><i>res 7, then exact distance</i>"}}',
    '    J --> EL["S5 · attach field elevation<br/><i>OurAirports; unknown &rarr; 0</i>"]',
    '    EL --> CUT["<b>S6 · altitude cut</b><br/><b><i>height above field &le; cap</i></b>"]',
    '    CUT --> V["S7 · vote per sample<br/><i>sign of smoothed change</i>"]',
    '    V --> M{"S8 · descending votes<br/>lead by the margin?"}',
    '    M -->|yes| R["E3 · rank survivors<br/><i>distance + scheduled penalty</i>"]',
    '    M -->|no| U(["undetermined"])',
    '    R --> A(["ADES = nearest survivor"])'
  )
)
```

:::: {.content-visible when-format="pdf"}
| Step | What it does |
|---|---|
| S1 | Read the track's state vectors |
| S2 | Smooth `baro_altitude` over a centred 5-sample window, before any cut |
| S3 | Pre-filter on pressure altitude at the cap plus the highest field elevation |
| S4 | Join the H3 aerodrome zones at resolution 7, then cut on exact distance |
| S5 | Attach the candidate aerodrome's field elevation; unknown becomes zero |
| **S6** | **Cut on height above field elevation** — the step this version changes |
| S7 | Each surviving sample votes on the sign of its smoothed altitude change |
| S8 | If descending votes lead by the margin the track is arriving; else *undetermined* |
| E3 | Rank survivors on distance plus the scheduled-service penalty; nearest becomes `ADES` |

::::

### `trend`, departures

The same machinery reading the same samples; only the direction of the winning
vote differs. `trend` has **no out-of-area concept** — it names an aerodrome or
says nothing — which is why the shipped configuration does not use it for
departures.

```{r fig-trend-departure-src}
#| echo: false
#| results: asis
mermaid(
  "fig-trend-departure",
  "`trend` naming a departure aerodrome (ADEP). Identical to the arrival case except at S8.",
  c(
    'flowchart TD',
    '    T[("S1 · track<br/>state vectors")] --> SM["S2 · smooth baro_altitude<br/><i>centred 5-sample mean</i>"]',
    '    SM --> PRE["S3 · pre-filter<br/><i>cap + highest field elevation</i>"]',
    '    PRE --> J{{"S4 · join H3 aerodrome zones<br/><i>res 7, then exact distance</i>"}}',
    '    J --> EL["S5 · attach field elevation<br/><i>OurAirports; unknown &rarr; 0</i>"]',
    '    EL --> CUT["<b>S6 · altitude cut</b><br/><b><i>height above field &le; cap</i></b>"]',
    '    CUT --> V["S7 · vote per sample<br/><i>sign of smoothed change</i>"]',
    '    V --> M{"S8 · climbing votes<br/>lead by the margin?"}',
    '    M -->|yes| R["E3 · rank survivors<br/><i>distance + scheduled penalty</i>"]',
    '    M -->|no| U(["undetermined<br/><i>never out-of-area</i>"])',
    '    R --> A(["ADEP = nearest survivor"])'
  )
)
```

:::: {.content-visible when-format="pdf"}
Identical to the arrival table above, with one difference: at **S8** the track
is a departure when *climbing* votes lead by the margin. Failing that, `trend`
returns *undetermined* — it has no out-of-area answer available to it.

::::

### `endpoint`, departures

```{r fig-endpoint-departure-src}
#| echo: false
#| results: asis
mermaid(
  "fig-endpoint-departure",
  "`endpoint` naming a departure aerodrome (ADEP). One fix, two gates, three possible answers.",
  c(
    'flowchart TD',
    '    T[("S1 · track<br/>state vectors")] --> E1["E1 · take the FIRST fix"]',
    '    E1 --> J{{"S4 · candidate aerodromes<br/><i>within the radius</i>"}}',
    '    J --> EL["S5 · attach field elevation<br/><i>OurAirports</i>"]',
    '    EL --> G{"E2 · within radius AND<br/>(on ground OR<br/><b>height above field &le; gate</b>)?"}',
    '    G -->|yes| R["E3 · rank survivors<br/><i>distance + scheduled penalty</i>"]',
    '    G -->|no| B{"E4 · fix at the edge<br/>of the observed area?"}',
    '    R --> A(["ADEP = nearest survivor"])',
    '    B -->|yes| OOA(["out-of-area"])',
    '    B -->|no| U(["undetermined"])'
  )
)
```

:::: {.content-visible when-format="pdf"}
| Step | What it does |
|---|---|
| S1 | Read the track's state vectors |
| E1 | Take the **first** fix — one sample, not many |
| S4 | Find candidate aerodromes within the radius |
| S5 | Attach each candidate's field elevation |
| **E2** | Gate: within the radius **and** either on the ground or **no higher than the gate above field elevation** |
| E3 | Passing: rank on distance plus the scheduled-service penalty; nearest becomes `ADEP` |
| E4 | Failing, with the fix at the edge of the observed area: **out-of-area**. Otherwise *undetermined* |

Precedence is aerodrome first, border second — the other order labelled Ponta
Delgada's departures out-of-area with the aircraft still on the runway.

::::

### `endpoint`, arrivals

```{r fig-endpoint-arrival-src}
#| echo: false
#| results: asis
mermaid(
  "fig-endpoint-arrival",
  "`endpoint` naming an arrival aerodrome (ADES). The last fix rather than the first.",
  c(
    'flowchart TD',
    '    T[("S1 · track<br/>state vectors")] --> E1["E1 · take the LAST fix"]',
    '    E1 --> J{{"S4 · candidate aerodromes<br/><i>within the radius</i>"}}',
    '    J --> EL["S5 · attach field elevation<br/><i>OurAirports</i>"]',
    '    EL --> G{"E2 · within radius AND<br/>(on ground OR<br/><b>height above field &le; gate</b>)?"}',
    '    G -->|yes| R["E3 · rank survivors<br/><i>distance + scheduled penalty</i>"]',
    '    G -->|no| B{"E4 · fix at the edge<br/>of the observed area?"}',
    '    R --> A(["ADES = nearest survivor"])',
    '    B -->|yes| OOA(["out-of-area"])',
    '    B -->|no| U(["undetermined"])'
  )
)
```

:::: {.content-visible when-format="pdf"}
Identical to the departure table above, with one difference: at **E1** the
method reads the track's **last** fix rather than its first.

::::

### Where the two datums part company

Only **S6** and **E2** read an altitude, and until this version they read it
against different references:

| Test | Threshold | Datum, before v6.1 | Datum, from v6.1 |
|---|---|---|---|
| `trend` S6 | `trend_max_height_ft` | flight level (sea level) | **above field elevation** |
| `endpoint` E2 | `endpoint_height_ft` | above field elevation | above field elevation |
| step 04 ground phase | `phase_ground_ceiling_ft` | above field elevation | above field elevation |
| `nearest` | *none* | — | — |

`nearest` applies no altitude condition at all, on any datum. That is
deliberate: it is the naive `traffic`-faithful baseline the other methods are
scored against, and giving it a gate would defeat its purpose.

::: {.callout-important}
## A flight level is not a height

`FL60` is not the same cut as `6,000 ft`. The pipeline derives the flight level
with an integer cast, so `flight_level <= 60` admits everything below **6,100
ft**. The datum comparison in this report therefore holds the ceiling at 6,100
ft above field, not 6,000 — otherwise it would move the ceiling and the datum
at the same time and could not attribute the result to either.
:::

## Results

```{r results-guard}
#| echo: false
#| results: asis
swap <- read_if("datum_swap_2025.csv")
bands <- read_if("elevation_bands_2025.csv")
if (is.null(swap)) {
  cat("::: {.callout-warning}\n## Not yet measured\n\n",
      "The arms defined in `regenerate_v61.py` have not been run against this ",
      "checkout, so this section is deliberately empty rather than populated ",
      "with numbers carried over from another study.\n:::\n", sep = "")
}
```

### The band breakdown, read first {#sec-bands}

This section comes before the headline deliberately. The headline is a single
pair of numbers and it is consistent with several explanations; the band
breakdown is the only part of this study that can distinguish between them. Read
in the other order, it becomes a search for confirmation.

The prediction the design committed to, before any of this was run: **no
movement at sea-level aerodromes, and gains concentrated at elevated ones.**

```{r bands-table}
#| echo: false
if (!is.null(bands)) {
  b <- bands[bands$role == "arrivals",
             c("band", "aerodromes", "n_truth", "n_correct_msl",
               "n_correct_field", "delta_correct", "delta_correct_loo",
               "delta_correct_ex_busiest")]
  b$relative <- sprintf("%+.2f%%", 100 * b$delta_correct / pmax(b$n_correct_msl, 1))
  names(b) <- c("Band", "Aerodromes", "Truth", "Correct MSL", "Correct field",
                "Δ", "Δ less top mover", "Δ less busiest", "Δ relative")
  kable(b, row.names = FALSE, format.args = list(big.mark = ","),
        caption = "Arrivals, 2025 sample. Δ is field minus sea level.")
}
```

**The effect is where it was predicted to be.** Measured against each band's own
base of correct arrivals — which is the honest denominator, since the bands
differ in size by two orders of magnitude — the sea-level control band moves by
+0.12% and the 1,500–3,000 ft band by **+5.7%**. That is a difference of roughly
forty-fold between the band where the datum change is arithmetically a no-op and
the band where it bites hardest.

It also survives both robustness checks. The census run before this study began
showed each treatment band resting on a single aerodrome, so a band mean alone
could not separate an elevation effect from a Madrid effect. Removing the
largest mover leaves +70 of the +109; removing Madrid, the busiest field in the
band, leaves +110 — very slightly *more*. The result is not one aerodrome's
result.

```{r departures-control}
#| echo: false
#| results: asis
if (!is.null(bands)) {
  dep <- bands[bands$role == "departures", ]
  if (all(dep$delta_correct == 0)) {
    cat("::: {.callout-note}\n## The control that had to hold, and did\n\n",
        "**Departures move by exactly zero in every band.** The shipped ",
        "configuration takes departures from `endpoint`, which never reads ",
        "`trend`'s altitude cut — so a non-zero number anywhere in that row ",
        "would have meant this change was touching something it has no ",
        "business touching. It is the cheapest available check that the ",
        "intervention is the one described.\n:::\n", sep = "")
  }
}
```

### What it costs

```{r swap-table}
#| echo: false
if (!is.null(swap)) {
  s <- swap[swap$run %in% c("legacy", "datum_msl", "datum_field"),
            c("run", "ades_coverage", "ades_accuracy", "ades_correct",
              "ades_wrong", "ades_score")]
  s$ades_coverage <- sprintf("%.2f%%", 100 * s$ades_coverage)
  s$ades_accuracy <- sprintf("%.2f%%", 100 * s$ades_accuracy)
  names(s) <- c("Run", "Coverage", "Accuracy", "Correct", "Wrong", "Score (k=2)")
  kable(s, row.names = FALSE, format.args = list(big.mark = ","),
        caption = "Arrivals through `process_dai` itself, 2025 sample.")
}
```

Moving to the field datum buys **194 more correct arrivals and 105 more wrong
ones**. Coverage rises 0.31 pp; accuracy falls 0.16 pp; the score at *k* = 2
moves from 62,185 to 62,169 — sixteen points the wrong way. In net terms it is
close to a wash, tilted toward answering more flights and being fractionally
less certain about each.

### Why it ships anyway {#sec-adoption}

The adoption criteria fixed in the design were: coverage up by at least 0.5 pp,
accuracy not falling, and the gain concentrated at elevated fields. **The third
holds and the first two do not.** On a cost-benefit reading, that would be the
end of it.

**The change is adopted regardless**, and the reason is worth stating plainly
because it is not a benchmark result.

A ceiling measured from sea level is *not the same test at different
aerodromes*. At Amsterdam it admits 6,000 ft of climb and descent; at Ankara,
2,975; at Erzurum, roughly 240. Nobody chose that gradient — it is an artefact
of which datum the threshold happened to be written against, and it makes the
method's willingness to answer depend on the terrain under the aerodrome rather
than on the evidence in the trajectory. `endpoint` was corrected on exactly this
reasoning in version 6, and step 04's ground detection in the same release.
`trend` was the last one left.

So this is a **correctness fix, not an optimisation**, and it is not the
benchmark's place to veto it. What the benchmark is for is telling us what the
fix costs, which it has done: a third of a percentage point of coverage gained,
a sixth of a point of accuracy given up, and the movement concentrated where the
mechanism predicts. That is the honest accounting, and it is reported here
rather than buried because a change adopted on principle still has to declare
its price.

The measured null in criteria 1 and 2 does carry one design implication: the
ceiling itself, inherited from a sweep conducted on the datum now abandoned, is
very likely no longer optimal. @sec-outstanding returns to that.

### The band that should have gained most, and did not

The `>3000` band is the one where a fixed sea-level ceiling ought to hurt worst,
and it shows **no change at all: 5 correct arrivals out of 705, on both
datums.**

That is not a null result about the datum. It is `trend` failing at those
aerodromes for a different reason — a recall of 0.7% is not a method being
mis-tuned, it is a method receiving almost nothing to work with. The likeliest
cause is ADS-B coverage over the Anatolian plateau, where most of this band's
traffic sits; an aerodrome the feed barely sees cannot be named on any datum.

This matters for how the change should be argued elsewhere. The motivating
example in the design was Erzurum at 5,763 ft, where a FL60 cut leaves about
240 ft in which to vote. That example is arithmetically correct and doubly
unrepresentative: such fields are rare in the sample, **and** they sit in the
one band where the method is independently broken. The population that actually
benefits is the far less dramatic 1,500–3,000 ft — Madrid, Ankara, Sofia,
Tbilisi — where a FL60 cut still leaves three to four thousand feet of usable
climb and descent.

### The ceiling survives the move {#sec-ceiling}

`trend_max_height_ft` ships at the FL60 equivalent because that is what the
datum comparison had to hold fixed to isolate one variable. That left an obvious
worry: FL60 was tuned by V6 *on the datum this version abandons*, so there was
no reason its optimum should have survived. The two sweeps settle it.

```{r ceiling-sweeps}
#| echo: false
hs <- read_if("height_sweep_2025.csv")
fs <- read_if("fl_sweep_2025.csv")
shipped <- function(d) d[d$radius_nm == 30 & d$margin == 0 & d$penalty_nm == 10, ]
if (!is.null(hs) && !is.null(fs)) {
  h <- shipped(hs); f <- shipped(fs)
  out <- rbind(
    data.frame(Datum = "above field", Ceiling = sprintf("%s ft", format(h$fl_cap, big.mark = ",")),
               Coverage = sprintf("%.2f%%", 100 * h$ades_coverage),
               Accuracy = sprintf("%.2f%%", 100 * h$ades_accuracy),
               Score = h$ades_score),
    data.frame(Datum = "flight level", Ceiling = sprintf("FL%s", f$fl_cap),
               Coverage = sprintf("%.2f%%", 100 * f$ades_coverage),
               Accuracy = sprintf("%.2f%%", 100 * f$ades_accuracy),
               Score = f$ades_score)
  )
  kable(out, row.names = FALSE, format.args = list(big.mark = ","),
        caption = "Arrivals, shipped geometry (30 NM, margin 0, penalty 10 NM). Both curves swept out of the same vote cache.")
}
```

**Each datum peaks at the equivalent ceiling** — 6,100 ft above field, FL60 on
sea level — and 6,100 ft wins at *every* radius from 20 NM outward, production's
30 NM included. The shipped value is therefore tuned rather than inherited, and
no change is indicated.

Two honest limits. The grid is coarse near the peak (4,000 → 6,100 → 8,000), so
the true optimum could sit anywhere in that span; and these are harness numbers,
which V6 showed can differ from the pipeline's in level even when they agree in
shape. What the sweep establishes is the *shape*: an interior optimum, falling
away on both sides, at the value already shipping.

### And it replicates on the second period {#sec-second-period}

```{r second-period}
#| echo: false
h24 <- read_if("height_sweep_2024.csv")
f24 <- read_if("fl_sweep_2024.csv")
if (!is.null(h24) && !is.null(f24)) {
  h <- shipped(h24)
  kable(
    data.frame(
      Ceiling  = sprintf("%s ft", format(h$fl_cap, big.mark = ",")),
      Coverage = sprintf("%.2f%%", 100 * h$ades_coverage),
      Accuracy = sprintf("%.2f%%", 100 * h$ades_accuracy),
      Score    = h$ades_score
    ),
    row.names = FALSE, format.args = list(big.mark = ","),
    caption = "2024 sample, above-field datum, arrivals at the shipped geometry."
  )
}
```

**6,100 ft is the optimum on 2024 as well**, at every radius from 20 NM outward
— the same value, the same interior peak, on an independent sample a year
apart. A ceiling that survives that is tuned, not fitted.

The datum comparison replicates too, including its sign: the best above-field
cell scores 0.02% below the best sea-level cell on 2025 and 0.25% below on 2024.
Both are fractions of a percent, and both are diluted for the reason
@sec-dilution gives. What matters is that the second period tells the same story
rather than a different one — the finding is the algorithm's and not the
sample's.

::: {.callout-note}
## Why this period is sweeps only

The 2024 tracks pre-date this study's H3 indexing, and `process_dai` reads
`h3_res_7` off the track table directly, so the pipeline arms cannot run here at
all. V6 met the same wall and ran the pipeline on 2025 alone. The sweeps get
around it because their vote cache is built with the index computed at read
time.

So the second period confirms the *ceiling* and the *datum comparison*, which is
what it is able to confirm. It does not independently reproduce the band
breakdown, and this report does not claim that it does.
:::

### Why the aggregate was never going to show much {#sec-dilution}

The two curves are also nearly identical in level — 60,952 against 60,857 at the
shipped geometry, and 61,069 against 61,056 at each datum's own best cell. That
looks like a change doing nothing, and it is worth being precise about why it is
not.

**Roughly seventy per cent of the ground-truth arrivals sit in the `<500 ft`
band**, where the two datums are the same test by construction. A change that
cannot move the majority of the sample cannot move an average over it, however
well it works on the rest. The aggregate score is diluted by design, and judging
a targeted correctness fix by it is using the wrong instrument — the reading is
dominated by the aerodromes the fix was never about.

That is precisely why @sec-bands exists and why it is read first. The signal is
in the 1,500–3,000 ft band, and the aggregate is where that signal goes to be
averaged away.

## What is still open {#sec-outstanding}

**The second period confirms through the sweeps, not the pipeline.** Everything
above is 2025-06-05…07. The 2024 sample cannot run the pipeline arms at all:
`process_dai` reads `h3_res_7` off the track table, and the 2024 tracks were
built before this study indexed them, so the column is absent. V6 met the same
wall and ran the pipeline on 2025 alone.

What 2024 *can* do is the paired sweeps, out of a vote cache built with the
index computed at read time — so the second period tests whether the ceiling and
the datum behave the same way there, which is the part most at risk of being a
one-sample artefact. Those two jobs are the whole of this period's evidence, and
the report should say so rather than implying a symmetry the data does not
support.

**The `>3000` band needs a different question.** A recall of 0.7% on both datums
is a coverage problem upstream of detection, and no amount of threshold work
will move it. Whether that is receiver geometry over the Anatolian plateau or
something in ingestion is worth knowing, but it is not this study's question.

**The pipeline ceiling walk was not run, deliberately.** `height_pipeline_2025`
would step the above-field ceiling through `process_dai` itself, as V6 did for
its flight-level cap — the check that catches a harness optimum which does not
survive contact with production.

It is not run here because it would confirm a value nothing is changing. The
sweeps put the optimum at 6,100 ft on both periods and at every radius from
20 NM outward, and 6,100 ft is what already ships. Six pipeline runs to agree
with the shipped configuration is a poor use of a shared cluster.

It becomes worth running the moment someone proposes moving the ceiling. Until
then the honest statement is that the ceiling is confirmed *by the harness on
two periods* and not independently by the pipeline, and the job stays in the
registry so that check is one command away.
