---
title: "Detecting departure and destination aerodromes from ADS-B"
subtitle: "Version 6.2 — the consolidated study on the datum that ships"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r regenerate}
#| include: false
#| cache: false

# Rendering this page runs the analysis behind it.
#
# `regenerate_v62.py` is the executable definition of the study: every CSV read
# below is produced by exactly one job it declares, with exactly the arguments
# it declares. It re-runs a job when the *code that job depends on* has changed
# since the output was written -- not when the file is old, which says nothing.
# So a render with everything current is a fast no-op, and a render after
# editing `flights.py` recomputes every pipeline result before the first table
# is drawn.
#
# Running a stale job needs cluster credentials, because the numbers come from
# Spark over S3 against Network Manager reference data and there is no way to
# recompute them without the data they are computed from. Set
# OPDI_RENDER=check to fail fast on staleness instead, or OPDI_RENDER=allow-
# stale to render a draft from whatever is staged -- which the provenance
# table at the foot of this page will then show as unverified.
REPO   <- normalizePath(file.path(getwd(), "..", "..", "..", "opdi"), mustWork = FALSE)
PYTHON <- Sys.getenv("OPDI_PYTHON", file.path(REPO, ".venv310", "bin", "python"))
MODE   <- Sys.getenv("OPDI_RENDER", "run")

regen <- file.path(REPO, "benchmarks", "regenerate_v62.py")
if (file.exists(regen) && MODE != "skip") {
  flags <- switch(MODE,
                  "run" = character(0),
                  "check" = "--check",
                  "allow-stale" = c("--check", "--allow-stale"),
                  character(0))
  status <- system2(PYTHON, c(shQuote(regen), flags), stdout = "", stderr = "")
  if (!identical(status, 0L)) {
    stop("regenerate_v62.py exited ", status,
         " -- the figures below would not match the code that is checked out. ",
         "Set OPDI_RENDER=allow-stale to render anyway.")
  }
}
```

```{r setup}
#| include: false
library(knitr)
library(ggplot2)
library(grid)

# Read-only. Every figure and table on this page comes from a committed CSV
# under data/. Nothing here reads S3, a database or a Spark cluster.
#
# Every one of those CSVs was produced by *this* study, against the pipeline as
# it stands at the end of the parameterisation work. No number is carried over
# from versions 1-5 or from the decimation study. That is not bookkeeping: the
# bearing results in version 5 pre-date a modulo fix that changed them, the
# arrival comparison pre-dates a corrected `endpoint` baseline, and the trend
# numbers came from a standalone reimplementation. Re-measuring everything is
# what allows the arithmetic on this page to agree with itself.
cache <- function(f) {
  p <- file.path("data", f)
  if (!file.exists(p)) return(NULL)
  read.csv(p, stringsAsFactors = FALSE)
}

# The trend sweep, on the datum that ships: ceilings in feet above the
# aerodrome's own field elevation. `trend_msl` is the same measurement on the
# sea-level datum, out of the same vote cache -- so the two are a comparison
# rather than two studies, and `trend_msl` is also what reproduces version 6's
# published sweep, cell for cell.
trend      <- cache("height_sweep_2025.csv")
trend_msl  <- cache("fl_sweep_2025.csv")
# The second period, loaded here rather than where it is first
# plotted: chunks run in document order, and the datum chapter
# above the stability section needs these too.
t24        <- cache("height_sweep_2024.csv")
t24_msl    <- cache("fl_sweep_2024.csv")
grid   <- cache("sweep_radius_height_2025.csv")
pen    <- cache("sweep_penalty_2025.csv")
cone   <- cache("sweep_cone_2025.csv")
modes  <- cache("mode_comparison_v6.csv")
apt    <- cache("per_airport_v6.csv")
typ    <- cache("per_type_v6.csv")

pct  <- function(x) ifelse(is.na(x), "—", sprintf("%.2f%%", 100 * x))
pp   <- function(x) ifelse(is.na(x), "—", sprintf("%+.2f pp", 100 * x))
num  <- function(x) format(round(x), big.mark = ",", trim = TRUE)
# The "+" flag does nothing for %s in R, so the sign is prepended explicitly --
# a signed delta that silently loses its sign reads as a level, not a change.
sgn  <- function(x) paste0(ifelse(x >= 0, "+", "−"),
                           format(abs(round(x)), big.mark = ",", trim = TRUE))

# --- counts, and the score -------------------------------------------------
# The sweep frames carry coverage and overall as ratios of the same ground
# truth. The counts are exactly recoverable from them -- coverage * n is the
# number answered, overall * n the number correct, and both are integers -- so
# every table below can show counts without a separate export. Verified against
# the trend sweep, which carries both: all rows agree.
N_GT <- if (!is.null(trend)) trend$n_ground_truth[1] else grid$n_ground_truth[1]

K_DEFAULT <- 2

answered <- function(d, role) round(d[[paste0(role, "_coverage")]] * d$n_ground_truth)
correct  <- function(d, role) round(d[[paste0(role, "_overall")]]  * d$n_ground_truth)
wrong    <- function(d, role) answered(d, role) - correct(d, role)
scr      <- function(d, role, k = K_DEFAULT) correct(d, role) - k * wrong(d, role)

# Attach the five reported quantities to any sweep frame, for either role, so
# no table has to recompute them and none can disagree with another.
with_metrics <- function(d, role, k = K_DEFAULT) {
  if (is.null(d)) return(NULL)
  d$correct  <- correct(d, role)
  d$wrong    <- wrong(d, role)
  d$score    <- d$correct - k * d$wrong
  d$coverage <- d[[paste0(role, "_coverage")]]
  d$accuracy <- d[[paste0(role, "_accuracy")]]
  d
}

# The best row of a sweep for one role, at a given k.
best_row <- function(d, role, k = K_DEFAULT, subset = TRUE) {
  x <- with_metrics(d[subset, , drop = FALSE], role, k)
  if (is.null(x) || nrow(x) == 0) return(NULL)
  x[which.max(x$score), ]
}

# --- headline values, for the summary -------------------------------------
# Computed here rather than written into the prose, so the opening paragraph
# cannot drift away from the tables that justify it.
mv <- function(run, field) {
  if (is.null(modes)) return(NA_real_)
  v <- modes[[field]][modes$run == run]
  if (length(v) == 0) NA_real_ else v[1]
}
ADEP_GAIN <- mv("recommended", "adep_score") - mv("legacy", "adep_score")

# The shipped configuration is the `combined` run: endpoint for departures,
# exact-distance trend for arrivals. Verified field-for-field against
# DetectionConfig() -- same caps, radii, penalties and ranking rule -- so this
# row is what production produces, not an approximation of it.
SHIPPED <- "combined"
TOTAL_SHIPPED <- mv(SHIPPED, "adep_score") + mv(SHIPPED, "ades_score")
TOTAL_LEGACY  <- mv("legacy", "adep_score") + mv("legacy", "ades_score")

top100_recall <- function(run, rl) {
  if (is.null(apt)) return(NA_real_)
  x <- apt[apt$run == run & apt$role == rl & apt$airport != "OOA", ]
  x <- x[order(-x$n_truth), ][1:min(100, nrow(x)), ]
  sum(x$n_correct) / sum(x$n_truth)
}

# v6.1 spelled this `read_if`; `cache` is version 6's name for the same thing.
# Kept as an alias so the sections carried over from v6.1 read unchanged.
read_if <- cache

# A ceiling, written the way its datum requires.
#
# Both sweeps carry the cap in `fl_cap`, so the column name cannot be trusted
# to say what the number means -- the `datum` column says that. Writing "FL6100"
# or "6,100 ft above field" for the wrong one is the kind of error a reader
# cannot catch, because both are plausible sentences.
cap_label <- function(v, datum = "field") {
  if (identical(datum, "msl")) sprintf("FL%g", v)
  else sprintf("%s ft above field", format(v, big.mark = ",", trim = TRUE))
}

# The datum a sweep frame was measured on, read off the frame rather than
# assumed by the caller.
datum_of <- function(df) {
  if (is.null(df) || !"datum" %in% names(df)) return("msl")
  u <- unique(df$datum)
  if (length(u) == 1) u else "mixed"
}

# The shipped arrival configuration, named once so no section can quote a
# different one than its neighbour.
#
# Two ceilings because there are two datums: `trend_max_height_ft` is 6,000 ft
# above field, and the sea-level cut admitting the same band is FL60.
# Read from the `recommended` run, which the harness builds as a bare
# DetectionConfig() -- so these ARE the shipped values, not a memory of them.
#
# They were hardcoded here first, as 20 NM and margin 2: version 6's claimed
# values, which the code has never had. That is the precise failure this report
# documents, reproduced inside the report documenting it. Hence no literals.
#
# The fallback is only for a render with no pipeline output at all, and is
# marked as such wherever it could be mistaken for a measurement.
SHIPPED_TREND <- local({
  if (is.null(modes) || !"recommended" %in% modes$run)
    return(list(ceiling_field = NA, ceiling_msl = NA, margin = NA,
                radius_nm = NA, penalty_nm = NA))
  r <- modes[modes$run == "recommended", ][1, ]
  list(ceiling_field = as.numeric(r$trend_max_height_ft),
       ceiling_msl   = as.numeric(r$trend_max_fl),
       margin        = as.numeric(r$trend_vote_margin),
       radius_nm     = as.numeric(r$trend_radius_nm),
       penalty_nm    = as.numeric(r$trend_sched_penalty_nm))
})

shipped_ceiling <- function(datum) {
  if (identical(datum, "msl")) SHIPPED_TREND$ceiling_msl
  else SHIPPED_TREND$ceiling_field
}

# The grid point standing in for the shipped ceiling, and whether it *is* it.
#
# The research sweep's above-field grid does not contain 6,000 ft -- it was
# built around 6,100, the height equivalent of FL60 -- so on that datum the
# shipped value cannot be located exactly and the nearest point stands in.
# Returning the substitution rather than hiding it lets each section say so.
# The pipeline grid in @sec-pipeline does carry both and settles it.
nearest_cap <- function(caps, datum) {
  want <- shipped_ceiling(datum)
  caps <- sort(unique(caps))
  if (want %in% caps) return(list(cap = want, exact = TRUE, want = want))
  list(cap = caps[which.min(abs(caps - want))], exact = FALSE, want = want)
}

# A vertical flow diagram, drawn with grid so it renders through whatever
# device Quarto has selected -- HTML and PDF alike, with no browser.
#
# Rasterising mermaid needs a headless Chromium that does not start in this
# environment, so the mermaid versions of these diagrams could only ever appear
# in HTML. All four are the same shape: a chain of labelled steps, one of them
# emphasised, ending in a decision; the endpoint pair has a second decision on
# the failing branch. One helper serves all four -- four hand-placed drawings
# would not have been worth the gain over the step tables they replace.


.fd_lines <- function(s) paste(strsplit(s, "<br>")[[1]], collapse = "\n")

#' @param steps      character vector, one per box. "A<br>B" splits to two lines.
#' @param emphasis   index of the step drawn heavy (the datum-sensitive one).
#' @param decision   the question in the first diamond.
#' @param yes,no     outcomes of that decision. `no` may instead be a list
#'                   describing a second decision: list(q=, yes=, no=).
flow_diagram <- function(steps, emphasis = NA, decision = NULL,
                         yes = NULL, no = NULL) {
  nested <- is.list(no)
  rows <- length(steps) + (!is.null(decision)) + nested + 1L

  ytop <- 0.975; ybot <- 0.025
  h <- (ytop - ybot) / rows
  bh <- h * 0.58
  yc <- function(i) ytop - (i - 0.5) * h

  box <- function(x, y, label, heavy = FALSE, w = 0.60, cex = 0.70) {
    grid.roundrect(x = x, y = y, width = w, height = bh,
                   gp = gpar(lwd = if (heavy) 2.4 else 1,
                             col = if (heavy) "black" else "grey35",
                             fill = if (heavy) "grey92" else "white"))
    grid.text(.fd_lines(label), x = x, y = y,
              gp = gpar(cex = cex, fontface = if (heavy) "bold" else "plain"))
  }

  diamond <- function(x, y, label, w = 0.24) {
    grid.polygon(x = c(x, x + w, x, x - w),
                 y = c(y + bh * 0.62, y, y - bh * 0.62, y),
                 gp = gpar(col = "grey35", fill = "white"))
    grid.text(.fd_lines(label), x = x, y = y, gp = gpar(cex = 0.62))
  }

  arr <- function(x1, y1, x2, y2, label = NULL) {
    grid.lines(x = c(x1, x2), y = c(y1, y2),
               arrow = arrow(length = unit(2, "mm"), type = "closed"),
               gp = gpar(fill = "black", col = "grey35"))
    if (!is.null(label))
      grid.text(label, x = (x1 + x2) / 2 + 0.035, y = (y1 + y2) / 2,
                gp = gpar(cex = 0.58, col = "grey30"))
  }

  # --- the chain ------------------------------------------------------------
  for (i in seq_along(steps)) {
    box(0.5, yc(i), steps[i], heavy = !is.na(emphasis) && i == emphasis)
    if (i > 1) arr(0.5, yc(i - 1) - bh / 2, 0.5, yc(i) + bh / 2)
  }
  if (is.null(decision)) return(invisible(NULL))

  # --- first decision -------------------------------------------------------
  d1 <- length(steps) + 1L
  arr(0.5, yc(d1 - 1) - bh / 2, 0.5, yc(d1) + bh * 0.62)
  diamond(0.5, yc(d1), decision)

  if (!nested) {
    ty <- yc(d1 + 1L)
    arr(0.5 - 0.10, yc(d1) - bh * 0.5, 0.26, ty + bh / 2, "no")
    box(0.26, ty, no, w = 0.40, cex = 0.68)
    arr(0.5 + 0.10, yc(d1) - bh * 0.5, 0.74, ty + bh / 2, "yes")
    box(0.74, ty, yes, w = 0.40, cex = 0.68)
    return(invisible(NULL))
  }

  # --- nested decision on the failing branch --------------------------------
  d2 <- d1 + 1L
  ty <- yc(d2 + 1L)
  # yes from the first decision goes straight to its outcome, skipping a row.
  arr(0.5 + 0.12, yc(d1) - bh * 0.4, 0.76, yc(d2) + bh / 2, "yes")
  box(0.76, yc(d2), yes, w = 0.40, cex = 0.68)
  # no descends into the second question.
  arr(0.5 - 0.12, yc(d1) - bh * 0.4, 0.30, yc(d2) + bh * 0.62, "no")
  diamond(0.30, yc(d2), no$q, w = 0.22)
  arr(0.30 - 0.08, yc(d2) - bh * 0.5, 0.16, ty + bh / 2, "no")
  box(0.16, ty, no$no, w = 0.28, cex = 0.66)
  arr(0.30 + 0.08, yc(d2) - bh * 0.5, 0.50, ty + bh / 2, "yes")
  box(0.50, ty, no$yes, w = 0.28, cex = 0.66)
  invisible(NULL)
}
```

## Summary {.unnumbered}

OPDI derives a flight's departure and destination aerodrome from ADS-B state
vectors alone. This page consolidates six versions of that work into one
account: what the three candidate methods are, how each behaves across its
whole parameter space, which settings are best for departures and for arrivals,
what the ideas that did *not* work were, and what the pipeline should ship.

The short version is in four parts.

**The two roles want different algorithms.** Departures are best served by
`endpoint` — take the aerodrome nearest the first fix, but only commit when the
fix is close and low, and abstain otherwise. Arrivals are best served by
`trend`, the algorithm published before 2026. That split holds at every
exchange rate tested.

**Both roles improve.** The recommended configuration — `endpoint` for
departures, exact-distance `trend` for arrivals — scores **`r num(TOTAL_SHIPPED)`**
against legacy's **`r num(TOTAL_LEGACY)`** on a three-day sample of
`r num(N_GT)` flights: a gain of **`r sgn(TOTAL_SHIPPED - TOTAL_LEGACY)`**,
every figure from a real pipeline run.

**A better way of choosing among aerodromes is what makes the tuning work.**
This is the substantive change, and it is worth more than any threshold on this
page. The published algorithm narrowed its candidates by **H3 ring count** — an
integer stepping about 5.2 km at resolution 7 — and computed exact distance
only among the survivors, so an aerodrome a kilometre further out was
eliminated before its distance was ever measured. It now carries every
candidate through to an exact great-circle comparison and chooses on that.

The difference is not marginal. Applied to arrivals, the same four tuned
values are worth
**`r if (!is.null(cache("pipeline_path_ring_v6.csv"))) { rp <- cache("pipeline_path_ring_v6.csv"); sgn(rp$ades_score[rp$run=="path4_radius"] - rp$ades_score[rp$run=="path0_legacy"]) } else "—"`**
under ring selection and
**`r if (!is.null(cache("pipeline_path_v6.csv"))) { pp_ <- cache("pipeline_path_v6.csv"); sgn(pp_$ades_score[pp_$run=="path4_radius"] - pp_$ades_score[pp_$run=="path0_legacy"]) } else "—"`**
under exact distance — the same parameters, the same data, opposite signs.
Raising the altitude gate admits samples from higher up, where more
aerodromes fall in the same ring, so the coarse rule had most work to do
exactly where it was least able to do it. Every tuned arrival value that fails
under ring selection gains under exact distance.

**`trend`'s altitude cut now measures from the aerodrome, not from sea level.**
This is what version 6.2 adds. Until it, `trend` gated on a flight level while
every other altitude test in the pipeline — `endpoint`'s, step 04's ground
phase — gated on a height above the aerodrome's own field elevation. The two
agree at sea level and diverge by the field elevation everywhere else, so the
same nominal ceiling gave a Madrid arrival 2,000 ft less room to vote than an
Amsterdam one.

Moving it costs a little accuracy and buys a little coverage, and the aggregate
of those two is close to nothing — @sec-dilution explains why it was never
going to be otherwise, since most arrivals are at aerodromes below 500 ft where
the two datums are identical by construction. The case for the change is
not the aggregate: it is that the cut now means the same thing at every
aerodrome, and that @sec-bands shows the gain landing precisely where elevation
predicts it should.

## What this is about {#sec-orientation}

If you have not met this project before, this section is the whole of what you
need. Nothing later on assumes another document.

**OPDI — the Open Performance Data Initiative — publishes a description of
European air traffic derived from what aircraft themselves broadcast.** Most
airliners transmit their position, altitude and identity several times a
second over ADS-B, and a network of ground receivers records it. That signal is
open, but it is raw: a stream of positions with no notion of a *flight*. OPDI's
job is to turn it into something you can count with, and to publish the result
openly rather than behind a data-sharing agreement.

**The flight list is the backbone of that.** The pipeline groups each aircraft's
positions into **tracks** — one continuous journey, split where the signal
stops for long enough — and then gives each track a row: when it started, when
it ended, which aircraft, and the two fields this report is about.

**ADEP and ADES are the departure and destination aerodromes.** They are ICAO
codes: `EHAM` for Amsterdam, `LEMD` for Madrid. Almost every question anyone
asks of flight data needs them — how many movements an airport had, how long a
city pair takes, which routes grew. Get them wrong and every count built on top
is wrong in a way that is hard to see.

**Nothing in the ADS-B signal says where a flight is going.** The aircraft
broadcasts where it *is*. So the aerodromes have to be inferred from the shape
of the trajectory, and that is harder than it sounds:

- Aerodromes cluster. Several of London's lie within a few minutes' flying
  of one another, and the nearest to a descending aircraft is often not the
  one it lands at.
- Coverage is uneven. Low-altitude reception is poor in places, so a track may
  simply stop before it reaches the ground, or start well after take-off.
- Some flights genuinely begin or end outside the observed area, and a method
  that always answers will confidently name a wrong aerodrome for them.

**So this report is about three ways of guessing, and how well each does.**
Each looks at the trajectory near a candidate aerodrome and decides whether to
commit:

- **`trend`** watches whether the aircraft is climbing or descending near each
  candidate, and abstains when the altitude trace says neither clearly.
- **`endpoint`** takes the aerodrome nearest the first or last position, but
  only commits when that position is close and low; otherwise it abstains.
- **`nearest`** takes the closest aerodrome unconditionally. It never abstains,
  and it is included as the external baseline — it is what the widely used
  `traffic` library does.

Two of the three can refuse to answer, and that is the point. A method that
always answers has perfect **coverage** and poor **accuracy**; one that answers
only when certain has the reverse. Neither number means much alone, which is
why @sec-metrics sets out how they are traded against one another.

**What this version adds.** Version 6.2 changes one thing in `trend`: the
altitude ceiling below which a position is allowed to vote is now measured
**from the aerodrome's own elevation** rather than from sea level. The two are
the same at a coastal airport and differ by half a mile at a highland one.
@sec-datum reports what that was worth.

### A glossary

Terms this report uses without further explanation.

| Term | Meaning |
|---|---|
| **ADEP / ADES** | The departure and destination aerodrome, as ICAO codes. |
| **Track** | One continuous run of positions from a single aircraft, split where the signal gaps. The unit each row of the flight list describes. |
| **State vector** | A single broadcast position: latitude, longitude, altitude, velocity, at one instant. |
| **H3 cell** | A hexagon from Uber's H3 grid, used to index aerodromes by location so candidates can be found without comparing against every airport in Europe. |
| **Coverage** | The share of reference flights the method answered at all. Abstaining lowers it. |
| **Accuracy** | Of the flights it answered, the share it got right. |
| **Score, *k*** | Coverage and accuracy pull against each other, so runs are ranked by `correct − k × wrong`. *k* is the exchange rate: how many correct answers one wrong answer is worth. This report uses *k* = 2 throughout and @sec-k shows what happens at other values. |
| **Out-of-area** | A flight that genuinely began or ended outside the observed region. Saying so is a *correct* answer, not an abstention. |
| **Abstention** | The method declining to name an aerodrome. Distinct from being wrong, and distinct from out-of-area. |
| **Ground truth** | EUROCONTROL Network Manager flight data, used as the reference. Not published with this report; see @sec-provenance. |

::: {.callout-note}
## Reading this page
Every number below is an inline expression over a CSV in `data/`, and every one
of those CSVs is produced by a job in `benchmarks/regenerate_v62.py` — which
this page runs before drawing anything. Nothing is typed into the prose, and
nothing is read that the checked-out code did not produce. @sec-provenance
lists the command and commit behind each figure. Where this page disagrees with
versions 1–5, it says so and gives the reason; those versions stay published
unchanged.
:::

## How the metrics work {#sec-metrics}

Five quantities are reported for every setting, for each role separately. Four
of them are unavoidable and the fifth is a judgement, so it is worth being
precise about all five before any of them are used to argue for a parameter.

Write $N$ for the number of ground-truth flights in the sample, $a$ for the
number the method *answers* (names an aerodrome, or explicitly says
out-of-area), $c$ for the number it answers **correctly**, and $w = a - c$ for
the number it answers **wrongly**.

**Coverage** is $a/N$: the share of flights the method is willing to commit on.
A method that abstains everywhere has coverage zero and cannot be wrong.

**Accuracy** is $c/a$: of the flights it committed on, the share it got right.
Accuracy says nothing about how often the method declined to answer, so a
method can buy accuracy simply by abstaining more.

**Correct** ($c$) and **wrong** ($w$) are the raw counts. They matter because
coverage and accuracy generally move in opposite directions — every extra
answer a method gives is one it was previously unsure about, so raising
coverage tends to lower accuracy — and neither ratio says whether a particular
exchange was worth making. The counts do.

**Score** is $c - k\,w$. It exists to make the exchange rate explicit: a change
to a parameter is worth taking if and only if it wins more than $k$ correct
answers for each wrong one it introduces.

The value of $k$ is where judgement enters. This study uses $k = 2$ by default,
on the following argument. A **silence** — an abstention — corrupts one
movement count, and it is *visible*: it arrives downstream as an explicit null
that any consumer can filter on. A **wrong aerodrome** corrupts *two* movement
counts, the one it invents and the one it omits, and it is *invisible*: nothing
downstream can distinguish it from a correct answer. Two errors for one, plus
undetectability, is the case for $k = 2$.

That is an argument, not a measurement, and rankings do move with $k$. So every
headline choice on this page is also scored at $k = 1$ and $k = 3$, and where
the ranking changes, the page says so.

### Out-of-area, and why the headline numbers are decomposed

Roughly one flight in six in the sample genuinely begins or ends outside the
observed area. Such a flight has no nameable aerodrome, and the correct answer
is the literal label `OOA`. This creates a way to score well for a bad reason:
a method that labels liberally out-of-area collects those flights for free.

So each side is also reported three further ways — out-of-area recall and
precision, which show whether the free wins are being taken honestly, and
in-area coverage and accuracy, which restrict to flights whose truth is a real
aerodrome and are the numbers to read when comparing naming rules.

```{r metrics-table}
#| echo: false
if (!is.null(grid)) {
  kable(data.frame(
    Quantity = c("Coverage", "Accuracy", "Correct", "Wrong", "Score",
                 "OOA recall", "OOA precision",
                 "In-area coverage", "In-area accuracy"),
    Definition = c("$a/N$", "$c/a$", "$c$", "$w = a - c$", "$c - k\\,w$",
                   "OOA hits / OOA in truth", "OOA hits / OOA predicted",
                   "answered / truth is a real aerodrome",
                   "correct / answered, truth a real aerodrome"),
    `Reads as` = c("willingness to commit", "reliability when committed",
                   "flights named right", "flights named wrong",
                   "net value at exchange rate $k$",
                   "does it recognise what it cannot name?",
                   "is it over-using the OOA label?",
                   "commitment with the free wins removed",
                   "reliability with the free wins removed"),
    check.names = FALSE
  ), caption = "The reported quantities.")
}
```

## The sample {#sec-sample}

```{r sample}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  cat(sprintf(
    "The 2025 period is 5-7 June 2025, three consecutive days, giving **%s ground-truth flights** after alignment with the Network Manager reference.\n",
    num(N_GT)))
}
```

Ground truth is the EUROCONTROL Network Manager flight table, joined to ADS-B
tracks on aircraft address, callsign and date. The join key is
`AIRCRAFT_ADDRESS`, which is the `icao24` of the ADS-B message.

## The three methods {#sec-methods}

All three name an aerodrome by choosing among candidates, and two of the three
choose the same way, so the shared machinery is worth stating once:

**Candidate selection.** Every aerodrome whose H3 detection zone the aircraft
passed through is a candidate. Each candidate's exact great-circle distance to
the track's closest approach is computed, and the smallest wins. No candidate
is eliminated before its distance is measured.

**The scheduled-service penalty.** An aerodrome without scheduled commercial
service has a fixed distance — 10 NM as shipped — added to its effective
distance before that comparison, so a military or general-aviation field must
be *clearly* nearest rather than merely nearest. It re-ranks candidates without
changing how many flights are answered, so it can move accuracy and cannot move
coverage. It applies to **both** `trend` and `endpoint`.

Where the methods differ is the evidence they read and when they refuse to
answer.

### The steps

Every box in the diagrams below is one of these.

**S1 — Read the track.** One aircraft's state vectors for one flight, already
split into a track by step 02 and cleaned by step 02a.

**S2 — Smooth the altitude.** A centred five-sample rolling mean over
barometric altitude, computed over the *whole* track before any altitude cut.
Smoothing after the cut would truncate the window exactly where altitude is
changing fastest, which is the worst place to lose half of it.

**S3 — Pre-filter on pressure altitude.** A cheap, deliberately loose cut whose
only job is to keep the aerodrome join affordable. Its ceiling is the
configured cap *plus the highest field elevation in the reference*, so it
cannot discard a sample that the exact test at S6 would have kept.

**S4 — Join the aerodrome zones.** Each sample is matched to candidate
aerodromes through an H3 index at resolution 7, then narrowed to those within
the detection radius by exact great-circle distance.

**S5 — Attach the field elevation.** Each candidate aerodrome's elevation, from
OurAirports. A candidate whose elevation is unknown gets zero, which reduces the
next step to the sea-level behaviour rather than dropping the aerodrome.

**S6 — Cut on altitude.** *This is the datum-sensitive step.* Samples
above the ceiling are excluded from the vote. `trend` now measures that ceiling
**above field elevation**; before this version it was a flight level.

**S7 — Vote.** Each surviving sample votes on the sign of its smoothed altitude
change: rising for a departure, falling for an arrival.

**S8 — Decide, or abstain.** If one direction leads the other by more than the
margin, the track is departing or arriving. Otherwise `trend` says nothing.

**E1 — Take the endpoint fix.** `endpoint` reads one sample, not many: the
track's first fix for a departure, its last for an arrival.

**E2 — Test the gates.** The fix must be within the radius, and either flagged
on the ground or no higher than the height gate **above field elevation**.

**E3 — Rank the survivors.** Nearest wins, on exact great-circle distance, plus
a penalty for aerodromes without scheduled service so that a military or
general-aviation field must be *clearly* nearest rather than merely nearest.

**E4 — Or say where it came from.** Where the gates fail and the fix sits at
the edge of the observed area, `endpoint` reports **out-of-area** rather than
guessing. Precedence is aerodrome first, border second: Ponta Delgada lies
about 8 NM inside the western edge, and the other order labelled its departures
out-of-area with the aircraft still on the runway.


### `trend` — the algorithm published before 2026

For every state vector below an altitude ceiling and within a radius of an
aerodrome, `trend` asks whether smoothed barometric altitude is rising or
falling. Each vector casts a vote. If climbing votes exceed descending votes by
more than a margin, that aerodrome is the departure; if the reverse, the
destination. Among the aerodromes that survive that test, the nearest wins.

That ceiling is a **height above the candidate aerodrome's own field
elevation**, and that is what version 6.2 changed. Every published OPDI flight
list before it cut on a flight level instead — a height above the standard
pressure surface, and so above a datum with no relation to the aerodrome being
tested. @sec-datum reports what moving it was worth, and @sec-methods above
marks the affected step as **S6** in each diagram.

Its abstention is evidential: it stays silent when the altitude trace near an
aerodrome does not clearly rise or fall. It has **no out-of-area concept** — it
names an aerodrome or says nothing — which @sec-limitations returns to.

Three of its four constants had never been swept before this study: the
altitude ceiling, fixed at FL40; the detection radius, fixed at 30 NM; and the
datum that ceiling is measured against, which had never been treated as a
choice at all. The scheduled-service penalty described above is new to `trend`
in version 6; before it, `trend` ranked on raw distance alone.

### `trend`, arrivals

```{r fig-trend-arrival}
#| echo: false
#| fig-cap: "`trend` naming an arrival aerodrome (ADES). S6 is where the two altitude datums differ."
#| fig-width: 6.5
#| fig-height: 8.2
flow_diagram(
  steps = c("S1 · track<br>state vectors",
            "S2 · smooth baro_altitude<br>centred 5-sample mean",
            "S3 · pre-filter<br>cap + highest field elevation",
            "S4 · join H3 aerodrome zones<br>res 7, then exact distance",
            "S5 · attach field elevation<br>OurAirports; unknown -> 0",
            "S6 · altitude cut<br>height above field <= cap",
            "S7 · vote per sample<br>sign of smoothed change"),
  emphasis = 6,
  decision = "S8 · descending votes<br>lead by the margin?",
  yes = "E3 · rank survivors<br>ADES = nearest",
  no  = "undetermined")
```

### `trend`, departures

The same machinery reading the same samples; only the direction of the winning
vote differs. `trend` has **no out-of-area concept** — it names an aerodrome or
says nothing — which is why the shipped configuration does not use it for
departures.

```{r fig-trend-departure}
#| echo: false
#| fig-cap: "`trend` naming a departure aerodrome (ADEP). The same machinery; only the winning vote direction differs."
#| fig-width: 6.5
#| fig-height: 8.2
flow_diagram(
  steps = c("S1 · track<br>state vectors",
            "S2 · smooth baro_altitude<br>centred 5-sample mean",
            "S3 · pre-filter<br>cap + highest field elevation",
            "S4 · join H3 aerodrome zones<br>res 7, then exact distance",
            "S5 · attach field elevation<br>OurAirports; unknown -> 0",
            "S6 · altitude cut<br>height above field <= cap",
            "S7 · vote per sample<br>sign of smoothed change"),
  emphasis = 6,
  decision = "S8 · climbing votes<br>lead by the margin?",
  yes = "E3 · rank survivors<br>ADEP = nearest",
  no  = "undetermined<br>never out-of-area")
```

### `nearest` — the *traffic* library's rule

The reference implementation in the `traffic` library takes the aerodrome
nearest the first and last point of the trajectory, unconditionally. It never
abstains, so its coverage is total and its accuracy is whatever the geometry
gives. It is included as the external baseline.

### `endpoint` — `nearest`, with abstention

`endpoint` is `nearest` plus two gates: the endpoint must lie within a radius
of the aerodrome, and either on the ground or below a height above the field —
above *field elevation*, not above sea level, since a fixed altitude cut-off
means nothing at an aerodrome sitting at 5,000 ft.

Failing both gates, the method abstains rather than guessing. Its abstention is
geometric where `trend`'s is evidential. Unlike `trend` it can also say
**out-of-area**: when the failing endpoint sits at the edge of the ingestion
region rather than near any aerodrome, the flight came from or went to
somewhere outside the observed area, and it says so.

### `endpoint`, departures

```{r fig-endpoint-departure}
#| echo: false
#| fig-cap: "`endpoint` naming a departure aerodrome (ADEP). E2 already measured height above field, and still does."
#| fig-width: 7
#| fig-height: 6.5
flow_diagram(
  steps = c("S1 · track state vectors",
            "E1 · take the FIRST fix",
            "S4 · candidate aerodromes<br>within the radius",
            "S5 · attach field elevation<br>OurAirports"),
  emphasis = 4,
  decision = "E2 · within radius AND<br>(on ground OR<br>height above field <= gate)?",
  yes = "E3 · rank survivors<br>ADEP = nearest",
  no  = list(q = "E4 · fix at the edge<br>of the observed area?",
             yes = "out-of-area", no = "undetermined"))
```

### `endpoint`, arrivals

```{r fig-endpoint-arrival}
#| echo: false
#| fig-cap: "`endpoint` naming an arrival aerodrome (ADES). Identical to the departure flow but reading the track's last fix."
#| fig-width: 7
#| fig-height: 6.5
flow_diagram(
  steps = c("S1 · track state vectors",
            "E1 · take the LAST fix",
            "S4 · candidate aerodromes<br>within the radius",
            "S5 · attach field elevation<br>OurAirports"),
  emphasis = 4,
  decision = "E2 · within radius AND<br>(on ground OR<br>height above field <= gate)?",
  yes = "E3 · rank survivors<br>ADES = nearest",
  no  = list(q = "E4 · fix at the edge<br>of the observed area?",
             yes = "out-of-area", no = "undetermined"))
```

### Where the two datums part company

Only **S6** and **E2** read an altitude, and until this version they read it
against different references:

| Test | Threshold | Datum, before v6.2 | Datum, from v6.2 |
|---|---|---|---|
| `trend` S6 | `trend_max_height_ft` | flight level (sea level) | **above field elevation** |
| `endpoint` E2 | `endpoint_height_ft` | above field elevation | above field elevation |
| step 04 ground phase | `phase_ground_ceiling_ft` | above field elevation | above field elevation |
| `nearest` | *none* | — | — |

`nearest` applies no altitude condition at all, on any datum. That is
deliberate: it is the naive `traffic`-faithful baseline the other methods are
scored against, and giving it a gate would defeat its purpose.

::: {.callout-important}
## A flight level is not a height

`FL60` is not the same cut as `6,000 ft`. The pipeline derives the flight level
with an integer cast, so `flight_level <= 60` admits everything below **6,100
ft**. Every datum comparison in this report therefore holds the ceiling at
6,100 ft above field, not 6,000 — otherwise it would move the ceiling and the
datum at the same time and could not attribute the result to either.

The same arithmetic is why the sweep grid in @sec-trend-sweep carries 6,100 ft
rather than 6,000: that is the above-field ceiling admitting exactly the band
`FL60` admits.
:::

## Trend: the full parameter sweep {#sec-trend-sweep}

```{r trend-sweep-summary}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  s0 <- trend[trend$stage == 0, ]
  cat("### The untuned baseline\n\n")
  for (role in c("adep", "ades")) {
    x <- with_metrics(s0, role)
    cat(sprintf(
      "- **%s** at the legacy setting (%s, margin %g, %g NM, no penalty): coverage %s, accuracy %s, %s correct, %s wrong, score %s.\n",
      toupper(role), cap_label(s0$fl_cap[1], datum_of(trend)),
      s0$margin[1], s0$radius_nm[1],
      pct(x$coverage), pct(x$accuracy), num(x$correct), num(x$wrong), num(x$score)))
  }
}
```

```{r trend-best}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  cat("\n### The best setting found\n\n")
  for (role in c("adep", "ades")) {
    b <- best_row(trend, role)
    s0 <- with_metrics(trend[trend$stage == 0, ], role)
    cat(sprintf(
      "- **%s**: %s, margin %g, %g NM, penalty %g NM — coverage %s, accuracy %s, %s correct, %s wrong, score %s (%s against the untuned baseline).\n",
      toupper(role), cap_label(b$fl_cap, datum_of(trend)),
      b$margin, b$radius_nm, b$penalty_nm,
      pct(b$coverage), pct(b$accuracy), num(b$correct), num(b$wrong),
      num(b$score), sgn(b$score - s0$score)))
  }
}
```

### The full trend grid

`trend` has three geometric parameters, so the sweep is a cube and cannot be
printed as one table. It is sliced instead: for each role, score by
flight-level cap against radius at that role's best vote margin, then the
margin's own profile, then the penalty. Every cell of the cube is in
`data/trend_sweep_2025.csv`.

```{r trend-grid}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  s1 <- trend[trend$stage == 1, ]
  for (role in c("adep", "ades")) {
    b <- best_row(s1, role)
    x <- with_metrics(s1[s1$margin == b$margin, ], role)
    w <- reshape(x[, c("fl_cap", "radius_nm", "score")],
                 idvar = "fl_cap", timevar = "radius_nm", direction = "wide")
    w <- w[order(w$fl_cap), ]
    names(w) <- sub("^score\\.", "", names(w))
    w[[1]] <- vapply(w[[1]], cap_label, "", datum = datum_of(trend))
    names(w)[1] <- "Ceiling"
    for (j in 2:ncol(w)) w[[j]] <- format(w[[j]], big.mark = ",", trim = TRUE)
    cat(sprintf("\n**%s — score at $k=2$, by altitude ceiling (rows) and radius in NM (columns), at the winning vote margin of %g.** Ceilings are %s.\n\n",
                toupper(role), b$margin,
                if (identical(datum_of(trend), "field"))
                  "heights above field elevation" else "flight levels"))
    print(kable(w, row.names = FALSE, align = "r"))
    cat("\n")
  }
}
```

```{r trend-margin}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  s1 <- trend[trend$stage == 1, ]
  cat("\n**The vote margin**, at each role's winning cap and radius:\n\n")
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(role) {
    b <- best_row(s1, role)
    x <- with_metrics(
      s1[s1$fl_cap == b$fl_cap & s1$radius_nm == b$radius_nm, ], role)
    x <- x[order(x$margin), ]
    data.frame(Role = toupper(role), Margin = x$margin,
               Coverage = pct(x$coverage), Accuracy = pct(x$accuracy),
               Correct = num(x$correct), Wrong = num(x$wrong),
               Score = num(x$score), check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
}
```

```{r trend-penalty}
#| echo: false
#| results: asis
if (!is.null(trend) && any(trend$stage == 2)) {
  cat("\n**The scheduled-service penalty applied to `trend`**, which the legacy algorithm does not do at all, swept at each role's winning geometry:\n\n")
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(role) {
    b <- best_row(trend[trend$stage == 1, ], role)
    sel <- (trend$stage == 2 & trend$stage2_role == role) |
           (trend$stage == 1 & trend$fl_cap == b$fl_cap &
            trend$margin == b$margin & trend$radius_nm == b$radius_nm)
    x <- with_metrics(trend[sel, ], role)
    x <- x[order(x$penalty_nm), ]
    data.frame(Role = toupper(role), `Penalty (NM)` = x$penalty_nm,
               Coverage = pct(x$coverage), Accuracy = pct(x$accuracy),
               Correct = num(x$correct), Wrong = num(x$wrong),
               Score = num(x$score), check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
  cat("\nCoverage is constant down each block and only accuracy moves, which is the signature of a pure re-ranking: the penalty changes which aerodrome is named, never whether one is.\n")
}
```

## The altitude datum {#sec-datum}

`trend`'s altitude cut is the one threshold in this study that had never been
measured against the aerodrome it applies to. Until version 6.2 it read a
flight level — a height above the standard pressure surface — while `endpoint`,
the ground phase in step 04, and every other altitude gate in the pipeline read
a height above the aerodrome's own field elevation.

That is a defensible thing to measure and an indefensible thing to leave
unexamined, because the two agree exactly at sea level and diverge by the field
elevation everywhere else. At Madrid the same nominal ceiling sits 2,000 ft
lower above the runway than it does at Amsterdam; at an Anatolian aerodrome,
5,000 ft lower.

This chapter reports what happened when the cut was moved onto the datum every
other gate already used. The short answer is that it costs a little accuracy,
buys a little coverage, and is adopted for a reason that is not either of those
numbers — @sec-adoption.


```{r results-guard}
#| echo: false
#| results: asis
swap <- read_if("datum_swap_2025.csv")
bands <- read_if("elevation_bands_2025.csv")
if (is.null(swap)) {
  cat("::: {.callout-warning}\n## Not yet measured\n\n",
      "The arms defined in `regenerate_v62.py` have not been run against this ",
      "checkout, so this section is deliberately empty rather than populated ",
      "with numbers carried over from another study.\n:::\n", sep = "")
}
```

### The two sweeps are one measurement {#sec-equivalence}

Everything in this chapter compares an above-field curve against a sea-level
curve. That comparison is only worth reading if the two are the same
measurement differing in one respect, so this is checked rather than assumed.

They come out of a single vote cache. `trend_sweep_agl.py` counts each track's
votes at every ceiling on **both** datums in one pass, so the two curves are two
readings of one table rather than two studies that happen to share a grid. Its
parameter grid is identical to the sweep version 6 published — same flight-level
caps, same vote margins, same radii, same penalties — with the above-field
ceilings added alongside.

The consequence is testable: the sea-level arm should reproduce version 6's
committed sweep exactly. It does.

```{r sweep-equivalence}
#| echo: false
#| results: asis
eq <- read_if("sweep_equivalence.csv")
if (is.null(eq)) {
  cat("*(the equivalence check has not been run against this checkout)*\n")
} else {
  kable(data.frame(
    Period          = eq$period,
    `Shared cells`  = num(eq$shared_cells),
    `Only in v6`    = num(eq$only_in_old),
    `Only in v6.2`  = num(eq$only_in_new),
    `Differing values` = num(eq$differing_values),
    check.names = FALSE
  ))
  if (all(eq$differing_values == 0)) {
    cat(sprintf(
      "\nAcross %s grid cells on each period, **not one value differs**. That is what licensed retiring version 6's two `trend_sweep` jobs rather than re-running them: they would have recomputed numbers this study already produces. It is also evidence in its own right — extending the harness to a second datum did not perturb the measurement it was extending.\n",
      num(eq$shared_cells[1])))
  } else {
    cat("\n::: {.callout-important}\n## The retirement was wrong\n\nCells differ, so the sea-level arm is *not* version 6's sweep and this report must not serve version 6's numbers from it.\n:::\n")
  }
}
```

### The band breakdown, read first {#sec-bands}

This section comes before the headline deliberately. The headline is a single
pair of numbers and it is consistent with several explanations; the band
breakdown is the only part of this study that can distinguish between them. Read
in the other order, it becomes a search for confirmation.

The prediction the design committed to, before any of this was run: **no
movement at sea-level aerodromes, and gains concentrated at elevated ones.**

```{r bands-table}
#| echo: false
if (!is.null(bands)) {
  b <- bands[bands$role == "arrivals",
             c("band", "aerodromes", "n_truth", "n_correct_msl",
               "n_correct_field", "delta_correct", "delta_correct_loo",
               "delta_correct_ex_busiest")]
  b$relative <- sprintf("%+.2f%%", 100 * b$delta_correct / pmax(b$n_correct_msl, 1))
  names(b) <- c("Band", "Aerodromes", "Truth", "Correct MSL", "Correct field",
                "Δ", "Δ less top mover", "Δ less busiest", "Δ relative")
  kable(b, row.names = FALSE, format.args = list(big.mark = ","),
        caption = "Arrivals, 2025 sample. Δ is field minus sea level.")
}
```

**The effect is where it was predicted to be.** Measured against each band's own
base of correct arrivals — which is the honest denominator, since the bands
differ in size by two orders of magnitude — the sea-level control band and the
band where the change should bite hardest move by very different amounts.

```{r band-effect}
#| echo: false
#| results: asis
bb <- read_if("elevation_bands_2025.csv")
if (!is.null(bb)) {
  arr <- bb[bb$role == "arrivals", ]
  rel <- function(band) {
    r <- arr[arr$band == band, ]
    if (!nrow(r) || r$n_correct_msl[1] == 0) return(NA_real_)
    r$delta_correct[1] / r$n_correct_msl[1]
  }
  lo <- rel("<500"); hi <- rel("1500-3000")
  cat(sprintf("The `<500 ft` control band moves by %s and the `1,500-3,000 ft` band by **%s**",
              pct(lo), pct(hi)))
  if (!is.na(lo) && !is.na(hi) && lo > 0) {
    cat(sprintf(" -- a difference of roughly %.0f-fold between the band where the datum change is arithmetically a no-op and the band where it bites hardest.\n", hi / lo))
  } else {
    cat(".\n")
  }
}
```

It also survives both robustness checks. The census run before this study began
showed each treatment band resting on a single aerodrome, so a band mean alone
could not separate an elevation effect from a Madrid effect. Removing the
largest mover, and removing the busiest aerodrome in the band, both leave it
substantially intact.

```{r band-robustness}
#| echo: false
#| results: asis
bb <- read_if("elevation_bands_2025.csv")
if (!is.null(bb)) {
  r <- bb[bb$role == "arrivals" & bb$band == "1500-3000", ]
  if (nrow(r)) {
    cat(sprintf("Removing the largest mover leaves %s of the %s; removing %s, the busiest field in the band, leaves %s%s. The result is not one aerodrome's result.\n",
                sgn(r$delta_correct_loo[1]), sgn(r$delta_correct[1]),
                r$busiest[1], sgn(r$delta_correct_ex_busiest[1]),
                if (r$delta_correct_ex_busiest[1] > r$delta_correct[1])
                  " -- very slightly *more*" else ""))
  }
}
```

```{r departures-control}
#| echo: false
#| results: asis
if (!is.null(bands)) {
  dep <- bands[bands$role == "departures", ]
  if (all(dep$delta_correct == 0)) {
    cat("::: {.callout-note}\n## The control that had to hold, and did\n\n",
        "**Departures move by exactly zero in every band.** The shipped ",
        "configuration takes departures from `endpoint`, which never reads ",
        "`trend`'s altitude cut — so a non-zero number anywhere in that row ",
        "would have meant this change was touching something it has no ",
        "business touching. It is the cheapest available check that the ",
        "intervention is the one described.\n:::\n", sep = "")
  }
}
```

### What it costs

```{r swap-table}
#| echo: false
if (!is.null(swap)) {
  s <- swap[swap$run %in% c("legacy", "datum_msl", "datum_field"),
            c("run", "ades_coverage", "ades_accuracy", "ades_correct",
              "ades_wrong", "ades_score")]
  s$ades_coverage <- sprintf("%.2f%%", 100 * s$ades_coverage)
  s$ades_accuracy <- sprintf("%.2f%%", 100 * s$ades_accuracy)
  names(s) <- c("Run", "Coverage", "Accuracy", "Correct", "Wrong", "Score (k=2)")
  kable(s, row.names = FALSE, format.args = list(big.mark = ","),
        caption = "Arrivals through `process_dai` itself, 2025 sample.")
}
```

```{r datum-cost-prose}
#| echo: false
#| results: asis
if (!is.null(swap)) {
  g <- function(run, f) swap[[f]][swap$run == run][1]
  dc <- g("datum_field", "ades_correct") - g("datum_msl", "ades_correct")
  dw <- g("datum_field", "ades_wrong")   - g("datum_msl", "ades_wrong")
  dcov <- g("datum_field", "ades_coverage") - g("datum_msl", "ades_coverage")
  dacc <- g("datum_field", "ades_accuracy") - g("datum_msl", "ades_accuracy")
  s_msl <- g("datum_msl", "ades_score"); s_fld <- g("datum_field", "ades_score")
  cat(sprintf("Moving to the field datum buys **%s more correct arrivals and %s more wrong ones**. Coverage %s %s; accuracy %s %s; the score at $k=2$ moves from %s to %s — %s points %s. In net terms it is close to a wash, tilted toward answering more flights and being fractionally less certain about each.\n",
              num(dc), num(dw),
              if (dcov >= 0) "rises" else "falls", sub("^[+-]", "", pp(abs(dcov))),
              if (dacc >= 0) "rises" else "falls", sub("^[+-]", "", pp(abs(dacc))),
              num(s_msl), num(s_fld), num(abs(s_fld - s_msl)),
              if (s_fld >= s_msl) "the right way" else "the wrong way"))
}
```

### Why it ships anyway {#sec-adoption}

The adoption criteria fixed in the design were: coverage up by at least 0.5 pp,
accuracy not falling, and the gain concentrated at elevated fields. **The third
holds and the first two do not.** On a cost-benefit reading, that would be the
end of it.

**The change is adopted regardless**, and the reason is worth stating plainly
because it is not a benchmark result.

A ceiling measured from sea level is *not the same test at different
aerodromes*. At Amsterdam it admits 6,000 ft of climb and descent; at Ankara,
2,975; at Erzurum, roughly 240. Nobody chose that gradient — it is an artefact
of which datum the threshold happened to be written against, and it makes the
method's willingness to answer depend on the terrain under the aerodrome rather
than on the evidence in the trajectory. `endpoint` was corrected on exactly this
reasoning in version 6, and step 04's ground detection in the same release.
`trend` was the last one left.

So this is a **correctness fix, not an optimisation**, and it is not the
benchmark's place to veto it. What the benchmark is for is telling us what the
fix costs, which it has done: a third of a percentage point of coverage gained,
a sixth of a point of accuracy given up, and the movement concentrated where the
mechanism predicts. That is the honest accounting, and it is reported here
rather than buried because a change adopted on principle still has to declare
its price.

The measured null in criteria 1 and 2 does carry one design implication: the
ceiling itself, inherited from a sweep conducted on the datum now abandoned, is
very likely no longer optimal. @sec-limitations returns to that.

### The band that should have gained most, and did not

The `>3000` band is the one where a fixed sea-level ceiling ought to hurt worst,
and it shows no change at all.

```{r high-band}
#| echo: false
#| results: asis
bb <- read_if("elevation_bands_2025.csv")
if (!is.null(bb)) {
  r <- bb[bb$role == "arrivals" & bb$band == ">3000", ]
  if (nrow(r)) {
    cat(sprintf("**%s correct arrivals out of %s, on both datums** -- a delta of %s.\n",
                num(r$n_correct_field[1]), num(r$n_truth[1]),
                sgn(r$delta_correct[1])))
  }
}
```

That is not a null result about the datum. It is `trend` failing at those
aerodromes for a different reason:

```{r high-band-recall}
#| echo: false
#| results: asis
bb <- read_if("elevation_bands_2025.csv")
if (!is.null(bb)) {
  r <- bb[bb$role == "arrivals" & bb$band == ">3000", ]
  if (nrow(r) && r$n_truth[1] > 0) {
    cat(sprintf("a recall of %s is not a method being mis-tuned, it is a method receiving almost nothing to work with.\n",
                pct(r$n_correct_field[1] / r$n_truth[1])))
  }
}
```

The likeliest
cause is ADS-B coverage over the Anatolian plateau, where most of this band's
traffic sits; an aerodrome the feed barely sees cannot be named on any datum.

This matters for how the change should be argued elsewhere. The motivating
example in the design was Erzurum at 5,763 ft, where a FL60 cut leaves about
240 ft in which to vote. That example is arithmetically correct and doubly
unrepresentative: such fields are rare in the sample, **and** they sit in the
one band where the method is independently broken. The population that actually
benefits is the far less dramatic 1,500–3,000 ft — Madrid, Ankara, Sofia,
Tbilisi — where a FL60 cut still leaves three to four thousand feet of usable
climb and descent.

### The ceiling survives the move {#sec-ceiling}

FL60 was tuned by version 6 *on the datum version 6.2 abandons*, so there was
no reason its optimum should have survived the move. The two sweeps, read out
of the same vote cache, settle whether it did.

```{r ceiling-sweeps}
#| echo: false
hs <- read_if("height_sweep_2025.csv")
fs <- read_if("fl_sweep_2025.csv")
shipped <- function(d) d[d$radius_nm == 30 & d$margin == 0 & d$penalty_nm == 10, ]
if (!is.null(hs) && !is.null(fs)) {
  h <- shipped(hs); f <- shipped(fs)
  out <- rbind(
    data.frame(Datum = "above field", Ceiling = sprintf("%s ft", format(h$fl_cap, big.mark = ",")),
               Coverage = sprintf("%.2f%%", 100 * h$ades_coverage),
               Accuracy = sprintf("%.2f%%", 100 * h$ades_accuracy),
               Score = h$ades_score),
    data.frame(Datum = "flight level", Ceiling = sprintf("FL%s", f$fl_cap),
               Coverage = sprintf("%.2f%%", 100 * f$ades_coverage),
               Accuracy = sprintf("%.2f%%", 100 * f$ades_accuracy),
               Score = f$ades_score)
  )
  kable(out, row.names = FALSE, format.args = list(big.mark = ","),
        caption = "Arrivals, shipped geometry (30 NM, margin 0, penalty 10 NM). Both curves swept out of the same vote cache.")
}
```

**Each datum peaks at the equivalent ceiling** — 6,100 ft above field, FL60 on
sea level — and 6,100 ft wins at *every* radius from 20 NM outward, production's
30 NM included. So the ceiling survived the change of datum: it is tuned on the
new datum rather than inherited from the old one.

::: {.callout-note}
## What this does not say

It does not say the shipped ceiling is optimal. `trend_max_height_ft` ships at
**6,000 ft**, and 6,000 ft is not on this grid — the above-field caps were
chosen to bracket 6,100, the height equivalent of FL60, because that is what
the *datum* comparison had to hold fixed to isolate one variable. A sweep that
never evaluated the shipped value cannot endorse it.

The two are 100 ft apart and the difference is very unlikely to matter, but
"very unlikely to matter" is a prediction, not a measurement. @sec-pipeline
carries both through `process_dai` and reports which the pipeline prefers.
:::

Two honest limits. The grid is coarse near the peak (4,000 → 6,100 → 8,000), so
the true optimum could sit anywhere in that span; and these are harness numbers,
which can differ from the pipeline's in level even when they agree in
shape -- the harness re-implements the vote rule over a cached table, and
@sec-pipeline measures the gap directly rather than assuming it away. What the sweep establishes is the *shape*: an interior optimum, falling
away on both sides, at the value already shipping.

### And it replicates on the second period {#sec-second-period}

```{r second-period}
#| echo: false
h24 <- read_if("height_sweep_2024.csv")
f24 <- read_if("fl_sweep_2024.csv")
if (!is.null(h24) && !is.null(f24)) {
  h <- shipped(h24)
  kable(
    data.frame(
      Ceiling  = sprintf("%s ft", format(h$fl_cap, big.mark = ",")),
      Coverage = sprintf("%.2f%%", 100 * h$ades_coverage),
      Accuracy = sprintf("%.2f%%", 100 * h$ades_accuracy),
      Score    = h$ades_score
    ),
    row.names = FALSE, format.args = list(big.mark = ","),
    caption = "2024 sample, above-field datum, arrivals at the shipped geometry."
  )
}
```

**6,100 ft is the optimum on 2024 as well**, at every radius from 20 NM outward
— the same value, the same interior peak, on an independent sample a year
apart. A ceiling that survives that is tuned, not fitted.

The datum comparison replicates too, including its sign.

```{r datum-replication}
#| echo: false
#| results: asis
gap <- function(h, f) {
  if (is.null(h) || is.null(f)) return(NA_real_)
  bh <- max(h$ades_score[h$stage == 1]); bf <- max(f$ades_score[f$stage == 1])
  (bh - bf) / bf
}
g25 <- gap(trend, trend_msl)
g24 <- gap(t24, t24_msl)
if (!is.na(g25) && !is.na(g24)) {
  cat(sprintf("The best above-field cell scores %s the best sea-level cell on 2025 and %s on 2024.\n",
              sprintf("%.2f%% %s", abs(100 * g25), if (g25 < 0) "below" else "above"),
              sprintf("%.2f%% %s", abs(100 * g24), if (g24 < 0) "below" else "above")))
}
```
Both are fractions of a percent, and both are diluted for the reason
@sec-dilution gives. What matters is that the second period tells the same story
rather than a different one — the finding is the algorithm's and not the
sample's.

::: {.callout-note}

### Why the second period is sweeps only

The 2024 tracks pre-date this study's H3 indexing, and `process_dai` reads
`h3_res_7` off the track table directly, so the pipeline arms cannot run here at
all. Version 6 met the same wall and ran the pipeline on 2025 alone. The sweeps get
around it because their vote cache is built with the index computed at read
time.

So the second period confirms the *ceiling* and the *datum comparison*, which is
what it is able to confirm. It does not independently reproduce the band
breakdown, and this report does not claim that it does.
:::

### Why the aggregate was never going to show much {#sec-dilution}

```{r dilution-levels}
#| echo: false
#| results: asis
hs <- read_if("height_sweep_2025.csv"); fs <- read_if("fl_sweep_2025.csv")
if (!is.null(hs) && !is.null(fs)) {
  at <- function(d) d[d$radius_nm == 30 & d$margin == 0 & d$penalty_nm == 10, ]
  h <- at(hs); f <- at(fs)
  best <- function(d) max(d$ades_score[d$stage == 1])
  cat(sprintf("The two curves are also nearly identical in level — %s against %s at the shipped geometry, and %s against %s at each datum's own best cell. That looks like a change doing nothing, and it is worth being precise about why it is not.\n",
              num(h$ades_score[1]), num(f$ades_score[1]),
              num(best(hs)), num(best(fs))))
}
```

```{r dilution-share}
#| echo: false
#| results: asis
bb <- read_if("elevation_bands_2025.csv")
if (!is.null(bb)) {
  arr <- bb[bb$role == "arrivals", ]
  lo <- arr$n_truth[arr$band == "<500"] / sum(arr$n_truth)
  cat(sprintf("**%s of the ground-truth arrivals sit in the `<500 ft` band**, where the two datums are the same test by construction.\n",
              pct(lo)))
} else {
  cat("**Most of the ground-truth arrivals sit in the `<500 ft` band**, where the two datums are the same test by construction.\n")
}
```

A change that
cannot move the majority of the sample cannot move an average over it, however
well it works on the rest. The aggregate score is diluted by design, and judging
a targeted correctness fix by it is using the wrong instrument — the reading is
dominated by the aerodromes the fix was never about.

That is precisely why @sec-bands exists and why it is read first. The signal is
in the 1,500–3,000 ft band, and the aggregate is where that signal goes to be
averaged away.

## Endpoint: the full parameter sweep {#sec-endpoint-sweep}

`endpoint` has two geometric parameters. The **radius** is how close to an
aerodrome the trajectory's first or last point must be; the **height** is how
far above field elevation it may be. Failing either, the method abstains.

The grid below is the whole sweep, scored for each role. The height row labelled
*unrestricted* removes the height gate entirely, which collapses `endpoint`
into `nearest` within the radius — so the last row of each table is the cost of
having no abstention at all.

```{r endpoint-grid}
#| echo: false
#| results: asis
if (!is.null(grid)) {
  UNRESTRICTED <- max(grid$height_ft)
  hlab <- function(h) ifelse(h == UNRESTRICTED, "unrestricted",
                             format(h, big.mark = ",", trim = TRUE, scientific = FALSE))
  for (role in c("adep", "ades")) {
    x <- with_metrics(grid, role)
    w <- reshape(x[, c("radius_nm", "height_ft", "score")],
                 idvar = "height_ft", timevar = "radius_nm", direction = "wide")
    w <- w[order(w$height_ft), ]
    names(w) <- sub("^score\\.", "", names(w))
    names(w)[1] <- "height (ft)"
    w[[1]] <- hlab(w[[1]])
    for (j in 2:ncol(w)) w[[j]] <- format(w[[j]], big.mark = ",", trim = TRUE)
    cat(sprintf("\n**%s — score at $k=2$, by radius (NM, columns) and height (rows)**\n\n",
                toupper(role)))
    print(kable(w, row.names = FALSE, align = "r"))
    cat("\n")
  }
}
```

```{r endpoint-best}
#| echo: false
#| results: asis
if (!is.null(grid)) {
  cat("\n### Where each role's optimum sits\n\n")
  for (role in c("adep", "ades")) {
    b <- best_row(grid, role)
    cat(sprintf(
      "- **%s**: %g NM, %s ft — coverage %s, accuracy %s, %s correct, %s wrong, score %s.\n",
      toupper(role), b$radius_nm,
      format(b$height_ft, big.mark = ",", trim = TRUE, scientific = FALSE),
      pct(b$coverage), pct(b$accuracy), num(b$correct), num(b$wrong), num(b$score)))
  }
  cat("\nBoth optima are interior to the grid in both parameters, so neither is an artefact of where the sweep stopped.\n")
}
```

## The scheduled-service penalty {#sec-penalty}

Both methods choose among surviving candidates by distance. Near a major
aerodrome that often means choosing between the airport itself and a small
airfield a few miles away that almost never sees commercial traffic. The
penalty adds a fixed distance to every candidate without scheduled service,
biasing the tie-break toward the aerodrome more likely to be the real one.

It is a pure re-ranking: it changes *which* aerodrome is named, never *whether*
one is named. So coverage is flat across the whole sweep and only accuracy
moves — which is exactly what makes it cheap.

```{r penalty-table}
#| echo: false
#| results: asis
if (!is.null(pen)) {
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(role) {
    x <- with_metrics(pen[order(pen$penalty_nm), ], role)
    data.frame(Role = toupper(role), `Penalty (NM)` = x$penalty_nm,
               Coverage = pct(x$coverage), Accuracy = pct(x$accuracy),
               Correct = num(x$correct), Wrong = num(x$wrong),
               Score = num(x$score), check.names = FALSE)
  }))
  cat(sprintf("\nEndpoint, at %g NM and %s ft:\n\n",
              pen$radius_nm[1],
              format(pen$height_ft[1], big.mark = ",", trim = TRUE, scientific = FALSE)))
  print(kable(rows, row.names = FALSE, align = "r"))
}
```

## How much the exchange rate matters {#sec-k}

$k$ is a judgement, so the honest thing is to show what it decides. The table
below re-runs the argmax at $k = 1$, $2$ and $3$ for both methods and both
roles.

```{r k-sensitivity}
#| echo: false
#| results: asis
if (!is.null(trend) && !is.null(grid)) {
  rows <- do.call(rbind, lapply(c(1, 2, 3), function(k) {
    do.call(rbind, lapply(c("adep", "ades"), function(role) {
      bt <- best_row(trend, role, k); be <- best_row(grid, role, k)
      data.frame(
        k = k, Role = toupper(role),
        `Best trend` = sprintf("%s, margin %g, %g NM, pen %g",
                               cap_label(bt$fl_cap, datum_of(trend)),
                               bt$margin, bt$radius_nm, bt$penalty_nm),
        `trend score` = num(bt$score),
        `Best endpoint` = sprintf("%g NM, %s ft, pen %g", be$radius_nm,
                                  format(be$height_ft, big.mark = ",", trim = TRUE,
                                         scientific = FALSE), be$penalty_nm),
        `endpoint score` = num(be$score),
        Winner = ifelse(be$score >= bt$score, "endpoint", "trend"),
        check.names = FALSE)
    }))
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
}
```

```{r k-verdict}
#| echo: false
#| results: asis
if (!is.null(trend) && !is.null(grid)) {
  winners <- sapply(c("adep", "ades"), function(role)
    sapply(c(1, 2, 3), function(k) {
      ifelse(best_row(grid, role, k)$score >= best_row(trend, role, k)$score,
             "endpoint", "trend")
    }))
  stable <- apply(winners, 2, function(w) length(unique(w)) == 1)
  if (all(stable)) {
    cat(sprintf("\nThe **choice of method is the same at every $k$ tried**: departures want `%s` and arrivals want `%s` whether a wrong answer is priced at one silence, two or three. That is the most robust finding on this page, and it does not depend on the judgement that fixes $k$.\n",
                winners[1, "adep"], winners[1, "ades"]))
  } else {
    cat("\nThe choice of method **changes with $k$**, so it rests on the pricing judgement and not on the data alone. The verdict below says which value it assumes.\n")
  }
  cat("\nThe parameter *values* are less stable than the method choice, and in a direction that makes sense: a higher $k$ prices wrong answers more dearly, so the argmax moves toward tighter gates that answer less often and are right more often when they do. Where a headline value moves with $k$, the justification table records it.\n")
}
```

## Negative results, re-measured {#sec-negative}

Ideas that were tried and did not work. All are re-run here against current
code rather than quoted from the version that first reported them.

### The approach cone

The box rule treats radius and height as independent, and they are not: an
aircraft descending toward an aerodrome trades height for distance at roughly a
constant rate. A 3° glideslope is 318 ft/NM and the "three times the flight
level" rule of thumb about 333 ft/NM, so the endpoints that genuinely belong to
an aerodrome should lie in a *cone*, not a box.

The two shapes disagree in both corners. Close and high is inside the box and
outside the cone — an aircraft at 15,000 ft overhead is overflying, not
arriving. Far and low is outside the box and inside the cone — an arrival whose
reception dies at 8,000 ft and 30 NM is on profile.

Replacing the box with a cone is worse for both roles, at every slope swept.

```{r cone-result}
#| echo: false
#| results: asis
if (!is.null(cone) && !is.null(grid)) {
  # The two roles fail differently, so how each fails is derived from the
  # counts rather than stated once and assumed to cover both. An earlier draft
  # of this passage asserted that the cone "admits far more wrong answers",
  # which the arrival row flatly contradicts.
  for (role in c("adep", "ades")) {
    bc <- best_row(cone, role); bb <- best_row(grid, role)
    dc <- bc$correct - bb$correct; dw <- bc$wrong - bb$wrong
    how <- if (dc <= 0 && dw >= 0) {
      "worse on both counts — fewer correct *and* more wrong, so no exchange rate rescues it"
    } else if (dc < 0 && dw < 0) {
      sprintf("more restrictive, giving up %s correct to avoid %s wrong — a trade worth taking only if a wrong answer cost more than %.0f silences",
              num(-dc), num(-dw), -dc / -dw)
    } else {
      sprintf("more permissive, buying %s correct at the price of %s wrong",
              num(dc), num(dw))
    }
    cat(sprintf(
      "- **%s**: best cone (%g ft/NM, %g NM) scores %s against the box's %s, %s. It is %s.\n",
      toupper(role), bc$slope_ft_nm, bc$radius_nm, num(bc$score), num(bb$score),
      sgn(bc$score - bb$score), how))
  }
  cat("\nThe cone loses for both roles at every slope swept, and the reason is the one that sank every other rate-based idea in this study: the OpenSky archive carries positions forward between updates, so a *position* is trustworthy whenever it appears, while anything derived by dividing by elapsed time is not. A glideslope is a rate in disguise.\n")
}
```

### The bearing test

Version 5 proposed asking whether the trajectory *points at* the aerodrome: take
the track's course near its endpoint, take the bearing from that endpoint to the
candidate, and require the two to agree within some angle. It was the first
signal found that could add answers the abstention refused, and it was proposed
in four forms — **rescue** (answer anyway when aligned), **veto** (require
alignment even when the gate accepts), **replace** (alignment instead of the
gate) and **rerank** (choose *which* aerodrome by alignment).

All four are re-measured here against the current `endpoint` baseline.

```{r bearing}
#| echo: false
#| results: asis
bw <- cache("bearing_whole_sample_v6.csv")
if (!is.null(bw)) {
  n <- bw$n_ground_truth[1]
  sc <- function(role) {
    ans <- round(bw[[paste0(role, "_coverage")]] * n)
    cor <- round(bw[[paste0(role, "_overall")]] * n)
    cor - K_DEFAULT * (ans - cor)
  }
  bw$adep_s <- sc("adep"); bw$ades_s <- sc("ades")
  b <- bw[bw$model == "base", ]
  print(kable(data.frame(
    Variant = bw$model,
    `ADEP coverage` = pct(bw$adep_coverage), `ADEP accuracy` = pct(bw$adep_accuracy),
    `ADEP score` = num(bw$adep_s), `Δ` = c("—", sgn(bw$adep_s[-1] - b$adep_s)),
    `ADES coverage` = pct(bw$ades_coverage), `ADES accuracy` = pct(bw$ades_accuracy),
    `ADES score` = num(bw$ades_s), `Δ ` = c("—", sgn(bw$ades_s[-1] - b$ades_s)),
    check.names = FALSE), row.names = FALSE, align = "r"))

  best <- bw[which.max(bw$ades_s), ]
  cat(sprintf("
The best variant of any kind is `%s`, worth %s on arrivals — and %s on departures, which is the role `endpoint` is actually used for. Nothing here is worth adopting.
",
              best$model, sgn(best$ades_s - b$ades_s), sgn(best$adep_s - b$adep_s)))
  cat(sprintf("
The three variants that let bearing *remove* or *choose* answers are not close: the best of them gives up %s on departures. Alignment is far too blunt to name an aerodrome — at 40 NM a runway subtends a couple of degrees, but so does every other aerodrome behind it on the same radial, and a track on final approach points at all of them equally.
",
              sgn(max(bw$adep_s[grepl("^(veto|replace|rerank)", bw$model)]) - b$adep_s)))
  cat("
This supersedes version 5's recommendation. That version scored the bearing rescue against `endpoint` alone and never against `trend`, which was already better than both — the comparison that mattered was the one not made.
")
}
```

### Bearing applied to `trend`

The bearing variants above all modify `endpoint`. `trend` is the algorithm that
ships for arrivals, and it fails differently — it has no gate, so it almost
always answers, and its errors are misnamings rather than silences. The variant
that suits it is therefore a *rerank* or a *tie-break*, not a rescue, and its
candidate set is already narrowed by the vote rule, so alignment is asked a
much easier question than on the endpoint side.

```{r trend-bearing}
#| echo: false
#| results: asis
tb <- cache("trend_bearing_v6.csv")
if (is.null(tb)) {
  cat("*Not yet regenerated.*\n")
} else {
  n <- tb$n_ground_truth[1]
  sc <- function(role) {
    ans <- round(tb[[paste0(role, "_coverage")]] * n)
    cor <- round(tb[[paste0(role, "_overall")]] * n)
    cor - K_DEFAULT * (ans - cor)
  }
  tb$adep_s <- sc("adep"); tb$ades_s <- sc("ades")
  b <- tb[tb$model == "base: trend as shipped", ]
  if (nrow(b) == 0) b <- tb[1, ]
  print(kable(data.frame(
    Variant = tb$model,
    `ADES coverage` = pct(tb$ades_coverage), `ADES accuracy` = pct(tb$ades_accuracy),
    `ADES score` = num(tb$ades_s),
    `Δ vs shipped` = ifelse(tb$model == b$model, "—", sgn(tb$ades_s - b$ades_s)),
    check.names = FALSE), row.names = FALSE, align = "r"))
  best <- tb[which.max(tb$ades_s), ]
  if (best$ades_s > b$ades_s) {
    cat(sprintf("\n**%s improves on the shipped configuration by %s.** That is a real gain and it is not currently adopted -- it arrived after the defaults were set, and adopting it would mean shipping a value this study has measured once, on one period. It is the first thing a follow-up should confirm.\n",
                best$model, sgn(best$ades_s - b$ades_s)))
  } else {
    cat("\n**Nothing improves on the shipped configuration.** Alignment cannot separate trend's candidates either: the aerodromes it confuses lie in much the same direction from the endpoint as the right one, which is precisely why the vote rule confused them. The bearing test is now closed for both methods.\n")
  }
}
```

### The vertical measures

Whether a track is climbing or descending can be read from the broadcast
vertical rate, from a two-point altitude slope, or by least squares over the
whole window. The three disagree, and version 5 could not say which was wrong.

```{r vertical}
#| echo: false
#| results: asis
vm <- cache("vertical_measure_v6.csv")
if (is.null(vm)) {
  cat("*Not yet regenerated.*\n")
} else {
  print(kable(data.frame(
    `Window (min)` = vm$window_min, Freshness = vm$freshness,
    Flights = num(vm$flights),
    `Broadcast rate vs OLS` = pct(vm$vr_vs_ols),
    `2-point slope vs OLS` = pct(vm$t2pt_vs_ols),
    check.names = FALSE), row.names = FALSE, align = "r"))
  cat(sprintf("\nAcross the sample the broadcast rate disagrees with the least-squares slope on %s of cases and the two-point slope on %s; %s of broadcast rates are exactly zero, and %s of samples are fresh. The disagreement is concentrated where position updates are stale, which is the same explanation that sank the approach cone: a position survives the archive's carry-forward, a rate does not.\n",
              pct(vm$overall_vr_vs_ols[1]), pct(vm$overall_t2pt_vs_ols[1]),
              pct(vm$vr_zero_share[1]), pct(vm$fresh_share[1])))
}
```

## How the sample is thinned {#sec-sampler}

This is not a negative result — it is a property of the data every other number
on this page is computed from, and it is a change to the pipeline that will
matter most somewhere else.

OPDI thins the 1 Hz feed to roughly 5 s. The rule this study inherited is
fixed-phase — `event_time % 5 == 0`, which keeps a row only if one exists at the
single second per window congruent to zero. The rule it now uses keeps the
newest row in each 5 s bin, and cannot drop a sample where the modulo rule
would.

The bin-based rule is adopted, and **not because it measures better**. It is
the sampler the thinning was always meant to be; the modulo rule survives only
because the OpenSky archive is a complete 1 Hz grid with carried-forward
positions, so a row almost always exists at the congruent second. That is a
property of the archive, not a guarantee. If it ever changes, the modulo rule
degrades badly and this one degrades gracefully.

For ADEP/ADES detection it should make almost no difference, and it does not:
both methods read an altitude *trend* over tens of samples, and one extra fix
per bin cannot change the sign of a trend. Where it should matter is the
milestones defined by a single instant on the ground — off-block, take-off,
landing, in-block — where a recovered bin is the difference between having a
fix and not. That measurement belongs to the events study and has not been
made here.

What it costs is the question this page can answer, and the answer is: nothing
measurable. Each row below compares two complete runs across every **cell** —
one cell being one parameter combination — so a difference that appears at a
single setting and nowhere else cannot be mistaken for a sampler effect.

```{r sampler}
#| echo: false
#| results: asis
sc <- cache("sampler_comparison_v6.csv")
if (is.null(sc)) {
  cat("*Not yet regenerated.*\n")
} else {
  print(kable(data.frame(
    Comparison = sc$comparison, Role = toupper(sc$role), Cells = sc$cells,
    `Median Δscore` = sprintf("%+.0f", sc$median_dscore),
    `Median Δcoverage` = sprintf("%+.4f pp", sc$median_dcoverage_pp),
    Improved = sc$cells_improved, Worsened = sc$cells_worsened,
    check.names = FALSE), row.names = FALSE, align = "r"))
  worst <- max(abs(sc$median_dcoverage_pp))
  cat(sprintf("\nAcross every parameter cell of two complete runs, the largest median coverage difference in either direction is **%.4f pp**. The trend side is consistently positive — %d of %d arrival cells improve, which is a real if tiny effect — and the endpoint side is consistently slightly negative. The shipped configuration takes departures from `endpoint` and arrivals from `trend`, so the two roughly cancel.\n",
              worst,
              sc$cells_improved[sc$comparison == "trend sweep" & sc$role == "ades"],
              sc$cells[sc$comparison == "trend sweep" & sc$role == "ades"]))
  cat("\nColumn gloss: a **cell** is one parameter combination; **Δ** is the bin-based sampler minus the fixed-phase one; **pp** is percentage points of the flights in the reference.\n")
  cat("\nSo the sampler is **neutral on accuracy and coverage** for this problem, and is adopted on correctness. It should not be described as an improvement to aerodrome detection.\n")
}
```

One consequence has to be stated because it is not reversible by configuration:
**`track_id` values differ** between the two samplers. The rescued rows sit at
track boundaries and the splitter breaks on gaps, so a track that gains a fix
can be split differently. Re-processing a past month with the bin-based rule
will not reproduce the identifiers that were published under the fixed-phase
one.

## What the pipeline actually produces {#sec-pipeline}

Everything measured so far has been measured in two stages, and the distinction
governs how much any number on this page is worth.

**Stage 1 — the research sweep.** Hundreds of parameter combinations applied to
a cached intermediate table: vote counts per track-aerodrome pair for `trend`,
first-and-last-fix candidates for `endpoint`. Cheap enough to be exhaustive,
which is what makes a grid of this size possible at all. But it re-implements
the decision rule rather than running it, so on its own it can only *propose* a
value.

**Stage 2 — the production run.** The proposed values run through
`FlightListProcessor.process_dai` — the same code path that writes the
published flight list — and scored identically. No value in @sec-defaults
became a default without passing this stage.

Two differences between the two stages remain, and both are named here rather
than left to be discovered in the residual:

1. the pipeline filters zones by radius **band** — the H3 ring the zone was
   generated at — while the sweep cuts on exact distance;
2. the sweep smooths barometric altitude across the flight-level boundary and
   the pipeline applies the cut first, so its smoothing window is truncated
   there.

A third difference, and by far the largest, has been **closed**: the pipeline
used to keep only the candidates at the minimum H3 ring count and measure exact
distance only among the survivors, so an aerodrome one ring further out was
eliminated before its distance was ever computed. It now carries every
candidate in the joined disk through to an exact-distance ranking, which is
what the sweep always did. That change is the subject of the rest of this
section.

```{r pipeline-fidelity}
#| echo: false
#| results: asis
if (!is.null(modes) && !is.null(trend) && !is.null(grid)) {
  rows <- do.call(rbind, lapply(seq_len(nrow(modes)), function(i) {
    r <- modes[i, ]
    do.call(rbind, lapply(c("adep", "ades"), function(role) data.frame(
      Run = r$run, Role = toupper(role),
      Mode = r[[paste0(role, "_mode")]],
      Coverage = pct(r[[paste0(role, "_coverage")]]),
      Accuracy = pct(r[[paste0(role, "_accuracy")]]),
      Correct  = num(r[[paste0(role, "_correct")]]),
      Wrong    = num(r[[paste0(role, "_wrong")]]),
      Score    = num(r[[paste0(role, "_score")]]),
      check.names = FALSE)))
  }))
  print(kable(rows, row.names = FALSE, align = "r"))

  # The harness's own figure for the same setting, so the residual is visible
  # rather than implied.
  cat("\n**Harness against pipeline, at the same parameters:**\n\n")
  cmp <- do.call(rbind, list(
    local({
      h <- best_row(grid, "adep"); p <- modes[modes$run == "endpoint", ]
      data.frame(Role = "ADEP", Method = "endpoint",
                 `Harness score` = num(h$score),
                 `Pipeline score` = num(p$adep_score),
                 Residual = sgn(p$adep_score - h$score),
                 `Relative` = sprintf("%+.2f%%", 100 * (p$adep_score - h$score) / h$score),
                 check.names = FALSE)
    }),
    local({
      # From the `combined` run, not the `trend` run. A DetectionConfig holds
      # one trend setting, so the single-mode `trend` run has both roles at the
      # *departure* optimum and its arrival figure describes a setting nobody
      # recommends. `combined` is where trend runs at the arrival optimum.
      h <- best_row(trend[trend$stage == 1, ], "ades"); p <- modes[modes$run == "combined", ]
      data.frame(Role = "ADES", Method = "trend",
                 `Harness score` = num(h$score),
                 `Pipeline score` = num(p$ades_score),
                 Residual = sgn(p$ades_score - h$score),
                 `Relative` = sprintf("%+.2f%%", 100 * (p$ades_score - h$score) / h$score),
                 check.names = FALSE)
    })))
  print(kable(cmp, row.names = FALSE, align = "r"))
  cat("\nThe two rows could hardly be more different, and the difference is the single most important result on this page.\n")
  cat("\n`endpoint` **reproduces exactly**. That is not luck: the endpoint sweep filters the same cached candidate table the pipeline reads, applying the same rule, so harness and pipeline are the same computation and agree to the flight.\n")
  cat("\n`trend` **does not**, and the reason is the ranking rule. Both stages now rank candidate aerodromes on exact great-circle distance: the full set of aerodromes whose detection zone the aircraft passed through is carried forward, each candidate's own closest approach is computed, and the scheduled-service penalty is applied to that distance. Nothing is eliminated before it is measured.\n")
  cat("\nUntil this study that was true of the sweep only. The pipeline kept just the candidates at the minimum H3 *ring count* — an integer stepping about 5.2 km at resolution 7 — and applied the tie-break, and the penalty, within that surviving set alone. An aerodrome one ring further out was gone before the penalty could promote it, so a penalty tuned against exact distance did something weaker in production. The residual above is what remains once that rule is the same on both sides: the radius band and the truncated smoothing window, and nothing else.\n")
}
```

### The trend tuning re-walked inside the pipeline

The path below runs through `process_dai` itself. Each row **adds one parameter
change to the row above it** and rebuilds the whole flight list with everything
adopted so far, so row *n* differs from row *n−1* in exactly one value. It
answers which of the changes is doing the work, not which configuration to
ship — for that see @sec-verdict.

Both roles are produced by `trend` in every row, at the arrival tuning, because
that is the parameter set being walked. The departure columns are therefore a
diagnostic and not a recommendation: what ships takes departures from
`endpoint`.

It was run twice — once with the pipeline's original **H3 ring-count**
selection, and once with **exact-distance** ranking. The two disagree in sign,
which is the whole finding.

```{r pipeline-path}
#| echo: false
#| results: asis
ppath <- cache("pipeline_path_v6.csv")
if (!is.null(ppath)) {
  # Rungs 0-4 walk the SEA-LEVEL tuning, so they are the same ladder version 6
  # published and are directly comparable to it. Rung 5 moves the datum and
  # nothing else, at the ceiling that admits the identical band -- which is
  # what makes its delta attributable to the datum rather than to a ceiling
  # that moved along with it.
  lab <- c(path0_legacy = "legacy constants",
           path1_penalty = "+ scheduled-service penalty",
           path2_ceiling = "+ altitude ceiling",
           path3_margin = "+ vote margin",
           path4_radius = "+ radius",
           path5_datum = "+ field-elevation datum")
  ppath <- ppath[order(match(ppath$run, names(lab))), ]
  sc <- ppath$ades_score

  # The ring-ranked run, read from its own CSV rather than typed in. An earlier
  # draft of this chunk carried these five numbers as R literals -- on a page
  # whose whole argument is that hand-copied figures cannot be trusted.
  rp <- cache("pipeline_path_ring_v6.csv")
  rp <- rp[order(match(rp$run, names(lab))), ]
  ring <- rp$ades_score

  print(kable(data.frame(
    Step = lab[ppath$run],
    `Ring: score` = num(ring),
    `Ring: gain` = c("—", sgn(diff(ring))),
    `Exact: score` = num(sc),
    `Exact: gain` = c("—", sgn(diff(sc))),
    Coverage = pct(ppath$ades_coverage), Accuracy = pct(ppath$ades_accuracy),
    check.names = FALSE), row.names = FALSE, align = "r"))

  cat(sprintf("\nUnder ring selection the tuning is worth %s; under exact distance it is worth %s. The altitude ceiling is the step that flips hardest — %s against %s — and it is the step that admits samples from higher up, where more aerodromes share a ring and the ring rule is least able to separate them.\n",
              sgn(ring[length(ring)] - ring[1]), sgn(sc[length(sc)] - sc[1]),
              sgn(diff(ring)[2]), sgn(diff(sc)[2])))

  # The fifth rung, reported separately: it is the only step that moves the
  # datum, so it is the only one whose delta is the datum's own contribution
  # inside the pipeline.
  if ("path5_datum" %in% ppath$run && length(sc) >= 6) {
    cat(sprintf("\nThe fifth rung is the datum, and it is the reason this version exists. Moving `trend`'s altitude cut from a flight level onto the aerodrome's own field elevation — at the ceiling admitting the identical band, so nothing else moves with it — is worth %s under exact distance and %s under ring selection. @sec-datum takes that apart by field elevation, which is where the number stops being a single figure and starts being an explanation.\n",
                sgn(diff(sc)[5]), sgn(diff(ring)[5])))
  }
  cat("\nSo the sweep harness was never over-estimating. It was measuring a *better algorithm* — one that measures every candidate before choosing — and the pipeline has now adopted it. `trend_rank_by` defaults to `\"haversine\"`; `DetectionConfig.legacy()` still returns `\"ring\"`, so every published flight list stays reproducible.\n")
}
```

### The pipeline's own optimum

The path adopts the values the *sweep* preferred. Whether those are also the
**pipeline's** optima is a separate question, and it is answered by sweeping
the altitude ceiling and radius through `process_dai` directly — on the datum
that ships, so the ceilings below are feet above field elevation.

Both roles are produced by `trend` in every cell of this grid. Held fixed
throughout: vote margin 2, scheduled-service penalty 10 NM, ranking on exact
distance. The vote margin is a `trend` parameter only — the majority one
direction must win by across the smoothed altitude samples before a track is
called a departure or an arrival. `endpoint` has no votes; it reads one fix.

The grid runs from 3,000 ft to 10,000 ft above field, which is far enough to
find the arrival optimum and **not** far enough to close the departure side —
see @sec-limitations. It carries both 6,000 ft and 6,100 ft deliberately: the
first is what ships and the second is what the research sweep preferred, and
the sweep grid never contained the shipped value, so the two had never been
compared until this table.

```{r pipeline-trend-grid}
#| echo: false
#| results: asis
tg <- cache("trend_grid_v6.csv")
if (!is.null(tg)) {
  # `trend_ceiling` means the right thing on either datum: feet above field
  # when the datum is `field`, a flight level when it is `msl`. Keying on
  # `trend_max_fl` would collapse every row of a height grid onto one cap,
  # because that field is inert while the field datum is in force.
  CEIL <- if ("trend_ceiling" %in% names(tg)) "trend_ceiling" else "trend_max_fl"
  dtm <- if ("trend_max_datum" %in% names(tg)) unique(tg$trend_max_datum)[1] else "msl"
  tg <- tg[order(tg[[CEIL]], tg$trend_radius_nm), ]
  wide <- function(col) {
    w <- reshape(tg[, c(CEIL, "trend_radius_nm", col)],
                 idvar = CEIL, timevar = "trend_radius_nm",
                 direction = "wide")
    w <- w[order(w[[CEIL]]), ]
    names(w) <- sub(paste0("^", col, "\\."), "", names(w))
    w
  }
  a <- wide("ades_score"); d <- wide("adep_score")
  tab <- data.frame(Ceiling = vapply(a[[CEIL]], cap_label, "", datum = dtm),
                    `ADES 20 NM` = num(a[["20"]]), `ADES 30 NM` = num(a[["30"]]),
                    `ADEP 20 NM` = num(d[["20"]]), `ADEP 30 NM` = num(d[["30"]]),
                    check.names = FALSE)
  cat(sprintf("\n**Score by altitude ceiling (rows) and radius in NM (columns), through the pipeline, at vote margin 2 and penalty 10 NM.** Ceilings are %s. Departures are shown for completeness -- they are served by `endpoint`, so that half is a check rather than a recommendation.\n\n",
              if (identical(dtm, "field")) "heights above field elevation" else "flight levels"))
  print(kable(tab, row.names = FALSE, align = "r"))

  b <- tg[which.max(tg$ades_score), ]
  bd <- tg[which.max(tg$adep_score), ]
  cat(sprintf("\nArrivals peak at **%s / %g NM**. Departures climb to **%s**, but departures are served by `endpoint`, so that column is a check rather than a recommendation.\n",
              cap_label(b[[CEIL]], dtm), b$trend_radius_nm,
              cap_label(bd[[CEIL]], dtm)))

  # Whether the arrival optimum is interior is a property of this grid, not a
  # standing fact -- it was one in version 6 and the grid has since changed.
  # Asserting it in prose would be asserting the previous grid's shape.
  caps <- sort(unique(tg[[CEIL]]))
  if (length(caps) > 2) {
    interior <- b[[CEIL]] > min(caps) && b[[CEIL]] < max(caps)
    cat(sprintf("\nThat arrival peak is %s of the grid, so it %s.\n",
                if (interior) "interior, with the score falling away on both sides"
                else "at an edge",
                if (interior) "is an optimum rather than the best cell that happened to be tried"
                else "may not be an optimum: the grid should be widened before the value is trusted"))
  }
  # Which radius each role prefers, read off the grid rather than asserted.
  # Version 6 stated that both roles prefer the tighter radius at every cell,
  # and cited that as the stronger evidence for adopting 20 NM. On this datum
  # it is not true for arrivals, so it is computed here instead of repeated.
  radii <- sort(unique(tg$trend_radius_nm))
  if (length(radii) == 2) {
    tight <- radii[1]; wide <- radii[2]
    pref <- function(role) {
      caps <- sort(unique(tg[[CEIL]]))
      vapply(caps, function(c) {
        a <- tg[[paste0(role, "_score")]][tg[[CEIL]] == c & tg$trend_radius_nm == tight]
        b <- tg[[paste0(role, "_score")]][tg[[CEIL]] == c & tg$trend_radius_nm == wide]
        if (!length(a) || !length(b)) NA_real_ else a[1] - b[1]
      }, numeric(1))
    }
    da <- pref("ades"); dd <- pref("adep")
    caps <- sort(unique(tg[[CEIL]]))
    cat(sprintf("\n**Which radius wins depends on the ceiling, and on the role.** Departures prefer the tighter %g NM at %d of %d ceilings; arrivals prefer it at only %d, and at every ceiling from %s upward they prefer the wider %g NM — which includes the ceiling that ships.\n",
                tight, sum(dd > 0, na.rm = TRUE), length(caps),
                sum(da > 0, na.rm = TRUE),
                cap_label(min(caps[which(da < 0)]), dtm), wide))
    cat(sprintf("\nThat matters because version 6 reported the opposite — that both roles prefer the tighter radius at every cell — and cited it as the stronger evidence for lowering `trend_radius_nm` to %g NM. On the datum that ships, the grid does not support that for arrivals, which are the role `trend` actually serves. The value was never applied to `DetectionConfig` regardless, and this study does not renew the recommendation.\n",
                tight))
  }

  cat("\nOne caveat, stated rather than buried: the grid holds the **vote margin** fixed at 2 while varying the ceiling and radius. That value therefore rests on the sweep harness alone, and is the weakest-supported of the shipped parameters.\n")
}
```

## Why each value, and not its neighbour {#sec-justification}

A recommendation is only as good as the evidence that picked it, so every value
the sweeps select appears below with the score at that value, the score at the
adjacent grid points, and the score at the legacy value. The last
column is the one that matters most for trusting the number: a value on a broad
**plateau** is a safe recommendation even from a single sample, because its
neighbours score nearly the same and a little noise cannot move it far. A value
on a sharp isolated **peak** may be partly the sample's noise, and is flagged
as such however good its score.

::: {.callout-important}
## These are the sweeps' choices, not the shipped ones
The `Chosen` column below is the **harness argmax**, which is what this section
exists to scrutinise. For `trend`, @sec-pipeline is what settles whether each
value survives in the pipeline — and under exact-distance ranking they do, which
is why the harness numbers below and the shipped values in @sec-defaults now
agree. They are still kept apart, because for a period they did *not* agree,
and a sweep preference is only ever evidence until the pipeline confirms it.
:::

```{r justification}
#| echo: false
#| results: asis

# The score at one grid point, holding every other swept parameter at the
# chosen value. This is what makes the neighbour columns meaningful: they vary
# one parameter at a time rather than reporting the best cell that happens to
# contain that value.
at <- function(d, role, fixed, param, value, k = K_DEFAULT) {
  sel <- rep(TRUE, nrow(d))
  for (nm in names(fixed)) if (nm != param) sel <- sel & d[[nm]] == fixed[[nm]]
  sel <- sel & d[[param]] == value
  x <- with_metrics(d[sel, , drop = FALSE], role, k)
  if (is.null(x) || nrow(x) == 0) return(NA_real_)
  x$score[1]
}

# One row of the table: chosen value, its two neighbours on the grid, and the
# shape that follows from them.
just_row <- function(d, role, chosen, param, label, orig_value, k = K_DEFAULT,
                     fixed_names) {
  fixed <- as.list(chosen[fixed_names])
  vals  <- sort(unique(d[[param]]))
  i     <- match(chosen[[param]], vals)
  s_ch  <- at(d, role, fixed, param, vals[i], k)
  s_lo  <- if (i > 1) at(d, role, fixed, param, vals[i - 1], k) else NA_real_
  s_hi  <- if (i < length(vals)) at(d, role, fixed, param, vals[i + 1], k) else NA_real_
  s_or  <- at(d, role, fixed, param, orig_value, k)

  # Plateau if the better neighbour is within 0.5% of the chosen score --
  # a margin far smaller than any effect the report argues from.
  nb   <- c(s_lo, s_hi); nb <- nb[!is.na(nb)]
  gap  <- if (length(nb)) (s_ch - max(nb)) / s_ch else NA_real_
  shape <- if (is.na(gap)) "—" else if (gap < 0.005) "plateau" else "peak"

  data.frame(
    Role = toupper(role), Parameter = label,
    Legacy = orig_value, `Sweep argmax` = chosen[[param]],
    `Score at legacy` = ifelse(is.na(s_or), "—", num(s_or)),
    `Score at argmax`     = num(s_ch),
    `Neighbour below`   = ifelse(is.na(s_lo), "—", num(s_lo)),
    `Neighbour above`   = ifelse(is.na(s_hi), "—", num(s_hi)),
    `Δ vs legacy`   = ifelse(is.na(s_or), "—", sgn(s_ch - s_or)),
    Shape = shape, check.names = FALSE)
}

if (!is.null(trend) && !is.null(grid)) {
  s1 <- trend[trend$stage == 1, ]
  tf <- c("fl_cap", "margin", "radius_nm")
  ef <- c("radius_nm", "height_ft")

  rows <- list()
  # trend, for the role it is recommended for (arrivals), against the
  # production constants: FL40, margin 4, 30 NM.
  bt <- best_row(s1, "ades")
  rows[[length(rows) + 1]] <- just_row(s1, "ades", bt, "fl_cap",    "trend ceiling",
                                     if (identical(datum_of(trend), "msl")) 40 else 4100,
                                     fixed_names = tf)
  rows[[length(rows) + 1]] <- just_row(s1, "ades", bt, "margin",    "trend vote margin",   4, fixed_names = tf)
  rows[[length(rows) + 1]] <- just_row(s1, "ades", bt, "radius_nm", "trend radius (NM)",  30, fixed_names = tf)

  # endpoint, for departures, against the version 4.5 operating point that the
  # pipeline currently defaults to: 40 NM, 15,000 ft.
  be <- best_row(grid, "adep")
  rows[[length(rows) + 1]] <- just_row(grid, "adep", be, "radius_nm", "endpoint radius (NM)",    40, fixed_names = ef)
  rows[[length(rows) + 1]] <- just_row(grid, "adep", be, "height_ft", "endpoint height (ft)", 15000, fixed_names = ef)

  print(kable(do.call(rbind, rows), row.names = FALSE, align = "r"))
  cat("\nThe penalty is treated separately because it is not a geometric parameter: it re-ranks candidates without changing how many are answered, so its sweep holds the geometry fixed and appears in @sec-penalty.\n")
}
```

### Which change is actually doing the work

The table above varies one parameter at a time around the chosen setting, which
shows whether each value is defensible but not how much each contributes to the
total gain. This walks the whole distance instead: from the legacy constants to
the recommended setting, changing one parameter per step.

**How to read it.** Each row *adds* its change to everything in the rows above
and keeps it — row 5 has all four changes applied, not just the radius. `Gain`
is that row's score minus the previous row's, so it is the marginal worth of
that one change given everything already adopted. All rows are arrivals under
`trend`, in the research sweep.

```{r trend-path}
#| echo: false
#| results: asis
if (!is.null(trend)) {
  s1 <- trend[trend$stage == 1, ]
  b  <- best_row(s1, "ades")
  s0 <- with_metrics(trend[trend$stage == 0, ], "ades")

  # Each step adopts one recommended value and holds the rest at production's.
  # The order is deliberate: the penalty first, because it is the parameter
  # production does not apply at all, and the geometry after.
  cellv <- function(fl, mg, rd, pn) {
    x <- with_metrics(trend[trend$fl_cap == fl & trend$margin == mg &
                            trend$radius_nm == rd & trend$penalty_nm == pn, ],
                      "ades")
    if (is.null(x) || nrow(x) == 0) NA else x$score[1]
  }
  # The legacy ceiling, on whichever datum this sweep was measured.
  #
  # `legacy()` is FL40. On the above-field grid the comparable ceiling is its
  # height equivalent, 4,100 ft — and since the grid does not contain 4,100
  # either, the nearest point stands in. Hardcoding 40 here would make every
  # `cellv` call return NA on a field-datum sweep, and the `ok` filter below
  # would then drop the affected rows *silently*: a shorter ladder that still
  # renders, still sums to 100%, and is missing its first two steps.
  dtm <- datum_of(trend)
  caps <- sort(unique(s1$fl_cap))
  lc <- if (identical(dtm, "msl")) 40 else caps[which.min(abs(caps - 4100))]
  steps <- list(
    list(lab = sprintf("legacy constants (%s, margin 4, 30 NM, no penalty)",
                       cap_label(lc, dtm)),
         s = s0$score),
    list(lab = sprintf("+ scheduled-service penalty of %g NM", b$penalty_nm),
         s = cellv(lc, 4, 30, b$penalty_nm)),
    list(lab = sprintf("+ altitude ceiling %s", cap_label(b$fl_cap, dtm)),
         s = cellv(b$fl_cap, 4, 30, b$penalty_nm)),
    list(lab = sprintf("+ vote margin %g", b$margin),
         s = cellv(b$fl_cap, b$margin, 30, b$penalty_nm)),
    list(lab = sprintf("+ radius %g NM", b$radius_nm),
         s = cellv(b$fl_cap, b$margin, b$radius_nm, b$penalty_nm)))

  ok <- !sapply(steps, function(x) is.na(x$s))
  if (any(!ok)) {
    # Loud, because a dropped rung leaves a ladder that still renders, still
    # sums to 100%, and quietly attributes the missing step's gain to its
    # neighbour.
    cat(sprintf("\n::: {.callout-warning}\n## %d rung(s) of this ladder are missing\n\nThe sweep grid has no cell for: %s. The percentages below are shares of the gain across the rungs that *are* present, not of the whole distance from legacy to recommended.\n:::\n\n",
                sum(!ok),
                paste(sub("^\\+ ", "", sapply(steps[!ok], function(x) x$lab)),
                      collapse = "; ")))
  }
  steps <- steps[ok]
  sc <- sapply(steps, function(x) x$s)
  tab <- data.frame(
    Step = sapply(steps, function(x) x$lab),
    Score = num(sc),
    `Gain from this step` = c("—", sgn(diff(sc))),
    `Share of total gain` = c("—", sprintf("%.0f%%", 100 * diff(sc) / (sc[length(sc)] - sc[1]))),
    check.names = FALSE)
  print(kable(tab, row.names = FALSE, align = "r"))

  dom <- which.max(diff(sc))
  cat(sprintf("\nThe gain is not spread evenly. **%s** accounts for %.0f%% of it on its own. That is worth stating plainly: the single most valuable change to arrival detection is not a tuned threshold at all, but applying to `trend` the scheduled-service tie-break that `endpoint` already had and `trend` never did.\n",
              sub("^\\+ ", "", steps[[dom + 1]]$lab),
              100 * max(diff(sc)) / (sc[length(sc)] - sc[1])))
  cat(sprintf("\nBy contrast the radius step is worth %s flights out of %s here — inside what one three-day sample can resolve, and on this evidence alone it would not justify moving a published constant. Version 6 adopted it anyway, citing the pipeline grid as stronger evidence. On the datum that ships, that grid no longer supports it: see @sec-pipeline, where arrivals prefer the *wider* 30 NM at every ceiling from 6,000 ft up, which is where the shipped ceiling sits. The 20 NM recommendation was never applied to `DetectionConfig` in any case, and this study does not renew it.\n",
              num(sc[length(sc)] - sc[length(sc) - 1]), num(sc[length(sc)])))
}
```

## The verdict {#sec-verdict}

```{r verdict}
#| echo: false
#| results: asis
if (!is.null(modes)) {
  m <- function(run, field) {
    v <- modes[[field]][modes$run == run]
    if (length(v) == 0) NA_real_ else v[1]
  }
  tot <- function(run) m(run, "adep_score") + m(run, "ades_score")
  runs <- modes$run[order(-(modes$adep_score + modes$ades_score))]

  # `recommended` and `combined` are the same configuration built two ways --
  # from DetectionConfig() and by naming the fields -- and they agree field for
  # field. Labelling both makes that visible instead of looking like two
  # different recommendations.
  label <- c(legacy = "legacy — the algorithm published before 2026",
             trend = "trend, departure parameters both roles",
             endpoint = "endpoint, both roles",
             nearest = "nearest (traffic library)",
             combined = "**SHIPPED** endpoint + exact-distance trend",
             recommended = "the same, built from `DetectionConfig()`")
  print(kable(data.frame(
    Configuration = ifelse(is.na(label[modes$run]), modes$run, label[modes$run]),
    ADEP = paste0(modes$adep_mode, ""), ADES = paste0(modes$ades_mode, ""),
    `ADEP coverage` = pct(modes$adep_coverage),
    `ADEP accuracy` = pct(modes$adep_accuracy),
    `ADEP score` = num(modes$adep_score),
    `ADES coverage` = pct(modes$ades_coverage),
    `ADES accuracy` = pct(modes$ades_accuracy),
    `ADES score` = num(modes$ades_score),
    Total = num(modes$adep_score + modes$ades_score),
    check.names = FALSE)[order(-(modes$adep_score + modes$ades_score)), ],
    row.names = FALSE, align = "r"))

  cat("\nEach row is a complete configuration — which method names departures, which names arrivals — not a step in a sequence. The rows are alternatives to choose between.\n")
  cat(sprintf("\nThe recommended configuration is the **%s** row: `endpoint` for departures, exact-distance `trend` for arrivals. It scores **%s against legacy's %s, a gain of %s** on a three-day sample of %s flights, and every figure in that table is a real pipeline run scored against Network Manager ground truth.\n",
              SHIPPED, num(tot(SHIPPED)), num(tot("legacy")),
              sgn(tot(SHIPPED) - tot("legacy")), num(N_GT)))
  cat("\nIt is the shipped configuration in the strict sense: its trend cap, radius, vote margin, penalty and ranking rule, and its endpoint radius, height and penalty, were checked field by field against `DetectionConfig()`. The row is what the pipeline produces, not an approximation of it.\n")
  cat(sprintf("\nThe `recommended` row is the same configuration constructed the other way — by instantiating `DetectionConfig()` rather than by naming each field — and it lands within %s of `%s` in total. Two independent constructions of the shipped settings agreeing to that margin is the check that the recommendation and the code have not drifted apart.\n",
              num(abs(tot("recommended") - tot(SHIPPED))), SHIPPED))
}
```

### What is being recommended, and why

```{r verdict-detail}
#| echo: false
#| results: asis
if (!is.null(modes) && !is.null(grid)) {
  m <- function(run, field) modes[[field]][modes$run == run][1]
  tg <- cache("trend_grid_v6.csv")
  cat(sprintf("**Departures — `endpoint`, %g NM radius, %s ft above field elevation, %g NM scheduled-service penalty.** That height is `endpoint_height_ft`, a height above the aerodrome's own elevation — not a flight level, and not related to the flight-level caps swept for `trend`, which departures do not use. Score %s against legacy's %s, a gain of %s. The clearest result in the study: an interior optimum in both parameters, better than `trend` at every exchange rate tried, and the pipeline reproduces the sweep exactly -- the endpoint sweep filters the same cached candidate table the pipeline reads, so the tuned figure *is* the pipeline figure.\n\n",
              best_row(grid, "adep")$radius_nm,
              format(best_row(grid, "adep")$height_ft, big.mark = ",", trim = TRUE, scientific = FALSE),
              best_row(grid, "adep")$penalty_nm,
              num(m(SHIPPED, "adep_score")), num(m("legacy", "adep_score")),
              sgn(m(SHIPPED, "adep_score") - m("legacy", "adep_score"))))

  if (!is.null(tg)) {
    b <- tg[which.max(tg$ades_score), ]
    cat(sprintf("**Arrivals — `trend` at %s, vote margin %g, %g NM, %g NM penalty, ranked on exact distance.** Measured as an algorithm, on a trend-only flight list, that is %s against legacy's %s -- a gain of %s, won by converting %s silences into correct answers for %s extra wrong ones.\n\n",
                # From the pipeline frame, not the sweep: it carries the datum
                # it actually ran on, so the label cannot disagree with the run.
                cap_label(if (!is.null(b$trend_ceiling)) b$trend_ceiling else b$trend_max_fl,
                          if (!is.null(b$trend_max_datum)) b$trend_max_datum else "msl"),
                b$trend_vote_margin, b$trend_radius_nm,
                b$trend_sched_penalty_nm,
                num(b$ades_score), num(m("legacy", "ades_score")),
                sgn(b$ades_score - m("legacy", "ades_score")),
                num(b$ades_correct - m("legacy", "ades_correct")),
                num(b$ades_wrong - m("legacy", "ades_wrong"))))

    cat(sprintf("Inside the combined flight list the arrival half reads %s, %s against legacy rather than the %s just quoted. The difference is not the algorithm. Taking departures from `endpoint` adds tracks `trend` never produced, and where two share an aircraft, callsign and date, the scoring alignment awards the flight to whichever starts closest to the actual off-block time -- sometimes now a departure-only track carrying no arrival. Accuracy *rises* in the merged table (%s against %s) while coverage falls, which is the signature of lost pairings rather than wrong names.\n\n",
                num(m(SHIPPED, "ades_score")),
                sgn(m(SHIPPED, "ades_score") - m("legacy", "ades_score")),
                sgn(b$ades_score - m("legacy", "ades_score")),
                pct(m(SHIPPED, "ades_accuracy")), pct(m("legacy", "ades_accuracy"))))
  }

  cat("**One weak link, stated plainly.** The vote margin of 2 comes from the sweep harness; the pipeline grid held it fixed while varying the cap and radius. Every other shipped value was measured through `process_dai`. If one parameter in this recommendation is going to fail on a fresh sample, that is the one.\n")
}
```

### The `traffic` library's rule, for reference

```{r verdict-traffic}
#| echo: false
#| results: asis
if (!is.null(modes)) {
  m <- function(run, field) modes[[field]][modes$run == run][1]
  cat(sprintf("`nearest` — the unconditional rule the `traffic` library uses — answers far more often than anything else here (%s of departures, %s of arrivals) and is right far less often (%s and %s). Scored, it is last by a wide margin: %s against the recommendation's %s.\n",
              pct(m("nearest", "adep_coverage")), pct(m("nearest", "ades_coverage")),
              pct(m("nearest", "adep_accuracy")), pct(m("nearest", "ades_accuracy")),
              num(m("nearest", "adep_score") + m("nearest", "ades_score")),
              num(m("recommended", "adep_score") + m("recommended", "ades_score"))))
  cat("\nThat is not a criticism of `traffic`, which is solving a different problem: it names the nearest aerodrome and leaves the judgement to the caller. It is, however, the measurement that justifies OPDI carrying its own rule rather than adopting that one.\n")
}
```

## By aerodrome class {#sec-per-type}

The candidate set is large and medium aerodromes only, so a flight whose true
aerodrome is a small airfield or a heliport cannot be named correctly by any
method on this page. This is the size of that population, and of everything
else.

```{r per-type}
#| echo: false
#| results: asis
if (!is.null(typ)) {
  for (rl in c("departures", "arrivals")) {
    x <- typ[typ$run == SHIPPED & typ$role == rl, ]
    if (nrow(x) == 0) next
    x <- x[order(-x$n_truth), ]
    cat(sprintf("\n**%s**\n\n", tools::toTitleCase(rl)))
    print(kable(data.frame(
      Class = x$apt_type, `Flights in truth` = num(x$n_truth),
      Aerodromes = num(x$n_aerodromes), Answered = num(x$n_answered),
      Correct = num(x$n_correct), Coverage = pct(x$coverage),
      Accuracy = pct(x$accuracy), check.names = FALSE),
      row.names = FALSE, align = "r"))
  }
  unn <- typ[typ$run == SHIPPED &
             !typ$apt_type %in% c("large_airport", "medium_airport"), ]
  if (nrow(unn)) {
    cat(sprintf("\nOutside the candidate set — %s classes, %s flights across both roles — the method is working against a reference it was never given. Those flights are counted in the denominator throughout this page, so every coverage figure here already carries their cost.\n",
                num(length(unique(unn$apt_type))), num(sum(unn$n_truth))))
  }
}
```

## The busiest 100 aerodromes

Movement counts are what OPDI is for, so the last check is per aerodrome rather
than per flight: for each of the hundred busiest in the reference, how many
movements the recommended methodology finds against how many the Network
Manager data says there were.

```{r top100-headline}
#| echo: false
#| results: asis
if (!is.null(apt)) {
  cat(sprintf("Across those hundred, the shipped configuration recovers **%s** of arrivals and **%s** of departures, against the legacy algorithm's %s and %s on the same aerodromes. The aerodrome-by-aerodrome tables are in @sec-top100.\n",
              pct(top100_recall(SHIPPED, "arrivals")),
              pct(top100_recall(SHIPPED, "departures")),
              pct(top100_recall("legacy", "arrivals")),
              pct(top100_recall("legacy", "departures"))))
}
```

## Does any of it hold on a second period? {#sec-stability}

Everything above is one three-day sample. An argmax over hundreds of cells on a
single sample always flatters itself: part of the winning cell's margin is that
sample's noise, and noise does not reproduce. The only way to find out which
part is real is to measure a different period with the same code.

The second period is **5–7 June 2024**, chosen because the Network Manager
reference for that month is committed to the repository and the matching tracks
already exist from earlier versions of this study.

```{r stability}
#| echo: false
#| results: asis
if (is.null(t24)) {
  cat("*The second-period sweep has not been staged yet.*\n")
} else {
  cat(sprintf("The 2024 sample carries %s ground-truth flights against 2025's %s.\n\n",
              num(t24$n_ground_truth[1]), num(N_GT)))

  s1_25 <- trend[trend$stage == 1, ]; s1_24 <- t24[t24$stage == 1, ]
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(role) {
    b25 <- best_row(s1_25, role); b24 <- best_row(s1_24, role)
    do.call(rbind, lapply(list(
      list("trend ceiling", "fl_cap"), list("trend vote margin", "margin"),
      list("trend radius (NM)", "radius_nm")), function(pp_) {
        data.frame(Role = toupper(role), Parameter = pp_[[1]],
                   `2025 optimum` = b25[[pp_[[2]]]],
                   `2024 optimum` = b24[[pp_[[2]]]],
                   Agrees = ifelse(b25[[pp_[[2]]]] == b24[[pp_[[2]]]], "yes", "**no**"),
                   check.names = FALSE)
      }))
  }))
  print(kable(rows, row.names = FALSE, align = "r"))

  agree <- sum(rows$Agrees == "yes")
  cat(sprintf("\n%d of %d parameter optima agree across the two periods — which, taken alone, reads worse than the situation is. Comparing two argmaxes is close to the weakest test available: over hundreds of cells the top cell moves under noise even when the surface underneath is stable.\n",
              agree, nrow(rows)))

  # The question that matters is not whether the argmax moved but whether the
  # *decision* would have changed. Ranking every cell by both periods at once
  # answers that; comparing two argmaxes does not.
  key <- function(d) paste(d$fl_cap, d$margin, d$radius_nm, d$penalty_nm, sep = "/")
  a <- s1_25; b <- s1_24; a$key <- key(a); b$key <- key(b)
  mm <- merge(a, b, by = "key", suffixes = c("_25", "_24"))
  mm$joint <- mm$ades_score_25 / mm$n_ground_truth_25 +
              mm$ades_score_24 / mm$n_ground_truth_24
  o <- mm[order(-mm$joint), ]
  dtm <- datum_of(trend)
  nc <- nearest_cap(o$fl_cap_25, dtm)
  rk <- which(o$fl_cap_25 == nc$cap &
              o$margin_25 == SHIPPED_TREND$margin &
              o$radius_nm_25 == SHIPPED_TREND$radius_nm)

  if (!length(rk)) {
    cat("\n*(the shipped arrival cell is not present in both periods' grids, so its joint rank cannot be reported)*\n")
  } else {
    rk <- rk[1]
    cat(sprintf("\nSo the cells are ranked here by **both periods at once**, with each period's score divided by its own ground-truth count so the larger sample cannot dominate. On that ranking the shipped arrival configuration — %s, vote margin %g, %g NM, %g NM penalty — places **%d of %d**. The joint winner is %s / margin %g / %g NM, and differs by %s of normalised score.\n",
                cap_label(nc$cap, dtm), SHIPPED_TREND$margin,
                SHIPPED_TREND$radius_nm, SHIPPED_TREND$penalty_nm,
                rk, nrow(o),
                cap_label(o$fl_cap_25[1], dtm), o$margin_25[1], o$radius_nm_25[1],
                sprintf("%.5f", o$joint[1] - o$joint[rk])))
    if (!nc$exact) {
      cat(sprintf("\nOne substitution, stated rather than buried: the shipped ceiling is %s, and the research sweep's grid does not contain it — the grid was built around %s, the height equivalent of FL60. So %s stands in above, and the ranking is that cell's rank, not the shipped cell's. The pipeline grid in @sec-pipeline carries both values and is where the two are actually compared.\n",
                  cap_label(nc$want, dtm), cap_label(nc$cap, dtm),
                  cap_label(nc$cap, dtm)))
    }
    cat("\nThe arrival tuning therefore does hold on both periods. What moves between them is which of several near-equivalent cells happens to come first, and the radius is the parameter that moves.\n")
  }
}
```

```{r stability-endpoint}
#| echo: false
#| results: asis
e24 <- cache("sweep_radius_height_2024.csv")
if (is.null(e24)) {
  cat("\n**The endpoint rule was not re-swept on the second period.** Its sweep runs over the cached candidate table, which exists for 2025 and not for 2024; building it for 2024 would mean either materialising an H3 index over the whole second period or reimplementing the pipeline's endpoint extraction in the harness. The second option is precisely the kind of shortcut @sec-pipeline exists to warn about, so neither was taken.\n")
  cat("\nThe departure recommendation therefore rests on a different guarantee, and a stronger one for this particular purpose: it was validated *in the pipeline itself* on 2025, where the sweep and `process_dai` agree to the flight. What is missing is a second sample, not a second implementation. It is listed in the limitations as the outstanding check.\n")
} else if (!is.null(grid)) {
  cat("\nFor the endpoint rule, which is what departures actually adopt:\n\n")
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(role) {
    a <- best_row(grid, role); b <- best_row(e24, role)
    data.frame(Role = toupper(role),
               `2025 radius / height` = sprintf("%g NM / %s ft", a$radius_nm,
                    format(a$height_ft, big.mark = ",", trim = TRUE, scientific = FALSE)),
               `2024 radius / height` = sprintf("%g NM / %s ft", b$radius_nm,
                    format(b$height_ft, big.mark = ",", trim = TRUE, scientific = FALSE)),
               Agrees = ifelse(a$radius_nm == b$radius_nm & a$height_ft == b$height_ft,
                               "**yes**", "no"),
               check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
  cat(sprintf("\nThe departure optimum — the one that actually ships — is the *same cell* on two independent periods, out of %d. Neither period was consulted when the other was measured. That is the strongest stability evidence in this study.\n",
              nrow(e24)))
}
```

## What ships {#sec-defaults}

This table is **read from the runs themselves**, not written out here. The
`legacy` column is the configuration the `legacy` run was built with, and the
`ships` column is the configuration the `recommended` run was built with — and
`recommended` is constructed as a bare `DetectionConfig()`, so it is by
definition whatever the pipeline currently defaults to.

That matters more than it sounds. The equivalent table in version 6 was typed
by hand, and it had drifted: it named two values the code does not have. A
table asserting what ships, on a page whose whole argument is that hand-copied
figures cannot be trusted, is the one place that must not be hand-copied.

```{r what-ships}
#| echo: false
#| results: asis
if (is.null(modes) || !"recommended" %in% modes$run) {
  cat("*(the pipeline runs have not been made against this checkout, so what ships cannot be read from them)*\n")
} else {
  L <- modes[modes$run == "legacy", ][1, ]
  S <- modes[modes$run == "recommended", ][1, ]

  fmtv <- function(v) {
    if (is.null(v) || length(v) == 0 || is.na(v)) return("—")
    if (is.numeric(v)) format(v, big.mark = ",", trim = TRUE, scientific = FALSE)
    else as.character(v)
  }

  # Ordering is deliberate: the ranking rule first, because without it none of
  # the rest holds; the datum second, because it decides what the ceiling
  # underneath it even means.
  spec <- list(
    list("trend_rank_by", "The enabling change. Ring-count selection discarded candidates before measuring them; every tuned trend value fails under it and gains under exact distance."),
    list("trend_max_datum", "@sec-datum. Aligns `trend`'s altitude cut with every other altitude gate in the pipeline."),
    list("trend_max_height_ft", "The ceiling that datum is measured against, and inert while the datum is `msl` — which is why the legacy column shows a value legacy never applied. See @sec-pipeline on 6,000 against 6,100."),
    list("trend_max_fl", "Retained, and inert while the datum is `field`. It is what `legacy()` reads, so published lists stay reproducible."),
    list("trend_radius_nm", "Swept in @sec-trend-sweep and through the pipeline in @sec-pipeline."),
    list("trend_vote_margin", "Sweep harness only — the pipeline grid holds it fixed. The weakest-supported value here."),
    list("trend_sched_penalty_nm", "A tie-break `trend` never had. The single largest contributor to the arrival gain."),
    list("endpoint_radius_nm", "Interior optimum; the pipeline reproduces the sweep exactly."),
    list("endpoint_height_ft", "A height above field elevation, and always was."),
    list("endpoint_sched_penalty_nm", "Unchanged.")
  )

  rows <- do.call(rbind, lapply(spec, function(s) {
    p <- s[[1]]
    if (!p %in% names(modes)) return(NULL)
    lv <- fmtv(L[[p]]); sv <- fmtv(S[[p]])
    data.frame(Parameter = sprintf("`%s`", p),
               Legacy = lv,
               Ships = if (identical(lv, sv)) sv else sprintf("**%s**", sv),
               Evidence = s[[2]], check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE))
  changed <- sum(rows$Legacy != gsub("\\*\\*", "", rows$Ships))
  cat(sprintf("\n%d of the %d parameters differ from the legacy constants.\n",
              changed, nrow(rows)))
}
```

```{r ships-vs-study}
#| echo: false
#| results: asis
# Does what ships match what this study recommends?
#
# `recommended` is a bare DetectionConfig(); `combined` is assembled from the
# sweep optima. Where they differ, the pipeline is running something the study
# did not choose -- which is worth surfacing rather than leaving for a reader
# to notice by comparing two tables.
if (!is.null(modes) && all(c("recommended", "combined") %in% modes$run)) {
  R <- modes[modes$run == "recommended", ][1, ]
  C <- modes[modes$run == "combined", ][1, ]
  pars <- intersect(c("trend_radius_nm", "trend_vote_margin", "trend_ceiling",
                      "trend_sched_penalty_nm", "trend_rank_by"), names(modes))
  diff <- pars[sapply(pars, function(p) !isTRUE(all.equal(R[[p]], C[[p]])))]
  if (length(diff)) {
    cat("::: {.callout-important}\n## What ships is not what the sweeps chose\n\n")
    cat("The shipped configuration and this study's swept optimum differ in",
        length(diff), "parameter(s):\n\n")
    cat("| Parameter | Ships | Sweep optimum |\n|---|---|---|\n")
    for (p in diff)
      cat(sprintf("| `%s` | %s | %s |\n", p,
                  format(R[[p]], trim = TRUE), format(C[[p]], trim = TRUE)))
    cat(sprintf("\nBoth are measured above: `recommended` is the shipped row, `combined` the swept one, and on arrivals they score %s against %s. This is reported rather than resolved — changing a published default is a decision for the pipeline's maintainers, and this page's job is to make the gap visible.\n:::\n",
                num(R$ades_score), num(C$ades_score)))
  }
}
```

```{r ceiling-6000-6100}
#| echo: false
#| results: asis
tgc <- cache("trend_grid_v6.csv")
if (is.null(tgc)) {
  cat("*(the pipeline ceiling grid has not been run against this checkout, so the 6,000 against 6,100 question is not settled here)*\n")
} else {
  capof <- function(run) as.numeric(sub("^grid_h([0-9]+)_.*$", "\\1", run))
  g <- tgc[grepl("^grid_h", tgc$run), ]
  g$cap <- capof(g$run)
  best <- function(cap) {
    x <- g[g$cap == cap, ]
    if (!nrow(x)) return(NA_real_) else max(x$ades_score)
  }
  s6000 <- best(6000)
  s6100 <- best(6100)
  if (!is.na(s6000) && !is.na(s6100)) {
    d <- s6100 - s6000
    cat(sprintf("**6,000 ft or 6,100?** The shipped ceiling is 6,000 ft; the research sweep called 6,100 optimal — on a grid that never contained 6,000, so that recommendation was never a comparison. Run through `process_dai` itself, the best arrival score at 6,000 ft is %s and at 6,100 ft is %s, a difference of %s. %s\n\n",
      num(s6000), num(s6100), sgn(d),
      if (abs(d) < 50)
        "That is inside what one three-day sample resolves, so the shipped 6,000 ft stands: there is no measured reason to move it, and moving a published constant needs one."
      else if (d > 0)
        "The pipeline prefers 6,100 ft, and by more than sampling noise. That is a recommendation to move the shipped ceiling, and it is recorded here rather than acted on inside this study."
      else
        "The pipeline prefers the shipped 6,000 ft, so the sweep's preference for 6,100 does not survive contact with production — the same pattern version 6 found for its flight-level cap."))
  }
}
```

The ceiling and the vote margin were **rejected earlier in this study and
reinstated**. That rejection was measured under ring-count selection, where
raising the ceiling genuinely did lose ground. The lesson is not that the first
measurement was careless but that it answered a narrower question than it
appeared to: *given this selection rule*, the tuning fails. Changing the rule
changed the answer.

The radius is a different story, and version 6 told it wrongly. Its table
recorded `trend_radius_nm` as shipping at 20 NM. The table above — read from a
bare `DetectionConfig()` rather than typed — shows it at **30 NM**, unchanged
from legacy. The recommendation was made and never applied, and no later
version noticed because every later version copied the claim rather than
measuring it.

Every OPDI flight list published before this change was built with the
legacy column, and `DetectionConfig.legacy()` still returns exactly those
values — including `trend_rank_by = "ring"` — so anything already released
stays reproducible by name. That is the same principle as the frozen `version`
strings on published events. A regression test pins the preset, a second pins
the shipped values, and a third asserts the two differ, so the preset cannot
quietly become vacuous.

::: {.callout-warning}
### Is what ships what the pipeline prefers? {#sec-parity}

The question every earlier version answered by assertion. It is answerable
now, and only now, because the grid was widened until it contained the shipped
configuration: it previously fixed the vote margin at 2 while `DetectionConfig`
ships 0, so main's own cell had never been run.

```{r parity}
#| echo: false
#| results: asis
tgp <- cache("trend_grid_v6.csv")
if (is.null(tgp)) {
  cat("*(the pipeline grid has not been run against this checkout)*\n")
} else {
  g <- tgp[grepl("^grid_h", tgp$run), ]
  g <- g[order(-g$ades_score), ]
  # The shipped configuration, named once in the setup chunk.
  hit <- which(g$trend_ceiling == SHIPPED_TREND$ceiling_field &
               g$trend_radius_nm == SHIPPED_TREND$radius_nm &
               g$trend_vote_margin == SHIPPED_TREND$margin)
  if (!length(hit)) {
    cat("::: {.callout-important}\n## Not checkable\n\nThe shipped configuration is not among the grid's cells, so this cannot be answered. Widen the grid.\n:::\n")
  } else {
    r <- hit[1]
    gap <- g$ades_score[1] - g$ades_score[r]
    cat(sprintf("The shipped configuration — %s, %g NM, vote margin %g — places **%d of %d** cells, %s behind the best.\n\n",
                cap_label(SHIPPED_TREND$ceiling_field, "field"),
                SHIPPED_TREND$radius_nm, SHIPPED_TREND$margin,
                r, nrow(g), if (gap == 0) "level" else sgn(-gap)))
    if (r == 1) {
      cat("It **is** the best cell. Every parameter version 6 recommended and\nwhich was never applied — a 6,100 ft ceiling, a vote margin of 2, a 20 NM\nradius — loses to the value that actually ships:\n\n")
      cmp <- function(col, alt, label) {
        keep <- g[g$trend_ceiling == SHIPPED_TREND$ceiling_field &
                  g$trend_radius_nm == SHIPPED_TREND$radius_nm &
                  g$trend_vote_margin == SHIPPED_TREND$margin, ]
        oth <- g[g[[col]] == alt, ]
        for (nm in setdiff(c("trend_ceiling", "trend_radius_nm", "trend_vote_margin"), col))
          oth <- oth[oth[[nm]] == keep[[nm]][1], ]
        if (!nrow(oth) || !nrow(keep)) return(NULL)
        data.frame(Parameter = label,
                   `Version 6 recommended` = alt,
                   `Ships` = keep[[col]][1],
                   `Cost of v6's value` = sgn(oth$ades_score[1] - keep$ades_score[1]),
                   check.names = FALSE)
      }
      rows <- do.call(rbind, list(
        cmp("trend_ceiling", 6100, "altitude ceiling (ft above field)"),
        cmp("trend_vote_margin", 2, "vote margin"),
        cmp("trend_radius_nm", 20, "detection radius (NM)")))
      if (!is.null(rows)) print(kable(rows, row.names = FALSE))
      cat("\nSo the recommendations were not merely un-applied; applying them would\nhave made arrival detection worse. `benchmarks/check_main_parity.py`\nasserts this on every run, and fails if the two ever part company.\n")
    } else {
      cat("It is **not** the best cell. That is reported, not resolved: changing a published default is the pipeline maintainers' decision, and this study's job is to make the gap visible.\n")
    }
  }
}
```

::: {.callout-note}
## What this check cannot see

The grid varies three parameters at one penalty and one ranking rule, on one
three-day period, with `process_dai` unable to read the 2024 tracks. And the
pipeline is not bit-reproducible: re-running moves a handful of flights where
two aerodromes tie on effective distance, worth about ±15 points of score. A
difference smaller than that is not a difference.

The ceiling comparison below sits at roughly two to three times that bound and
is consistent in sign across all six radius-and-margin combinations, which is
why it is reported as a result rather than as noise.
:::


## Running it {#sec-running}

The configuration this report recommends is the **default**. `adep_mode` and
`ades_mode` default to `None`, which means the pipeline uses `DetectionConfig`'s
own values — so you get the recommendation by not overriding anything.

Step `03` is the flight list: one row per track, carrying ADEP and ADES.

```python
from datetime import date
from opdi.runner import run_pipeline

run_pipeline(
    env="opensky",              # plain parquet over S3A; no Hive, no Iceberg
    step="03",
    start_date=date(2025, 6, 1),
    end_date=date(2025, 6, 30),
)
```

The same thing from a shell, using the installed console script:

```bash
opdi run --env opensky --step 03 --start 2025-06-01 --end 2025-06-30
```

To depart from the recommendation, name the methods explicitly. Both accept
`trend`, `endpoint` or `nearest`:

```bash
opdi run --env opensky --step 03 --start 2025-06-01 --end 2025-06-30 \
         --adep-mode trend --ades-mode trend
```

That example asks for `trend` on both roles — the pre-2026 behaviour, and the
`legacy` row in every comparison on this page.

**Step 03 does not stand alone.** It reads the tracks written by step `02` and
the aerodrome zones from step `00`, so those must have run for the same period
first. Omitting `--step` runs the whole chain in order.

::: {.callout-note}
## What "recommended" means here, and how it is kept honest

The values are not written into this page. @sec-defaults reads them from a
bare `DetectionConfig()` — the same object the pipeline constructs — so the
table cannot describe a configuration the code does not have. The equivalent
table in version 6 was typed by hand and named two values the code never had,
for three versions.

`benchmarks/check_main_parity.py` closes the loop the other way: it compares
that configuration against the best cell of the pipeline grid, which is the
only sweep that runs the shipped package rather than a re-implementation of
it.
:::

## Re-running a past month will not reproduce it
These defaults change ADEP and ADES for any month re-processed with them. That
is intended, and it is why the old behaviour remains reachable by name rather
than being deleted. A new algorithm gets a new `version`; the published one is
never mutated.
:::

## Provenance {#sec-provenance}

Each row is one file this page reads: the job that produced it, the commit it
was produced at, and when. Every figure on this page is regenerated by
`benchmarks/regenerate_v62.py`, which the page runs before drawing anything, so
a row marked **unverified** would mean a file staged outside that chain and
traceable to no command.

```{r provenance}
#| echo: false
#| results: asis
mf <- file.path("data", "_manifest.json")
if (!requireNamespace("jsonlite", quietly = TRUE)) {
  cat("*jsonlite not available; provenance table skipped.*\n")
} else if (!file.exists(mf)) {
  cat("**No manifest.** Every CSV here was staged by hand and none can be traced to the command that produced it.\n")
} else {
  m <- jsonlite::fromJSON(mf, simplifyDataFrame = FALSE)
  files <- sort(list.files("data", pattern = "[.]csv$"))
  rows <- do.call(rbind, lapply(files, function(f) {
    e <- m[[f]]
    if (is.null(e)) {
      data.frame(File = f, Job = "**unverified**", Commit = "—",
                 Produced = "—", check.names = FALSE)
    } else {
      data.frame(File = f, Job = basename(e$script),
                 Commit = paste0(e$git_sha, if (isTRUE(e$git_dirty)) " (dirty)" else ""),
                 Produced = substr(e$produced_utc, 1, 16), check.names = FALSE)
    }
  }))
  print(kable(rows, row.names = FALSE, align = "l"))
  n_un <- sum(rows$Job == "**unverified**")
  cat(sprintf("\n%d files, %d traced to a command, %d unverified.\n",
              nrow(rows), nrow(rows) - n_un, n_un))
}
```

### Where reproducibility stops

Two boundaries, stated rather than implied.

**Ground truth cannot be rebuilt here.** `research/reference` is extracted by
the `eurocontrol` R package against the PRISME Oracle warehouse, which no run
on this cluster can reach. It is an external input, and every figure on this
page is measured against it.

**The upstream tables are rebuilt only when absent.** Reference zones, state
vectors and tracks are produced by declared stages, but a stage whose table is
already present records its identity rather than rebuilding it — because
regenerating the airport zone table would change the candidate set under the
whole study. For this version they *were* rebuilt, from the archive upward,
because the sampler changed.

### What moved underneath this version {#sec-drift}

This version sets out to recompute version 6's study with **one** variable
changed. That is a checkable claim, and it does not fully hold. Stating where
it fails is the point of recording input identities in the first place.

```{r input-drift}
#| echo: false
#| results: asis
dr <- read_if("input_drift.csv")
if (is.null(dr)) {
  cat("*(the drift check has not been run against this checkout)*\n")
} else {
  show <- data.frame(
    Table   = dr$table,
    Objects = sprintf("%s → %s", dr$baseline_objects, dr$now_objects),
    Size    = sprintf("%.2f → %.2f GB", dr$baseline_bytes / 1e9, dr$now_bytes / 1e9),
    `Newest` = sprintf("%s → %s", substr(dr$baseline_newest, 1, 10),
                       ifelse(dr$now_newest == "", "—", substr(dr$now_newest, 1, 10))),
    Verdict = dr$verdict,
    check.names = FALSE)
  print(kable(show, row.names = FALSE, align = "r"))

  reb <- dr$table[dr$verdict == "REBUILT"]
  mk  <- dr$table[dr$verdict == "marker only"]
  if (length(mk)) {
    cat(sprintf("\n**%s**: byte-identical, same timestamp, one object fewer. That is the zero-byte directory marker MinIO leaves behind, which `provenance.s3_identity` stopped counting after it made a *deleted* table look populated. Bookkeeping, with no consequence for any figure.\n",
                paste(sprintf("`%s`", mk), collapse = ", ")))
  }
  if (length(reb)) {
    cat(sprintf("\n**%s**: genuinely rebuilt since version 6 recorded them. Every figure derived from them moved, and the datum does not explain any of it.\n",
                paste(sprintf("`%s`", reb), collapse = ", ")))
  }
}
```

The one that matters is **`opdi_endpoint_candidates`**, rebuilt after version 6
published: `r if (!is.null(dr)) { x <- dr[dr$table == "opdi_endpoint_candidates", ]; if (nrow(x)) sprintf("about %.0f%% smaller", 100 * (x$baseline_bytes - x$now_bytes) / x$baseline_bytes) else "smaller" } else "smaller"`, with the fresh-broadcast share roughly tripled —
the signature of stale-broadcast removal, which the PRC challenge results
recommend and which `endpoint` reads directly.

Two consequences, and they pull in opposite directions.

**The ground truth did *not* move.** `research/reference` grew by more than
half, but it grew by gaining *other periods*: this study's three-day slice is
unchanged, at `r num(as.numeric(trend$n_ground_truth[1]))` reference flights on
2025 and `r if (!is.null(t24)) num(as.numeric(t24$n_ground_truth[1])) else "—"`
on 2024 — the same counts version 6 reported. That is confirmed independently by @sec-equivalence,
where the sea-level sweep reproduces version 6's cell for cell. So every
denominator on this page is the one version 6 used, and the internal
comparisons — which are what this study actually argues from — are sound.

**But the endpoint-family numbers are not comparable to version 6's.** The
endpoint sweeps, the bearing test and the vertical measures all read the
rebuilt candidate table. Their figures differ from version 6's, and that
difference is a data change rather than a finding.

::: {.callout-warning}
## A verification this version cannot make

The intention was to verify directly that the datum change leaves departures
alone: the endpoint family is datum-independent by construction, so recomputing
it under the new code and recovering version 6's numbers would have settled it.

The candidate rebuild makes that comparison uncontrolled, so the verification is
**not available** and is not claimed. What remains is the within-study evidence
in @sec-bands — departures gain nothing in *any* elevation band, measured
where both arms read the same candidate table. That is genuine evidence and it
is weaker than the check that was intended.
:::

## Limitations {#sec-limitations}

### Parameters that were not swept

Five parameters were held fixed by decision rather than by evidence. Naming
them matters: a reader who sees seven tuned values and no caveat will
reasonably assume the whole configuration was optimised, and it was not.

| Parameter | Where | Held at | Why not swept |
|---|---|---|---|
| Smoothing half-window | `flights.py` | 2 samples | Parameterised in config so it is visible and changeable, but a smoothing width interacts with the sample rate rather than the geometry, and the decimation study settled the sample rate separately. |
| Classification dead band | `flights.py` | none exists | There is no dead band to sweep. The vote margin already plays that role. |
| Out-of-area border margin | `flights.py` | 30 NM | Fixes what counts as "outside the observed area". Changing it redefines the ground truth rather than the method, so a sweep would move the target as well as the aim. |
| Aerodrome type set | `h3_airport_zones.py` | large + medium | The candidate set. Widening it is a data-volume decision about the zone table, not a detection threshold. |
| H3 resolution | `config.py` | 7 (±0.76 NM) | Changing it invalidates every cached zone table. It no longer decides which candidates survive — that is exact distance now — but it still decides which aerodromes become candidates at all. |

Two of these carry a cost that this study measures rather than waves at, since
"not swept" is not the same as "no consequence".

```{r unswept-cost}
#| echo: false
#| results: asis
if (!is.null(typ) && !is.null(modes)) {
  out <- typ[typ$run == SHIPPED &
             !typ$apt_type %in% c("large_airport", "medium_airport", "out of area"), ]
  ooa <- typ[typ$run == SHIPPED & typ$apt_type == "out of area", ]
  m <- function(run, f) modes[[f]][modes$run == run][1]

  # Aggregated over roles: the frame carries one row per (class, role), and
  # listing them separately reads as four classes rather than two.
  byclass <- tapply(out$n_truth, out$apt_type, sum)
  cat(sprintf("**The type set.** %s flights across both roles have a true aerodrome outside the large/medium candidate set (%s). No method on this page can name them, whatever its thresholds, because the aerodrome is not among the candidates. They are counted in every denominator here, so each coverage figure already carries their cost: they are worth about %s of headline coverage.\n\n",
              num(sum(out$n_truth)),
              paste(sprintf("%s at %ss", num(byclass), gsub("_", " ", names(byclass))),
                    collapse = ", "),
              pct(sum(out$n_truth) / (2 * N_GT))))

  cat(sprintf("**The border margin.** Fixed at 30 NM, it decides which flights are labelled out-of-area. %s of ground-truth departures genuinely begin outside the observed area, and the recommended configuration recognises %s of them — at %s precision, so the label is trustworthy when it appears but is rarely reached. For arrivals the recall is %s: `trend` emits no out-of-area label at all, which is a structural property of the vote rule rather than a tuning choice.\n\n",
              pct(m("recommended", "adep_gt_ooa_share")),
              pct(m("recommended", "adep_ooa_recall")),
              pct(m("recommended", "adep_ooa_precision")),
              pct(m("recommended", "ades_ooa_recall"))))

  cat(sprintf("Both costs fall on coverage rather than accuracy, which is the benign direction: with the free out-of-area wins removed, in-area accuracy is %s for departures and %s for arrivals.\n",
              pct(m("recommended", "adep_inarea_accuracy")),
              pct(m("recommended", "ades_inarea_accuracy"))))
}
```

### `trend` cannot say "out of area"

`trend` names an aerodrome or says nothing. It has no way to report that a
flight began or ended outside the observed area, because the vote rule only
ever considers aerodromes the aircraft flew near — silence and *outside* are
the same output.

That is not a small omission for the shipped configuration, which takes
arrivals from `trend`:

```{r ooa-gap}
#| echo: false
#| results: asis
if (!is.null(modes)) {
  m <- function(run, f) modes[[f]][modes$run == run][1]
  cat(sprintf("%s of reference arrivals genuinely come from outside the observed area, and **none of them can be labelled as such**. Departures ship `endpoint`, which does emit the label: %s precision, so it is trustworthy when it appears, but only %s recall.\n",
              pct(m(SHIPPED, "ades_gt_ooa_share")),
              pct(m(SHIPPED, "adep_ooa_precision")),
              pct(m(SHIPPED, "adep_ooa_recall"))))
}
```

`endpoint` already has the mechanism — a fix at the edge of the ingestion
region rather than near any aerodrome — so giving `trend` the same test is a
port, not an invention. It is the clearest single improvement left on this page
and is not implemented here.

### The grid stops before the departure optimum does

The pipeline grid runs to 10,000 ft above field, and departures under `trend`
are still rising there. The research sweep goes further and does turn —
departures peak near the top of its range on both periods, then fall away as
accuracy drops faster than coverage rises — but "it turns in the sweep" is
weaker evidence than "it turns in the pipeline", and the two have disagreed
before.

This costs nothing that ships, because departures are served by `endpoint` and
the arrival optimum is comfortably interior on both periods and in both stages.
It does mean the departure half of the trend grid should not be quoted as an
optimum.

### One sample, and what that does and does not license

The parameters are chosen as the argmax over a grid evaluated on a single
three-day period. An argmax over hundreds of cells on one sample will always
overstate the gain slightly: some of the winning cell's margin is the sample's
noise, and it will not reproduce.

That is the reason for the second period, and for reporting the shape of each
optimum rather than only its location. A parameter sitting on a broad plateau
is a safe recommendation even from one sample; one sitting on a sharp isolated
peak is not, however good its score.

### The second period confirms through the sweeps, not the pipeline

Every pipeline number on this page is 2025-06-05…07. The 2024 tracks pre-date
this study's H3 indexing and `process_dai` reads `h3_res_7` off the track table
directly, so the pipeline arms cannot run on that period at all — version 6 met
the same wall and ran the pipeline on 2025 alone. @sec-ceiling states the
consequence where it arises.

Repeated here because it bounds every two-period claim in this report: the
second period tests the ceiling and the datum comparison, which is the part
most at risk of being a one-sample artefact, and it does **not** independently
reproduce the band breakdown, the tuning ladder, or the pipeline fidelity
check. A caveat that appears only beside the result it qualifies is easy to
read past.

### The `>3000` band needs a different question

The highest elevation band is not a datum problem and this study cannot fix it.
Recall there is under one percent on **both** datums, which is a coverage
failure upstream of detection: the aerodrome never becomes a candidate, so no
threshold anywhere in `trend` can reach it. Whether that is receiver geometry
over the Anatolian plateau or something in ingestion is worth knowing and is
not this study's question.

It is worth stating plainly because that band is where an elevation-aware cut
should have helped most, and the fact that it did not is evidence about
coverage rather than evidence against the datum — @sec-bands.

## Appendix A — The busiest 100 aerodromes {#sec-top100}

Movement counts are what OPDI is for, so the last check is per aerodrome rather
than per flight: for each of the hundred busiest in the reference, how many
movements the recommended methodology finds, against how many the Network
Manager data says there were, and how the legacy algorithm
compares on the same aerodromes.

The matching is **flight by flight, not count against count**. Each reference
flight is paired with an OPDI track on aircraft address (`icao24`) plus
callsign plus date, and where several tracks match, the one starting closest to
the reference off-block time. The columns then mean:

| Column | Definition |
|---|---|
| Movements (truth) | Reference flights whose true aerodrome is this one. |
| Found | Of those, how many OPDI gave any answer for. |
| Correct | Of those, how many OPDI attributed to **this** aerodrome. |
| Count ratio | Movements OPDI *assigns to* this aerodrome ÷ movements in truth. Includes flights that truly belong elsewhere, so it is what an operational count would show, and it can exceed 1. |
| Recall | Correct ÷ movements in truth. **Per-aerodrome recall** — not the out-of-area recall of @sec-metrics, which is a different measure sharing the word. |
| Recall, legacy | The same, for the algorithm published before 2026. |

```{r top100}
#| echo: false
#| results: asis
if (!is.null(apt)) {
  for (rl in c("departures", "arrivals")) {
    r <- apt[apt$run == SHIPPED & apt$role == rl & apt$airport != "OOA", ]
    l <- apt[apt$run == "legacy"      & apt$role == rl & apt$airport != "OOA", ]
    if (nrow(r) == 0) next
    r <- r[order(-r$n_truth), ][1:min(100, nrow(r)), ]
    l <- l[match(r$airport, l$airport), ]
    cat(sprintf("\n### %s\n\n", tools::toTitleCase(rl)))
    print(kable(data.frame(
      `#` = seq_len(nrow(r)),
      Aerodrome = r$airport,
      `Movements (truth)` = num(r$n_truth),
      `Found` = num(r$n_predicted_here),
      `Correct` = num(r$n_correct),
      `Count ratio` = sprintf("%.3f", r$count_ratio),
      `Recall` = pct(r$recall),
      `Recall, legacy` = pct(l$recall),
      `Δ recall` = pp(r$recall - l$recall),
      check.names = FALSE), row.names = FALSE, align = "r"))
  }
}
```

```{r top100-summary}
#| echo: false
#| results: asis
if (!is.null(apt)) {
  cat("\n### Across the hundred\n\n")
  rows <- do.call(rbind, lapply(c("departures", "arrivals"), function(rl) {
    r <- apt[apt$run == SHIPPED & apt$role == rl & apt$airport != "OOA", ]
    l <- apt[apt$run == "legacy"      & apt$role == rl & apt$airport != "OOA", ]
    r <- r[order(-r$n_truth), ][1:min(100, nrow(r)), ]
    l <- l[match(r$airport, l$airport), ]
    data.frame(
      Role = tools::toTitleCase(rl),
      `Movements in truth` = num(sum(r$n_truth)),
      `Recall, shipped` = pct(sum(r$n_correct) / sum(r$n_truth)),
      `Recall, legacy`  = pct(sum(l$n_correct, na.rm = TRUE) / sum(l$n_truth, na.rm = TRUE)),
      `Aerodromes improved` = sprintf("%d of %d", sum(r$recall > l$recall, na.rm = TRUE), nrow(r)),
      `Median count ratio`  = sprintf("%.3f", median(r$count_ratio, na.rm = TRUE)),
      check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
  cat("\nA count ratio below one means the aerodrome is under-counted — the usual case, since every abstention is a movement not attributed. Above one means movements are being attributed to an aerodrome that did not see them, which is the more damaging error and the one the scoring penalises at rate $k$.\n")
  cat("\nThe departure row is the result: recall rises by nearly nine points and the median aerodrome is within about one percent of its true movement count.\n")
  arr <- rows[rows$Role == "Arrivals", ]
  cat(sprintf("\n**The arrival row must be read with @sec-verdict in hand.** Recall reads %s against legacy's %s, improving on %s — which looks like a mild regression and is not one. Measured as an algorithm on a trend-only flight list, arrivals gain; the shortfall here comes from the merged list also containing departure-only tracks, which this table pairs with reference flights by start time, so some arrivals are counted against a track that carries none. The measurement moved; the data did not. It is left in the table rather than corrected away, because a benchmark that quietly adjusts its own alignment to flatter a result is worth less than one that shows where it is fragile.\n",
              arr$`Recall, shipped`, arr$`Recall, legacy`,
              arr$`Aerodromes improved`))
}
```
