---
title: "A new rule builds OPDI's flight tracks"
subtitle: "Version 2 — the release note: what changed in the pipeline, what it recovers, and why every `track_id` you hold stops joining"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r regenerate}
#| include: false
#| cache: false

# Rendering this page runs the analysis behind it.
#
# `regenerate_track_v2.py` is the executable definition of the study: each CSV
# read below is produced by exactly one job it declares, with exactly the
# arguments it declares. It re-runs a job when the *code that job depends on*
# has changed since the output was written -- not when the file is old, which
# says nothing. A render with everything current is a fast no-op; a render after
# editing `segmentation/methods.py` re-runs both pipeline jobs before the first
# table is drawn.
#
# Re-running one is expensive here in a way it was not for V1. V2's jobs drive
# the *production* steps 01 -> 02 -> 02a -> 03 over a full day of state vectors,
# three arms per period, which is hours of cluster time and needs credentials.
# OPDI_RENDER selects what happens instead:
#
#   run          (default) regenerate anything stale, then render
#   check        fail fast naming every stale job, regenerate nothing
#   allow-stale  render a draft from whatever is staged, after naming staleness
#   skip         do not invoke the entrypoint at all
#
# `skip` exists for the case this page was last rendered under: the cluster was
# fully committed to another job, and starting a Spark session would have taken
# quota from it. It is the only mode that touches neither the cluster nor S3,
# and it is honest about what it buys -- nothing is checked against the code, so
# the CSVs are trusted as staged.
REPO   <- Sys.getenv("OPDI_REPO_DIR",
                     normalizePath(file.path(getwd(), "..", "..", "..", "opdi"),
                                   mustWork = FALSE))
PYTHON <- Sys.getenv("OPDI_PYTHON", file.path(REPO, ".venv310", "bin", "python"))
MODE   <- Sys.getenv("OPDI_RENDER", "run")

# The pipeline repo is a sibling of the portal. That resolution is wrong inside
# a git worktree, where the paper sits under .claude/worktrees/<name>/papers and
# the sibling walk lands somewhere that does not exist -- so the regeneration
# step would be skipped silently and the page would render whatever is staged.
# OPDI_REPO_DIR overrides it; the default is unchanged for a normal checkout.
#
# A missing entrypoint must not read as "nothing to regenerate". Skipping it
# silently is how a page ends up publishing numbers no command can account for.
regen <- file.path(REPO, "benchmarks", "regenerate_track_v2.py")
if (!file.exists(regen) && MODE != "skip") {
  stop("regenerate_track_v2.py not found under REPO=", REPO,
       " -- set OPDI_REPO_DIR, or OPDI_RENDER=skip to render without it.")
}

# `skip` short-circuits before system2, so no Python process starts and no Spark
# session is created. Every other mode reaches the entrypoint.
if (MODE != "skip" && file.exists(regen)) {
  flags <- switch(MODE,
                  "run" = character(0),
                  "check" = "--check",
                  "allow-stale" = c("--check", "--allow-stale"),
                  character(0))
  status <- system2(PYTHON, c(shQuote(regen), flags), stdout = "", stderr = "")
  if (!identical(status, 0L)) {
    stop("regenerate_track_v2.py exited ", status,
         " -- the figures below would not match the code that is checked out. ",
         "Set OPDI_RENDER=allow-stale to render anyway.")
  }
}
```

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every figure and table on this page comes from a committed CSV.
# Nothing here reads S3, a database or a Spark cluster.

D <- function(f) read.csv(file.path("data", f), stringsAsFactors = FALSE)

pipe <- rbind(D("pipeline_2025.csv"), D("pipeline_2024.csv"))
pipe$period <- as.character(pipe$period)

# V1's arms, read from V1's own committed CSVs rather than retyped here. The
# cross-validation compares two measurements of the same quantity, and a
# transcribed comparand is exactly the thing that goes stale silently.
V1DIR <- file.path("..", "track-construction-v1", "data")
if (!dir.exists(V1DIR)) {
  stop("V1's data directory not found at ", V1DIR,
       " -- the cross-validation section reads its arms_*.csv directly.")
}
D1 <- function(f) read.csv(file.path(V1DIR, f), stringsAsFactors = FALSE)
v1 <- rbind(D1("arms_2025.csv"), D1("arms_2024.csv"))
v1$period <- as.character(v1$period)

# One accessor per file, so no two tables on this page can disagree about what
# an arm scored. Both stop rather than return a zero-length value: a missing arm
# would otherwise propagate into a table cell as a blank and read as a measured
# result.
P <- function(m, p, col) {
  v <- pipe[pipe$method == m & pipe$period == p, col]
  if (length(v) != 1) stop("no single pipeline row for ", m, "/", p)
  v
}
# V1 named the shipped rule `recommended`; the pipeline names it `standard`.
# Same rule, renamed when it stopped being a recommendation and became a
# default. Mapped once, here, so nothing below has to remember it.
V1ARM <- c(legacy = "legacy", airframe_only = "airframe_only",
           standard = "recommended")
V <- function(m, p, col) {
  v <- v1[v1$arm == V1ARM[[m]] & v1$period == p, col]
  if (length(v) != 1) stop("no single V1 row for ", m, "/", p)
  v
}

PERIODS <- c("2025", "2024")
ARMS    <- c("legacy", "airframe_only", "standard")
ARMLAB  <- c(legacy = "legacy", airframe_only = "airframe_only",
             standard = "standard (ships)")

num  <- function(x) formatC(x, format = "d", big.mark = ",")
pct  <- function(x) sprintf("%.2f%%", x)
pp   <- function(x) sprintf("%+.2f", x)
p1   <- function(x) sprintf("%.1f", x)
p4   <- function(x) sprintf("%.4f", x)
# A magnitude, for sentences that already carry the direction in a verb.
# "falls -0.66 points" is a double negative; "falls 0.66 points" is not.
mag  <- function(x) sprintf("%.2f", abs(x))
# Every cell that can be absent goes through this. `airframe_only` carries no
# ADEP/ADES numbers, and an NA formatted by default prints the literal "NA" into
# a table -- which reads as a measured value of nothing rather than as a cell no
# run produced.
dash <- function(x, f = pct) ifelse(is.na(x), "—", f(x))

# --- headline values, computed here rather than written into the prose, so the
# --- opening paragraph cannot drift away from the tables that justify it.
GAIN      <- function(p) P("standard", p, "clean_match_pct") -
                         P("legacy", p, "clean_match_pct")
GATE_GAIN <- function(p) P("standard", p, "gate_clean_match_pct") -
                         P("legacy", p, "gate_clean_match_pct")
V1_GAIN      <- function(p) V("standard", p, "clean_match_pct") -
                            V("legacy", p, "clean_match_pct")
V1_GATE_GAIN <- function(p) V("standard", p, "gate_clean_match_pct") -
                            V("legacy", p, "gate_clean_match_pct")
# The largest disagreement between the pipeline's answer and the harness's, over
# both intervals and both periods. This is the number the cross-validation is.
CROSS <- max(abs(c(sapply(PERIODS, function(p) GAIN(p) - V1_GAIN(p)),
                   sapply(PERIODS, function(p) GATE_GAIN(p) - V1_GATE_GAIN(p)))))

# The ablation, split at its midpoint. Both halves are in points of clean
# matching, so they sum to the total by construction.
STEP1 <- function(p) P("airframe_only", p, "clean_match_pct") -
                     P("legacy", p, "clean_match_pct")
STEP2 <- function(p) P("standard", p, "clean_match_pct") -
                     P("airframe_only", p, "clean_match_pct")

# The partition change. `n_tracks` counts tracks carrying at least one
# ground-truth sample, which is the count V1 also reports and the only one the
# two studies can be compared on.
DROP    <- function(p) 100 * (1 - P("standard", p, "n_tracks") /
                                  P("legacy", p, "n_tracks"))
V1_DROP <- function(p) 100 * (1 - V("standard", p, "n_tracks") /
                                  V("legacy", p, "n_tracks"))
EXPDROP <- function(p) 100 * (1 - P("standard", p, "n_tracks_exported") /
                                  P("legacy", p, "n_tracks_exported"))

# The flight list's own score for one arm: correct aerodromes minus twice the
# wrong ones, summed over both ends. Written once because the payoff section
# quotes the total, the delta and the share, and three transcriptions of one
# formula is three chances to use a different k.
SCORE <- function(m, p) P(m, p, "adep_score") + P(m, p, "ades_score")
# How many of the period's ground-truth flights got an aerodrome at all on one
# side. The denominator the payoff has to be read against.
NAMED <- function(m, p, side) P(m, p, paste0(side, "_correct")) +
                              P(m, p, paste0(side, "_wrong"))

# The gate interval admits flights the airborne interval does not, so the two
# rates have different denominators in principle. Here they do not, and the
# difference is measured rather than asserted.
DENOM_GAP <- function(p) P("legacy", p, "n_flights") -
                         P("legacy", p, "gate_n_flights")

# Wall-clock minutes per arm, as the pipeline job recorded them. Quoted as a
# range because the ordering is not stable between periods and a single
# difference would read as a systematic cost.
MIN_LO <- min(pipe$minutes)
MIN_HI <- max(pipe$minutes)

# What ships, from `SegmentationConfig` in `src/opdi/config.py`. Configuration
# values, not measurements -- named here so the config table and the prose read
# one source.
SHIP <- list(method = "standard", gap = 30, low_gap = 15, low_ft = 5000,
             lookback = "None")

# How many parameter combinations V1 swept before leaving the thresholds alone.
# Counted from V1's own grid rather than quoted, for the same reason its arms
# are read rather than transcribed.
SWEEP_CELLS <- nrow(D1("sweep_recommended_2025_stage1.csv"))
```

## Summary {.unnumbered}

**OPDI now builds `track_id` with a different rule, and no identifier from any
earlier release joins to any identifier in this one.**

Segmentation is the pipeline's decision about which state vectors belong to the
same flight. It is made once, in step 02, and everything downstream inherits it:
the flight list groups by `track_id`, events fire per track, measurements
accumulate from a track's first sample. Every OPDI release up to v0.0.2 made
that decision by grouping on `(icao24, callsign)` and cutting on time gaps. From
this release forward the pipeline groups on `icao24` alone and cuts when the
airframe's callsign genuinely changes. The configuration key is
`segmentation.method`, and its default moved from `"legacy"` to `"standard"`.

**What it recovers.** On a full day of state vectors through the production
steps, the share of Network Manager flights that come out as exactly one track
carrying nothing else rises from `r pct(P("legacy", "2025", "clean_match_pct"))`
to `r pct(P("standard", "2025", "clean_match_pct"))` on 2025-06-05, and from
`r pct(P("legacy", "2024", "clean_match_pct"))` to
`r pct(P("standard", "2024", "clean_match_pct"))` on 2024-06-05. Score the
turnaround as part of the flight — block to block, so stand and taxi samples
count — and the same change is
`r pct(P("legacy", "2025", "gate_clean_match_pct"))` to
`r pct(P("standard", "2025", "gate_clean_match_pct"))`, a gain of
`r pp(GATE_GAIN("2025"))` points.

**Why this page exists.** Version 1 of this study measured eight candidate rules
in a harness that read a pre-built track table, and recommended this one. A
harness is not the pipeline, and the question a recommendation cannot answer is
whether its result survives the real thing. It does. The pipeline and the
harness disagree by at most `r sprintf("%.2f", CROSS)` points across two periods
and two scoring intervals — four independent comparisons of the same quantity,
agreeing to about a point.

**What it costs.** The flight list names slightly fewer arrival aerodromes:
`r pct(100 * P("standard", "2025", "ades_coverage"))` of ground-truth flights
against `r pct(100 * P("legacy", "2025", "ades_coverage"))`, at essentially
unchanged accuracy. Across both ends the flight list's own score falls
`r pct(100 * (1 - SCORE("standard", "2025") / SCORE("legacy", "2025")))` on 2025
and `r pct(100 * (1 - SCORE("standard", "2024") / SCORE("legacy", "2024")))` on
2024. The mechanism is merging, and [the payoff section](#sec-payoff) measures
it.

**What breaks.** Every `track_id` changes. The two rules hash different group
keys and use different suffixes, so the count of identifiers surviving the
release is zero for reasons visible in the code rather than in any measurement.
The partition itself shrinks by `r pct(DROP("2025"))` on 2025 and
`r pct(DROP("2024"))` on 2024. A consumer joining on `track_id` across the
release boundary gets an empty result, not an error. **Nothing in the published
data says which rule produced a row** — `osn_tracks` carries no version column,
so the release date is the only discriminator. That is
[what breaks](#sec-continuity), and it is the section to read if you read only
one.

::: {.callout-note}
## How to read this page

This is a release note, not a study. It reports one pipeline run per period,
three segmentation arms each, through the production steps 01 → 02 → 02a → 03.
[What changed](#sec-changed) is the consumer-visible list.
[How a segmentation is scored](#sec-metrics) defines every term used later, so
this page stands alone and does not send you to V1 for a definition.
[What it improves](#sec-results) and [which change does the work](#sec-ablation)
are the measurement. [What it costs](#sec-payoff) is downstream.
[What breaks](#sec-continuity) and [what ships](#sec-defaults) are the
operational sections.

Every number here is an expression evaluated at render time over a committed CSV
— `data/pipeline_2025.csv` and `data/pipeline_2024.csv` for the pipeline run,
and V1's own `arms_*.csv` for the harness figures the cross-validation compares
against. Both pipeline files are declared by
`benchmarks/regenerate_track_v2.py`, which this page runs before drawing
anything, which invokes `benchmarks/track_pipeline_v2.py` to produce them, and
which stamps each output with the code that produced it in
`data/_manifest.json` — script, arguments, git SHA, and a fingerprint over the
source files the job depends on. Nothing is typed into the prose, which is what stops the
text and the tables drifting apart. Two exceptions are named where they appear:
the configuration defaults, which are code rather than measurement, and one
diagnostic quoted from `config.py`.

**V1's figures are a different measurement of the same quantity, not a baseline
these are differences from.** V1 scored a harness over three days; V2 scores the
pipeline over one. Where both are quoted, they are quoted as two answers to one
question.
:::

## What changed {#sec-changed}

```{r changed-table}
#| echo: false
ch <- data.frame(
  What = c("`segmentation.method`",
           "`FLIGHT_LIST_VERSION`",
           "`FLT_ID` (flight list)",
           "`FLT_ID` on unlabelled overflights",
           "Airport events (step 04)"),
  From = c("`\"legacy\"`", "`v2.0.0` / `v3.0.0`",
           "lexicographically smallest callsign in the track",
           "`NULL`",
           "one event per callsign broadcast during a zone crossing"),
  To = c("`\"standard\"`", "`v5.0.0`",
         "the track's dominant non-blank callsign",
         "`\"\"`",
         "one event per zone crossing"),
  check.names = FALSE)
kable(ch, row.names = FALSE, align = "lll",
      col.names = c("What", "Was", "Is now"))
```

**The two rules, stated in full.** `legacy` groups samples by
`SHA2(icao24 || callsign)`, orders them by time, and starts a new track when the
gap to the previous sample exceeds `r SHIP$gap` minutes, or exceeds
`r SHIP$low_gap` minutes while the aircraft is below `r num(SHIP$low_ft)` ft.
The identifier it stamps carries a `_{year}_{month}` suffix. `standard` groups
by `SHA2(icao24)`, keeps both gap rules unchanged, and adds one break: the last
*non-blank* callsign seen on this airframe differs from the current one. Its
identifier carries a sample offset instead of a month suffix.

The break compares the last non-blank callsign rather than the raw column
because ADS-B broadcasts the callsign in its own message type and it is blank
until the crew sets it. Comparing raw values would fire on every blank sample
and reproduce exactly the fragmentation the change removes. The lookback is
bounded by the gap threshold, so it cannot reach back across a gap break into
the previous flight.

**Three changes reach data consumers directly, and none of them raises an
error.**

`FLT_ID` is now the dominant callsign of the track rather than the
lexicographically smallest, because a blank sorts first and a `standard` track
can carry a blank. Under legacy segmentation the two aggregates agree on every
track — callsign is in the group key, so a legacy track carries one value by
construction — so no historical data is affected.

**Unlabelled overflights now carry `""` where they carried `NULL`.** The column
previously used `""` for unlabelled detected flights and `NULL` for unlabelled
overflights: two spellings of "no callsign" in one column, so
`WHERE FLT_ID IS NULL` and `WHERE FLT_ID = ''` answered the same question
differently. **A consumer filtering overflights on `IS NULL` now silently gets
nothing back.**

Step 04 emits one airport event per zone crossing. It previously grouped on
`flight_id` without aggregating it, so an aircraft that broadcast two callsigns
while crossing one runway emitted two entry-runway events for one crossing. On
legacy tracks that could not happen, and the two behaviours are identical on
every published release.

**`osn_tracks` gains no version column.** A row does not say which segmentation
produced it. The release date is the only discriminator, which is why
[what breaks](#sec-continuity) is written the way it is.

## What has already happened to a sample {#sec-pipeline}

*Side information: if you know the pipeline steps, skip to
[the metrics](#sec-metrics).*

Segmentation is step 02 of a chain, and the samples reaching it are not raw
OpenSky rows. Step 01 restricts them to the OPDI bounding box and thins them to
one sample per aircraft per 5 s interval, so every gap threshold in this study
is measured against a 5 s grid.

```{r pipeline-diagram}
#| echo: false
#| fig-width: 7
#| fig-height: 4.2
#| fig-cap: "The production chain. Step 02 is where the rule changed; every step below it inherits the result, and this release ran the whole chain per arm rather than a harness over step 02's output."
# Drawn with ggplot rather than as a mermaid block, deliberately. A mermaid
# diagram written as an executable cell renders in HTML but needs a headless
# browser to rasterise for PDF, and that browser does not start in this
# environment -- the render then hangs with no error. A mermaid fence emitted
# from knitr avoids the hang but arrives without the JavaScript that draws it,
# so it shows as literal text. A plot goes through whatever device Quarto has
# already selected and is correct in both formats.
nodes <- data.frame(
  y = 5:1,
  label = c(
    "step 01 — ingest OSN state vectors: bbox + thinning",
    "step 02 — tracks: assign track_id",
    "step 02a — cleaning: mask bad values to NULL",
    "step 03 — flight list: ADEP / ADES per track",
    "step 04 — events and measurements per track"),
  changed = c(FALSE, TRUE, FALSE, FALSE, FALSE))
h <- 0.34
ggplot(nodes) +
  geom_rect(aes(xmin = 0, xmax = 1, ymin = y - h, ymax = y + h,
                fill = changed, colour = changed, linewidth = changed)) +
  geom_text(aes(x = 0.5, y = y, label = label, fontface = ifelse(changed, 2, 1)),
            size = 3.1) +
  geom_segment(data = nodes[nodes$y > 1, ],
               aes(x = 0.5, xend = 0.5, y = y - h, yend = y - 1 + h),
               arrow = arrow(length = unit(1.8, "mm"), type = "closed"),
               colour = "grey40", linewidth = 0.3) +
  annotate("text", x = 1.04, y = 4, hjust = 0, size = 2.9, colour = "grey25",
           label = "the rule inside step 02\nis what this release changes") +
  annotate("segment", x = 1.02, xend = 1.02, y = 5 + h, yend = 2 - h,
           colour = "grey55", linewidth = 0.3) +
  annotate("text", x = 1.04, y = 2, hjust = 0, size = 2.9, colour = "grey25",
           label = "run per arm, both periods\n(step 03 for legacy and\nstandard only)") +
  coord_cartesian(xlim = c(0, 1.6), ylim = c(0.4, 5.6), clip = "off") +
  scale_fill_manual(values = c(`FALSE` = "white", `TRUE` = "grey90")) +
  scale_colour_manual(values = c(`FALSE` = "grey55", `TRUE` = "grey15")) +
  scale_linewidth_manual(values = c(`FALSE` = 0.3, `TRUE` = 0.8)) +
  theme_void() +
  theme(legend.position = "none")
```

```{r pipeline-table}
#| echo: false
steps <- data.frame(
  Step = c("01", "02", "02a", "03", "04"),
  Does = c(
    "Downloads OpenSky state vectors, restricts them to the OPDI bounding box, and thins them to one sample per aircraft per interval.",
    "Groups samples into tracks and stamps `track_id`. **This is the step that changed.**",
    "Masks values that cannot be right — duplicates, out-of-range positions, stale broadcasts, derivative spikes, isolated points — to NULL, keeping the row. Writes a second table rather than editing the first.",
    "One flight per `track_id`: departure and destination aerodrome, and the flight's callsign label.",
    "Milestones per track — take-off, top-of-climb, level segments, landing — and the measurements attached to them."
  ),
  Output = c("`osn_statevectors`", "`osn_tracks`", "`osn_tracks_clean`",
             "`opdi_flight_list`", "events, measurements"),
  check.names = FALSE)
kable(steps, row.names = FALSE, align = "lll",
      col.names = c("Step", "What it does", "Writes"))
```

Each arm ran 01 → 02 → 02a for its period, and `legacy` and `standard` ran 03 as
well. The three arms of a period share one ingested copy of the day, so a
difference between arms cannot be an artefact of having ingested the day three
times.

## How a segmentation is scored {#sec-metrics}

*Nothing after this section introduces a metric.*

Ground truth is Network Manager's flight table, joined to ADS-B on `icao24` —
the 24-bit airframe address, which Network Manager carries as
`AIRCRAFT_ADDRESS`. A ground-truth flight is an airframe and a span of time; a
state vector is attributed to the flight of the same airframe whose span holds
its timestamp. A track and a flight are *matched* when at least one of the
track's samples is attributed to that flight. The join is never on callsign,
because callsign is exactly what the rules under comparison disagree about.

**Every ground-truth flight falls in one of three classes, which sum to 100%.**

- **clean** — the flight's samples are in one track, and that track carries no
  other flight. This is `clean_match_pct`, the headline.
- **fragmented** — the flight's samples span several tracks, none of which
  carries another flight.
- **merged** — the flight's dominant track also carries at least one other
  flight.

A flight that is both counts as **merged**, because merging is the worse
failure: a merged flight is unrecoverable downstream, where a fragmented one is
at least present in pieces. One stray sample in a second track costs a flight
its clean match, so `clean_match_pct` is a lower bound and is read as one.

Two entropy-based clustering measures name the two failure modes on a 0–1 scale.
**Homogeneity** asks whether each track holds samples from only one flight, so
it falls with merging. **Completeness** asks whether each flight's samples all
landed in one track, so it falls with fragmentation. Both are weighted over
every sample in the period, and most samples sit in tracks that every arm gets
right, so they span a narrow range and name the failure modes rather than rank
the arms. The three-class rate is what ranks them.

**Each arm is scored twice, over two intervals of the same partition.**

- The **airborne** interval runs from take-off to landing, `[t_off, t_land]`. A
  sample taken on stand or on a taxiway falls outside it and is attributed to no
  flight.
- The **gate-to-gate** interval runs off-block to in-block, so the turnaround
  counts. This matters because an OPDI track is supposed to carry it: first- and
  last-seen events fire on it, taxi phases are extracted from it, and every
  cumulative measurement starts at the track's first sample. A rule that
  shatters a flight between the gate and the runway threshold scores a clean
  airborne match and hands over a broken product.

The gate interval is where a rule carrying callsign in its group key comes
apart, because that is when the callsign is blank or rotating: the aircraft is
on stand, the crew has not set the next leg's callsign, and every blank sample
forms a group — and therefore a track — of its own.

Block times come from APDF where a movement was measured. The departure side
falls back to Network Manager's `AOBT_3` and so exists for substantially every
flight; the arrival side has no fallback, because `ARVT_3` is a landing time and
not a gate arrival, and is therefore APDF-only. Where a block time is missing
the interval is extended past the airborne endpoint by the median measured taxi
on that side, over the flights of that period that measured one.

**The two intervals score the same flights here.** The gate interval can in
principle admit flights the airborne one does not — a departure seen only on the
taxiway — so the denominators are reported rather than assumed. On 2025 they
differ by `r num(abs(DENOM_GAP("2025")))` flights out of
`r num(P("legacy", "2025", "n_flights"))`, and on 2024 by
`r num(abs(DENOM_GAP("2024")))`. Nothing below is a change of denominator
wearing the clothes of a change of rate.

**The sample.** One day per period, run whole:
`r num(P("legacy", "2025", "n_samples"))` state vectors on 2025-06-05 and
`r num(P("legacy", "2024", "n_samples"))` on 2024-06-05.
`r num(P("legacy", "2025", "n_ground_truth"))` Network Manager flights are in
scope for 2025 and `r num(P("legacy", "2024", "n_ground_truth"))` for 2024, of
which `r num(P("legacy", "2025", "n_flights"))` and
`r num(P("legacy", "2024", "n_flights"))` respectively have at least one ADS-B
sample inside their airborne interval and are what the rates are computed over.

## What it improves {#sec-results}

```{r results-table}
#| echo: false
res <- do.call(rbind, lapply(PERIODS, function(p) {
  do.call(rbind, lapply(ARMS, function(m) data.frame(
    Period = p,
    Arm = ARMLAB[[m]],
    Clean = pct(P(m, p, "clean_match_pct")),
    Frag = pct(P(m, p, "fragmented_pct")),
    Merged = pct(P(m, p, "merged_pct")),
    GClean = pct(P(m, p, "gate_clean_match_pct")),
    GFrag = pct(P(m, p, "gate_fragmented_pct")),
    GMerged = pct(P(m, p, "gate_merged_pct")),
    check.names = FALSE)))
}))
kable(res, row.names = FALSE, align = "llrrrrrr",
      col.names = c("Period", "Arm", "clean", "fragmented", "merged",
                    "clean", "fragmented", "merged"),
      caption = "Airborne (columns 3–5) and gate-to-gate (columns 6–8). Each triple sums to 100%.")
```

`legacy` fragments `r pct(P("legacy", "2025", "fragmented_pct"))` of flights on
2025 and merges `r pct(P("legacy", "2025", "merged_pct"))`. Almost all of its
failure is the splitting and almost none is the opposite error. `standard`
fragments `r pct(P("standard", "2025", "fragmented_pct"))` and merges
`r pct(P("standard", "2025", "merged_pct"))`: a fragmentation fix that pays for
itself several times over in a small amount of new merging.

Scored gate to gate, `legacy` falls to
`r pct(P("legacy", "2025", "gate_clean_match_pct"))` while `standard` holds
`r pct(P("standard", "2025", "gate_clean_match_pct"))`. The wider interval costs
`standard` `r mag(P("standard", "2025", "gate_clean_match_pct") - P("standard", "2025", "clean_match_pct"))`
points on 2025 and `legacy`
`r mag(P("legacy", "2025", "gate_clean_match_pct") - P("legacy", "2025", "clean_match_pct"))`.
Adding the turnaround adds samples, and added samples can only fragment a
partition further; the size of the two penalties is the difference between a
rule that carries callsign in its group key and one that does not.

```{r hc-table}
#| echo: false
hc <- do.call(rbind, lapply(PERIODS, function(p) {
  do.call(rbind, lapply(ARMS, function(m) data.frame(
    Period = p, Arm = ARMLAB[[m]],
    Hom = p4(P(m, p, "homogeneity")),
    Comp = p4(P(m, p, "completeness")),
    Tracks = num(P(m, p, "n_tracks")),
    Exported = num(P(m, p, "n_tracks_exported")),
    check.names = FALSE)))
}))
kable(hc, row.names = FALSE, align = "llrrrr",
      col.names = c("Period", "Arm", "homogeneity", "completeness",
                    "matched tracks", "tracks written"),
      caption = "Homogeneity falls with merging, completeness with fragmentation. `matched tracks` counts tracks carrying at least one ground-truth sample; `tracks written` counts every track step 02 wrote for the day.")
```

Completeness rises from `r p4(P("legacy", "2025", "completeness"))` to
`r p4(P("standard", "2025", "completeness"))` on 2025 while homogeneity falls
from `r p4(P("legacy", "2025", "homogeneity"))` to
`r p4(P("standard", "2025", "homogeneity"))`: less fragmentation, marginally
more merging, the same finding the three-class rates give and the reason both
are reported.

Running `standard` costs no more cluster time than running `legacy`. Arm
runtimes across the six runs span `r p1(MIN_LO)` to `r p1(MIN_HI)` minutes, and
which arm is slowest reverses between the two periods — on 2025 `legacy` took
`r p1(P("legacy", "2025", "minutes"))` minutes against `standard`'s
`r p1(P("standard", "2025", "minutes"))`, and on 2024 the order is the other way
round. These are wall-clock times on a shared cluster and settle nothing finer
than that.

## Does the harness result survive the pipeline? {#sec-crossval}

This is the question the release was made to answer, and it is why the arms were
re-run through the production steps rather than scored again in the harness that
recommended them.

V1 measured segmentation by applying each candidate rule to an already-built,
already-cleaned track table. That design bought eight rules for the price of one
pipeline run, and it left one thing open: a re-implementation of a rule scored
over a re-used table is not the rule the pipeline runs over the table the
pipeline builds. Cleaning runs *after* segmentation in production and *before*
it in the harness. If the recommendation depended on that ordering, or on any
other difference between the two, the recommendation was an artefact.

```{r crossval-table}
#| echo: false
cv <- do.call(rbind, lapply(PERIODS, function(p) {
  data.frame(
    Period = p,
    Interval = c("airborne", "gate-to-gate"),
    Pipeline = c(pp(GAIN(p)), pp(GATE_GAIN(p))),
    Harness = c(pp(V1_GAIN(p)), pp(V1_GATE_GAIN(p))),
    Diff = c(pp(GAIN(p) - V1_GAIN(p)), pp(GATE_GAIN(p) - V1_GATE_GAIN(p))),
    check.names = FALSE)
}))
kable(cv, row.names = FALSE, align = "llrrr",
      col.names = c("Period", "Interval", "pipeline (V2)", "harness (V1)",
                    "difference"),
      caption = "`standard` minus `legacy`, in points of clean matching. Two independent measurements of one quantity over different samples: V1 scored three days in a harness, V2 scores one day through the production steps.")
```

The two measurements agree to within `r sprintf("%.2f", CROSS)` points on all
four comparisons. The pipeline run confirms the harness result: the gain V1
recommended on is the gain production delivers, and the harness's shortcuts did
not manufacture it.

The agreement also disciplines the claim. Four numbers this close, from
different samples and different code paths, say that the effect is a property of
the segmentation rule rather than of the day, the ordering of cleaning, or the
implementation.

## Which change is doing the work {#sec-ablation}

`standard` is two changes to `legacy`, and either could have been carrying the
result. Dropping callsign from the group key stops blank callsigns from forming
tracks. Adding the callsign-change break puts back the separation the group key
used to supply. The middle arm, `airframe_only`, is the first change without the
second, and it exists so the gain can be split.

```{r ablation-table}
#| echo: false
ab <- do.call(rbind, lapply(PERIODS, function(p) data.frame(
  Period = p,
  Step = c("legacy → airframe_only", "airframe_only → standard", "total"),
  What = c("drop callsign from the group key", "add the callsign-change break", ""),
  Points = c(pp(STEP1(p)), pp(STEP2(p)), pp(GAIN(p))),
  Merged = c(pp(P("airframe_only", p, "merged_pct") - P("legacy", p, "merged_pct")),
             pp(P("standard", p, "merged_pct") - P("airframe_only", p, "merged_pct")),
             pp(P("standard", p, "merged_pct") - P("legacy", p, "merged_pct"))),
  check.names = FALSE)))
kable(ab, row.names = FALSE, align = "lllrr",
      col.names = c("Period", "Step", "What it changes",
                    "clean matching (points)", "merged (points)"),
      caption = "The gain split at its midpoint, airborne interval. The two steps sum to the total by construction.")
```

**Dropping callsign from the group key is nearly all of it**:
`r mag(STEP1("2025"))` of the `r mag(GAIN("2025"))` points on 2025, and
`r mag(STEP1("2024"))` of `r mag(GAIN("2024"))` on 2024. The mechanism is blunt.
With callsign in the key, a sample decoded without one forms its own group and
therefore its own track; a diagnostic taken while the rules were implemented and
recorded in `config.py` — not read from a CSV here — puts 42.4% of legacy's
tracks in that state. Fragmentation is what that produces, and it is an artefact
of the key rather than anything about the flight.

**The callsign-change break earns the rest by removing merging.**
`airframe_only` alone merges `r pct(P("airframe_only", "2025", "merged_pct"))`
of flights on 2025 against `legacy`'s
`r pct(P("legacy", "2025", "merged_pct"))`: grouping on the airframe alone joins
two consecutive legs of the same aircraft into one track, and the second leg
stops existing downstream. Restoring the break cuts that to
`r pct(P("standard", "2025", "merged_pct"))` and adds `r pp(STEP2("2025"))`
points of clean matching on 2025 and `r pp(STEP2("2024"))` on 2024. The trade is
visible in the fragmentation column too — `standard` fragments
`r mag(P("standard", "2025", "fragmented_pct") - P("airframe_only", "2025", "fragmented_pct"))`
points more than `airframe_only`, the cost of splitting on a signal that is
sometimes noise — and it is worth paying because merging is the failure that
loses a flight outright.

Both halves earn their place on both periods, which is why the shipped rule is
the combination and not either piece.

## What it costs downstream {#sec-payoff}

Step 03 — the expensive step — ran for `legacy` and `standard` only.
`airframe_only` is the segmentation ablation's midpoint and has no downstream
question of its own: what a consumer needs is what the release changes, which is
the before and the after. **Its ADEP/ADES cells are blank because no run
produced them, not because a run failed.**

```{r payoff-table}
#| echo: false
pay <- do.call(rbind, lapply(PERIODS, function(p) {
  do.call(rbind, lapply(ARMS, function(m) data.frame(
    Period = p, Arm = ARMLAB[[m]],
    AdepCov = dash(100 * P(m, p, "adep_coverage")),
    AdepAcc = dash(100 * P(m, p, "adep_accuracy")),
    AdesCov = dash(100 * P(m, p, "ades_coverage")),
    AdesAcc = dash(100 * P(m, p, "ades_accuracy")),
    Score = dash(SCORE(m, p), num),
    check.names = FALSE)))
}))
kable(pay, row.names = FALSE, align = "llrrrrr",
      col.names = c("Period", "Arm", "ADEP coverage", "ADEP accuracy",
                    "ADES coverage", "ADES accuracy", "score"),
      caption = "Coverage is the share of ground-truth flights given an aerodrome at all; accuracy is the share of those that are right. `score` is correct minus twice wrong, summed over both ends — the flight list's own trade-off between naming an aerodrome and naming it wrong.")
```

**Accuracy is unchanged and coverage falls slightly.** ADEP accuracy moves
`r pp(100 * (P("standard", "2025", "adep_accuracy") - P("legacy", "2025", "adep_accuracy")))`
points on 2025 and
`r pp(100 * (P("standard", "2024", "adep_accuracy") - P("legacy", "2024", "adep_accuracy")))`
on 2024. The loss is on the arrival side: ADES coverage falls
`r mag(100 * (P("standard", "2025", "ades_coverage") - P("legacy", "2025", "ades_coverage")))`
points on 2025 and
`r mag(100 * (P("standard", "2024", "ades_coverage") - P("legacy", "2024", "ades_coverage")))`
on 2024. Over both ends the score falls
`r num(SCORE("legacy", "2025") - SCORE("standard", "2025"))` points on 2025 and
`r num(SCORE("legacy", "2024") - SCORE("standard", "2024"))` on 2024, which is
`r pct(100 * (1 - SCORE("standard", "2025") / SCORE("legacy", "2025")))` and
`r pct(100 * (1 - SCORE("standard", "2024") / SCORE("legacy", "2024")))` of the
total.

**The mechanism is merging, the only failure mode the new rule makes worse.** A
merged track holds two legs and gets one arrival aerodrome named for it, so the
second leg's arrival is not named at all — a coverage loss rather than an
accuracy loss, which is what the table shows. `standard` merges
`r pct(P("standard", "2025", "merged_pct"))` of flights on 2025 against
`legacy`'s `r pct(P("legacy", "2025", "merged_pct"))`, and that difference is
where the arrival coverage went.

**Why the downstream effect is so much smaller than the segmentation gain.**
Read the denominator. Of `r num(P("legacy", "2025", "n_ground_truth"))`
ground-truth flights on 2025, `legacy` already names an ADEP for
`r num(NAMED("legacy", "2025", "adep"))` of them and `standard` for
`r num(NAMED("standard", "2025", "adep"))`. The flight list resolves an
aerodrome from the samples near the ground at either end of a track, and a
fragmented flight usually keeps those samples in *some* fragment — the piece
holding the departure roll still sees the departure aerodrome. Segmentation
moves the flights where that is not true, a small share of a large denominator.
A `r pp(GAIN("2025"))`-point gain in clean matching beside a
`r pct(100 * (1 - SCORE("standard", "2025") / SCORE("legacy", "2025")))` change
in aerodrome score is not a null result; it is two metrics measuring different
things, one of which was already close to its ceiling.

The gain is real for anything that consumes a track rather than a flight-list
row. Events, taxi phases and every cumulative measurement are computed per track
and are wrong in exactly the way segmentation was wrong.

## What breaks {#sec-continuity}

::: {.callout-important}
## A `track_id` is not a flight

It is the identifier of a partition, and this release repartitions. Every
`track_id` published before it and every `track_id` published after it are drawn
from disjoint sets. A join across the boundary returns an empty result rather
than an error, which reads as missing data rather than as a broken key — and
nothing in the data will tell you, because `osn_tracks` carries no column saying
which rule built a row.
:::

**No identifier survives, and no run measured that.** It is true by
construction, from two facts in `segmentation/methods.py`. `legacy` hashes
`icao24 || callsign` and suffixes `_{year}_{month}`; `standard` hashes `icao24`
alone and suffixes a sample offset. Different hash inputs and different suffix
shapes mean the count of identifiers common to the two releases is zero, and no
measurement could return anything else.

**The partition change was measured.**

```{r continuity-table}
#| echo: false
co <- do.call(rbind, lapply(PERIODS, function(p) data.frame(
  Period = p,
  Legacy = num(P("legacy", p, "n_tracks")),
  Standard = num(P("standard", p, "n_tracks")),
  Change = pct(-DROP(p)),
  V1 = pct(-V1_DROP(p)),
  Exported = pct(-EXPDROP(p)),
  check.names = FALSE)))
kable(co, row.names = FALSE, align = "lrrrrr",
      col.names = c("Period", "legacy tracks", "standard tracks",
                    "change", "same change, V1 harness",
                    "change in tracks written"),
      caption = "Tracks carrying at least one ground-truth sample. The last column is the same comparison over every track step 02 wrote, whose denominator includes tracks no ground-truth flight touches.")
```

The number of tracks falls `r pct(DROP("2025"))` on 2025 and
`r pct(DROP("2024"))` on 2024, and V1's harness measured the same reduction
independently at `r pct(V1_DROP("2025"))` and `r pct(V1_DROP("2024"))`. Both
arms see the same airframes, so tracks-per-airframe moves by the same ratio: a
consumer counting tracks per aircraft per day sees that proportional fall, and
none of the tracks it loses are lost flights.

**What a consumer has to do.** Re-key anything that joins on `track_id` across
the boundary, and treat the release date as the discriminator between the two
regimes, because the data carries no other. Re-processing a past month with
`segmentation.method = "legacy"` reproduces that month's identifiers; running
the default over it does not.

::: {.callout-warning}
## The cleaned table cannot reproduce its own identifiers

`track_id` is assigned in step 02 from **raw** altitudes. Step 02a then masks
the values that fail its filters to NULL, including
`r pct(P("standard", "2025", "null_baro_pct"))` of `baro_altitude` on 2025 and
`r pct(P("standard", "2024", "null_baro_pct"))` on 2024.

Legacy's low-altitude split tests `altitude < 5,000 ft`, and a NULL compared
against 5,000 is neither true nor false, so re-deriving the partition from
`osn_tracks_clean` misses splits the original run made. Anyone doing so gets a
different answer and concludes the segmentation is non-deterministic. It is not.
The input it was computed from is no longer in the table being read.

Reproducing a published `track_id` needs `segmentation.method = "legacy"` **and**
the raw altitudes, which means re-ingesting rather than reading the cleaned
table.
:::

## What ships {#sec-defaults}

```{r defaults-table}
#| echo: false
df <- data.frame(
  Setting = c("`segmentation.method`", "`segmentation.gap_minutes`",
              "`segmentation.low_alt_gap_minutes`", "`segmentation.low_alt_ft`",
              "`segmentation.callsign_lookback_minutes`"),
  Value = c(paste0("`\"", SHIP$method, "\"`"), num(SHIP$gap), num(SHIP$low_gap),
            num(SHIP$low_ft), paste0("`", SHIP$lookback, "`")),
  Means = c(
    "Group on `icao24`; break on a gap, or on a genuine non-blank callsign change. `\"legacy\"` stays selectable and reproduces published identifiers.",
    "Split when the gap to the previous sample exceeds this, at any altitude.",
    "Split on this shorter gap when the aircraft is low.",
    "The altitude below which the shorter gap applies.",
    "Bound on how far back the callsign comparison looks. `None` follows `gap_minutes`, so the lookback cannot reach across a gap break into the previous flight."
  ),
  check.names = FALSE)
kable(df, row.names = FALSE, align = "lll",
      col.names = c("Setting", "Ships as", "What it means"))
```

The three gap thresholds are unchanged from `legacy`. V1 swept them over
`r num(SWEEP_CELLS)` combinations and found the shipped cell within a point of the best cell
anywhere, with the ordering of the one axis worth moving reversing between the
two periods — so nothing was retuned, and this release changes the group key and
the break condition only.

**Re-running a past month with the defaults will not reproduce it.** It will
produce the same aerodromes for substantially the same flights, and different
identifiers for all of them.

## Limitations

**One day per period.** Every rate on this page is 2025-06-05 or 2024-06-05, run
whole through the production steps. Two independent days agreeing, and agreeing
with V1's three-day harness sample, is the whole of the evidence that the result
is not a property of a day.

**Five of V1's eight rules were not re-run.** The pipeline arms are `legacy`,
`airframe_only` and `standard`. The ground-anchored, airport-anchored,
vertical-profile, month-suffix and traffic-style rules exist only as harness
measurements, and nothing here says how they would behave in production.

**Step 04 was not measured.** Events and measurements inherit the new partition,
and no arm here scored them. The airport-event change described in
[what changed](#sec-changed) is a code change with a stated mechanism, not a
measured delta.

**Ground truth is Network Manager flights only.** A track that merges a Network
Manager flight with a military or general-aviation flight the ground truth does
not carry is scored as clean, so the merge rates here are lower bounds.

**V1's figures are a second measurement, not a baseline.** Where the
cross-validation sets V1 beside V2 it is comparing two answers to one question
over different samples. Neither column is the other's error bar.

**Published data carries no marker of which segmentation produced it.**
Distinguishing the two regimes depends on knowing the release date. This is
stated three times on this page because it is the failure a consumer cannot
detect from the data.

**The release's safety guards are keyed on configuration, because the data
carries no provenance.** Both `flights.py` and `events.py` skip callsign
resolution when the run stamps a legacy version string, so a correctly
configured legacy re-run reproduces its released month byte for byte. Neither
guard can ask the question it actually needs answered — *were the tracks I am
reading built with the legacy rule?* — because no column records it. A
legacy-stamped run over an `osn_tracks` rebuilt with `standard` therefore fans
out while stamping the frozen version, and nothing in the output reveals it.
`flights.py`'s `tracks_table == "osn_tracks"` clause distinguishes raw tracks
from *clean* ones, not legacy-built from standard-built. This is the cost of
shipping without a segmentation marker, and it already affects one shipped
configuration: `benchmarks/event_bench.py`'s rungs `L00`–`L12` inherit
`events_v0.0.2` and so run unresolved.

**Production and the benchmark resolve the callsign over different rows at a
month boundary.** Production resolves inside the reader, *after* the month
filter, because moving it earlier defeats partition pruning. The benchmark
resolves over its whole redirected table. The two therefore agree for every
track that lies inside one month and can disagree for one that straddles a
boundary. No track in either sample straddles one — V1's payoff runs three June
days, V2 runs 2025-06-05 and 2024-06-05 — so nothing on this page is affected. A
whole-month comparison would meet it, which is why it is written down rather
than left for someone to rediscover.

## Provenance note {#sec-provenance}

`pipeline_2024.csv` and `pipeline_2025.csv` were produced at `opdi` commit
`2f95ec4`. Two commits landed afterward that touch a declared dependency of
both files. `e88516a` edits `benchmarks/track_pipeline_v2.py` to pass
`args.period` instead of the `PERIODS` dict into `extents_name`, fixing the
filename of the extents export — an output that same commit stops declaring.
`090db8a` adds three paragraphs to a docstring in `benchmarks/track_score.py`
and changes no executable line. Neither touches the code that computes
`n_tracks`, `clean_match_pct`, or any other value these two CSVs carry.

`benchmarks/provenance.py` fingerprints a declared dependency by hashing its
full file bytes, so it cannot tell a filename argument or a docstring from a
change to the arithmetic. Both commits mark `pipeline_2024.csv` and
`pipeline_2025.csv` stale, and `--check` reports them that way. Reproducing
them to clear the flag costs about 3.8 hours per period, 7.6 hours total, and
would write files identical to the ones committed here, because nothing that
ran actually changed. This page renders with `OPDI_RENDER=skip` and reports
these two outputs as staged rather than paying that cost for an identical
result.
