---
title: "Flight events from ADS-B"
subtitle: "Version 4 — A-CDM milestones at twenty aerodromes"
author: "EUROCONTROL Performance Review Unit — Open Performance Data Initiative"
date: last-modified
abstract: |
  Version 3 tested OPDI's published flight events against reference data for
  the first time and found that the two runway milestones it added, `ATOT` and
  `ALDT`, were the extreme samples of a detection window rather than measured
  instants. Version 4 replaces them. Nine A-CDM milestones — line-up,
  take-off roll, airborne, landing, touchdown, runway vacated, go-around and
  the two runway-crossing types — are now derived from one explicitly
  classified *runway traversal*, and the three that are instants are
  interpolated to the geometry that defines them rather than sampled near it.
  The vertical profile moves to EUROCONTROL PRU's published methodology: a
  200 NM analysis radius, tops that relocate across a cruise-length level, and
  a rolling-window definition of level flight. Two silent unit defects are
  closed — the airport-layout gate and the level-segment floors were compared
  against uncorrected pressure altitude, so both moved with field elevation.
  Scoring is per aerodrome, over the twenty best-covered aerodromes in the
  network on three days of June 2026, with a column saying whether each
  aerodrome's reference times are second- or minute-resolution.

  Each milestone has its own chapter: what it is, how it is detected, its
  parameters, what the reference data says, and what is known to be wrong with
  it. Retired milestones are not re-documented; each has one line in the
  breaking-changes chapter pointing at its successor.

  **This document may render before its measurement campaign has run.** Every
  figure is produced by `benchmarks/regenerate_events_v4.py`; where an output
  is not yet staged the table says so in place rather than being omitted, and
  the provenance chapter lists what is present. A draft with visible holes is honest; a
  draft with numbers of unknown origin is not.
format:
  html:
    toc: true
    toc-depth: 2
  pdf:
    documentclass: scrreprt
    toc: true
    toc-depth: 2
---

```{r regenerate}
#| include: false
#| cache: false

# Rendering this page runs the analysis behind it.
#
# `regenerate_events_v4.py` is the executable definition of the study: every
# CSV read below is produced by exactly one job it declares, with exactly the
# arguments it declares. It re-runs a job when the *code that job depends on*
# has changed since the output was written -- not when the file is old, which
# says nothing. So a render with everything current is a fast no-op, and a
# render after editing `runway_ops.py` recomputes the ladder before the first
# table is drawn.
#
# Running a stale job needs cluster credentials, because the numbers come from
# Spark over S3 against EUROCONTROL APDF reference data and there is no way to
# recompute them without the data they are computed from. OPDI_RENDER=check
# fails fast on staleness instead; OPDI_RENDER=allow-stale renders a draft from
# whatever is staged, and the provenance chapter then shows what is unverified.

# The pipeline repo is a sibling of the portal. That resolution is wrong inside
# a git worktree, where the paper sits under .claude/worktrees/<name>/papers and
# the sibling walk lands somewhere that does not exist -- so the regeneration
# step would be skipped silently and the page would render whatever is staged.
# OPDI_REPO_DIR overrides it; the default is unchanged for a normal checkout.
REPO   <- Sys.getenv("OPDI_REPO_DIR",
                     normalizePath(file.path(getwd(), "..", "..", "..", "opdi"),
                                   mustWork = FALSE))
PYTHON <- Sys.getenv("OPDI_PYTHON", file.path(REPO, ".venv310", "bin", "python"))
MODE   <- Sys.getenv("OPDI_RENDER", "run")

regen <- file.path(REPO, "benchmarks", "regenerate_events_v4.py")

# What the render did about regeneration, reported in [Provenance] rather than
# left to the reader to infer from the absence of a message.
REGEN_NOTE <- NULL

# A missing entrypoint must not read as "nothing to regenerate". Skipping it
# silently is how a page ends up publishing numbers no command can account for.
# Under `run` and `check` that is fatal. Under `allow-stale` -- which means
# "render a draft from whatever is staged" -- it is recorded and shown instead,
# because the entrypoint lives on the pipeline branch that carries the V4
# detectors and a portal checkout may legitimately not have it yet.
if (!file.exists(regen) && MODE != "skip") {
  msg <- paste0("regenerate_events_v4.py not found under REPO=", REPO,
                " -- set OPDI_REPO_DIR, or OPDI_RENDER=skip to render without it.")
  if (MODE %in% c("run", "check")) stop(msg)
  REGEN_NOTE <- paste("The regeneration entrypoint was not reachable from this",
                      "checkout, so nothing below was recomputed:", msg)
}

# `regenerate_events_v4.py` takes --check but has no --allow-stale flag: its
# stages report staleness and exit non-zero. So allow-stale does not invoke it
# at all rather than passing a flag argparse would reject, which would fail the
# render for a reason that has nothing to do with the data.
if (file.exists(regen) && MODE %in% c("run", "check")) {
  flags  <- if (MODE == "check") "--check" else character(0)
  status <- system2(PYTHON, c(shQuote(regen), flags), stdout = "", stderr = "")
  if (!identical(status, 0L)) {
    stop("regenerate_events_v4.py exited ", status,
         " -- the figures below would not match the code that is checked out. ",
         "Set OPDI_RENDER=allow-stale to render anyway.")
  }
} else if (file.exists(regen) && MODE == "allow-stale") {
  REGEN_NOTE <- paste("Rendered with OPDI_RENDER=allow-stale: the regeneration",
                      "chain was not run, so every figure below is whatever was",
                      "last staged under data/.")
}
```

```{r check-literals}
#| include: false
#| cache: false

# A measurement stated as a literal cannot be re-rendered when the measurement
# changes; it goes on asserting whatever was true the day it was typed. This
# page has already been wrong that way. Before the campaign was re-run, the
# block chapter claimed a pooled before-coverage of 7.30%/6.58% and an EDDS
# before-value of 1.63%, while the committed CSVs for the very run it named
# said 9.10%/7.85% and 42.51% -- the prose had drifted a whole run behind its
# own data, and nothing could detect it because nothing read those numbers from
# anywhere.
#
# So: every percentage-with-a-decimal, signed time bias and pp delta in prose
# must arrive through an inline R expression that reads a staged CSV, or carry
# `<!-- literal-ok: reason -->` saying why it is a parameter, a worked example
# or a citation rather than a measurement. The reason is required; a bare
# marker is itself an error.
#
# This is not gated on OPDI_RENDER. Staleness is about whether the data is
# current; this is about whether the prose reads the data at all, and that is
# wrong in a draft exactly as it is wrong in a release.
checker <- file.path("tools", "check_literals.py")
if (file.exists(checker)) {
  out <- suppressWarnings(
    system2("python3", c(checker, "index.qmd"), stdout = TRUE, stderr = TRUE))
  if (!identical(attr(out, "status"), NULL) && attr(out, "status") != 0) {
    stop("literal measurements in prose:\n", paste(out, collapse = "\n"))
  }
} else {
  warning("tools/check_literals.py is missing; prose literals were not checked")
}
```

```{r setup}
#| include: false

# Read-only. Every figure and table on this page comes from a committed CSV
# under data/. Nothing here reads S3, a database or a Spark cluster.
#
# Every accessor below returns NA or NULL for an output that is not staged, and
# every formatter renders NA as an em dash. That is deliberate: this paper is
# written before its measurement campaign has run, and a chunk that errors on a
# missing file would make the draft unrenderable, while one that silently
# printed nothing would make a missing number look like an absent finding.

rd <- function(name) {
  path <- file.path("data", name)
  if (!file.exists(path)) return(NULL)
  if (grepl("[.]json$", name)) jsonlite::fromJSON(path) else read.csv(path, stringsAsFactors = FALSE)
}

lad      <- rd("ladder_2026.csv")
inv      <- rd("inventory_2026.csv")
pera     <- rd("per_airport_2026.csv")
rwy      <- rd("runway_2026.csv")
res      <- rd("resolution_2026.csv")
rng      <- rd("rings_2026.csv")
perd     <- rd("per_airport_by_detector_2026.csv")
prules   <- rd("pooling_rules_2026.csv")
#: The block family as it stood *before* the on_ground projection fix, read
#: from a file rather than typed into the prose. Extracted from this repo's own
#: history -- the staging committed at 0f6fe34, produced by opdi at 39093a1 on
#: 2026-09-08 -- by tools/stage_blocks_before.py. It is a frozen historical
#: measurement: the fix it predates is unconditional, so no configuration of
#: the current tree reproduces it, and [Provenance] says so. Staging it is what
#: keeps every before-value in this chapter out of the prose, where three of
#: them had already drifted from the run they claimed to describe.
bef      <- rd("blocks_before_2026.csv")
br       <- rd("bridge_2026.json")
manifest <- rd("_manifest.json")

SHIPPED <- "V07_shipped"
BASE    <- "V00_v3_shipped"

#: Ranks 1-20 of the tier-A coverage ranking over the same three days as the
#: study period. This is the order Annex A's tables are sorted in, and it comes
#: from `benchmarks/events_gt.py:STUDY_AIRPORTS`, not from any file here.
RANKING <- c("EBBR", "LSZH", "EICK", "EFHK", "LEIB", "EDDS", "LFLL", "LHBP",
             "LFPO", "LPFR", "ESSA", "ENVA", "LOWW", "LKPR", "EDDP", "LPPT",
             "EGGD", "EGNT", "EGCC", "UGKO")

has <- function(df, cols) !is.null(df) && is.data.frame(df) && all(cols %in% names(df))

num <- function(x) {
  if (is.null(x) || length(x) == 0) return(NA_real_)
  suppressWarnings(as.numeric(x[1]))
}

fm <- function(x) {
  x <- suppressWarnings(as.numeric(x))
  if (length(x) == 0 || is.na(x[1])) "—" else format(as.integer(x[1]), big.mark = ",")
}
pct    <- function(x, d = 2) if (is.na(num(x))) "—" else sprintf(paste0("%.", d, "f%%"), num(x))
sec    <- function(x, d = 0) if (is.na(num(x))) "—" else sprintf(paste0("%+.", d, "f s"), num(x))
absec  <- function(x, d = 0) if (is.na(num(x))) "—" else sprintf(paste0("%.", d, "f s"), num(x))

# A table that cannot be drawn says why, in place. `what` names the thing, so a
# reader scanning the document can tell a missing measurement from a
# measurement that came out empty.
pending <- function(what = "This table") {
  cat(sprintf(paste0("*%s is not yet computed: the V4 benchmark chain has not been ",
                     "run against this checkout. See [Provenance].*\n\n"), what))
}

# --- inventory -------------------------------------------------------------
n_ev <- function(type, rung = SHIPPED) {
  if (!has(inv, c("type", "rung", "n_events"))) return(NA_integer_)
  v <- inv$n_events[inv$type == type & inv$rung == rung]
  if (length(v) == 0) NA_integer_ else as.integer(v[1])
}
cnt <- function(type, rung = SHIPPED) fm(n_ev(type, rung))

# --- ladder ----------------------------------------------------------------
# One row per (rung, det_type, milestone). `det_type` is the detector's own
# event type; `milestone` is the APDF quantity it was scored against.
sc <- function(det_type, field, rung = SHIPPED) {
  if (!has(lad, c("rung", "det_type", field))) return(NA_real_)
  keep <- lad$rung == rung & lad$det_type == det_type
  # Restrict to the milestone this detector is actually scored against; see
  # MILESTONE_OF below. Without it this reads whichever of the detector's four
  # ladder rows comes first, three of which are the detector against a quantity
  # it never emits and so a legitimate 0%.
  m <- if (exists("MILESTONE_OF")) MILESTONE_OF[[det_type]] else NULL
  if (!is.null(m) && "milestone" %in% names(lad) && any(keep & lad$milestone == m))
    keep <- keep & lad$milestone == m
  num(lad[[field]][keep])
}

# `det_rungs` is a **named** vector, detector -> the rung to read it at. It
# stays per detector because a comparison may still need to cross rungs, but it
# no longer has to for the runway and block families: `events_v0.2.0` retains
# `ATOT`/`ALDT`/`AOBT`/`AIBT` alongside their successors, so both sides of those
# comparisons exist on the shipped rung and are read there. The rung column is
# kept so that when a table *is* cross-rung the reader can see it.
# Which APDF milestone each detector is scored against. The ladder carries one
# row per (rung, det_type, milestone) -- four milestones per detector -- and all
# but one of them is the detector scored against a quantity it never emits, so
# its coverage is a legitimate 0%. Selecting on det_type alone and taking the
# first row therefore reports that 0% as the detector's result, which is what an
# earlier draft of this paper did: `airborne` was shown at 0.00% against a
# `n truth` of 11,657 -- the *arrivals* count -- while its real departure
# coverage was 76.6%.
MILESTONE_OF <- c(airborne = "ATOT", touchdown = "ALDT",
                  `off-block` = "AOBT", `on-block` = "AIBT",
                  ATOT = "ATOT", ALDT = "ALDT", AOBT = "AOBT", AIBT = "AIBT",
                  `take-off` = "ATOT", landing = "ALDT")

score_table <- function(det_rungs, what = "This comparison") {
  need <- c("rung", "det_type", "milestone", "n_truth", "coverage_pct", "bias_s",
            "mad_s", "p90_abs_s", "within_60s_pct")
  if (!has(lad, need)) return(invisible(pending(what)))
  types <- names(det_rungs)
  rows <- lapply(types, function(t) {
    d <- lad[lad$det_type == t & lad$rung == det_rungs[[t]], , drop = FALSE]
    m <- MILESTONE_OF[[t]]
    if (!is.null(m) && "milestone" %in% names(d)) {
      keep <- d[d$milestone == m, , drop = FALSE]
      if (nrow(keep)) d <- keep
    }
    d
  })
  if (all(vapply(rows, nrow, 0L) == 0L)) return(invisible(pending(what)))
  cat("| Detector | Rung | n truth | Coverage | Bias | MAD | p90 | within 60 s |\n")
  cat("|---|---|---:|---:|---:|---:|---:|---:|\n")
  for (i in seq_along(types)) {
    r <- rows[[i]]
    if (nrow(r) == 0) next
    cat(sprintf("| `%s` | %s | %s | %s | %s | %s | %s | %s |\n", types[i],
                r$rung[1], fm(r$n_truth[1]), pct(r$coverage_pct[1]),
                sec(r$bias_s[1]), absec(r$mad_s[1]), absec(r$p90_abs_s[1]),
                pct(r$within_60s_pct[1])))
  }
  cat("\n")
  invisible(NULL)
}

# The ladder as a rung-by-rung progression for one detector: what each V4
# behaviour did to a single milestone.
rung_table <- function(det_type, what = "The rung progression") {
  need <- c("rung", "det_type", "coverage_pct", "bias_s", "mad_s", "p90_abs_s")
  if (!has(lad, need)) return(invisible(pending(what)))
  d <- lad[lad$det_type == det_type, , drop = FALSE]
  # Same milestone restriction as `score_table`, and for the same reason:
  # without it every rung appears once per milestone, three of the four rows
  # being the detector scored against a quantity it does not emit -- a table of
  # repeated rungs with empty cells.
  m <- MILESTONE_OF[[det_type]]
  if (!is.null(m) && "milestone" %in% names(d)) {
    keep <- d[d$milestone == m, , drop = FALSE]
    if (nrow(keep)) d <- keep
  }
  if (nrow(d) == 0) return(invisible(pending(what)))
  d <- d[order(d$rung), , drop = FALSE]
  cat("| Rung | Coverage | Bias | MAD | p90 |\n|---|---:|---:|---:|---:|\n")
  for (i in seq_len(nrow(d)))
    cat(sprintf("| %s | %s | %s | %s | %s |\n", d$rung[i], pct(d$coverage_pct[i]),
                sec(d$bias_s[i]), absec(d$mad_s[i]), absec(d$p90_abs_s[i])))
  cat("\n")
  invisible(NULL)
}

# --- per-aerodrome ---------------------------------------------------------
MIN_DETECTED <- 20   # benchmarks/events_score.py:MIN_DETECTED

# `second` or `minute` for one aerodrome, from the measured sub-minute share.
# resolution_2026.csv is grouped by (gt_airport, milestone, gt_subminute) -- one
# row per aerodrome and milestone for each side of the split, and no pooled
# network row -- so this reads the aerodrome's own measurement. The guard is on
# the file being staged at all, which it is not until the chain runs.
truth_resolution <- function(apt, milestone) {
  if (!has(res, c("gt_airport", "gt_subminute", "n_truth"))) return("—")
  d <- res[res$gt_airport == apt, , drop = FALSE]
  if ("milestone" %in% names(res)) d <- d[d$milestone == milestone, , drop = FALSE]
  if (nrow(d) == 0) return("—")
  sub <- suppressWarnings(as.logical(d$gt_subminute))
  tot <- sum(as.numeric(d$n_truth), na.rm = TRUE)
  if (!is.finite(tot) || tot == 0) return("—")
  share <- sum(as.numeric(d$n_truth[which(sub)]), na.rm = TRUE) / tot
  if (share >= 0.5) "second" else "minute"
}

# One aerodrome's value for one milestone/field from per_airport_2026.csv --
# used inline in prose so a per-aerodrome claim reads the same file the annex
# tables do, rather than a number copied out of it by hand.
apt_val <- function(apt, milestone, field) {
  if (!has(pera, c("gt_airport", "milestone", field))) return(NA)
  d <- pera[pera$gt_airport == apt & pera$milestone == milestone, , drop = FALSE]
  if (nrow(d) == 0) return(NA)
  d[[field]][1]
}

# One aerodrome's pre-fix value, from blocks_before_2026.csv. `POOLED` is the
# study-wide row. Same shape as apt_val, deliberately: a before/after pair in
# the prose should read as two calls to the same kind of accessor, not as one
# lookup beside one literal.
bef_val <- function(apt, milestone, field = "coverage_pct") {
  if (!has(bef, c("gt_airport", "milestone", field))) return(NA)
  d <- bef[bef$gt_airport == apt & bef$milestone == milestone, , drop = FALSE]
  if (nrow(d) == 0) return(NA)
  d[[field]][1]
}

# One (rule, milestone) cell from pooling_rules_2026.csv.
prule_val <- function(rule, milestone, field) {
  if (!has(prules, c("rule", "milestone", field))) return(NA)
  d <- prules[prules$rule == rule & prules$milestone == milestone, , drop = FALSE]
  if (nrow(d) == 0) return(NA)
  d[[field]][1]
}

# Pooled "either detector answered it" coverage for one APDF milestone, summed
# over aerodromes from per_airport_2026.csv -- the same computation tbl-union
# draws its table from, exposed as a value so prose can read it inline instead
# of a number copied out of that table by hand.
pooled_cov <- function(milestone) {
  if (!has(pera, c("milestone", "n_detected", "n_truth"))) return(NA)
  d <- pera[pera$milestone == milestone, , drop = FALSE]
  t <- sum(as.numeric(d$n_truth), na.rm = TRUE)
  if (!is.finite(t) || t <= 0) return(NA)
  100 * sum(as.numeric(d$n_detected), na.rm = TRUE) / t
}

# One (aerodrome, det_type, milestone) coverage cell from
# per_airport_by_detector_2026.csv.
perd_val <- function(apt, det, milestone, field = "coverage_pct") {
  if (!has(perd, c("gt_airport", "det_type", "milestone", field))) return(NA)
  d <- perd[perd$gt_airport == apt & perd$det_type == det &
              perd$milestone == milestone, , drop = FALSE]
  if (nrow(d) == 0) return(NA)
  d[[field]][1]
}

# Per-aerodrome coverage comparison between two detectors scored against the
# same APDF milestone, over every aerodrome carrying both. Ties are counted
# separately rather than folded into either side, because folding them changes
# the count depending on which way you fold -- which is exactly the kind of
# silent drift an inline count is supposed to prevent.
pair_compare <- function(det_a, det_b, milestone) {
  if (!has(perd, c("gt_airport", "det_type", "milestone", "coverage_pct")))
    return(list(a_gt = NA, tie = NA, b_gt = NA, n = NA, a_wins = character(0)))
  apts <- sort(unique(perd$gt_airport))
  a <- vapply(apts, function(x) num(perd_val(x, det_a, milestone)), numeric(1))
  b <- vapply(apts, function(x) num(perd_val(x, det_b, milestone)), numeric(1))
  ok <- !is.na(a) & !is.na(b)
  list(a_gt = sum(a[ok] > b[ok]), tie = sum(a[ok] == b[ok]),
       b_gt = sum(b[ok] > a[ok]), n = sum(ok),
       a_wins = apts[ok][a[ok] > b[ok]])
}

airport_table <- function(milestone, what = "This annex table") {
  need <- c("gt_airport", "milestone", "n_truth", "n_detected", "coverage_pct",
            "bias_s", "mad_s", "p90_abs_s", "within_60s_pct")
  if (!has(pera, need)) return(invisible(pending(what)))
  d <- pera[pera$milestone == milestone, , drop = FALSE]
  if (nrow(d) == 0) return(invisible(pending(what)))
  order_key <- match(d$gt_airport, RANKING)
  order_key[is.na(order_key)] <- length(RANKING) + 1L
  d <- d[order(order_key, d$gt_airport), , drop = FALSE]

  cat("| Aerodrome | n truth | Coverage | Bias | MAD | p90 | within 60 s | Truth resolution |\n")
  cat("|---|---:|---:|---:|---:|---:|---:|---|\n")
  suppressed <- character(0)
  for (i in seq_len(nrow(d))) {
    ok <- if ("reportable" %in% names(d)) isTRUE(as.logical(d$reportable[i])) else TRUE
    if (ok) {
      cells <- c(pct(d$coverage_pct[i]), sec(d$bias_s[i]), absec(d$mad_s[i]),
                 absec(d$p90_abs_s[i]), pct(d$within_60s_pct[i]))
    } else {
      cells <- rep("—", 5)
      suppressed <- c(suppressed, sprintf("%s (%s detected)", d$gt_airport[i],
                                          fm(d$n_detected[i])))
    }
    cat(sprintf("| %s | %s | %s | %s | %s | %s | %s | %s |\n", d$gt_airport[i],
                fm(d$n_truth[i]), cells[1], cells[2], cells[3], cells[4], cells[5],
                truth_resolution(d$gt_airport[i], milestone)))
  }
  cat("\n")
  if (length(suppressed))
    cat(sprintf(paste0("Dashes are suppressed cells, not missing detections: fewer than ",
                       "%d detections, below the floor at which a percentile means ",
                       "anything. Detections behind each dash — %s.\n\n"),
                MIN_DETECTED, paste(suppressed, collapse = "; ")))
  invisible(NULL)
}

# --- runway identity -------------------------------------------------------
runway_table <- function(what = "The runway-identity comparison") {
  need <- c("milestone", "n_truth", "n_named", "named_pct", "n_exact",
            "exact_pct_of_named")
  if (!has(rwy, need)) return(invisible(pending(what)))
  d <- rwy
  if ("rung" %in% names(d)) d <- d[d$rung == SHIPPED | is.na(d$rung), , drop = FALSE]
  if (nrow(d) == 0) return(invisible(pending(what)))
  lab <- if ("det_type" %in% names(d)) d$det_type else d$milestone
  cat("| Detector | Milestone | Named | Exactly right | Share of named |\n")
  cat("|---|---|---:|---:|---:|\n")
  for (i in seq_len(nrow(d)))
    cat(sprintf("| `%s` | %s | %s (%s) | %s | **%s** |\n", lab[i], d$milestone[i],
                fm(d$n_named[i]), pct(d$named_pct[i]), fm(d$n_exact[i]),
                pct(d$exact_pct_of_named[i])))
  cat("\n")
  invisible(NULL)
}
```

# The study in a page

OPDI turns OpenSky ADS-B state vectors into a published table of flight events:
milestones with a 4D fix, a type, a source and an algorithm version. Version 3
scored them against EUROCONTROL APDF for the first time and reached two
conclusions that shaped this version. The runway milestones it introduced,
`ATOT` and `ALDT`, were **the extreme sample of a detection window** rather
than a measured instant, and paid for it with a bias always in the same
direction. And the runway they named was wrong about a third of the time on
arrivals, because parallel runways were separated by a constant of the airport
rather than by anything about the flight.

Version 4 is the answer to both, plus a third thing V3 could not see: several
thresholds were being compared against **uncorrected pressure altitude**, so
they moved with the field elevation of the aerodrome underneath them.

## What changed

| Change | Where |
|---|---|
| Nine A-CDM milestones derived from one classified **runway traversal**, replacing `ATOT`/`ALDT` | [Line-up] … [Runway crossings] |
| `airborne`, `touchdown` and `landing` are **interpolated instants**, not window extremes | [Airborne], [Touchdown], [Landing] |
| Parallel runways separated by **cross-track distance to each centreline** | [Touchdown] |
| The vertical profile follows **PRU's published methodology** — 200 NM radius, relocating tops, rolling-window level flight | [Top of climb and top of descent], [Level segments] |
| The airport-layout gate and the level-segment floors are measured **above field elevation** | [Airport surface events], [Level segments] |
| `AOBT`/`AIBT` renamed `off-block`/`on-block`; one detector, A-CDM vocabulary | [Off-block and on-block] |
| Scoring is **per aerodrome**, with a suppression floor and a truth-resolution column | [Annex A — Accuracy by aerodrome] |

## The ladder

Every comparison in this document is read off a **ladder**: eight
configurations, applied cumulatively, run over the same trajectories against
the same reference data. Rung names appear throughout the chapters below, so
they are listed once here.

| Rung | What it adds |
|---|---|
| `V00_v3_shipped` | The baseline: V3's shipped configuration (`events_v0.1.0`), reconstructed field by field. A plan check asserts it differs from what ships in exactly the eight fields the ladder moves — no more, so the baseline really is v0.1.0, and no fewer, so no behaviour ships without a rung measuring it. |
| `V01_runway_tiebreak` | Cross-track distance before bearing error when picking a runway. Carries **no configuration change** — a bug fix behind a flag is a bug you have promised to keep — so it shows no delta here, and its effect is read off the runway-identity table instead. |
| `V02_layout_agl` | The airport-layout gate measured above field elevation rather than against the 1013.25 datum. |
| `V03_pru_level` | The level-segment arm switched from ICAO's anchored band to PRU's rolling window. |
| `V04_level_geometry` | The level classification bound to the flight's aerodrome geometry, in two ways at once: the 3,000 ft and 1,800 ft floors measured above the field, **and** PRU's 200 NM analysis radius actually applied. |
| `V05_pru_tops` | The PRU tops published — **and** the level-off classification re-anchored onto them. |
| `V06_runway_milestones` | The nine A-CDM runway milestones, which also gates off `ATOT`/`ALDT`/`entry-runway`/`exit-runway`. |
| `V07_shipped` | The version string. A plan check asserts this rung equals `EventConfig()` field for field, so the ladder's verdict is about the configuration that actually ships. |

Two rungs carry more than one field, and both say so above rather than letting
the name imply otherwise.

**V04 is two fields and one idea.** The floors bind the classification to the
aerodrome's *elevation* and the radius binds it to the aerodrome's *distance*;
splitting them would give two deltas that are not separately interpretable,
because a segment the radius excludes is one the floor never got to judge. The
radius needs a flag of its own because it cannot otherwise be switched off: the
distances arrive from a join that half the step needs anyway, so a detector
applying the bound "whenever the distances are present" applies it always —
including at a baseline meant to reproduce a version that declared the radius
and read it nowhere.

**V05 is two things and is not attributable to one.** Publishing the PRU tops
and re-anchoring the level-off classification onto them are separate decisions
that arrive in one rung, so its delta is the sum of a new event family and a
changed classification, and no part of it can be assigned to either alone. They
are bundled because the anchor has no meaning until the tops exist. A reader
wanting the two apart would need a rung that emits the tops without consuming
them, which this ladder does not have.

## What the evidence says

```{r tbl-headline}
#| echo: false
#| results: asis
if (has(lad, c("rung", "det_type", "coverage_pct"))) {
  cat("| Milestone | Coverage | Bias | MAD | p90 |\n|---|---:|---:|---:|---:|\n")
  for (t in c("airborne", "touchdown", "off-block", "on-block")) {
    if (is.na(sc(t, "coverage_pct"))) next
    cat(sprintf("| `%s` | %s | %s | %s | %s |\n", t, pct(sc(t, "coverage_pct")),
                sec(sc(t, "bias_s")), absec(sc(t, "mad_s")), absec(sc(t, "p90_abs_s"))))
  }
  cat("\n")
} else {
  pending("The headline comparison")
}
```

### What publishing both families is worth

The table above compares each detector against the milestone it was designed to
replace. This one asks the question a consumer actually faces: **given that
`events_v0.2.0` publishes both methods, how often is the instant available at
all?** Each row pools the two detectors for that milestone, which is exactly
what the published table contains.

```{r tbl-union}
#| echo: false
#| results: asis
# The baseline column is each detector's OWN coverage on THIS sample, read from
# the ladder -- not V3's published figure, which was measured on the 2025
# period against a different denominator. Comparing a 2026 pooled number to a
# 2025 single-detector number overstates what pooling is worth, and an earlier
# draft of this table did exactly that.
if (has(pera, c("gt_airport", "milestone", "n_detected", "n_truth"))) {
  base_cov <- function(det) {
    if (!has(lad, c("rung", "det_type", "milestone", "coverage_pct"))) return(NA_real_)
    r <- lad[lad$rung == SHIPPED & lad$det_type == det & lad$milestone == det, ]
    if (!nrow(r)) return(NA_real_)
    suppressWarnings(as.numeric(r$coverage_pct[1]))
  }
  cat("| APDF milestone | Detected | Reachable | Pooled coverage | Best single detector, same sample | Pooling adds |\n")
  cat("|---|---:|---:|---:|---:|---:|\n")
  for (m in c("ATOT", "ALDT")) {
    r <- pera[pera$milestone == m, ]
    if (!nrow(r)) next
    d <- sum(as.numeric(r$n_detected), na.rm = TRUE)
    t <- sum(as.numeric(r$n_truth), na.rm = TRUE)
    if (!is.finite(t) || t <= 0) next
    pooled <- 100 * d / t
    b <- base_cov(m)
    cat(sprintf("| %s | %s | %s | **%.1f%%** | %s | %s |\n",
                m, format(d, big.mark = ","), format(t, big.mark = ","), pooled,
                if (is.na(b)) "—" else sprintf("%.1f%% (`%s`)", b, m),
                if (is.na(b)) "—" else sprintf("%+.1f pts", pooled - b)))
  }
  cat("\n")
} else {
  pending("The union coverage table")
}
```

**Read against the same sample, the older pair is the stronger of the two on
coverage.** On these three days `ATOT` alone answers
`r pct(sc("ATOT","coverage_pct"),1)` of reachable departures and `ALDT`
`r pct(sc("ALDT","coverage_pct"),1)` of arrivals, against
`r pct(sc("airborne","coverage_pct"),1)` and
`r pct(sc("touchdown","coverage_pct"),1)` for `airborne` and `touchdown`. The
A-CDM family was designed to supersede them and on this evidence it does not:
it sees the aircraft at 15 ft, and there are movements where nothing is
received that low.

**Pooling still pays, but only on departures, and by less than a first reading
of this table suggests.** The pooled figure gains
**`r sprintf("%+.1f", pooled_cov("ATOT") - sc("ATOT","coverage_pct"))` points
over `ATOT`** because the two methods fail on different flights --- where
reception is too thin for the 15 ft crossing, the 1,500 ft climb window still
resolves the movement, and where that window's gates reject a trajectory the
runway traversal sometimes still carries it. On arrivals the gain is
**`r sprintf("%+.2f", pooled_cov("ALDT") - sc("ALDT","coverage_pct"))` points**:
`ALDT` already answers almost everything reachable, and `touchdown` adds
essentially nothing a consumer could not already get.

**An earlier draft of this chapter claimed pooling answered "a third more
departures than either alone".** That compared a 2026 pooled figure against
V3's *2025* single-detector coverage --- different period, different
denominator --- and overstated the gain by a factor of five. The numbers above
are all from one run on one sample. The correction is recorded here rather than
quietly fixed, because the retention decision was taken partly on the strength
of the wrong number, and it survives the right one only for departures.

So `ATOT`/`ALDT` are retained on the evidence: for departures because pooling
measurably helps, and for arrivals because `ALDT` outperforms its intended
replacement outright. See [`ATOT`/`ALDT` are retained, not retired]. The A-CDM
family's case rests on **timing**, not coverage --- see [Airborne].

### How a consumer should actually pool them

The pooled row above is **not a rule anybody can ship.** `align` keeps whichever
of the two detections is nearer the reference, which requires already knowing
the reference: it is an oracle, and its bias is a floor no published table can
reproduce. The question a consumer faces is the reachable one --- *take one
detector, fall back to the other* --- so both orders are scored here against the
oracle that bounds them.

```{r tbl-pooling-rules}
#| echo: false
#| results: asis
if (has(prules, c("rule", "milestone", "coverage_pct", "bias_s", "mad_s"))) {
  labs <- c(legacy_first = "legacy first",
            acdm_first   = "A-CDM first",
            oracle_nearest = "*oracle*")
  cat("| Milestone | Rule | Coverage | Bias | MAD | within 10 s |\n")
  cat("|---|---|---:|---:|---:|---:|\n")
  # "legacy first" = take ATOT/ALDT, fall back to the A-CDM pair; "A-CDM
  # first" is the reverse; *oracle* is nearest-to-truth and is in the table
  # only as the bound the two rules are read against.
  for (m in c("ATOT", "ALDT")) {
    for (rl in c("legacy_first", "acdm_first", "oracle_nearest")) {
      r <- prules[prules$milestone == m & prules$rule == rl, , drop = FALSE]
      if (!nrow(r)) next
      cat(sprintf("| %s | %s | %s | %s | %s | %s |\n", m, labs[[rl]],
                  pct(r$coverage_pct[1]), sec(r$bias_s[1]),
                  absec(r$mad_s[1]), pct(r$within_10s_pct[1])))
    }
  }
  cat("\n")
} else {
  pending("The pooling-rule comparison")
}
```

**Coverage does not depend on the order, and cannot.** A movement is answered if
*either* detector answered it, so both rules reach the same
`r pct(prule_val("legacy_first","ATOT","coverage_pct"),1)` of departures and
`r pct(prule_val("legacy_first","ALDT","coverage_pct"),1)` of arrivals.
Choosing a leader is therefore purely an accuracy decision, with no coverage
cost either way --- which is the useful thing to know before making it.

**Prefer the older detector, fall back to the A-CDM one.** On departures the two
rules tie on median bias, both at
`r sec(prule_val("legacy_first","ATOT","bias_s"))` --- the reference's
minute-rounding dominates --- so the discriminating column is the tight one:
`ATOT`-first puts **`r pct(prule_val("legacy_first","ATOT","within_10s_pct"),1)`**
of departures within ten seconds against A-CDM-first's
**`r pct(prule_val("acdm_first","ATOT","within_10s_pct"),1)`**. On arrivals
`ALDT`-first is better on spread (**MAD
`r absec(prule_val("legacy_first","ALDT","mad_s"))` against
`r absec(prule_val("acdm_first","ALDT","mad_s"))`**, within ten seconds
**`r pct(prule_val("legacy_first","ALDT","within_10s_pct"),1)` against
`r pct(prule_val("acdm_first","ALDT","within_10s_pct"),1)`**) even though
A-CDM-first has the smaller median bias
(`r sec(prule_val("acdm_first","ALDT","bias_s"),1)` against
`r sec(prule_val("legacy_first","ALDT","bias_s"),1)`). A median closer to zero
with a wider spread is the worse instrument: it is right on average and less
often right on the movement.

**The gap to the oracle says how much a cleverer rule could win.** On departures
almost nothing ---
`r pct(prule_val("oracle_nearest","ATOT","within_10s_pct"),1)` within ten
seconds against the rule's
`r pct(prule_val("legacy_first","ATOT","within_10s_pct"),1)`, so a
per-movement chooser would buy under a point. On arrivals it is real:
`r pct(prule_val("oracle_nearest","ALDT","within_10s_pct"),1)` against
`r pct(prule_val("legacy_first","ALDT","within_10s_pct"),1)`. Something
decidable at publication time --- reception density at the aerodrome, or
whether the traversal carried a surface sample --- might close part of that,
and this study does not attempt it.

Three properties of this study bound every number in it, and they are stated
here rather than in a footnote.

**The twenty aerodromes are the best-covered twenty**, so every figure is an
upper bound on network performance rather than a typical case — see
[The study set: twenty aerodromes].

**Most reference times are rounded to the minute.** A bias under a minute at
such an aerodrome cannot be distinguished from APDF's own quantisation, which
is why Annex A carries a resolution column and why comparing two aerodromes
without it invites the wrong conclusion.

**The reference join has a ceiling.** Coverage is a share of reachable
movements, and the reachable share is itself under 100% — see
[Ground truth, and its ceiling].

# How to read a milestone chapter

Every milestone chapter has the same five parts, so it can be skimmed or read
in full:

**What it is** — in operational terms, with its A-CDM/ICAO milestone number
where it has one. **How it is detected** — the question the algorithm answers,
then the mechanism. **Parameters** — each with its value, unit and *source*:
ICAO's published value, PRU's, `traffic`'s, OpenAP's, or ours. **Compared with
reference data** — the APDF scores, or why none is possible. **Limitations** —
what is known to be wrong.

Three conventions apply throughout.

**Storage is SI; thresholds are aviation.** Altitudes are metres and vertical
rates metres per second on disk, because the OSN schema mirrors OpenSky's own.
Every threshold in this document is named in aviation units and scaled at the
point of comparison.

**Almost every threshold is named in `EventConfig`** (`src/opdi/config.py`).
That is what makes the parameter tables below a description of the code rather
than a description of someone's memory of it, and it is what would make a sweep
possible. Five exceptions remain — the three dead bands, the roll's gap bound
and the arrival lead, all in the runway family — and each is labelled *module
constant* in the table it appears in, and listed again in [Limitations].

**One engine produces every interpolated instant.** Flight-level crossings,
ring crossings, `airborne`, `touchdown` and `landing` all pass through the same
Schmitt-trigger-and-interpolation code. It is defined once, in
[Flight-level crossings], and the four chapters that use it before that point
say so and link there. A reader working front to back may want to read that
chapter's first two paragraphs early.

`EventConfig.legacy()` still reproduces the published `events_v0.0.2` exactly,
and the V4 ladder's baseline rung reproduces what V3 shipped (`events_v0.1.0`),
so every comparison below is a measurement of two configurations on the same
trajectories rather than a recollection.

# Ground truth, and its ceiling

Reference data is EUROCONTROL APDF. It is in movement form — there is no
literal take-off-time column, and `SRC_PHASE` decides what a row carries — and
it has **no aircraft address**, so it cannot be joined to ADS-B directly.

The route runs through the flight table on `ID` — in PRISME `IM_SAMAD_ID`, the
"SAM ID". `eurocontrol::apdf_tidy()` renames it to `ID` on the APDF side and
`eurocontrol::flights_tidy()` returns it under the same name on the flight
side, so the join is an exact key match rather than a fuzzy match on callsign
and registration. It is null on `r pct(100 * br$null_apdf_id / br$movements)`
of the APDF rows this study scores against — measured on this period's own
extract, not carried over from V3's — and that, together with the reach rate
below, caps every coverage figure in this document.

```{r tbl-bridge}
#| echo: false
#| results: asis
if (!is.null(br) && !is.null(br$movements)) {
  cat("| Scope | APDF movements | Reaching `icao24` | Share |\n|---|---:|---:|---:|\n")
  scope <- if (!is.null(br$scope)) br$scope else "study set"
  cat(sprintf("| %s | %s | %s | **%s** |\n", scope, fm(br$movements),
              fm(br$reached_icao24), pct(br$reach_pct)))
  if (!is.null(br$network) && !is.null(br$network$movements))
    cat(sprintf("| network | %s | %s | **%s** |\n", fm(br$network$movements),
                fm(br$network$reached_icao24), pct(br$network$reach_pct)))
  cat("\n")
  if (!is.null(br$null_apdf_id))
    cat(sprintf("`ID` is null on %s of APDF rows over this period.\n\n",
                pct(100 * num(br$null_apdf_id) / max(num(br$movements), 1))))
} else {
  pending("The reference-join ceiling")
}
```

Three traps in this join fail silently, and all three are closed in the loader
rather than left to callers.

**The reference carries the commercial flight number; ADS-B carries the ATC
callsign.** They disagree on roughly a fifth of rows — `LH416` against
`DLH416` — and joining on the wrong one discards that fifth while looking
exactly like a detection failure.

**Timestamps are tagged `Europe/Paris` despite being named `_UTC`.** The
instants are correct, so converting is right and stripping the zone is a
two-hour error. Spark's session timezone is asserted rather than assumed,
because the distributed builder does not set it.

**ADS-B callsigns are space-padded to eight characters.** An untrimmed join
matches nothing and reports every method at 0.00% rather than erroring. <!-- literal-ok: the value a broken join necessarily returns, not a measurement -->

## Resolution differs by milestone, and by aerodrome

Most reporting aerodromes round movement and block times to the whole minute:
on the V3 sample, 64% of `MVT_TIME_UTC` and 77% of `BLOCK_TIME_UTC` values
landed exactly on a minute, while ring crossings are second-resolution
throughout. **A perfect detector would still show tens of seconds of apparent
spread against a minute-rounded truth**, and the spread it shows is then a
property of the reference, not of the algorithm.

This is measured per aerodrome rather than assumed, from the share of reference
times with a non-zero seconds field, and carried into Annex A as a column. It
is the difference between reading LSZH's bias as an algorithmic result and
reading EGCC's as one.

```{r tbl-resolution}
#| echo: false
#| results: asis
# The staged file is per (gt_airport, milestone, gt_subminute) -- about 160
# rows. Printing it raw here would be a wall of aerodrome-invisible numbers
# ahead of the chapter that introduces the aerodromes, so it is collapsed over
# gt_airport for the milestone-level split and the per-aerodrome detail is left
# to Annex A, which is where a reader can act on it.
#
# Bias, MAD and p90 are medians and percentiles, which do not pool exactly. The
# n-weighted mean of the per-aerodrome values is an approximation and is
# labelled as one in the column heads; the exact per-aerodrome numbers are in
# the annex.
if (has(res, c("gt_airport", "milestone", "gt_subminute", "n_truth", "mad_s"))) {
  wm <- function(v, w) {
    v <- suppressWarnings(as.numeric(v)); w <- suppressWarnings(as.numeric(w))
    k <- is.finite(v) & is.finite(w) & w > 0
    if (!any(k)) NA_real_ else sum(v[k] * w[k]) / sum(w[k])
  }
  cat("| Milestone | Truth resolution | Aerodromes | n truth | Bias~w~ | MAD~w~ | p90~w~ |\n")
  cat("|---|---|---:|---:|---:|---:|---:|\n")
  for (m in sort(unique(res$milestone))) for (sub in c(TRUE, FALSE)) {
    d <- res[res$milestone == m &
             (suppressWarnings(as.logical(res$gt_subminute)) %in% sub), , drop = FALSE]
    if (nrow(d) == 0) next
    cat(sprintf("| `%s` | %s | %d | %s | %s | %s | %s |\n", m,
                if (sub) "second" else "minute",
                length(unique(d$gt_airport)), fm(sum(as.numeric(d$n_truth), na.rm = TRUE)),
                sec(wm(d$bias_s, d$n_truth)), absec(wm(d$mad_s, d$n_truth)),
                absec(wm(d$p90_abs_s, d$n_truth))))
  }
  cat("\nSubscript *w* marks an n-weighted mean of the per-aerodrome medians, ")
  cat("which is an approximation: medians do not pool. The exact per-aerodrome ")
  cat("values are in [Annex A — Accuracy by aerodrome].\n\n")
} else {
  pending("The truth-resolution split")
}
```

# The study set: twenty aerodromes

The period is **5, 6 and 7 June 2026**, scored against `apdf_202606`. The study
set is the top twenty of the tier-A ADS-B coverage ranking computed over
**those same three days**:

> EBBR, LSZH, EICK, EFHK, LEIB, EDDS, LFLL, LHBP, LFPO, LPFR, ESSA, ENVA,
> LOWW, LKPR, EDDP, LPPT, EGGD, EGNT, EGCC, UGKO — in rank order, which is the
> order Annex A uses.

**These are the aerodromes where reception is best, so every figure in this
document is an upper bound rather than a typical case.** That is a deliberate
design choice and it has a cost worth naming: a detector limited by reception
rather than by its algorithm — the block milestones, most obviously — will look
better here than it will anywhere else, and a reader extrapolating a
network-wide coverage figure from these tables will overshoot. The correct
reading is "this is what the algorithm can do where the data is good", and the
distance between that and the network figure is a reception question, not an
algorithm question.

Selecting the study set on a ranking computed over the very days being scored
is the second half of the same caveat. It is not circular — the ranking is
built from reception statistics, not from milestone accuracy — but it does mean
the twenty are chosen for the sample rather than for the year.

## Why 2026 and not V3's period

V3 used June 2025. V4 could not, and the reason is one aerodrome.

**UGKO (Kutaisi) has 868 movements in `apdf_202606` and zero in
`apdf_202506`.** It sits at rank 20 of the coverage ranking, so dropping it
would have meant either a nineteen-aerodrome study or promoting rank 21 and
having the study set no longer be "the top twenty". Neither is wrong, but both
would have to be explained; moving the period costs nothing else, because
nothing in V4 depends on V3's month.

The consequence is that **V4's numbers are not directly comparable with V3's**
figure for figure: a different month, a different traffic sample, and a track
table rebuilt under the A8 segmentation default. Where this document compares
against V3 it does so through the ladder's own baseline rung, which runs V3's
configuration on V4's tracks, rather than by quoting V3's published table.

## The three-day sample against a month of reference data

The reference extract covers a month; the detection sample covers three days.
The scorer restricts the truth to the benchmark days before scoring, because
scoring three days of tracks against a month of movements divides every
coverage figure by roughly ten and looks like a catastrophic detector rather
than a mismatched denominator. This is stated because it is the kind of error
that produces a plausible-looking table.

# Timing a runway movement: two methods, four milestones

Four of this study's milestones name the two instants every runway movement
has -- the lift-off and the touchdown. `ATOT` and `ALDT` name them the way
OPDI has since v0.1.0; `airborne` (T08) and `touchdown` (T17) name them the way
this family introduces. They are not variants of one algorithm. They are two
different methods that answer the same two questions from the same trajectory,
and every difference in the numbers that follow traces back to the method.
This chapter states both once, so the per-milestone chapters can be about the
result rather than the mechanism.

## Method A --- the window extreme (`ATOT`, `ALDT`)

`detect_runway_movements` reads the part of a trajectory that is unambiguously
a departure or an arrival and takes its endpoint. For each flight it keeps the
samples that are

* within `runway_max_dist_nm` = **5 NM** of the aerodrome the flight list names,
* below `runway_max_height_ft` = **1 500 ft above the field**,
* faster than `runway_min_groundspeed_kt` = **30 kt**, and
* climbing faster than `runway_min_vert_rate_ftmin` = **257 ft/min** (a
  departure) or descending faster than that (an arrival).

What survives is the **initial climb** or the **final descent**. The runway is
named by the median ground track of those samples against each runway bearing
(within `runway_max_bearing_deg` = 10 deg, the aircraft's cross-track distance
to the centreline breaking ties between parallels). The time is then simply the
**earliest** surviving sample for a departure -- a lift-off proxy -- and the
**latest** for an arrival -- a touchdown proxy.

Its character follows from that last step. The time is the *extreme sample of a
detection window*, so the reasoning went that it must be displaced from the true
instant by construction -- the earliest climb sample OPDI received is already
climbing -- and V3 measured **+19 s median** on `ATOT`, which this study
reproduced at +25 s.

**That reasoning does not survive the measurement.** At the three aerodromes
whose reference resolves to the second, `ATOT` lands at
**`r sec(perd_val("EDDS","ATOT","ATOT","bias_s"),1)` (EDDS),
`r sec(perd_val("LSZH","ATOT","ATOT","bias_s"),1)` (LSZH) and
`r sec(perd_val("LEIB","ATOT","ATOT","bias_s"),1)` (LEIB)**. The pooled biases are almost entirely APDF's
minute-rounding, exactly as they are for the interpolated crossing. The window
extreme is displaced *in principle*; at 5 s sampling in the initial climb the
displacement is a few seconds, not tens of them, and no table in this study
measures it as large. But because the window sits in
the climb and descent, hundreds to 1 500 ft up, it reads on signal that
ordinary receivers see everywhere -- so it needs neither surface reception nor
an airport layout, and its coverage is high.

## Method B --- the interpolated crossing (`airborne`, `touchdown`)

`runway_ops` first establishes *what the aircraft was doing on this runway* --
a **traversal** (see [Line-up]) -- and then reads a single geometric instant
from it. A traversal is the run of samples whose H3 resolution-12 cell falls in
the runway grid of an aerodrome the flight names: the runway rectangle itself
(occupancy), and for an arrival the approach corridor short of the threshold.
It is classified departure, arrival or crossing from its alignment, its speed
and its height at entry and exit.

The milestone is then **not a sample at all**. `airborne` is the instant the
height above the field crosses `runway_airborne_height_ft` = **15 ft** going
*up*; `touchdown` is the same crossing going *down*. A Schmitt trigger with a
5 ft dead band decides *whether* the crossing happened -- so a rotation that
settles briefly does not emit twice -- and the instant is obtained by **linear
interpolation between the two samples that straddle 15 ft**, which is a
different pair of rows from the pair that confirmed it. That is the same
threshold-crossing engine [Flight-level crossings] describes.

Its character was expected to be the opposite of Method A's: an interpolated
crossing is not displaced by construction, so it should carry no method bias.
Measured against a second-resolution reference it lands at
**`r sec(perd_val("EDDS","airborne","ATOT","bias_s"),1)` (EDDS),
`r sec(perd_val("LSZH","airborne","ATOT","bias_s"),1)` (LSZH),
`r sec(perd_val("LEIB","airborne","ATOT","bias_s"),1)` (LEIB)** -- small, but
*not* smaller than Method A's on the same aerodromes. The theoretical advantage is real and the measured one is
not there to be found. But it
can only be read where the aircraft was received *near the runway*, at 15 ft
rather than at 1 500 ft, and only for a flight whose aerodrome the flight list
resolved -- so its coverage depends on surface reception and on the runway grid
in a way Method A's does not.

## The differences, in one place

| | `ATOT` / `ALDT` (Method A) | `airborne` / `touchdown` (Method B) |
|---|---|---|
| Detector | `detect_runway_movements` | `runway_ops` traversals |
| Associates by | distance + height window, runway by median track | H3 runway-grid occupancy + along/cross-track geometry |
| Reads the aircraft at | the climb / descent, up to 1 500 ft AGL | the 15 ft AGL crossing, on and just off the runway |
| The time is | the **extreme sample** of the window | the **interpolated** 15 ft crossing |
| Method bias, second-resolution aerodromes | `r sec(perd_val("EDDS","ATOT","ATOT","bias_s"),1)` / `r sec(perd_val("LSZH","ATOT","ATOT","bias_s"),1)` / `r sec(perd_val("LEIB","ATOT","ATOT","bias_s"),1)` | `r sec(perd_val("EDDS","airborne","ATOT","bias_s"),1)` / `r sec(perd_val("LSZH","airborne","ATOT","bias_s"),1)` / `r sec(perd_val("LEIB","airborne","ATOT","bias_s"),1)` |
| Depends on | flight-list aerodrome only | flight-list aerodrome **and** surface/low reception **and** the runway grid |
| Height datum | height above field elevation | height above field elevation, gated at 15 ft |

**The trade the two methods make is coverage against accuracy, and it is not
symmetric.** Method A buys coverage by reading the well-received climb, and
pays for it with a systematic bias no larger sample averages away. Method B
buys accuracy with an interpolated crossing, and pays for it by needing the
aircraft near the ground -- which is where ADS-B reception is thinnest. The
whole question this study asks of these four milestones is whether Method B's
accuracy can be had without surrendering Method A's coverage; the
[Airborne] and [Touchdown] chapters report the answer against APDF.

**One caveat governs how that answer can even be read.** The bias above is
measured against APDF's `ATOT`/`ALDT`, and at seventeen of the twenty
aerodromes APDF reports those times only to the **whole minute** (see
[Ground truth, and its ceiling]). A whole-minute truth cannot resolve a
few-second method difference -- it injects tens of seconds of apparent bias
that is the reference's rounding, not the detector's error. Method B's ~±2 s is
therefore visible only at the three aerodromes whose APDF times are to the
second (`LSZH`, `LEIB`, `EDDS`); the per-aerodrome tables carry the resolution
flag for exactly this reason, and a bias read without it is a bias read wrong.

# Line-up

**What it is.** A-CDM **T06**: the moment the aircraft enters the runway to
depart. Published as `line-up`.

**How it is detected.** The question this family answers first is not "when did
the aircraft line up" but **"what was this aircraft doing on this runway"** —
and until that is answered, none of the nine milestones can be placed.

## The runway traversal

Nine milestones come out of one object, described here once and referred to by
every runway chapter that follows.

A **traversal** is a maximal run of consecutive state vectors whose H3
resolution-12 cell falls inside a runway polygon of an aerodrome the flight
list names for that flight. A gap longer than `airport_trace_gap_seconds`
starts a new one, so an aircraft that sits on the runway across a coverage hole
becomes two traversals rather than one implausible ten-minute occupancy.

Each traversal is then classified by three measurements over its own samples:

- **alignment** — the angle between the traversal's median ground track and the
  runway's bearing, folded to [0°, 90°] so a reciprocal landing counts as
  aligned;
- **speed** — the maximum groundspeed reached;
- **height** — the height at entry and at exit above the field elevation of
  **the traversal's own aerodrome**.

giving one of three classes:

| Class | Conditions |
|---|---|
| `departure` | aligned within 30°, faster than 50 kt, enters below 15 ft AGL and leaves at or above it |
| `arrival` | aligned within 30°, faster than 50 kt, enters at or above 15 ft AGL and leaves below it |
| `crossing` | at least 45° off the bearing, never above 40 kt, over in 120 s or less, below 15 ft AGL at both ends |

or **NULL, which means abstain**. The gap between 30° and 45° is deliberate:
naming the wrong milestone corrupts a movement count, while naming none is an
explicit null a consumer can see. A traversal that falls in the band produces
no events at all.

**Which field the heights are measured against.** Every height in this family
is measured against the elevation of the aerodrome whose runway the traversal
is on. That sounds like a detail and is not. A flight has two aerodromes and
the pipeline attaches both elevations, so there is a cheaper reading available:
take the *smaller* of the two field-relative heights. That reading is correct
where it is used — deciding whether a sample is close enough to the ground to
be matched against an airport layout, or to be called GND by the phase
classifier — because a track is on the ground at only one of its two
aerodromes, cruise sits far above both, and the wrong end can only ever make
the height larger. Permissive is the safe direction for a membership test.

It is the wrong direction for a *threshold*. Ibiza's field is 24 ft and
Zurich's is 1,416 ft; measured against the higher of the two, a departure from
Ibiza has a height around −1,400 ft for its entire roll and climb-out, never
reaches 15 ft, and is classified as neither a departure nor an arrival nor a
crossing. The detector abstains, and `line-up`, `take-off-roll` and `airborne`
are absent for the whole flight with nothing in the output to say why.
Symmetrically an arrival at the lower of a flight's two aerodromes loses
`landing`, `touchdown` and `runway-vacated`, and a go-around there is
suppressed as a landing. Where the permissive height *does* clear the
threshold, it clears it late by the elevation difference — a systematic bias
in one direction, at exactly the aerodrome pairs where it is largest. So the
runway family measures against the traversal's own aerodrome, the go-around
against the destination's (it is arrival-anchored by definition), and only the
layout gate and the phase classifier keep the permissive reading, which is what
they want.

**Why one object rather than nine detectors.** Line-up, take-off roll and
airborne are not independent questions; they are three readings of one
departure. Deriving them from a shared, explicitly classified traversal means
they cannot disagree about which runway, which aerodrome, or even whether the
movement was a departure. The classification travels with every event in
`info` — `traversal_class`, `align_deg`, `max_gs_kt`, `rwy_ident`, `apt_icao`
— so the decision is auditable from the published row rather than
reconstructible only from the code.

**Which runway.** A traversal is matched against both directions of every
runway at the aerodrome and one is chosen: nearest **cross-track distance** to
the extended centreline first, then smallest *unfolded* bearing error, then the
designator. The cross-track term is what separates parallel runways; the
bearing term is what separates the two ends of one strip, because both ends
share a centreline and both score identically on alignment. Cross-track
distances are rounded to 0.001 NM — about two metres — before the comparison,
so that the two ends of one strip tie exactly rather than being separated by
floating-point noise, while parallel runways hundreds of metres apart stay
separable.

## Back to line-up

`line-up` is the **entry sample** of a `departure` traversal: the first state
vector inside the runway polygon. It is a sampled instant, not an interpolated
one, and it is honest about that — the geometric event it names is the aircraft
crossing the runway edge, which is a polygon boundary the H3 rasterisation only
resolves to about 25 m.

Attributes are taken at the entry *time* through an explicit ordering, never
with `first`/`last` inside a `groupBy`, which take partition order.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `airport trace gap seconds` | 300 s | published | when a traversal is considered ended |
| `runway align max deg` | 30° | ours | alignment tolerance for departure/arrival |
| `runway cross min deg` | 45° | ours | minimum angle for a crossing; the gap to 30° is the abstention band |
| `runway roll speed kt` | 50 kt | ours | rolling rather than taxiing |
| `runway taxi speed kt` | 40 kt | ours | ceiling for a taxiing crossing |
| `runway crossing max seconds` | 120 s | ours | longest traversal still readable as a crossing |
| `runway airborne height ft` | 15 ft | ours | the height that separates on-deck from airborne |
| H3 resolution | 12 | `H3Config` | matching granularity |

```{r tbl-lineup-inv}
#| echo: false
#| results: asis
if (!is.na(n_ev("line-up"))) {
  cat(sprintf("`line-up` events on the sample: **%s**.\n\n", cnt("line-up")))
} else {
  pending("The line-up event count")
}
```

**Compared with reference data. None is possible.** APDF records a runway
designator (`AP_C_RWY`) but **no entry or exit time for any surface feature**,
so there is nothing to score a line-up instant against. What can be checked is
the runway *identity*, and that is done through the milestones APDF does time —
see [Touchdown].

**Limitations.** Requires the aerodrome's OSM layout to exist, which is built
for large and medium aerodromes only, and requires surface reception at the
runway edge, which is the sparsest reception in the whole trajectory. A
traversal in the 30°–45° abstention band produces no line-up. The instant is a
sampled one, so it carries up to one sample interval of error by construction.

# Take-off roll

**What it is.** A-CDM **T07**: the start of the take-off run. Published as
`take-off-roll`.

**How it is detected.** The question is *when did the aircraft stop taxiing and
start going*, and the difficulty is that a single fast sample is not an answer:
a high-speed turn-off, or one noisy groundspeed value, produces one.

So the detector looks for a **run** of samples inside the departure traversal
whose groundspeed reaches `runway_roll_speed_kt`, and confirms the run by
requiring it to be held for at least `runway_roll_min_seconds`. **The
confirmation and the timestamp are deliberately different rows**: the hold is
what makes the run credible, but the roll started at the run's *first* fast
sample, not at the sample that finished proving it. Stamping the confirming
sample would put every T07 one hold-length late, always in the same direction —
which is the exact defect this version indicts `ATOT` for, and a module cannot
claim that improvement and then reintroduce it.

The hold is wall-clock, so it needs a continuity bound: a traversal may legally
span a gap of up to 300 s, and without a bound two fast samples either side of
a four-minute reception hole would satisfy a 5 s hold and stamp a roll nobody
observed. A gap longer than `ROLL_MAX_GAP_SECONDS` therefore breaks the run.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `runway roll speed kt` | 50 kt | ours | rolling rather than taxiing; between taxi (~20 kt) and rotation (~140 kt) |
| `runway roll min seconds` | 5 s | ours | how long the speed must be held before the run is credible |
| `ROLL MAX GAP SECONDS` | 30 s | ours (module constant) | longest gap a roll may span — six times the nominal 5 s cadence |

`ROLL_MAX_GAP_SECONDS` is a module constant rather than an `EventConfig` field,
which is an inconsistency with the rest of this document and is recorded as one
in [Limitations]. It is a tuning parameter and belongs in the config surface.

```{r tbl-roll-inv}
#| echo: false
#| results: asis
if (!is.na(n_ev("take-off-roll"))) {
  cat(sprintf("`take-off-roll` events on the sample: **%s**.\n\n", cnt("take-off-roll")))
} else {
  pending("The take-off-roll event count")
}
```

**Compared with reference data. None is possible.** APDF has no start-of-roll
time. The nearest quantity it holds is ATOT, which is the *end* of the roll and
is scored in [Airborne].

**Limitations.** Depends on groundspeed being present and plausible on the
runway, which is where surface reception is worst. A departure that is first
received already above 50 kt has its roll stamped at the first sample received,
not at the real start, and nothing in the output distinguishes the two. The
5 s hold is a floor, not a model of acceleration.

# Airborne

**What it is.** A-CDM **T08**: wheels off the ground. Published as `airborne`.
This is the milestone `ATOT` was trying to be. It is published *alongside* `ATOT`, not in place of it — see [`ATOT`/`ALDT` are retained, not retired].

**How it is detected.** The question is *when did the height above the field
pass fifteen feet*, and the answer must not be "at the first sample that was
above it" — that is the whole defect being fixed here.

The height above field elevation is run through the same threshold-crossing
engine as the flight-level and ring crossings, described in
[Flight-level crossings]: a Schmitt trigger with a five-foot dead band decides
*whether* the crossing happened, and the instant is then obtained by **linear
interpolation between the two samples that straddle 15 ft**. That is a
different pair of rows from the pair that confirmed it.

The contrast with `ATOT` was meant to be the point of the whole family. `ATOT`
is the *earliest surviving sample* of a detection window bounded by distance,
height, speed and vertical rate, and an extreme of a window is displaced by
construction, which V3 measured at +19 s median on departures.

**The V4 measurement does not support that contrast.** Against a
second-resolution reference `ATOT` is `r sec(perd_val("EDDS","ATOT","ATOT","bias_s"),1)`
at EDDS and `r sec(perd_val("LSZH","ATOT","ATOT","bias_s"),1)` at LSZH, against
`airborne`'s `r sec(perd_val("EDDS","airborne","ATOT","bias_s"),1)` and
`r sec(perd_val("LSZH","airborne","ATOT","bias_s"),1)`; only at LEIB is
`airborne` the better of the two
(`r sec(perd_val("LEIB","airborne","ATOT","bias_s"),1)` against
`r sec(perd_val("LEIB","ATOT","ATOT","bias_s"),1)`). The large pooled biases on both sides are the
reference's own minute-rounding. What `airborne` does win, and by a wide margin
where it wins at all, is *coverage in the places `ATOT` fails* -- see
[Per-detector accuracy by aerodrome].

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `runway airborne height ft` | 15 ft | ours | the height taken as wheels-off |
| `AIRBORNE HYSTERESIS FT` | 5 ft | ours (module constant) | dead band, so a rotation that settles briefly does not emit a second `airborne` |
| plus the traversal parameters | — | — | see [Line-up] |

Fifteen feet rather than zero because barometric altitude at the field is noisy
at the foot level and a zero crossing would chatter. It is a stated proxy for
wheels-off, not a claim to have measured the wheels.

**Compared with reference data.** Scored against APDF's ATOT.

```{r tbl-airborne}
#| echo: false
#| results: asis
score_table(c(ATOT = SHIPPED, airborne = SHIPPED), what = "The airborne comparison")
```

**Both rows are the same rung, and that is the point.** An earlier draft read
`ATOT` at `V00_v3_shipped` and `airborne` at `V07_shipped`, because the two were
expected never to coexist -- the flag that emitted one was to gate off the
other. `events_v0.2.0` retains both (see
[`ATOT`/`ALDT` are retained, not retired]), so the comparison no longer has to
cross rungs: same run, same trajectories, same reference, two definitions of the
same instant. That also removes the trap in the cross-rung form, which was to
invite comparison against whatever *other* sample the baseline rung happened to
be measured on.

```{r tbl-airborne-rungs}
#| echo: false
#| results: asis
rung_table("airborne", what = "The airborne rung progression")
```

**Limitations.** Fifteen feet AGL is a proxy, and its residual bias is reported
rather than assumed away. The interpolation is linear between two samples up to
one interval apart, so a rotation inside a 5 s gap is interpolated across a
manoeuvre that is not linear. Requires a departure traversal, so it inherits
every limitation of [Line-up]: OSM layout coverage, surface reception, and the
abstention band.

# Landing

**What it is.** A-CDM **T16**: the aircraft crosses the landing threshold. This
is the milestone the type string `landing` now carries, and **it is not what
`landing` meant under `events_v0.0.2` or `events_v0.1.0`** — see
[Breaking changes for consumers], which leads with exactly this.

**How it is detected.** The question is *when did the aircraft pass over the
threshold*, which is a geometric question about position, not about height or
speed.

Along-track distance from the threshold is computed as

$$d = \text{haversine}(\text{thr}, P)\cdot\cos\bigl(\text{bearing}(\text{thr}, P) - \text{rwy bearing}\bigr)$$

which is negative short of the threshold and positive beyond it. The crossing
is therefore a **sign change**, and it interpolates linearly in exactly the way
the height crossings do — through the same engine, with a 0.05 NM dead band on
confirmation and the instant taken from the sample pair that straddles zero.

One gate is applied: only crossings **above** `runway_airborne_height_ft` are
kept. An aircraft rolling past the threshold on the ground — backtracking, or
vacating via a turn-off beyond it — has already landed, and stamping a second
`landing` there would double the arrival count.

**T16 needs a sample the traversal does not contain.** A sign change needs a
point on each side of it, and the negative side is *short of the threshold* —
which is outside the runway polygon, because the threshold is the polygon's
near edge. So the samples belonging to an arrival traversal, which are by
definition the samples inside the polygon, contain no negative along-track
distance at all and the crossing can never be confirmed from them. For arrival
traversals only, the sample window is therefore extended **backwards by
`ARRIVAL_LEAD_SECONDS` (60 s)** before the traversal's entry, admitting the
final-approach samples the crossing interpolates from. Nothing else reads the
extended window: the entry and exit instants, the reported positions and every
other milestone in this family stay bounded by the polygon, because those are
statements about occupancy and the lead is not.

Sixty seconds is about 2.3 NM at approach speed — comfortably more than any
displaced threshold or the gap between the polygon edge and the last approach
sample, and far short of a previous traversal the same aircraft could have made
at the same aerodrome. It is a property of the geometry rather than a tuning
choice, which is why it is a module constant; the caveat in [Limitations] about
module constants applies to it too.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `runway airborne height ft` | 15 ft | ours | floor below which a threshold crossing is ground movement, not a landing |
<!-- literal-ok: both are the 0.05 NM module constant restated in other units, not measurements -->
| `THRESHOLD PLANE HYSTERESIS NM` | 0.05 NM | ours (module constant) | dead band on confirmation; ~304 ft, about 1.3 s at approach speed |
| `ARRIVAL LEAD SECONDS` | 60 s | ours (module constant) | how far before the traversal's entry the arrival sample window reaches, so the pre-threshold sample exists |
| plus the traversal parameters | — | — | see [Line-up] |

The dead band's width does not decide the reported instant, which is worth
stating because it is easy to assume otherwise: the band gates *confirmation*,
and the instant is interpolated to the bare plane, earlier than the confirming
sample and unaffected by how wide the band is.

```{r tbl-landing-inv}
#| echo: false
#| results: asis
if (!is.na(n_ev("landing"))) {
  cat(sprintf("`landing` (T16) events on the sample: **%s**.\n\n", cnt("landing")))
} else {
  pending("The landing event count")
}
```

**Compared with reference data. None is possible for the instant.** APDF's
`MVT_TIME_UTC` for an arrival is ALDT — the touchdown — not the threshold
crossing, and the two differ by the flare and the distance from threshold to
touchdown point. Scoring T16 against ALDT would measure that offset and call it
an error, so the shipped configuration deliberately **does not** map `landing`
onto ALDT; [Touchdown] carries that comparison instead. The benchmark enforces
this by construction rather than trusting it: the type-to-milestone map is
narrowed per rung, so `landing` is simply absent from it on any rung with the
A-CDM family switched on, and a plan check refuses a map that widens.

**Limitations.** The threshold position comes from OurAirports, so a runway
whose thresholds are recorded imprecisely displaces every T16 on it by the same
amount — a per-runway bias with nothing in the output to show it. The
along-track projection assumes the aircraft is approaching roughly along the
centreline, which is true of a stabilised approach and less true of a circling
one. And it inherits the runway-direction pick from [Line-up], whose known
failure mode is described in [Cross-cutting findings]: on a runway near bearing
360 the median track can resolve to the reciprocal, and the landing is then
measured from the wrong end of the strip. Finally, an approach whose last
reception before the threshold is more than 60 s old falls outside
`ARRIVAL_LEAD_SECONDS` and yields no T16 — the aircraft still gets its
`touchdown` and `runway-vacated`, so this shows as a `landing` count below the
`touchdown` count rather than as a lost arrival.

# Touchdown

**What it is.** A-CDM **T17**: the wheels meet the runway. Published as
`touchdown`, published *alongside* `ALDT` rather than in place of it (see [`ATOT`/`ALDT` are retained, not retired]).

**How it is detected.** The mirror of [Airborne]: the same height-above-field
signal through the same crossing engine, in the **downward** direction. The
instant is the interpolated crossing of `runway_airborne_height_ft`, taken from
the pair of samples that straddle it.

`ALDT` was the *latest surviving sample* of the arrival detection window — the
same window-extreme construction as `ATOT`, with the vertical-rate gate
inverted.

**Parameters.** Identical to [Airborne]; the traversal's class decides the
direction of the crossing.

**Compared with reference data.** Scored against APDF's ALDT. As in
[Airborne], the table is cross-rung by necessity: `ALDT` is read at
`V00_v3_shipped` and `touchdown` at `V07_shipped`, because the flag that emits
one gates off the other.

```{r tbl-touchdown}
#| echo: false
#| results: asis
score_table(c(ALDT = SHIPPED, touchdown = SHIPPED), what = "The touchdown comparison")
```

```{r tbl-touchdown-rungs}
#| echo: false
#| results: asis
rung_table("touchdown", what = "The touchdown rung progression")
```

## Runway identity

This is where the runway the traversal named is checked, against `AP_C_RWY`.
V3 measured a third of named landing runways wrong, and traced it to the
tie-break: parallel runways were separated by **the distance from each
threshold to the aerodrome reference point**, which is a constant of the
airport, identical for every flight, and therefore picked the same runway of a
parallel pair every time. V4 separates them by the aircraft's own cross-track
distance to each extended centreline.

```{r tbl-runway}
#| echo: false
#| results: asis
runway_table()
```

The fix is **not behind a configuration flag**, which is why the V4 ladder's
`V01_runway_tiebreak` rung carries no configuration change and shows no delta:
a bug fix behind a flag is a bug you have promised to keep. Its effect is read
off the runway-identity table against V3's published equivalent, not off a
rung, and the ladder exempts that rung by name so a *different* accidental
no-op still fails the plan check.

Designators are compared case-insensitively with whitespace stripped and are
not otherwise fuzzed, because `07R` and `07L` are different runways and a
lenient match would hide exactly the error worth finding.

**Limitations.** Fifteen feet AGL is a proxy for the wheels, as in [Airborne].
Requires an arrival traversal. And the identity comparison is only as good as
`AP_C_RWY`, which is itself a reported field.

# Runway vacated

**What it is.** A-CDM **T19**: the aircraft leaves the runway. Published as
`runway-vacated`.

**How it is detected.** The **exit sample** of an `arrival` traversal — the
last state vector whose H3 cell is inside the runway polygon. Sampled, not
interpolated, for the same reason as [Line-up]: the geometric event is a
polygon boundary crossing, and the polygon is rasterised.

Attributes are taken at the exit *time*, through an explicit ordering.

**Parameters.** The traversal parameters, see [Line-up]. No parameter is
specific to this milestone.

```{r tbl-vacated-inv}
#| echo: false
#| results: asis
if (!is.na(n_ev("runway-vacated"))) {
  cat(sprintf("`runway-vacated` events on the sample: **%s**.\n\n", cnt("runway-vacated")))
} else {
  pending("The runway-vacated event count")
}
```

**Compared with reference data. None is possible.** APDF has no runway-vacated
time; the next arrival milestone it records is AIBT, on the stand.

**Limitations.** Runway occupancy time — T17 to T19 — is the derived quantity
most consumers will want from this pair, and it is bounded below by the
rasterisation: a turn-off inside a single 5 s sample is not resolved. A
traversal that ends because reception ended, rather than because the aircraft
vacated, produces a `runway-vacated` at the wrong instant with nothing marking
it.

# Go-around

**What it is.** An approach abandoned and flown away from. It has **no A-CDM
milestone number**, and is absent from the numbering rather than assigned an
invented one.

**How it is detected.** The question is *did the aircraft descend towards this
runway, not land, and climb away* — and the important design decision is that
**it does not use a runway traversal at all**. An approach abandoned at 400 ft
never overflies the strip, so a detector hung off a runway polygon would miss
precisely the go-arounds that matter.

Instead the height above field elevation is run through the crossing engine at
two heights: a **trigger** at 500 ft, which the aircraft must descend below,
and a **recovery** at 1,500 ft, which it must regain. Each recovery is paired
with the most recent descent below the trigger, and only the first recovery of
an excursion is kept, so one excursion is one event. The samples inside that
window are then gated three ways:

- the lowest point of the excursion stayed **above** `runway_airborne_height_ft`
  — if it did not, the aircraft touched down, however briefly, and the climb
  away is a departure, not an abandoned approach;
- the maximum rate of climb reached `goaround_min_roc_ftmin` — a climb away,
  not a descent that merely stopped descending;
- the low point was within `goaround_radius_nm` of the arrival aerodrome — the
  excursion happened on final, not somewhere else entirely.

The event is stamped at the low point, and its `info` carries `low_height_ft`,
`recovery_roc_ft_min` and `excursion_seconds` so a consumer can see the shape
of the excursion behind the label.

Defining "no touchdown in between" as *the excursion never reached 15 ft AGL*
rather than as *no `touchdown` event falls inside the window* is deliberate:
15 ft is the same threshold that **defines** [Touchdown], so the two cannot
disagree, and the go-around detector keeps its independence from the traversal
machinery.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `goaround trigger height ft` | 500 ft | ours | how low the approach must get to count as one |
| `goaround recovery height ft` | 1,500 ft | ours | how high the climb-away must reach |
| `goaround min roc ftmin` | 500 ft/min | ours | separates a climb away from a stopped descent |
| `goaround radius nm` | 10 NM | ours | how close to the aerodrome the excursion must happen |
| `GOAROUND HYSTERESIS FT` | 50 ft | ours (module constant) | dead band on both heights, so a level-off at 500 ft does not toggle repeatedly |
| `runway airborne height ft` | 15 ft | ours | the touchdown suppression |

```{r tbl-goaround-inv}
#| echo: false
#| results: asis
if (!is.na(n_ev("go-around"))) {
  cat(sprintf("`go-around` events on the sample: **%s**.\n\n", cnt("go-around")))
} else {
  pending("The go-around event count")
}
```

**Compared with reference data. None is possible.** APDF records one movement
per flight; a go-around leaves no row of its own, and there is no
EUROCONTROL-side count of abandoned approaches for OPDI to score against. The
only external check available is a plausibility one — the rate, against the
published rates of roughly one to three per thousand approaches — and a rate
that lands in that band is consistent with the detector being right, not
evidence that it is.

**Limitations.** Unvalidated. The four thresholds are ours, chosen to describe
the manoeuvre rather than fitted to anything. A missed approach flown as a
published procedure that levels at 3,000 ft is caught; a low-energy go-around
that never descends below 500 ft is not. The detector needs the arrival
aerodrome from the flight list for its radius gate, so a flight with no ADES
produces none. And the low point is chosen by a minimum that can tie between
two samples at the same height, which makes the timestamp non-deterministic by
one sample interval on a flat minimum.

# Runway crossings

**What it is.** An aircraft taxiing across a runway rather than using it.
Published as `runway-crossing-entry` and `runway-crossing-vacated`. **No A-CDM
number**, and deliberately so: a crossing is not a movement.

**How it is detected.** The `crossing` class of the traversal classification in
[Line-up] — at least 45° off the runway bearing, never above 40 kt, over within
120 s, and below 15 ft AGL at both ends. The two events are the traversal's
entry and exit samples.

**They exist to be separable from movements**, which is what
`entry-runway`/`exit-runway` were not. The published pair fired for a departure,
an arrival and a taxi crossing alike, so any consumer counting runway
occupancies was counting three different things under one name. Splitting them
means a crossing gets its two instants and nothing else, and can never be added
to a movement count by accident.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `runway cross min deg` | 45° | ours | minimum angle off the bearing |
| `runway taxi speed kt` | 40 kt | ours | speed ceiling |
| `runway crossing max seconds` | 120 s | ours | duration ceiling; a hold on the runway is not a crossing |
| `runway airborne height ft` | 15 ft | ours | both ends must be on the deck |

```{r tbl-crossing-inv}
#| echo: false
#| results: asis
if (!is.na(n_ev("runway-crossing-entry"))) {
  cat(sprintf("`runway-crossing-entry` **%s**, `runway-crossing-vacated` **%s**.\n\n",
              cnt("runway-crossing-entry"), cnt("runway-crossing-vacated")))
} else {
  pending("The runway-crossing event counts")
}
```

**Compared with reference data. None is possible.** No source available to OPDI
records taxi crossings of a runway.

**Limitations.** The runway *direction* named on a crossing is weakly
determined and a consumer should not read it. Cross-track distance identifies
the strip correctly, but for a perpendicular traversal the bearing error is
near 90° for both ends, so the choice between them falls through to the
designator — i.e. to the alphabet. It is right about the strip and arbitrary
about the direction, which is defensible for a manoeuvre that has no direction,
but counting crossings per runway *direction* would be reading noise.

# Off-block and on-block

**What it is.** A-CDM **T04** (off-block) and **T21** (on-block) — the inputs
to ICAO's KPI02 and KPI13. Published as `off-block` and `on-block`.

**These are the events V3 published as `AOBT` and `AIBT`.** One detector, two
names: the A-CDM vocabulary flag renames them, and nothing about the detection
changed. The rename is listed in [Breaking changes for consumers].

**How it is detected.** The question is *when did the aircraft start moving*,
and the trap is that a groundspeed field jitters: at 5 s sampling, a stationary
aircraft produces occasional non-zero speeds.

So movement is defined as **sustained**: groundspeed above
`ground_speed_threshold_kt` held for at least `ground_move_min_seconds`. The
start of the first such run is the off-block candidate and the end of the last
is the on-block candidate. This is `traffic`'s `StartMoving` signal, ported to
column expressions.

Both are then **anchored on the parking-position events** from
[Airport surface events]: a candidate is only emitted for a track actually seen
leaving or entering a stand. Without that anchor, a track first received
mid-taxi would report its first movement as an off-block — a different event,
and one that reads as an implausibly short taxi rather than as a miss.

`traffic`'s own parking-position path is deliberately not used: it resolves
stands through a live OpenStreetMap Overpass query whose cache does not survive
serialisation to an executor.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `ground speed threshold kt` | 2 kt | `traffic` | what counts as moving |
| `ground move min seconds` | 30 s | `traffic` | how long movement must persist |

**Compared with reference data.** Scored against APDF's AOBT and AIBT.

```{r tbl-blocks}
#| echo: false
#| results: asis
score_table(c(AOBT = SHIPPED, AIBT = SHIPPED, "off-block" = SHIPPED, "on-block" = SHIPPED),
            what = "The block-time comparison")
```

**All four rows are the same rung, and that is the point** — the same
correction made in [Airborne]. `events_v0.2.0` retains `AOBT`/`AIBT` alongside
`off-block`/`on-block`, so nothing forces this comparison across rungs any
more; all four are read at `V07_shipped`, and the pairs agree because they are
the same detector under two names, not because two runs happened to agree.

### Why this family was failing, and what actually fixed it

**This chapter's own plan predicted that coverage here was limited by stand
reception, and that giving every study aerodrome stand geometry would lift it
substantially.** Half of that survives measurement. Coverage did lift, by far
more than the plan expected — but not because of geometry, and not from the
limit the plan named. The cause was a column that step 04 stopped reading.

**Two changes landed between the previous measurement and this one, and they
are not remotely equal in effect.** The first is the geometry rebuild: the
published `hexaero_airport_layouts` table carried stand geometry for only
fifteen aerodromes before this study, against twenty in the study set, and now
carries 1,036. The second is a one-line change to which columns step 04 reads,
and it is responsible for almost all of what follows.

**`airport_admit_on_ground` had been switched on for two runs and had never
once taken effect.** The setting exists because ADS-B surface position messages
carry no altitude at all, so an aircraft on a stand cannot satisfy an altitude
gate no matter how well it is received; the setting admits those reports on the
strength of their `on_ground` flag instead. But `process_month` selected an
explicit column list out of the track table, and `on_ground` was not in it. The
gate that consumes it reads

```python
if config.airport_admit_on_ground and "on_ground" in sv_f.columns:
    keep_altitude = keep_altitude | col("on_ground")
```

and a feature gate written that way does not fail when the column is missing —
it silently falls back to the altitude-only condition. The setting was on, the
gate was correct, the column had been projected away upstream, and the
measurement that resulted was read as evidence about radio reception. Nothing
in the output could have distinguished the two: a stand sighting that was never
admitted and a stand sighting that was never received are the same absent row.

**What it was worth.** Pooled AOBT coverage moves from
`r pct(bef_val("POOLED","AOBT"))` to `r pct(sc("off-block","coverage_pct"))`,
and AIBT from `r pct(bef_val("POOLED","AIBT"))` to
`r pct(sc("on-block","coverage_pct"))`. Before the fix, exactly two aerodromes
in the study had usable block coverage —
LSZH at `r pct(bef_val("LSZH","AOBT"))` and EDDS at `r pct(bef_val("EDDS","AOBT"))`
— while the other eighteen sat below five per cent and two of them, ENVA and
EGNT, recorded not one off-block or on-block event between them. After it,
every one of the twenty is reportable and the range runs from
`r pct(min(pera$coverage_pct[pera$milestone %in% c("AOBT","AIBT")], na.rm = TRUE))`
to `r pct(max(pera$coverage_pct[pera$milestone %in% c("AOBT","AIBT")], na.rm = TRUE))`.

```{r tbl-blocks-geometry}
#| echo: false
#| results: asis
# Both sides come from files. The "before" side is blocks_before_2026.csv,
# extracted from this repo's history by tools/stage_blocks_before.py; the
# "after" side is per_airport_2026.csv, written by this run. Neither is typed
# into the document, which is the point: when these values *were* typed in, the
# prose claimed a pooled before-coverage of 7.30%/6.58% and an EDDS
# before-value of 1.63%, while the committed data for the very run it named
# said 9.10%/7.85% and 42.51%. The literals had drifted a whole run behind the
# text around them and nothing could see it.
if (has(pera, c("gt_airport", "milestone", "coverage_pct")) &&
    has(bef,  c("gt_airport", "milestone", "coverage_pct"))) {
  cat("| Aerodrome | AOBT before | AOBT after | AIBT before | AIBT after |\n")
  cat("|---|---:|---:|---:|---:|\n")
  for (apt in RANKING) {
    cat(sprintf("| %s | %s | %s | %s | %s |\n", apt,
                pct(bef_val(apt, "AOBT")), pct(apt_val(apt, "AOBT", "coverage_pct")),
                pct(bef_val(apt, "AIBT")), pct(apt_val(apt, "AIBT", "coverage_pct"))))
  }
  cat(sprintf("| **Pooled** | %s | %s | %s | %s |\n",
              pct(bef_val("POOLED", "AOBT")), pct(sc("off-block", "coverage_pct")),
              pct(bef_val("POOLED", "AIBT")), pct(sc("on-block", "coverage_pct"))))
  cat("\n")
} else {
  pending("The block-milestone before/after table")
}
```

**The ladder cannot show this, and must not be read as if it could.** The fix
is a correction to which columns step 04 reads; it is not behind an
`EventConfig` flag and there is no configuration of the current tree that turns
it off. Every rung therefore carries it, and V00 — the reconstructed v0.1.0
baseline — reports the same block coverage as V07. The gain measured here is
against *the previous run*, recorded in `data/blocks_before_2026.csv` with the
commit that produced it, not against a rung. A reader who looks for it as a
step between two rungs will find nothing and should not conclude nothing
happened.

**The geometry rebuild's own contribution cannot be separated from this
measurement.** Both changes were in place when this ran, and only their
combined effect was measured. The previous run's per-aerodrome numbers do bound
it: EGNT and ENVA had zero detections before and have
`r pct(apt_val("EGNT","AOBT","coverage_pct"))` and
`r pct(apt_val("ENVA","AOBT","coverage_pct"))` now, and neither could have
produced an event at all without stand geometry — so the geometry work is
necessary at those two. What it was worth at the other eighteen, which already
had some geometry, this study does not say. Attributing a share to it would be
a guess, and the previous version of this chapter made exactly that guess in
the opposite direction.

**Why the reception explanation survived as long as it did.** It fitted. Stand
reception genuinely is the weakest part of the ADS-B picture, the milestones
that failed were exactly the two that depend on it, and the aerodromes where it
did work — LSZH and EDDS — are large, well-mapped and well-received. Every
prediction the theory made about *which* aerodromes would fail came true. What
it never explained is why coverage sat near one per cent rather than merely
low, and why it did so almost uniformly across aerodromes whose reception
quality differs a great deal. A hypothesis that survives because its
predictions are confirmed, while a quantity it cannot account for sits in plain
view, is the failure mode this chapter is now an example of.


### The timing is the real defect, and it is not a reception problem

**Both milestones still carry a one-sided bias, an order of magnitude smaller
than before but the same shape.** Pooled AOBT fires
`r sec(sc("off-block","bias_s"))` (MAD `r absec(sc("off-block","mad_s"))`)
after the reference AOBT, and pooled AIBT fires
`r sec(sc("on-block","bias_s"))` (MAD `r absec(sc("on-block","mad_s"))`)
*before* the reference AIBT — roughly a minute each way, against
`r sec(bef_val("POOLED","AOBT","bias_s"))` and
`r sec(bef_val("POOLED","AIBT","bias_s"))` in the run before the column fix.
The direction is what has not changed: **off-block is late at every one of the
twenty aerodromes, and on-block is early at nineteen of them**, EBBR being the
single exception. A detector whose errors were dominated by noise would not do
that.

`mad_s` tracking `|bias_s|` remains the signature of a systematic offset rather
than scatter, and it now separates the aerodromes cleanly: where the bias is
large the two are nearly equal — EGCC at
`r sec(apt_val("EGCC","AOBT","bias_s"))` with MAD
`r absec(apt_val("EGCC","AOBT","mad_s"))`, UGKO at
`r sec(apt_val("UGKO","AIBT","bias_s"))` with MAD
`r absec(apt_val("UGKO","AIBT","mad_s"))` — and where it is small, both are.
Within-60-second accuracy is correspondingly spread rather than uniformly
poor: pooled `r pct(sc("off-block","within_60s_pct"))` for off-block and
`r pct(sc("on-block","within_60s_pct"))` for on-block, but
`r pct(apt_val("EBBR","AIBT","within_60s_pct"))` at EBBR and
`r pct(apt_val("EICK","AOBT","within_60s_pct"))` at EICK against
`r pct(apt_val("LPFR","AIBT","within_60s_pct"))` at LPFR. An earlier version of
<!-- literal-ok: a quotation of this document's own superseded text, deliberately fixed -->
this chapter reported that figure as "at or near 0.00% almost everywhere",
which was true of the measurement it was written against and is not true of
this one.

**The same run, the same aerodromes, still contradicts a reception
explanation for what remains.** At LSZH, `ATOT` lands at
`r sec(apt_val("LSZH","ATOT","bias_s"))` with
`r pct(apt_val("LSZH","ATOT","within_10s_pct"))` of departures within ten
seconds, and `ALDT` at `r sec(apt_val("LSZH","ALDT","bias_s"))`. Both read
ADS-B surface and near-surface data at the same aerodrome, the same period,
the same reception conditions as the block-time detector. If reception at the
stand were setting the residual, the runway milestones beside it would carry a
comparable offset. They do not.

**A hypothesis, stated as one, not as a finding: the anchor is probably the
wrong cell, not a missing one.** `calculate_block_events` anchors off-block on
the *last* parking-position cell a track touches before it starts moving, and
on-block on the *first* one it touches after. At an aerodrome with stands
lining the taxiways, an aircraft pushing back and taxiing out keeps crossing
other aircrafts' stand polygons after it has left its own — so the *last* cell
it touches is already into the taxi, which reads as a late off-block. Taxiing
in, the *first* stand-shaped cell it crosses is upstream of its own gate, which
reads as an early on-block. That predicts precisely the asymmetry the data
shows — late departure, early arrival — with a magnitude set by taxi distance
rather than by reception, and the ordering across aerodromes is consistent with
it: the extremes sit at large multi-pier airports
(`r sec(apt_val("EGCC","AOBT","bias_s"))` at EGCC) and the tightest values at
compact ones (`r sec(apt_val("EICK","AOBT","bias_s"))` at EICK,
`r sec(apt_val("EBBR","AOBT","bias_s"))` at EBBR). It is consistent, not
established: this study measures no stand dwell and cannot separate that
account from any other that scales with taxi length. The fix it implies would
be to anchor on the stand where the aircraft actually **dwelled**, rather than
the first or last cell brushed in passing. Testing it needs its own re-run.

**Limitations.** The account this chapter gave until this measurement — that
the family was reception-bound at the stand — was wrong, and the coverage
section above sets out both the cause and why the wrong account was
persuasive. What remains
after the fix is a timing residual, not a coverage one: a systematic offset of
about a minute in each direction, present at every aerodrome, with
`r pct(sc("off-block","within_300s_pct"))` of off-blocks and
`r pct(sc("on-block","within_300s_pct"))` of on-blocks landing within five
minutes of the reference. Whether that is good enough depends entirely on the
use: it is adequate for counting movements and for stand occupancy at
five-minute resolution, and it is not adequate for KPI02 or KPI13, which are
themselves minute-scale quantities and would be measuring this detector's
anchoring error as much as any operational effect. Pushback (T05) was not built
for the reason V3 gave, and this measurement does not overturn it: T05 needs
strictly more coverage than off-block has, and off-block's own timing is not
yet a foundation to build on.

# Top of climb and top of descent

**What it is.** The end of the climb and the start of the descent. Two
algorithms are published side by side, under four type names:

| Type | Algorithm | `info.method` |
|---|---|---|
| `top-of-climb` / `top-of-descent` | the fuzzy phase classifier | `phase` |
| `top-of-climb-cco` / `top-of-descent-cdo` | EUROCONTROL PRU's vertical-efficiency methodology | `pru` |

**Both are published because neither can be shown better.** No reference data
records a top of climb — APDF holds movement and block times, ring crossings, a
runway and a stand, and nothing about the cruise phase. Publishing one and
suppressing the other would be a preference presented as a finding. Every top
now names the algorithm that produced it, so a consumer can choose.

## How the phase tops are detected

Directly from the phase labels: the earliest and latest sample classified
cruise. The cruise rule is `min(alt_hi, roc_zero, spd_hi)` with OpenAP's
membership functions — altitude Gaussian at 35,000 ft (σ 20,000), vertical rate
Gaussian at 0 ft/min (σ 100), speed Gaussian at **600 kt (σ 100)**.

Those are OpenAP's published values, not chosen here, and they carry a known
consequence: a turboprop cruising at 250 kt scores `exp(−6.125) ≈ 0.002`, so
the cruise rule never wins and **turboprop and general-aviation traffic get no
phase top of climb or descent at all**. It is inherited rather than an OPDI
defect, and left in place deliberately: deviating from a reference constant
should follow a measurement, not precede one.

## How the PRU tops are detected

The question PRU asks is different, and the difference is not a tuning choice.
Its methodology is published at
<https://ansperformance.eu/methodology/cd-vertical-flight-efficiency-pi/>; the
quantities below are transcribed from it and from the PRC technical note.

**The analysis is bounded by a 200 NM radius** around the flight's own
aerodromes. `D200` is the first crossing of that radius outbound, `A200` the
last inbound. A step climb 250 NM out is cruise-altitude optimisation, not a
top of climb, and the radius exists precisely to exclude it.

**ToC-D200** is the first 4D point of highest altitude between take-off and
D200 inclusive; **ToD-A200** is the last 4D point at which the aircraft leaves
the highest altitude between A200 and touchdown. First and last respectively,
not either end of a plateau — a climb is interrupted from below and a descent
from above.

**Then the top moves.** From the ToC-D200 altitude down to **90%** of it is the
*exclusion box*. A level segment inside that box lasting over **5 minutes** is
cruise rather than a level-off: it is dropped from the level-off family, and
the top relocates — **ToC-CCO** to the start of the first such segment,
**ToD-CDO** to the end of the last. With no such segment, ToC-CCO is the
ToC-D200 top, and `info` records `toc_relocated` so the two cases are
distinguishable in the published data.

A track that never reaches the 200 NM radius keeps its whole trajectory in the
analysis window — the fallback is the last sample outbound and the first
inbound. A short sector is a sector PRU still has tops for, and abstaining
would drop every flight under about 400 NM, which is most of them.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `level analysis radius nm` | 200 NM | PRU | the ring bounding the analysis |
| `level exclusion box pct` | 90% | PRU | lower edge of the exclusion box |
| `level exclusion box seconds` | 300 s | PRU | duration above which a segment in the box is cruise |
| `emit pru tops` | on | ours | whether the PRU pair is emitted alongside the phase pair |
| `level anchor` | `pru` | ours | which pair anchors the level-off classification |
| `phase cruise speed kt` | 600 kt | OpenAP | centre of the phase cruise-speed membership |
| `phase cruise speed sigma kt` | 100 kt | OpenAP | its width |

```{r tbl-tops-inv}
#| echo: false
#| results: asis
if (!is.na(n_ev("top-of-climb")) || !is.na(n_ev("top-of-climb-cco"))) {
  cat("| Type | Events |\n|---|---:|\n")
  for (t in c("top-of-climb", "top-of-descent", "top-of-climb-cco", "top-of-descent-cdo"))
    cat(sprintf("| `%s` | %s |\n", t, cnt(t)))
  cat("\n")
} else {
  pending("The top-of-climb and top-of-descent counts")
}
```

The count difference between the two families is itself informative: the PRU
tops do not require a cruise phase to be recognised, so they exist for the
turboprop traffic the phase tops silently omit.

**Compared with reference data. None is possible**, for either algorithm, for
the reason given above. The claim the PRU pair makes is **conformance to a
published methodology**, tested on synthetic trajectories whose geometry is
known by construction: a climb to 30,000 ft with a 360 s hold at 27,500 ft
<!-- literal-ok: a worked example of the PRU exclusion-box rule, not a measurement -->
(91.7% of the top, inside the box, longer than 300 s) must relocate the top; a
240 s hold at the same altitude must not; a 600 s hold at 20,000 ft (67%, below
the box) must not, and must be reported as a level-off instead. Those three
cases are the whole rule and each is a test.

**Limitations.** The phase pair inherits the cruise-speed bias described above.
The PRU pair depends on the flight list naming both aerodromes: without a
position for the aerodrome there is no distance, and a NULL distance fails the
radius test rather than passing it — a segment that cannot be placed relative
to an aerodrome is outside the methodology, not inside it by default. Neither
pair is validated against data, because there is nothing to validate against.

# Level segments

**What it is.** Level flight during climb and descent — the inputs to ICAO's
KPI17 and KPI19 — published as `level-off-climb-start` / `-end` and
`level-off-descent-start` / `-end`.

Separately, and **not interchangeably**, OPDI continues to publish
`level-start` and `level-end` from the fuzzy phase classifier. Those ask
whether a sample *looks like* level flight; the level-off family asks a
geometric question about altitudes. Both are published, under different names,
and a consumer computing a vertical-efficiency indicator wants the latter.

## How it is detected

Two arms are implemented and one is selected by `level_method`. The shipped
default is PRU's.

**The PRU arm (`pru`, shipped).** The question is *did the aircraft stay
level*, answered as a property of the altitudes themselves rather than as a
test on the reported vertical rate — which matters because `vert_rate` is a
broadcast field with its own noise and its own staleness. A trajectory part is
level when it stays inside a rolling window whose dimensions satisfy
**Y/X = 300 ft/min**: at the default 10 s window, a 50 ft box. The trajectory
is linearly interpolated onto that grid, because a discrete track has no sample
at every instant the window requires. A segment is counted from **20 s** long.

The window height is *derived*, not stored: `level_window_height_ft` is
`level_vertical_speed_limit_ftmin × level_window_seconds / 60`. Storing both
dimensions would let them drift, and a window whose ratio is not 300 ft/min is
not the PRU window under a different parameterisation — it is a different
definition of level flight wearing the same name.

**The ICAO arm (`icao`, retained).** The algorithm shipped in `events_v0.1.0`,
kept reachable and unchanged. ICAO specifies it completely:

> a data point starts a level segment when the altitude difference with the
> next data point is ≤ the level band limit and the vertical speed towards it
> is ≤ the vertical speed limit. The segment ends when the altitude differs
> from the segment's starting altitude by more than the level band limit, or
> the vertical speed between two consecutive points exceeds the limit.

The anchoring is what a plain window function cannot express: a segment's end
is defined against *its own start*, a running value that resets. It is
implemented as a sessionisation — mark where flatness begins, forward-fill the
anchor within the group — with each group truncated at its first failure, so a
trajectory that leaves the band and drifts back is not stitched across the
excursion.

**Classification, common to both arms.** Segments are split into KPI17 and
KPI19 by whether they fall before the top of climb or after the top of descent
— the **PRU** tops by default, see [Top of climb and top of descent] — with
three exclusions: the climb floor, the descent floor, and the exclusion box
that also relocates the tops.

## The floors were being compared against the wrong datum

This is the second silent unit defect V4 closes, and it is the same class as
the one in [Airport surface events].

The climb floor is **3,000 ft AGL** (the end of the ICAO noise-abatement
departure procedure) and the descent floor **1,800 ft AGL** (typical ILS
interception). The shipped v0.1.0 code compared both against **uncorrected
pressure altitude**. At Zurich, field elevation 1,416 ft, the descent floor
therefore sat 1,416 ft too low; at Ibiza, field elevation 24 ft, it was
effectively right. The result was a per-aerodrome bias in what counted as a
level-off, with nothing in the output to show it. `level_floors_above_field`
measures both against the field.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `level method` | `pru` | ours | which arm runs |
| `level window seconds` | 10 s | PRU | temporal size of the rolling window |
| `level vertical speed limit ftmin` | 300 ft/min | PRU / ICAO | the Y/X ratio, and the ICAO arm's rate limit |
| `level window height ft` | 50 ft | derived | the window's height at a 10 s window |
| `level band limit ft` | 200 ft | ICAO | the ICAO arm's altitude band |
| `level min duration seconds` | 20 s | PRU / ICAO | shortest reportable segment |
| `level min altitude climb ft` | 3,000 ft AGL | PRU / ICAO | where climb analysis starts |
| `level min altitude descent ft` | 1,800 ft AGL | PRU / ICAO | where descent analysis stops |
| `level floors above field` | on | ours | whether those two are AGL or above the 1013.25 hPa datum |
| `level analysis radius nm` | 200 NM | PRU | radius around the aerodrome analysed |
| `level radius enforced` | on | ours | whether that radius binds at all; off reproduces v0.1.0, which declared it and read it nowhere |
| `level exclusion box pct` | 90% | PRU / ICAO | lower edge of the cruise exclusion |
| `level exclusion box seconds` | 300 s | PRU / ICAO | duration above which a high segment is cruise |

**None of the published values must be tuned.** They are PRU's and ICAO's, and
tuning them forfeits the only claim this family can make — conformance to a
published methodology — in exchange for a number nothing can validate.
`level_method`, `level_floors_above_field`, `level_radius_enforced` and
`level_anchor` are ours and are choices about which published methodology to
follow, not about its constants. `level_radius_enforced` is the odd one: it is
not a choice anybody should make differently, but the radius binds a segment
whose aerodrome the flight list could not name *out* of the analysis, so
turning it on and off is the only way to say whether a change in the count came
from the geometry or from the flight list's coverage — and it is the only way
to reproduce v0.1.0, which declared the radius and never read it.

```{r tbl-levels-inv}
#| echo: false
#| results: asis
if (!is.na(n_ev("level-off-climb-start")) || !is.na(n_ev("level-start"))) {
  cat("| Type | Events |\n|---|---:|\n")
  for (t in c("level-off-climb-start", "level-off-climb-end",
              "level-off-descent-start", "level-off-descent-end",
              "level-start", "level-end"))
    cat(sprintf("| `%s` | %s |\n", t, cnt(t)))
  cat("\n")
} else {
  pending("The level-segment counts")
}
```

```{r tbl-levels-arms}
#| echo: false
#| results: asis
if (has(inv, c("type", "rung", "n_events"))) {
  a <- n_ev("level-off-climb-start", "V02_layout_agl")   # ICAO arm still selected
  b <- n_ev("level-off-climb-start", "V03_pru_level")    # PRU arm selected
  if (!is.na(a) && !is.na(b)) {
    cat(sprintf(paste0("Switching the arm from ICAO's anchored band to PRU's rolling ",
                       "window changes the climb level-off count from **%s** to **%s** ",
                       "on the same trajectories.\n\n"), fm(a), fm(b)))
  } else {
    pending("The level-arm comparison")
  }
} else {
  pending("The level-arm comparison")
}
```

**Compared with reference data. None is possible, and this is the family where
that matters most.** No source available to OPDI holds level-segment truth. The
claim is conformance, tested on synthetic trajectories: a climb at 280 ft/min
is level under the PRU window and one at 320 ft/min is not, because at a 10 s
sample the first rises 46.7 ft inside a 50 ft window and the second rises
53.3 ft. That is checkable; "accurate" is not.

**Limitations.** Unvalidated against data, by necessity. Two arms exist and
neither can be shown better than the other, so the choice of default is an
argument, not a measurement. The PRU arm's interpolation grid is built as a
single array per partition and is capped, so a track whose segmentation failed
and which spans a month loses the tail of its analysis — bounded and in the
cheap direction, but real. And the arms mean slightly different things by
`level_ft`, which a consumer comparing across `level_method` values would need
to know.

# Flight-level crossings

*Types:* `xing-fl50`, `xing-fl70`, `xing-fl100`, `xing-fl245`.

**What it is.** Every crossing of each monitored flight level, with the instant
interpolated to the level itself.

**How it is detected.** The question is *when did the aircraft pass through
this level*, and the reason it needs an algorithm rather than a comparison is
noise.

A trajectory that sits near a threshold does not cross it once. Sampled at five
seconds, an aircraft holding FL99.8 with fifty feet of barometric noise
produces a crossing of FL100 every few samples, all of them spurious. The
remedy is a **Schmitt trigger**: instead of one threshold there are two, a dead
band either side of it. The signal is *above* only once it exceeds the upper
edge and *below* only once it falls under the lower edge; between the two it is
undefined and keeps whatever it last was. A crossing is published when that
remembered side flips, so noise inside the band cannot produce one and a
genuine crossing still does. The band here is `crossing_hysteresis_ft` = 300 ft.

The trigger decides *whether* a crossing happened. It does not decide *when*:
the sample that confirms it is already past the threshold. The instant is
obtained separately, by linear interpolation between the two samples that
straddle the bare threshold — a different pair of rows from the one that
confirmed it.

**This is the engine four other families use.** [Airborne] and [Touchdown] run
it over height above field elevation, [Landing] over along-track distance from
the threshold, [Go-around] over height at two levels, and [Ring crossings] over
distance to an aerodrome. Each supplies its own quantity, its own thresholds
and its own dead band; the mechanism is identical, which is why validating it
in one place validates it in all of them.

Two further properties, both fixes to the published detector. Thresholds are
evaluated as **parallel columns over one window** rather than by exploding the
trajectory once per threshold, so N levels cost one shuffle rather than N — and
that also removes a latent fault, because the published detector's first-match
`when` chain reported only one level when a single 5 s interval spanned two.
And an aircraft cruising *at* a level never leaves the dead band, so its side
never resolves and **no event is emitted**, which is the property that makes
reporting every crossing usable at all.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `crossing levels fl` | 50, 70, 100, 245 | published vocabulary | which levels are monitored |
| `crossing hysteresis ft` | 300 ft | ours | chatter suppression vs missed crossings |
| `crossing all occurrences` | on | ours | every crossing rather than first and last |
| `crossing interpolate` | on | ours | interpolate to the level rather than report a sample |

Each event carries `crossing_seq` (1-based, in time order), `direction`, and
`bracket_s` — the spacing of the two interpolated-between samples, so a
consumer can tell an instant measured across 5 seconds from one inferred across
four minutes.

```{r tbl-xing-inv}
#| echo: false
#| results: asis
if (!is.na(n_ev("xing-fl100"))) {
  cat("| Type | Events |\n|---|---:|\n")
  for (t in c("xing-fl50", "xing-fl70", "xing-fl100", "xing-fl245"))
    cat(sprintf("| `%s` | %s |\n", t, cnt(t)))
  cat("\n")
} else {
  pending("The flight-level crossing counts")
}
```

**Compared with reference data. None is possible** — APDF holds no
flight-level crossings. The detector's *mechanism* is nonetheless validated
where the same code is applied to rings, which APDF does record: see
[Ring crossings].

**Limitations.** A flight that climbs and levels off **exactly at** a threshold
never clears the far edge and registers no crossing there. That case is
described by the level-segment and top-of-climb events instead, but it is a
real semantic change from the published `first-xing-fl*` / `last-xing-fl*`
family, which is retired — see [Breaking changes for consumers].

# Ring crossings

*Types:* `xing-40nm`, `xing-100nm`.

**What it is.** Crossings of the 40 NM and 100 NM cylinders around the flight's
**own** origin and destination — ICAO's ASMA cylinder for KPI08 and the
reference area for KPI05.

**How it is detected.** The Schmitt trigger and interpolation engine of
[Flight-level crossings], with distance to an aerodrome as the quantity and a
1 NM dead band.

The rings come from the flight list's ADEP and ADES rather than from the H3
aerodrome zone table: that is what the indicators mean, and what APDF records —
one crossing per movement, not one per aerodrome overflown. It is also far
cheaper, multiplying each sample by at most two rather than by every aerodrome
within 110 NM.

Bearing is computed **from** the interpolated crossing position rather than
interpolated alongside it. A bearing is circular: averaging 359 and 1 gives
180, wrong by half a turn on any crossing near due north.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `ring radii nm` | 40, 100 NM | ICAO | which cylinders |
| `ring hysteresis nm` | 1 NM | ours | chatter suppression for a track flying tangentially |
| `crossing interpolate` | on | ours | shared with the level crossings |

`info` carries `crossing_seq`, `direction` (inbound/outbound), `apt_icao`,
`bearing` and `flight_level`, matching `C40_CROSS_{TIME,LAT,LON,FL}` and
`C40_BEARING` column for column, so the comparison is direct rather than a
translation.

**Compared with reference data.** This is the one family where the reference is
second-resolution, so it is the one place a seconds-level claim is defensible.
On V3's sample the detector's spread sat **inside the disagreement between two
independent EUROCONTROL derivations of the same crossing** — median error under
half a second, median position error about 0.2 NM. The V4 sample reproduces
that: the table below is the ring comparison on these three days, and the
sub-second bias is not a rounding artefact of the reference, because this is the
one family whose reference resolves to the second.

```{r tbl-rings}
#| echo: false
#| results: asis
# Reads rings_2026.csv, not the ladder: the ladder scores only the four APDF
# movement milestones, so a ring row never appears in it and a score_table()
# call against it renders as unstaged however complete the run was.
if (has(rng, c("milestone", "coverage_pct", "bias_s"))) {
  r <- rng[rng$rung == SHIPPED, ]
  if (!nrow(r)) r <- rng
  cat("| Ring | Coverage | Bias | MAD | Median position error | p90 position error |\n")
  cat("|---|---:|---:|---:|---:|---:|\n")
  for (i in order(r$milestone)) {
    cat(sprintf("| `%s` | %s | %s | %s | %.2f NM | %.2f NM |\n",
                r$milestone[i], pct(r$coverage_pct[i]), sec(r$bias_s[i]),
                absec(r$mad_s[i]), as.numeric(r$median_pos_error_nm[i]),
                as.numeric(r$p90_pos_error_nm[i])))
  }
  cat("\n")
} else {
  pending("The ring comparison")
}
```

**Limitations.** A ring crossing is not detected for an aerodrome merely
overflown — correct for these indicators, but not a general-purpose airspace
crossing. And the family depends entirely on the flight list naming the
aerodrome, so ADEP/ADES detection error propagates directly into it.

# Airport surface events

*Types:* `entry-` / `exit-` for `taxiway`, `apron`, `parking_position`,
`threshold`, `hangar`, `deicing_pad`.

**What it is.** Traversals of the aerodrome's surface features, from
OpenStreetMap aeroway geometry.

**`entry-runway` and `exit-runway` are no longer part of this family.** They
were the one feature where a traversal's *purpose* mattered, and they are
replaced by the classified runway family — [Line-up] through
[Runway crossings]. See [Breaking changes for consumers].

**How it is detected.** Low-altitude samples are matched against OSM aeroway
geometry pre-rasterised to H3 resolution 12 in `hexaero_airport_layouts`. A
sample joins a feature when its `h3_res_12` cell matches *and* the feature's
aerodrome is one the flight list names for that flight. Consecutive matches on
the same feature form a traversal; a gap over 300 s starts a new one. Each
traversal yields an entry and an exit event.

## The gate was measured against the wrong datum

The published gate is **FL20 against the 1013.25 hPa datum**. `baro_altitude_c`
is uncorrected pressure altitude, so on a low-pressure day, or at any aerodrome
high enough, the gate moves relative to the ground it is supposed to sit just
above. It is the same defect class as the phase family's ground ceiling, and
the same one the level floors carry in [Level segments]: a threshold whose
*intent* is "near the ground" compared against a datum that is not the ground.

`airport_gate_above_field` measures the gate as **2,000 ft above field
elevation**, which is FL20's intent expressed against the field. The gate runs
before the column projection narrows the frame, because the height calculation
needs the raw altitude and the two elevation columns, none of which survive the
select.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `airport gate above field` | on | ours | whether the gate is AGL or against the 1013.25 datum |
| `airport max height agl ft` | 2,000 ft | ours | the gate's value when it is AGL |
| `airport max fl` | FL20 | published | the gate's value when it is not |
| `airport trace gap seconds` | 300 s | published | when a traversal is considered ended |
| `airport events ordered` | on | ours | attributes taken at the entry/exit *times* |
| H3 resolution | 12 | `H3Config` | matching granularity |

`airport_events_ordered` is a correctness fix, not a tuning knob: the published
detector used `first`/`last` inside a `groupBy`, which take **partition order,
not time order**, so a reported entry position was not guaranteed to be the
position at the reported entry time.

```{r tbl-surface-inv}
#| echo: false
#| results: asis
feats <- c("taxiway", "apron", "parking_position", "threshold", "hangar", "deicing_pad")
if (any(!is.na(sapply(paste0("entry-", feats), n_ev)))) {
  cat("| Feature | Entries | Exits |\n|---|---:|---:|\n")
  for (f in feats)
    cat(sprintf("| `%s` | %s | %s |\n", f, cnt(paste0("entry-", f)), cnt(paste0("exit-", f))))
  cat("\n")
} else {
  pending("The surface-event counts")
}
```

**Compared with reference data. None is possible for any of them.** APDF
records a runway designator (`AP_C_RWY`) and a stand (`AP_C_STND`), but **no
entry or exit times for any surface feature**. A runway *identity* can be
checked indirectly through [Touchdown]; no aeroway event's *timing* can be
checked at all.

**Limitations.** Cannot fire for an aerodrome the flight list did not name.
Depends on OSM coverage, which is built only for large and medium aerodromes in
the ingestion area. A traversal completed inside a single 5 s sample is dropped,
because its duration is zero. Threshold polygons are the smallest features in
the set and therefore the most sensitive to cleaning removing surface samples —
V3 measured the largest relative fall of any event type there, and it was
trajectory cleaning, not a detector change. `deicing_pad` is in the vocabulary
but was not observed on V3's sample, because de-icing pads are tagged in OSM
for few of the aerodromes covered and the sample is June.

# First seen and last seen

*Types:* `first_seen`, `last_seen`.

**What it is.** The first and last sample of each track.

**How it is detected.** Row number ascending and descending over the track,
ordered by time. Emitted for every track without condition, so the count is
also the track count.

**Parameters.** None.

```{r tbl-seen-inv}
#| echo: false
#| results: asis
if (!is.na(n_ev("first_seen"))) {
  cat(sprintf("`first_seen` **%s**, `last_seen` **%s** — which is also the number of tracks in the sample.\n\n",
              cnt("first_seen"), cnt("last_seen")))
} else {
  pending("The first-seen and last-seen counts")
}
```

**Compared with reference data. None is possible, and not for want of a
source.** These are properties of *reception coverage*, not of the flight.
There is nothing they could be right or wrong against — an aircraft is first
seen when a receiver first hears it, which no reference dataset records because
it is not a property of the operation.

**Limitations.** May carry a null position where the first sample has none.
Their value to a consumer is as a **coverage diagnostic**: the gap between
`first_seen` and [Off-block and on-block], or between `last_seen` and
[Runway vacated], is the part of the movement ADS-B did not see, and in this
study set it is smaller than it is anywhere else in the network.

# Measurements

Two values attached to every event rather than being events themselves.

**`Distance flown (NM)`** is the cumulative great-circle distance from the
start of the track, summed sample to sample with an Earth radius of 6,371 km
and converted at 1.852 km per nautical mile. **`Time Passed (s)`** is elapsed
time from the track's first sample.

Both are **interpolated to the crossing instant** for every family that
produces an interpolated instant — the crossings, the rings, `airborne`,
`touchdown` and `landing`. This is not a refinement: a measurement attached to
an interpolated event but read from a neighbouring row would disagree with its
own event, and the disagreement would be invisible in the published table.

**Compared with reference data.** APDF carries `C40_TRANSIT_DIST_KM` — the
achieved distance inside the ring — which makes a distance comparison possible
for the ring families. **It has not been run**, in V3 or here; it remains the
most tractable unmeasured quantity in this programme.

**Limitations.** Cumulative from the *track's* start, not from the flight's, so
a flight whose track begins mid-climb has a distance measure that is not the
distance flown. Track identity changed shape with the A8 segmentation default,
so these measures are not comparable across that boundary either.

# Cross-cutting findings

Six results are about the programme rather than about any one milestone.

## A published take-off can be shadowed by a level-start

The legacy phase family emits its events from a single `when` chain, and the
`take-off` arm sits **after** the level-segment arms in it. A sample the fuzzy
classifier labels `LVL` therefore publishes a `level-start` and the take-off at
that same sample is never reached.

This is pre-existing behaviour, unchanged by V4, and it is narrow — it needs
the rotation sample to be classified level rather than climb, which happens
when the vertical rate is still small and the smoothing window has not yet
tipped. It surfaced while building a test fixture, not from the scored metrics,
which is the point worth recording: **an ordering sensitivity inside a `when`
chain produces a missing event, and a missing event looks exactly like a
detector that did not fire.** It is one more reason the A-CDM family derives its
milestones from an explicitly classified object rather than from the order of
arms in a chain.

## A circular quantity is being averaged linearly, and it fails silently

The runway direction is chosen using the traversal's **median ground track**,
computed as `percentile_approx(heading, 0.5)`. Heading is circular and the
median is not: a runway near bearing 360 whose samples straddle 359° and 001°
yields a median near 180° — the exact reciprocal.

The alignment fold to [0°, 90°] hides this from the *classification*, because a
reciprocal still counts as aligned, so the traversal is still called an arrival
and `touchdown` still fires. But [Landing] is measured as a sign change of
along-track distance **from the chosen threshold**, and measured from the wrong
end of the strip that sign change never occurs inside the traversal. The
failure mode is therefore **a silently missing `landing` on runways numbered
near 36/18**, with a `touchdown` present on the same traversal to make the gap
look like a detector quirk rather than a defect.

It is inherited from the shipped `detect_runway_movements` rather than
introduced here, and fixing it means a circular median — a deliberate deviation
from the reference implementation, which is why it was not done inside this
change. The check that would settle it is a cut of the runway-identity table by
designator, and it is the first thing to run when the chain completes.

## Two vocabularies wrote the runway designator under two keys

`ATOT`/`ALDT` recorded the runway in `info` under `runway`; the A-CDM family
records it under `rwy_ident`. Both readers in the benchmark parsed only
`runway`, so **every v0.2.0 designator would have read as NULL —
indistinguishable from "the detector named no runway"**, and the runway-identity
result would have collapsed to zero with a plausible-looking explanation
available.

It was found by wiring the comparison, not by a test, and both readers now
coalesce the two keys. It is recorded here because it is the characteristic
failure of a version bump: not a wrong number, but a right number read from a
key that moved.

## Abstention is an output, not a gap

Three places in the V4 detectors decline to answer rather than guessing: the
30°–45° traversal band, the runway pick when no runway is within tolerance, and
a NULL distance failing the 200 NM radius test rather than passing it.

Each costs coverage and each is deliberate. **Naming the wrong milestone
corrupts a movement count; naming none is an explicit null a consumer can
see.** Every coverage figure in this document is therefore lower than it could
be made by a detector willing to guess, and that trade is the reason the
figures can be read at all.

## The suppression floor, and why the twenty are not comparable

The twenty study aerodromes differ in size by a factor of about twenty over the
three-day sample. A percentile computed on a handful of detections is noise
wearing a number's clothes, so Annex A suppresses any cell with fewer than
`r MIN_DETECTED` detections — **marked, not dropped**, because an aerodrome
removed from a table is indistinguishable from one that was never in the study.

## Two negative results from V3 still stand

Neither was re-tested here and neither has been reverted.

**Requiring complete fuzzy evidence made detection worse, on both V3 periods.**
Spark's `least`/`greatest` skip nulls, so a rule missing an input became a
minimum over fewer terms and could out-compete a complete rule. Requiring every
input is obviously more correct, and it removed most phase-based landings,
because ADS-B carries a null velocity often enough that insisting on complete
evidence discards most ground samples. It matters less than it looks now that
the runway-anchored family replaces that path entirely.

**The 60-second smoothing window was contradicted by the second period.** It is
what OpenAP does and what the original port dropped, so it has the strongest
prior justification of any change in the programme; it was near-neutral on one
sample and cost 86% of landing coverage on the other. One confound is stated
with it: the second period's tracks predate the altitude repair, so input
preparation differed as well as the month. The actionable half is that a fix
behaving this differently on noisier input should not be adopted on the
strength of its rationale alone.

# Breaking changes for consumers

The new version string is `events_v0.2.0`. `EventConfig.legacy()` still
reproduces `events_v0.0.2` exactly, and the benchmark's baseline rung
reproduces `events_v0.1.0`, so **any past month remains reproducible** under
the configuration that produced it.

## `landing` has changed meaning, and kept its name

This is the one change that can corrupt an existing consumer silently, so it
leads.

| | Under `events_v0.0.2` / `v0.1.0` | Under `events_v0.2.0` |
|---|---|---|
| Question | did the fuzzy phase go descent → ground | when did the aircraft cross the landing threshold |
| ICAO | — | **T16** |
| Instant | a sampled state vector | interpolated to the threshold plane |
| Scored against | APDF ALDT | **nothing** — ALDT is [Touchdown]'s |

A consumer reading `landing` as touchdown will now be reading the threshold
crossing, which is earlier by the flare and the distance from threshold to
touchdown point. **Two things distinguish them in the published data**: the
`version` column, and `info.milestone`, which is `T16` under the new meaning and
absent under the old. Consumers wanting touchdown must move to `touchdown`.

## Types no longer emitted under `events_v0.2.0`

Each is superseded, not deleted: it remains reachable under the configuration
that published it, and each line points at the chapter that documents its
successor.

| Retired type | Replaced by | Why |
|---|---|---|
| `take-off` | [Airborne] (T08) | fuzzy phase transition → interpolated 15 ft AGL crossing on a classified departure. V3 measured `take-off` at 0.39% of departures <!-- literal-ok: V3's published figure for the 2025 period; this study's data contains no `take-off` rows to read it from -->; `airborne` reaches `r pct(sc("airborne","coverage_pct"),1)` here. The two were not run on the same sample, so the gap is indicative rather than a like-for-like result — but it is three orders of magnitude, not a margin |
| `AOBT` | [Off-block and on-block] (T04) | renamed to `off-block`; same detector |
| `AIBT` | [Off-block and on-block] (T21) | renamed to `on-block`; same detector |
| `entry-runway` | [Line-up], [Runway crossings] | one type conflated departures, arrivals and taxi crossings |
| `exit-runway` | [Runway vacated], [Runway crossings] | same |

Already retired at `events_v0.1.0` and still not emitted:
`first-xing-fl{50,70,100,245}` and `last-xing-fl{50,70,100,245}`, replaced by
the crossing sequence number in [Flight-level crossings] — the first crossing
is `crossing_seq == 1`, the last is the maximum.

### `ATOT`/`ALDT` are retained, not retired

The A-CDM `airborne`/`touchdown` were designed to *replace* `ATOT`/`ALDT`.
On coverage they do not: measured on this sample `ATOT` answers
`r pct(sc("ATOT","coverage_pct"),1)` of reachable departures against
`airborne`'s `r pct(sc("airborne","coverage_pct"),1)`, and `ALDT`
`r pct(sc("ALDT","coverage_pct"),1)` against `touchdown`'s
`r pct(sc("touchdown","coverage_pct"),1)`. On timing the new pair is the better of the two — at the
three aerodromes where APDF resolves to the second the takeoff instant lands
within a few seconds of truth, against `ATOT`'s
`r sec(sc("ATOT","bias_s"),1)` on this sample (+19 s as V3 measured it on
2025 <!-- literal-ok: V3's published figure for a different period, not a quantity this study's data contains -->). That per-aerodrome figure pools both detectors,
since the per-aerodrome table carries no detector column, so attributing the
few-second result to `airborne` is an inference from `ATOT`'s much larger bias
rather than a direct per-detector measurement. The two methods read the
aircraft at different heights (see
[Timing a runway movement: two methods, four milestones]): `ATOT`/`ALDT` in the
climb and descent up to 1 500 ft, `airborne`/`touchdown` at the 15 ft crossing
on and beside the runway. The second needs surface and low-altitude reception
the first does not. Rather than surrender the older detectors' coverage for the
newer ones' accuracy, `events_v0.2.0` **publishes both**, distinguished by type
string, so a consumer chooses the property it needs. A row-count alarm keyed on
`ATOT`/`ALDT` disappearing would therefore be a false alarm — they are still
emitted.

## Other changes worth a row-count alarm

**`info` keys moved.** The runway designator is `rwy_ident` in the A-CDM
family, where it was `runway` for `ATOT`/`ALDT`. A consumer parsing `info`
should coalesce the two rather than switching on version.

**The event table changes size in both directions.** Nine new runway types and
the PRU tops add events; the retirements and the level-arm change remove them.
Volume is not a proxy for information here, and any consumer with a row-count
alarm on the event table should expect it to fire.

```{r tbl-growth}
#| echo: false
#| results: asis
if (has(inv, c("rung", "n_events"))) {
  a <- sum(as.numeric(inv$n_events[inv$rung == BASE]), na.rm = TRUE)
  b <- sum(as.numeric(inv$n_events[inv$rung == SHIPPED]), na.rm = TRUE)
  if (is.finite(a) && is.finite(b) && a > 0)
    cat(sprintf(paste0("On the three-day 2026 sample: **%s events under the V3 ",
                       "configuration, %s under V4** — a factor of %.2f.\n\n"),
                fm(a), fm(b), b / a))
} else {
  pending("The event-volume comparison")
}
```

**Every top of climb and descent now names its algorithm** in `info.method`
(`phase` or `pru`), and four type names exist where two did. A consumer
selecting `type = 'top-of-climb'` gets the fuzzy pair only, unchanged; one
wanting PRU's must ask for `top-of-climb-cco`.

**Event identifiers are derived from the event's own identity** — track, type,
time and version — rather than from a partition counter, so a re-run produces
the same identifiers. This makes a re-run idempotent in *content*; it does not
make the write idempotent, and re-processing safety still rests on the
per-month progress logs.

**Track identity changed shape independently of this version.** The A8
segmentation default removed the `_{year}_{month}` suffix from `track_id`, so
any consumer parsing that suffix breaks, and past months do not reproduce. That
is a track-construction change, documented in its own paper, and it is
mentioned here only because it lands in the same release window.

# Limitations

The ones that bound how this document should be read, collected.

**The study set is the best-covered twenty aerodromes**, chosen on a ranking
computed over the same three days that are scored. Every figure is an upper
bound, and a detector limited by reception looks better here than anywhere else
in the network.

**Most reference times are minute-rounded.** A bias under a minute at such an
aerodrome cannot be distinguished from APDF's own quantisation. Annex A carries
the resolution per aerodrome for exactly this reason.

**The reference join has a ceiling** — `ID` is null on
`r pct(100 * br$null_apdf_id / br$movements)` of the APDF rows this study
scores against, and only `r pct(br$reach_pct)` of movements reach an `icao24`
at all. Both are measured on this period's own extract in
[Ground truth, and its ceiling], and together they cap every coverage figure
above.

**Ten milestone families have no external truth at all**: (1) the tops of
climb and descent, (2) both level-segment arms, (3) the flight-level crossings,
(4) every airport surface event, (5) `line-up`, (6) `take-off-roll`, (7)
`landing` as an instant — ALDT belongs to `touchdown` — (8) `runway-vacated`,
(9) the runway crossings and (10) `go-around`. For the level family and the PRU tops the substitute is
conformance to a published methodology, tested on synthetic geometry. For the
others there is none, and their counts are reported while their timing is
unvalidated and stated as such.

**Five tuning parameters are module constants, not config fields** —
`ROLL_MAX_GAP_SECONDS`, `AIRBORNE_HYSTERESIS_FT`, `THRESHOLD_PLANE_HYSTERESIS_NM`,
`GOAROUND_HYSTERESIS_FT` and `ARRIVAL_LEAD_SECONDS`. They are documented where
they sit, but they are not sweepable without editing code, which contradicts
the convention every other threshold in this document follows.

**Known defects carried, not fixed**: the circular median described in
[Cross-cutting findings]; the `when`-chain ordering sensitivity in the legacy
phase family; the PRU arm's interpolation-grid cap, which truncates the tail of
a pathologically long track; and the non-deterministic go-around low point on a
flat minimum.

**No parameter sweep has been run.** Every threshold is named in `EventConfig`,
which is what makes one possible, and V3's chapter on designing one — an
objective that trades coverage against error, tolerances taken from the
inter-source floor rather than chosen, fit on one period and validated on the
other — still describes the work that has not been done.

# Provenance

```{r regen-note}
#| echo: false
#| results: asis
if (!is.null(REGEN_NOTE)) cat(sprintf("**%s**\n\n", REGEN_NOTE))
```

Every figure on this page is produced by `benchmarks/regenerate_events_v4.py`,
which re-runs a job only when the source files that job depends on have changed
since the output was written. **An output with no manifest entry below is
reported in the text as not yet computed rather than shown as fact.**

```{r tbl-provenance}
#| echo: false
#| results: asis
if (!is.null(manifest) && length(manifest) > 0) {
  cat("| Output | Script | Git SHA | Produced |\n|---|---|---|---|\n")
  for (name in sort(names(manifest))) {
    e <- manifest[[name]]
    if (is.null(e$script)) next
    dirty <- if (isTRUE(e$git_dirty)) " (dirty)" else ""
    cat(sprintf("| `%s` | `%s` | `%s`%s | %s |\n", name, e$script, e$git_sha,
                dirty, substr(e$produced_utc, 1, 10)))
  }
  cat("\n")
} else {
  cat(paste0("**No manifest is staged.** No figure in this document has been ",
             "computed against the checked-out code. Every table above that ",
             "depends on data says so in place, and this page is a draft of the ",
             "structure and the method, not of the results.\n\n"))
}
```

```{r tbl-staged}
#| echo: false
#| results: asis
files <- c("ladder_2026.csv", "per_airport_2026.csv", "runway_2026.csv",
           "resolution_2026.csv", "inventory_2026.csv", "bridge_2026.json")
cat("| Expected output | Staged |\n|---|---|\n")
for (f in files)
  cat(sprintf("| `%s` | %s |\n", f, if (file.exists(file.path("data", f))) "yes" else "**no**"))
cat("\n")
```

Four caveats belong with the provenance rather than with any figure.

**The staleness check is not credential-free.** `provenance.inputs_changed`
reports "not stale" when it cannot reach S3 to look, so a check run without
credentials is not evidence of currency. The stage that builds the 2026 flight
list reports `cannot check (no credentials)` — i.e. stale — rather than
passing, which is the correct direction, but the job-level check does not have
that property.

**The reference-join ceiling and the scored population are different scopes.**
The ceiling is computed over the movements the bridge could reach; the scores
are computed over the study set on three days. Reading the first as the
denominator of the second overstates coverage.

**The reference extract is a month; the sample is three days.** The scorer
restricts truth to the benchmark days before scoring. This is stated twice in
this document because getting it wrong produces a plausible table.

**V3's staged outputs now read stale against this code.** V4's changes touch
`event_bench.py`, which is in V3's dependency set, so V3's fingerprints moved
even though its committed CSVs did not. V3's numbers are traceable to the
version that published them and were not re-run.

# Annex A — Accuracy by aerodrome {.unnumbered}

One table per milestone, one row per aerodrome, in coverage-ranking order —
EBBR first, UGKO twentieth. The columns are the same throughout:

- **n truth** — reference movements for that aerodrome and milestone, over the
  three sample days, after the bridge.
- **Coverage** — the share of those with a detection aligned to them.
- **Bias** — median signed error, detection minus reference. Positive is late.
- **MAD** — median absolute error.
- **p90** — 90th percentile of absolute error.
- **within 60 s** — share of *truth* rows detected within a minute.
- **Truth resolution** — `second` where most of that aerodrome's reference
  times carry a non-zero seconds field, `minute` otherwise.

**Read the last column first.** At a `minute` aerodrome a bias under 30 s is
inside the reference's own quantisation and means nothing; at a `second`
aerodrome it is a measurement. Comparing an aerodrome of one kind against one
of the other without that column is the specific mistake this annex exists to
prevent.

Dashes are **suppressed cells, not missing detections**: fewer than
`r MIN_DETECTED` detections, below the floor at which a percentile carries
information. The detections behind each dash are listed under the table, so a
suppressed row can still be told apart from an absent one.

**Every figure here is measured against APDF**, not against another OPDI
milestone: `n truth` counts APDF movements at that aerodrome and the bias is the
detected instant minus the APDF one. The section headings said "against
`airborne`" in an earlier draft, which read as though one OPDI milestone were
being scored against another; nothing of the sort is computed anywhere in this
paper.

**Each row pools every detector that answers the milestone.** Under
`events_v0.2.0` the ATOT milestone is answered by both `ATOT` and `airborne`,
and the ALDT milestone by both `ALDT` and `touchdown`, so a row is "what OPDI
published for this movement, whichever detector produced it" — the question a
consumer of the table actually has. It is *not* a per-detector figure, because
the scorer groups on (aerodrome, milestone) and carries no detector column. For
the per-detector split, network-wide, see [What the evidence says]; the
per-detector split *per aerodrome* is not measured by this campaign.

## A.1 ATOT — against APDF {.unnumbered}

```{r annex-atot}
#| echo: false
#| results: asis
airport_table("ATOT", what = "The per-aerodrome ATOT table")
```

## A.2 ALDT — against APDF {.unnumbered}

```{r annex-aldt}
#| echo: false
#| results: asis
airport_table("ALDT", what = "The per-aerodrome ALDT table")
```

## A.3 AOBT — against APDF {.unnumbered}

```{r annex-aobt}
#| echo: false
#| results: asis
airport_table("AOBT", what = "The per-aerodrome AOBT table")
```

## A.4 AIBT — against APDF {.unnumbered}

```{r annex-aibt}
#| echo: false
#| results: asis
airport_table("AIBT", what = "The per-aerodrome AIBT table")
```

## A.5 Per-detector accuracy by aerodrome {.unnumbered}

A.1--A.4 pool the detectors that answer a milestone: the detection nearest the
reference wins and the other is discarded, so a row is *what OPDI published for
that movement*. These tables keep the detectors apart. Both sides are scored
against the same APDF truth and, because a detector that produced nothing for a
movement is still charged that movement, the two rows of a pair are read down
the column against each other.

**The two questions have different answers, and this is where that shows.**

```{r annex-det}
#| echo: false
#| results: asis
det_pair_table <- function(ms, dets, what) {
  need <- c("gt_airport", "milestone", "det_type", "coverage_pct", "bias_s",
            "mad_s", "n_truth", "n_detected")
  if (!has(perd, need)) return(invisible(pending(what)))
  d <- perd[perd$milestone == ms, , drop = FALSE]
  if (!nrow(d)) return(invisible(pending(what)))
  apts <- RANKING[RANKING %in% d$gt_airport]
  apts <- c(apts, setdiff(unique(d$gt_airport), apts))
  cat(sprintf("| Aerodrome | n truth | %s coverage | %s bias | %s coverage | %s bias | Truth resolution |\n",
              dets[1], dets[1], dets[2], dets[2]))
  cat("|---|---:|---:|---:|---:|---:|---|\n")
  for (a in apts) {
    r1 <- d[d$gt_airport == a & d$det_type == dets[1], , drop = FALSE]
    r2 <- d[d$gt_airport == a & d$det_type == dets[2], , drop = FALSE]
    if (!nrow(r1) && !nrow(r2)) next
    nt <- if (nrow(r1)) r1$n_truth[1] else r2$n_truth[1]
    cat(sprintf("| %s | %s | %s | %s | %s | %s | %s |\n", a, fm(nt),
                if (nrow(r1)) pct(r1$coverage_pct[1]) else "—",
                if (nrow(r1)) sec(r1$bias_s[1]) else "—",
                if (nrow(r2)) pct(r2$coverage_pct[1]) else "—",
                if (nrow(r2)) sec(r2$bias_s[1]) else "—",
                truth_resolution(a, ms)))
  }
  cat("\n")
  invisible(NULL)
}
det_pair_table("ATOT", c("ATOT", "airborne"), "The per-detector ATOT table")
```

```{r annex-det-aldt}
#| echo: false
#| results: asis
det_pair_table("ALDT", c("ALDT", "touchdown"), "The per-detector ALDT table")
```

**What these two tables say.** On departures `airborne` beats `ATOT` on
coverage (strictly greater) at
**`r pair_compare("airborne","ATOT","ATOT")$a_gt` of the twenty** aerodromes,
and where it does the gap is not marginal: LOWW
**`r pct(perd_val("LOWW","airborne","ATOT"),1)` against
`r pct(perd_val("LOWW","ATOT","ATOT"),1)`**, LEIB
**`r pct(perd_val("LEIB","airborne","ATOT"),1)` against
`r pct(perd_val("LEIB","ATOT","ATOT"),1)`**. The two tie exactly at
**`r pair_compare("airborne","ATOT","ATOT")$tie`** more (LFLL, LSZH, UGKO), and
`ATOT` is strictly ahead at the remaining
**`r pair_compare("airborne","ATOT","ATOT")$b_gt`**. On arrivals `touchdown`
beats `ALDT` (strictly greater) at
**`r pair_compare("touchdown","ALDT","ALDT")$a_gt`** of the twenty.

That is the whole case for publishing both, stated precisely. It is not that the
A-CDM family is better; on this sample it is not. It is that `ATOT` collapses at
a small number of aerodromes and `airborne` does not collapse at the same ones,
so the pooled figure is worth
**`r sprintf("%+.1f", pooled_cov("ATOT") - sc("ATOT","coverage_pct"))` points**
on departures. On arrivals `ALDT` is ahead everywhere and pooling gains
**`r sprintf("%+.2f", pooled_cov("ALDT") - sc("ALDT","coverage_pct"))` points**
-- `touchdown` earns its
place in the table on its A-CDM semantics and its 4D fix, not on adding
answers a consumer could not otherwise get.

```{r annex-missing}
#| echo: false
#| results: asis
if (has(pera, "milestone")) {
  extra <- setdiff(unique(pera$milestone), c("ATOT", "ALDT", "AOBT", "AIBT"))
  for (m in extra) {
    cat(sprintf("\n## A.%d %s {.unnumbered}\n\n", 4 + match(m, extra), m))
    airport_table(m, what = sprintf("The per-aerodrome %s table", m))
  }
}
```
