---
title: "Detecting departure and destination aerodromes from ADS-B"
subtitle: "Version 4.1 — detection inside the pipeline"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every figure and table comes from a committed CSV under data/,
# refreshed by running benchmarks/benchmark_modes.py in the opdi repository.
cache <- function(f) {
  p <- file.path("data", f)
  if (!file.exists(p)) return(NULL)
  read.csv(p, stringsAsFactors = FALSE)
}
modes  <- cache("mode_comparison.csv")
grid   <- cache("sweep_radius_height.csv")
pen    <- cache("sweep_penalty.csv")
apt    <- cache("per_airport.csv")

pct <- function(x) ifelse(is.na(x), "—", sprintf("%.2f%%", 100 * x))
pp  <- function(x) ifelse(is.na(x), "—", sprintf("%+.2f pp", 100 * x))

# One accessor, so prose and tables cannot drift apart.
m <- function(mode, field) {
  v <- modes[[field]][modes$mode == mode]
  if (length(v) == 0) NA_real_ else v[1]
}
# Sweep cell lookup, likewise.
g <- function(r, h, field) {
  v <- grid[[field]][grid$radius_nm == r & grid$height_ft == h]
  if (length(v) == 0) NA_real_ else v[1]
}
UNRESTRICTED <- max(grid$height_ft)
N_GT <- modes$n_ground_truth[1]

# The operating point the three flight lists were produced at. Read out of the
# sweep frames rather than typed, so a re-run cannot leave the prose behind:
# the penalty sweep holds radius and height fixed, the radius x height sweep
# holds the penalty fixed, and between them they pin all three.
OP_RADIUS  <- pen$radius_nm[1]
OP_HEIGHT  <- pen$height_ft[1]
OP_PENALTY <- grid$penalty_nm[1]
OP <- sprintf("%g NM detection radius, %s ft abstention height, %g NM scheduled-service penalty",
              OP_RADIUS, format(OP_HEIGHT, big.mark = ",", trim = TRUE), OP_PENALTY)
```

## Summary {.unnumbered}

Versions 1–3 of this study ran outside the OPDI pipeline. They reimplemented
aerodrome proximity with their own coordinate grid, which made every
recommendation awkward to adopt: the findings were expressed in nautical miles
while production ranks by H3 cell membership, so acting on them meant
re-deriving rather than transcribing.

Version 4 moves detection **into the pipeline**. Three modes now share one
candidate-detection stage — H3 as the index, exact great-circle distance as the
comparison — and are selected by a `mode` parameter on the production function.
The pipeline ran end to end on three days of 2025-06, producing a real flight
list scored against Network Manager ground truth.

Three results:

**Endpoint abstention buys `r pp(m("endpoint","adep_coverage") - m("trend","adep_coverage"))`
of departure coverage for `r pp(m("endpoint","adep_accuracy") - m("trend","adep_accuracy"))`
of accuracy.** A real trade, not a free win — and a different conclusion from
version 3, for a reason worth stating up front.

**The detection radius barely matters; the height threshold is everything.**
Extending the radius from 40 NM to 110 NM changes departure coverage by
`r pp(g(110, 15000, "adep_coverage") - g(40, 15000, "adep_coverage"))`. Raising
the height cut-off from on-ground-only to unrestricted changes it by more than
fifty points. A distance cap costs almost nothing, which is worth knowing
because it is the parameter most likely to be tuned.

**Preferring aerodromes with scheduled service is close to free.** Ranking
candidates on a distance penalised for aerodromes without scheduled service
gains
**`r pp(pen$adep_accuracy[pen$penalty_nm == 10] - pen$adep_accuracy[pen$penalty_nm == 0])`
of departure accuracy** and
**`r pp(pen$ades_accuracy[pen$penalty_nm == 10] - pen$ades_accuracy[pen$penalty_nm == 0])`
of arrival accuracy**, at no measurable cost in coverage.

::: {.callout-important}
## The production baseline improved too

Version 3 concluded that endpoint abstention beat production on *both* coverage
and accuracy. That no longer holds, and the reason is not that abstention got
worse — it is that **the baseline got better**. Running the altitude-trend
algorithm on H3 candidates with exact distances lifts its departure accuracy
from 97.45% in version 3 to `r pct(m("trend","adep_accuracy"))` here.

The comparison in this version is therefore harder and fairer than the one
before it. Version 3's numbers are not comparable to these and remain published
unchanged.
:::

::: {.callout-note collapse="true"}
## What changed in 4.1

No new computation — 4.1 reports more of the same sweep. Every headline number
is unchanged.

* Results are now given for **both roles on all four measures**: arrival
  overall correctness and both in-area figures were computed but not printed.
* The sweep is plotted **on accuracy as well as coverage**, for departures and
  arrivals. This is the substantive addition: accuracy peaks near the ground
  and falls as height is admitted, so the coverage view alone overstates the
  case for a high threshold (@sec-accuracy-sweep).
* The per-aerodrome cross-check names the **mode and parameters** it was
  produced at, extends to the **30 busiest** aerodromes, and separates the
  out-of-area bucket from the movement-count ratio it was distorting.
* Figure captions are generated from the data rather than written by hand.
:::

## What the pipeline does before any method runs {#sec-preprocessing}

The population a method is judged on is fixed long before the method executes.
Each stage below discards something, and what survives is what detection sees.

**1. Ingest** (`ingestion/osn_statevectors.py`). Hourly partitions are read
from the OpenSky archive and immediately filtered to the European bounding box
`(-25.87, 26.75, 49.66, 70.26)` and decimated to one sample every 5 s. Both
filters are applied *before* anything is written. For the three days here that
turned roughly 60 GB of raw feed into **`r format(142630930, big.mark=",")`
state vectors**.

*What is discarded:* everything outside Europe, and four samples in five. The
5 s decimation is the reason a track's "first sample" can be several seconds
into the take-off roll rather than at the stand.

**2. Track construction** (`pipeline/tracks.py`). State vectors are grouped
into tracks by `SHA2(icao24 ‖ callsign)` and split on a gap of over 30 minutes,
or over 15 minutes below 5,000 ft. Each sample is also indexed into H3 cells at
resolutions 7 and 12.

*What is discarded:* nothing, but the grouping decides what counts as one
flight. Version 2 measured the consequences — roughly one matched flight in ten
is split across more than one track, and a few percent of tracks cover more
than one flight.

**3. Aerodrome zones** (`reference/h3_airport_zones.py`). Each aerodrome is
surrounded by concentric rings of H3 res-7 cells: 5 NM steps out to 40 NM, then
10 NM steps to 110 NM. Every cell carries the ring band it belongs to, so a
consumer chooses its own radius at read time. **`r format(33751619, big.mark=",")`
(aerodrome, cell) rows** for the 1,353 large and medium aerodromes in the area.

*Why it reaches 110 NM:* the same table serves ASMA ring crossings at 40 and
100 NM. Detection uses a fraction of it.

**4. Endpoint candidates** (`pipeline/flights.py`). For each track, the first
and last sample are matched against the zone table on `h3_res_7`, and the exact
great-circle distance to each candidate aerodrome is computed. The result is
cached as `opdi_endpoint_candidates` — **`r format(7491655, big.mark=",")` rows**.

This is the step that makes the sweeps in this paper affordable. Deriving
candidates from state vectors is expensive; once cached, changing a threshold
is a filter and a re-rank over a few million rows rather than a pass over 142
million. Reducing to the two endpoint samples *before* the join is also what
makes the full 110 NM reach cheap — about a million rows to match, not the
whole month.

::: {.callout-note}
## H3 as an index, distance as the comparison

The H3 join answers "which aerodromes are plausibly near this sample" in one
equi-join, without a spatial cross product. It does not answer "how near",
because a cell inside a 40 NM ring can still be more than 40 NM from the
aerodrome. So every candidate is then measured exactly, and the distance — not
the cell — decides.
:::

## How the metrics work {#sec-metrics}

This study turns on the difference between measures that are easy to conflate,
so each is defined before any number appears.

Let $G$ be the ground-truth flights and $A \subseteq G$ those the method
answered.

| Measure | Definition | The question it answers |
|---|---|---|
| **Coverage** | $\lvert A \rvert / \lvert G \rvert$ | How often does it say *something*? |
| **Accuracy** | $\lvert \text{correct} \rvert / \lvert A \rvert$ | When it speaks, how often is it right? |
| **Overall correctness** | $\lvert \text{correct} \rvert / \lvert G \rvert$ | Of everything, how much did it get right? |

A flight whose trajectory was never seen counts against coverage. That is
deliberate: ground truth is the denominator throughout, so "we had no data" is
a miss rather than an excluded row.

**Accuracy alone is close to meaningless here.** It says nothing about how
often a method speaks, so a method that answers rarely and cautiously can post
a better accuracy than one doing far more useful work. Consider two methods on
the same flights:

```{r metric-example}
ex <- data.frame(
  Method = c("Cautious", "Eager"),
  Coverage = c("40.00%", "95.00%"),
  Accuracy = c("99.00%", "88.00%"),
  `Overall correctness` = c("39.60%", "83.60%"),
  check.names = FALSE
)
kable(ex, align = "lrrr",
      caption = "Accuracy favours the cautious method; overall correctness shows the eager one labels more than twice as many flights correctly.")
```

Neither is simply better — they suit different uses. A flight list that
downstream users treat as authoritative may well prefer the cautious one. The
point is that **no single number decides it**, which is why all three are
reported throughout.

### Precision and recall, for out-of-area

Some flights have an origin or destination outside the observed area entirely.
"Out of area" is therefore a legitimate *answer*, not a failure — and how well a
method identifies those cases is measured with the standard pair:

| Measure | Definition | The question it answers |
|---|---|---|
| **Precision** | of flights *labelled* out-of-area, the fraction that were | When it says out-of-area, is it right? |
| **Recall** | of flights that *were* out-of-area, the fraction it labelled | Of the cases out there, how many did it catch? |

Precision is accuracy computed on one label instead of all of them. Recall has
no counterpart in the table above, and it is the one that matters here: **high
precision with low recall means a method is right when it fires but rarely
fires**.

In the standard confusion-matrix vocabulary, with "out-of-area" as the positive
class: precision is $TP/(TP+FP)$ and recall is $TP/(TP+FN)$. Coverage has no
confusion-matrix equivalent, because it measures whether the method produced
*any* label rather than which one.

### In-area coverage and accuracy

Because roughly one flight in twelve is genuinely out of area, a method could
raise its apparent accuracy simply by labelling out-of-area liberally. So the
same two measures are also reported **restricted to flights whose truth is a
real aerodrome** — detection skill with the free wins removed. When headline and
in-area figures diverge, the gap is exactly the contribution of out-of-area
labelling.

## The three modes {#sec-modes}

All three read the same cached candidates. They differ in what evidence they
consult and, more importantly, in when they are willing to stay silent.

### `trend` — the current production algorithm

*Evidence:* every state vector below **FL40** that falls inside an aerodrome's
detection zone, currently 30 NM.

*Rule:* for each (track, aerodrome) pair, smooth altitude over a 5-sample
window, count how often the smoothed value rises versus falls, and require a
margin of 4 either way. A positive margin is a take-off, negative a landing.

*Abstains when:* the margin is within ±4 — classified `ambiguous` and **the
flight is dropped entirely**.

*Failure mode:* a flight seen only above FL40 near its origin, or with a noisy
altitude trace, silently loses its aerodrome. The FL40 gate is the binding
constraint, not the 30 NM radius.

### `nearest` — the *traffic* library's rule

*Evidence:* the track's first and last sample. Nothing else — no altitude, no
ground state, no trend.

*Rule:* name the nearest aerodrome by effective distance.

*Abstains when:* never, except when no candidate exists at all.

*Failure mode:* it answers just as confidently for a track that begins in the
cruise as for one that begins at a stand. Its coverage is the highest of any
method and its accuracy the lowest, and both facts have the same cause.

### `endpoint` — nearest, with abstention

*Evidence:* the same two samples, plus their height above **field elevation**
and their `on_ground` flag.

*Rule:* name the nearest aerodrome, but only if the endpoint is within
`radius_nm` **and** either below `height_ft` above that aerodrome's elevation or
flagged on the ground. Otherwise, if the endpoint lies near the boundary of the
observed area, answer out-of-area. Otherwise answer nothing.

*Why height above field elevation:* a fixed altitude cut-off means nothing at
an aerodrome sitting at 5,000 ft. Elevation comes from OurAirports and is
missing for about 3% of large and medium aerodromes; where it is missing the
height test is skipped and only the ground flag can satisfy the rule, rather
than silently comparing against sea level.

*Failure mode:* it inherits `nearest`'s naming and therefore its confusions
between neighbouring aerodromes; the abstention only decides *whether* to
answer, never *what* to answer.

## Results

```{r mode-table}
# Split by role rather than widening: departures and arrivals each get the full
# set -- coverage, accuracy, overall correctness and the in-area restriction --
# so no measure is reported for one role and silently omitted for the other.
tab <- do.call(rbind, lapply(c(adep = "departures", ades = "arrivals"), function(rl) {
  pre <- if (rl == "departures") "adep" else "ades"
  do.call(rbind, lapply(c("trend", "nearest", "endpoint"), function(md) data.frame(
    Movements = rl, Mode = md,
    Coverage = pct(m(md, paste0(pre, "_coverage"))),
    Accuracy = pct(m(md, paste0(pre, "_accuracy"))),
    Overall  = pct(m(md, paste0(pre, "_overall"))),
    `In-area cov` = pct(m(md, paste0(pre, "_inarea_coverage"))),
    `In-area acc` = pct(m(md, paste0(pre, "_inarea_accuracy"))),
    check.names = FALSE)))
}))
kable(tab, row.names = FALSE, align = "llrrrrr",
      caption = sprintf("The three modes on %s ground-truth flights, 2025-06-05 to 07. Overall correctness is coverage x accuracy. Produced at %s.",
                        format(N_GT, big.mark = ","), OP))
```

`endpoint` sits between the two by construction, and that is the whole design:
it keeps `nearest`'s naming rule and adds `trend`'s willingness to stay silent.
It answers `r pp(m("endpoint","adep_coverage") - m("trend","adep_coverage"))`
more departures than production for
`r pp(m("endpoint","adep_accuracy") - m("trend","adep_accuracy"))` of accuracy.

Whether that trade is worth taking is a policy question rather than a technical
one. A flight list treated as authoritative may prefer production's caution;
an analysis needing volume may not.

Arrivals remain weaker than departures on every measure, as in all previous
versions. A departure's first sample is at or near a stand, with the aircraft
stationary shortly before. An arrival's last sample is more often a loss of
reception during descent than a landing, and can be anywhere on the approach.

## What a distance cap costs {#sec-radius}

The sweep varies the detection radius and the abstention height independently,
90 combinations in all, over the cached candidates.

Both measures are plotted for both roles, because the two axes do not merely
differ in strength — they pull in opposite directions, and a coverage-only view
hides that entirely.

```{r sweep-setup}
#| include: false
d <- grid[grid$height_ft < UNRESTRICTED, ]
d$radius <- factor(d$radius_nm, levels = sort(unique(d$radius_nm)),
                   labels = paste0(sort(unique(d$radius_nm)), " NM"))

# Long form, so departures and arrivals become facets of one figure and can be
# read against each other on a shared pair of axes.
sweep_plot <- function(measure, ylab) {
  x <- rbind(
    data.frame(d[c("height_ft", "radius")], Movements = "Departures",
               value = d[[paste0("adep_", measure)]]),
    data.frame(d[c("height_ft", "radius")], Movements = "Arrivals",
               value = d[[paste0("ades_", measure)]]))
  x$Movements <- factor(x$Movements, levels = c("Departures", "Arrivals"))
  ggplot(x, aes(height_ft, value, colour = radius, group = radius)) +
    geom_line(linewidth = 0.6) + geom_point(size = 1.1) +
    facet_wrap(~ Movements) +
    scale_x_continuous("Abstention height above field elevation (ft)",
                       breaks = c(0, 10000, 20000), labels = scales::comma) +
    scale_y_continuous(ylab, labels = scales::percent) +
    scale_colour_viridis_d("Detection radius", end = 0.9) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank())
}

# Captions are built from the data for the same reason the tables are: an
# earlier revision of this page carried a hand-written caption that the numbers
# had already contradicted.
sat <- function(pre, r) g(110, OP_HEIGHT, paste0(pre, "_coverage")) -
                        g(r,   OP_HEIGHT, paste0(pre, "_coverage"))
peak <- function(pre) {
  v <- d[d$radius_nm == OP_RADIUS, ]
  v <- v[order(v$height_ft), ]
  h <- v$height_ft[which.max(v[[paste0(pre, "_accuracy")]])]
  c(height = h, acc = max(v[[paste0(pre, "_accuracy")]]),
    drop = max(v[[paste0(pre, "_accuracy")]]) - g(OP_RADIUS, OP_HEIGHT, paste0(pre, "_accuracy")))
}
pk_dep <- peak("adep"); pk_arr <- peak("ades")
spread <- function(pre) g(5, OP_HEIGHT, paste0(pre, "_accuracy")) -
                        g(110, OP_HEIGHT, paste0(pre, "_accuracy"))

cap_cov <- sprintf(
  "Coverage against abstention height, one line per detection radius. The shape of the curve belongs to the height axis in both panels; the radius only sets how much of it is reachable, and arrivals need more of it than departures. At %s ft a 20 NM radius still leaves %s of departure coverage and %s of arrival coverage on the table against the full 110 NM reach, while at %g NM both are within half a point of it.",
  format(OP_HEIGHT, big.mark = ",", trim = TRUE),
  pp(sat("adep", 20)), pp(sat("ades", 20)), OP_RADIUS)

cap_acc <- sprintf(
  "Accuracy against abstention height, same axes. This is the figure the coverage plot hides. Accuracy peaks close to the ground -- at %g NM, departures at %s ft and arrivals at %s ft -- and every foot admitted above that is bought with correctness: %s for departures by %s ft, %s for arrivals. The radius, which barely touches coverage above 40 NM, spreads accuracy at %s ft by %s for departures and %s for arrivals.",
  OP_RADIUS,
  format(pk_dep["height"], big.mark = ",", trim = TRUE),
  format(pk_arr["height"], big.mark = ",", trim = TRUE),
  pp(-pk_dep["drop"]), format(OP_HEIGHT, big.mark = ",", trim = TRUE),
  pp(-pk_arr["drop"]),
  format(OP_HEIGHT, big.mark = ",", trim = TRUE),
  pp(spread("adep")), pp(spread("ades")))
```

```{r sweep-plot-cov, fig.cap=cap_cov, fig.width=8, fig.height=4}
sweep_plot("coverage", "Coverage")
```

```{r sweep-plot-acc, fig.cap=cap_acc, fig.width=8, fig.height=4}
sweep_plot("accuracy", "Accuracy")
```

```{r radius-cost}
r <- grid[grid$height_ft == OP_HEIGHT, c("radius_nm","adep_coverage","adep_accuracy",
                                         "ades_coverage","ades_accuracy")]
r <- r[order(r$radius_nm), ]
kable(data.frame(
  `Max radius` = paste0(r$radius_nm, " NM"),
  `ADEP cov` = pct(r$adep_coverage), `ADEP acc` = pct(r$adep_accuracy),
  `ADES cov` = pct(r$ades_coverage), `ADES acc` = pct(r$ades_accuracy),
  check.names = FALSE), align = "lrrrr",
  caption = sprintf("Detection radius at a fixed %s ft abstention height. Coverage saturates by about 30 NM for departures and 40 NM for arrivals; beyond that a wider radius trades accuracy for almost no coverage.",
                    format(OP_HEIGHT, big.mark = ",", trim = TRUE)))
```

**A cap costs almost nothing.** Going from 40 NM to the full 110 NM reach
changes departure coverage by
`r pp(g(110,15000,"adep_coverage") - g(40,15000,"adep_coverage"))` and accuracy
by `r pp(g(110,15000,"adep_accuracy") - g(40,15000,"adep_accuracy"))`. The
useful range is 20–40 NM; below that coverage falls away quickly, above it
nothing happens.

Every radius at or above 20 NM traces effectively the same coverage curve: at
5,000 ft they span `r pct(g(20,5000,"adep_coverage"))` to
`r pct(g(110,5000,"adep_coverage"))`, about a point across a five-fold change.
Only the 5 NM and 10 NM lines fall away, and they fall further as the height
threshold rises, because a tight radius costs more once more of the sky is
admitted.

**Height is the parameter that matters.** At `r OP_RADIUS` NM, departure
coverage runs from `r pct(g(OP_RADIUS,0,"adep_coverage"))` with the
on-ground-only rule, through `r pct(g(OP_RADIUS,2000,"adep_coverage"))` at
2,000 ft, to `r pct(g(OP_RADIUS,OP_HEIGHT,"adep_coverage"))` at
`r format(OP_HEIGHT, big.mark=",")` ft. That is over fifty points from one axis
against a fraction of a point from the other.

Arrivals behave the same way with less room: from
`r pct(g(OP_RADIUS,0,"ades_coverage"))` to
`r pct(g(OP_RADIUS,OP_HEIGHT,"ades_coverage"))` over the same range. They also
saturate later in radius — at `r format(OP_HEIGHT, big.mark=",")` ft a 20 NM cap
costs departures `r pp(sat("adep", 20))` of coverage but arrivals
`r pp(sat("ades", 20))`, because an arrival's last recorded sample is further
from the aerodrome than a departure's first.

Two endpoints of the sweep serve as sanity checks, and both behave as they
should: at height 0 the rule reduces to "endpoint on the ground" and gives
`r pct(g(OP_RADIUS,0,"adep_coverage"))`, close to what a surface-samples-only
method achieved in earlier versions; with height unrestricted and a wide radius
it reaches `r pct(g(110,UNRESTRICTED,"adep_coverage"))`, close to `nearest`'s
`r pct(m("nearest","adep_coverage"))`, which is what it should converge to.

### What the coverage view hides {#sec-accuracy-sweep}

Read on accuracy, the same grid tells a different story, and the two together
are what actually determines an operating point.

**Accuracy peaks near the ground, then declines.** At `r OP_RADIUS` NM,
departure accuracy is highest at `r format(pk_dep["height"], big.mark=",")` ft
(`r pct(pk_dep["acc"])`) and arrival accuracy at
`r format(pk_arr["height"], big.mark=",")` ft (`r pct(pk_arr["acc"])`). By
`r format(OP_HEIGHT, big.mark=",")` ft they have given up
`r pp(-pk_dep["drop"])` and `r pp(-pk_arr["drop"])` respectively. Every foot of
extra height admits endpoints that are further into the climb or descent, and
those are exactly the endpoints most likely to sit nearer a neighbouring
aerodrome than the real one.

**Arrivals pay roughly three times as much for the same height.** That ratio is
the single clearest statement of why this study has never recommended changing
the arrivals algorithm: the same parameter that is nearly free for departures
is expensive for arrivals, so one operating point cannot serve both well.

**The radius matters for accuracy even where it does not for coverage.** Above
40 NM a wider radius buys essentially no coverage, but it keeps costing
accuracy: at `r format(OP_HEIGHT, big.mark=",")` ft, going from 5 NM to 110 NM
costs `r pp(-spread("adep"))` of departure accuracy and
`r pp(-spread("ades"))` of arrival accuracy. This is the real argument for a
cap. Version 4 justified one on the grounds that it costs almost no coverage;
the accuracy view shows it is mildly *positive* rather than merely harmless.

**Overall correctness is what reconciles them.** Coverage rises steeply with
height and accuracy falls gently, so their product keeps rising for departures
across the whole range — which is why
`r pct(g(OP_RADIUS,OP_HEIGHT,"adep_coverage") * g(OP_RADIUS,OP_HEIGHT,"adep_accuracy"))`
at `r format(OP_HEIGHT, big.mark=",")` ft beats
`r pct(g(OP_RADIUS,2000,"adep_coverage") * g(OP_RADIUS,2000,"adep_accuracy"))`
at 2,000 ft despite the lower accuracy. A user who cares about a trustworthy
label rather than a complete one should read the accuracy panel and pick a much
lower height than the one recommended here.

## Preferring aerodromes with scheduled service {#sec-penalty}

Version 3 found that the most frequent departure errors formed one pattern: a
scheduled civil airport losing to an adjacent military airfield — Izmir to
Çiğli Airbase, Naples to Grazzanise, Cagliari to Decimomannu, Catania to
Sigonella, Gdańsk to Cewice, all between 7 and 25 NM apart.

The mechanism is straightforward. With 5 s sampling a departing aircraft's
first recorded fix is already into the climb and displaced along the departure
track. Where a military field lies between the airport and that track, it is
genuinely the nearest aerodrome — the rule is not malfunctioning, it is
answering the question it was asked.

Aviation knowledge resolves it without new data: a commercial flight almost
certainly used the aerodrome with scheduled service. Candidates are therefore
ranked on **effective distance** — true distance plus a fixed penalty for
aerodromes without scheduled service — so a military field must be *clearly*
nearest rather than merely nearest.

```{r penalty-table}
p <- pen[order(pen$penalty_nm), ]
kable(data.frame(
  Penalty = paste0(p$penalty_nm, " NM"),
  `ADEP cov` = pct(p$adep_coverage), `ADEP acc` = pct(p$adep_accuracy),
  `ADES cov` = pct(p$ades_coverage), `ADES acc` = pct(p$ades_accuracy),
  check.names = FALSE), align = "lrrrr",
  caption = "Scheduled-service penalty at 40 NM and 15,000 ft. A penalty of 0 reproduces the unbiased ranking exactly and is the control.")
```

The gain is
**`r pp(p$adep_accuracy[p$penalty_nm==10] - p$adep_accuracy[p$penalty_nm==0])`
of departure accuracy** and
**`r pp(p$ades_accuracy[p$penalty_nm==10] - p$ades_accuracy[p$penalty_nm==0])`
of arrival accuracy**, for a coverage change of
`r pp(p$adep_coverage[p$penalty_nm==10] - p$adep_coverage[p$penalty_nm==0])`.

Two features make this trustworthy rather than a lucky constant. The control at
penalty 0 reproduces the unbiased ranking, so the effect is attributable. And
the curve **peaks and turns** — accuracy is best around 10 NM and declines
above it. A penalty that merely favoured larger aerodromes would improve
monotonically; one that corrects a specific bias should overshoot, and this
does.

::: {.callout-warning}
## It is not free — the cost is a mirror error

Version 3 measured what this preference breaks: genuine movements at
non-scheduled aerodromes get pulled to the nearby hub — Biggin Hill to London
City, Northolt to Heathrow. In that analysis the mirror error became the
**largest single class of remaining departure error**, at roughly 59% of them.

The net is favourable, and the accuracy figures above include the cost. But a
flight list built this way will systematically under-report movements at
non-scheduled aerodromes, and anyone analysing general aviation or military
traffic should know that before using it.
:::

## Out-of-area {#sec-ooa}

A flight from Dubai to Amsterdam has a departure aerodrome that no European
ADS-B feed can name: the aircraft entered the observed area already airborne.
Scoring that as a detection failure understates every method. Treating "out of
area" as an answer is an accounting correction, not a detection improvement.

### How it is derived

**On the ground-truth side**, an aerodrome is relabelled `OOA` when its
coordinates fall outside the ingestion bounding box, **or** when its ICAO code
does not resolve in OurAirports. The unresolvable codes are overwhelmingly
non-European, and a code that cannot be located is one no trajectory endpoint
could have been matched to. This is what puts prediction and truth in the same
alphabet.

**On the prediction side**, an endpoint within 30 NM of the bounding box edge
is labelled out-of-area. The longitude margin is scaled by $1/\cos(\text{lat})$:
a degree of longitude shrinks toward the pole, so an unscaled margin would be
roughly twice as strict in northern Norway as in the Canaries.

**Precedence is aerodrome first, border second.** The opposite order looks
equivalent and is not — version 3 found Ponta Delgada, which sits about 8 NM
inside the western edge, having its departures labelled out-of-area with the
aircraft still on the runway. Being near the boundary means "out of area" only
when there is nothing better to say.

**The border test carries no height condition.** A flight leaving the observed
area is at cruise level by definition, so any height cut-off would reject
exactly the endpoints the branch exists to catch.

### What it does to the numbers

```{r ooa-table}
o <- do.call(rbind, lapply(c("trend","nearest","endpoint"), function(md) data.frame(
  Mode = md,
  `truly OOA` = pct(m(md, "adep_gt_ooa_share")),
  `labelled OOA` = pct(m(md, "adep_pred_ooa_share")),
  `OOA recall` = pct(m(md, "adep_ooa_recall")),
  `OOA precision` = pct(m(md, "adep_ooa_precision")),
  `headline acc` = pct(m(md, "adep_accuracy")),
  `in-area acc` = pct(m(md, "adep_inarea_accuracy")),
  check.names = FALSE)))
kable(o, align = "lrrrrrr",
      caption = "Out-of-area for departures. Only `endpoint` emits the label; the others never do, so their recall is zero by construction.")
```

`truly OOA` is a property of the sample, not of the method, so it must be
identical across all three rows — that it is confirms the decomposition is
being computed correctly rather than being a per-method artefact.

The gap between headline and in-area accuracy shows how much out-of-area
labelling contributes. It is small here, which is the honest reading: the
mechanism is right but the geometric test catches only part of what is there.
Version 3 measured this directly at **92–94% precision against 10–15% recall** —
when the rule fires it is almost always correct, but it rarely fires. An
inbound flight's first recorded sample appears wherever a receiver happens to
catch it, which is usually well inside the boundary rather than at it.

**The concept is right and the instrument is wrong.** A phase-based test would
fit the physics better: a track whose first sample is already in the cruise,
neither near an aerodrome nor descending toward one, has an origin outside the
observed area regardless of where it sits. That is the clearest single
improvement available and is not implemented here.

## Cross-check against the reference {#sec-crosscheck}

Flight-level accuracy can hide errors that cancel: a departure misattributed
from A to B leaves both counts wrong and the total right. Movement counts per
aerodrome are what an operational user reads, so they are checked separately.

```{r per-airport}
if (!is.null(apt)) {
  ia <- apt[!is.na(apt$n_truth) & apt$n_truth > 0, ]
  s <- do.call(rbind, lapply(c("departures","arrivals"), function(rl) {
    d <- ia[ia$role == rl, ]
    a <- d[d$airport != "OOA", ]         # real aerodromes only
    b <- a[a$n_truth >= 50, ]
    data.frame(Movements = rl,
      Aerodromes = nrow(a),
      `Count ratio` = sprintf("%.3f", sum(d$n_assigned) / sum(d$n_truth)),
      `Excl. OOA` = sprintf("%.3f", sum(a$n_assigned) / sum(a$n_truth)),
      `Median recall` = pct(median(a$recall, na.rm = TRUE)),
      `Median recall, 50+ movements` = pct(median(b$recall, na.rm = TRUE)),
      check.names = FALSE)
  }))
  kable(s, row.names = FALSE, align = "lrrrrr",
        caption = sprintf("Movement counts against Network Manager, from the `endpoint` flight list at %s. A count ratio below 1 is an undercount. The second ratio drops the out-of-area bucket, whose own recall is poor (@sec-ooa) and which otherwise carries the shortfall of every unlabelled flight.", OP))
}
```

```{r per-airport-top}
if (!is.null(apt)) {
  # OOA is excluded: it is not an aerodrome, and its counts are reported in
  # their own right in the out-of-area section.
  top <- apt[apt$role == "departures" & apt$airport != "OOA", ]
  top <- head(top[order(-top$n_truth), ], 30)
  kable(data.frame(
    Aerodrome = top$airport,
    `NM departures` = format(top$n_truth, big.mark = ","),
    `OPDI assigned` = format(top$n_assigned, big.mark = ","),
    `Count ratio` = sprintf("%.2f", top$count_ratio),
    Recall = pct(top$recall),
    check.names = FALSE), align = "lrrrr",
    caption = sprintf("The 30 busiest aerodromes by Network Manager departures, against the `endpoint` flight list at %s. Out-of-area is excluded -- it is not an aerodrome and is reported separately in @sec-ooa. *Count ratio* is OPDI movements divided by Network Manager movements, so it can sit near 1 even when the individual flights are wrong; *recall* is the share of Network Manager departures OPDI put at this aerodrome, and is the stricter measure.", OP))
}
```

The error is an **undercount** rather than an overcount, which is the
correctable direction: a systematic shortfall can be adjusted for, an inflated
count cannot. Roughly six points of it are the out-of-area bucket rather than
per-aerodrome detection failure — flights the method never labelled land
nowhere, so they show up as missing movements at whichever aerodrome the truth
names.

```{r recall-scale}
#| include: false
dep30 <- apt[apt$role == "departures" & apt$airport != "OOA", ]
dep30 <- dep30[order(-dep30$n_truth), ]
r30 <- head(dep30$recall, 30)
```

Recall also scales sharply with aerodrome size. Across the busiest thirty it
runs from `r pct(min(r30))` to `r pct(max(r30))`, against a median of
`r pct(median(dep30$recall, na.rm = TRUE))` over all `r nrow(dep30)` aerodromes
with departures in the sample. The flight list is therefore substantially more
complete at large aerodromes than small ones, which matters before comparing
across aerodrome classes.

## Recommendation

**Adopt `endpoint` for departures if coverage matters more than the last point
of accuracy.** It adds `r pp(m("endpoint","adep_coverage") - m("trend","adep_coverage"))`
of coverage for `r pp(m("endpoint","adep_accuracy") - m("trend","adep_accuracy"))`,
and it is a smaller algorithm than the one it replaces — no smoothing window,
no margin parameter, no per-aerodrome vote. If accuracy is paramount, keep
`trend`; it is genuinely the more accurate rule now.

**Use 20–40 NM and a scheduled-service penalty of 10 NM.** Both are on
plateaus, so neither is delicate.

**Do not change the arrivals algorithm on this evidence.** Arrivals are weaker
on every measure and every version of this study has found the same.

**Publish the provenance column.** `adep_source` / `ades_source` distinguish
"named an aerodrome", "outside the observed area" and "could not determine".
Today all three are a single null in the published flight list, and they mean
entirely different things to a consumer. This is the highest-value change per
line of code in the whole study.

## Limitations {#sec-limitations}

* **Three days**, one sample. Versions 2 and 3 used two samples a year apart;
  this version re-runs the whole pipeline, which is expensive, so it uses one.
  The mode ranking is robust across earlier versions, but the exact figures
  here have not been validated on a second period.
* **Out-of-area recall is low.** The geometric border test is the wrong
  instrument; the phase-based test in @sec-ooa is the successor.
* **The scheduled-service preference has a cost** that these accuracy figures
  include but do not itemise: movements at non-scheduled aerodromes are pulled
  to nearby hubs.
* **The sweep re-implements the endpoint rule** as a DataFrame operation
  instead of calling the pipeline's `classify_endpoints`, because a sweep cell
  needs only `(track_id, adep, ades)` while the pipeline method builds a whole
  flight list. The headline `endpoint` row *does* come from the pipeline path,
  so a divergence between the two would surface as the sweep disagreeing with
  its own mode — but this is a duplication and worth removing.
* **`track_id` is taken as given**, with the fragmentation and merging rates
  version 2 measured.

## Reproducing

Reference zones, state vectors and tracks, in the `opdi` repository:

```bash
RANGE="--env opensky --start 2025-06-05 --end 2025-06-08"
opdi run $RANGE --step 00a   # zones: 5 NM bands to 110 NM
opdi run $RANGE --step 01    # state vectors, bbox and 5 s applied at ingest
opdi run $RANGE --step 02    # tracks, with H3 res 7 and 12
```

Candidates once, then a flight list per mode:

```python
from datetime import date
from opdi.pipeline.flights import FlightListProcessor

fp = FlightListProcessor(spark, config)
fp.build_endpoint_candidates(date(2025, 6, 1))
for mode in ("trend", "nearest", "endpoint"):
    fp.process_dai(date(2025, 6, 1), mode=mode, skip_if_processed=False,
                   abstention_radius_nm=40, abstention_height_ft=15000,
                   sched_penalty_nm=10,
                   table_name=f"research/flight_list_{mode}",
                   write_mode="overwrite")
```

Scoring and both sweeps. This runs locally: the inputs are small once the
candidates are cached, and the cluster namespace is shared.

```bash
python benchmarks/benchmark_modes.py --months 202506 \
    --days 2025-06-05 2025-06-06 2025-06-07 --local --results-dir <dir>
```

Rendering this page needs no credentials, no database and no cluster.
