---
title: "Detecting departure and destination aerodromes from ADS-B"
subtitle: "Version 4.5 — detection inside the pipeline"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every figure and table comes from a committed CSV under data/,
# refreshed by running benchmarks/benchmark_modes.py in the opdi repository.
cache <- function(f) {
  p <- file.path("data", f)
  if (!file.exists(p)) return(NULL)
  read.csv(p, stringsAsFactors = FALSE)
}
modes  <- cache("mode_comparison.csv")
grid   <- cache("sweep_radius_height.csv")
pen    <- cache("sweep_penalty.csv")
apt    <- cache("per_airport.csv")
typ    <- cache("per_type.csv")
cone   <- cache("sweep_cone.csv")

pct <- function(x) ifelse(is.na(x), "—", sprintf("%.2f%%", 100 * x))
pp  <- function(x) ifelse(is.na(x), "—", sprintf("%+.2f pp", 100 * x))

# One accessor, so prose and tables cannot drift apart.
m <- function(mode, field) {
  v <- modes[[field]][modes$mode == mode]
  if (length(v) == 0) NA_real_ else v[1]
}
# Sweep cell lookup, likewise.
g <- function(r, h, field) {
  v <- grid[[field]][grid$radius_nm == r & grid$height_ft == h]
  if (length(v) == 0) NA_real_ else v[1]
}
UNRESTRICTED <- max(grid$height_ft)
N_GT <- modes$n_ground_truth[1]

# The operating point the three flight lists were produced at. Read out of the
# sweep frames rather than typed, so a re-run cannot leave the prose behind:
# the penalty sweep holds radius and height fixed, the radius x height sweep
# holds the penalty fixed, and between them they pin all three.
OP_RADIUS  <- pen$radius_nm[1]
OP_HEIGHT  <- pen$height_ft[1]
OP_PENALTY <- grid$penalty_nm[1]
OP <- sprintf("%g NM detection radius, %s ft abstention height, %g NM scheduled-service penalty",
              OP_RADIUS, format(OP_HEIGHT, big.mark = ",", trim = TRUE), OP_PENALTY)

# --- flight counts -------------------------------------------------------
# Rates hide the size of a trade; counts do not. Everything below is derived
# from the same coverage/overall pair the rest of the paper uses, so the two
# views can never disagree.
cnt <- function(cov, ovr) {
  ans <- round(cov * N_GT); cor <- round(ovr * N_GT)
  list(answered = ans, correct = cor, wrong = ans - cor, silent = N_GT - ans)
}
cell <- function(r, h, role) {
  x <- grid[grid$radius_nm == r & grid$height_ft == h, ]
  if (nrow(x) == 0) return(NULL)
  c(cnt(x[[paste0(role, "_coverage")]][1], x[[paste0(role, "_overall")]][1]),
    list(cov = x[[paste0(role, "_coverage")]][1],
         acc = x[[paste0(role, "_accuracy")]][1],
         ovr = x[[paste0(role, "_overall")]][1]))
}
prod_cell <- function(role) c(
  cnt(m("trend", paste0(role, "_coverage")), m("trend", paste0(role, "_overall"))),
  list(cov = m("trend", paste0(role, "_coverage")),
       acc = m("trend", paste0(role, "_accuracy")),
       ovr = m("trend", paste0(role, "_overall"))))
num <- function(x) format(x, big.mark = ",", trim = TRUE)
sgn <- function(x) sprintf("%+s", format(x, big.mark = ",", trim = TRUE))

#: The grid the operating-point discussion is argued over.
SEL_R <- c(20, 30, 40)
SEL_H <- c(15000, 20000, 25000, 30000)
```

## Summary {.unnumbered}

Versions 1–3 of this study ran outside the OPDI pipeline. They reimplemented
aerodrome proximity with their own coordinate grid, which made every
recommendation awkward to adopt: the findings were expressed in nautical miles
while production ranks by H3 cell membership, so acting on them meant
re-deriving rather than transcribing.

Version 4 moves detection **into the pipeline**. Three modes now share one
candidate-detection stage — H3 as the index, exact great-circle distance as the
comparison — and are selected by a `mode` parameter on the production function.
The pipeline ran end to end on three days of 2025-06, producing a real flight
list scored against Network Manager ground truth.

Three results:

**Endpoint abstention buys `r pp(m("endpoint","adep_coverage") - m("trend","adep_coverage"))`
of departure coverage for `r pp(m("endpoint","adep_accuracy") - m("trend","adep_accuracy"))`
of accuracy.** A real trade, not a free win — and a different conclusion from
version 3, for a reason worth stating up front.

**The detection radius barely matters; the height threshold is everything.**
Extending the radius from 40 NM to 110 NM changes departure coverage by
`r pp(g(110, 15000, "adep_coverage") - g(40, 15000, "adep_coverage"))`. Raising
the height cut-off from on-ground-only to unrestricted changes it by more than
fifty points. A distance cap costs almost nothing, which is worth knowing
because it is the parameter most likely to be tuned.

**Preferring aerodromes with scheduled service is close to free.** Ranking
candidates on a distance penalised for aerodromes without scheduled service
gains
**`r pp(pen$adep_accuracy[pen$penalty_nm == 10] - pen$adep_accuracy[pen$penalty_nm == 0])`
of departure accuracy** and
**`r pp(pen$ades_accuracy[pen$penalty_nm == 10] - pen$ades_accuracy[pen$penalty_nm == 0])`
of arrival accuracy**, at no measurable cost in coverage.

::: {.callout-important}
## The production baseline improved too

Version 3 concluded that endpoint abstention beat production on *both* coverage
and accuracy. That no longer holds, and the reason is not that abstention got
worse — it is that **the baseline got better**. Running the altitude-trend
algorithm on H3 candidates with exact distances lifts its departure accuracy
from 97.45% in version 3 to `r pct(m("trend","adep_accuracy"))` here.

The comparison in this version is therefore harder and fairer than the one
before it. Version 3's numbers are not comparable to these and remain published
unchanged.
:::

::: {.callout-note collapse="true"}
## What changed since version 4

Every headline number is unchanged throughout. The point releases report more
of what version 4 already computed, plus three new computations.

**4.5**

* A **result analysis** that picks an operating point and argues for it
  (@sec-operating-point), in flights rather than percentages, separately for
  departures and arrivals. It finds the currently published parameters are past
  break-even, and that no arrival setting beats production.
* **Why no single metric can choose** the parameters — each of coverage,
  accuracy and overall correctness is maximised at a degenerate corner
  (@sec-op-metrics).
* **An approach cone tested against the rectangle** (@sec-cone), and what
  `endpoint` could borrow from `trend` for arrivals (@sec-hybrid).

**4.3**

* The height sweep runs to **40,000 ft** rather than stopping at 20,000. At
  20,000 ft six points of departure coverage still separated the curve from
  the unrestricted case, so the old grid cut it off mid-rise and left the
  reader to guess where it flattened.

**4.2**

* A breakdown **by class of ground-truth aerodrome** (@sec-per-type) — the one
  new computation. It surfaces the sharpest finding in the study: on small
  aerodromes and heliports the method is wrong **every single time it answers**,
  because the detection reference contains large and medium aerodromes only.
* Overall correctness is reported in the **radius and penalty sweeps**, not
  only in the results table.
* The per-aerodrome cross-check extends to the **100 busiest**, and now covers
  **arrivals as well as departures**.

**4.1**

* Results given for **both roles on all four measures**: arrival overall
  correctness and both in-area figures were computed but not printed.
* The sweep plotted **on accuracy as well as coverage**, for both roles.
  Accuracy peaks near the ground and falls as height is admitted, so the
  coverage view alone overstates the case for a high threshold
  (@sec-accuracy-sweep).
* The cross-check names the **mode and parameters** it was produced at, and
  separates the out-of-area bucket from the movement-count ratio it was
  distorting.
* Figure captions generated from the data rather than written by hand.
:::

## What the pipeline does before any method runs {#sec-preprocessing}

The population a method is judged on is fixed long before the method executes.
Each stage below discards something, and what survives is what detection sees.

**1. Ingest** (`ingestion/osn_statevectors.py`). Hourly partitions are read
from the OpenSky archive and immediately filtered to the European bounding box
`(-25.87, 26.75, 49.66, 70.26)` and decimated to one sample every 5 s. Both
filters are applied *before* anything is written. For the three days here that
turned roughly 60 GB of raw feed into **`r format(142630930, big.mark=",")`
state vectors**.

*What is discarded:* everything outside Europe, and four samples in five. The
5 s decimation is the reason a track's "first sample" can be several seconds
into the take-off roll rather than at the stand.

**2. Track construction** (`pipeline/tracks.py`). State vectors are grouped
into tracks by `SHA2(icao24 ‖ callsign)` and split on a gap of over 30 minutes,
or over 15 minutes below 5,000 ft. Each sample is also indexed into H3 cells at
resolutions 7 and 12.

*What is discarded:* nothing, but the grouping decides what counts as one
flight. Version 2 measured the consequences — roughly one matched flight in ten
is split across more than one track, and a few percent of tracks cover more
than one flight.

**3. Aerodrome zones** (`reference/h3_airport_zones.py`). Each aerodrome is
surrounded by concentric rings of H3 res-7 cells: 5 NM steps out to 40 NM, then
10 NM steps to 110 NM. Every cell carries the ring band it belongs to, so a
consumer chooses its own radius at read time. **`r format(33751619, big.mark=",")`
(aerodrome, cell) rows** for the 1,353 large and medium aerodromes in the area.

*Why it reaches 110 NM:* the same table serves ASMA ring crossings at 40 and
100 NM. Detection uses a fraction of it.

**4. Endpoint candidates** (`pipeline/flights.py`). For each track, the first
and last sample are matched against the zone table on `h3_res_7`, and the exact
great-circle distance to each candidate aerodrome is computed. The result is
cached as `opdi_endpoint_candidates` — **`r format(7491655, big.mark=",")` rows**.

This is the step that makes the sweeps in this paper affordable. Deriving
candidates from state vectors is expensive; once cached, changing a threshold
is a filter and a re-rank over a few million rows rather than a pass over 142
million. Reducing to the two endpoint samples *before* the join is also what
makes the full 110 NM reach cheap — about a million rows to match, not the
whole month.

::: {.callout-note}
## H3 as an index, distance as the comparison

The H3 join answers "which aerodromes are plausibly near this sample" in one
equi-join, without a spatial cross product. It does not answer "how near",
because a cell inside a 40 NM ring can still be more than 40 NM from the
aerodrome. So every candidate is then measured exactly, and the distance — not
the cell — decides.
:::

## How the metrics work {#sec-metrics}

This study turns on the difference between measures that are easy to conflate,
so each is defined before any number appears.

Let $G$ be the ground-truth flights and $A \subseteq G$ those the method
answered.

| Measure | Definition | The question it answers |
|---|---|---|
| **Coverage** | $\lvert A \rvert / \lvert G \rvert$ | How often does it say *something*? |
| **Accuracy** | $\lvert \text{correct} \rvert / \lvert A \rvert$ | When it speaks, how often is it right? |
| **Overall correctness** | $\lvert \text{correct} \rvert / \lvert G \rvert$ | Of everything, how much did it get right? |

A flight whose trajectory was never seen counts against coverage. That is
deliberate: ground truth is the denominator throughout, so "we had no data" is
a miss rather than an excluded row.

**Accuracy alone is close to meaningless here.** It says nothing about how
often a method speaks, so a method that answers rarely and cautiously can post
a better accuracy than one doing far more useful work. Consider two methods on
the same flights:

```{r metric-example}
ex <- data.frame(
  Method = c("Cautious", "Eager"),
  Coverage = c("40.00%", "95.00%"),
  Accuracy = c("99.00%", "88.00%"),
  `Overall correctness` = c("39.60%", "83.60%"),
  check.names = FALSE
)
kable(ex, align = "lrrr",
      caption = "Accuracy favours the cautious method; overall correctness shows the eager one labels more than twice as many flights correctly.")
```

Neither is simply better — they suit different uses. A flight list that
downstream users treat as authoritative may well prefer the cautious one. The
point is that **no single number decides it**, which is why all three are
reported throughout.

### Precision and recall, for out-of-area

Some flights have an origin or destination outside the observed area entirely.
"Out of area" is therefore a legitimate *answer*, not a failure — and how well a
method identifies those cases is measured with the standard pair:

| Measure | Definition | The question it answers |
|---|---|---|
| **Precision** | of flights *labelled* out-of-area, the fraction that were | When it says out-of-area, is it right? |
| **Recall** | of flights that *were* out-of-area, the fraction it labelled | Of the cases out there, how many did it catch? |

Precision is accuracy computed on one label instead of all of them. Recall has
no counterpart in the table above, and it is the one that matters here: **high
precision with low recall means a method is right when it fires but rarely
fires**.

In the standard confusion-matrix vocabulary, with "out-of-area" as the positive
class: precision is $TP/(TP+FP)$ and recall is $TP/(TP+FN)$. Coverage has no
confusion-matrix equivalent, because it measures whether the method produced
*any* label rather than which one.

### In-area coverage and accuracy

Because roughly one flight in twelve is genuinely out of area, a method could
raise its apparent accuracy simply by labelling out-of-area liberally. So the
same two measures are also reported **restricted to flights whose truth is a
real aerodrome** — detection skill with the free wins removed. When headline and
in-area figures diverge, the gap is exactly the contribution of out-of-area
labelling.

## The three modes {#sec-modes}

All three read the same cached candidates. They differ in what evidence they
consult and, more importantly, in when they are willing to stay silent.

### `trend` — the current production algorithm

*Evidence:* every state vector below **FL40** that falls inside an aerodrome's
detection zone, currently 30 NM.

*Rule:* for each (track, aerodrome) pair, smooth altitude over a 5-sample
window, count how often the smoothed value rises versus falls, and require a
margin of 4 either way. A positive margin is a take-off, negative a landing.

*Abstains when:* the margin is within ±4 — classified `ambiguous` and **the
flight is dropped entirely**.

*Failure mode:* a flight seen only above FL40 near its origin, or with a noisy
altitude trace, silently loses its aerodrome. The FL40 gate is the binding
constraint, not the 30 NM radius.

### `nearest` — the *traffic* library's rule

*Evidence:* the track's first and last sample. Nothing else — no altitude, no
ground state, no trend.

*Rule:* name the nearest aerodrome by effective distance.

*Abstains when:* never, except when no candidate exists at all.

*Failure mode:* it answers just as confidently for a track that begins in the
cruise as for one that begins at a stand. Its coverage is the highest of any
method and its accuracy the lowest, and both facts have the same cause.

### `endpoint` — nearest, with abstention

*Evidence:* the same two samples, plus their height above **field elevation**
and their `on_ground` flag.

*Rule:* name the nearest aerodrome, but only if the endpoint is within
`radius_nm` **and** either below `height_ft` above that aerodrome's elevation or
flagged on the ground. Otherwise, if the endpoint lies near the boundary of the
observed area, answer out-of-area. Otherwise answer nothing.

*Why height above field elevation:* a fixed altitude cut-off means nothing at
an aerodrome sitting at 5,000 ft. Elevation comes from OurAirports and is
missing for about 3% of large and medium aerodromes; where it is missing the
height test is skipped and only the ground flag can satisfy the rule, rather
than silently comparing against sea level.

*Failure mode:* it inherits `nearest`'s naming and therefore its confusions
between neighbouring aerodromes; the abstention only decides *whether* to
answer, never *what* to answer.

## Results

```{r mode-table}
# Split by role rather than widening: departures and arrivals each get the full
# set -- coverage, accuracy, overall correctness and the in-area restriction --
# so no measure is reported for one role and silently omitted for the other.
tab <- do.call(rbind, lapply(c(adep = "departures", ades = "arrivals"), function(rl) {
  pre <- if (rl == "departures") "adep" else "ades"
  do.call(rbind, lapply(c("trend", "nearest", "endpoint"), function(md) data.frame(
    Movements = rl, Mode = md,
    Coverage = pct(m(md, paste0(pre, "_coverage"))),
    Accuracy = pct(m(md, paste0(pre, "_accuracy"))),
    Overall  = pct(m(md, paste0(pre, "_overall"))),
    `In-area cov` = pct(m(md, paste0(pre, "_inarea_coverage"))),
    `In-area acc` = pct(m(md, paste0(pre, "_inarea_accuracy"))),
    check.names = FALSE)))
}))
kable(tab, row.names = FALSE, align = "llrrrrr",
      caption = sprintf("The three modes on %s ground-truth flights, 2025-06-05 to 07. Overall correctness is coverage x accuracy. Produced at %s.",
                        format(N_GT, big.mark = ","), OP))
```

`endpoint` sits between the two by construction, and that is the whole design:
it keeps `nearest`'s naming rule and adds `trend`'s willingness to stay silent.
It answers `r pp(m("endpoint","adep_coverage") - m("trend","adep_coverage"))`
more departures than production for
`r pp(m("endpoint","adep_accuracy") - m("trend","adep_accuracy"))` of accuracy.

Whether that trade is worth taking is a policy question rather than a technical
one. A flight list treated as authoritative may prefer production's caution;
an analysis needing volume may not.

Arrivals remain weaker than departures on every measure, as in all previous
versions. A departure's first sample is at or near a stand, with the aircraft
stationary shortly before. An arrival's last sample is more often a loss of
reception during descent than a landing, and can be anywhere on the approach.

## What a distance cap costs {#sec-radius}

The sweep varies the detection radius and the abstention height independently,
`r nrow(grid)` combinations in all, over the cached candidates. Height runs from
on-ground-only to
`r format(max(grid$height_ft[grid$height_ft < UNRESTRICTED]), big.mark=",")` ft
above field elevation — past typical cruise, so the curve is followed to where
it flattens — plus an unrestricted case as the limit.

Both measures are plotted for both roles, because the two axes do not merely
differ in strength — they pull in opposite directions, and a coverage-only view
hides that entirely.

```{r sweep-setup}
#| include: false
d <- grid[grid$height_ft < UNRESTRICTED, ]
d$radius <- factor(d$radius_nm, levels = sort(unique(d$radius_nm)),
                   labels = paste0(sort(unique(d$radius_nm)), " NM"))

# Long form, so departures and arrivals become facets of one figure and can be
# read against each other on a shared pair of axes.
sweep_plot <- function(measure, ylab) {
  x <- rbind(
    data.frame(d[c("height_ft", "radius")], Movements = "Departures",
               value = d[[paste0("adep_", measure)]]),
    data.frame(d[c("height_ft", "radius")], Movements = "Arrivals",
               value = d[[paste0("ades_", measure)]]))
  x$Movements <- factor(x$Movements, levels = c("Departures", "Arrivals"))
  ggplot(x, aes(height_ft, value, colour = radius, group = radius)) +
    geom_line(linewidth = 0.6) + geom_point(size = 1.1) +
    facet_wrap(~ Movements) +
    scale_x_continuous("Abstention height above field elevation (ft)",
                       breaks = c(0, 20000, 40000), labels = scales::comma) +
    scale_y_continuous(ylab, labels = scales::percent) +
    scale_colour_viridis_d("Detection radius", end = 0.9) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank())
}

# Captions are built from the data for the same reason the tables are: an
# earlier revision of this page carried a hand-written caption that the numbers
# had already contradicted.
sat <- function(pre, r) g(110, OP_HEIGHT, paste0(pre, "_coverage")) -
                        g(r,   OP_HEIGHT, paste0(pre, "_coverage"))
peak <- function(pre) {
  v <- d[d$radius_nm == OP_RADIUS, ]
  v <- v[order(v$height_ft), ]
  h <- v$height_ft[which.max(v[[paste0(pre, "_accuracy")]])]
  c(height = h, acc = max(v[[paste0(pre, "_accuracy")]]),
    drop = max(v[[paste0(pre, "_accuracy")]]) - g(OP_RADIUS, OP_HEIGHT, paste0(pre, "_accuracy")))
}
pk_dep <- peak("adep"); pk_arr <- peak("ades")
spread <- function(pre) g(5, OP_HEIGHT, paste0(pre, "_accuracy")) -
                        g(110, OP_HEIGHT, paste0(pre, "_accuracy"))

cap_cov <- sprintf(
  "Coverage against abstention height, one line per detection radius. The shape of the curve belongs to the height axis in both panels; the radius only sets how much of it is reachable, and arrivals need more of it than departures. At %s ft a 20 NM radius still leaves %s of departure coverage and %s of arrival coverage on the table against the full 110 NM reach, while at %g NM both are within half a point of it.",
  format(OP_HEIGHT, big.mark = ",", trim = TRUE),
  pp(sat("adep", 20)), pp(sat("ades", 20)), OP_RADIUS)

cap_acc <- sprintf(
  "Accuracy against abstention height, same axes. This is the figure the coverage plot hides. Accuracy peaks close to the ground -- at %g NM, departures at %s ft and arrivals at %s ft -- and every foot admitted above that is bought with correctness: %s for departures by %s ft, %s for arrivals. The radius, which barely touches coverage above 40 NM, spreads accuracy at %s ft by %s for departures and %s for arrivals.",
  OP_RADIUS,
  format(pk_dep["height"], big.mark = ",", trim = TRUE),
  format(pk_arr["height"], big.mark = ",", trim = TRUE),
  pp(-pk_dep["drop"]), format(OP_HEIGHT, big.mark = ",", trim = TRUE),
  pp(-pk_arr["drop"]),
  format(OP_HEIGHT, big.mark = ",", trim = TRUE),
  pp(spread("adep")), pp(spread("ades")))
```

```{r sweep-plot-cov, fig.cap=cap_cov, fig.width=8, fig.height=4}
sweep_plot("coverage", "Coverage")
```

```{r sweep-plot-acc, fig.cap=cap_acc, fig.width=8, fig.height=4}
sweep_plot("accuracy", "Accuracy")
```

```{r radius-cost}
r <- grid[grid$height_ft == OP_HEIGHT, ]
r <- r[order(r$radius_nm), ]
kable(data.frame(
  `Max radius` = paste0(r$radius_nm, " NM"),
  `ADEP cov` = pct(r$adep_coverage), `ADEP acc` = pct(r$adep_accuracy),
  `ADEP ovr` = pct(r$adep_overall),
  `ADES cov` = pct(r$ades_coverage), `ADES acc` = pct(r$ades_accuracy),
  `ADES ovr` = pct(r$ades_overall),
  check.names = FALSE), align = "lrrrrrr",
  caption = sprintf("Detection radius at a fixed %s ft abstention height. *ovr* is overall correctness, coverage x accuracy. Coverage saturates by about 30 NM for departures and 40 NM for arrivals; beyond that a wider radius trades accuracy for almost no coverage, and overall correctness flattens rather than turning -- the radius has no interior optimum on this axis, unlike the height (@sec-accuracy-sweep).",
                    format(OP_HEIGHT, big.mark = ",", trim = TRUE)))
```

**A cap costs almost nothing.** Going from 40 NM to the full 110 NM reach
changes departure coverage by
`r pp(g(110,15000,"adep_coverage") - g(40,15000,"adep_coverage"))` and accuracy
by `r pp(g(110,15000,"adep_accuracy") - g(40,15000,"adep_accuracy"))`. The
useful range is 20–40 NM; below that coverage falls away quickly, above it
nothing happens.

Every radius at or above 20 NM traces effectively the same coverage curve: at
5,000 ft they span `r pct(g(20,5000,"adep_coverage"))` to
`r pct(g(110,5000,"adep_coverage"))`, about a point across a five-fold change.
Only the 5 NM and 10 NM lines fall away, and they fall further as the height
threshold rises, because a tight radius costs more once more of the sky is
admitted.

**Height is the parameter that matters.** At `r OP_RADIUS` NM, departure
coverage runs from `r pct(g(OP_RADIUS,0,"adep_coverage"))` with the
on-ground-only rule, through `r pct(g(OP_RADIUS,2000,"adep_coverage"))` at
2,000 ft, to `r pct(g(OP_RADIUS,OP_HEIGHT,"adep_coverage"))` at
`r format(OP_HEIGHT, big.mark=",")` ft. That is over fifty points from one axis
against a fraction of a point from the other.

Arrivals behave the same way with less room: from
`r pct(g(OP_RADIUS,0,"ades_coverage"))` to
`r pct(g(OP_RADIUS,OP_HEIGHT,"ades_coverage"))` over the same range. They also
saturate later in radius — at `r format(OP_HEIGHT, big.mark=",")` ft a 20 NM cap
costs departures `r pp(sat("adep", 20))` of coverage but arrivals
`r pp(sat("ades", 20))`, because an arrival's last recorded sample is further
from the aerodrome than a departure's first.

Two endpoints of the sweep serve as sanity checks, and both behave as they
should: at height 0 the rule reduces to "endpoint on the ground" and gives
`r pct(g(OP_RADIUS,0,"adep_coverage"))`, close to what a surface-samples-only
method achieved in earlier versions; with height unrestricted and a wide radius
it reaches `r pct(g(110,UNRESTRICTED,"adep_coverage"))`, close to `nearest`'s
`r pct(m("nearest","adep_coverage"))`, which is what it should converge to.

**The curve does flatten, and the grid now reaches the flat part.** At
`r OP_RADIUS` NM the last measured height is within
`r pp(g(OP_RADIUS,UNRESTRICTED,"adep_coverage") - g(OP_RADIUS,max(grid$height_ft[grid$height_ft < UNRESTRICTED]),"adep_coverage"))`
of the unrestricted case for departures and
`r pp(g(OP_RADIUS,UNRESTRICTED,"ades_coverage") - g(OP_RADIUS,max(grid$height_ft[grid$height_ft < UNRESTRICTED]),"ades_coverage"))`
for arrivals, so a threshold at cruise level is indistinguishable from no
threshold at all. The steepest remaining step is between 30,000 and 35,000 ft —
`r pp(g(OP_RADIUS,35000,"ades_coverage") - g(OP_RADIUS,30000,"ades_coverage"))`
of arrival coverage in one 5,000 ft band, which is where first and last contact
at cruise gets admitted.

### What the coverage view hides {#sec-accuracy-sweep}

Read on accuracy, the same grid tells a different story, and the two together
are what actually determines an operating point.

**Accuracy peaks near the ground, then declines.** At `r OP_RADIUS` NM,
departure accuracy is highest at `r format(pk_dep["height"], big.mark=",")` ft
(`r pct(pk_dep["acc"])`) and arrival accuracy at
`r format(pk_arr["height"], big.mark=",")` ft (`r pct(pk_arr["acc"])`). By
`r format(OP_HEIGHT, big.mark=",")` ft they have given up
`r pp(-pk_dep["drop"])` and `r pp(-pk_arr["drop"])` respectively. Every foot of
extra height admits endpoints that are further into the climb or descent, and
those are exactly the endpoints most likely to sit nearer a neighbouring
aerodrome than the real one.

**Arrivals pay roughly three times as much for the same height.** That ratio is
the single clearest statement of why this study has never recommended changing
the arrivals algorithm: the same parameter that is nearly free for departures
is expensive for arrivals, so one operating point cannot serve both well.

**The radius matters for accuracy even where it does not for coverage.** Above
40 NM a wider radius buys essentially no coverage, but it keeps costing
accuracy: at `r format(OP_HEIGHT, big.mark=",")` ft, going from 5 NM to 110 NM
costs `r pp(-spread("adep"))` of departure accuracy and
`r pp(-spread("ades"))` of arrival accuracy. This is the real argument for a
cap. Version 4 justified one on the grounds that it costs almost no coverage;
the accuracy view shows it is mildly *positive* rather than merely harmless.

```{r overall-peak}
#| include: false
fin  <- grid[grid$height_ft < UNRESTRICTED, ]
op_r <- fin[fin$radius_nm == OP_RADIUS, ]
op_r <- op_r[order(op_r$height_ft), ]
pk   <- function(f) op_r$height_ft[which.max(op_r[[f]])]
HMAX <- max(fin$height_ft)
```

**Overall correctness is what reconciles them, and it has a maximum.** Coverage
rises steeply with height while accuracy falls gently, so their product rises
for a long way: at `r OP_RADIUS` NM departure overall correctness runs from
`r pct(g(OP_RADIUS,2000,"adep_overall"))` at 2,000 ft to
`r pct(g(OP_RADIUS,OP_HEIGHT,"adep_overall"))` at
`r format(OP_HEIGHT, big.mark=",")` ft. But it does not rise forever. It peaks
at **`r format(pk("adep_overall"), big.mark=",")` ft**
(`r pct(max(op_r$adep_overall))`) and declines above that, as the accuracy loss
finally outruns the coverage gain.

::: {.callout-warning}
## Version 4.2 got this wrong, and the extended grid is why

Version 4.2 stated that the product "keeps rising for departures across the
whole range". That was true of the range it measured — the sweep stopped at
20,000 ft, below the turning point — and false of the parameter. Extending the
grid to `r format(HMAX, big.mark=",")` ft was what exposed it. The claim is
withdrawn.
:::

The maximum is shallow, which is the practical point:
`r pct(max(op_r$adep_overall))` at its peak against
`r pct(g(OP_RADIUS,OP_HEIGHT,"adep_overall"))` at the recommended
`r format(OP_HEIGHT, big.mark=",")` ft, a difference of
`r pp(max(op_r$adep_overall) - g(OP_RADIUS,OP_HEIGHT,"adep_overall"))`. Buying
it costs `r pp(g(OP_RADIUS,pk("adep_overall"),"adep_accuracy") - g(OP_RADIUS,OP_HEIGHT,"adep_accuracy"))`
of accuracy, so the recommendation does not move.

Arrivals never turn over inside the grid: overall correctness is still climbing
at `r format(HMAX, big.mark=",")` ft
(`r pct(g(OP_RADIUS,HMAX,"ades_overall"))` against
`r pct(g(OP_RADIUS,OP_HEIGHT,"ades_overall"))` at
`r format(OP_HEIGHT, big.mark=",")` ft). Arrival coverage is so far from its
ceiling that admitting more of it still pays, which is the same asymmetry once
more.

A user who cares about a trustworthy label rather than a complete one should
ignore all of this and read the accuracy panel, where the optimum sits near the
ground.

## Preferring aerodromes with scheduled service {#sec-penalty}

Version 3 found that the most frequent departure errors formed one pattern: a
scheduled civil airport losing to an adjacent military airfield — Izmir to
Çiğli Airbase, Naples to Grazzanise, Cagliari to Decimomannu, Catania to
Sigonella, Gdańsk to Cewice, all between 7 and 25 NM apart.

The mechanism is straightforward. With 5 s sampling a departing aircraft's
first recorded fix is already into the climb and displaced along the departure
track. Where a military field lies between the airport and that track, it is
genuinely the nearest aerodrome — the rule is not malfunctioning, it is
answering the question it was asked.

Aviation knowledge resolves it without new data: a commercial flight almost
certainly used the aerodrome with scheduled service. Candidates are therefore
ranked on **effective distance** — true distance plus a fixed penalty for
aerodromes without scheduled service — so a military field must be *clearly*
nearest rather than merely nearest.

```{r penalty-table}
p <- pen[order(pen$penalty_nm), ]
kable(data.frame(
  Penalty = paste0(p$penalty_nm, " NM"),
  `ADEP cov` = pct(p$adep_coverage), `ADEP acc` = pct(p$adep_accuracy),
  `ADEP ovr` = pct(p$adep_overall),
  `ADES cov` = pct(p$ades_coverage), `ADES acc` = pct(p$ades_accuracy),
  `ADES ovr` = pct(p$ades_overall),
  check.names = FALSE), align = "lrrrrrr",
  caption = sprintf("Scheduled-service penalty at %g NM and %s ft. *ovr* is overall correctness. A penalty of 0 reproduces the unbiased ranking exactly and is the control.",
                    OP_RADIUS, format(OP_HEIGHT, big.mark = ",", trim = TRUE)))
```

The gain is
**`r pp(p$adep_accuracy[p$penalty_nm==10] - p$adep_accuracy[p$penalty_nm==0])`
of departure accuracy** and
**`r pp(p$ades_accuracy[p$penalty_nm==10] - p$ades_accuracy[p$penalty_nm==0])`
of arrival accuracy**, for a coverage change of
`r pp(p$adep_coverage[p$penalty_nm==10] - p$adep_coverage[p$penalty_nm==0])`.
Overall correctness moves with it —
`r pp(p$adep_overall[p$penalty_nm==10] - p$adep_overall[p$penalty_nm==0])` and
`r pp(p$ades_overall[p$penalty_nm==10] - p$ades_overall[p$penalty_nm==0])` —
which is what makes this the one parameter in the study with no trade attached.

Two features make this trustworthy rather than a lucky constant. The control at
penalty 0 reproduces the unbiased ranking, so the effect is attributable. And
the curve **peaks and turns** — accuracy is best around 10 NM and declines
above it. A penalty that merely favoured larger aerodromes would improve
monotonically; one that corrects a specific bias should overshoot, and this
does.

::: {.callout-warning}
## It is not free — the cost is a mirror error

Version 3 measured what this preference breaks: genuine movements at
non-scheduled aerodromes get pulled to the nearby hub — Biggin Hill to London
City, Northolt to Heathrow. In that analysis the mirror error became the
**largest single class of remaining departure error**, at roughly 59% of them.

The net is favourable, and the accuracy figures above include the cost. But a
flight list built this way will systematically under-report movements at
non-scheduled aerodromes, and anyone analysing general aviation or military
traffic should know that before using it.
:::

## Out-of-area {#sec-ooa}

A flight from Dubai to Amsterdam has a departure aerodrome that no European
ADS-B feed can name: the aircraft entered the observed area already airborne.
Scoring that as a detection failure understates every method. Treating "out of
area" as an answer is an accounting correction, not a detection improvement.

### How it is derived

**On the ground-truth side**, an aerodrome is relabelled `OOA` when its
coordinates fall outside the ingestion bounding box, **or** when its ICAO code
does not resolve in OurAirports. The unresolvable codes are overwhelmingly
non-European, and a code that cannot be located is one no trajectory endpoint
could have been matched to. This is what puts prediction and truth in the same
alphabet.

**On the prediction side**, an endpoint within 30 NM of the bounding box edge
is labelled out-of-area. The longitude margin is scaled by $1/\cos(\text{lat})$:
a degree of longitude shrinks toward the pole, so an unscaled margin would be
roughly twice as strict in northern Norway as in the Canaries.

**Precedence is aerodrome first, border second.** The opposite order looks
equivalent and is not — version 3 found Ponta Delgada, which sits about 8 NM
inside the western edge, having its departures labelled out-of-area with the
aircraft still on the runway. Being near the boundary means "out of area" only
when there is nothing better to say.

**The border test carries no height condition.** A flight leaving the observed
area is at cruise level by definition, so any height cut-off would reject
exactly the endpoints the branch exists to catch.

### What it does to the numbers

```{r ooa-table}
o <- do.call(rbind, lapply(c("trend","nearest","endpoint"), function(md) data.frame(
  Mode = md,
  `truly OOA` = pct(m(md, "adep_gt_ooa_share")),
  `labelled OOA` = pct(m(md, "adep_pred_ooa_share")),
  `OOA recall` = pct(m(md, "adep_ooa_recall")),
  `OOA precision` = pct(m(md, "adep_ooa_precision")),
  `headline acc` = pct(m(md, "adep_accuracy")),
  `in-area acc` = pct(m(md, "adep_inarea_accuracy")),
  check.names = FALSE)))
kable(o, align = "lrrrrrr",
      caption = "Out-of-area for departures. Only `endpoint` emits the label; the others never do, so their recall is zero by construction.")
```

`truly OOA` is a property of the sample, not of the method, so it must be
identical across all three rows — that it is confirms the decomposition is
being computed correctly rather than being a per-method artefact.

The gap between headline and in-area accuracy shows how much out-of-area
labelling contributes. It is small here, which is the honest reading: the
mechanism is right but the geometric test catches only part of what is there.
Version 3 measured this directly at **92–94% precision against 10–15% recall** —
when the rule fires it is almost always correct, but it rarely fires. An
inbound flight's first recorded sample appears wherever a receiver happens to
catch it, which is usually well inside the boundary rather than at it.

**The concept is right and the instrument is wrong.** A phase-based test would
fit the physics better: a track whose first sample is already in the cruise,
neither near an aerodrome nor descending toward one, has an origin outside the
observed area regardless of where it sits. That is the clearest single
improvement available and is not implemented here.

## Cross-check against the reference {#sec-crosscheck}

Flight-level accuracy can hide errors that cancel: a departure misattributed
from A to B leaves both counts wrong and the total right. Movement counts per
aerodrome are what an operational user reads, so they are checked separately.

```{r per-airport}
if (!is.null(apt)) {
  ia <- apt[!is.na(apt$n_truth) & apt$n_truth > 0, ]
  s <- do.call(rbind, lapply(c("departures","arrivals"), function(rl) {
    d <- ia[ia$role == rl, ]
    a <- d[d$airport != "OOA", ]         # real aerodromes only
    b <- a[a$n_truth >= 50, ]
    data.frame(Movements = rl,
      Aerodromes = nrow(a),
      `Count ratio` = sprintf("%.3f", sum(d$n_assigned) / sum(d$n_truth)),
      `Excl. OOA` = sprintf("%.3f", sum(a$n_assigned) / sum(a$n_truth)),
      `Median recall` = pct(median(a$recall, na.rm = TRUE)),
      `Median recall, 50+ movements` = pct(median(b$recall, na.rm = TRUE)),
      check.names = FALSE)
  }))
  kable(s, row.names = FALSE, align = "lrrrrr",
        caption = sprintf("Movement counts against Network Manager, from the `endpoint` flight list at %s. A count ratio below 1 is an undercount. The second ratio drops the out-of-area bucket, whose own recall is poor (@sec-ooa) and which otherwise carries the shortfall of every unlabelled flight.", OP))
}
```

### By class of aerodrome {#sec-per-type}

Every headline figure averages over populations that behave nothing like each
other. The detection reference holds **large and medium aerodromes only**, so
flights whose true aerodrome is a small field or a heliport cannot be named
correctly at any radius or height. Grouping ground truth by its OurAirports
class separates that from detection skill.

```{r per-type}
if (!is.null(typ)) {
  t <- typ[order(typ$role, -typ$n_truth), ]
  kable(data.frame(
    Movements = t$role,
    Class = gsub("_", " ", t$apt_type),
    Aerodromes = format(t$n_aerodromes, big.mark = ","),
    Flights = format(t$n_truth, big.mark = ","),
    Coverage = pct(t$coverage),
    Accuracy = pct(t$accuracy),
    Overall = pct(t$overall),
    check.names = FALSE), row.names = FALSE, align = "llrrrrr",
    caption = sprintf("Coverage and accuracy by the class of the *ground-truth* aerodrome, from the `endpoint` flight list at %s. Classes are OurAirports types. The detection reference contains large and medium aerodromes only.", OP))
}
```

```{r per-type-vals}
#| include: false
tv <- function(rl, cl, f) {
  v <- typ[[f]][typ$role == rl & typ$apt_type == cl]
  if (length(v) == 0) NA_real_ else v[1]
}
small_n <- sum(typ$n_truth[typ$apt_type %in% c("small_airport", "heliport")])
small_ans <- sum(typ$n_answered[typ$apt_type %in% c("small_airport", "heliport")])
```

**Small aerodromes and heliports are answered confidently and wrongly, every
time.** Accuracy on both classes is exactly zero, on
`r pct(tv("departures","small_airport","coverage"))` coverage for small
airports and `r pct(tv("departures","heliport","coverage"))` for heliports.
That is not a near-miss rate: the reference cannot name these aerodromes, so
when the endpoint happens to fall within `r OP_RADIUS` NM of some large or
medium field below `r format(OP_HEIGHT, big.mark=",")` ft, the rule names that
one instead. Across both roles, `r format(small_ans, big.mark=",")` of
`r format(small_n, big.mark=",")` such flights get an answer and none of them
are right.

The absolute cost is small — under half a point of the sample — but the shape
of the error matters more than its size. **These are not abstentions; they are
false statements**, and nothing in the output distinguishes them from the
`r pct(tv("departures","large_airport","accuracy"))` of correct large-airport
departures. Two fixes are available and neither is implemented here: extend the
reference to small aerodromes so they can be named, or make abstention aware of
what the reference contains, so an endpoint nearest to an aerodrome the
reference omits produces silence rather than its neighbour.

**Between the classes the reference does hold, the difference is real but
ordinary.** Departures at large airports run
`r pct(tv("departures","large_airport","coverage"))` coverage at
`r pct(tv("departures","large_airport","accuracy"))` accuracy, against
`r pct(tv("departures","medium_airport","coverage"))` and
`r pct(tv("departures","medium_airport","accuracy"))` at medium ones. Since
large airports are `r pct(tv("departures","large_airport","n_truth") / N_GT)`
of the sample, the headline figures are very close to the large-airport figures
and should not be read as applying uniformly.

```{r per-airport-top}
if (!is.null(apt)) {
  # OOA is excluded: it is not an aerodrome, and its counts are reported in
  # their own right in the out-of-area section.
  top_n <- function(rl, n = 100) {
    d <- apt[apt$role == rl & apt$airport != "OOA", ]
    head(d[order(-d$n_truth), ], n)
  }
  top_table <- function(rl, label) {
    top <- top_n(rl)
    kable(data.frame(
      Aerodrome = top$airport,
      `NM movements` = format(top$n_truth, big.mark = ","),
      `OPDI assigned` = format(top$n_assigned, big.mark = ","),
      `Count ratio` = sprintf("%.2f", top$count_ratio),
      Recall = pct(top$recall),
      check.names = FALSE), row.names = FALSE, align = "lrrrr",
      caption = sprintf("The 100 busiest aerodromes by Network Manager %s, against the `endpoint` flight list at %s. Out-of-area is excluded -- it is not an aerodrome and is reported separately in @sec-ooa. *Count ratio* is OPDI movements divided by Network Manager movements, so it can sit near 1 even when the individual flights are wrong; *recall* is the share of Network Manager %s OPDI put at this aerodrome, and is the stricter measure.",
                        label, OP, label))
  }
}
```

```{r per-airport-top-dep}
if (!is.null(apt)) top_table("departures", "departures")
```

```{r per-airport-top-arr}
if (!is.null(apt)) top_table("arrivals", "arrivals")
```

The error is an **undercount** rather than an overcount, which is the
correctable direction: a systematic shortfall can be adjusted for, an inflated
count cannot. Roughly six points of it are the out-of-area bucket rather than
per-aerodrome detection failure — flights the method never labelled land
nowhere, so they show up as missing movements at whichever aerodrome the truth
names.

```{r recall-scale}
#| include: false
depall <- apt[apt$role == "departures" & apt$airport != "OOA", ]
depall <- depall[order(-depall$n_truth), ]
r100 <- head(depall$recall, 100)
share100 <- sum(head(depall$n_truth, 100)) / sum(depall$n_truth)
```

Recall also scales sharply with aerodrome size. Across the hundred busiest —
`r pct(share100)` of all departures in the sample — the median is
`r pct(median(r100))` and the quartiles are `r pct(quantile(r100, 0.25))` and
`r pct(quantile(r100, 0.75))`, against a median of
`r pct(median(depall$recall, na.rm = TRUE))` over all `r nrow(depall)`
aerodromes with departures. The flight list is therefore substantially more
complete at large aerodromes than small ones, which matters before comparing
across aerodrome classes — and @sec-per-type says why.

### Busy aerodromes that fail outright {#sec-outliers}

Extending the listing from thirty rows to a hundred exposes something the
shorter view could not: a handful of genuinely busy aerodromes where the method
does not merely lose a few movements but fails almost completely.

```{r outliers}
#| include: false
top100 <- do.call(rbind, lapply(c("departures","arrivals"), function(rl) {
  d <- apt[apt$role == rl & apt$airport != "OOA", ]
  head(d[order(-d$n_truth), ], 100)
}))
top100$answered <- top100$n_predicted_here / top100$n_truth
bad      <- top100[top100$recall < 0.5, ]
unseen   <- bad[bad$answered < 0.25, ]
misnamed <- bad[bad$answered >= 0.25, ]
# One accessor for the worked examples, so no figure is transcribed by hand.
w <- function(ap, rl, f) apt[[f]][apt$role == rl & apt$airport == ap]
```

`r nrow(bad)` of the two hundred rows above sit below 50% recall —
`r sum(bad$role == "departures")` departure aerodromes and
`r sum(bad$role == "arrivals")` arrival ones. They split cleanly into two kinds
of failure:

* **Never seen** (`r nrow(unseen)` rows). Under a quarter of their flights get
  an answer at all: Gran Canaria departures return
  `r format(w("GCLP","departures","n_predicted_here"), big.mark=",")` answers
  against `r format(w("GCLP","departures","n_truth"), big.mark=",")` movements,
  and Hurghada none at all. These cluster where ADS-B receiver coverage is
  thin — the Canaries, North Africa, the Caucasus, Anatolia. No detection rule
  can recover a flight with no low-altitude trajectory.
* **Seen and misnamed** (`r nrow(misnamed)` rows). The trajectory is there and
  the answer is wrong. Bilbao arrivals are answered
  `r pct(w("LEBB","arrivals","n_predicted_here") / w("LEBB","arrivals","n_truth"))`
  of the time and only
  `r format(w("LEBB","arrivals","n_correct"), big.mark=",")` of
  `r format(w("LEBB","arrivals","n_truth"), big.mark=",")` land on Bilbao.

Tenerife North is the cleanest warning about reading count ratios. It is
assigned `r format(w("GCXO","arrivals","n_assigned"), big.mark=",")` arrivals
against `r format(w("GCXO","arrivals","n_truth"), big.mark=",")` true ones — a
count ratio of `r sprintf("%.2f", w("GCXO","arrivals","count_ratio"))`, which
looks like a well-measured aerodrome — while its recall is
`r pct(w("GCXO","arrivals","recall"))`. Almost every movement it is credited
with belongs somewhere else, and the totals cancel the error exactly as
@sec-crosscheck warned they could.

Every aerodrome in both groups is a `large_airport` with scheduled service and
is present in the detection reference, so neither failure is the reference gap
of @sec-per-type. The first group is a data limitation and is not fixable here.
The second is a detection failure at aerodromes large enough to matter, it is
concentrated rather than scattered, and **this study has not diagnosed it** —
it is the most promising lead the cross-check produces.

## Choosing the operating point {#sec-operating-point}

The sections above establish how each parameter behaves. This one picks values,
and argues for them in flights rather than in percentages — because a rate
conceals the size of what is being traded. "−0.84 pp of accuracy" and "792 more
wrong aerodromes in three days" are the same fact, and only one of them can be
weighed against anything.

All counts are out of `r num(N_GT)` ground-truth flights per role, over
2025-06-05 to 07. *Answered* is what the method labelled, *correct* is what it
labelled right, *wrong* is what it labelled wrong, *silent* is what it declined.
The last three columns compare against production (`trend`).

```{r opt-table-fn}
#| include: false
optable <- function(role) {
  p <- prod_cell(role)
  rows <- list(data.frame(
    Parameters = "trend (production)",
    Cov = pct(p$cov), Acc = pct(p$acc), Ovr = pct(p$ovr),
    Answered = num(p$answered), Correct = num(p$correct),
    Wrong = num(p$wrong), Silent = num(p$silent),
    `d correct` = "—", `d wrong` = "—", `Gain:err` = "—",
    check.names = FALSE))
  for (h in SEL_H) for (r in SEL_R) {
    x <- cell(r, h, role); if (is.null(x)) next
    dc <- x$correct - p$correct; dw <- x$wrong - p$wrong
    ratio <- if (dw > 0 && dc > 0) sprintf("%.1f:1", dc / dw) else "—"
    rows[[length(rows) + 1]] <- data.frame(
      Parameters = sprintf("%g NM / %s ft", r, format(h, big.mark = ",")),
      Cov = pct(x$cov), Acc = pct(x$acc), Ovr = pct(x$ovr),
      Answered = num(x$answered), Correct = num(x$correct),
      Wrong = num(x$wrong), Silent = num(x$silent),
      `d correct` = sgn(dc), `d wrong` = sgn(dw), `Gain:err` = ratio,
      check.names = FALSE)
  }
  do.call(rbind, rows)
}
```

::: {.content-visible when-format="pdf"}
```{=latex}
\small
```
:::

```{r opt-table-dep}
kable(optable("adep"), row.names = FALSE, align = "lrrrrrrrrrr",
      caption = sprintf("Departures: the parameter grid in rates and in flights, against production. Penalty fixed at %g NM. *Gain:err* is extra correct labels per extra wrong label, against production.", OP_PENALTY))
```

```{r opt-table-arr}
kable(optable("ades"), row.names = FALSE, align = "lrrrrrrrrrr",
      caption = sprintf("Arrivals: the same grid. Note that most cells produce *fewer* correct labels than production while adding wrong ones. Penalty fixed at %g NM.", OP_PENALTY))
```

::: {.content-visible when-format="pdf"}
```{=latex}
\normalsize
```
:::

### Departures: every cell wins, but the marginal return collapses {#sec-op-dep}

```{r dep-ladder}
#| include: false
LADDER <- list(c(20,5000), c(20,8000), c(20,10000), c(20,15000),
               c(30,15000), c(40,15000), c(40,20000), c(40,25000), c(40,30000))
ladder <- function(role) {
  p <- prod_cell(role); prev <- p; lab <- "production"
  out <- list()
  for (rh in LADDER) {
    x <- cell(rh[1], rh[2], role)
    dc <- x$correct - prev$correct; dw <- x$wrong - prev$wrong
    out[[length(out)+1]] <- data.frame(
      Step = sprintf("%s to %g NM / %s ft", lab, rh[1], format(rh[2], big.mark=",")),
      `d correct` = sgn(dc), `d wrong` = sgn(dw),
      Marginal = if (dc <= 0) "loss" else if (dw <= 0) "free" else sprintf("%.2f:1", dc/dw),
      check.names = FALSE)
    prev <- x; lab <- sprintf("%g NM / %s ft", rh[1], format(rh[2], big.mark=","))
  }
  do.call(rbind, out)
}
dl <- ladder("adep")
```

Every cell in the table beats production. That is not the question. The question
is what each successive loosening *buys*, and the answer is that it collapses.

```{r dep-ladder-table}
kable(dl, row.names = FALSE, align = "lrrr",
      caption = "Departures: each step of loosening, and its marginal return. *Marginal* is extra correct labels per extra wrong label for that step alone, not against production.")
```

Two readings of the same table. Against production, every cell looks excellent —
the worst ratio in the grid is still better than three correct labels per error.
Step by step, the picture inverts: the first move off production is nearly free,
and by the wide-and-high end each additional correct label costs more than one
additional error.

Note where the currently published operating point sits. Going from 30 NM to
40 NM at `r format(OP_HEIGHT, big.mark=",")` ft adds
`r sgn(cell(40,OP_HEIGHT,"adep")$correct - cell(30,OP_HEIGHT,"adep")$correct)`
correct labels and
`r sgn(cell(40,OP_HEIGHT,"adep")$wrong - cell(30,OP_HEIGHT,"adep")$wrong)`
wrong ones. **The published parameters are already past break-even**, which the
percentage view never showed because it is a 0.18 pp change in accuracy.

### Arrivals: nothing in the grid earns its place {#sec-op-arr}

```{r arr-cost}
#| include: false
p_arr <- prod_cell("ades")
arrcost <- do.call(rbind, lapply(list(c(30,15000),c(40,15000),c(40,20000),c(40,25000),c(40,30000)), function(rh) {
  x <- cell(rh[1], rh[2], "ades")
  dc <- x$correct - p_arr$correct; dw <- x$wrong - p_arr$wrong
  data.frame(Cell = sprintf("%g NM / %s ft", rh[1], format(rh[2], big.mark=",")),
             `d correct` = sgn(dc), `d wrong` = sgn(dw),
             `Errors per correct gained` = if (dc > 0) sprintf("%.1f", dw/dc) else "gains none",
             check.names = FALSE)
}))
n_worse <- sum(sapply(SEL_H, function(h) sum(sapply(SEL_R, function(r) {
  x <- cell(r, h, "ades"); !is.null(x) && x$correct < p_arr$correct }))))
```

Arrivals invert the departure result completely.
**`r n_worse` of the `r length(SEL_R)*length(SEL_H)` cells produce fewer correct
arrival labels than production**, while every one of them produces more wrong
ones. The cells that do gain correct labels buy them at a ruinous rate:

```{r arr-cost-table}
kable(arrcost, row.names = FALSE, align = "lrrr",
      caption = "Arrivals: what each cell costs against production. Cells omitted from the ratio column produce fewer correct labels than production and so gain nothing to price.")
```

The best arrival cell in the entire grid, 40 NM / 30,000 ft, gains
`r sgn(cell(40,30000,"ades")$correct - p_arr$correct)` correct labels and adds
`r sgn(cell(40,30000,"ades")$wrong - p_arr$wrong)` wrong ones. There is no
exchange rate under which that is a good trade. **`endpoint` should not replace
`trend` for arrivals at any setting measured here** — and @sec-hybrid asks why,
and what could be done about it.

### Why no single metric can make this choice {#sec-op-metrics}

```{r argmax}
#| include: false
best_of <- function(role, metric) {
  f <- paste0(role, "_", metric)
  x <- grid[which.max(grid[[f]]), ]
  list(lab = sprintf("%g NM / %s", x$radius_nm,
                     if (x$height_ft >= UNRESTRICTED) "unrestricted"
                     else paste0(format(x$height_ft, big.mark=","), " ft")),
       cov = x[[paste0(role,"_coverage")]], acc = x[[paste0(role,"_accuracy")]],
       ovr = x[[paste0(role,"_overall")]])
}
argrow <- function(role, metric, label) {
  b <- best_of(role, metric)
  data.frame(Optimising = label, Parameters = b$lab,
             Coverage = pct(b$cov), Accuracy = pct(b$acc), Overall = pct(b$ovr),
             check.names = FALSE)
}
argtab <- function(role) rbind(
  argrow(role, "coverage", "coverage"),
  argrow(role, "accuracy", "accuracy"),
  argrow(role, "overall",  "overall"),
  data.frame(Optimising = "—", Parameters = "trend (production)",
             Coverage = pct(m("trend", paste0(role,"_coverage"))),
             Accuracy = pct(m("trend", paste0(role,"_accuracy"))),
             Overall  = pct(m("trend", paste0(role,"_overall"))), check.names = FALSE),
  data.frame(Optimising = "—", Parameters = "nearest (no abstention)",
             Coverage = pct(m("nearest", paste0(role,"_coverage"))),
             Accuracy = pct(m("nearest", paste0(role,"_accuracy"))),
             Overall  = pct(m("nearest", paste0(role,"_overall"))), check.names = FALSE))
cr <- function(role, a, b) cor(grid[[paste0(role,"_",a)]], grid[[paste0(role,"_",b)]])
flat <- function(role) {
  f <- grid[[paste0(role,"_overall")]]
  near <- grid[max(f) - f <= 0.005, ]
  list(n = nrow(near), rmin = min(near$radius_nm), rmax = max(near$radius_nm),
       hmin = min(near$height_ft), hmax = max(near$height_ft[near$height_ft < UNRESTRICTED]))
}
fd <- flat("adep")
```

A natural instinct is to let one number decide. It cannot, and it is worth
seeing why before trusting any headline figure.

```{r argmax-dep}
kable(argtab("adep"), row.names = FALSE, align = "llrrr",
      caption = "Departures: the best cell in the whole grid under each metric taken alone.")
```

```{r argmax-arr}
kable(argtab("ades"), row.names = FALSE, align = "llrrr",
      caption = "Arrivals: the same. Note that no `endpoint` cell reaches `nearest`'s overall correctness.")
```

Each of these is degenerate in its own way.

**Maximising accuracy produces a rule quieter than the one it replaces.** The
best-accuracy departure cell answers `r pct(best_of("adep","coverage")$cov*0+best_of("adep","accuracy")$cov)`
of flights against production's `r pct(m("trend","adep_coverage"))`. It is more
accurate and *less useful*: fewer flights labelled, for a fraction of a point.
For arrivals it is starker still, at
`r pct(best_of("ades","accuracy")$cov)` coverage.

**Maximising coverage switches abstention off.** The best-coverage cell reaches
`r pct(best_of("adep","coverage")$cov)` against `nearest`'s
`r pct(m("nearest","adep_coverage"))` — the residual is only the aerodromes
with no elevation in OurAirports. At that setting `endpoint` *is* `nearest`,
and the reason for choosing it has been configured away.

**Maximising overall correctness is maximising coverage in disguise.** Across
the grid, overall correctness correlates
**`r sprintf("%+.3f", cr("adep","overall","coverage"))` with coverage** and
**`r sprintf("%+.3f", cr("adep","overall","accuracy"))` with accuracy** for
departures (`r sprintf("%+.3f", cr("ades","overall","coverage"))` and
`r sprintf("%+.3f", cr("ades","overall","accuracy"))` for arrivals). Coverage
spans `r pct(max(grid$adep_coverage) - min(grid$adep_coverage))` across the
grid and accuracy only
`r pct(max(grid$adep_accuracy) - min(grid$adep_accuracy))`, so the product
tracks the wider factor. And for arrivals the verdict of overall correctness is
explicitly *do not abstain*: `nearest` scores
`r pct(m("nearest","ades_overall"))`, above every `endpoint` cell.

**It is also too flat to discriminate.** `r fd$n` of the `r nrow(grid)`
departure cells sit within half a point of the best overall correctness,
spanning `r fd$rmin`–`r fd$rmax` NM and
`r format(fd$hmin, big.mark=",")`–`r format(fd$hmax, big.mark=",")` ft. Those
statistically indistinguishable cells differ by **thousands of wrong labels**.
Overall correctness cannot see the thing that separates them.

### The exchange rate, and the choice {#sec-op-choice}

What remains is a policy question that the data can inform but not settle: how
many extra wrong labels is one extra correct label worth?

There is a structural asymmetry that argues for pricing errors above correct
answers. **A wrong aerodrome corrupts two movement counts** — the aerodrome that
loses the flight and the aerodrome that wrongly gains it — and it is invisible
to a consumer, indistinguishable from a correct label. **A silent flight
depresses one count and is explicit**: it arrives as a null the consumer can
filter, and @sec-crosscheck's undercount is correctable in a way an inflated
count is not. Tenerife North makes the point concretely: a count ratio of
`r sprintf("%.2f", apt$count_ratio[apt$role=="arrivals" & apt$airport=="GCXO"])`
that looks healthy, on a recall of
`r pct(apt$recall[apt$role=="arrivals" & apt$airport=="GCXO"])`.

On that reasoning an error is worth at least two correct labels, and the
departure ladder stops at the last step above 2:1.

::: {.callout-tip}
## The operating point this study recommends

**Departures: `endpoint` at 30 NM, `r format(OP_HEIGHT, big.mark=",")` ft,
`r OP_PENALTY` NM scheduled-service penalty.**
`r sgn(cell(30,OP_HEIGHT,"adep")$correct - prod_cell("adep")$correct)` correct
departure labels against production for
`r sgn(cell(30,OP_HEIGHT,"adep")$wrong - prod_cell("adep")$wrong)` wrong ones.

**Arrivals: keep `trend`.** Nothing measured here improves on it, and the
cells that add correct labels add errors seven to seventeen times faster.

At a stricter exchange rate the departure answer moves inward — 5:1 gives
20 NM / 10,000 ft, 10:1 gives 20 NM / 8,000 ft — and at 1:1 it moves out to
40 NM / 20,000 ft. The rate is the decision; the parameters follow from it.
:::

The penalty of `r OP_PENALTY` NM is carried over from @sec-penalty, where it was
swept at `r OP_RADIUS` NM and `r format(OP_HEIGHT, big.mark=",")` ft. It has not
been re-verified at 30 NM, and should be before this is published.

## What `endpoint` could borrow from `trend` {#sec-hybrid}

@sec-op-arr leaves an uncomfortable result: for arrivals, the algorithm this
study has been arguing for loses to the one it proposed to replace, at every
setting measured. That is worth understanding rather than working around,
because the reason points at what a better rule would look like.

### Why `trend` wins arrivals {#sec-why-trend}

The two rules differ in how much of the trajectory they read.

**`endpoint` reads one sample.** The first or last fix of the track, its
distance to each candidate, its height above field elevation, its ground flag.
Nothing else.

**`trend` reads every sample below FL40 inside the zone.** For each
(track, aerodrome) pair it smooths altitude over five samples and counts rises
against falls. It sees a descent, not a point.

For departures that difference barely matters. A departure's first sample is
near the field whatever else is true, so one sample locates it almost as well as
fifty — which is why `endpoint` beats `trend` on departures by
`r sgn(cell(30,OP_HEIGHT,"adep")$correct - prod_cell("adep")$correct)` correct
labels.

For arrivals it matters a great deal, for the reason @sec-radius identified:
**an arrival's last sample is where reception died, which is a property of the
receiver network rather than of the flight.** The samples *before* it still
describe an approach. `endpoint` throws them away; `trend` uses them.

There is a second difference, and it is the more important one. `trend`'s
altitude vote is evaluated **per candidate aerodrome** — it asks "is this
aircraft descending toward *this* aerodrome", which discriminates between
candidates. `endpoint`'s thresholds are a **gate**: as @sec-modes puts it, the
abstention decides *whether* to answer, never *what* to answer. Naming is left
entirely to effective distance.

That matters because the dominant arrival failure is misnaming, not silence.
@sec-outliers found Bilbao answering
`r pct(apt$n_predicted_here[apt$role=="arrivals" & apt$airport=="LEBB"] / apt$n_truth[apt$role=="arrivals" & apt$airport=="LEBB"])`
of its arrivals and placing only
`r num(apt$n_correct[apt$role=="arrivals" & apt$airport=="LEBB"])` of
`r num(apt$n_truth[apt$role=="arrivals" & apt$airport=="LEBB"])` correctly.
**No gate can fix a naming error.** Tightening the thresholds makes the rule
answer less; it does not make it answer better.

### Tested: an approach cone instead of a rectangle {#sec-cone}

The first idea to try is the cheapest, because it needs no new data — only a
different shape for the same two numbers.

The current rule accepts an endpoint inside `radius_nm` **and** below
`height_ft`, treating distance and height as independent. On an approach they
are not: an aircraft descending toward an aerodrome trades height for distance
at roughly a constant rate. The 3° glideslope is 318 ft/NM and the "three times
the flight level" descent rule of thumb is about 333 ft/NM. So the endpoints
that genuinely belong to an aerodrome should lie in a **cone**,
`agl_ft ≤ k · dist_nm`, not a box.

The two shapes disagree in both corners:

* **close and high** is inside the box, outside the cone. An aircraft at
  15,000 ft overhead an aerodrome is overflying it. The box accepts this, and it
  is a pure source of wrong labels.
* **far and low** is outside the box, inside the cone. An arrival whose
  reception dies at 8,000 ft and 30 NM is on profile and is a good candidate.

```{r cone-table}
if (!is.null(cone)) {
  cn <- cone[cone$radius_nm == 40, ]
  cn <- cn[order(cn$slope_ft_nm), ]
  # The only fair comparison between two rule shapes is at equal coverage:
  # interpolate the box's accuracy along height at the same radius.
  boxacc <- function(role, cov) {
    b <- grid[grid$radius_nm == 40, ]
    b <- b[order(b[[paste0(role, "_coverage")]]), ]
    cs <- b[[paste0(role, "_coverage")]]; as_ <- b[[paste0(role, "_accuracy")]]
    if (cov <= min(cs) || cov >= max(cs)) return(NA_real_)
    approx(cs, as_, xout = cov)$y
  }
  ba_d <- sapply(cn$adep_coverage, function(c) boxacc("adep", c))
  ba_a <- sapply(cn$ades_coverage, function(c) boxacc("ades", c))
  kable(data.frame(
    `Slope` = paste0(cn$slope_ft_nm, " ft/NM"),
    `ADEP cov` = pct(cn$adep_coverage), `ADEP acc` = pct(cn$adep_accuracy),
    `box acc` = pct(ba_d), `cone gain` = pp(cn$adep_accuracy - ba_d),
    `ADES cov` = pct(cn$ades_coverage), `ADES acc` = pct(cn$ades_accuracy),
    `box acc ` = pct(ba_a), `cone gain ` = pp(cn$ades_accuracy - ba_a),
    check.names = FALSE), row.names = FALSE, align = "lrrrrrrrr",
    caption = "The approach cone at a 40 NM outer cap, against the rectangle **at matched coverage** -- the only fair comparison between two acceptance shapes. *box acc* is the rectangle's accuracy interpolated to the cone's coverage; *cone gain* is positive where the cone is better.")
}
```

```{r cone-worst}
#| include: false
worst_dep <- NA_real_
if (!is.null(cone)) {
  b <- grid[grid$radius_nm == 40, ]; b <- b[order(b$adep_coverage), ]
  cd <- cone[cone$radius_nm == 40, ]
  ba <- approx(b$adep_coverage, b$adep_accuracy, xout = cd$adep_coverage)$y
  worst_dep <- max(ba - cd$adep_accuracy, na.rm = TRUE)
}
```

**It does not work.** For departures the cone is worse at every slope — by up to
`r sprintf("%.2f pp", 100 * worst_dep)` of accuracy at matched coverage. For arrivals it is a wash: negative at most slopes, and the largest
positive figure is a tenth of a point, which is noise on this sample.

The departure failure is physical and worth stating, because it explains why the
box is the right shape there. **Climb gradients are far steeper than descent
gradients.** A departure 10 NM out is routinely at 6,000–8,000 ft — 600 to
800 ft/NM, against the glideslope's 318 — so a cone tuned to descent excludes
precisely the departures it should keep. The box's permissiveness close to the
field is not an oversight; it is what departures need. To reach the coverage the
box gives at `r format(OP_HEIGHT, big.mark=",")` ft, the cone needs a slope
above 1,000 ft/NM, at which point it is no longer a cone in any meaningful sense.

::: {.callout-important}
## The negative result that matters most

Neither shape closes the arrival gap. At the coverage where production sits, the
cone reaches
`r if (!is.null(cone)) pct(cone$ades_accuracy[cone$radius_nm==40 & cone$slope_ft_nm==650]) else "—"`
accuracy against `trend`'s `r pct(m("trend","ades_accuracy"))`.

Box and cone are both **acceptance regions defined on the distance and height of
a single endpoint**, and this experiment suggests that family is exhausted. No
rearrangement of those two numbers recovers what `trend` has. The remaining
gains have to come from somewhere else: **more evidence**, or **a naming rule**.
:::

### Untested, in order of expected value {#sec-untested}

**1. A descent-direction tie-break — a naming rule, not a gate.** Rank
candidates by effective distance as now, but when the top two are within a small
margin, break the tie on which aerodrome the aircraft is descending *toward*,
using the vertical rate at the endpoint and the bearing to each candidate. This
is the smallest piece of `trend` that can be borrowed, and it is the only
proposal here that attacks misnaming. It fires only on the ambiguous pairs, so
it cannot damage the flights that are already unambiguous — which is exactly the
property @sec-penalty found made the scheduled-service penalty safe. The
candidate cache would need `vert_rate` and a bearing, both available in the
state vectors at no join cost.

**2. Closest approach below FL40, instead of the last sample.** Replace "the
last fix" with "the fix of closest approach among those below FL40". For an
arrival whose reception fails on the approach, the closest approach is a far
better locator than the last recorded point. The cost is real: the candidate
builder currently emits two rows per track, which is what makes the sweeps in
this paper affordable (@sec-preprocessing), and this would emit more. A middle
course is to keep the two endpoints and add the closest-approach fix as a third.

**3. Re-testing the cascade — but read version 1 first.** The obvious move is to
take `trend`'s answer when it has one and fall through to `endpoint`. **Version 1
tested exactly this and it failed**, and the way it failed is the caution: every
rung had *negative lift*, meaning the trend rungs fired on easy flights where
nearest was already right, and overwrote better answers. The cascade collapsed
onto its fallback rung.

Whether that still holds is an open question rather than a settled one, because
version 1 measured it in the pre-H3 architecture and on departures, and this
version's arrival result is new. The diagnostic is cheap and decisive: attribute
each cascade answer to the rung that supplied it, and score that rung against
`endpoint` **on the same flights**. If `trend`'s arrival advantage comes from
answering *different* flights, a cascade will help; if it comes from being
better on *shared* flights, `trend` should simply replace `endpoint` for
arrivals, which is what @sec-op-arr already recommends. Version 1's harness has
the attribution code (`--diagnose-cascade`).

::: {.callout-warning}
## Not this: extrapolating the trajectory

The intuitive fix for an arrival whose reception dies at 8,000 ft is to project
the descent forward to the ground. **Do not.** Version 2 of this study
implemented trajectory extrapolation and it failed, and the 2025 PRC Data
Challenge reached the same conclusion independently and more strongly:
*"attempts to complete the trajectory always lead to worse result"*. Two
independent negative results on the same idea is enough.
:::

### Should any of this apply to departures too? {#sec-hybrid-dep}

The tie-break of proposal 1 should, and for the same reason it helps arrivals:
@sec-penalty showed the largest identified departure error class was a
neighbouring-aerodrome confusion, and the scheduled-service penalty only fixed
the subset where the neighbour was military. A descent-direction test —
inverted, a *climb*-direction test — discriminates on physics rather than on
aerodrome metadata, and would catch confusions the penalty cannot.

Proposals 2 and 3 should not. Departures do not have the evidence problem: one
sample already locates them, `endpoint` already beats `trend` there by
`r sgn(cell(30,OP_HEIGHT,"adep")$correct - prod_cell("adep")$correct)` correct
labels, and adding trajectory evidence to a rule that is not short of evidence
is how the version 1 cascade lost accuracy.

## Recommendation

**Adopt `endpoint` for departures if coverage matters more than the last point
of accuracy.** It adds `r pp(m("endpoint","adep_coverage") - m("trend","adep_coverage"))`
of coverage for `r pp(m("endpoint","adep_accuracy") - m("trend","adep_accuracy"))`,
and it is a smaller algorithm than the one it replaces — no smoothing window,
no margin parameter, no per-aerodrome vote. If accuracy is paramount, keep
`trend`; it is genuinely the more accurate rule now.

**Use 20–40 NM and a scheduled-service penalty of 10 NM.** Both are on
plateaus, so neither is delicate.

**Do not change the arrivals algorithm on this evidence.** Arrivals are weaker
on every measure and every version of this study has found the same.

**Publish the provenance column.** `adep_source` / `ades_source` distinguish
"named an aerodrome", "outside the observed area" and "could not determine".
Today all three are a single null in the published flight list, and they mean
entirely different things to a consumer. This is the highest-value change per
line of code in the whole study.

**Make abstention aware of what the reference contains.** Flights to and from
small aerodromes and heliports are answered about half the time and are wrong
every time (@sec-per-type). They are a small share of movements but they are
false statements rather than gaps, and they are indistinguishable in the output
from correct answers. Either extend the reference to small aerodromes or abstain
when the nearest candidate is far enough away to suggest the real aerodrome is
one the reference does not hold.

## Limitations {#sec-limitations}

* **Three days**, one sample. Versions 2 and 3 used two samples a year apart;
  this version re-runs the whole pipeline, which is expensive, so it uses one.
  The mode ranking is robust across earlier versions, but the exact figures
  here have not been validated on a second period.
* **Out-of-area recall is low.** The geometric border test is the wrong
  instrument; the phase-based test in @sec-ooa is the successor.
* **The detection reference holds large and medium aerodromes only**, so
  accuracy on any other class is zero by construction rather than by
  measurement. The per-class table quantifies this; it does not fix it.
* **The scheduled-service preference has a cost** that these accuracy figures
  include but do not itemise: movements at non-scheduled aerodromes are pulled
  to nearby hubs.
* **The sweep re-implements the endpoint rule** as a DataFrame operation
  instead of calling the pipeline's `classify_endpoints`, because a sweep cell
  needs only `(track_id, adep, ades)` while the pipeline method builds a whole
  flight list. The headline `endpoint` row *does* come from the pipeline path,
  so a divergence between the two would surface as the sweep disagreeing with
  its own mode — but this is a duplication and worth removing.
* **`track_id` is taken as given**, with the fragmentation and merging rates
  version 2 measured.

## Reproducing

Reference zones, state vectors and tracks, in the `opdi` repository:

```bash
RANGE="--env opensky --start 2025-06-05 --end 2025-06-08"
opdi run $RANGE --step 00a   # zones: 5 NM bands to 110 NM
opdi run $RANGE --step 01    # state vectors, bbox and 5 s applied at ingest
opdi run $RANGE --step 02    # tracks, with H3 res 7 and 12
```

Candidates once, then a flight list per mode:

```python
from datetime import date
from opdi.pipeline.flights import FlightListProcessor

fp = FlightListProcessor(spark, config)
fp.build_endpoint_candidates(date(2025, 6, 1))
for mode in ("trend", "nearest", "endpoint"):
    fp.process_dai(date(2025, 6, 1), mode=mode, skip_if_processed=False,
                   abstention_radius_nm=40, abstention_height_ft=15000,
                   sched_penalty_nm=10,
                   table_name=f"research/flight_list_{mode}",
                   write_mode="overwrite")
```

Scoring and both sweeps. This runs locally: the inputs are small once the
candidates are cached, and the cluster namespace is shared.

```bash
python benchmarks/benchmark_modes.py --months 202506 \
    --days 2025-06-05 2025-06-06 2025-06-07 --local --results-dir <dir>
```

Rendering this page needs no credentials, no database and no cluster.
