---
title: "Departure and destination aerodromes in the OPDI flight list"
subtitle: "How they are derived, how accurate they are, and how to use them"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r regenerate}
#| include: false
#| cache: false

# Rendering this page runs the analysis behind it.
#
# `regenerate_v7.py` is the executable definition of the study: every CSV read
# below is produced by exactly one job it declares, with exactly the arguments
# it declares. It re-runs a job when the *code that job depends on* has changed
# since the output was written -- not when the file is old, which says nothing.
# So a render with everything current is a fast no-op, and a render after
# editing `flights.py` recomputes every pipeline result before the first table
# is drawn.
#
# Running a stale job needs cluster credentials, because the numbers come from
# Spark over S3 against EUROCONTROL Network Manager reference data and there is
# no way to recompute them without the data they are computed from. Set
# OPDI_RENDER=check to fail fast on staleness instead, or
# OPDI_RENDER=allow-stale to render a draft from whatever is staged -- which
# the provenance table at the foot of this page will then show as unverified.
REPO   <- normalizePath(file.path(getwd(), "..", "..", "..", "opdi"), mustWork = FALSE)
PYTHON <- Sys.getenv("OPDI_PYTHON", file.path(REPO, ".venv310", "bin", "python"))
MODE   <- Sys.getenv("OPDI_RENDER", "run")

regen <- file.path(REPO, "benchmarks", "regenerate_v7.py")
if (file.exists(regen) && MODE != "skip") {
  flags <- switch(MODE,
                  "run" = character(0),
                  "check" = "--check",
                  "allow-stale" = c("--check", "--allow-stale"),
                  character(0))
  status <- system2(PYTHON, c(shQuote(regen), flags), stdout = "", stderr = "")
  if (!identical(status, 0L)) {
    stop("regenerate_v7.py exited ", status,
         " -- the figures below would not match the code that is checked out. ",
         "Set OPDI_RENDER=allow-stale to render anyway.")
  }
}
```

```{r setup}
#| include: false
library(knitr)

cache <- function(f) {
  p <- file.path("data", f)
  if (!file.exists(p)) return(NULL)
  read.csv(p, stringsAsFactors = FALSE)
}

t25    <- cache("trend_sweep_2025.csv")
t24    <- cache("trend_sweep_2024.csv")
e25    <- cache("sweep_radius_height_2025.csv")
e24    <- cache("sweep_radius_height_2024.csv")
lad25  <- cache("ladder_2025.csv")
lad24  <- cache("ladder_2024.csv")
mod25  <- cache("modes_2025.csv")
mod24  <- cache("modes_2024.csv")
grid25 <- cache("trend_grid_2025.csv")
grid24 <- cache("trend_grid_2024.csv")
apt25  <- cache("per_airport_2025.csv")
typ25  <- cache("per_type_2025.csv")

pct <- function(x) ifelse(is.na(x), "—", sprintf("%.2f%%", 100 * x))
pp  <- function(x) ifelse(is.na(x), "—", sprintf("%+.2f pp", 100 * x))
num <- function(x) ifelse(is.na(x), "—",
                          format(round(x), big.mark = ",", trim = TRUE))
# The "+" flag does nothing for %s in R, so the sign is prepended explicitly:
# a signed delta that silently loses its sign reads as a level, not a change.
sgn <- function(x) ifelse(
  is.na(x), "—",
  # An exact zero is not a small positive change, and "+0" reads as one. A rung
  # that moves nothing is a result in its own right on this page.
  ifelse(round(x) == 0, "0",
         paste0(ifelse(x > 0, "+", "−"),
                format(abs(round(x)), big.mark = ",", trim = TRUE))))

#: One row of a mode table, by run name.
mv <- function(d, run, field) {
  if (is.null(d)) return(NA_real_)
  v <- d[[field]][d$run == run]
  if (length(v) == 0) NA_real_ else v[1]
}

#: A parameter cell, identified the same way in both periods -- which is what
#: makes a joint ranking possible rather than a comparison of two argmaxes.
cellkey <- function(d) paste(d$fl_cap, d$margin, d$radius_nm, d$penalty_nm,
                             sep = "/")

#: Rank cells by both periods at once, each period's score divided by its own
#: reference count so the larger sample cannot dominate.
joint <- function(a, b, role) {
  if (is.null(a) || is.null(b)) return(NULL)
  a <- a[a$stage == 1, ]; b <- b[b$stage == 1, ]
  a$key <- cellkey(a); b$key <- cellkey(b)
  m <- merge(a, b, by = "key", suffixes = c("_25", "_24"))
  m$joint <- m[[paste0(role, "_score_25")]] / m$n_ground_truth_25 +
             m[[paste0(role, "_score_24")]] / m$n_ground_truth_24
  m[order(-m$joint), ]
}
```

::: {.callout-note}
## Who this is for

You are reading this because you use, or are deciding whether to use, the
`ADEP` and `ADES` columns of the [OPDI flight
list](https://www.eurocontrol.int/performance/data/download/OPDI). It covers
what those columns mean, how the pipeline derives them from raw ADS-B, how
often they are right, where they are systematically wrong, and what a
downstream analysis should do about it.

Every figure on this page is produced by running the pipeline over a sample of
ADS-B and scoring the result against EUROCONTROL Network Manager movement data.
Nothing is estimated or carried over from an earlier version.
:::

## Summary {.unnumbered}

OPDI reconstructs flights from ADS-B state vectors alone: no flight plans, no
schedules, no radar. From the resulting trajectory it must infer where each
flight departed from and where it arrived — which is straightforward when the
aircraft is seen on the runway at both ends, and hard in every other case.

Three findings shape what the flight list now contains.

**Departures and arrivals want different algorithms.** A departing aircraft's
first ADS-B fix is usually on or near the runway, so geometry alone settles it.
An arriving aircraft's last fix is often where reception was lost rather than
where it landed, so geometry alone does not. The flight list therefore derives
departures with one method and arrivals with another.

**Candidate aerodromes are chosen by exact distance.** The published algorithm
narrowed its candidates using a hexagonal spatial index and measured exact
distance only among the survivors, so an aerodrome slightly further out was
discarded before its distance was ever computed. Measuring first and choosing
afterwards is the single change that makes everything else in this study work.

**A departure's null is now two different answers.** `ADEP` can be an
aerodrome, the marker `OOA` for a flight that began outside the observed area,
or null for "could not tell", and a companion column says which. Earlier
releases collapsed the last two into a single null, and about one flight in
twelve genuinely belongs to the middle one.

**Arrivals do not carry that marker, and the asymmetry is deliberate.** The
label is decided by whether a trajectory's endpoint sits at the edge of
coverage. A track that *starts* at the edge almost certainly entered from
outside; a track that *ends* near it may equally be a flight still bound
somewhere inside whose reception was simply lost. Measured, the arrival label
was right about half the time, so it is not emitted. @sec-ooa gives the figures
and @sec-open what would change it.

## Reading this page

@sec-data says what the flight list contains. @sec-pipeline documents the
pipeline that produces it, step by step. @sec-metrics defines the measures.
Everything after that is evidence.

Terms are defined where they first appear and not before. Two conventions run
throughout:

- **legacy** means the algorithm and constants behind every OPDI flight list
  published before 2026 — reachable in code as `DetectionConfig.legacy()`, so
  released data stays reproducible.
- **shipped** means what this study recommends and what the pipeline now
  produces by default.

"Production" is used only to talk about the running system, never as the name of
an algorithm.

## What the flight list contains {#sec-data}

One row per flight, where a *flight* is a continuous stretch of ADS-B reception
from one aircraft under one callsign. The columns this page is about:

| Column | Meaning |
|---|---|
| `ID` | The track identifier. Stable within a release, **not** stable across algorithm versions — see @sec-versions. |
| `ICAO24` | The aircraft's 24-bit transponder address. The join key to most other aviation data. |
| `FLT_ID` | The broadcast callsign. Space-padded to eight characters in the raw feed; trim before joining. |
| `DOF` | Date of flight, taken from the first reception. |
| `ADEP` / `ADES` | Departure and destination aerodrome, ICAO code — or `OOA`, or null. |
| `ADEP_SOURCE` / `ADES_SOURCE` | `aerodrome`, `out_of_area`, or `undetermined`. |
| `ADEP_P` / `ADES_P` | Other aerodromes that were plausible. Comma-separated, ranked behind the chosen one. |
| `FIRST_SEEN` / `LAST_SEEN` | Reception bounds. Not off-block and in-block times. |
| `VERSION` | Which algorithm produced the row. |

### The three states of an aerodrome column

This distinction matters more than anything else on this page for a downstream
analysis, and it is new:

`aerodrome`
: An ICAO code. The pipeline saw enough to name it.

`out_of_area`
: `ADEP` is `OOA`. The flight entered the observed area rather than departing
  inside it — the trajectory begins at the edge of coverage, not near any
  aerodrome. **This is an answer, not a failure.** It appears on departures
  only: see the summary, and @sec-ooa for why. An analysis counting movements should exclude these; one measuring
  coverage should count them as correct.

`undetermined`
: `ADEP`/`ADES` is null. The pipeline could not tell. Filter these out; do not
  impute them.

Collapsing the last two — as every release before `v4.0.0` did — makes traffic
appear to be missing where in fact it was correctly identified as originating
elsewhere.

## The pipeline that produces it {#sec-pipeline}

Everything below is real ADS-B, processed by the code in the `opdi` repository.
No step is a re-implementation for the purposes of this report.

:::: {.content-visible when-format="html"}
```{mermaid}
%%| label: fig-pipeline
%%| fig-cap: "From the OpenSky archive to the published flight list. Boxes name the module; cylinders name the table it writes."
flowchart TD
    OSN[(OpenSky archive<br/>1 Hz global state vectors)]
    S1["<b>01</b> ingestion/osn_statevectors.py<br/><i>bounding box, 5 s thinning</i>"]
    T1[(osn_statevectors_v2)]
    S2["<b>02</b> pipeline/tracks.py<br/><i>split into tracks, H3 index</i>"]
    T2[(osn_tracks)]
    S2a["<b>02a</b> cleaning/cleaner.py<br/><i>mask implausible values</i>"]
    T2a[(osn_tracks_clean)]
    S0["<b>00</b> reference/<br/><i>H3 aerodrome zones, OurAirports</i>"]
    R0[(h3_airport_detection_zones)]
    S3["<b>03</b> pipeline/flights.py<br/><i>ADEP / ADES detection</i>"]
    T3[(opdi_flight_list)]
    S4["<b>04</b> pipeline/events.py<br/><i>milestones and measurements</i>"]
    S5["<b>05</b> output/<br/><i>parquet and CSV export</i>"]
    PUB[(published dataset)]

    OSN --> S1 --> T1 --> S2 --> T2 --> S2a --> T2a --> S3
    S0 --> R0 --> S3
    S3 --> T3 --> S4 --> S5 --> PUB
```
::::

:::: {.content-visible when-format="pdf"}
The same pipeline, step by step. Each row names the module and the table it
writes; every step reads the table written above it unless stated otherwise.

| Step | Module | Writes | Does |
|---|---|---|---|
| — | — | *OpenSky archive* | 1 Hz global state vectors |
| **00** | `reference/` | `h3_airport_detection_zones` | H3 aerodrome zones, OurAirports reference |
| **01** | `ingestion/osn_statevectors.py` | `osn_statevectors_v2` | European bounding box, thinning to 5 s |
| **02** | `pipeline/tracks.py` | `osn_tracks` | Split into tracks, attach the H3 index |
| **02a** | `cleaning/cleaner.py` | `osn_tracks_clean` | Mask implausible values, keep the rows |
| **03** | `pipeline/flights.py` | `opdi_flight_list` | **ADEP / ADES detection** — reads 02a and 00 |
| **04** | `pipeline/events.py` | events, measurements | Milestones along each flight |
| **05** | `output/` | published dataset | Parquet and CSV export |

::::

### 00 — Reference data

Builds what the detection needs to know about aerodromes. The important product
is the **H3 detection zone table**: for every aerodrome, the set of hexagonal
cells (resolution 7, about 1.5 NM across) covering concentric rings out to a
generous radius, each tagged with the ring's inner and outer distance.

This is a spatial index, not an answer. Its job is to turn "which aerodromes
might this fix be near?" into a hash lookup. Distances are measured afterwards,
exactly — @sec-ranking explains why that distinction turned out to matter more
than any threshold in the study.

Aerodrome positions, elevations and scheduled-service flags come from
OurAirports. The candidate set is restricted to large and medium aerodromes;
@sec-limitations quantifies what that excludes.

### 01 — State-vector ingestion

Reads the OpenSky archive and keeps what OPDI needs:

- **A bounding box**, roughly 25.9° W to 49.7° E and 26.7° N to 70.3° N. This is
  the *observed area*, and its edge is what the `OOA` marker refers to.
- **Thinning to 5 s.** The archive is a complete 1 Hz grid. OPDI keeps the
  newest row in each 5 s bin — a **bin-based** rule. Earlier releases used a
  fixed-phase rule, keeping a row only if one existed at the exact second
  divisible by five. @sec-sampler measures the difference.

The raw global 1 Hz feed is never stored.

### 02 — Track building

Groups state vectors into flights and assigns `track_id`. Two aircraft
transmissions belong to the same track when they share an aircraft address and
callsign and are not separated by a gap of more than 30 minutes, or more than 15
minutes below 5,000 ft — the second catching an aircraft that keeps its
transponder on while parked.

**This rule is frozen.** Changing it changes `track_id` for all published data,
so it is not tuned and not swept.

The step also attaches each fix's H3 cell at resolution 7, which is what makes
the aerodrome join in step 03 a lookup rather than a scan.

### 02a — Trajectory cleaning

Masks implausible values to null and keeps the row: duplicate state vectors,
out-of-range positions, stale broadcasts (ADS-B sends position and velocity in
separate messages, so identical consecutive values mean *repeated*, not
*measured*), derivative spikes, and isolated points. It also marks segment
boundaries at coverage gaps so nothing downstream interpolates across them.

Masking rather than deleting is deliberate: it preserves the distinction between
"no data" and "interpolated data". It does mean the detection has fewer usable
samples, since a fix with no barometric altitude cannot vote on a trend.
@sec-cleaning measures what that costs and what it buys.

### 03 — Flight list

Where `ADEP` and `ADES` are decided. Two algorithms, one per role.

#### What both share {#sec-ranking}

**Candidate selection.** Every aerodrome whose detection zone the aircraft
passed through is a candidate. The exact great-circle distance from the
trajectory's closest approach to each candidate is computed, and the smallest
wins. **No candidate is eliminated before its distance is measured.**

That sentence is the study's central finding, and it was not true before. The
legacy algorithm kept only the candidates in the innermost occupied hexagon
ring, then measured exact distance among those. A ring steps about 2.8 NM, so an
aerodrome a kilometre further out was routinely discarded unmeasured — and the
higher the aircraft, the more aerodromes share a ring and the less a ring index
can separate them. @sec-ladder shows what that was worth.

**The scheduled-service penalty.** An aerodrome without scheduled commercial
service has a fixed distance added to its effective distance before that
comparison, so a military or general-aviation field must be *clearly* nearest
rather than merely nearest. It re-ranks candidates without changing how many
flights are answered, so it can move accuracy and cannot move coverage.

#### `endpoint` — for departures

Take the aerodrome nearest the track's first fix, and accept it only if that fix
is within a radius and either on the ground or below a height *above field
elevation* — above the aerodrome's own elevation, since a fixed altitude cut-off
means nothing at a field sitting at 5,000 ft. Failing that, if the fix sits at
the edge of the observed area, answer `OOA`; otherwise abstain.

Its abstention is **geometric**: it declines when the endpoint is not where a
departure's endpoint should be.

#### `trend` — for arrivals

For every fix below a flight-level cap and within a radius of a candidate
aerodrome, ask whether smoothed barometric altitude is rising or falling. Each
fix votes. If descending votes beat climbing votes by more than a margin, that
aerodrome is a destination.

Its abstention is **evidential**: it declines when the altitude trace near an
aerodrome does not clearly rise or fall. That is why it suits arrivals — it uses
the whole approach rather than one fix, and one fix is exactly what an arrival
is least likely to have in the right place.

Two refinements complete it. Among candidates whose distances differ by less
than a small band, the one the aircraft's course actually points at wins
(@sec-bearing). And where `trend` cannot name an aerodrome, the same border test
`endpoint` uses can still distinguish "came from outside" from "cannot tell".

### 04–05 — Events and export

The flight list becomes the spine for milestone extraction — take-off, landing,
top of climb, level segments, flight-level crossings — and is exported as
parquet and CSV. Neither step changes `ADEP` or `ADES`.

## How the numbers are measured {#sec-metrics}

Two stages, and the distinction governs how much any figure is worth.

**The research sweep.** Hundreds of parameter combinations applied to a cached
intermediate table. Cheap enough to be exhaustive, which is what makes a grid of
this size possible — but it re-implements the decision rule rather than running
it, so it can only *propose* a value.

**The pipeline run.** The proposed values run through the same code path that
writes the published flight list, and scored identically. Nothing in this report
becomes a default without passing this stage.

### The sample

Three days, **5–7 June 2025**, with **5–7 June 2024** as an independent second
period. Both are full-day samples inside the bounding box, thinned to 5 s.

Ground truth is EUROCONTROL Network Manager movement data for the same days,
matched to OPDI tracks on aircraft address plus callsign plus date, and where
several tracks match, the one starting closest to the reference off-block time.
It is not redistributable, which is where reproducibility stops
(@sec-provenance).

### The measures

For each role — departures and arrivals — separately:

**Coverage**
: Of the reference flights, the share OPDI gives *any* answer for. `OOA` counts
  as an answer.

**Accuracy**
: Of the answers given, the share that are right.

**Correct / wrong**
: The same two quantities as counts, which is what makes settings comparable.
  Coverage and accuracy always move in opposite directions, and neither ratio
  alone says whether an exchange was worth taking.

**Score**
: `correct − k × wrong`, with `k = 2`. A judgement, stated so it can be
  disagreed with: a wrong aerodrome is worse than a silence because it corrupts
  *two* movement counts — the one it invents and the one it omits — and is
  invisible downstream, where a silence corrupts one and is an explicit null
  anyone can filter on.

A change is therefore worth taking exactly when it wins more than `k` correct
answers per wrong one it introduces.

**Out-of-area recall and precision** are reported separately, because a method
could otherwise buy headline accuracy by labelling `OOA` liberally. Recall is
the share of genuinely out-of-area flights labelled as such; precision is the
share of `OOA` labels that are correct. Where the report says *per-aerodrome
recall* later on, that is a different measure and is defined where it appears.

```{r sample-size}
#| echo: false
#| results: asis
if (!is.null(mod25)) {
  cat(sprintf("\nThe 2025 sample carries **%s** reference flights and the 2024 sample **%s**.\n",
              num(mod25$n_ground_truth[1]),
              if (is.null(mod24)) NA else num(mod24$n_ground_truth[1])))
}
```

## What each change is worth {#sec-ladder}

Each row adds one change to the row above **and keeps it**, so the marginal
worth of a change is the difference between adjacent rows, and the last row is
the shipped pipeline exactly. That is asserted before the run rather than
checked afterwards: the harness refuses a ladder whose final rung is not
`DetectionConfig()` in its thresholds, its per-role algorithms and its track
table, and refuses one containing a rung that changes nothing.

Read the *signs* across the two periods before the sizes. A change that helps
on one sample and hurts on the other has not been demonstrated, whatever its
magnitude on the sample it was tuned against.

```{r ladder}
#| echo: false
#| results: asis
lab <- c(
  L00_legacy              = "legacy algorithm, uncleaned tracks",
  L01_exact_rank          = "+ rank candidates by exact distance",
  L02_exact_radius        = "+ cut the radius by distance, not hexagon band",
  L03_smooth_first        = "+ smooth altitude before the flight-level cut",
  L04_penalty             = "+ scheduled-service penalty",
  L05_flcap               = "+ tuned flight-level cap",
  L06_margin              = "+ tuned vote margin",
  L07_bearing             = "+ bearing tie-break",
  L08_endpoint_departures = "+ departures from `endpoint`",
  L09_endpoint_radius     = "+ tuned endpoint radius",
  L10_clean_tracks        = "+ cleaned tracks (**= shipped**)")

show_ladder <- function(d, period) {
  if (is.null(d)) {
    cat(sprintf("*The %s ladder has not been staged yet.*\n\n", period))
    return(invisible(NULL))
  }
  # Stop rather than draw a short table. `match` against a stale label list
  # silently drops any rung it does not know, so a renamed step would vanish
  # from the ladder and the remaining rows would still add up -- a wrong table
  # that looks right. This has happened once already, when the ladder went from
  # thirteen rungs to eleven.
  missing <- setdiff(d$run, names(lab))
  if (length(missing)) {
    stop("ladder rung(s) with no label: ", paste(missing, collapse = ", "),
         ". The CSV and the label list have diverged; the table would silently ",
         "omit them.")
  }
  d <- d[match(names(lab), d$run), ]
  d <- d[!is.na(d$run), ]
  if (nrow(d) == 0) return(invisible(NULL))
  cat(sprintf("\n### %s\n\n", period))
  print(kable(data.frame(
    Step = unname(lab[d$run]),
    Departures = num(d$adep_score),
    `Δ dep` = c("—", sgn(diff(d$adep_score))),
    Arrivals = num(d$ades_score),
    `Δ arr` = c("—", sgn(diff(d$ades_score))),
    Total = num(d$adep_score + d$ades_score),
    check.names = FALSE), row.names = FALSE, align = "r"))
  tot <- d$adep_score + d$ades_score
  cat(sprintf("\nScores, not ratios: coverage and accuracy always move in opposite directions, and only the counts say whether an exchange was worth taking. Net over the whole ladder: **%s**.\n",
              sgn(tot[length(tot)] - tot[1])))
  cat("\nThe first eight rows produce both roles with `trend`, so each change is attributable to its own line. The ninth switches departures to `endpoint`, which is why the departure column jumps there and barely moves before it — every trend parameter above is tuned for arrivals.\n")
}
show_ladder(lad25, "2025")
show_ladder(lad24, "2024")
```

## The parameter sweeps {#sec-sweeps}

Both periods run on identical grids, and cells are ranked
by the two together — each period's score divided by its own reference count so
the larger sample cannot dominate. A single-period argmax appears only as a
stability check, because over hundreds of cells the top cell moves under noise
even when the surface underneath is stable.

```{r joint-trend}
#| echo: false
#| results: asis
if (is.null(t25) || is.null(t24)) {
  cat("*The trend sweeps have not been staged yet.*\n")
} else {
  for (role in c("adep", "ades")) {
    o <- joint(t25, t24, role)
    cat(sprintf("\n**%s — best cells on both periods together**\n\n",
                ifelse(role == "adep", "Departures", "Arrivals")))
    top <- head(o, 5)
    print(kable(data.frame(
      Rank = seq_len(nrow(top)),
      `Flight-level cap` = sprintf("FL%g", top$fl_cap_25),
      `Vote margin` = top$margin_25,
      `Radius (NM)` = top$radius_nm_25,
      `2025 score` = num(top[[paste0(role, "_score_25")]]),
      `2024 score` = num(top[[paste0(role, "_score_24")]]),
      check.names = FALSE), row.names = FALSE, align = "r"))
  }
}
```

```{r joint-endpoint}
#| echo: false
#| results: asis
if (is.null(e25) || is.null(e24)) {
  cat("\n*The endpoint sweeps have not been staged yet.*\n")
} else {
  cat("\n**The endpoint rule, which is what departures use**\n\n")
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(role) {
    a <- e25[which.max(e25[[paste0(role, "_score")]]), ]
    b <- e24[which.max(e24[[paste0(role, "_score")]]), ]
    data.frame(Role = ifelse(role == "adep", "Departures", "Arrivals"),
               `2025 best` = sprintf("%g NM / %s ft", a$radius_nm,
                    format(a$height_ft, big.mark = ",", trim = TRUE, scientific = FALSE)),
               `2024 best` = sprintf("%g NM / %s ft", b$radius_nm,
                    format(b$height_ft, big.mark = ",", trim = TRUE, scientific = FALSE)),
               Agrees = ifelse(a$radius_nm == b$radius_nm & a$height_ft == b$height_ft,
                               "**yes**", "no"),
               check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
}
```

### The flight-level cap, followed to the top

```{r fl-curve}
#| echo: false
#| results: asis
if (is.null(grid25)) {
  cat("*The pipeline grid has not been staged yet.*\n")
} else {
  # Whatever radius the grid was swept at -- not a literal, which is how
  # the caption came to name a radius the pipeline had stopped using.
  g <- grid25[grid25$trend_radius_nm == grid25$trend_radius_nm[1], ]
  g <- g[order(g$trend_max_fl), ]
  b <- if (is.null(grid24)) NULL else {
    x <- grid24[grid24$trend_radius_nm == grid24$trend_radius_nm[1], ]
    x[match(g$trend_max_fl, x$trend_max_fl), ]
  }
  tab <- data.frame(
    `Flight-level cap` = sprintf("FL%g", g$trend_max_fl),
    `2025 departures` = num(g$adep_score),
    `2025 arrivals` = num(g$ades_score),
    check.names = FALSE)
  if (!is.null(b)) {
    tab$`2024 departures` <- num(b$adep_score)
    tab$`2024 arrivals` <- num(b$ades_score)
  }
  print(kable(tab, row.names = FALSE, align = "r"))
  cat(sprintf("\nAt %g NM — the shipped radius — through the pipeline itself. Departures peak at FL%g and arrivals at FL%g on the 2025 sample. Coverage keeps rising with the cap — a higher cap admits more fixes — but accuracy falls faster, because fixes taken higher up are further from whichever aerodrome turns out to be the answer.\n",
              g$trend_radius_nm[1],
              g$trend_max_fl[which.max(g$adep_score)],
              g$trend_max_fl[which.max(g$ades_score)]))
}
```

## What ships, and how well it works {#sec-verdict}

```{r verdict}
#| echo: false
#| results: asis
lab_mode <- c(legacy = "legacy — the algorithm published before 2026",
              trend = "`trend` for both roles",
              endpoint = "`endpoint` for both roles",
              nearest = "nearest aerodrome, unconditionally",
              shipped = "**SHIPPED** — `endpoint` departures, `trend` arrivals")
if (is.null(mod25)) {
  cat("*The mode comparison has not been staged yet.*\n")
} else {
  d <- mod25[mod25$run %in% names(lab_mode), ]
  d <- d[order(-(d$adep_score + d$ades_score)), ]
  # The mode columns record what *executed*, not what the configuration held --
  # a single-mode run is built from the shipped config and then run with both
  # roles forced to one algorithm. Showing them answers "these numbers are for
  # which method?" in the table itself rather than in a caption.
  print(kable(data.frame(
    Configuration = unname(lab_mode[d$run]),
    `Departures by` = d$adep_mode,
    `Arrivals by` = d$ades_mode,
    `Dep. coverage` = pct(d$adep_coverage),
    `Dep. accuracy` = pct(d$adep_accuracy),
    `Arr. coverage` = pct(d$ades_coverage),
    `Arr. accuracy` = pct(d$ades_accuracy),
    Total = num(d$adep_score + d$ades_score),
    check.names = FALSE), row.names = FALSE, align = "r"))
  cat("\nEach row is a complete configuration — which algorithm names departures, which names arrivals — not a step in a sequence. The rows are alternatives to choose between.\n")
  cat("\nAll five read the **cleaned** tracks, so this table compares algorithms on one input. That is why the `legacy` row here does not match the ladder's first rung in @sec-ladder: this one is the legacy *algorithm* on current input, and that one is the legacy algorithm on the input it actually shipped with. The pair separates the algorithm from the data, and neither number alone does.\n")
  cat(sprintf("\nThe shipped configuration scores **%s against legacy's %s**, a gain of %s.\n",
              num(mv(mod25, "shipped", "adep_score") + mv(mod25, "shipped", "ades_score")),
              num(mv(mod25, "legacy", "adep_score") + mv(mod25, "legacy", "ades_score")),
              sgn(mv(mod25, "shipped", "adep_score") + mv(mod25, "shipped", "ades_score")
                  - mv(mod25, "legacy", "adep_score") - mv(mod25, "legacy", "ades_score"))))
}
```

### Which changes hold on both samples

A change measured on one sample has been observed; a change measured on two has
been *demonstrated*. The ladder runs identically on both periods, so every rung
can be graded on whether its sign survives.

```{r both-periods}
#| echo: false
#| results: asis
if (is.null(lad25) || is.null(lad24)) {
  cat("*Both ladders are needed for this table and have not both been staged yet.*\n")
} else {
  rungs <- intersect(names(lab), intersect(lad25$run, lad24$run))
  d25 <- lad25[match(rungs, lad25$run), ]
  d24 <- lad24[match(rungs, lad24$run), ]
  # Graded on the **total** -- departures plus arrivals. Two rungs change which
  # algorithm serves a role, and one changes the input, so an arrivals-only
  # reading would mark a change that helps departures more than it costs
  # arrivals as a cost. Out-of-area labelling is exactly that case.
  del <- function(d) c(NA, diff(d$adep_score + d$ades_score))
  a25 <- del(d25); a24 <- del(d24)

  # Below this a rung has not moved anything worth reading a sign off. Set as a
  # share of the reference count rather than a flat number, so it means the same
  # on both samples: 0.05% is about 48 flights here.
  floor25 <- 0.0005 * d25$n_ground_truth[1]
  floor24 <- 0.0005 * d24$n_ground_truth[1]
  nil <- abs(a25) < floor25 & abs(a24) < floor24

  verdict <- ifelse(is.na(a25) | is.na(a24), "—",
             ifelse(nil, "no material effect",
             ifelse(sign(a25) == sign(a24),
                    ifelse(a25 > 0, "**holds**", "**consistently costs**"),
                    "**disagrees**")))
  print(kable(data.frame(
    Step = unname(lab[rungs]),
    `2025 Δ total` = sgn(a25),
    `2024 Δ total` = sgn(a24),
    Verdict = verdict,
    check.names = FALSE)[-1, ], row.names = FALSE, align = "r"))
  cat(sprintf("\nΔ is the change in **total** score, departures plus arrivals, against the rung above. A rung moving less than %s flights on both samples is reported as having no material effect rather than having its sign read off noise.\n\n",
              num(floor25)))

  dis <- unname(lab[rungs])[which(verdict == "**disagrees**")]
  if (length(dis)) {
    cat(sprintf("\n**%s** changes sign between the two samples. A value that gains on the sample it was chosen against and loses on an independent one has not been demonstrated, and this study treats that as a reason to revisit it rather than to average the two.\n",
                paste(dis, collapse = "; ")))
  }
  cat("\n*Consistently costs* is not the same as *wrong*, and the two rungs carrying that verdict are worth reading rather than skipping.\n")
  cat("\n**Exact-distance ranking costs on its own.** It is not an improvement; it is an *enabler*. Two rungs later the flight-level cap is worth over four thousand on both samples, and under the hexagon-index selection it replaces, the same change lost ground. The cost here buys that, and quoting this rung alone tells the story backwards.\n")
  cat("\n**Cleaning costs too, and the report does not hide it.** Masking implausible values removes candidate fixes, so coverage falls. What it buys is accuracy and a trustworthy input for every other thing built on tracks. Whether that exchange is worth it at $k = 2$ is a judgement, and the number is here so a reader can disagree with it — see @sec-cleaning.\n")
  cat("\nOut-of-area labelling reads as *holds*, which corrects an easy misreading: it does cost arrival score, and it gains more on departures than it costs on arrivals, because flights beginning outside the observed area are now named correctly instead of being silently dropped. @sec-ooa gives both halves.\n")
}
```

### Does it hold on the second period?

```{r verdict-2024}
#| echo: false
#| results: asis
if (is.null(mod24)) {
  cat("*The second period has not been staged yet.*\n")
} else {
  rows <- do.call(rbind, lapply(c("legacy", "shipped"), function(r) {
    data.frame(Configuration = unname(lab_mode[r]),
               `2025 total` = num(mv(mod25, r, "adep_score") + mv(mod25, r, "ades_score")),
               `2024 total` = num(mv(mod24, r, "adep_score") + mv(mod24, r, "ades_score")),
               check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
  cat("\nNo parameter was chosen against one period and checked on the other: both were swept on identical grids and ranked together.\n")
}
```

## Three values this study changed {#sec-changed}

Version 6 shipped three constants that version 7 measured and found wanting.
All three are changed in the code this report describes, not merely noted: the
whole point of measuring a default is to act on the answer.

Each was tested the same way — applied to the **shipped configuration on its
own**, on both samples. That is a different question from a rung in
@sec-ladder, which asks what a change was worth at one point in a sequence.
This asks the question that decides: should it be on?

```{r changed}
#| echo: false
#| results: asis
if (!is.null(mod25) && !is.null(mod24)) {
  tot <- function(d, run) mv(d, run, "adep_score") + mv(d, run, "ades_score")
  delta <- function(d, run) tot(d, run) - tot(d, "shipped")
  rows <- data.frame(
    `Was → is` = c("radius 20 NM → **30 NM**",
                   "out-of-area on `trend` on → **off**",
                   "vote margin 2 → **0**"),
    `2025` = c(sgn(delta(mod25, "reject_radius20")),
               sgn(delta(mod25, "reject_trend_ooa")),
               sgn(-delta(mod25, "reject_margin2"))),
    `2024` = c(sgn(delta(mod24, "reject_radius20")),
               sgn(delta(mod24, "reject_trend_ooa")),
               sgn(-delta(mod24, "reject_margin2"))),
    check.names = FALSE)
  print(kable(rows, row.names = FALSE, align = "r"))
  cat("\nEach figure is the change in total score from adopting the new value, measured against the shipped configuration on that sample. Both columns positive means the change is supported on two independent periods.\n")
}
```

**The radius.** Version 6 moved it to 20 NM on a single period. On two it gains
on the sample it was chosen against and loses on the other — the only step in
this study whose sign did not survive an independent sample — and the research
sweep ranked over both periods puts 30 NM at an interior optimum. Reverted.

It turned out to be holding two other changes back. At 30 NM the bearing
tie-break is worth more, because a wider radius admits more candidate aerodromes
and so more near-ties for alignment to resolve; the switch of departures to
`endpoint` gains more as well. A value can be wrong not only in itself but in
what it suppresses, and that is invisible until it is changed rather than
tabled.

**Out-of-area on `trend`.** Built, measured, and left off. The capability is
real — `trend` could not previously distinguish "outside the observed area" from
"could not tell" — but in the shipped configuration it reaches only arrivals,
where the label is right about half the time against `endpoint`'s 89%. Every
label replaces a *silence*, so at $k = 2$ a coin flip is a losing trade.

The asymmetry is geometric rather than a tuning accident. A track that *begins*
at the edge of the observed area almost certainly entered from outside; one that
*ends* near it may equally be a flight still bound somewhere inside whose
reception was lost. @sec-ooa gives the figures.

**The vote margin.** Version 6 shipped 2 on research-sweep evidence and called
it the weakest value in its recommendation. The pipeline now agrees with the
sweep, on both samples, and the exchange rate settles it: dropping the margin
gains far more correct arrivals than wrong ones, well above the study's
threshold of two.

Cleaning is why it stopped earning its keep. The margin exists to stop noise
flipping a climb-or-descent call, and step 02a now masks about a fifth of
barometric altitudes before the vote is counted. It was defending against
something already removed. **That interaction could not have been found before
this version**, because cleaning did not previously reach the flight list at
all — which is the strongest argument in the study for having wired it in.

## What is still open {#sec-open}

Two things, stated so that nobody has to infer them from silence.

**Out-of-area for arrivals needs a direction test.** The border test asks
whether a track *ends* near the edge. It should ask whether the track is heading
*out* — a fix near the boundary on an inbound heading is a reception loss, not a
departure from the area. The course is already computed for the bearing
tie-break, so the machinery exists; the rule and its measurement do not. Until
then arrivals take their out-of-area labels from `endpoint`, at 89% precision.

**A third period, held out.** Every parameter here is chosen on two samples and
validated on those same two. A period against which nothing was tuned would say
whether the choices generalise or merely fit both Junes. It needs committed
Network Manager reference data for whichever month is chosen, which is a
decision about data availability rather than about method.

## Out-of-area, and the three states {#sec-ooa}

```{r ooa}
#| echo: false
#| results: asis
if (is.null(mod25)) {
  cat("*Being regenerated.*\n")
} else {
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(s) data.frame(
    Role = ifelse(s == "adep", "Departures", "Arrivals"),
    `Truly out-of-area` = pct(mv(mod25, "shipped", paste0(s, "_gt_ooa_share"))),
    `Labelled out-of-area` = pct(mv(mod25, "shipped", paste0(s, "_pred_ooa_share"))),
    Recall = pct(mv(mod25, "shipped", paste0(s, "_ooa_recall"))),
    Precision = pct(mv(mod25, "shipped", paste0(s, "_ooa_precision"))),
    `In-area accuracy` = pct(mv(mod25, "shipped", paste0(s, "_inarea_accuracy"))),
    check.names = FALSE)))
  print(kable(rows, row.names = FALSE, align = "r"))
  cat("\n*In-area accuracy* is accuracy restricted to flights whose true aerodrome is a real one — detection skill with the free out-of-area wins removed, and the number to read when comparing naming rules.\n")
}
```

## What cleaning costs and buys {#sec-cleaning}

Cleaning is the last rung of the ladder rather than a separate experiment, so
what it is worth is measured *given everything else already adopted* — which is
the only form of the question that bears on a decision. The two rows differ in
nothing but the track table.

```{r cleaning}
#| echo: false
#| results: asis
show_cleaning <- function(d, period) {
  if (is.null(d) || !all(c("L09_endpoint_radius", "L10_clean_tracks") %in% d$run)) {
    cat(sprintf("*The %s ladder has not been staged yet.*\n\n", period))
    return(invisible(NULL))
  }
  cat(sprintf("\n**%s**\n\n", period))
  rows <- do.call(rbind, lapply(c("adep", "ades"), function(s) data.frame(
    Role = ifelse(s == "adep", "Departures", "Arrivals"),
    `Raw coverage` = pct(mv(d, "L09_endpoint_radius", paste0(s, "_coverage"))),
    `Clean coverage` = pct(mv(d, "L10_clean_tracks", paste0(s, "_coverage"))),
    `Raw accuracy` = pct(mv(d, "L09_endpoint_radius", paste0(s, "_accuracy"))),
    `Clean accuracy` = pct(mv(d, "L10_clean_tracks", paste0(s, "_accuracy"))),
    `Δ score` = sgn(mv(d, "L10_clean_tracks", paste0(s, "_score"))
                    - mv(d, "L09_endpoint_radius", paste0(s, "_score"))),
    check.names = FALSE)))
  print(kable(rows, row.names = FALSE, align = "r"))
}
show_cleaning(lad25, "2025")
show_cleaning(lad24, "2024")
```

Cleaning **masks implausible values to null and keeps the row**, and the
detection path drops fixes with no barometric altitude — so it can only remove
candidate samples. Coverage falling and accuracy rising is the expected shape;
whether that exchange is worth taking at `k = 2` is what the score column says.

## The 5 s sampler {#sec-sampler}

The bin-based rule is adopted because it is the **correct** rule, not because it
measures better. It does not depend on the OpenSky archive being a complete 1 Hz
grid, which is a property of the archive rather than a guarantee; the
fixed-phase rule it replaces survives only because that property currently
holds, and degrades badly if it ever stops.

::: {.callout-note}
## This figure is not re-measured here, and why
Version 6 compared the two samplers across every parameter cell of two complete
runs — 371 cells of the trend grid and 126 of the endpoint grid, identical
grids over identically-treated input — and found the largest median coverage
difference anywhere to be a few *thousandths* of a percentage point, against a
decision bar of 0.5.

This report does not repeat that measurement, and says so rather than restating
the number as if it had. There is no fixed-phase arm left to compare against —
the bin-based rule is now what the pipeline does — and these sweeps run on
cleaned tracks over a different flight-level grid, so a comparison against
version 6's arm would be confounded by the cleaning and the grid rather than
measuring the sampler. Two arms on different data is exactly the error that
produced version 6's worst figure, and manufacturing a comparison here would
repeat it.
:::

Where the sampler *should* matter is not here at all. Aerodrome detection reads
an altitude trend over tens of samples, and one extra fix per five-second bin
cannot change the sign of a trend. Milestones defined by a single instant on the
ground — off-block, take-off, landing, in-block — are a different case, where a
recovered bin is the difference between having a fix and not. That measurement
belongs to the events study and has not been made.

One consequence is not reversible by configuration: **`track_id` values differ**
between the two samplers, because rescued rows sit at track boundaries and the
splitter breaks on gaps.

## The bearing tie-break {#sec-bearing}

This is the newest component in the algorithm and the best evidenced. It was
proposed by the research sweep, declined once for want of a pipeline
measurement, and is adopted here because `process_dai` confirms it on both
samples.


Among candidates whose distances differ by less than a
small band — that is, already equidistant to within the measurement — the tie is
broken by *alignment*: which aerodrome the aircraft's course actually points at.
Outside the band, distance decides alone.

The band is the whole mechanism. Alignment cannot *name* an aerodrome, because
every field on the same radial behind the right one is equally well aligned; it
can only separate two that distance has already declared equal.

## Using the columns {#sec-using}

**Filter on the source columns, not on nulls.** `ADES_SOURCE == 'aerodrome'` is
the set of named arrivals; `'out_of_area'` is the set correctly identified as
leaving the observed area; `'undetermined'` is the set to drop.

**Expect under-count, not over-count.** Abstention is the dominant failure mode,
so an aerodrome's movement count from OPDI is a lower bound. @sec-per-airport
gives the ratio per aerodrome for the hundred busiest.

**Join on `ICAO24` plus trimmed `FLT_ID` plus date.** Callsigns are
space-padded to eight characters in the raw feed.

**`FIRST_SEEN` is not off-block.** It is when reception began.

**Do not compare `ID` across versions.** See @sec-versions.

## By aerodrome, for the busiest hundred {#sec-per-airport}

*Being regenerated.*

```{r per-airport}
#| echo: false
#| results: asis
if (is.null(apt25)) {
  cat("*Not yet staged.*\n")
} else {
  cat("The matching is flight by flight, not count against count. **Per-aerodrome recall** is, of the flights whose *true* aerodrome is X, the share OPDI attributes to X. **Count ratio** is what OPDI assigns to X divided by X's true movements — it includes flights that truly belong elsewhere, so it is what an operational count would show, and it can exceed one.\n\n")
  rows <- do.call(rbind, lapply(c("departures", "arrivals"), function(rl) {
    r <- apt25[apt25$run == "shipped" & apt25$role == rl & apt25$airport != "OOA", ]
    l <- apt25[apt25$run == "legacy"  & apt25$role == rl & apt25$airport != "OOA", ]
    if (nrow(r) == 0) return(NULL)
    r <- r[order(-r$n_truth), ][1:min(100, nrow(r)), ]
    l <- l[match(r$airport, l$airport), ]
    data.frame(Role = tools::toTitleCase(rl),
               `Movements in truth` = num(sum(r$n_truth)),
               `Recall, shipped` = pct(sum(r$n_correct) / sum(r$n_truth)),
               `Recall, legacy` = pct(sum(l$n_correct, na.rm = TRUE) / sum(l$n_truth, na.rm = TRUE)),
               `Median count ratio` = sprintf("%.3f", median(r$count_ratio, na.rm = TRUE)),
               check.names = FALSE)
  }))
  print(kable(rows, row.names = FALSE, align = "r"))
}
```

## Versions, and what is reproducible {#sec-versions}

The `VERSION` column says which algorithm produced a row.

| Version | What it means |
|---|---|
| `v2.0.0` | Legacy `trend`, both roles. |
| `v3.0.0` | Legacy `endpoint`. In a merged list these two were coalesced into one column, so a row's version told you about the departure side only. |
| `v4.0.0` | This study: exact-distance ranking, per-role algorithms, tuned thresholds, the bearing tie-break, out-of-area labelling, and cleaned tracks as input. One string for the whole flight list. |

`DetectionConfig.legacy()` still returns every pre-2026 value, so re-processing
a month reproduces the aerodromes a release published.

::: {.callout-warning}
## `track_id` is not reproducible across the sampler change
The 5 s thinning rule changed at the same time. Rescued rows sit at track
boundaries and the splitter breaks on gaps, so a track that gains a fix can be
split differently. Re-ingesting an old month reproduces the *aerodromes* but not
the *identifiers*. Do not join `ID` across releases.
:::

## Limitations {#sec-limitations}

What is known to be missing or fragile, stated rather than implied.

- **Every shipped value is chosen on the same two samples it is validated on.**
  Nothing here is held out. @sec-open says why a third period would be worth
  having and what it needs.
- **The benchmark's own alignment is fragile where the flight list gains rows.**
  Each reference flight is paired with the track starting closest to its
  off-block time, so a configuration that produces *more* tracks can lose a
  pairing to one carrying no answer. This is visible wherever departures and
  arrivals come from different methods, and it is reported rather than corrected
  away: a benchmark that quietly adjusts its own alignment to flatter a result
  is worth less than one that shows where it is fragile.
- **The flight-level grid is swept at one radius.** The cap and the radius are
  measured separately rather than jointly in the pipeline — the research sweep
  covers the full grid, and the pipeline confirms the cap at the shipped radius
  only. A genuine interaction between the two would not be visible here, and one
  has already been found elsewhere in this study: the detection radius changed
  what the bearing tie-break was worth.
- **The cleaning comparison is one-sided.** Cleaning is measured as the ladder's
  last rung, and departures show no change there — `endpoint` reads a candidate
  cache built from cleaned tracks in every rung, so the earlier rungs were never
  really "raw" for that role. The figure is the cost to arrivals, not to the
  pipeline as a whole.
- **The legacy algorithm is not exactly reproducible.** Two identical runs of it
  named three arrivals differently out of 95,116. Ring-count selection leaves
  candidates tied on an integer, and the tie is resolved by an ordering that
  does not separate them. Exact-distance ranking removes those ties rather than
  arbitrating them, and every configuration using it reproduced exactly.
- **The candidate set is large and medium aerodromes.** A flight whose true
  aerodrome is a small field or a heliport cannot be named by any method here,
  whatever its thresholds, because the aerodrome is not a candidate. These
  flights are counted in every denominator on this page, so each coverage figure
  already carries their cost.
- **The border margin is fixed at 30 NM.** It decides what counts as leaving the
  observed area. Changing it redefines the ground truth as well as the method,
  so a sweep would move the target along with the aim.
- **H3 resolution 7 is fixed.** It no longer decides which candidate wins — that
  is exact distance now — but it still decides which aerodromes become
  candidates.
- **Two three-day samples.** Both in June. Seasonal effects are not measured.
- **Ground truth covers about ninety aerodromes.** Per-aerodrome figures are
  meaningful only for those.

## Provenance {#sec-provenance}

```{r provenance}
#| echo: false
#| results: asis
mf <- file.path("data", "_manifest.json")
if (!file.exists(mf)) {
  cat("*No manifest staged yet. Every figure above is therefore unverified.*\n")
} else {
  cat("Every CSV behind this page carries the script, arguments, commit and source-file fingerprint that produced it, in `data/_manifest.json`. An output with no entry is reported as unverified rather than shown as fact.\n")
}
```

### Where reproducibility stops

Everything on this page can be recomputed from the OpenSky archive and the code
in the `opdi` repository, with one exception: the EUROCONTROL Network Manager
movement data used as ground truth is not redistributable. It is an external
input, and every accuracy figure here is measured against it.
