---
title: "Flight events from ADS-B: fixing what OPDI publishes, and reaching the GANP KPIs"
subtitle: "Version 2 — with the extraction methodology for every event"
date: last-modified
author: "EUROCONTROL Performance Review Unit — Open Performance Data Initiative"
abstract: |
  OPDI has published flight events since v0.0.2 without a single test or a
  single comparison against reference data. This study audits the four
  detectors behind them, finds nine defects — one of which silently suppresses
  every take-off and landing at any aerodrome above roughly 200 ft — fixes
  them, and adds four milestone families chosen because they are what the ICAO
  GANP key performance indicators actually require. It reports what has been
  measured and, equally, what has not. Version 2 adds a full account of how
  each of the thirty-four published event types is extracted: the algorithm,
  its parameters and where they come from, the code that implements it, and
  what is known to be wrong with it.
format:
  html:
    toc: true
  pdf:
    toc: true
---

```{r regenerate}
#| include: false
#| cache: false

# Rendering this page runs the analysis behind it, as the version 7 report
# does. `regenerate_events.py` is the executable definition of the study:
# every figure below is produced by exactly one job it declares, and it
# re-runs a job only when the code that job depends on has changed since the
# output was written. A render with everything current is a fast no-op.
#
# OPDI_RENDER=check fails fast on staleness; allow-stale renders a draft from
# whatever is staged, which the provenance table then shows as unverified.
REPO   <- normalizePath(file.path(getwd(), "..", "..", "..", "opdi"), mustWork = FALSE)
PYTHON <- Sys.getenv("OPDI_PYTHON", file.path(REPO, ".venv310", "bin", "python"))
MODE   <- Sys.getenv("OPDI_RENDER", "allow-stale")

if (MODE %in% c("run", "check") && file.exists(PYTHON)) {
  args <- c(file.path(REPO, "benchmarks", "regenerate_events.py"))
  if (MODE == "check") args <- c(args, "--check")
  status <- system2(PYTHON, args, stdout = "", stderr = "")
  if (MODE == "check" && status != 0) {
    stop("regenerate_events.py --check reports stale outputs")
  }
}

read_json_if <- function(name) {
  path <- file.path("data", name)
  if (file.exists(path)) jsonlite::fromJSON(path) else NULL
}

read_csv_if <- function(name) {
  path <- file.path("data", name)
  if (file.exists(path)) read.csv(path, stringsAsFactors = FALSE) else NULL
}

bridge_2025 <- read_json_if("bridge_2025.json")
bridge_2024 <- read_json_if("bridge_2024.json")
manifest    <- read_json_if("_manifest.json")
ladder_2025 <- read_csv_if("ladder_2025.csv")
ladder_2024 <- read_csv_if("ladder_2024.csv")
inv_2025    <- read_csv_if("inventory_2025.csv")
inv_2024    <- read_csv_if("inventory_2024.csv")
```

## What changed in version 2

Version 1 reported *what* the detectors produce and how accurate the four
milestones APDF can score turned out to be. It never said *how* any event is
extracted, which left a reader unable to tell what triggers a `top-of-climb`,
what `entry-runway` matches against, or which parameter would change either.

That gap mattered more than it might elsewhere, because this study's findings
are about algorithm *shape* — that requiring complete evidence from a fuzzy
rule is harmful, that a smoothing window behaves differently on noisier input,
that a threshold detector is unusable without a dead band. None of those is
legible without the method beside it.

Version 2 adds that account and changes nothing else. **The data, the figures
and the findings are version 1's**, carried across with their provenance
records intact rather than regenerated.

## What this study is

Two things prompted it.

**The events were never checked.** OPDI's flight-event table is the
least-examined thing the project publishes. Departure and destination
aerodrome detection has been through seven successive studies and a full
rewrite; the event detectors had no tests at all, and no number in them had
ever been compared against anything.

**The event vocabulary did not reach the indicators.** The stated goal is to
compute as many ICAO GANP key performance indicators as ADS-B allows. Mapping
the published KPI list against what OPDI emitted showed the gap concentrated in
four milestone families — and that ICAO itself names ADS-B data providers as a
feed for two indicators OPDI could not compute at all.

## What was wrong

Nine defects, found by reading the detectors. None of them fails. Each produces
fewer events, or events displaced in one direction — the failure mode that
survives an eyeball check of the totals.

| | Defect | Consequence |
|---|---|---|
| D1 | Ground membership measured against **uncorrected pressure altitude** | The membership reaches zero at 200 ft *pressure* altitude, so at any aerodrome above roughly 200 ft AMSL — or anywhere on a low-pressure day — no sample is ever classified as on-ground. Since take-off requires a preceding ground sample and landing requires a ground sample, **neither event has ever been emitted for those flights.** The loss is biased against high-elevation aerodromes. |
| D2 | The reference implementation's 60-second majority smoothing was **not ported** | Phases were decided per state vector at 5-second spacing with no temporal aggregation, so one misclassified sample injects a spurious level segment or destroys a take-off. |
| D3 | Cruise speed membership centred on 600 kt | A turboprop at 250 kt scores 0.002, so it never reaches cruise and gets no top-of-climb or top-of-descent. **Inherited from OpenAP, not introduced here** — left unchanged pending measurement. |
| D4 | `least`/`greatest` skip nulls | A rule missing an input became a minimum over fewer terms, which can only *raise* its activation — so incomplete evidence out-competed complete evidence. |
| D5 | `first`/`last` inside a group-by | Take partition order, not time order, so a reported entry position was not guaranteed to be the position at the reported entry time. |
| D6 | Events read the **raw** track table | Trajectory cleaning reached the flight list and nothing else, so no published event has ever been derived from a cleaned trajectory. |
| D7 | Non-reproducible event identifiers; one version string for every type | The same event got a different identifier on every run, so two runs of a month could not be reconciled. |
| D8 | Flight-level crossings kept **only the first and last** per level, uninterpolated | A flight that levels off and re-climbs through a level lost every crossing in between — exactly the crossings a vertical-efficiency indicator is about — and every reported instant was biased by up to one sample interval in the same direction. |
| D9 | No tests | Not one detector was covered. |

D1 and D2 are the two that matter most, because they mean the published
take-off, landing, top-of-climb and top-of-descent events are missing
**non-randomly**.

## What was built

Eight defects are fixed and the ninth (D3) is left in place deliberately,
because changing a constant inherited from the reference implementation should
follow a measurement rather than precede one. Four milestone families were
added:

- **Threshold crossings**, rewritten. Every crossing rather than the first and
  last, with a hysteresis dead band so an aircraft cruising *at* a level emits
  none, and the instant interpolated to the exact threshold.
- **ASMA ring crossings** at 40 and 100 NM, built around the flight's own
  origin and destination — which is what the indicators mean, and what the
  reference data records.
- **ICAO level segments**, implementing the published algorithm rather than
  reusing the fuzzy-phase level events. The two are **not interchangeable**.
- **Runway identification with ATOT and ALDT**, and **off-block/on-block**.

Every change is reversible: a `legacy()` configuration reproduces published
`events_v0.0.2` exactly, and the new algorithm carries a new version string.

## Which KPIs this reaches

| KPI | Title | Reachable from ADS-B? |
|---|---|---|
| KPI02 | Taxi-out additional time | Yes |
| KPI05 | Actual en-route extension | Yes — ICAO names ADS-B as the feed |
| KPI08 | Additional time in terminal airspace | Yes |
| KPI10 | Airport peak throughput | Yes |
| KPI13 | Taxi-in additional time | Yes, coverage-limited |
| KPI15 | Flight time variability | Yes |
| KPI17 | Level-off during climb | Yes |
| KPI18 | Level capping during cruise | Yes (variant 2) |
| KPI19 | Level-off during descent | Yes |
| KPI01, KPI14 | Departure / arrival punctuality | Actual half only — the scheduled half needs a schedule feed |
| KPI11 | Airport throughput efficiency | Numerator only |
| Others | ATFM, capacity, flight plan, fuel, safety, noise | No |

Nine fully and three partially, from four milestone families.

## Ground truth, and its ceiling

Reference data is EUROCONTROL APDF. It is in movement form — there is no
literal take-off-time column, and `SRC_PHASE` decides what a row carries — and
crucially it has **no aircraft address**, so it cannot be joined to ADS-B
directly. The route runs through the flight table on an internal key.

That bridge is the ceiling on every coverage figure in this study: a milestone
that cannot be reached is indistinguishable from one that was missed, unless
the ceiling is known. So it is measured first.

```{r tbl-bridge}
#| echo: false
#| results: asis
if (!is.null(bridge_2025) && !is.null(bridge_2024)) {
  rows <- list(list("2024-06", bridge_2024), list("2025-06", bridge_2025))
  cat("| Period | APDF movements | Reaching `icao24` | Share |\n")
  cat("|---|---:|---:|---:|\n")
  for (r in rows) {
    b <- r[[2]]
    cat(sprintf("| %s | %s | %s | **%s%%** |\n", r[[1]],
                format(b$movements, big.mark = ","),
                format(b$reached_icao24, big.mark = ","), b$reach_pct))
  }
} else {
  cat("*Bridge measurement not present in the committed data.*\n")
}
```

Both periods clear 93%, so the ceiling is high enough not to dominate any
result. The residual loss is roughly three per cent of movements carrying no
internal key, and a further two per cent whose key has no counterpart in the
flight table.

Two traps in this join are worth recording, because both fail silently:

**The reference carries the commercial flight number, ADS-B carries the ATC
callsign.** The two disagree on 20.4% of rows — `LH416` against `DLH416` — and
joining on the wrong one discards a fifth of the sample while looking exactly
like a detection failure.

**Timestamps are tagged Europe/Paris despite being named UTC.** The instants
are correct, so converting is right and stripping the zone is a two-hour error.
Spark's session timezone is set to UTC in the local session builder and not in
the distributed one, so the loader asserts it rather than assuming it.


## How each event is extracted

Every event OPDI publishes is produced by one of eleven detectors. This section
states, for each, what the event means, how it is found, which parameters
govern it and where they come from, which module implements it, and what is
known to be wrong with it.

Three conventions apply throughout. **Storage is SI and thresholds are
aviation** — altitudes are metres and vertical rates metres per second on disk,
scaled at the point of comparison into feet and feet per minute, because those
are the units the thresholds are stated in. **Every threshold is named** in
`EventConfig` (`src/opdi/config.py`); none is an inline literal, so any of them
can be swept by the benchmark. And **`EventConfig.legacy()` reproduces
`events_v0.0.2`**, so the detectors below describe the new version while the
published one remains reachable.

### 1. Flight phase classification

Not an event type itself: it is the substrate for take-off, landing, top of
climb and descent, and the phase-derived level segments.

Each state vector is assigned a phase — ground, climb, descent, cruise or level
— by fuzzy logic. Nine membership functions score altitude, rate of climb and
speed; five rules combine them with a minimum t-norm; the phase is the rule
with the highest activation.

| Membership | Function | Parameters |
|---|---|---|
| ground altitude | `zmf` | 0 → 200 ft |
| low altitude | `gaussmf` | µ 10,000 ft, σ 10,000 |
| high altitude | `gaussmf` | µ 35,000 ft, σ 20,000 |
| zero climb rate | `gaussmf` | µ 0 ft/min, σ 100 |
| positive climb rate | `smf` | 10 → 1,000 ft/min |
| negative climb rate | `zmf` | −1,000 → −10 ft/min |
| high speed | `gaussmf` | µ 600 kt, σ 100 |
| medium speed | `gaussmf` | µ 300 kt, σ 100 |
| low speed | `gaussmf` | µ 0 kt, σ 50 |

**These constants are OpenAP's published values, not chosen here.** That
matters for one known limitation: a turboprop cruising at 250 kt scores 0.002
on the high-speed membership, so the cruise rule never wins and **turboprop and
general-aviation traffic get no top-of-climb or top-of-descent at all**. It is
inherited from the reference implementation rather than an OPDI defect, and it
is left in place deliberately — changing a reference constant should follow a
measurement rather than precede one.

Two things differ from a naive port:

**A 60-second majority smoothing** (`phase_twindow_seconds`) is applied to the
per-sample labels. OpenAP calls `phaselabel(twindow=60)`; the original OPDI
port carried the membership functions across and dropped the smoothing, so
phases were decided per state vector at 5-second spacing with no temporal
aggregation. One misclassified sample was enough to inject a spurious level
segment or destroy a take-off. Ties keep the unsmoothed label, which makes this
a de-flicker rather than a relabel. **This study found the change is not robust
across periods** — see the results section.

**Ground membership is measured above field elevation**
(`phase_ground_above_field`). `baro_altitude_c` is uncorrected pressure
altitude, so a 200 ft ceiling is unreachable at any aerodrome above roughly 200
ft AMSL or on a low-pressure day anywhere. The elevation of **both** the
track's ADEP and ADES is attached and the more permissive membership taken: a
track is on the ground at only one of its ends, and cruise sits far above both,
so the wrong end can only contribute a zero. A missing elevation falls back to
zero, which is the published behaviour rather than a loss of phases.

*Implemented in* `pipeline/events.py`: `calculate_horizontal_segment_events`,
`_smooth_phase`, `attach_field_elevation`.

### 2. `take-off` and `landing`

The instant the aircraft leaves or meets the ground, as the phase sequence sees
it.

`take-off` fires where the previous sample was ground and the next is climb.
`landing` fires where the previous sample was descent and the current is
ground. The two definitions are deliberately asymmetric — one looks forward,
one does not — which is inherited from the published detector and preserved so
the type keeps its meaning.

**Known weaknesses.** Both depend entirely on a ground sample existing, so
sparse surface reception removes them silently. Both are measured here against
APDF and shown to have low coverage (1.8% and 2.1%) with a very long tail — the
90th-percentile take-off error is 71 minutes. The runway-anchored `ATOT`/`ALDT`
below are the intended replacement; these keep their published type strings so
existing consumers are undisturbed.

*Implemented in* `pipeline/events.py`: `calculate_horizontal_segment_events`.

### 3. `top-of-climb` and `top-of-descent`

The first and last sample classified as cruise, per track.

Simple by construction, and therefore inheriting everything above: a flight
that never reaches the cruise rule — a turboprop, or anything with missing
velocity — has neither event. **Neither has ever been validated against an
external source**, because none exists that records them.

*Implemented in* `pipeline/events.py`: `calculate_horizontal_segment_events`.

### 4. `level-start` and `level-end` (phase-derived)

The boundaries of a run of samples classified cruise or level.

A `level-start` is emitted where the phase enters cruise or level from
something else, a `level-end` where it leaves. The first cruise sample carries
both `level-start` and `top-of-climb`; the last carries `level-end` and
`top-of-descent`.

**These are not the ICAO level segments** in section 11, and the two are not
interchangeable. This pair asks "does this sample look like level flight"
through the fuzzy classifier; ICAO asks a geometric question about an altitude
band anchored at a segment's own start. Both are published, under different
names.

Their volume is the clearest single measure of phase flicker: 1,270,412
`level-start` events under the published detector against 370,466 under the
new one, on the same three days.

*Implemented in* `pipeline/events.py`: `calculate_horizontal_segment_events`.

### 5. `first-xing-fl*` and `last-xing-fl*` — the published crossings

The first and last time a track crosses FL50, FL70, FL100 or FL245.

A crossing is detected by comparing each sample's flight level with the next
one's; the sample *before* the crossing is reported. A guard requires the
sample to be within 10 flight levels of the boundary, which rejects crossings
inferred across a data gap. The first and last per level are kept and
everything between is discarded.

**Three faults, which is why section 6 replaces this.** Keeping only first and
last discards every crossing in between — a flight that levels off and
re-climbs through FL100 loses exactly the crossings a vertical-efficiency
indicator is about. Reporting the bracketing sample biases every timestamp by
up to one sample interval, always in the same direction. And the level is
chosen by a first-match chain, so a single 5-second gap spanning FL50 *and*
FL70 reports only one of them.

**These types are emitted only under `events_v0.0.2`.** The new version does
not re-emit them; consumers migrate to `crossing_seq` in section 6.

*Implemented in* `pipeline/events.py`: `calculate_vertical_crossing_events`.

### 6. `xing-fl50`, `xing-fl70`, `xing-fl100`, `xing-fl245`

Every crossing of each monitored flight level, with the instant interpolated to
the level itself.

The detector is a Schmitt trigger. For a level `L` and dead band `h`, a sample
is assigned `+1` above `L + h`, `−1` below `L − h`, and null inside the band.
That side is forward-filled, and a crossing is confirmed where the filled value
flips. An aircraft cruising *at* the level never leaves the band, so its side
never resolves and **no event is emitted** — without this, barometric noise at
a cruise level produces a crossing on almost every sample.

The consequence has to be stated with the benefit: **a flight that climbs and
levels off exactly at FL100 registers no FL100 crossing**, because it never
clears the upper edge. That case is described by the level-segment and
top-of-climb events instead.

The instant is then interpolated to exactly `L` **between the two samples that
straddle it** — which is a different row from the one that confirmed the
crossing. Confirmation happens at the band edge; the crossing itself is
earlier. Conflating them reintroduces the one-sided bias this replaces. Each
event carries `crossing_seq` (1-based, in time order), `direction`, and
`bracket_s`, the spacing of the two interpolated-between samples, so a consumer
can tell an instant measured across 5 seconds from one inferred across four
minutes.

| Parameter | Value | Source |
|---|---|---|
| `crossing_levels_fl` | 50, 70, 100, 245 | the published vocabulary |
| `crossing_hysteresis_ft` | 300 ft | ours |
| `crossing_all_occurrences` | on | new |
| `crossing_interpolate` | on | new |

Thresholds are evaluated as parallel columns over one window specification
rather than by exploding the trajectory once per level, so four levels cost one
shuffle rather than four, and only rows where a crossing fired are expanded
into events.

*Implemented in* `pipeline/crossings.py`: `threshold_crossings`,
`flight_level_crossings`; wired by `events.calculate_threshold_crossing_events`.

### 7. `xing-40nm` and `xing-100nm` — the ASMA rings

Crossings of the 40 NM and 100 NM cylinders around the flight's **own** origin
and destination.

The same detector as section 6, with distance to an aerodrome as the quantity
instead of altitude, and a dead band of 1 NM. The rings come from the flight
list's ADEP and ADES rather than from the H3 aerodrome zone table, for two
reasons: it is what the indicators mean — KPI08's ASMA is a cylinder around the
destination and KPI05's reference area one around each end — and it is what
APDF records, one crossing per movement rather than one per aerodrome
overflown. It is also far cheaper, multiplying each sample by at most two
rather than by every aerodrome within 110 NM.

`info` carries `crossing_seq`, `direction` (`inbound`/`outbound`), `apt_icao`,
`bearing` and `flight_level`, matching `C40_CROSS_{TIME,LAT,LON,FL}` and
`C40_BEARING` column for column so the comparison against reference data is
direct rather than a translation.

**Bearing is computed from the interpolated crossing position, not interpolated
alongside it.** A bearing is circular: averaging 359 and 1 gives 180, which
would be wrong by half a turn on any crossing near due north.

**A bug worth recording, because it shaped this section.** The first run
produced 2.85 million 40 NM crossings and 3.20 million at 100 NM against
236,899 tracks — with *identical* counts for both rings on the worst flights, a
signature real geometry cannot produce. The cause was in a shared helper:
`haversine_nm` clamped the argument of `asin` with `F.least(a, 1.0)`, and
`F.least` skips nulls, so a missing coordinate became `asin(1)` — 10,807 NM,
the antipodal distance. A transponder parked at Budapest transmitted for the
whole 72-hour sample; cleaning nulled its repeated positions as stale
broadcasts, and its distance series then alternated between 0.7 NM and the
antipode, crossing both rings on nearly every sample. After the fix the counts
are 151,526 and 138,512, a 19- and 23-fold reduction.

*Implemented in* `pipeline/crossings.py`: `ring_crossings`; wired by
`events.calculate_ring_crossing_events`.

### 8. `entry-*` and `exit-*` — airport layout events

*Types emitted:* `entry-runway`, `exit-runway`, `entry-taxiway`,
`exit-taxiway`, `entry-apron`, `exit-apron`, `entry-hangar`, `exit-hangar`,
`entry-threshold`, `exit-threshold`, `entry-parking_position`,
`exit-parking_position`, `entry-deicing_pad`, `exit-deicing_pad`.

Entry to and exit from runway, taxiway, apron, hangar, threshold, parking
position and de-icing pad.

Low-altitude samples are matched against OpenStreetMap aeroway geometry that
has been pre-rasterised to H3 resolution 12 hexagons in
`hexaero_airport_layouts`. A sample joins a feature when its `h3_res_12` cell
matches and the feature's aerodrome is one the flight list names for that
flight. Consecutive matches on the same feature form a traversal; a gap of more
than 300 seconds starts a new one. Each traversal yields an entry and an exit
event.

| Parameter | Value | Source |
|---|---|---|
| `airport_max_fl` | FL20 | the published gate |
| `airport_trace_gap_seconds` | 300 s | the published gate |
| H3 resolution | 12 | `H3Config.airport_layout_resolution` |

**Entry and exit attributes are taken from the samples at the entry and exit
times** (`airport_events_ordered`). The published detector used `first`/`last`
inside a group-by, which take partition order rather than time order — so a
reported entry position was not guaranteed to be the position at the reported
entry time. This is a correctness fix, not a tuning knob.

**Two limitations that bound coverage.** The detector cannot fire for an
aerodrome the flight list did not name, and it depends on OSM coverage, which
`hexaero_airport_layouts` builds only for large and medium aerodromes within
the ingestion area. A traversal completed inside a single 5-second sample is
also dropped, because its duration is zero.

`entry-threshold` and `exit-threshold` fall from 774 to 9 between the published
and new configurations. The threshold polygons are the smallest features in the
layout set, so they are the most sensitive to which samples survive cleaning —
this is the clearest instance of trajectory cleaning removing the samples a
detector depends on, and it is not a detector change.

*Implemented in* `pipeline/events.py`: `calculate_airport_events`.

### 9. `ATOT` and `ALDT` — runway identification and touchdown

Take-off and landing times, anchored on the runway the movement actually used.

A native port of `traffic`'s `TrackBasedRunwayDetection`. Samples are kept when
they are within 5 NM of the aerodrome, below 1,500 ft above field elevation,
faster than 30 kt over the ground, and climbing or descending faster than 257
ft/min in the direction the role implies. At least four must survive. The
median track of the survivors is matched against each runway's bearing, and the
nearest within 10 degrees wins; parallel runways are separated by cross-track
distance to each centreline, in closed form where `traffic` uses a geometry
library.

Runway geometry comes from `oa_runways` — the OurAirports table step 00d has
been generating since it was written and which nothing had ever read. It is the
same data `traffic` downloads at runtime, so this adds no dependency and
nothing to warm on an offline executor. **Bearings are computed from the two
threshold positions** rather than taken from `le_heading_degT`, which is often
null and sometimes magnetic rather than true.

| Parameter | Value | Source |
|---|---|---|
| `runway_max_dist_nm` | 5 NM | `traffic` |
| `runway_max_height_ft` | 1,500 ft | `traffic` |
| `runway_min_vert_rate_ftmin` | 257 ft/min | `traffic` |
| `runway_min_groundspeed_kt` | 30 kt | `traffic` |
| `runway_max_bearing_deg` | 10° | `traffic` |

The detector **abstains** when no runway falls within the bearing tolerance.
Naming the wrong runway corrupts a movement count; naming none is an explicit
null.

**Both times are proxies**, and are documented as such: `ATOT` is the earliest
surviving sample of a departure and `ALDT` the latest of an arrival. Their bias
against reference data is measured and reported in the results rather than
assumed to be zero.

*Implemented in* `pipeline/runways.py`: `runway_thresholds`,
`detect_runway_movements`; wired by `events.calculate_runway_events`.

### 10. `AOBT` and `AIBT` — block times

Off-block and on-block.

The aircraft is moving once its groundspeed stays above 2 kt for a sustained
30 seconds — a port of the signal in `traffic`'s `StartMoving`, where the
sustained part is what separates a push from a jitter in the speed field. The
start of the first such run is the off-block candidate and the end of the last
the on-block candidate.

Both are **anchored on the parking-position events** from section 8: a
candidate is only emitted for a track that was actually seen leaving or
entering a stand. Without that anchor, a track first received mid-taxi would
report its first movement as an off-block — a different event, and one that
reads as an implausibly short taxi rather than as a miss.

`traffic`'s own parking-position path is deliberately **not** used: it resolves
stands through a live OpenStreetMap Overpass query whose cache does not survive
serialisation to an executor, so it cannot run without network access. OPDI's
`hexaero_airport_layouts` already carries the same geometry.

| Parameter | Value | Source |
|---|---|---|
| `ground_speed_threshold_kt` | 2 kt | `traffic` |
| `ground_move_min_seconds` | 30 s | `traffic` |

**Coverage is the limitation, and it is measured**: 2.6% of reference
movements, with the off-block proxy firing around twelve minutes late. Surface
reception is sparse, and sustained movement is detectable only once reception
begins. Both numbers are reported as the finding rather than tuned away, and
they are the reason pushback (T05) was not built — it needs strictly more
coverage than this.

*Implemented in* `pipeline/ground.py`: `movement_window`, `block_times`; wired
by `events.calculate_block_events`.

### 11. ICAO level segments

*Types emitted:* `level-off-climb-start`, `level-off-climb-end`,
`level-off-descent-start`, `level-off-descent-end`.

Level flight during climb and descent, as ICAO defines it for KPI17 and KPI19.

ICAO specifies these with a complete algorithm rather than leaving detection to
the implementer, and this is that algorithm:

> a data point starts a level segment when the altitude difference with the
> next data point is ≤ the level band limit and the vertical speed towards it
> is ≤ the vertical speed limit. The segment ends when the altitude differs
> from the segment's starting altitude by more than the level band limit, or
> the vertical speed between two consecutive points exceeds the limit.

The anchoring is what a plain window function cannot express: a segment's end
is defined against *its own start*, a running value that resets. It is
implemented as a sessionisation — mark where flatness begins, forward-fill the
anchor within the group — and each group is truncated at its first failure, so
a trajectory that leaves the band and drifts back is not stitched across the
excursion. Anchoring against the segment start rather than the previous sample
is what stops a slow drift being read as one long level segment.

Segments are then split into KPI17 and KPI19 by whether they fall before
top-of-climb or after top-of-descent, with two exclusions: a **minimum
altitude** below which the trajectory is not analysed, and an **exclusion box**
that removes a segment sitting above 90% of the cruise altitude and lasting
longer than five minutes — that is cruise, not a level-off, and without the box
every cruise would be counted as the largest level-off in the flight.

| Parameter | Value | Source |
|---|---|---|
| `level_band_limit_ft` | 200 ft | ICAO |
| `level_vertical_speed_limit_ftmin` | 300 ft/min | ICAO |
| `level_min_duration_seconds` | 20 s | ICAO |
| `level_exclusion_box_pct` | 90% | ICAO |
| `level_exclusion_box_seconds` | 300 s | ICAO |
| `level_min_altitude_climb_ft` | 3,000 ft | ICAO |
| `level_min_altitude_descent_ft` | 1,800 ft | ICAO |
| `level_analysis_radius_nm` | 200 NM | ICAO |

Top-of-climb and top-of-descent are taken from the phase detector's own output
rather than recomputed, so the two families cannot disagree about where cruise
began — which matters because the exclusion box is defined against the cruise
altitude.

**No source available to OPDI holds level-segment truth**, so the claim made
for this family is conformance to the published algorithm, tested on synthetic
trajectories whose geometry is known by construction. That is checkable;
"accurate" is not.

*Implemented in* `pipeline/level_segments.py`: `level_segments`,
`classify_level_offs`; wired by `events.calculate_level_off_events`.

### `first_seen` and `last_seen`

The first and last sample of each track, by time. Emitted for every track
without condition, which is why their count — 236,899 on the sample — is also
the track count.

*Implemented in* `pipeline/events.py`: `calculate_firstseen_lastseen_events`.

### Measurements

Two values are attached to every event rather than being events themselves.

**`Distance flown (NM)`** is the cumulative great-circle distance from the
start of the track, summed sample to sample with an Earth radius of 6,371 km
and converted at 1.852 km per nautical mile. **`Time Passed (s)`** is the
elapsed time from the track's first sample.

Both are **interpolated to the crossing instant** for the crossing families
rather than read from a neighbouring sample — a measurement attached to an
interpolated event but taken from a different row would disagree with its own
event.

*Implemented in* `pipeline/events.py`: `add_distance_measure`,
`add_time_measure`.

### Event identity

Each event's identifier is derived from the event's own identity — track,
type, time and version — rather than from a partition counter. The published
detector used `monotonically_increasing_id()`, which encodes the partition
index, so the same event received a different identifier on every run and two
runs of a month could not be reconciled.

This makes a re-run idempotent in *content*. It does **not** make the write
idempotent: the storage layer appends, and the event tables are partitioned by
`(type, version)` rather than by month, so an overwrite would destroy every
other month. Re-processing safety currently rests on the per-month progress
logs, and a month partition on the published tables is recorded as deferred
work.

## What the detectors produce

Every event type step 04 emits, counted on the sample. Most of these have no
counterpart in any reference dataset, so this table is the only place they are
visible at all — a family that stopped being emitted would otherwise show up
not as a regression but as an absence, which is harder to notice.

```{r tbl-inventory}
#| echo: false
#| results: asis
if (!is.null(inv_2025)) {
  w <- reshape(inv_2025, idvar = "type", timevar = "rung", direction = "wide")
  names(w) <- sub("^n_events\\.", "", names(w))
  keep <- intersect(c("type", "L00_legacy", "L13_shipped"), names(w))
  w <- w[order(w$type), keep, drop = FALSE]
  cat("| Event type |", paste(keep[-1], collapse = " | "), "|\n")
  cat("|---", paste(rep("---:", length(keep) - 1), collapse = "|"), "|\n", sep = "")
  for (i in seq_len(nrow(w))) {
    vals <- vapply(keep[-1], function(k) {
      v <- w[i, k]
      if (is.na(v)) "—" else format(v, big.mark = ",")
    }, character(1))
    cat(sprintf("| `%s` | %s |\n", w$type[i], paste(vals, collapse = " | ")))
  }
} else {
  cat("*Extraction inventory not present in the committed data — the ladder has not been run.*\n")
}
```

## How accurate the milestones are

Scored against APDF on the milestones it can reach. Coverage is the share of
reference milestones with any detection; bias is the median signed error, so a
negative value means the detector fires early.

```{r tbl-ladder}
#| echo: false
#| results: asis
if (!is.null(ladder_2025)) {
  d <- ladder_2025
  # det_type is the OPDI event type; milestone is the APDF milestone it claims
  # to be. Both are shown because `take-off` and `ATOT` are scored against the
  # same truth, and which of them wins is the question the ladder answers.
  # Each OPDI type is scored against every truth milestone, so `take-off`
  # also produces rows against ALDT, AOBT and AIBT with no detections. Those
  # are true but meaningless -- a take-off event is not a failed on-block.
  # Keep only the pairing each type actually claims.
  claims <- c("take-off" = "ATOT", "landing" = "ALDT", "ATOT" = "ATOT",
              "ALDT" = "ALDT", "AOBT" = "AOBT", "AIBT" = "AIBT")
  if (!is.null(d$det_type)) {
    d <- d[!is.na(claims[d$det_type]) & claims[d$det_type] == d$milestone, ]
  }
  d <- d[order(d$milestone, d$rung), ]
  cat("| Rung | OPDI type | vs | Truth | Detected | Coverage | Bias (s) | MAD (s) | ±60 s |\n")
  cat("|---|---|---|---:|---:|---:|---:|---:|---:|\n")
  for (i in seq_len(nrow(d))) {
    dt <- if (!is.null(d$det_type)) d$det_type[i] else d$milestone[i]
    cat(sprintf("| %s | `%s` | %s | %s | %s | %.2f%% | %.1f | %.1f | %.2f%% |\n",
        d$rung[i], dt, d$milestone[i],
        format(d$n_truth[i], big.mark = ","),
        format(d$n_detected[i], big.mark = ","),
        d$coverage_pct[i], d$bias_s[i], d$mad_s[i], d$within_60s_pct[i]))
  }
} else {
  cat("*Ladder results not present in the committed data — see below.*\n")
}
```

## What the numbers say

### Landing time is transformed; take-off trades sharpness for reach

Both the published fuzzy-phase pair and the new runway-anchored pair are scored
against the same APDF truth, because which of them wins is the question this
study exists to answer.

The published `landing` is beaten on every axis at once: the new `ALDT` covers
forty times as many flights, its median error is two seconds early rather than
thirty-four late, and four in five detections fall within a minute against one
in sixty.

`take-off` is the more interesting case, and the honest reading is a trade
rather than a win. The published detector has the *better median* — a one
second bias against nineteen — but it fires on 1.8% of flights and its 90th
percentile error is **4,273 seconds**. Seventy-one minutes. So it is often
excellent and occasionally absurd, and nothing in its output distinguishes the
two. The new `ATOT` covers thirty-five times more flights with a 90th
percentile of 59 seconds. For an indicator averaged over a population, a
detector that is never wildly wrong is worth more than one that is sometimes
perfect.

### Requiring complete evidence made detection worse

The ladder's purpose is attribution, and it earned that here by contradicting
one of this study's own fixes.

`least`/`greatest` skip nulls, so a fuzzy rule missing an input became a
minimum over fewer terms and could out-compete a complete rule. Requiring every
input to be present is obviously more correct. On real data it removes 90% of
phase-based landings — coverage falls from 6.49% to 0.69% — because ADS-B
carries a null velocity often enough that insisting on complete evidence
discards most ground samples.

It matters less than it appears, since the runway-anchored detectors replace
that path entirely. But it is a genuine negative result, it would have been
invisible in a two-configuration comparison, and it is recorded here rather
than quietly reverted.

### Off-block and on-block are limited by reception, as expected

Coverage of 2.6% and a bias of +744 seconds. The bias is not noise: sustained
movement is detectable only once reception begins, which is typically well
after the aircraft actually leaves the stand. Both numbers are reported as the
finding rather than tuned away, and they are why T05 (pushback) was not built —
it needs strictly more coverage than this.

### Does it hold on a second period?

Both periods were run. What 2024 can check is narrower than 2025: no flight
list exists for that period, and the runway, ring, airport and block detectors
all need one, so 2024 exercises the phase and crossing families only. The ALDT
and ATOT headline figures are therefore **single-period**.

For what it does cover, the two disagree in an instructive way.

| Rung | `landing` coverage, 2025 | 2024 |
|---|---:|---:|
| `L01_clean_tracks` | 6.39% | 6.66% |
| `L02_smoothing` | 6.49% | **0.96%** |
| `L03_complete_rules` | 0.69% | 0.23% |

**Requiring complete fuzzy rules reproduces.** A large relative collapse on
both periods -- 89% and 76% -- so the finding is a property of the change
rather than of a month.

**The smoothing does not.** Restoring the reference implementation's 60-second
majority vote is near-neutral on 2025 and costs 86% of landing coverage on
2024. This was the change with the strongest prior justification in the whole
study, and it is the one the second period contradicts.

One confound has to be stated with it. The 2024 tracks predate step 02's
altitude repair, so that run substitutes the raw barometric altitude. The
period therefore differs in input preparation as well as in date, and the
smoothing operates on phase labels derived from exactly that column. The
divergence may be the substitution rather than the month, and this study cannot
separate the two. What it can say is that the change is **not robust across the
conditions tested**, which is the useful half of the claim: a fix that behaves
this differently on noisier input should not be adopted on the strength of its
rationale alone.

### What has still not been measured

**The vertical family cannot be scored against data at all.** No source
available to OPDI holds level-segment truth. The claim made for KPI17 and
KPI19 is conformance to ICAO's published algorithm, tested on trajectories
whose geometry is known by construction. That is checkable; "accurate" is not.

**Top-of-climb and top-of-descent timing rests on no external evidence.** The
counts move — 114,099 to 100,925 for TOC — but nothing says which is closer to
the truth.

**The 2025 ladder is stale against the current code.** It was produced before
the changes that made the 2024 period runnable -- the track redirect, the
column backfills, the identity source. Those cannot alter a 2025 result, but
the provenance fingerprint covers the file rather than the behaviour, so the
chain reports it stale and this page says so rather than re-recording it as
current. Clearing it honestly means re-running it.

**Ring crossings are not yet scored against `C40_CROSS_TIME`.** The detector
runs and its volumes are now plausible, but the comparison against the
reference crossing times is not in this version.

## Breaking change for consumers

The new version does **not** re-emit `first-xing-fl*` and `last-xing-fl*`.
Those types keep their meaning under the frozen `events_v0.0.2`, so published
data stays reproducible, but consumers moving to the new version must migrate
to the crossing sequence number carried in each event's `info`: the first
crossing is sequence 1, the last is the maximum.

Consumers should also know that `level-start`/`level-end` and the new ICAO
`level-off-*` family answer different questions and are not interchangeable,
and that the event table grows materially — every crossing is now emitted where
previously only two per level were.

## Provenance

```{r tbl-provenance}
#| echo: false
#| results: asis
if (!is.null(manifest) && length(manifest) > 0) {
  cat("| Output | Script | Git SHA | Produced |\n")
  cat("|---|---|---|---|\n")
  for (name in sort(names(manifest))) {
    e <- manifest[[name]]
    if (is.null(e$script)) next
    dirty <- if (isTRUE(e$git_dirty)) " (dirty)" else ""
    cat(sprintf("| `%s` | `%s` | `%s`%s | %s |\n", name, e$script,
                e$git_sha, dirty, substr(e$produced_utc, 1, 10)))
  }
} else {
  cat("*No manifest present.*\n")
}
```

Every figure above is produced by `benchmarks/regenerate_events.py`, which
re-runs a job only when the source files that job depends on have changed.
Outputs with no manifest entry are reported as unverified rather than shown as
fact. The ladder outputs are absent from this table because they have not been
produced.
