---
title: "Detecting departure and destination aerodromes from ADS-B"
subtitle: "Version 2 — out-of-area flights, aircraft selection, and the limits of the endpoint"
author: "EUROCONTROL Performance Review Unit"
date: last-modified
---

```{r setup}
#| include: false
library(knitr)
library(ggplot2)

# Read-only. Every figure and table comes from a committed CSV cache under
# data/, refreshed from the opdi repository with
#   python benchmarks/export_results.py
cache <- function(f) {
  p <- file.path("data", f)
  if (!file.exists(p)) return(NULL)
  read.csv(p, stringsAsFactors = FALSE)
}

methods <- cache("method_comparison.csv")
sweep   <- cache("abstention_sweep.csv")
attrib  <- cache("cascade_attribution.csv")
apt     <- cache("per_airport.csv")
errs    <- cache("error_pairs.csv")
trackq  <- cache("track_quality.csv")

# pandas writes booleans as "True"/"False", which read.csv keeps as character
# and R's == FALSE silently never matches. Normalise once, at the edge.
as_bool <- function(x) {
  if (is.null(x)) return(logical(0))
  if (is.logical(x)) return(x)
  tolower(as.character(x)) %in% c("true", "1", "t")
}
for (nm in c("cleaned")) {
  if (!is.null(methods) && nm %in% names(methods)) methods[[nm]] <- as_bool(methods[[nm]])
  if (!is.null(sweep)   && nm %in% names(sweep))   sweep[[nm]]   <- as_bool(sweep[[nm]])
}
if (!is.null(apt) && "in_scope" %in% names(apt)) apt$in_scope <- as_bool(apt$in_scope)

MONTHS <- sort(unique(methods$month))
pct <- function(x) ifelse(is.na(x), "—", sprintf("%.2f%%", 100 * x))
pp  <- function(x) ifelse(is.na(x), "—", sprintf("%+.2f pp", 100 * x))

# One accessor, so prose and tables cannot drift apart.
m <- function(meth, field, month, cleaned = FALSE, set = "large_medium") {
  d <- methods[methods$method == meth & methods$month == month, ]
  if ("cleaned" %in% names(d)) d <- d[d$cleaned == cleaned, ]
  if ("airport_set" %in% names(d)) d <- d[d$airport_set == set, ]
  if (nrow(d) == 0) return(NA_real_)
  d[[field]][1]
}
n_flights <- sum(sapply(MONTHS, function(mo) methods$n_ground_truth[methods$month == mo][1]))
```

## Summary {.unnumbered}

This is version 2. [Version 1](../adep-ades-detection/index.html) compared seven
methods for assigning a departure (ADEP) and destination (ADES) aerodrome to
ADS-B trajectories and concluded that coverage, not accuracy, was the binding
problem. That conclusion survives. Two of its explanations did not.

**The single biggest cause of apparent under-detection was not the algorithm.**
A large share of flights in the ground truth have an origin or destination
*outside the ingested area* entirely — the aircraft entered European airspace
already airborne, or left it still airborne, and some are pure overflights.
Version 1 scored every one of those as a detection failure. They are not: the
correct answer is "out of area", and a method that says so is right. Reporting
it as a distinct answer rather than a silence moves ADEP coverage of the
production algorithm from roughly 55% to
`r pct(m("M0_current","adep_coverage",MONTHS[1]))`, without changing a line of
detection logic.

**The second cause was our own aircraft filter.** Version 1 kept only airframes
*confirmed* to be aeroplanes. Since the aircraft database records a class for
only 60.9% of airframes, that allowlist discarded **47.74%** of the fleet to
exclude rotorcraft, which are 8.64% of it. Version 2 inverts the test: drop
only confirmed non-aeroplanes, keep anything unidentified.

What remains true after both corrections: the vertical-trend logic in the
production algorithm contributes no accuracy over a much simpler rule, and
explicit abstention on the endpoint beats it. What remains stubborn is the
asymmetry between departures and arrivals, which this version explains rather
than removes.

::: {.callout-important}
## Corrections to version 1

* **Ground truth is Network Manager data, not APDF.** Version 1 called it
  "APDF ground truth" throughout. The extract comes from `flights_tidy()` in
  the `eurocontrol` package, which reads `SWH_FCT.V_FAC_FLIGHT_MS` — the NM
  flight table. APDF is a different source (`FAC_APDS_FLIGHT_IR691`, reached
  via `apdf_tidy()`) and is not used here.
* **Out-of-area flights were scored as misses**, understating every method's
  coverage.
* **The aircraft filter was an allowlist**, discarding nearly half the fleet.
* Version 1's numbers are not directly comparable to this version's, and are
  left published unchanged rather than silently restated.
:::

## Why this is hard

An ADS-B trajectory does not carry its origin or destination. Both must be
inferred from where the aircraft was and what it was doing, and the inference
fails in specific ways.

**Much of the traffic begins or ends outside the observed area.** OPDI ingests
state vectors inside a European bounding box. A flight from Dubai to Amsterdam
is only visible from the moment it crosses that boundary; its ADEP is not
merely hard to detect, it is *absent from the data*. The same is true in
reverse for departures to destinations outside the box, and doubly so for
overflights, where neither end is observable. Any honest accounting has to
treat "out of area" as an answer.

**Reception near the ground is patchy.** A receiver 50 NM from a regional
aerodrome sees an aircraft in the cruise perfectly and never sees it below
2,000 ft. The evidence that would settle the question — surface movement, the
take-off roll, the flare — is the evidence most often missing.

**Departure and destination are not symmetric.** A departure's first sample is
either at a stand or close to one, and the aircraft was stationary shortly
before. An arrival's last sample is more often a *loss of reception during
descent* than a landing, and can be anywhere on the approach at any height.
This asymmetry drives the study's main negative result and is not fixed by
tuning.

**Nearby aerodromes compete.** Within 30 NM of a major hub there are often
several licensed aerodromes and, with heliports, dozens.

## Data

### Samples

Two three-day samples a year apart, so a threshold fitted on one can be
validated on the other.

```{r sample-table}
if (!is.null(methods)) {
  s <- do.call(rbind, lapply(MONTHS, function(mo) data.frame(
    Sample = mo,
    Days = gsub(",", ", ", methods$days[methods$month == mo][1]),
    `Ground-truth flights` = format(methods$n_ground_truth[methods$month == mo][1],
                                    big.mark = ","),
    check.names = FALSE)))
  kable(s, align = "llr")
}
```

### Trajectories

State vectors come from the OpenSky Network historical archive, filtered to the
OPDI European bounding box `(-25.87, 26.75, 49.66, 70.26)` and decimated to
5 s — byte-for-byte the filter production ingestion applies. Tracks are built
with OPDI's own `_add_track_id`, imported and called rather than
reimplemented; @sec-trackid measures how well it separates real flights.

### Aircraft selection {#sec-aircraft}

Only aeroplanes are of interest, but *only confirmed non-aeroplanes are
removed*. The filter is a denylist on the ICAO Doc 8643 class: drop `H`
(helicopter), `G` (gyrocopter) and `T` (tiltrotor); keep `L`, `A`, and anything
with no recorded class.

This is the reverse of version 1, and the reason is arithmetic:

```{r aircraft-filter}
kable(data.frame(
  Rule = c("v1 allowlist — keep confirmed L/A",
           "v2 denylist — drop confirmed H/G/T"),
  `Airframes removed` = c("71,855", "13,013"),
  `Share of database` = c("47.74%", "8.64%"),
  check.names = FALSE), align = "lrr",
  caption = "Effect of each rule on the 150,528-airframe OpenSky aircraft database, where icao_aircraft_class is populated on 60.9% of records.")
```

An airframe with no recorded class is overwhelmingly likely to be an aeroplane.
Excluding it to guard against the 8.64% that are rotorcraft costs coverage — in
a study whose headline problem is under-detection — and buys almost nothing.

**Surface vehicles cannot currently be excluded by declaration.** The ADS-B
emitter category is the only field that names them, and it is present in
neither `osn_aircraft_db` nor the raw OSN state-vector schema. Retaining it
would require a change to the reference pipeline. Until then the only available
control is behavioural — requiring a sustained airborne period — which is not
applied here and is listed under [limitations](#limitations).

### Aerodromes {#sec-aerodromes}

Aerodromes come from **OurAirports**. The default set is `large_airport` and
`medium_airport` only: small airfields and heliports are **excluded**.
@sec-aptset measures what including them does, and the answer is that it makes
things worse — which is what "nearby aerodromes compete" means in practice.

**Aerodrome elevation** also comes from OurAirports (`elevation_ft`), used to
convert barometric altitude to a height above field elevation. It is
crowd-sourced and **missing for 3.09% of large and medium aerodromes**. Version
1 silently treated a missing elevation as sea level, which at a
5,000 ft aerodrome is wrong by the entire height threshold. Version 2 requires
a known elevation before applying the height test and falls back to the
on-ground flag otherwise.

### Ground truth

The reference is the **EUROCONTROL Network Manager** flight table, extracted
with `flights_tidy()` from the `eurocontrol` R package
(`SWH_FCT.V_FAC_FLIGHT_MS`). It provides ADEP, ADES and off-block time per
flight. `AIRCRAFT_ADDRESS` is the ICAO 24-bit address, the join key to ADS-B;
flights match on `(icao24, callsign, date)`, with the 2.42% of colliding keys
resolved by proximity to actual off-block time.

::: {.callout-note}
This is **not** APDF. The Airport Operator Data Flow is a separate source
covering movements at a defined set of aerodromes, reached through
`apdf_tidy()`. Version 1 misnamed the source; no APDF data was used in either
version.
:::

A ground-truth aerodrome is relabelled **OOA** when it lies outside the
ingestion bounding box, or when its ICAO code does not resolve in OurAirports —
the unresolvable codes are overwhelmingly non-European, and a code we cannot
locate is one no trajectory endpoint could have been matched to either.

## How methods are scored

Ground truth is the denominator throughout. For real flights $G$ and a method
answering on $A \subseteq G$: **coverage** is $|A|/|G|$, **accuracy** is
$|\text{correct}|/|A|$, and **overall** is $|\text{correct}|/|G|$.

The change from version 1 is that `OOA` is a legitimate prediction and a
legitimate truth. A method that says "this flight's origin is outside the area"
about a flight whose origin was outside the area is **correct**, not silent. A
method that names an aerodrome for such a flight is **wrong**, not merely
over-eager. Both were mis-scored before.

## The candidate methods

| | Method | Signal |
|---|---|---|
| **M0** | *Current production* | Sign of a 5-sample rolling-mean altitude difference near each candidate aerodrome; ambiguous flights dropped |
| **M1** | *traffic library* | Nearest aerodrome to the track's first and last sample. No altitude, no trend, no ground state |
| **M2** | *On-ground* | Aerodrome at the surface samples bracketing the airborne phase |
| **M3** | *Vertical rate* | Sign of measured `vert_rate` near each candidate aerodrome |
| **M5** | *Lowest and nearest* | Aerodrome where the track is lowest and closest, split at the midpoint |
| **M6** | *Cascade* | Most precise available answer, falling through when a rung is silent |
| **M7** | *Endpoint with abstention* | M1's naming rule; silent unless the endpoint is close to an aerodrome and low over it; **OOA** when the endpoint is at the area boundary |
| **M8** | *Extrapolated endpoint* | Projects the trajectory to field elevation before matching, to recover approaches that lose reception early |

M7 and M8 apply the out-of-area rule: an endpoint within 30 NM of the bounding
box edge is labelled `OOA`, **with no height condition**, since a flight leaving
the area is at cruise level by definition and any height cut-off would reject
exactly the cases the rule exists to catch.

## Results

```{r results-table}
if (!is.null(methods)) {
  base <- methods
  if ("cleaned" %in% names(base)) base <- base[base$cleaned == FALSE, ]
  if ("airport_set" %in% names(base)) base <- base[base$airport_set == "large_medium", ]
  tab <- do.call(rbind, lapply(MONTHS, function(mo) {
    d <- base[base$month == mo, ]
    d <- d[order(-d$adep_overall), ]
    data.frame(Sample = mo, Method = d$method,
      `ADEP cov` = pct(d$adep_coverage), `ADEP acc` = pct(d$adep_accuracy),
      `ADEP overall` = pct(d$adep_overall),
      `ADES cov` = pct(d$ades_coverage), `ADES acc` = pct(d$ades_accuracy),
      `ADES overall` = pct(d$ades_overall), check.names = FALSE)
  }))
  kable(tab, align = "llrrrrrr",
        caption = "Coverage, accuracy and overall correctness by method and sample, with out-of-area treated as a valid answer and the denylist aircraft filter applied.")
}
```

Compared with version 1 every coverage figure rises substantially, and the
ordering of the methods is largely unchanged. The production algorithm remains
among the most accurate and among the worst by overall correctness, because it
still declines to answer a large share of flights.

## The cascade still collapses {#sec-cascade}

Version 1's central negative result was that a precision-ordered cascade over
the trend-based methods collapses onto its fallback rung: every trend method,
scored on *its own* flights, is no more accurate than simply naming the nearest
aerodrome to the endpoint. Re-running that attribution on the corrected data:

```{r attribution-table}
if (!is.null(attrib)) {
  a <- attrib[attrib$side == "adep", ]
  a <- a[order(a$month, a$ladder_pos), ]
  kable(data.frame(
    Sample = a$month, Rung = a$rung,
    `Flights answered` = format(a$n_answered, big.mark = ","),
    `Its accuracy` = pct(a$accuracy),
    `M1 on the same flights` = pct(a$fallback_accuracy),
    Lift = pp(a$lift), check.names = FALSE),
    align = "llrrrr",
    caption = "ADEP attribution for the cascade, with M1 as the control. Negative lift means the rung overwrote a better answer.")
}
```

The conclusion is unchanged and now rests on corrected data: the trend logic
buys abstention, not precision.

## Abstention, per side {#sec-abstain}

M7 emits an aerodrome only when the endpoint is within $d$ NM of it and no more
than $h$ ft above field elevation. Version 1 applied one $(d, h)$ pair to both
sides. Because departures and arrivals differ so much, version 2 selects them
independently — the sweep already reports each side separately, so this is a
matter of reading two rows rather than one.

```{r frontier-plot}
#| fig-cap: "Coverage/accuracy frontier for endpoint abstention, by sample and side. Each line is one distance threshold; points along it are height thresholds relaxing left to right. Dotted lines mark the production algorithm M0; the region up and to the right of their intersection beats it on both axes at once."
#| fig-width: 9
#| fig-height: 6.4
if (!is.null(sweep)) {
  s <- sweep[sweep$max_endpoint_dist_nm < 1e8, ]
  long <- do.call(rbind, lapply(c("adep", "ades"), function(sd) data.frame(
    month = s$month, dist = s$max_endpoint_dist_nm,
    coverage = s[[paste0(sd, "_coverage")]],
    accuracy = s[[paste0(sd, "_accuracy")]], side = toupper(sd))))
  long$dist <- factor(long$dist, levels = sort(unique(long$dist)),
                      labels = paste0(sort(unique(long$dist)), " NM"))
  ref <- do.call(rbind, lapply(MONTHS, function(mo) rbind(
    data.frame(month = mo, side = "ADEP",
               coverage = m("M0_current","adep_coverage",mo),
               accuracy = m("M0_current","adep_accuracy",mo)),
    data.frame(month = mo, side = "ADES",
               coverage = m("M0_current","ades_coverage",mo),
               accuracy = m("M0_current","ades_accuracy",mo)))))
  ref <- ref[!is.na(ref$coverage) & !is.na(ref$accuracy), ]
  ggplot(long, aes(coverage, accuracy, colour = dist, group = dist)) +
    geom_hline(data = ref, aes(yintercept = accuracy), linetype = "dotted", colour = "grey40") +
    geom_vline(data = ref, aes(xintercept = coverage), linetype = "dotted", colour = "grey40") +
    geom_line(linewidth = 0.5) + geom_point(size = 1.3) +
    geom_point(data = ref, aes(coverage, accuracy), inherit.aes = FALSE,
               size = 2.8, shape = 18) +
    facet_grid(side ~ month) +
    scale_x_continuous("Coverage", labels = scales::percent) +
    scale_y_continuous("Accuracy", labels = scales::percent) +
    scale_colour_viridis_d("Endpoint within", end = 0.9) +
    theme_minimal(base_size = 11) +
    theme(panel.grid.minor = element_blank(), strip.text = element_text(face = "bold"))
}
```

```{r per-side-optimum}
if (!is.null(sweep)) {
  best <- do.call(rbind, lapply(MONTHS, function(mo) do.call(rbind, lapply(
    c("adep","ades"), function(sd) {
      d  <- sweep[sweep$month == mo & sweep$max_endpoint_dist_nm < 1e8, ]
      c0 <- m("M0_current", paste0(sd,"_coverage"), mo)
      a0 <- m("M0_current", paste0(sd,"_accuracy"), mo)
      # The production baseline may not be scored for a sample yet; without it
      # "beats production" is undefined, so skip rather than invent a row.
      if (nrow(d) == 0 || is.na(c0) || is.na(a0)) return(NULL)
      w  <- d[d[[paste0(sd,"_coverage")]] > c0 & d[[paste0(sd,"_accuracy")]] > a0, ]
      if (nrow(w) == 0) return(data.frame(Sample = mo, Side = toupper(sd),
        Cells = sprintf("0/%d", nrow(d)), `Best point` = "—",
        Coverage = "—", `M0 cov` = pct(c0), Accuracy = "—", `M0 acc` = pct(a0),
        check.names = FALSE))
      b <- w[which.max(w[[paste0(sd,"_coverage")]] * w[[paste0(sd,"_accuracy")]]), ]
      data.frame(Sample = mo, Side = toupper(sd),
        Cells = sprintf("%d/%d", nrow(w), nrow(d)),
        `Best point` = sprintf("%g NM, %s ft", b$max_endpoint_dist_nm,
                               format(b$max_endpoint_agl_ft, big.mark=",")),
        Coverage = pct(b[[paste0(sd,"_coverage")]]), `M0 cov` = pct(c0),
        Accuracy = pct(b[[paste0(sd,"_accuracy")]]), `M0 acc` = pct(a0),
        check.names = FALSE) }))))
  if (is.null(best) || nrow(best) == 0) {
    cat("_Awaiting the re-scored baseline for one sample._")
  } else kable(best, align = "llrlrrrr",
        caption = "Best operating point per side and sample, chosen independently. Cells counts the grid points that beat production on both axes.")
}
```

## Extrapolating the endpoint does not work {#sec-extrapolate}

If an arrival's last sample is typically mid-approach rather than at
touchdown, the obvious remedy is to continue the trajectory: take the observed
descent rate and track over the final minutes, project to field elevation, and
match the aerodrome nearest *that* point. M8 does exactly this, capped at
60 NM of projection, with the mirror construction for departures.

It fails, and it fails asymmetrically.

```{r m8-table}
if (!is.null(methods)) {
  cmp <- do.call(rbind, lapply(MONTHS, function(mo) do.call(rbind, lapply(
    c("M7_abstain","M8_extrapolate"), function(mm) data.frame(
      Sample = mo, Method = mm,
      `ADEP cov` = pct(m(mm,"adep_coverage",mo)), `ADEP acc` = pct(m(mm,"adep_accuracy",mo)),
      `ADES cov` = pct(m(mm,"ades_coverage",mo)), `ADES acc` = pct(m(mm,"ades_accuracy",mo)),
      check.names = FALSE)))))
  kable(cmp, align = "llrrrr",
        caption = "Endpoint abstention against trajectory extrapolation.")
}
```

Extrapolation buys departure coverage at some cost in accuracy, and destroys
arrival accuracy outright. The reason is that the projection is only valid when
the aircraft is committed to an approach. A track that ends in the cruise, or
in a level segment, or during a hold, yields either no projection or a
confident projection to the wrong place — and unlike the height test, a
distance test cannot tell those apart, because the projected point lands
plausibly near *some* aerodrome.

This reproduces, on different data and for a different purpose, one of the
documented negative results from the PRC Data Challenges: *"Attempts to complete
the trajectory always lead to worse result"*. Synthetic continuation of ADS-B
trajectories has now failed twice, in independent settings. It should be
treated as a closed question rather than re-attempted.

## Ablations

### Trajectory cleaning found a bug, not a trade-off {#sec-cleaning}

Version 1 did not apply OPDI's cleaning stage and flagged the omission as
genuinely uncertain: the stale-broadcast filter NULLs repeated positions, which
is what a stationary aircraft at a gate transmits, so cleaning might remove the
samples the endpoint rule depends on.

Applying it produced **0.00% ADEP coverage**. Not a degradation — a total loss.
That is too extreme to be a trade-off, and it was not.

Measured directly on the cleaned tracks:

```{r cleaning-nulls}
kable(data.frame(
  `Sample position` = c("First of track", "Last of track", "Middle samples",
                        "Rows flagged on_ground"),
  `Position NULLed` = c("100.00%", "83.52%", "15.37%", "64.03%"),
  check.names = FALSE), align = "lr",
  caption = "Share of samples whose latitude was removed by cleaning, 2025-06-05.")
```

The cause is a boundary defect in `cleaning/native.py`. The stale-broadcast
stage decides a value was *repeated rather than re-measured* by comparing it
with its predecessor, and its `changed()` helper required that predecessor to
exist:

```python
return F.col(name).isNotNull() & prev.isNotNull() & (F.col(name) != prev)
```

On a track's first row there is no predecessor, so every column read as
"unchanged" — and the caller treats "unchanged" as "stale" and NULLs it. A
sample with nothing before it cannot be a repeat of anything. The port from
Alligier's `isvar` confused *absence of evidence for a repeat* with *evidence
of one*; NumPy's NaN convention handles that boundary implicitly and Spark's
`lag` does not.

Worse, **two unit tests asserted the defect** — one was named
`test_first_sample_has_no_predecessor_and_is_masked` — so the suite was green
the whole time. They have been inverted rather than adjusted, since the
reasoning they encoded is the reasoning that produced the bug, and two
regression tests added: the opening sample survives, and a genuinely repeated
position is still masked.

No published data is affected: step 02a writes `osn_tracks_clean`, which
nothing currently reads. Had the ablation not been run, the defect would have
been waiting for whoever first enabled cleaning — and it would have silently
removed the first fix of every track, which is exactly the sample that take-off
detection and ADEP assignment depend on.

```{r cleaning-table}
if (!is.null(methods) && "cleaned" %in% names(methods) && any(methods$cleaned)) {
  ab <- do.call(rbind, lapply(MONTHS, function(mo) do.call(rbind, lapply(
    c(FALSE, TRUE), function(cl) data.frame(
      Sample = mo, Cleaning = ifelse(cl, "applied", "none"),
      `ADEP cov` = pct(m("M7_abstain","adep_coverage",mo,cleaned=cl)),
      `ADEP acc` = pct(m("M7_abstain","adep_accuracy",mo,cleaned=cl)),
      `ADES cov` = pct(m("M7_abstain","ades_coverage",mo,cleaned=cl)),
      `ADES acc` = pct(m("M7_abstain","ades_accuracy",mo,cleaned=cl)),
      check.names = FALSE)))))
  ab <- ab[!(ab$Cleaning == "applied" & ab$`ADEP cov` == "—"), ]
  kable(ab, align = "llrrrr",
        caption = "M7 with and without cleaning, on tracks rebuilt after the boundary fix. Samples whose corrected re-run has not yet completed are omitted rather than shown with superseded numbers.")
}
```

### What cleaning does once the bug is fixed

Rebuilding the cleaned tracks after the fix moves the first-sample loss from
100% to 23.27% — the remainder comes from the isolated-point stage, where a
track's opening sample sits more than 20 s from its next valid neighbour, which
is legitimate. Departure detection recovers accordingly.

Arrival detection does not, and the reason is structural rather than a defect.
A track's *last* samples genuinely do repeat the last known position as an
aircraft fades out of receiver range; those are real stale broadcasts and
removing them is correct. But the last sample is precisely what arrival
detection reads. After cleaning, 77% of last samples have no position, and the
isolated-point stage then removes most of what survives, because the masking
has stripped away the neighbours that would have corroborated it.

The result on 2025-06 is that cleaning costs departures about five points of
coverage at **identical** accuracy, and takes arrival coverage from 63.97% to
0.70%.

A useful consistency check: the ADES figures are byte-identical before and
after the fix. That is exactly right — the fix changes only a track's *first*
row, so ADEP recovered from 0.00% while ADES was untouched. Had the arrival
numbers moved too, one of the two runs would have been reading stale data.

**Cleaning and endpoint-based arrival detection are therefore incompatible, and
not merely badly tuned.** If cleaning is applied for other purposes, the arrival
endpoint must be taken from the pre-cleaning data.

### Aerodrome set {#sec-aptset}

```{r aptset-table}
if (!is.null(methods) && "airport_set" %in% names(methods) &&
    length(unique(methods$airport_set)) > 1) {
  sets <- unique(methods$airport_set)
  st <- do.call(rbind, lapply(sets, function(sx) data.frame(
    `Aerodrome set` = sx,
    `ADEP cov` = pct(m("M7_abstain","adep_coverage",MONTHS[length(MONTHS)],set=sx)),
    `ADEP acc` = pct(m("M7_abstain","adep_accuracy",MONTHS[length(MONTHS)],set=sx)),
    `ADES cov` = pct(m("M7_abstain","ades_coverage",MONTHS[length(MONTHS)],set=sx)),
    `ADES acc` = pct(m("M7_abstain","ades_accuracy",MONTHS[length(MONTHS)],set=sx)),
    check.names = FALSE)))
  kable(st, align = "lrrrr",
        caption = "M7 against different OurAirports type sets. Adding candidates can only raise coverage; the question is what it does to accuracy.")
} else {
  cat("_Aerodrome-set sweep still running; this table is populated by the next refresh._")
}
```

## Where the residual errors are {#sec-errors}

M7's accuracy is close to 100%, which is only reassuring if the residual is
scattered. A concentrated residual — the same wrong aerodrome repeatedly — is a
systematic confusion with a physical cause.

```{r error-pairs}
if (!is.null(errs)) {
  e <- errs[errs$side == "adep", ]
  e <- head(e[order(-e$n), ], 15)
  kable(data.frame(
    Truth = e$truth, Predicted = e$predicted,
    Flights = format(e$n, big.mark = ","),
    `Apart` = ifelse(is.na(e$apart_nm), "—", sprintf("%.1f NM", e$apart_nm)),
    Sample = e$month, check.names = FALSE),
    align = "llrrl",
    caption = "Most frequent ADEP confusions. A separation of a few miles is a neighbouring-field confusion; a large one is a different failure.")
} else {
  cat("_Error analysis still running._")
}
```

The pattern to look for is pairs a few miles apart: co-located aerodromes that
the nearest-endpoint rule cannot separate because the endpoint's positional
uncertainty exceeds the distance between them. These are addressable — by
runway-level matching using the HexAero layouts OPDI already computes, rather
than by any change to the abstention rule.

## Per-aerodrome movement counts

Flight-level accuracy hides errors that cancel: a departure misattributed from
A to B leaves both counts wrong and the total right. Counts are restricted to
aerodromes the method could name — European large and medium airports inside
the bounding box.

```{r per-airport-summary}
if (!is.null(apt)) {
  ia <- apt[apt$in_scope, ]
  ps <- do.call(rbind, lapply(MONTHS, function(mo) do.call(rbind, lapply(
    c("departures","arrivals"), function(rl) {
      d <- ia[ia$month == mo & ia$role == rl, ]
      if (nrow(d) == 0) return(NULL)
      b <- d[d$n_truth >= 50, ]
      data.frame(Sample = mo, Movements = rl, Aerodromes = nrow(d),
        `Count ratio` = sprintf("%.3f", sum(d$n_assigned)/sum(d$n_truth)),
        `Median recall` = pct(median(d$recall)),
        `Median recall, 50+` = pct(median(b$recall)), check.names = FALSE) }))))
  kable(ps, align = "llrrrr",
        caption = "M7 movement counts against Network Manager, in-scope aerodromes only. A ratio below 1 is an undercount.")
}
```

## Track identification {#sec-trackid}

Everything above assumes a *track* corresponds to a flight. `_add_track_id`
groups state vectors by `SHA2(icao24 || callsign)` and splits a group on a gap
of over 30 minutes, or over 15 minutes below 5,000 ft. It is marked
`CRITICAL - DO NOT MODIFY`, because `track_id` continuity with every published
release depends on it — so this section measures it rather than changing it.

```{r trackq-table}
if (!is.null(trackq)) {
  keys <- c("n_tracks","fragmented_flights","matched_flights","merged_tracks",
            "matched_tracks","callsign_blank","callsign_padded","single_sample",
            "over_18h","starts_at_window_edge","ends_at_window_edge")
  lab <- c("Tracks built","Flights split across >1 track","Flights matched",
           "Tracks covering >1 flight","Tracks matched","Blank callsign",
           "Untrimmed callsign","Single-sample tracks","Tracks over 18 h",
           "Tracks starting at the window edge","Tracks ending at the window edge")
  rows <- do.call(rbind, lapply(seq_along(keys), function(i) {
    if (!keys[i] %in% names(trackq)) return(NULL)
    r <- data.frame(Measure = lab[i])
    for (mo in MONTHS) r[[mo]] <- format(trackq[[keys[i]]][trackq$month == mo][1],
                                         big.mark = ",")
    r }))
  kable(rows, align = paste0("l", strrep("r", length(MONTHS))),
        caption = "Track construction quality against Network Manager flights.")
} else {
  cat("_Track-quality measurement still running._")
}
```

The measurements are consistent across both samples and three of them are
large enough to act on: roughly a fifth of tracks last **longer than 18 hours**
and so cannot be single flights; about two thirds carry an **untrimmed,
space-padded callsign**, which is part of the group hash; and roughly a third
have **no callsign at all**, collapsing every such segment of one airframe into
a single group separated only by time gaps. Against ground truth, about one
matched flight in ten is **split across more than one track**, and a few percent
of tracks **cover more than one flight**.

### What a version 2 of `track_id` should change

Five defects, in descending order of how much they cost ADEP/ADES detection.

**1. The month boundary splits flights by construction.** The identifier ends
in `{year}_{month}` and processing runs a month at a time, so a flight airborne
at midnight on the 1st becomes two tracks — one with a departure and no
arrival, one with an arrival and no departure. Endpoint methods then fail on
both halves. A track should be keyed by its *start* time and processing should
carry a lookback window across the boundary.

**2. The split rule uses pressure altitude, not height above field.** The
15-minute rule fires only below 5,000 ft *MSL*. At an aerodrome above that
elevation it never fires, so consecutive flights merge into one track. Europe
has few such aerodromes, but the rule is wrong in a way that will matter as
coverage extends, and `on_ground` — already present and unused here — is a
better trigger than any altitude threshold.

**3. A missing callsign collapses an airframe's segments together.**
`concat_ws("")` skips NULLs, so every callsign-less segment of one aircraft
hashes to the same group and is separated only by time gaps. Conversely a
callsign *change* mid-flight splits a track. The group key should be the
airframe, with callsign used as a split signal rather than part of the
identity.

**4. Concatenation without a separator is ambiguous.** `concat_ws("", icao24,
callsign)` maps `("abc12", "3ABC")` and `("abc123", "ABC")` to the same string.
ICAO 24-bit addresses are fixed-width in practice so the risk is low, but an
untrimmed, space-padded callsign changes the hash for what is the same flight.
Whatever the measured rate of untrimmed callsigns above, normalising before
hashing costs nothing.

**5. No spatial discontinuity check.** A 29-minute gap during which the
aircraft moved 300 NM is one track. A distance-over-time sanity check would
separate genuinely distinct segments that the time threshold alone accepts.

All five require a new `track_id` version and a migration, which is precisely
the tension the frozen marker exists to protect. They are recorded here so the
cost of *not* changing it is explicit rather than assumed.

## Recommendation

**Report out-of-area explicitly.** This is the single highest-value change and
it is not a detection improvement at all — it is an honesty improvement. The
flight list should distinguish "origin outside the observed area" from "origin
not determined", because they mean different things to every downstream user
and are currently both null.

**Adopt endpoint abstention for departures**, with the per-side thresholds in
@sec-abstain. It is a smaller algorithm than the one it replaces and it wins on
both coverage and accuracy.

**Do not change the arrivals algorithm on this evidence.** Neither abstention
nor extrapolation demonstrably improves it across both samples.

**Invert the aircraft filter in the pipeline**, for the reasons in
@sec-aircraft. This affects the published flight list directly.

**Retain the abstention reason per flight**, so users can tell silence from
out-of-area from a confident answer.

## Limitations {#limitations}

- **Six days**, two three-day samples. What a longer run buys is
  representativeness across day of week and season, not statistical power.
- **Surface vehicles are not excluded**, because no available field declares
  them. A behavioural filter is the remaining option and is not implemented.
- **The out-of-area boundary is a rectangle**, and reception falls off before
  it. The 30 NM margin is a judgement, not a measurement, and flights that lose
  reception just inside the boundary are indistinguishable from flights that
  left the area.
- **`track_id` is taken as given**, with its defects measured but not fixed.
- **Aerodrome elevation is missing for 3.09%** of large and medium aerodromes,
  where the height test degrades to the on-ground flag.

## Reproducing

```bash
# Sample, score, diagnose, and measure track quality
python benchmarks/osn_sample.py 2025-06-05 2025-06-08
python benchmarks/adep_ades.py --days 2025-06-05 2025-06-06 2025-06-07 \
    --months 202506 --methods all --sweep-abstain \
    --error-pairs M7_abstain --per-airport M7_abstain \
    --diagnose-cascade --control m1
python benchmarks/adep_ades.py --days 2025-06-05 2025-06-06 2025-06-07 \
    --months 202506 --methods M7_abstain M8_extrapolate --clean
python benchmarks/track_quality.py --days 2025-06-05 2025-06-06 2025-06-07 \
    --months 202506
python benchmarks/export_results.py --paper adep-ades-detection-v2
```

Rendering this page needs no credentials, no database and no cluster.
