---
title: "Flight events from ADS-B: fixing what OPDI publishes, and reaching the GANP KPIs"
subtitle: "Version 1"
date: last-modified
author: "EUROCONTROL Performance Review Unit — Open Performance Data Initiative"
abstract: |
  OPDI has published flight events since v0.0.2 without a single test or a
  single comparison against reference data. This study audits the four
  detectors behind them, finds nine defects — one of which silently suppresses
  every take-off and landing at any aerodrome above roughly 200 ft — fixes
  them, and adds four milestone families chosen because they are what the ICAO
  GANP key performance indicators actually require. It reports what has been
  measured and, equally, what has not.
format:
  html:
    toc: true
  pdf:
    toc: true
---

```{r regenerate}
#| include: false
#| cache: false

# Rendering this page runs the analysis behind it, as the version 7 report
# does. `regenerate_events.py` is the executable definition of the study:
# every figure below is produced by exactly one job it declares, and it
# re-runs a job only when the code that job depends on has changed since the
# output was written. A render with everything current is a fast no-op.
#
# OPDI_RENDER=check fails fast on staleness; allow-stale renders a draft from
# whatever is staged, which the provenance table then shows as unverified.
REPO   <- normalizePath(file.path(getwd(), "..", "..", "..", "opdi"), mustWork = FALSE)
PYTHON <- Sys.getenv("OPDI_PYTHON", file.path(REPO, ".venv310", "bin", "python"))
MODE   <- Sys.getenv("OPDI_RENDER", "allow-stale")

if (MODE %in% c("run", "check") && file.exists(PYTHON)) {
  args <- c(file.path(REPO, "benchmarks", "regenerate_events.py"))
  if (MODE == "check") args <- c(args, "--check")
  status <- system2(PYTHON, args, stdout = "", stderr = "")
  if (MODE == "check" && status != 0) {
    stop("regenerate_events.py --check reports stale outputs")
  }
}

read_json_if <- function(name) {
  path <- file.path("data", name)
  if (file.exists(path)) jsonlite::fromJSON(path) else NULL
}

read_csv_if <- function(name) {
  path <- file.path("data", name)
  if (file.exists(path)) read.csv(path, stringsAsFactors = FALSE) else NULL
}

bridge_2025 <- read_json_if("bridge_2025.json")
bridge_2024 <- read_json_if("bridge_2024.json")
manifest    <- read_json_if("_manifest.json")
ladder_2025 <- read_csv_if("ladder_2025.csv")
ladder_2024 <- read_csv_if("ladder_2024.csv")
inv_2025    <- read_csv_if("inventory_2025.csv")
inv_2024    <- read_csv_if("inventory_2024.csv")
```

## What this study is

Two things prompted it.

**The events were never checked.** OPDI's flight-event table is the
least-examined thing the project publishes. Departure and destination
aerodrome detection has been through seven successive studies and a full
rewrite; the event detectors had no tests at all, and no number in them had
ever been compared against anything.

**The event vocabulary did not reach the indicators.** The stated goal is to
compute as many ICAO GANP key performance indicators as ADS-B allows. Mapping
the published KPI list against what OPDI emitted showed the gap concentrated in
four milestone families — and that ICAO itself names ADS-B data providers as a
feed for two indicators OPDI could not compute at all.

## What was wrong

Nine defects, found by reading the detectors. None of them fails. Each produces
fewer events, or events displaced in one direction — the failure mode that
survives an eyeball check of the totals.

| | Defect | Consequence |
|---|---|---|
| D1 | Ground membership measured against **uncorrected pressure altitude** | The membership reaches zero at 200 ft *pressure* altitude, so at any aerodrome above roughly 200 ft AMSL — or anywhere on a low-pressure day — no sample is ever classified as on-ground. Since take-off requires a preceding ground sample and landing requires a ground sample, **neither event has ever been emitted for those flights.** The loss is biased against high-elevation aerodromes. |
| D2 | The reference implementation's 60-second majority smoothing was **not ported** | Phases were decided per state vector at 5-second spacing with no temporal aggregation, so one misclassified sample injects a spurious level segment or destroys a take-off. |
| D3 | Cruise speed membership centred on 600 kt | A turboprop at 250 kt scores 0.002, so it never reaches cruise and gets no top-of-climb or top-of-descent. **Inherited from OpenAP, not introduced here** — left unchanged pending measurement. |
| D4 | `least`/`greatest` skip nulls | A rule missing an input became a minimum over fewer terms, which can only *raise* its activation — so incomplete evidence out-competed complete evidence. |
| D5 | `first`/`last` inside a group-by | Take partition order, not time order, so a reported entry position was not guaranteed to be the position at the reported entry time. |
| D6 | Events read the **raw** track table | Trajectory cleaning reached the flight list and nothing else, so no published event has ever been derived from a cleaned trajectory. |
| D7 | Non-reproducible event identifiers; one version string for every type | The same event got a different identifier on every run, so two runs of a month could not be reconciled. |
| D8 | Flight-level crossings kept **only the first and last** per level, uninterpolated | A flight that levels off and re-climbs through a level lost every crossing in between — exactly the crossings a vertical-efficiency indicator is about — and every reported instant was biased by up to one sample interval in the same direction. |
| D9 | No tests | Not one detector was covered. |

D1 and D2 are the two that matter most, because they mean the published
take-off, landing, top-of-climb and top-of-descent events are missing
**non-randomly**.

## What was built

Eight defects are fixed and the ninth (D3) is left in place deliberately,
because changing a constant inherited from the reference implementation should
follow a measurement rather than precede one. Four milestone families were
added:

- **Threshold crossings**, rewritten. Every crossing rather than the first and
  last, with a hysteresis dead band so an aircraft cruising *at* a level emits
  none, and the instant interpolated to the exact threshold.
- **ASMA ring crossings** at 40 and 100 NM, built around the flight's own
  origin and destination — which is what the indicators mean, and what the
  reference data records.
- **ICAO level segments**, implementing the published algorithm rather than
  reusing the fuzzy-phase level events. The two are **not interchangeable**.
- **Runway identification with ATOT and ALDT**, and **off-block/on-block**.

Every change is reversible: a `legacy()` configuration reproduces published
`events_v0.0.2` exactly, and the new algorithm carries a new version string.

## Which KPIs this reaches

| KPI | Title | Reachable from ADS-B? |
|---|---|---|
| KPI02 | Taxi-out additional time | Yes |
| KPI05 | Actual en-route extension | Yes — ICAO names ADS-B as the feed |
| KPI08 | Additional time in terminal airspace | Yes |
| KPI10 | Airport peak throughput | Yes |
| KPI13 | Taxi-in additional time | Yes, coverage-limited |
| KPI15 | Flight time variability | Yes |
| KPI17 | Level-off during climb | Yes |
| KPI18 | Level capping during cruise | Yes (variant 2) |
| KPI19 | Level-off during descent | Yes |
| KPI01, KPI14 | Departure / arrival punctuality | Actual half only — the scheduled half needs a schedule feed |
| KPI11 | Airport throughput efficiency | Numerator only |
| Others | ATFM, capacity, flight plan, fuel, safety, noise | No |

Nine fully and three partially, from four milestone families.

## Ground truth, and its ceiling

Reference data is EUROCONTROL APDF. It is in movement form — there is no
literal take-off-time column, and `SRC_PHASE` decides what a row carries — and
crucially it has **no aircraft address**, so it cannot be joined to ADS-B
directly. The route runs through the flight table on an internal key.

That bridge is the ceiling on every coverage figure in this study: a milestone
that cannot be reached is indistinguishable from one that was missed, unless
the ceiling is known. So it is measured first.

```{r tbl-bridge}
#| echo: false
#| results: asis
if (!is.null(bridge_2025) && !is.null(bridge_2024)) {
  rows <- list(list("2024-06", bridge_2024), list("2025-06", bridge_2025))
  cat("| Period | APDF movements | Reaching `icao24` | Share |\n")
  cat("|---|---:|---:|---:|\n")
  for (r in rows) {
    b <- r[[2]]
    cat(sprintf("| %s | %s | %s | **%s%%** |\n", r[[1]],
                format(b$movements, big.mark = ","),
                format(b$reached_icao24, big.mark = ","), b$reach_pct))
  }
} else {
  cat("*Bridge measurement not present in the committed data.*\n")
}
```

Both periods clear 93%, so the ceiling is high enough not to dominate any
result. The residual loss is roughly three per cent of movements carrying no
internal key, and a further two per cent whose key has no counterpart in the
flight table.

Two traps in this join are worth recording, because both fail silently:

**The reference carries the commercial flight number, ADS-B carries the ATC
callsign.** The two disagree on 20.4% of rows — `LH416` against `DLH416` — and
joining on the wrong one discards a fifth of the sample while looking exactly
like a detection failure.

**Timestamps are tagged Europe/Paris despite being named UTC.** The instants
are correct, so converting is right and stripping the zone is a two-hour error.
Spark's session timezone is set to UTC in the local session builder and not in
the distributed one, so the loader asserts it rather than assuming it.


## What the detectors produce

Every event type step 04 emits, counted on the sample. Most of these have no
counterpart in any reference dataset, so this table is the only place they are
visible at all — a family that stopped being emitted would otherwise show up
not as a regression but as an absence, which is harder to notice.

```{r tbl-inventory}
#| echo: false
#| results: asis
if (!is.null(inv_2025)) {
  w <- reshape(inv_2025, idvar = "type", timevar = "rung", direction = "wide")
  names(w) <- sub("^n_events\\.", "", names(w))
  keep <- intersect(c("type", "L00_legacy", "L13_shipped"), names(w))
  w <- w[order(w$type), keep, drop = FALSE]
  cat("| Event type |", paste(keep[-1], collapse = " | "), "|\n")
  cat("|---", paste(rep("---:", length(keep) - 1), collapse = "|"), "|\n", sep = "")
  for (i in seq_len(nrow(w))) {
    vals <- vapply(keep[-1], function(k) {
      v <- w[i, k]
      if (is.na(v)) "—" else format(v, big.mark = ",")
    }, character(1))
    cat(sprintf("| `%s` | %s |\n", w$type[i], paste(vals, collapse = " | ")))
  }
} else {
  cat("*Extraction inventory not present in the committed data — the ladder has not been run.*\n")
}
```

## How accurate the milestones are

Scored against APDF on the milestones it can reach. Coverage is the share of
reference milestones with any detection; bias is the median signed error, so a
negative value means the detector fires early.

```{r tbl-ladder}
#| echo: false
#| results: asis
if (!is.null(ladder_2025)) {
  d <- ladder_2025
  # det_type is the OPDI event type; milestone is the APDF milestone it claims
  # to be. Both are shown because `take-off` and `ATOT` are scored against the
  # same truth, and which of them wins is the question the ladder answers.
  # Each OPDI type is scored against every truth milestone, so `take-off`
  # also produces rows against ALDT, AOBT and AIBT with no detections. Those
  # are true but meaningless -- a take-off event is not a failed on-block.
  # Keep only the pairing each type actually claims.
  claims <- c("take-off" = "ATOT", "landing" = "ALDT", "ATOT" = "ATOT",
              "ALDT" = "ALDT", "AOBT" = "AOBT", "AIBT" = "AIBT")
  if (!is.null(d$det_type)) {
    d <- d[!is.na(claims[d$det_type]) & claims[d$det_type] == d$milestone, ]
  }
  d <- d[order(d$milestone, d$rung), ]
  cat("| Rung | OPDI type | vs | Truth | Detected | Coverage | Bias (s) | MAD (s) | ±60 s |\n")
  cat("|---|---|---|---:|---:|---:|---:|---:|---:|\n")
  for (i in seq_len(nrow(d))) {
    dt <- if (!is.null(d$det_type)) d$det_type[i] else d$milestone[i]
    cat(sprintf("| %s | `%s` | %s | %s | %s | %.2f%% | %.1f | %.1f | %.2f%% |\n",
        d$rung[i], dt, d$milestone[i],
        format(d$n_truth[i], big.mark = ","),
        format(d$n_detected[i], big.mark = ","),
        d$coverage_pct[i], d$bias_s[i], d$mad_s[i], d$within_60s_pct[i]))
  }
} else {
  cat("*Ladder results not present in the committed data — see below.*\n")
}
```

## What the numbers say

### Landing time is transformed; take-off trades sharpness for reach

Both the published fuzzy-phase pair and the new runway-anchored pair are scored
against the same APDF truth, because which of them wins is the question this
study exists to answer.

The published `landing` is beaten on every axis at once: the new `ALDT` covers
forty times as many flights, its median error is two seconds early rather than
thirty-four late, and four in five detections fall within a minute against one
in sixty.

`take-off` is the more interesting case, and the honest reading is a trade
rather than a win. The published detector has the *better median* — a one
second bias against nineteen — but it fires on 1.8% of flights and its 90th
percentile error is **4,273 seconds**. Seventy-one minutes. So it is often
excellent and occasionally absurd, and nothing in its output distinguishes the
two. The new `ATOT` covers thirty-five times more flights with a 90th
percentile of 59 seconds. For an indicator averaged over a population, a
detector that is never wildly wrong is worth more than one that is sometimes
perfect.

### Requiring complete evidence made detection worse

The ladder's purpose is attribution, and it earned that here by contradicting
one of this study's own fixes.

`least`/`greatest` skip nulls, so a fuzzy rule missing an input became a
minimum over fewer terms and could out-compete a complete rule. Requiring every
input to be present is obviously more correct. On real data it removes 90% of
phase-based landings — coverage falls from 6.49% to 0.69% — because ADS-B
carries a null velocity often enough that insisting on complete evidence
discards most ground samples.

It matters less than it appears, since the runway-anchored detectors replace
that path entirely. But it is a genuine negative result, it would have been
invisible in a two-configuration comparison, and it is recorded here rather
than quietly reverted.

### Off-block and on-block are limited by reception, as expected

Coverage of 2.6% and a bias of +744 seconds. The bias is not noise: sustained
movement is detectable only once reception begins, which is typically well
after the aircraft actually leaves the stand. Both numbers are reported as the
finding rather than tuned away, and they are why T05 (pushback) was not built —
it needs strictly more coverage than this.

### Does it hold on a second period?

Both periods were run. What 2024 can check is narrower than 2025: no flight
list exists for that period, and the runway, ring, airport and block detectors
all need one, so 2024 exercises the phase and crossing families only. The ALDT
and ATOT headline figures are therefore **single-period**.

For what it does cover, the two disagree in an instructive way.

| Rung | `landing` coverage, 2025 | 2024 |
|---|---:|---:|
| `L01_clean_tracks` | 6.39% | 6.66% |
| `L02_smoothing` | 6.49% | **0.96%** |
| `L03_complete_rules` | 0.69% | 0.23% |

**Requiring complete fuzzy rules reproduces.** A large relative collapse on
both periods -- 89% and 76% -- so the finding is a property of the change
rather than of a month.

**The smoothing does not.** Restoring the reference implementation's 60-second
majority vote is near-neutral on 2025 and costs 86% of landing coverage on
2024. This was the change with the strongest prior justification in the whole
study, and it is the one the second period contradicts.

One confound has to be stated with it. The 2024 tracks predate step 02's
altitude repair, so that run substitutes the raw barometric altitude. The
period therefore differs in input preparation as well as in date, and the
smoothing operates on phase labels derived from exactly that column. The
divergence may be the substitution rather than the month, and this study cannot
separate the two. What it can say is that the change is **not robust across the
conditions tested**, which is the useful half of the claim: a fix that behaves
this differently on noisier input should not be adopted on the strength of its
rationale alone.

### What has still not been measured

**The vertical family cannot be scored against data at all.** No source
available to OPDI holds level-segment truth. The claim made for KPI17 and
KPI19 is conformance to ICAO's published algorithm, tested on trajectories
whose geometry is known by construction. That is checkable; "accurate" is not.

**Top-of-climb and top-of-descent timing rests on no external evidence.** The
counts move — 114,099 to 100,925 for TOC — but nothing says which is closer to
the truth.

**The 2025 ladder is stale against the current code.** It was produced before
the changes that made the 2024 period runnable -- the track redirect, the
column backfills, the identity source. Those cannot alter a 2025 result, but
the provenance fingerprint covers the file rather than the behaviour, so the
chain reports it stale and this page says so rather than re-recording it as
current. Clearing it honestly means re-running it.

**Ring crossings are not yet scored against `C40_CROSS_TIME`.** The detector
runs and its volumes are now plausible, but the comparison against the
reference crossing times is not in this version.

## Breaking change for consumers

The new version does **not** re-emit `first-xing-fl*` and `last-xing-fl*`.
Those types keep their meaning under the frozen `events_v0.0.2`, so published
data stays reproducible, but consumers moving to the new version must migrate
to the crossing sequence number carried in each event's `info`: the first
crossing is sequence 1, the last is the maximum.

Consumers should also know that `level-start`/`level-end` and the new ICAO
`level-off-*` family answer different questions and are not interchangeable,
and that the event table grows materially — every crossing is now emitted where
previously only two per level were.

## Provenance

```{r tbl-provenance}
#| echo: false
#| results: asis
if (!is.null(manifest) && length(manifest) > 0) {
  cat("| Output | Script | Git SHA | Produced |\n")
  cat("|---|---|---|---|\n")
  for (name in sort(names(manifest))) {
    e <- manifest[[name]]
    if (is.null(e$script)) next
    dirty <- if (isTRUE(e$git_dirty)) " (dirty)" else ""
    cat(sprintf("| `%s` | `%s` | `%s`%s | %s |\n", name, e$script,
                e$git_sha, dirty, substr(e$produced_utc, 1, 10)))
  }
} else {
  cat("*No manifest present.*\n")
}
```

Every figure above is produced by `benchmarks/regenerate_events.py`, which
re-runs a job only when the source files that job depends on have changed.
Outputs with no manifest entry are reported as unverified rather than shown as
fact. The ladder outputs are absent from this table because they have not been
produced.
