---
title: "Flight events from ADS-B"
subtitle: "Version 3 — a chapter per event"
date: last-modified
author: "EUROCONTROL Performance Review Unit — Open Performance Data Initiative"
abstract: |
  OPDI publishes thirty-four flight event types. Until this study none had been
  tested or compared against reference data. This version is organised as one
  chapter per event: what it is, how it is detected, the parameters that govern
  it, how it changed from the published algorithm, and whether any external
  source agrees with it. Where no comparison is possible the chapter says so
  and says why, because the reasons differ and are informative. A closing
  chapter designs a parameter sweep without running one.
format:
  html:
    toc: true
    toc-depth: 2
  pdf:
    toc: true
    toc-depth: 2
---

```{r regenerate}
#| include: false
#| cache: false

# Rendering this page runs the analysis behind it. `regenerate_events.py` is
# the executable definition of the study: every figure below is produced by
# exactly one job it declares, and it re-runs a job only when the code that job
# depends on has changed. OPDI_RENDER=check fails fast on staleness.
REPO   <- normalizePath(file.path(getwd(), "..", "..", "..", "opdi"), mustWork = FALSE)
PYTHON <- Sys.getenv("OPDI_PYTHON", file.path(REPO, ".venv310", "bin", "python"))
MODE   <- Sys.getenv("OPDI_RENDER", "allow-stale")

if (MODE %in% c("run", "check") && file.exists(PYTHON)) {
  args <- c(file.path(REPO, "benchmarks", "regenerate_events.py"))
  if (MODE == "check") args <- c(args, "--check")
  status <- system2(PYTHON, args, stdout = "", stderr = "")
  if (MODE == "check" && status != 0) stop("regenerate_events.py --check reports stale outputs")
}

rd <- function(name) {
  path <- file.path("data", name)
  if (!file.exists(path)) return(NULL)
  if (grepl("[.]json$", name)) jsonlite::fromJSON(path) else read.csv(path, stringsAsFactors = FALSE)
}

inv25 <- rd("inventory_2025.csv"); inv24 <- rd("inventory_2024.csv")
lad25 <- rd("ladder_2025.csv");    lad24 <- rd("ladder_2024.csv")
rings  <- rd("rings_2025.csv");    rwy   <- rd("runway_2025.csv")
floor  <- rd("floor_2025.csv")
br25   <- rd("bridge_2025.json");  br24  <- rd("bridge_2024.json")
manifest <- rd("_manifest.json")

# Every count in this document is looked up from the staged inventory rather
# than typed into the prose, so a re-run moves the text with the data.
n <- function(type, rung = "L13_shipped", inv = inv25) {
  v <- inv$n_events[inv$type == type & inv$rung == rung]
  if (length(v) == 0) 0L else as.integer(v[1])
}
fm <- function(x) format(x, big.mark = ",")
# legacy -> shipped, as "8,483 -> 2,570"
delta <- function(type, inv = inv25) {
  sprintf("%s → %s", fm(n(type, "L00_legacy", inv)), fm(n(type, "L13_shipped", inv)))
}
# a scored row from the ladder
sc <- function(rung, det_type, field) {
  v <- lad25[lad25$rung == rung & lad25$det_type == det_type &
             lad25$milestone == c("take-off"="ATOT","landing"="ALDT","ATOT"="ATOT",
                                  "ALDT"="ALDT","AOBT"="AOBT","AIBT"="AIBT")[det_type], field]
  if (length(v) == 0) NA else v[1]
}
```

# The study in a page

OPDI turns OpenSky ADS-B state vectors into a published table of flight events —
milestones with a 4D fix, a type, a source and an algorithm version. Thirty-four
types ship. Before this study, **none of them had a test, and no number in any
of them had been compared against anything.**

Reading the four detectors turned up nine defects. None of them fails: each
produces fewer events, or events displaced in one direction, which is the
failure mode that survives an eyeball check of the totals. The worst suppressed
every take-off and landing at any aerodrome above roughly 200 ft.

Eight are fixed, four milestone families added, and everything measurable is
measured against EUROCONTROL APDF on two three-day samples.

## What the evidence says

| Finding | Where |
|---|---|
| **Ring crossings agree with the reference as closely as the reference agrees with itself** — median error under half a second, spread inside the inter-source floor | [Ring crossings] |
| Landing time transformed: coverage 2.12% → 86.06%, bias +34 s → −2 s | [ALDT — touchdown] |
| Take-off trades a sharp median for a usable tail: p90 4,273 s → 59 s | [ATOT — lift-off] |
| Runway identity is the weak spot: a third of named landing runways are wrong | [ATOT — lift-off], [ALDT — touchdown] |
| Two thirds of published level segments were phase flicker | [level-start / level-end] |
| **A fix of ours is harmful and reproduces on both periods** — requiring complete fuzzy evidence | [Cross-cutting findings] |
| **A fix of ours is contradicted by the second period** — the smoothing window | [Cross-cutting findings] |
| Ground milestones are limited by reception, not algorithm: 2.6% coverage | [AOBT / AIBT — block times] |

Nine ICAO GANP indicators become fully computable and three half — KPI02, 05,
08, 10, 13, 15, 17, 18, 19, plus the actual halves of KPI01, KPI11 and KPI14.

# How to read a chapter

Every chapter has the same six parts, so it can be skimmed or read in full:

**What it is** — in operational terms. **How it is detected** — the algorithm.
**Parameters** — each with its value, unit, and *source*: ICAO's published
value, `traffic`'s, OpenAP's, or ours. **Compared with the published
algorithm** — what changed and by how much. **Compared with reference data** —
the APDF scores, or why none is possible. **Limitations** — what is known to be
wrong.

Two conventions apply throughout. Storage is SI and thresholds are aviation:
altitudes are metres and vertical rates metres per second on disk, scaled at
the point of comparison. And every threshold is named in `EventConfig`
(`src/opdi/config.py`) — none is an inline literal, which is what makes a sweep
possible at all.

`EventConfig.legacy()` reproduces the published `events_v0.0.2` exactly, so
every "compared with the published algorithm" figure below is a real
measurement of two configurations on the same trajectories, not a recollection.

# Ground truth, and its ceiling

Reference data is EUROCONTROL APDF. It is in movement form — there is no
literal take-off-time column, and `SRC_PHASE` decides what a row carries — and
it has **no aircraft address**, so it cannot be joined to ADS-B directly. The
route runs through the flight table on an internal key, and that bridge caps
every coverage figure in this document.

```{r tbl-bridge}
#| echo: false
#| results: asis
if (!is.null(br25) && !is.null(br24)) {
  cat("| Period | APDF movements | Reaching `icao24` | Share |\n|---|---:|---:|---:|\n")
  for (r in list(list("2024-06", br24), list("2025-06", br25))) {
    b <- r[[2]]
    cat(sprintf("| %s | %s | %s | **%s%%** |\n", r[[1]], fm(b$movements),
                fm(b$reached_icao24), b$reach_pct))
  }
}
```

Three traps in this join fail silently, and all three are closed in the loader
rather than left to callers.

**The reference carries the commercial flight number; ADS-B carries the ATC
callsign.** They disagree on 20.4% of rows — `LH416` against `DLH416` — and
joining on the wrong one discards a fifth of the sample while looking exactly
like a detection failure.

**Timestamps are tagged `Europe/Paris` despite being named `_UTC`.** The
instants are correct, so converting is right and stripping the zone is a
two-hour error. Spark's session timezone is set in the local builder and not
the distributed one, so the loader asserts it rather than assuming it.

**ADS-B callsigns are space-padded to eight characters.** An untrimmed join
matches nothing and reports every method at 0.00% rather than erroring.

## The yardstick

APDF contains two independent EUROCONTROL derivations of each ring crossing —
the shipped column and its `_CTFM` twin. How far apart *they* are sets the
floor on what agreement can mean here.

```{r tbl-floor}
#| echo: false
#| results: asis
if (!is.null(floor)) {
  cat("| Ring | p10 | median | p90 |\n|---|---:|---:|---:|\n")
  for (i in seq_len(nrow(floor)))
    cat(sprintf("| `%s` | %+.0f s | %+.0f s | %+.0f s |\n", floor$milestone[i],
                floor$p10_s[i], floor$median_s[i], floor$p90_s[i]))
}
```

**No tolerance tighter than this means anything.** It is a better basis than a
round number chosen for looking reasonable, and the ring chapter is read
against it.

## Resolution differs by milestone

64% of `MVT_TIME_UTC` values and 77% of `BLOCK_TIME_UTC` land on a whole
minute — most reporting airports round to the minute — while ring crossings are
second-resolution. **The rings are therefore the precise target and the
movement times the coarse one**, and a perfect take-off detector would still
show tens of seconds of apparent spread against a minute-rounded truth.

## What 2024 can and cannot check

The second period exists to test whether a finding survives a different sample.
It covers less: no flight list exists for it, and every detector needing an
aerodrome — runway, rings, airport layout, blocks — depends on one. So 2024
emits 16 types against 2025's 34, and exercises the **phase and crossing
families only**. Each chapter says which of its figures are single-period.

---

# The chapters

## take-off

**What it is.** The instant the aircraft leaves the ground, as the phase
sequence sees it.

**How it is detected.** Each state vector is classified into a flight phase by
fuzzy logic — nine membership functions over altitude, vertical rate and speed,
combined by five rules with a minimum t-norm, the phase being the highest
activation. `take-off` fires where the previous sample is ground and the next is
climb.

Ground membership is the part that matters. `baro_altitude_c` is *uncorrected
pressure altitude*, so the published 200 ft ceiling is unreachable at any
aerodrome above roughly 200 ft AMSL, or anywhere on a low-pressure day — and
without a ground sample there is no take-off at all. The fix measures height
above **field elevation**, attaching the elevation of both the track's ADEP and
ADES and taking the more permissive membership: a track is on the ground at
only one of its ends, and cruise sits far above both, so the wrong end can only
contribute a zero.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `phase_ground_ceiling_ft` | 200 ft | published | where ground membership reaches zero |
| `phase_ground_above_field` | on | ours | whether that ceiling is above field or above the ellipsoid |
| `phase_twindow_seconds` | 60 s | OpenAP | majority smoothing of per-sample labels |
| `phase_require_complete_rules` | on | ours | whether a rule with a missing input may win |

**Compared with the published algorithm.**
`r delta("take-off")` events on the 2025 sample. The count *falls*, which is
the opposite of the field-elevation fix's intent — the ladder attributes it to
two other changes, and that is the subject of [Cross-cutting findings].

**Compared with reference data.** Scored against APDF's ATOT.

```{r tbl-takeoff}
#| echo: false
#| results: asis
if (!is.null(lad25)) {
  cat("| Configuration | Coverage | Bias | MAD | p90 |\n|---|---:|---:|---:|---:|\n")
  for (r in c("L00_legacy", "L13_shipped"))
    cat(sprintf("| %s | %.2f%% | %+.0f s | %.0f s | %.0f s |\n", r,
        as.numeric(sc(r,"take-off","coverage_pct")), as.numeric(sc(r,"take-off","bias_s")),
        as.numeric(sc(r,"take-off","mad_s")), as.numeric(sc(r,"take-off","p90_abs_s"))))
}
```

Under the published configuration the median is excellent and the tail is not:
a bias of one second against a 90th-percentile error of over an hour, on 1.81%
of departures. It is often nearly exact and occasionally absurd, with nothing in
the output distinguishing the two. The new configuration tightens the tail to
39 s but detects fewer take-offs still — the collapse traced in
[Cross-cutting findings]. Either way this is a detector that answers for well
under one flight in fifty. [ATOT — lift-off] is the intended replacement; this
type keeps its published string so existing consumers are undisturbed.

**Limitations.** Depends entirely on a ground sample existing, so sparse
surface reception removes it silently. Inherits every property of the phase
classifier, including its speed bias — see [top-of-climb / top-of-descent].

## landing

**What it is.** The instant the aircraft meets the ground.

**How it is detected.** As above, but fires where the previous phase is descent
and the current is ground. The two definitions are deliberately asymmetric —
one looks forward, one does not — inherited from the published detector and
preserved so the type keeps its meaning.

**Parameters.** As [take-off].

**Compared with the published algorithm.** `r delta("landing")` on 2025.

**Compared with reference data.** Scored against APDF's ALDT.

```{r tbl-landing}
#| echo: false
#| results: asis
if (!is.null(lad25)) {
  cat("| Configuration | Coverage | Bias | MAD | p90 |\n|---|---:|---:|---:|---:|\n")
  for (r in c("L00_legacy", "L13_shipped"))
    cat(sprintf("| %s | %.2f%% | %+.0f s | %.0f s | %.0f s |\n", r,
        as.numeric(sc(r,"landing","coverage_pct")), as.numeric(sc(r,"landing","bias_s")),
        as.numeric(sc(r,"landing","mad_s")), as.numeric(sc(r,"landing","p90_abs_s"))))
}
```

**Limitations.** As [take-off]. Superseded in practice by [ALDT — touchdown],
which covers forty times as many flights.

## top-of-climb / top-of-descent

**What it is.** The first and last sample classified as cruise.

**How it is detected.** Directly from the phase labels: the earliest and latest
`CR` sample per track.

**Parameters.** The cruise rule is `min(alt_hi, roc_zero, spd_hi)` with
OpenAP's memberships — altitude Gaussian at 35,000 ft (σ 20,000), vertical rate
Gaussian at 0 ft/min (σ 100), speed Gaussian at **600 kt (σ 100)**.

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `phase_cruise_speed_kt` | 600 kt | OpenAP | centre of the cruise speed membership |
| `phase_cruise_speed_sigma_kt` | 100 kt | OpenAP | its width |

**These are OpenAP's published values, not chosen here**, and they carry a
known consequence: a turboprop cruising at 250 kt scores `exp(−6.125) ≈ 0.002`,
so the cruise rule never wins and **turboprop and general-aviation traffic get
no top-of-climb or top-of-descent at all**. It is inherited rather than an OPDI
defect, and left in place deliberately — deviating from a reference constant
should follow a measurement, not precede one. See [The parameter sweep,
designed].

**Compared with the published algorithm.** TOC `r delta("top-of-climb")`,
TOD `r delta("top-of-descent")`.

**Compared with reference data.** **None is possible.** No source available to
OPDI records a top of climb or descent. APDF holds movement and block times,
ring crossings, a runway and a stand — nothing about the cruise phase. These
counts are reported; their timing is unvalidated and stated as such.

**Limitations.** A flight that never satisfies the cruise rule has neither
event, and the speed bias above makes that systematic rather than random.

## level-start / level-end

**What it is.** The boundaries of a run of samples classified cruise or level.

**How it is detected.** A `level-start` where the phase enters cruise or level
from something else, a `level-end` where it leaves. The first cruise sample
carries both `level-start` and `top-of-climb`; the last carries `level-end` and
`top-of-descent`.

**These are not the ICAO level segments.** This pair asks "does this sample look
like level flight" through the fuzzy classifier; ICAO asks a geometric question
about an altitude band anchored at a segment's own start. Both are published,
under different names, and they are **not interchangeable** — see
[level-off-climb / level-off-descent].

**Parameters.** As [take-off]; `phase_twindow_seconds` dominates.

**Compared with the published algorithm.** `level-start` `r delta("level-start")`
and `level-end` `r delta("level-end")` on 2025.

This is the clearest single measure of phase flicker in the study: **roughly
two thirds of published level segments were an artefact of deciding the phase
per state vector with no temporal aggregation.** One misclassified sample at
5-second spacing injects a spurious start-and-end pair, and the reference
implementation's 60-second majority smoothing — which the original port carried
the membership functions across but dropped — removes most of them.

**Compared with reference data.** **None is possible**, as for
[top-of-climb / top-of-descent].

**Limitations.** The count fell by two thirds, and nothing external says which
count is closer to the truth. That the smoothed number is more plausible is an
argument, not a measurement.

## Published flight-level crossings

*Types:* `first-xing-fl50`, `first-xing-fl70`, `first-xing-fl100`,
`first-xing-fl245`, `last-xing-fl50`, `last-xing-fl70`, `last-xing-fl100`,
`last-xing-fl245`.

**What it is.** The first and last time a track crosses each monitored flight
level.

**How it is detected.** Each sample's flight level is compared with the next
one's; the sample *before* the crossing is reported. A guard requires it to be
within 10 flight levels of the boundary, rejecting crossings inferred across a
gap. The first and last per level are kept; everything between is discarded.

**Parameters.** The level set — FL50, FL70, FL100, FL245 — and the 10 FL guard,
both published values.

**Compared with the published algorithm.** These types are emitted **only**
under `events_v0.0.2`: `r fm(n("first-xing-fl100","L00_legacy"))` `first-xing-fl100`
events under legacy, `r fm(n("first-xing-fl100"))` under the new version. They
are replaced by [Flight-level crossings], and the migration is
`crossing_seq == 1`.

**Compared with reference data.** **None is possible.** APDF records no
flight-level crossings at all. Their only comparison is against their
replacement.

**Limitations — three, which is why they were replaced.** Keeping only first
and last discards every crossing between, which are exactly the ones a
vertical-efficiency indicator is about. Reporting the bracketing sample biases
every timestamp by up to one sample interval, always in the same direction. And
the level is chosen by a first-match chain, so a 5-second gap spanning FL50 *and*
FL70 reports only one of them.

## Flight-level crossings

*Types:* `xing-fl50`, `xing-fl70`, `xing-fl100`, `xing-fl245`.

**What it is.** Every crossing of each monitored flight level, with the instant
interpolated to the level itself.

**How it is detected.** A Schmitt trigger. For a level `L` and dead band `h`, a
sample is `+1` above `L + h`, `−1` below `L − h`, and null inside the band; that
side is forward-filled, and a crossing is confirmed where the filled value
flips. An aircraft cruising *at* the level never leaves the band, so its side
never resolves and **no event is emitted** — without this, barometric noise at a
cruise level produces a crossing on almost every sample.

The instant is then interpolated to exactly `L` **between the two samples that
straddle it**, which is a *different row* from the one that confirmed the
crossing: confirmation happens at the band edge, the crossing itself is earlier.
Conflating them reintroduces the one-sided bias this replaces.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `crossing_levels_fl` | 50, 70, 100, 245 | published vocabulary | which levels are monitored |
| `crossing_hysteresis_ft` | 300 ft | ours | chatter suppression vs missed crossings |
| `crossing_all_occurrences` | on | ours | every crossing rather than first and last |
| `crossing_interpolate` | on | ours | interpolate to the level rather than report a sample |

Each event carries `crossing_seq` (1-based, in time order), `direction`, and
`bracket_s` — the spacing of the two interpolated-between samples, so a consumer
can tell an instant measured across 5 seconds from one inferred across four
minutes.

**Compared with the published algorithm.** `r fm(n("xing-fl100"))`
`xing-fl100` events against `r fm(n("first-xing-fl100","L00_legacy") + n("last-xing-fl100","L00_legacy"))`
first-and-last events under legacy — a factor of
`r sprintf("%.2f", n("xing-fl100") / (n("first-xing-fl100","L00_legacy") + n("last-xing-fl100","L00_legacy")))`.

**That factor is the finding, and it is smaller than expected.** Most flights
cross a given level exactly twice — once up, once down — so for most
trajectories the published first-and-last rule was already capturing everything
there was. The extra 13% is concentrated in the flights that step through a
level repeatedly, which are precisely the ones a vertical-efficiency indicator
is about. **The case for this detector is therefore not volume.** It is that the
timestamps are interpolated rather than displaced by up to a sample interval,
that a gap spanning two levels now reports both, and that the flights where the
old rule lost information are exactly the interesting ones.

**Compared with reference data.** **None is possible** — APDF holds no
flight-level crossings. The detector's *mechanism* is nonetheless validated
where the same code is applied to rings, which APDF does record: see
[Ring crossings].

**Limitations.** A flight that climbs and levels off **exactly at** a threshold
never clears the far edge and registers no crossing there. That case is
described by the level-segment and top-of-climb events instead, but it is a real
semantic change from the published detector.

## Ring crossings

*Types:* `xing-40nm`, `xing-100nm`.

**What it is.** Crossings of the 40 NM and 100 NM cylinders around the flight's
**own** origin and destination — ICAO's ASMA cylinder for KPI08 and the
reference area for KPI05.

**How it is detected.** The same Schmitt trigger as [Flight-level crossings],
with distance to an aerodrome as the quantity and a 1 NM dead band. The rings
come from the flight list's ADEP and ADES rather than from the H3 aerodrome
zone table: that is what the indicators mean, and what APDF records — one
crossing per movement, not one per aerodrome overflown. It is also far cheaper,
multiplying each sample by at most two rather than by every aerodrome within
110 NM.

Bearing is computed **from** the interpolated crossing position rather than
interpolated alongside it. A bearing is circular: averaging 359 and 1 gives 180,
wrong by half a turn on any crossing near due north.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `ring_radii_nm` | 40, 100 NM | ICAO | which cylinders |
| `ring_hysteresis_nm` | 1 NM | ours | chatter suppression |
| `crossing_interpolate` | on | ours | shared with the level crossings |

`info` carries `crossing_seq`, `direction` (inbound/outbound), `apt_icao`,
`bearing` and `flight_level`, matching `C40_CROSS_{TIME,LAT,LON,FL}` and
`C40_BEARING` column for column, so the comparison is direct rather than a
translation.

**Compared with the published algorithm.** New in this version:
`r fm(n("xing-40nm"))` and `r fm(n("xing-100nm"))` events where the published
algorithm emitted none.

**Compared with reference data — the strongest result in this study.**

```{r tbl-rings}
#| echo: false
#| results: asis
if (!is.null(rings)) {
  cat("| Ring | Coverage | Bias | MAD | p90 | within 10 s | Median position error |\n")
  cat("|---|---:|---:|---:|---:|---:|---:|\n")
  for (i in order(rings$milestone))
    cat(sprintf("| `%s` | %.2f%% | %+.1f s | %.1f s | %.1f s | %.2f%% | %.3f NM |\n",
        rings$milestone[i], rings$coverage_pct[i], rings$bias_s[i], rings$mad_s[i],
        rings$p90_abs_s[i], rings$within_10s_pct[i], rings$median_pos_error_nm[i]))
}
```

Read against the yardstick: the inter-source floor is p10 −9 s / p90 +11 s at
40 NM and −6 s / +7 s at 100 NM. **The detector's spread sits inside the
disagreement between two EUROCONTROL derivations of the same crossing.** A
median error under half a second and a median position error of about 0.2 NM
say the interpolation is doing exactly what it was built to do.

This is also the only place a seconds-level claim in this study is defensible,
because ring reference times are second-resolution where the movement times are
not.

**Limitations.** Single-period: 2024 emits no ring events because it has no
flight list. And a ring crossing is not detected for an aerodrome merely
overflown — correct for these indicators, but not a general-purpose airspace
crossing.

**A bug worth recording, because it shaped this detector.** The first run
produced 2.85 million 40 NM crossings against 236,899 tracks, with *identical*
counts for both rings on the worst flights — a signature real geometry cannot
produce. `haversine_nm` clamped the argument of `asin` with `F.least(a, 1.0)`,
and `F.least` skips nulls, so a missing coordinate became `asin(1)` = 10,807 NM,
the antipodal distance. A transponder parked at Budapest transmitted for the
whole 72-hour sample; cleaning nulled its repeated positions as stale
broadcasts, and its distance series then alternated between 0.7 NM and the
antipode. The fix reduced the counts nineteen- and twenty-three-fold. It was
found by the extraction inventory — not by the tests, and not by the scored
metrics.

## Airport layout events

The next seven chapters share one detector, described once here.

**How they are detected.** Low-altitude samples are matched against
OpenStreetMap aeroway geometry pre-rasterised to H3 resolution 12 in
`hexaero_airport_layouts`. A sample joins a feature when its `h3_res_12` cell
matches *and* the feature's aerodrome is one the flight list names for that
flight. Consecutive matches on the same feature form a traversal; a gap over 300
seconds starts a new one. Each traversal yields an entry and an exit event.

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `airport_max_fl` | FL20 | published | altitude gate for layout matching |
| `airport_trace_gap_seconds` | 300 s | published | when a traversal is considered ended |
| `airport_events_ordered` | on | ours | attributes taken at the entry/exit *times* |
| H3 resolution | 12 | `H3Config` | matching granularity |

`airport_events_ordered` is a correctness fix, not a tuning knob: the published
detector used `first`/`last` inside a group-by, which take **partition order,
not time order**, so a reported entry position was not guaranteed to be the
position at the reported entry time.

**Compared with reference data — none is possible for any of them.** APDF
records a runway designator (`AP_C_RWY`) and a stand (`AP_C_STND`), but **no
entry or exit times for any surface feature**. So a runway *identity* can be
checked indirectly through [ATOT — lift-off], but no aeroway event's *timing*
can be checked at all.

**Shared limitations.** The detector cannot fire for an aerodrome the flight
list did not name; it depends on OSM coverage, built only for large and medium
aerodromes in the ingestion area; and a traversal completed inside a single
5-second sample is dropped, because its duration is zero. All counts below are
single-period — 2024 emits none of these.

### entry-runway / exit-runway

The runway surface. `r delta("entry-runway")` entries and
`r delta("exit-runway")` exits. The fall is the cleaned track table: cleaning
nulls repeated positions as stale broadcasts, which removes surface samples.

### entry-taxiway / exit-taxiway

`r delta("entry-taxiway")` entries, `r delta("exit-taxiway")` exits. The most
numerous aeroway feature and the most sensitive to surface reception.

### entry-apron / exit-apron

`r delta("entry-apron")` entries, `r delta("exit-apron")` exits.

### entry-parking_position / exit-parking_position

The stand. `r delta("entry-parking_position")` entries,
`r delta("exit-parking_position")` exits. These anchor
[AOBT / AIBT — block times]: without them a track first received mid-taxi would
report its first movement as an off-block.

### entry-threshold / exit-threshold

`r delta("entry-threshold")` entries, `r delta("exit-threshold")` exits.

**The largest relative fall of any event type, and it deserves an explanation
rather than a row in a table.** Threshold polygons are the smallest features in
the layout set, so they are the most sensitive to which samples survive
cleaning. This is trajectory cleaning removing the samples a detector depends
on — not a detector change — and it is the clearest instance of that effect in
the study.

### entry-hangar / exit-hangar

`r delta("entry-hangar")` entries, `r delta("exit-hangar")` exits. Rare by
nature; included for completeness.

### entry-deicing_pad / exit-deicing_pad

In the emitted vocabulary but **not observed on either sample** — de-icing pads
are tagged in OSM for few of the aerodromes covered, and the sample is June.

## ATOT — lift-off

**What it is.** The take-off time, anchored on the runway the movement actually
used. ICAO's T08 in all but name; APDF's `MVT_TIME_UTC` for a departure.

**How it is detected.** A native port of `traffic`'s
`TrackBasedRunwayDetection`. Samples are kept when they are within 5 NM of the
aerodrome, below 1,500 ft above field elevation, faster than 30 kt over the
ground, and climbing faster than 257 ft/min. At least four must survive. The
median track of the survivors is matched against each runway's bearing and the
nearest within 10° wins; parallel runways are separated by cross-track distance
to each centreline, in closed form where `traffic` uses a geometry library.
**ATOT is the earliest surviving sample** of that window.

Runway geometry comes from `oa_runways` — the OurAirports table step 00d has
generated since it was written and which nothing had ever read. Bearings are
computed from the two threshold positions rather than from `le_heading_degT`,
which is often null and sometimes magnetic rather than true.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `runway_max_dist_nm` | 5 NM | `traffic` | how far from the aerodrome samples count |
| `runway_max_height_ft` | 1,500 ft | `traffic` | ceiling of the detection window |
| `runway_min_vert_rate_ftmin` | 257 ft/min | `traffic` | separates the climb from ground movement |
| `runway_min_groundspeed_kt` | 30 kt | `traffic` | excludes taxiing |
| `runway_max_bearing_deg` | 10° | `traffic` | abstention threshold |

**Compared with the published algorithm.** New: `r fm(n("ATOT"))` events where
the published algorithm emitted none. Its predecessor is [take-off], and the
comparison is the point of this study — see the two tables above and below.

**Compared with reference data.**

```{r tbl-atot}
#| echo: false
#| results: asis
if (!is.null(lad25)) {
  cat("| Detector | Coverage | Bias | MAD | p90 | ±60 s |\n|---|---:|---:|---:|---:|---:|\n")
  for (d in c("take-off", "ATOT"))
    cat(sprintf("| `%s` | %.2f%% | %+.0f s | %.0f s | %.0f s | %.2f%% |\n", d,
        as.numeric(sc("L13_shipped",d,"coverage_pct")), as.numeric(sc("L13_shipped",d,"bias_s")),
        as.numeric(sc("L13_shipped",d,"mad_s")), as.numeric(sc("L13_shipped",d,"p90_abs_s")),
        as.numeric(sc("L13_shipped",d,"within_60s_pct"))))
}
```

**This is a trade, not a clean win.** The published `take-off` has the better
median — one second, against nineteen. But it has it on 0.39% of departures,
where `ATOT` has 63.30%, and even the published algorithm's own configuration
only reached 1.81% with a 90th-percentile error of over an hour. `ATOT` gets a
90th-percentile error of about a minute on two thirds of all departures. For an
indicator averaged over a population, a detector that is never wildly wrong is
worth more than one that is sometimes perfect and almost always absent.

Runway identity, against `AP_C_RWY`:

```{r tbl-rwy}
#| echo: false
#| results: asis
if (!is.null(rwy)) {
  d <- rwy[rwy$n_named > 0, ]
  cat("| Milestone | Named | Exactly right | Share of named |\n|---|---:|---:|---:|\n")
  for (i in order(d$milestone))
    cat(sprintf("| `%s` | %s (%.2f%%) | %s | **%.2f%%** |\n", d$milestone[i],
        fm(d$n_named[i]), d$named_pct[i], fm(d$n_exact[i]), d$exact_pct_of_named[i]))
}
```

**This is the weak spot of the family.** Roughly a quarter of the runways
`ATOT` names are wrong. Designators are compared case-insensitively with
whitespace stripped and not otherwise fuzzed, because `07R` and `07L` are
different runways and a lenient match would hide exactly the error worth
finding.

**Limitations.** The time is a **proxy** — the extreme of a detection window,
not a measured wheels-off — and its bias is reported rather than assumed.
Single-period. The detector abstains where no runway is within tolerance, which
is deliberate: naming the wrong runway corrupts a movement count, naming none
is an explicit null.

## ALDT — touchdown

**What it is.** The landing time with its runway. ICAO's T17; APDF's
`MVT_TIME_UTC` for an arrival.

**How it is detected.** As [ATOT — lift-off], with the vertical-rate gate
inverted — descending faster than 257 ft/min — and **ALDT taken as the latest
surviving sample** of the window.

**Parameters.** Identical to [ATOT — lift-off]; the role decides the sign of
the vertical-rate gate.

**Compared with the published algorithm.** New: `r fm(n("ALDT"))` events. Its
predecessor is [landing].

**Compared with reference data.**

```{r tbl-aldt}
#| echo: false
#| results: asis
if (!is.null(lad25)) {
  cat("| Detector | Coverage | Bias | MAD | p90 | ±60 s |\n|---|---:|---:|---:|---:|---:|\n")
  for (d in c("landing", "ALDT"))
    cat(sprintf("| `%s` | %.2f%% | %+.0f s | %.0f s | %.0f s | %.2f%% |\n", d,
        as.numeric(sc("L13_shipped",d,"coverage_pct")), as.numeric(sc("L13_shipped",d,"bias_s")),
        as.numeric(sc("L13_shipped",d,"mad_s")), as.numeric(sc("L13_shipped",d,"p90_abs_s")),
        as.numeric(sc("L13_shipped",d,"within_60s_pct"))))
}
```

**An unambiguous win, unlike take-off** — better on every axis at once: forty-six
times the coverage, a median two seconds early rather than twenty-one late, and
four detections in five within a minute against fewer than two in a hundred.

Runway identity is in the table under [ATOT — lift-off]: `ALDT` names a runway
for 86.06% of arrivals and gets **66.73%** of those exactly right. **A third of
named landing runways are wrong**, which is worse than the departure case and
is the single most improvable number in the study.

**Limitations.** As [ATOT — lift-off]. The touchdown time is a proxy; the
runway identity is measured and mediocre.

## AOBT / AIBT — block times

**What it is.** Off-block and on-block — ICAO's T04 and T21, and the inputs to
KPI02 and KPI13.

**How it is detected.** The aircraft is moving once its groundspeed stays above
2 kt for a sustained 30 seconds — the signal from `traffic`'s `StartMoving`,
where the *sustained* part separates a push from a jitter in the speed field.
The start of the first such run is the off-block candidate and the end of the
last the on-block candidate.

Both are **anchored on the parking-position events**: a candidate is only
emitted for a track actually seen leaving or entering a stand. Without that
anchor a track first received mid-taxi would report its first movement as an
off-block — a different event, and one that reads as an implausibly short taxi
rather than as a miss.

`traffic`'s own parking-position path is deliberately not used: it resolves
stands through a live OpenStreetMap Overpass query whose cache does not survive
serialisation to an executor.

**Parameters.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `ground_speed_threshold_kt` | 2 kt | `traffic` | what counts as moving |
| `ground_move_min_seconds` | 30 s | `traffic` | how long movement must persist |

**Compared with the published algorithm.** New: `r fm(n("AOBT"))` AOBT and
`r fm(n("AIBT"))` AIBT events where the published algorithm emitted none.

**Compared with reference data.**

```{r tbl-blocks}
#| echo: false
#| results: asis
if (!is.null(lad25)) {
  cat("| Milestone | Coverage | Bias | MAD | p90 |\n|---|---:|---:|---:|---:|\n")
  for (d in c("AOBT", "AIBT"))
    cat(sprintf("| `%s` | %.2f%% | %+.0f s | %.0f s | %.0f s |\n", d,
        as.numeric(sc("L13_shipped",d,"coverage_pct")), as.numeric(sc("L13_shipped",d,"bias_s")),
        as.numeric(sc("L13_shipped",d,"mad_s")), as.numeric(sc("L13_shipped",d,"p90_abs_s"))))
}
```

**Coverage is the limitation and it is not an algorithm problem.** ADS-B
surface reception is sparse — an aircraft on a stand is often not received at
all. The bias is also structural rather than noise: sustained movement is
detectable only once reception begins, which is typically well after the
aircraft actually leaves the stand, so the off-block proxy fires around twelve
minutes late.

Both numbers are reported as the finding rather than tuned away, and they are
why pushback (T05) was not built: it needs strictly more coverage than this.

**Limitations.** Single-period. The bias is large enough that these milestones
are not usable for KPI02 or KPI13 as they stand — that is the honest reading.

## level-off-climb / level-off-descent

*Types:* `level-off-climb-start`, `level-off-climb-end`,
`level-off-descent-start`, `level-off-descent-end`.

**What it is.** Level flight during climb and descent — the inputs to ICAO's
KPI17 and KPI19.

**How it is detected.** ICAO specifies these with a complete algorithm rather
than leaving detection to the implementer, and this is that algorithm:

> a data point starts a level segment when the altitude difference with the next
> data point is ≤ the level band limit and the vertical speed towards it is ≤
> the vertical speed limit. The segment ends when the altitude differs from the
> segment's starting altitude by more than the level band limit, or the vertical
> speed between two consecutive points exceeds the limit.

The anchoring is what a plain window function cannot express: a segment's end is
defined against *its own start*, a running value that resets. It is implemented
as a sessionisation — mark where flatness begins, forward-fill the anchor within
the group — with each group truncated at its first failure, so a trajectory that
leaves the band and drifts back is not stitched across the excursion. Anchoring
against the segment start rather than the previous sample is what stops a slow
drift being read as one long level segment.

Segments are split into KPI17 and KPI19 by whether they fall before
top-of-climb or after top-of-descent, with two exclusions: a **minimum
altitude** below which the trajectory is not analysed, and an **exclusion box**
removing a segment above 90% of the cruise altitude lasting over five minutes —
that is cruise, not a level-off, and without it every cruise would be the
largest level-off in the flight.

**Parameters — all eight are ICAO's published values.**

| Parameter | Value | Source | Governs |
|---|---|---|---|
| `level_band_limit_ft` | 200 ft | ICAO | altitude band defining "level" |
| `level_vertical_speed_limit_ftmin` | 300 ft/min | ICAO | rate defining "level" |
| `level_min_duration_seconds` | 20 s | ICAO | shortest reportable segment |
| `level_exclusion_box_pct` | 90% | ICAO | lower edge of the cruise exclusion |
| `level_exclusion_box_seconds` | 300 s | ICAO | duration above which a high segment is cruise |
| `level_min_altitude_climb_ft` | 3,000 ft | ICAO | where climb analysis starts |
| `level_min_altitude_descent_ft` | 1,800 ft | ICAO | where descent analysis stops |
| `level_analysis_radius_nm` | 200 NM | ICAO | radius around the aerodrome analysed |

**These must not be tuned** — see [The parameter sweep, designed].

**Compared with the published algorithm.** New:
`r fm(n("level-off-climb-start"))` climb level-offs and
`r fm(n("level-off-descent-start"))` descent level-offs where the published
algorithm emitted none. Descent level-offs outnumber climb ones by roughly
`r round(n("level-off-descent-start") / max(n("level-off-climb-start"), 1))`
to one, which matches the operational expectation: arrival sequencing produces
far more level flight than departure.

**Compared with reference data. None is possible, and this is the family where
that matters most.** No source available to OPDI holds level-segment truth. The
claim made here is therefore **conformance to ICAO's published algorithm**,
tested on synthetic trajectories whose geometry is known by construction — a
climb with an injected 90-second level-off at a known altitude must yield
exactly one segment of that duration. That is checkable; "accurate" is not.

**Limitations.** Unvalidated against data, by necessity. Inherits top-of-climb
and top-of-descent from the phase detector, so a flight with no cruise phase —
a turboprop — produces no level-offs either.

## first_seen / last_seen

**What it is.** The first and last sample of each track.

**How it is detected.** Row number ascending and descending over the track,
ordered by time. Emitted for every track without condition, which is why the
count — `r fm(n("first_seen"))` on the 2025 sample — is also the track count.

**Parameters.** None.

**Compared with the published algorithm.** Unchanged:
`r delta("first_seen")`.

**Compared with reference data. None is possible, and not for want of a
source.** These are properties of *reception coverage*, not of the flight.
There is nothing they could be right or wrong against — an aircraft is first
seen when a receiver first hears it, which no reference dataset records because
it is not a property of the operation.

**Limitations.** May carry a null position where the first sample has none.

## Measurements

Two values attached to every event rather than being events themselves.

**`Distance flown (NM)`** is the cumulative great-circle distance from the start
of the track, summed sample to sample with an Earth radius of 6,371 km and
converted at 1.852 km per nautical mile. **`Time Passed (s)`** is elapsed time
from the track's first sample.

Both are **interpolated to the crossing instant** for the crossing families
rather than read from a neighbouring sample: a measurement attached to an
interpolated event but taken from a different row would disagree with its own
event.

**Compared with reference data.** APDF carries `C40_TRANSIT_DIST_KM` — the
achieved distance inside the ring — which makes a distance comparison possible
for the ring families. **It has not been run**; it is the most tractable
remaining measurement in this study.

---

# Cross-cutting findings

Two results are about the programme rather than any one event.

## Requiring complete evidence made detection worse — and it reproduces

Spark's `least`/`greatest` skip nulls, so a fuzzy rule missing an input became a
minimum over fewer terms and could out-compete a complete rule. Requiring every
input to be present is obviously more correct.

On real data it removes most phase-based landings, on **both** periods:

```{r tbl-d4}
#| echo: false
#| results: asis
if (!is.null(lad25) && !is.null(lad24)) {
  g <- function(l, r) {
    v <- l$coverage_pct[l$rung == r & l$det_type == "landing" & l$milestone == "ALDT"]
    if (length(v)) v[1] else NA
  }
  cat("| Rung | `landing` coverage 2025 | 2024 |\n|---|---:|---:|\n")
  for (r in c("L01_clean_tracks", "L02_smoothing", "L03_complete_rules"))
    cat(sprintf("| %s | %.2f%% | %.2f%% |\n", r, g(lad25, r), g(lad24, r)))
}
```

A large relative collapse on both samples, so this is a property of the change
rather than of a month. ADS-B carries a null velocity often enough that
insisting on complete evidence discards most ground samples.

It matters less than it looks, because the runway-anchored detectors replace
that path entirely — but it is a genuine negative result about a fix this study
introduced, and it is recorded rather than quietly reverted.

## The smoothing window is contradicted by the second period

Restoring the reference implementation's 60-second majority vote is the change
with the strongest prior justification in the whole study: it is what OpenAP
does, and the original port simply dropped it. It is near-neutral on 2025 and
costs 86% of landing coverage on 2024.

**One confound must be stated with it.** The 2024 tracks predate step 02's
altitude repair, so that run substitutes the raw barometric altitude. The period
differs in *input preparation* as well as in date, and the smoothing operates on
phase labels derived from exactly that column. The divergence may be the
substitution rather than the month, and this study cannot separate them.

What it can say is that the change is **not robust across the conditions
tested** — which is the actionable half: a fix that behaves this differently on
noisier input should not be adopted on the strength of its rationale alone.

---

# The parameter sweep, designed

Every threshold in this pipeline is named in `EventConfig`, which is what makes
optimisation possible. This chapter says what should be optimised, against
what, and at what cost. **Nothing here has been run.**

## The objective does not exist yet, and that is the first task

The ADEP/ADES study could score a configuration with `correct − k·wrong`
because its answer was categorical: an aerodrome is right, wrong or absent. A
milestone is a **time**, so an objective has to trade coverage against error.

The shape that fits the evidence here is *count within tolerance, minus k times
count outside it*, with:

- **the tolerance taken from the inter-source floor**, not chosen — ±10 s at
  40 NM, ±7 s at 100 NM, and for the movement times something no tighter than
  the minute the reference is rounded to;
- **k stated and swept**, as the ADEP/ADES work did at k = 1, 2, 3, because the
  exchange rate between a wrong answer and a silence is a judgement that should
  be visible rather than buried;
- **medians throughout**, since one detection landing on the wrong flight
  carries an error of hours.

Defining this is a prerequisite. A sweep without an objective produces a table
nobody can rank.

## What is worth sweeping

| Parameter | Now | Why, and what to expect |
|---|---|---|
| `phase_twindow_seconds` | 60 s | **First priority.** The change the second period contradicted. Sweep 0–120 s on *both* periods: the question is not which value is best on one sample but whether any value is robust across both. |
| `phase_require_complete_rules` | on | Measured harmful on both periods. A boolean is not a sweep — the real question is a *partial* rule, requiring two of three inputs, which needs a small code change first. Highest expected value per unit of work. |
| `runway_min_groundspeed_kt` · `runway_max_height_ft` · `runway_max_dist_nm` | 30 kt · 1,500 ft · 5 NM | These three gates decide which samples define a movement, so they trade `ATOT` coverage against its +19 s bias directly. A 3×3×3 grid is the obvious start. |
| `runway_max_bearing_deg` | 10° | Controls abstention *and* runway-identity error. Widening raises coverage and risks naming the wrong parallel — and identity is already the family's weak point at 67–72% exact. |
| `ground_speed_threshold_kt` · `ground_move_min_seconds` | 2 kt · 30 s | Where the +744 s off-block bias lives. Unlikely to fix coverage, which is a reception limit, but may move the bias. |
| `crossing_hysteresis_ft` | 300 ft | Trades chatter against crossings missed for aircraft levelling near a threshold. Cheap, and the effect is easy to reason about. |
| `phase_cruise_speed_kt` | 600 kt | Quantifies what deviating from OpenAP would buy for turboprop coverage — which is the evidence needed *before* deviating, not after. |

## What must not be swept

**The eight `level_*` parameters are ICAO's published values.** Tuning them
would forfeit the only claim that family has — conformance to a published
algorithm — in exchange for a number nothing can validate. There is no external
truth for level segments, so a "better" tuned value would mean only "different".

**`crossing_levels_fl` is the published vocabulary**, not a knob. Changing it
changes what OPDI publishes rather than how well it publishes it.

## Making it affordable

A full ladder is about two hours per period. A 3×3×3 runway grid is 27
configurations — a week of cluster at that rate, which is not a sweep anyone
will run.

The way out is to separate parameters that change **what is detected** from
those that only change **post-processing**. The runway gates, the phase
parameters and the ground thresholds require re-running detection. But the
crossing hysteresis, the sequence numbering and the interpolation operate on a
trajectory that has already been read — a harness that caches the per-sample
distance and altitude series could sweep those without touching the pipeline.
Building that cache is a prerequisite for any grid worth the name.

## The trap

**Fit on one period, validate on the other.** This study has already produced
one finding that reversed between samples. A parameter chosen on a single
period is a parameter chosen on noise, and the smoothing result is the standing
proof.

---

# What is not measured, and why

The reasons differ, and lumping them together would hide that.

**No source records it.** Top of climb and descent, the phase-derived level
segments, and the ICAO level-offs. APDF holds movement and block times, ring
crossings, a runway and a stand — nothing about the cruise phase or level
flight. For the ICAO family the substitute is conformance testing against the
published algorithm; for the others there is none.

**The source records the identity but not the timing.** Every aeroway event.
`AP_C_RWY` and `AP_C_STND` name a runway and a stand, but no surface feature has
an entry or exit time in APDF.

**There is nothing to be right about.** `first_seen` and `last_seen` describe
reception, not the flight.

**Possible but not yet run.** The ring distance comparison against
`C40_TRANSIT_DIST_KM`, and the whole of the 2024 period for every
aerodrome-dependent family — 2024 has no flight list, so it emits 16 types
against 2025's 34.

**Provenance, honestly.** Both ladder outputs read **stale** against the
current code: they were produced before the changes that made the 2024 period
runnable. Those changes cannot alter a 2025 result, but the fingerprint covers
the file rather than the behaviour, so the chain reports staleness and this page
says so rather than re-recording them as current. Clearing it means re-running
them.

---

# Breaking changes for consumers

**`first-xing-fl*` and `last-xing-fl*` are not re-emitted** by the new version.
They keep their meaning under the frozen `events_v0.0.2`, so published data
stays reproducible, but consumers moving to `events_v0.1.0` must migrate to the
crossing sequence number in each event's `info`: the first crossing is
`crossing_seq == 1`, the last is the maximum.

**`level-start`/`level-end` and `level-off-*` are not interchangeable.** They
answer different questions and only the latter corresponds to ICAO's KPI17 and
KPI19.

**The event table shrinks, which is not what anyone expects from a version that
adds four families.**

```{r tbl-growth}
#| echo: false
#| results: asis
if (!is.null(inv25)) {
  a <- sum(inv25$n_events[inv25$rung == "L00_legacy"])
  b <- sum(inv25$n_events[inv25$rung == "L13_shipped"])
  cat(sprintf("On the three-day 2025 sample: **%s events under the published algorithm, %s under the new one** — a factor of %.2f.\n",
              fm(a), fm(b), b / a))
}
```

The new families add several hundred thousand events, and the phase-flicker
removal in [level-start / level-end] takes away roughly 1.5 million — so a
consumer sizing storage from the type list alone would get the direction wrong.
**Volume is not a proxy for information here**, and any consumer with a row-count
alarm on the event table should expect it to fire.

**Event identifiers are now derived from the event's own identity** — track,
type, time and version — rather than a partition counter, so a re-run produces
the same identifiers. This makes a re-run idempotent in *content*; it does not
make the write idempotent, and re-processing safety still rests on the
per-month progress logs.

---

# Provenance

```{r tbl-provenance}
#| echo: false
#| results: asis
if (!is.null(manifest) && length(manifest) > 0) {
  cat("| Output | Script | Git SHA | Produced |\n|---|---|---|---|\n")
  for (name in sort(names(manifest))) {
    e <- manifest[[name]]
    if (is.null(e$script)) next
    dirty <- if (isTRUE(e$git_dirty)) " (dirty)" else ""
    cat(sprintf("| `%s` | `%s` | `%s`%s | %s |\n", name, e$script, e$git_sha,
                dirty, substr(e$produced_utc, 1, 10)))
  }
}
```

Every figure is produced by `benchmarks/regenerate_events.py`, which re-runs a
job only when the source files that job depends on have changed. An output with
no manifest entry is reported as unverified rather than shown as fact.
